### 2022-01-27 paper-seb-main.tex
Rebuild of project for PLOS submission

\PassOptionsToPackage{hyphens}{url}
\input plos-header.tex

\renewcommand\thesubsection{(\thesection.\alph{subsection})}
\usepackage{hyperref}

\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}

\makeatletter
    \def\do@url@hyp{\do/}
    \usepackage{moreverb,url}
    %    \renewcommand{\thefootnote}{\arabic{footnote}}
\makeatother

\input paper-seb-macros.tex

\input generated-info-for-main.tex
%\IfFileExists{generated-info-for-main.tex}{\input{generated-info-for-main.tex}}{\typeout{No generated-info-for-main.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}

\input generated-page-lengths.tex

\begin{document}

\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Harold Thimbleby
\\
\bigskip
See Change Fellow in Digital Health
\\
\bigskip

% Current address notes

% Use the asterisk to denote corresponding authorship and provide email address in note below.
\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
&\texttt{harold@thimbleby.net}\\
&ORCID 0000-0003-2222-4243
\end{tabular}

%\orcid{0000-0003-2222-4243}

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
\noindent
\emph{Background:} Computer code underpins modern science. While scientific models are routinely peer reviewed, the algorithms and the quality of code implementing them often avoid scrutiny; therefore the details of many scientific conclusions cannot be rigorously justified.

\noindent\emph{Problem:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit. Additionally, both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is therefore fallible, and if the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.

\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.

\noindent\emph{Solutions:} Code can be improved using software engineering. This paper argues for specific solutions:

\newdimen \mywidth \mywidth=\textwidth
\advance \mywidth by -1em
\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 1.\hfill}Professional software engineers can help and should be involved, particularly in critical research;

\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 2.\hfill}``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;

\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 3.\hfill}Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent software engineers.

\noindent
The \supplement\ provides a summary of professional software engineering best practice relevant to scientific research and publication.

%\end{abstract}

\vfill\begin{quote}
\hfill``Criticism is the mother of methodology.''

\hfill Robert P Abelson, \emph{Statistics as Principled Argument\/} \cite{abelson}\end{quote}
\vfill
\paragraph*{Keywords:} 
Computational science; Software Engineering; Reproducibility; Scrutiny

\newpage\section*{Author summary}

%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.
\linenumbers

%\maketitle

\section{Introduction}\label{problems}
Computational methods are comparatively new, powerful, and very flexible tools for scientists. Computational methods rely on programming, but unfortunately poor programming is very easy use to apparently get results with; such results are likely to be misleading. The role of computational methods can be compared to statistics, another useful scientific tool. Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics. Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.

{I}{n good science}, all statistics, methods and results have to be reported very carefully and in precise detail. For example, a statistical claim might be summarized as follows:

\begin{quote}\raggedright
\setbox0=\hbox{``}
\sf\hskip -\wd0\box0 %Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 ~($\sf 95\% \mbox{\sf ~CI~} 0.78, 4.27, p = 0.0009$). Between group effect size $\sf d = 0.55 ~(95\% \mbox{\sf ~CI~} 0.32, 0.77)$.'' \cite{example-stats}\end{quote}

This typical wording briefly summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize. It is a \emph{lingua franca}. It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics.

Scientists write like this --- and conferences and journals require it --- because statistical claims need to be properly accountable and documented in a clear way. Speigelhalter \cite{Speigelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Speigelhalter-section}). Results should not be uncritically accepted just because they are claimed.

The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted. Scientists work closely with competent, often specialist, statisticians who engage with the research from experiment design through to analysis and publication. Further, it is assumed that statistics will be peer reviewed, and that review will often lead to improvement.

Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement. Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians. Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.

Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them will be generated by computer. Indeed, computers are now very widely used, not just to calculate statistics, but to run the models (the term is defined below), do the data sampling and processing, and even to operate the sensors that generate the data that is analyzed. Computers are used to run human-participant surveys, such as web-based surveys. The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers. Computers even help with the word processing and typesetting of research.

In short, computers, data, and computer code are central to modern science. However, using code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more. All are non-trivial concerns requiring expertise to do reliably. Results claimed, then, however nicely they may be expressed, are no more reliable than the code and data that was used to generate them. A common oversight in scientific papers is to present a mathematical model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes.

Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims. Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.

%When relevant details of models are omitted (or are not described in recognisable ways), it is impossible to properly scrutinise claims. Describing a model used, perhaps mathematically, or providing the code (e.g., to download) that implements it is as inadequate as just providing raw data without details of the statistics analysis supporting the claim.

Just as in statistics, code and results from code need to be discussed and presented in a way that ground belief in claims derived from using them. Specifically, code should be developed in a sufficiently professional and rigorous way that is able to support clear presentation and scrutiny. Developing justifiably reliable code is the concern of the field of \emph{software engineering}, which will be discussed further below in this paper, as well as more substantially in this paper's \supplement).

To make critical claims, to inform public policy or to support further research, models need to be run under varying assumptions \cite{whitty}. Being able to understand (at least in principle) the exact code used in implementing a model is critical to having confidence in the results that rely on it. Unfortunately code is rarely considered a substantial part of the science to which it contributes.

We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.

This paper will show that unreliable computational dependencies in science are widespread and serious. Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself not contribute to furthering science, for instance through replication or reproduction.

In short, the quality of much modern science is undermined because the code it relies on is rarely of adequate quality \emph{for the uses to which it is put}. This paper will explore the extent of these problems and reasons why they persist. The paper additionally suggests some ways forward. The \supplement\ is an integral part of the suggested solutions. In particular, the \supplement\ section~\ref{supplementary-Speigelhalter-section} summarizes Speigelhater's uncontroversial statistics probes and draws explicit analogies for the critical assessment of the quality of scientific code.

\subsection{The role of code in science and scientific publication}
%Yet without code, models could not be run.

For the purposes of this paper, models map theory and parameters to describe phenomena, typically to make predictions or to test and refine the models. With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.

Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive. The chance that a thousand lines of hand-written code is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability. How to do this well is the concern of software engineering, and is discussed throughout this paper.

While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are (certainly in all cases reviewed later in this paper) they are long, complex, inscrutable and (our survey shows) lack adequate documentation. Therefore the models are very likely to be unreliable \emph{in principle}. If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if has not been clearly documented? Without documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?

Without substantial documentation it is impossible to scrutinize code properly. Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications. What does it mean? What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation? Does the code mean what was intended? What are the assumptions on \texttt{y} and do they hold invariantly? Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown. A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of overflow).

Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code. Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim. If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed. It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.

Contrary to the views expressed in the present paper, some authors have asserted that code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality}. On the contrary, if code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways. Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.

Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes. The question is whether it matters, \emph{when\/} it matters, and when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work.

%This paper explores the role of code in scientific research. Initially focusing on examples from pandemic modelling (because of its relevance to serious matters of public health) this paper shows that scientific culture, including editorial processes, have not adapted to the increasingly dominant role of code and managing the risks of its fallibility. The paper then suggests how to improve, including building mutual engagement across the science and software engineering communities.

\subsection{Bugs, code and programming}\label{knowledge}
Critiques of data and model assumptions are very common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned. Yet data and program are formally equivalent (see further discussion in \supplement, section \ref{on-code-data-publication}). Program code has as great an affect on results as the data. Code is harder to scrutinize, which means errors in code can have more subtle effects on results.

It should be noted that almost all code contains ``magic numbers'' --- that is, data masquerading as code. This common practice ensures that published data is very rarely all of the data because it omits the magic numbers. This emphasizes the need for repositories to require the inclusion of code so all data is actually available.

Bugs can be understood as discrepancies between what code is intended to do and what it actually does. Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered. Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination. None of the code examined for this paper includes either.

If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs. Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things. For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the sense of producing incorrect results --- that is, what was wanted was wrong. Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.

\iffalse Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, unqualified programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering. Dunning and Kruger go on to say,

\begin{quote}\sf\raggedright\setbox0=\hbox{``}\hskip -\wd0\copy0 People usually choose what they think is the most reasonable and optimal option \hbox{[\hskip 2pt\ldots]} The failure to recognize that one has performed poorly will instead leave one to assume that one has performed well; as a result, the incompetent will tend to grossly overestimate their skills and abilities. \hbox{[\hskip 2pt\ldots]} Not only do these people reach erroneous conclusions and make unfortunate choices, but their incompetence robs them of the metacognitive ability to realize it.''\end{quote}

Unlike many skills (skating, brain surgery, \ldots) programming, typical of much engineering, is one where errors can go unnoticed for long periods of time --- things seem to work nicely right up to the moment they fail. The worse programmers are, the more trivial bugs they tend to make, but trivial bugs are easy to find so, ironically, being a poor programmer \emph{increases\/} one's self-assessment because debugging seems very productive. It is easy for poor programmers and their associates to believe they are better than they actually are. 
%This is fertile ground for the better-than-average bias \cite{dunning-kruger}.

It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}. We \emph{all\/} make mistakes we are unaware of. If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.

{Everyone is subject to Human Factors (including the author of the present paper, e.g., as discussed in \cite{enigma}): for instance, the standard cognitive bias of confirmation bias encourages us to look for bugs when code fails to do what is expected and then debug it to produce better results, but if code generates expected results not to bother to debug it further. This of course tends to make code increasingly conform to prior expectations, whether or not those expectations are scientifically justified. Typically, there was no prior specification of the code, so the code should be right, especially after all the debugging to make it ``correct''! Thus coding routinely suffers from HARKing (Hypothesizing After the Results are Known \cite{harking}), a methodological trap widely recognized in statistics.}

Computers themselves are also a part of the problem. Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable. Programs can be written so that it is not possible to determine what they do or how they do it (whether deliberate obfuscation or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}. The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
\fi

\section{State of the art in pandemic modeling}
A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no conscious role for code or computation as such. It appears that the new insights, advances, rigor, and problems that computers bring to research are not considered relevant.

A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}. (TRIPOD guidelines ignore code completely.) The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted. It should be noted that flowcharts, which the review did consider, are programs intended for direct human use. Flowcharts, too, should be designed as carefully as code, for exactly the same reason: it is hard to program reliably.

A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} for H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling. The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.

A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,

\begin{quote}\raggedright
\setbox0=\hbox{``}
\sf\hskip -\wd0\box0 
For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
'' \cite{ferguson-interview}\end{quote}

This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times. The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022. (Typical $h$ values vary by discipline.)} of 93. Ferguson should be well aware of the standards of coding used in at least his own field. This comment, quoted above, is therefore likely to be representative of the standards of the field as a whole.

Ferguson's admission is tantamount to saying that the published scientific findings are not reproducible.\footnote{A constructive discussion of software engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.} This is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis. If this C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation. It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified over fifteen years without becoming incoherent. If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results. In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.

The programming language C which was used is not a dependable language; to develop reliable code in C requires professional tools and skills. Moreover, C code is not portable, which limits making it available for other scientists to use safely (C notoriously gets different results with different compliers, libraries, or hardware). The \supplement\ discusses these issues further.

Ferguson, author of the code, says of the criticisms of his code, ``However, none of the criticisms of the code affects the mathematics or science of the simulation'' \cite{thumbs-up}. Indeed. The problem the current paper is raising is that the theoretical epidemiology may be fine, but if it is not mapped into code that correctly implements the models, then the program's output cannot be relied on without independent evidence. In science, this is the normal requirement of \emph{reproducibility}.

The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed the code could be run again and produced comparable results (compared, apparently, to an Excel spreadsheet!). That can be achieved at this low level of sophistication is hardly surprising, regardless of the quality of the code. There was no scientific insight that merits the use of the word ``reproduction.'' If reproducibility is to be a useful scientific criterion, an \emph{independently\/} developed model needs to produce equivalent results (called $N$ version programming, a standard software engineering practice \cite{NVP}) like public health surely requires --- as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues. Meanwhile, it is a shame that using software for scientific models has enabled the bar to reproducibility, as understood by journals such as \emph{Nature}, to be lowered to a mechanical level that is only sufficient to detect some forms of dishonesty, as opposed to methodological limitations, which is the point of reproducing work.

Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021). It is
\newcount\u
\newcount\v
\newcount\w
\newcount\frac
\input generated-covidsimSummary.tex
kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done. If nothing else, as the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models. It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place. The revised code has now been published, and it has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.

Some epidemiology papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming. Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications. Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it is arguably even worse than not publishing code at all. Pseudo-code is not precise enough to help anyone scrutinize a model; copying pseudo-code introduces bugs. An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}. {The \supplement\ provides further discussion of reproducibility.}

\subsection{Science beyond epidemiology}
Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims. People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place. The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even downright flaws.

Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}. World-class software can attain maybe 1 bug per 1,000 lines of code. Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new insights into their models. In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- can be subject to numerical errors \cite{hamming}. It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion). Such bugs undermine confidence in model results.

Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what researchers have done, but merely making code available (for instance, providing it in their \supplement\ with papers, putting it in repositories, or using open source) is not sufficient. It is noteworthy that some papers disclosed that they had access to special hardware.

Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available. While some \citeeg{unfinished,lancet-unfinished} do make what they call ``documented'' code available, they provide no more than superficial comments. This is \emph{not\/} documentation as properly understood. Such comments do not explain code, explain contracts, nor explain algorithms. Some readers of the present paper may not recognize these technical software terms; contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming {(see the \supplement\ for a checklist of many relevant, professional software engineering concepts and resources)}.

If full code is made available, it may be technically ``reproducible,'' but the scientific point is to be able to understand and challenge, potentially refute, the findings; to do that, much more is required than merely being able to run the code \cite{notebooks,popper}.

Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper {and indeed in computer science research itself \cite{machine-learning-reproducibility}}) is inscrutable. Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do. Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.

Making outline code (including pseudo-code) available without proper documentation and without discussing its limitations is unethical: it encourages others to reproduce and build on poor work.

\makeatletter
\long\def\@makecaption#1#2{%
  \vskip\abovecaptionskip
  \sbox\@tempboxa{#1: #2}%
  \ifdim \wd\@tempboxa >\hsize
    \textbf{#1}: #2\par
  \else
    \global \@minipagefalse
    \hb@xt@\hsize{\hfil\box\@tempboxa\hfil}%
  \fi
  \vskip\belowcaptionskip}
\makeatother

\begin{table*}
\begin{center}
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}

%\newcount \total
%\total=\dataN
%\multiply \total by 4 % three reviewers, one editor
%\advance \total by \countAuthors
%\the\total\
%leading scientists assuming authors, 1 editor, and 3 reviewers per paper, but not counting acknowledged colleagues et al, covering  (plus appendices and supporting material) in total.
%
\caption{Overview of peer-reviewed paper sample.}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\normalsize
\input generated-summary-table.tex
%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
\end{center}

\caption{Summary of survey results.}
\label{table-summary}
\end{table*}

\begin{table*}[t]

\def\reponame#1#2{{\tt #1} \csname cite-#2\endcsname}

\expandafter\def\csname pagelength-covid-sim\endcsname{20}
\expandafter\def\csname cite-covid-sim\endcsname{\cite{ICmodel}}

% divide #1 by #2
\newcount \uu
\newcount \vv
\def\makeAverage#1#2{%
	\uu=#1
	\vv=#2
	\divide \uu by \vv
	\the\uu
}

\input generated-clone-date.tex
\begin{center}

\expandafter \advance \gitPages by \csname pagelength-covid-sim\endcsname

\input generated-repos.tex
\vskip 1ex
\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
\\
\small Repository clones downloaded and automatically summarized \clonedate
\end{center}
\newcount \pubdelayinmonths
\pubdelayinmonths=\cloneyear
\advance \pubdelayinmonths by -2020
\multiply \pubdelayinmonths by 12
\advance \pubdelayinmonths by -4 % published in April
\advance \pubdelayinmonths by \clonemonth
\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison. Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting. Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper. Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the\pubdelayinmonths\ months had already elapsed since the paper referring to the code was published.}
\label{table-repo-summary}
\end{table*}

\section{A pilot survey of peer-reviewed research relying on code}
\label{survey-section}
The problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. For instance, a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model} provides no details of its code. Its \supplement, which might have provided code, no longer exists.

Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized. We therefore undertook a survey of recent papers in leading peer reviewed journals.

\subsection{Methodology}
A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.

The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal. All papers sampled are Open Access, although for some papers some or all of the associated data has restricted access. Table \ref{table-overview} is an overview of the sample.

Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research. Commentary, correspondence, and editorials were excluded. The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.

The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.

The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.

%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).

The \the\dataN\ papers cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install. Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment was made whether the code provided actually reproduced a paper's specific claims. Indeed, if we trust the papers that their code was actually run and provides the results as reported, then merely running their code (when provided in full) merely checks the paper/code consistency, but will not assess the quality or reliability of the code.

%Assessing code quality has been a lively topic in software engineering for decades, and has international standards (ISO/IEC 9126, updated to ISO 25000). Here, code is assessed\footnote{The author of this paper has been assessing undergraduate and postgraduate computer science student code since the 1970s, using double marking and moderation.} using the n criteria of \cite{software-quality} as a marking scheme with scores out of 100/n.

The full methodology, data and analysis, as well as many software engineering and Human Factors techniques to help improve reliability and reproducibility used in the preparation of the present paper, are provided in the \supplement, which is also available from a repository (see details at the end of this paper).

%\emph{It is important that this survey is not taken as an evaluation of any specific paper: individual measurements are inevitably noisy. Instead, the aim is to form a view of scientific publishing generally. Mean scores, as summarized in table \ref{table-summary} and \ref{table-quality}, are more reliable measurements than the individual paper score breakdowns provided in the \supplement.}

\subsection{Summary of results}
The sample selection criteria necessarily identified scientific research with software engineering contributions. No evidence of verification and validation was seen. There was only one example of basic software engineering methods, namely independent coding. There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend. No competent programmer --- who writes high quality code to start with!\ --- would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc. (See the \supplement\ for definitions and a list of standard methods.)

Much code depended on substantial manual intervention to compile it. All code (where actually provided) was sufficiently complex that, if it was supposed to be used or scrutinized, required more substantial documentation than was provided.

On the whole, on the basis of the sample evidence, scientists do not make their code available, and rarely provide adequate documentation (see table \ref{table-summary}).

It is noteworthy that the papers themselves, typically a few published pages, are very small compared to the code they rely on; the papers represent much less authoring effort than the code and data that they rely on for their claims and correctness (see table \ref{table-repo-summary}). With one minor exception (which may have been accidental), no papers reported anything on any software engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).

None of the papers used any specific software engineering methods, such as open source \cite{open-source} and other standard methodologies provided in the \supplement, to help manage their processes and help improve quality. Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reason this paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}

\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies. In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach. The implication of these results is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies, or they are technically unable to assess them --- and, if so, they also fail to say they are unable to assess them. This lack of expertise is consistent with the limited awareness of software engineering best practice that is manifest in the published papers (and resources) themselves.

Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \input generated-clone-date.tex
while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could not been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.

Sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).

Overall, there was no evidence that any code had been developed carefully, {let alone by using recognized professional software engineering methods}. In particular, 
\ifnum \countCodetested=0
no papers
\else only \plural{\countCodetested}{paper}
\fi in the survey provide any claims or evidence of effective testing, for instance with evidence that tests were run on clean builds. {While it may sound unrealistic to ask for evidence on software quality in a paper written for another field of science, the need is no less than the need for standard levels of rigor in statistics reporting, as discussed in the opening of this paper.}

Data repositories (the Dryad Digital Repository, Figshare or similar) were used by \plural{\counthasDataRepository}{paper} to provide structured access to their data. Unlike GitHub, which is a general purpose repository, Dryad has scientifically-informed guidelines on handling data, and all papers that used Dryad provided more than just their raw data --- they provided a little, sometimes substantial, documentation for their data. At the time of writing, Dryad is not helpful for managing code --- its model appears to be founded on the requirement that once published papers must refer to exactly the data they used, so further refinements on the data (or code) are taboo, even with version control.

Key findings from the survey are summarized in tables \ref{table-summary} and \ref{table-repo-summary}, and are discussed in greater detail in the \supplement.

\section{Discussion}
Effective access to quality code is not routine in science. Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers. The evidence suggests that, generally, some but not all manually developed code is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no clear evidence (over the survey period) that any published code was maintained using the repositories.

The \supplement\ provides a summary of the analysis and full citations for all papers in the sample.

\subsection{A call to action}\label{summary}
Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide. Given the urgency of rigorously understanding COVID-19, any epidemic for that matter, it is essential that epidemiologists engage professional software engineers to help develop reliable laboratory methodologies. For instance, code lab books can be generated and signed easily.

Software used for informing public health policy, medical research or other medical applications is \emph{critical software}. Professional critical software development, as used in aviation and the nuclear power industry, is (put briefly) based on \emph{correct by construction}: \cite{cbc} effectively, design it right first time, {supported by numerous rigorous techniques, such as formal methods, to manage error. (See extensive discussion in this paper's \supplement.)}\ Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable. Conversely, not following these practices undermines the rigor of the science.

An analogous situation arises in ethics. There are some sorts of research that are ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work. Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems. National funders, and others, therefore require Ethics Boards to formally review ethical quality. Medical journals will not publish research that has not undergone appropriate ethical review.

Analogously, and supplementing Ethics Boards, Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality software engineering. Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to epidemic modeling papers --- and in due course, all scientific papers --- being required to include Software Engineering Board clearance statements as appropriate. {Software Engineers have a code of ethics that applies to \emph{their\/} work in epidemic modeling \cite{ethics-code}.}

{The present paper did not explore data, because in almost all cases the code and data were explained so poorly and archived so haphazardly it would be impossible to know whether the author's intentions were being followed.\footnote{{For the present paper, all the code, data, analysis and documents are available for download in a single zip file.}} Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data. Ideally, the authors should test a clean deployed build of their code and save the results. Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims. This requirement is no more unreasonable than requesting good statistics, as discussed in the opening of the paper. And the solution is the same: relevant experts --- whether statisticians or software engineers --- need to be routinely available and engaged with the science. SEBs would be a straight forward way of helping achieve this.}

There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university. Active, professional software engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art. There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.

Open Source generally improves the quality of software. SEBs will take account of the fact that open source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers. Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis {and other quality control methods for which the lead scientist and software engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).}\ {Software engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer reviewed publications \cite{acm-artifacts}.}\ {See this paper's \supplement\ for further suggestions.}

A potential argument against SEBs is that they may become onerous, onerous to run and onerous to comply with their requirements. A more mature view is that SEBs need their processes to be adaptable and proportionate. If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation. Hence SEBs processes are likely to follow a risk analysis, perhaps starting with a simple checklist. {There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar. Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.}

A professional organization, such as the UK Royal Academy of Engineering ideally working in collaboration with other national international bodies such as IFIP, should be asked to develop and support a framework for SEBs. SEBs could be quickly established to provide direct access to mature software engineering expertise for both researchers and for journals seeking competent peer reviewers. In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations. Given the urgency, this paper recommends that \emph{ad hoc\/} SEBs should be established for this purpose.

SEBs are a new suggestion, providing a supportive, collaborative process. Methodological suggestions already made in the literature include open source and specific software engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}. While \cite{ABCs-SE} provides an insightful framework to conceptualize approaches and compare their merits, there is clearly scope for much further research to provide an evidence base to motivate and assess appropriate interventions to help scientists do more rigorous and effective software engineering to support their research and publishing. These and further issues are discussed at greater length in the \supplement.

\subsection{Suggestions for further work}

Although this study of software engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope. However, the insights appear to be general.

Further work to extend the scope of the survey beyond the basic requirements of the present paper is of course worthwhile, but  the following cases (listed in the next few paragraphs) suggest that the problem is widespread. We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor.

\label{jvs-policy}
\begin{enumerate}\raggedright
\item
The journal \emph{The Lancet\/} published and then subsequently retracted a paper on using hydroxychloroquine as a treatment for COVID \cite{lancet-retracted}. The paper was found to rely on fraudulent data \cite{science-lancet1,science-lancet2}. \emph{The Lancet\/} subsequently tightened its data policies \cite{lancet-learning}, for instance to require that more than one author must have directly accessed and verified the data reported in the manuscript. Curiously, the original (now retracted) paper declares 
\begin{quote}\sf
\setbox0=\hbox{``}\hskip -\wd0\copy0 \ldots\ all authors participated in critical revision of the manuscript for important intellectual content. MRM and ANP supervised the study. All authors approved the final manuscript and were responsible for the decision to submit for publication.'' \end{quote} 
which seems to suggest that several original authors of the paper would have been happy to make the new declarations --- and, of course, if there is fraud (as was established in this case) it seems likely that authors who make the new declarations of accessing and verifying data are unlikely to make reliable declarations.

\hskip 1em\emph{The Lancet\/} still has no code publication policy (see the \supplement), and for more than one author to have ``direct access'' to the data they are very likely to access the data through the same code. If the code is faulty or fraudulent, an additional author's confirmation of the data is insufficient, and there is at least as much reason for code to be fraudulent (not least because code is much harder to scrutinize than data). Code needs more than one author to check it, and ideally reviewers independent of the authors so they do not share the same assumptions and systems (for instance shared libraries, let alone potential collusion in fraud).

\item 
In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms. Notably, the paper fails to provide access to its anonymized data (with or without qualification), and fails to define the data anonymization algorithm, and also fails to even mention the code that it developed and used to perform its study. The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see \supplement\ section \ref{supplementary-journal-policies-section}). Ironically, the retracted article \cite{jvs1} is still online (as of August 2020) with no reference to any editorial statement to the effect that it has been retracted, despite this being trivial --- and necessary --- to achieve in the widely-accessed online medium.

\item
Medical research often aims to establish a formula to define a clinical parameter (such as body mass index, BMI) or to specify an optimal drug dose or other intervention for treatment. These formulas, for which there is conventional clinical evidence, are often used as the basis for computer code that provides advice or even directly controls interventions. Unfortunately a simple formula as may be published in a medical paper is \emph{never\/} sufficient to specify code to implement it safely. For example, clinical papers do not need to evaluate or manage user error when operating apps, and therefore the statistical results of the research will be idealistic compared to the outcomes using an app under real conditions --- which is what the clinical research is supposedly for. A widespread bug (and its fix) that is often overlooked is discussed in \cite{numerals}; the paper includes an example of a popular clinical calculator (based on published clinical research) that calculated nonsense, and potentially dangerous, results. The paper \cite{fda} summarizes evidence that such bugs, ignored by the clinical research literature, are commonplace in medical systems and devices.
\end{enumerate}

\section{Conclusions}
We need to improve the quality of software engineering that supports science. While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed in shows that current coding practice makes for poor science. In the current pandemic, scientific modeling, such as track and trace \cite{excel-fiasco}, balancing COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.

The challenge to scientific research is to manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices. Computer code should be explicit, accessible (well-structured, etc), and properly documented. Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}. Professional software methodologies should not be ignored.

Unfortunately, while programming is easy, and is often taken for granted, programming \emph{well\/} is very difficult \cite{fixit}. We know from software research than ordinary programming is very buggy and unreliable. Simply put, without quality code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect. Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion:  computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper}.

Software Engineering Boards (as introduced in this paper) are a straight forward, constructive and practical way to support and improve computer-based science. This paper's \supplement\ summarizes the relevant professional software engineering practice that Software Engineering Boards would use, including discussing how and why software engineering helps improve code reliability, dependability, and quality.

\section*{Supporting information}
\newcount\csrefcount \csrefcount=0
\def\csref{\global\advance\csrefcount by 1}
\def\ethics#1{\paragraph*{Ethics}#1}
\def\ack#1{\paragraph*{Acknowledgments}#1}
\def\dataaccess#1{\paragraph*{Data and code access}#1}
\def\aucontribute#1{\paragraph*{Author contribution}#1}
\def\competing#1{\paragraph*{Competing interests}#1}
\def\funding#1{\paragraph*{Funding}#1}

\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson provided very helpful comments.}
%\the\csrefcount\ computer science reviewers.

\ethics{This article presents research with ethical considerations but does not fall not within the usual scope of ethics policies.}

\dataaccess{There is a full discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}. All material is available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards} (which has been tested in a clean build, etc). The data is encoded in JSON\@. JavaScript code, conveniently in the same file as the JSON data, checks (with \input generated-data.js-errors.tex
possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset results reliably in this paper and its \supplement\ (e.g., table \ref{table-summary} in this paper) \emph{directly\/} from the data. In addition, a standard CSV file is generated in case this is more convenient to use, for instance to browse in Excel or other spreadsheet application. All data, analysis, and details of the methods used can be found in the \supplement.}

\aucontribute{Harold Thimbleby is the sole author. An preliminary outline of this paper, containing no supplementary material or data, was submitted to the UK Parliamentary Science and Technology Select Committee's inquiry into UK Science, Research and Technology Capability and Influence in Global Disease Outbreaks, under reference LAS905222, 7 April, 2020. The evidence, which was not peer reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper. It is available to the public \cite{parliamentary-evidence,my-parliamentary-evidence}.}

\competing{The author declares no competing interests.}

\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], and by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186]. The funders had no involvement in the research or in this paper.}

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

{\raggedright
\initialiseBibliography{References}{1}{}
\bibliographystyle{plos2015}
\bibliography{paper-seb-main.bib}

% Note that the supplementary material calculates how many references there are,
% because it reads in this .tex file's .aux file so that it can cite the references 
% above with the same number sequence. 
%
% It simple redefines \bibitem to work it out while it
% defines all the citations from the bibliography above.
%
% The same technique allows the supplementary material to have 2 more bibliographies,
% with consecutive numbering.
}

\end{document}

%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}

%\def \allInOne {don't include header in supplemental material}
%input paper-seb-supplementary-material.tex

\label{LastPage}
\end{document}

### 2022-01-27 paper-seb-supplementary-material.tex
Rebuild of project for PLOS submission

\PassOptionsToPackage{hyphens}{url}
\documentclass{article}
\renewcommand\thesubsection{\thesection.\alph{subsection}}
%\usepackage{hyperref}
%\usepackage{refcount}
\usepackage{url}
\usepackage{doi}
\usepackage{longtable}
\usepackage{color}
\usepackage{amsmath,amssymb}
\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}

\input paper-seb-macros.tex

\makeatletter
      \renewcommand{\thefootnote}{\arabic{footnote}}
      {\let\oldbibcite=\bibcite
      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
      \input paper-seb-main.aux
      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
      }
\makeatother

\begin{document}

\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}

\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}

%\orcid{0000-0003-2222-4243}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

\def\href#1{#1}

\maketitle

% this will be used to save cross references that can be used in the main paper too
\newwrite\infofile
\immediate\openout\infofile=generated-info-for-main.tex

\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}

% writing to the aux file, this file generates labels like
% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
% whereas the main file using IEEEtrans class generates
% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc

\makeatletter
\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
% as well as in the current document....
\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
}
\makeatother

% start supplementary material's page counters to continue on from the paper's page numbering
%\setcounterpageref{page}{LastPage}
%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!

\setcounter{section}{0}
\def\n#1{{\sf\bfseries ({#1})}}

\newdimen \www 
\setbox0=\hbox{\n{viii}~}
\www=\wd0 \advance \www by .4em
\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}

\setbox0=\hbox{\sfdefault\large(00)}
\newdimen\howwide \howwide=\wd0 \howwide=2.5em

\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
%\begin{flushleft}
\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
%\end{flushleft}
\leavevmode
\vskip 1ex
\par\leftskip 0in}

\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
\def\mysection#1{\section{#1}\sectioncontents{#1}}
\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}

%\newpage

\makeatletter
\let \bslash = \@backslashchar
\makeatother

%\renewcommand{\thesubsection}{\alph{subsection}}
\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}

\noindent\vbox{% The table of contents
\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
\noindent\input generated-supplementary.toc
%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
}
\newwrite\contentsfile
\immediate\openout\contentsfile=generated-supplementary.toc

\newpage
\mysection{Further issues for Software Engineering Boards (SEBs)}

\highlightmysubsection{Brief definition}
Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.

Further details of the SEB proposal is in the main paper. Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:

\begin{enumerate}\raggedright
\item 
Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.

\item
When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}. Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves. Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}. Often the software engineers will be solving entirely new problems and contributing to the research. How this is handled needs exploring. How software engineers are appropriately credited and cited for their contributions also needs exploring.

\item 
SEBs require policies on professional issues such as membership, transparency, and accountability.

\item
There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).

\item
Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs. For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?

\item
In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling. More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.

\item
SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice. For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.

\item
Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large. Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
\end{enumerate}
There are other ideas to help make SEBs work, but it is clear they are part of the solution. We must not let perfection be the enemy of the good. SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.

\mysubsection{Relationships of SEBs to Ethics Boards}

\begin{enumerate}\raggedright
\item 
Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on). SEBs may also provide mentoring and training.

\item
Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software. During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.

\item 
Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious). On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own. SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is. As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.

\item
SEBs should have a two-way collaboration with Ethics Boards.

\begin{itemize}
\item SEBs have to deal with ethical concerns, and how they may be implemented in code. One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.

\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively. The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
\end{itemize}

\item
Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway. It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem. 
\end{enumerate}

\changemysubsection{SEBs are necessary but not sufficient}
The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.

SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results. As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced. In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication. SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed. It needs doing.

\mysection{Software engineering best practice}
\label{supplementary-best-practice}
\mysubsection{Introduction and standard references}

This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.

The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.

The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible. Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer. Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}. Further suggestions for background reading can be found throughout this section.

\mysubsection{Essential components of best practice}
Software Engineering includes the following topics, which are discussed at more length below:
\vskip 2ex

\vbox{\noindent\input generated-minicontents.tex
%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
}

\newwrite\minicontentsfile
\immediate\openout\minicontentsfile=generated-minicontents.tex

\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations. On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it. The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.

In practice, requirements and expectations change. Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance. Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}. In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented. This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.

\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}

{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy. To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe. Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}

{The type and use of mathematics used in software engineering is formal methods. Not using formal methods ensures the resulting code is unsafe and unreliable. Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}

Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others). Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone. Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.

Crucially, tools are designed to catch common human errors that we are all prone to. Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.

Many programming languages and programming environments have integrated features that support formal methods. For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@. In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program. Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since. Hoare is also a very good writer.

Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way. Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy). Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do. Unfortunately, because formal methods are mathematical, few programmers have experience of using them. Fortunately tools are widely available to help use formal methods very effectively.

\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc. Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}. The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code). Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.

A special case of defensive programming appropriate for pandemic modeling is mixing methods. Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.

\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
Many popular languages are popular because they are easy to use, which is not the same as being reliable to use. The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do. C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers. (This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer. SPARK Ada is one example of a much more appropriate programming language to use. SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.

\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
It is appreciated that the models may change and be adapted as new data and insights become available. Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable. Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used. Version control is supported by many tools (such as Git, Subversion, etc).

If version control is not used, one has no idea what the current program actually is. {Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository. Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}

{If results cannot be reproduced, has anything reliable been contributed? When a modeling paper presents results from a model, it is important to reproduce those results without using the same code. Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}). Reproducing the same results relying on the same codebase tells you little. The more independent reproductions of results the greater the evidence for belief in the implications.}

Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes. Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.

Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community. Open source is not a panacea, however; it raises many trade-offs. Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration. Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.

Open source raises important licensing and management questions to ensure the quality of contributions. A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository. This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.

Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate. Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.

\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them. Rigorous testing includes methods like fault injection. Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs. Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).

It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again. For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).

It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested. If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.

There are many other important testing methods \cite{sommerville,knight,NVP}.

\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).

{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories. One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
}

For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works. This is different from developer documentation, which is how to \emph{use\/} the code in other programs. For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have? In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it. How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist. An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}

There are many tools to help manage documentation (Javadoc, Doxygen, \ldots). Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}. Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.

Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature). All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work. In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.

A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless. The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.

Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project. Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.

\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
\globalLabel{supplementary-material-makefiles}
Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results? Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team. Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.

{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described. When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible. Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine. Makefiles typically also provide demo and test runs and other helpful features. Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}

\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs. The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).

Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure. Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods. Indeed, reuse, scrutiny, and consensus are the foundations of good science.

Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.

A special case of reuse is to use software tools to help with software development. The tools (if appropriately chosen) have been carefully developed and widely tested. Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.

\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs. This way lies over-fitting, a problem familiar from statistics (and machine learning). Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident. Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct. A simpler program is easier to prove correct, easier to program, and easier to debug. A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened. Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.

\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}

To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}. However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate. Publications should then cite the  standards to which their computer models comply.

Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.

\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}

As this long list illustrates, Software Engineering is a complex and wide-ranging subject. Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc). Multidisciplinary teamwork is essential.

Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs. Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand). The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.

Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.

Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software. A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be. In short, users have to be brought into the software team too.

\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}

As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving. Continuous Professional Development (CPD) is essential.

Ironically, the more organized CPD the more likely the content itself will lag behind. There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees. Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.

The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.

Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves. {In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}

\highlightsec{13}{Other factors \ldots}{Security and other factors}
Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.

For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point. Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.

{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}

\mysection{Code, data and publication}
\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}

There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes. There is no distinction one can imagine that cannot easily, even accidentally, be circumvented. In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code. It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.

{Significant cyber-vulnerabilities result from there being no difference between code and data. For example: an email arrives, which brings data to a user. The user opens an attachment, perhaps a word processor text document, which is more data. The word processor runs macros in the text document --- but now it is code. The macros move data onto the user's disc. The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files. And so on. Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}

This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}. Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.

The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.

A fragment of program code might say

\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}

This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment. The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula? Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically. The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.

Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage). You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever. Confusingly, it is generally a different function when the code is run on a different computer.

It is impossible to tell. More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data. Code can even read in formulas from data and compile them to perform further calculations, and so on. Equally, data can control the flow of code. For example, data summarizing patients may include their gender, but the program processes males and females differently. Then data starts becoming like code.

{Many computer programs blur the distinctions deliberately, to create virtual machines. Data is then run on the virtual machine as program. Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions. Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program. There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal. In short, any distinctions between code and data are impossible to maintain.}

In the present paper, we knowingly built on this blur between data and code. However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.

The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language. \LaTeX\ not only has text (as you are reading right now) but it also has code. For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted. When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters? The distinctions are all a bit moot. In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs. Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.

The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above). As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data. A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer. In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.

{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood). The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}

Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.

\changemysection{The Speigelhalter trustworhiness questions}
\globalLabel{supplementary-Speigelhalter-section}

David Speigelhalter is concerned how statistics is often misused and misunderstood. In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable. Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence. Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.

What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.

\def\questionsection#1{\changemysubsection{#1}}
\def\question#1{\item \emph{#1\/}}
\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}

\questionsection{How trustworthy are the numbers?}
\newcounter{resumeCounter}

\begin{enumerate}
\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.

\sequestion{How rigorously has the software engineering been done? Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}

\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it. Were different methods used and compared, or was all confidence put into a single implementation? What internal consistency checks does the implementation have? Were invariants and assertions defined and checked? }

\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.

\sequestion{How are the claims presented that give us confidence in the code that they are based on? Are there discussions of invariants, independent checks for errors, and so on? Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}

\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.

\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
\setcounter{resumeCounter}{\value{enumi}}
\end{enumerate}
\questionsection{How trustworthy is the source?}
\begin{enumerate}
\setcounter{enumi}{\value{resumeCounter}}
\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed. Ask yourself, `Why does this source want me to hear this story?'

\sequestion{The source of many science stories is the output of running some code. How reliable is this code? What evidence is there that the code was well-engineered so its reliability can be trusted?}

\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.

\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}

\question{What am I not being told?} This is perhaps the most important question of all. Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.

\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly. So, what evidence is there that the code was rigorously developed and cherry picking avoided?}

\setcounter{resumeCounter}{\value{enumi}}
\end{enumerate}
\questionsection{How trustworthy is the interpretation?}
\begin{enumerate}
\setcounter{enumi}{\value{resumeCounter}}
\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.

\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}

\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.

\sequestion{These are all good statistical questions. The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation? The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}

\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.

\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}

\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
\end{enumerate}

\newcount\temporary \temporary=\mainNumberOfReferences
\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
\advance \temporaryPlusOne by 1

\bibliographystyle{plos2015}

\global \newdimen \shortercol 
\renewenvironment{thebibliography}[1]{%
	\section*{\uppercase{\refname}}
    \startBibliography
    \raggedright
    \parindent=0em
    \setbox0=\hbox{\hskip 1ex[#1]}
    \labelwidth=\wd0
    \advance \labelwidth by 1ex
    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
}{}

\makeatletter    
    \renewcommand{\bibitem}[2][DEFAULT]{%
    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
        \parshape 2 0em \columnwidth \labelwidth \shortercol
        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
        \global\advance\bibciten by 1
        \ignorespaces 
    }
\makeatother

\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}

\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}

\bibliography{paper-seb-supplementary-material.bib}

\highlightmysection{Summary of pilot survey}
\label{data-methods-section}
\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown. Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research. Commentary, correspondence and editorials were excluded. The sample represents what the editorial and the broader peer review community considers to be good practice.

The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution. Indeed, all sampled papers used code in their research.  Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable. It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).

Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed. Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code. Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.

There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties. Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on). For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey. Note that comments used for explanation does not preclude any other uses.

The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted. Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey. What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized. Obviously being able to run the code will help, but clarity in documentation and comments is critical. It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''

As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.

The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):

\input generated-example-data.tex

The data was written out by hand, after reading and reviewing each paper in the survey. In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate! (A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)

A JavaScript program sanity checks the JSON data. The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors. Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.

Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file. Hence, running the data generates the core human-readable information used in this paper.

The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission. If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):

\begin{quote}
\journalBreakdown 
\end{quote}

--- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.

An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:

\begin{quote}
\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
\end{quote}

it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data. The phrase above is now in fact written in \LaTeX\ in the paper as follows:

\begin{quote}\tt\small
Code repositories were used by \\
\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
\end{quote}

where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository). When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.

Some of the files generated from the JSON data are Unix shell scripts. For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.

The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.

\highlightmysubsection{Detecting and defending against error}
\label{detecting-against-error}
It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.

% check it works for other numbers 
%\countUsesVersionControlRepository=296

Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
\newcount\t
\t=\countUsesVersionControlRepository
\divide \t by 10
\newcount\fraction
\fraction=\countUsesVersionControlRepository
\newcount\tent \tent=\t \multiply \tent by 10
\advance\fraction by -\tent
\the\t.\the\fraction,
\t=\countUsesVersionControlRepository
\advance \t by -1
\the\t,
\t=\countUsesVersionControlRepository
\advance \t by 1
\the\t, or
\t=\countUsesVersionControlRepository
\multiply \t by 10
\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.

\begin{itemize}\raggedright
\item
The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
\divide \t by 10
\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
\t=\countUsesVersionControlRepository
\multiply \t by 10
\the\t) \cite{fixit}.

\item
The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
\end{itemize}

All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository. This approach, too, considerably helps defend against common Human Factors errors. For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.

If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors. For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties. For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.

\newcount\min \min=5
\newcount\max \max=20
%\countUsesVersionControlRepository=4
\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
\ifnum \countUsesVersionControlRepository>\max
	\else 
		\ifnum \countUsesVersionControlRepository<\min
			\else
								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
		\fi
\fi

The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.

\highlightmysubsection{Code policies of sampled journals}
\globalLabel{supplementary-journal-policies-section}

It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}. In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
\\

{\sf
\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
--- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available. These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance. [\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article. Data and code should be deposited in a form that will allow maximum reuse. As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.

\vskip 3mm
{\raggedleft\scriptsize
\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
accessed 29 July 2020.\par
}

\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
--- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications. [\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.

{\vskip 3mm\scriptsize
\raggedleft
\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
accessed 29 July 2020.\par
}

\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
--- Has detailed data policies, but no code policy.
{\vskip 3mm
\raggedleft\scriptsize
\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
\\
accessed 29 July 2020.\par
}

\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
--- Has detailed data policies, but no code policy. While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:

The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it. [\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
{\vskip 3mm
\raggedleft\scriptsize
\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
\\
accessed 29 July 2020.\par
}
}

\highlightmysubsection{Assessments and criteria}

\def\flagStyle#1{\textcolor{red}{\sf #1}}

Legend:\\
\input generated-legend.tex

\newpage

{
\def\citenum#1{\cite{#1}}
\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
\input generated-assessments.tex
\end{longtable}
}

\highlightmysubsection{Summary of assessments}

\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}

\begin{center}
\input generated-summary-table.tex
\end{center}

This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience. See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.

\stepcounter{subsection}
\def\reftitle{References for sampled papers}
\def\unnumberedrefname{{\reftitle}}
\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}

\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}

\begin{thebibliography}{999}
\subsectioncontents{\unnumberedrefname}
\raggedright

% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
% figure summarising github repos, so we do it specially....

\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}

\input generated-supplementary-references.tex

\end{thebibliography}

\immediate\closeout\infofile
\immediate\closeout\contentsfile
\immediate\closeout\minicontentsfile
\end{document}

### 2022-01-27 data.js

@@ -1,0 +1,1288 @@
+// Harold Thimbleby, 2020-2021
+// JSON data and Javascript code to process it
+
+'use strict';
+
+// documentation for fields in the JSON data
+// you don't need to use all fields in the data (flags default to unset)
+// first we start with definitions of all the data fields
+var descriptionsOfFields = {
+    accessed: {
+        documentation: "Date when accessed",
+        essential: 1
+    },
+    doubleChecked: {
+        documentation: "Date when double checked",
+        essential: 1
+    },
+    authors: {
+        // surname initials comma-separated list
+        // code can be modified to generate author lists in other formats as required by journal
+        // search for function fixAuthors(s) to see what we do:
+        //     list comma-separated authors, final one with 'and' but then say et al if more than maxAuthors authors
+        documentation: "Authors",
+        essential: 1
+    },
+    codeComment: {
+        documentation: "Evaluator's notes about code",
+        essential: 1
+    },
+    dataComment: {
+        documentation: "Evaluator's notes about data",
+        essential: 1
+    },
+    doi: {
+        documentation: "DOI"
+    },
+    hasBreach: {
+        documentation: "Paper breaches journal code policy (see section~\\ref{supplementary-journal-policies-section})",
+        flag: "$\\sf P_{{\\mbox{\\scriptsize c-breach}}}$" // how a flag will be abbreviated in the paper text
+    },
+    hasCodeInPrinciple: {
+        documentation: "Paper says source code is available in principle",
+        flag: "$\\sf S_p$"
+    },
+    hasCodePolicy: { // flag value set automatically from journal name
+        documentation: "Journal has a code policy (see section~\\ref{supplementary-journal-policies-section})",
+        flag: "$\\sf P_c$"
+    },
+    hasDirectCode: {
+        documentation: "Paper or URL provides source code",
+        flag: "$\\sf S_{+}$"
+    },
+    codeURL: {
+        documentation: "URL (typically GitHub) for code repository"
+    },
+    hasDataRepo: {
+        documentation: "Paper uses a data repository (e.g., Dryad, Figshare, GitHub)",
+        flag: "$\\sf R_d$",
+        dataFlag: 1 // this flag is for the data comment (defaults to code flag)
+    },
+    hasCodeRepo: {
+        documentation: "Paper uses a code repository (e.g., GitHub)",
+        flag: "$\\sf R_c$"
+    },
[…]

### 2022-01-27 expand.js

@@ -1,0 +1,63 @@
+// Harold Thimbleby, 2020-2022
+// Javascript code to expand TeX \input commands
+
+var TeXFiles = ["paper-seb-main.tex", "paper-seb-supplementary-material.tex"];
+
+'use strict';
+
+var fs = require('fs');
+
+function saveFile(fileName, string) {
+    fs.open(fileName, 'w', function (err, fd) {
+        if (err) {
+            if (err.code === 'EEXIST') {
+                console.error(fileName + " can't be over-written");
+                return;
+            }
+            console.error("** Some error trying to write " + fileName);
+            throw err;
+        }
+        fs.writeFileSync(fd, string, function (err) {
+            if (err) {
+                return console.log("** Saving " + fileName + " got error: " + err);
+            }
+        })
+    });
+}
+
+function getFile(file) {
+    try {
+        return fs.readFileSync(file, 'utf8');
+    } catch (err) {
+        console.log("** Can't read file " + file + " - " + err);
+        return "\\error couldn't read " + file + "\n";
+    }
+}
+
+var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}");
+
+var bufferExpand, bufferInsert;
+
+for (var i = 0; i < TeXFiles.length; i++) {
+    console.log("Recursively expand all \\input and \\bibliography macro calls in " + TeXFiles[i]);
+    bufferExpand = getFile(TeXFiles[i]); // puts in buffer
+    var replaced;
+    do {
+        replaced = false;
+        bufferExpand = bufferExpand.replace(re, function (match, filename) {
+            replaced = true;
+            console.log("        \\input " + filename);
+            bufferInsert = getFile(filename);
+            return "% expand " + filename + "\n" + bufferInsert + "% end expanding " + filename + "\n";
+        });
+        bufferExpand = bufferExpand.replace(rebib, function (match, filename) {
+        	replaced = true;
+            console.log("      * \\bibliography{"+filename+".bib} replaced with " + filename+".bbl");
+            bufferInsert = getFile(filename+".bbl");
+            return "% expand bibliography " + filename + ".bbl\n" + bufferInsert + "\n% end expanding bibliography " + filename + ".bbl\n";
+        });
+    }
+    while (replaced);
+    saveFile("expanded-" + TeXFiles[i], bufferExpand);
+    console.log("  ->  Saved as: " + "expanded-" + TeXFiles[i] + "\n");
+}

### 2022-01-27 paper-seb-main.tex
Fixing final issues for PLOS

@@
 \input paper-seb-macros.tex
-
-\input generated-info-for-main.tex
-%\IfFileExists{generated-info-for-main.tex}{\input{generated-info-for-main.tex}}{\typeout{No generated-info-for-main.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
 
@@
 However, using code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
-All are non-trivial concerns requiring expertise to do reliably.
-Results claimed, then, however nicely they may be expressed, are no more reliable than the code and data that was used to generate them.
-A common oversight in scientific papers is to present a mathematical model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes.
+Software code, like statistics, is also subject to unintentional bias \cite{se-bias}.
+All these issues are non-trivial concerns requiring technical expertise to manage well.
+
+A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.
 
@@
 
-Just as in statistics, code and results from code need to be discussed and presented in a way that ground belief in claims derived from using them.
-Specifically, code should be developed in a sufficiently professional and rigorous way that is able to support clear presentation and scrutiny.
-Developing justifiably reliable code is the concern of the field of \emph{software engineering}, which will be discussed further below in this paper, as well as more substantially in this paper's \supplement).
+We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
+An interesting overlap with statistical and software engineering scrutiny is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)\@.'' yet \emph{how\/} those analyses might have been performed is not discussed.
+The problem is that admitting using SAS 9.2 or any other named system does not really help scrutiny, as such systems can do anything with the data.
+A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is reliable.
 
-To make critical claims, to inform public policy or to support further research, models need to be run under varying assumptions \cite{whitty}.
+It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements the models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases.
 Being able to understand (at least in principle) the exact code used in implementing a model is critical to having confidence in the results that rely on it.
@@
 
-We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
+This paper reviews a selection of papers in leading international journals, and finds that both papers and journal policies take code for granted.
+This paper then argues that, just as is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.
+Specifically, code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny.
+Developing justifiably reliable code is the concern of the field of \emph{software engineering}, which will be discussed further below, as well as more substantially in this paper's \supplement).
 
-This paper will show that unreliable computational dependencies in science are widespread and serious.
-Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself not contribute to furthering science, for instance through replication or reproduction.
+This paper shows that unreliable computational dependencies in science are widespread.
+Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction.
+In short, the quality of much modern science seems to be undermined because the code it relies on is rarely of adequate quality \emph{for the uses to which it is put}.
 
-In short, the quality of much modern science is undermined because the code it relies on is rarely of adequate quality \emph{for the uses to which it is put}.
-This paper will explore the extent of these problems and reasons why they persist.
+This paper explores the extent of these software engineering problems in published science.
 The paper additionally suggests some ways forward.
@@
 \initialiseBibliography{References}{1}{}
-\bibliographystyle{plos2015}
+% \bibliographystyle{plos2015} - already provided in plos-header.tex
 \bibliography{paper-seb-main.bib}

### 2022-01-27 paper-seb-supplementary-material.tex
Fixing final issues for PLOS

@@
       \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
+      \IfFileExists{paper-seb-main.aux}{%
       \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
+      }{\typeout{You will need to run Latex on paper-seb-main.tex because we need its aux file}}
       }
@@
 \def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
+\noindent
+\IfFileExists{generated-supplementary.toc}{%
+      \input generated-supplementary.toc
+      }{\typeout{You will need to run Latex again on paper-seb-supplementary.tex because we need its .toc file}}
 }
+
 \newwrite\contentsfile
@@
 
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
+\vbox{\noindent\IfFileExists{generated-minicontents.tex}{%
+      \input generated-minicontents.tex
+      }{\typeout{You will need to run Latex again on paper-seb-supplementary.tex because we need its generated-minicontents.tex file}}}
 
@@
 \immediate\closeout\minicontentsfile
+
 \end{document}

### 2022-01-27 data.js

@@ -1051,7 +1051,7 @@ s += "( echo \"% date generated by running generated-allGitRepos.sh downloading 
     "echo \\\\\\\\def\\\\\\\\clonedate {`date \"+%e %B %Y\"`}\n" +
     "echo \\\\\\\\\def\\\\\\\\cloneyear {`date \"+%Y\"`}\n" +
     "echo \\\\\\\\def\\\\\\\\clonemonth {`date \"+%m\"`}\n" +
-    ") > .. / generated-clone-date.tex\n";
+    ") > ../generated-clone-date.tex\n";
 
 s += "\n# how many errors can data.js find?\ngrep \"error(\" ../data.js|wc -l>../generated-data.js-errors.tex\n";
 

### 2022-01-27 paper-seb-main.tex
Final, as submitted to PLOS ONE

@@
 
-\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.
+\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed papers published in leading scientific journals.
 
-\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.
+\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to understand or use code reliably.
 

### 2022-01-28 data.js
Improvements to README and makefile

@@ -1025,7 +1025,7 @@ for (var i = 0; i < data.length; i++) {
     for (var f in descriptionsOfFields)
         s += csvise(data[i][f], f == lastflag);
 }
-saveFile("generated-allData.csv", s, "All data expressed in CSV");
+saveFile("CSV-generated-data.csv", s, "All data from data.js expressed in CSV for convenience");
 
 s = "";
 var spagelengths = "";

### 2022-02-28 expanded-paper-seb-main.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-main.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\input plos-header.tex
+% expand plos-header.tex
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+%\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+% end expanding plos-header.tex
 
@@
 
-\input paper-seb-macros.tex
-
-\input generated-page-lengths.tex
+% expand paper-seb-macros.tex
+% These macros allow the paper and the supplementary material to share a single sequence of bibliography citations
+
+\DeclareUrlCommand\doi{\def\UrlLeft{{\textrm{DOI}}~}}
+\DeclareUrlCommand\url{\def\UrlLeft{{\textrm{URL}}~}}
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
+\def\mytitle{Computational science: \hbox{A (fixable) failure of software~engineering}}
+%Achieving professional software engineering \hbox{in scientific research}}
+
+\def \citeeg#1{(e.g., \cite{#1})}
+
+\def\inputifexists#1#2{\IfFileExists{#1}{\input{#1}}{\typeout{No file #1. #2}}}
+
+% calculate percentages
+\newcount \t
+\newcount \tu
+\def\pc#1#2{\t = #1%
+\multiply \t by 100000%
+\divide \t by #2% ?? 0 times percentage
+\tu = \t
+\divide \t by 1000% percentage, but no idea of remainder
+\multiply \t by 1000% 1000 times percentage, with 00 as bottom digits
+\advance \tu by -\t% bottom two digits
+\divide \t by 1000\relax%
+% round to nearest integer
+% if ending in > .5 round up
+% if ending in exactly .5 round towards nearest even number
+\ifnum \tu > 500 % round up
+	\advance \t by 1 
+\else
+	\ifnum \tu = 500 
+		\ifodd \t % it's odd, so round up to even number 
+			\advance \t by 1
+		\else % leave rounded to even number 	
+		\fi
+	\fi
+\fi
+\the\t\%}
+
+% insert commas into large numbers
+\def\commarise#1{%[#1]
+	\t=#1
+    \divide \t by 1000
+    \ifnum \t > 0
+    	\the\t,%
+    	\tu=#1
+    	\multiply \t by 1000
+    	\advance \tu by -\t
+		\ifnum \tu < 100 0\fi
+		\ifnum \tu < 10 0\fi
+		\the\tu
+    \else
+    	#1%
+    \fi
+}
+
+% simple macro to use a number register to pluralise (add an s) to words....
+% \plural{\t}{fred} => \the\t\ freds if plural, or => one fred if singular
+\def\plural#1#2{\ifnum #1=1 
+	one #2%
+\else
+	\the#1\ #2s%
+\fi}
+
+% initialisBibliography <title> <starting number> <introductory text>
+\newcount\bibciten \bibciten=0
+\def\initialiseBibliography#1#2#3{%
+	\global\def\refname{#1}
+    \global\bibciten=#2
+    \global\def\startBibliography{#3}
+}
+
+\def\bibskip{\vskip 1ex} % the gap between bib items generated by data.js
+
+% read in any constants defined from the JSON data
+% expand generated-constants.tex
+\def\journalBreakdown{\emph{Lancet Digital Health\/} ($N=6$), \emph{Nature Digital Medicine\/} ($N=12$) and \emph{Royal Society Open Science\/} ($N=14$)}
+\def\tabularJournalBreakdown{&\hbox to 3em {\hfill 6}\hskip 1em \emph{Lancet Digital Health}\\&\hbox to 3em {\hfill 12}\hskip 1em \emph{Nature Digital Medicine}\\&\hbox to 3em {\hfill 14}\hskip 1em \emph{Royal Society Open Science}\\}
+\global\newcount \dataN \global\dataN=32
+\global\newcount \countAuthors \global\countAuthors=264
+\global\newcount \countHasBreach \global\countHasBreach=11
+\global\newcount \countHasPolicy \global\countHasPolicy=26
+\global\newcount \countUsesVersionControlRepository \global\countUsesVersionControlRepository=10
+\global\newcount \counthasDataRepository \global\counthasDataRepository=9
+\global\newcount \countNoCodeInRepo \global\countNoCodeInRepo=1
+\global\newcount \numberOfJournals \global\numberOfJournals=3
+\global\newcount \countCodetested \global\countCodetested=0
+\global\newcount \hasDevelopedRigorously \global\hasDevelopedRigorously=0
+% end expanding generated-constants.tex
+% expand generated-info-for-main.tex
+\newlabel{supplementary-material-makefiles}{{2}{11}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{on-code-data-publication}{{3}{14}{XXX}{YYYY}{ZZZZ}}
+\newlabel{supplementary-Speigelhalter-section}{{4}{17}{DUMMYA}{DUMMYB}{DUMMYC}}
+\def\MaxMainPaperCitationNumber{64}
+\newlabel{supplementary-journal-policies-section}{{5}{25}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementary-summary-table}{{5}{30}{DUMMY1}{DUMMY2}{DUMMY3}}
+\expandafter\def \csname cite-MetricSelectionFramework\endcsname{[97]}
+\expandafter\def \csname cite-PENet\endcsname{[101]}
+\expandafter\def \csname cite-PostoperativeOutcomes.RiskNet\endcsname{[103]}
+\expandafter\def \csname cite-philter-ucsf\endcsname{[104]}
+\expandafter\def \csname cite-AI-CDSS-Cardiovascular-Silo\endcsname{[105]}
+\expandafter\def \csname cite-SiameseChange\endcsname{[107]}
+\expandafter\def \csname cite-dryad.1g1jwstrw\endcsname{[108]}
+\expandafter\def \csname cite-rsos.192210\endcsname{[109]}
+\expandafter\def \csname cite-ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ\endcsname{[110]}
+\expandafter\def \csname cite-rsos.200566\endcsname{[111]}
+\expandafter\def \csname cite-?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c\endcsname{[112]}
+\expandafter\def \csname cite-dryad.vx0k6djnr\endcsname{[115]}
+\expandafter\def \csname cite-lactModel\endcsname{[116]}
+\expandafter\def \csname cite-LRM\endcsname{[118]}
+\expandafter\def \csname cite-1\endcsname{[119]}
+\expandafter\def \csname cite-manifold-ga\endcsname{[124]}
+\expandafter\def \csname cite-blast-ct\endcsname{[126]}
+% end expanding generated-info-for-main.tex
+
+\def\supplement{Supplemental Material}
+% end expanding paper-seb-macros.tex
+
+% expand generated-page-lengths.tex
+\expandafter\def\csname pagelength-MetricSelectionFramework\endcsname{17}
+\expandafter\def\csname pagelength-PENet\endcsname{9}
+\expandafter\def\csname pagelength-PostoperativeOutcomes.RiskNet\endcsname{10}
+\expandafter\def\csname pagelength-philter-ucsf\endcsname{8}
+\expandafter\def\csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname{6}
+\expandafter\def\csname pagelength-SiameseChange\endcsname{9}
+\expandafter\def\csname pagelength-lactModel\endcsname{13}
+\expandafter\def\csname pagelength-LRM\endcsname{22}
+\expandafter\def\csname pagelength-manifold-ga\endcsname{7}
+\expandafter\def\csname pagelength-blast-ct\endcsname{8}
+\newcount \gitPages \gitPages=109
+\newcount \totalPages \totalPages=341
+% end expanding generated-page-lengths.tex
 
@@
 
-\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.
-
-\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.
+\noindent\emph{Methodology:} This paper critiques the quality of software engineering practice (particularly as relied on in COVID-19 pandemic research driving international public health interventions), and undertakes a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
+
+\noindent\emph{Results:} Scientists rarely assure the quality of code they rely on, rarely make complete code available, and rarely provide adequate documentation to understand or use their code reliably.
 
@@
 However, using code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
-Software code, like statistics, is also subject to unintentional bias \cite{se-bias}.
+Software code, like statistics, is also subject to unintentional bias \cite{Ben,se-bias}.
 All these issues are non-trivial concerns requiring technical expertise to manage well.
@@
 \newcount\frac
-\input generated-covidsimSummary.tex
-kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
+% expand generated-covidsimSummary.tex
+25
+% end expanding generated-covidsimSummary.tex
+kLOC (thousands of lines of code) written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
 Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done.
@@
 \subsection{Science beyond epidemiology}
-Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims.
+A short 2022 summary of typical problems of software engineering impacting science generally appears in \emph{Nature} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration.
+The paper \cite{nature-review} unfortunately makes some misleading comments about the simplicity of software engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''
+
+Guest and Martin \cite{psychological-modeling} in another 2022 paper promote the use of computational modeling, arguing that through writing code, one debugs scientific thinking.
+Psychology, Guest and Martin's focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}.
+In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived.
+Computational models can even be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks.
+It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely (e.g., ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973 and is now a 120 kLOC Common LISP and Python system \cite{actr}; and of course any paper using ACT-R would require additional code on top of the basic framework).
+
+The psychology paper \cite{psychological-modeling} demonstrates building an example computational model from scratch to illustrate their own framework of computational science.
+In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not explored.
+Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.
+
+The Python code is presented without derivation or discussion, as if software engineering is trivial.
+The program listed in the paper certainly runs without error (ignoring typographical errors due to the journal's publishers), but ironically it does not implement the mathematical specification explicitly provided for it, thus unintentionally undermining the argument of the paper.
+One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses.
+Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct (at least when finished).
+Otherwise, computational modeling with inadequate software engineering will very likely introduce errors into scientific thinking.
+
+    Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims.
 People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place.
@@
 \begin{center}\normalsize
-\input generated-summary-table.tex
-%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
+% expand generated-summary-table.tex
+\begin{tabular}{|rrrc|}\hline
+Number of papers sampled relying on code&32&100\%&\\\hline\hline
+\multicolumn{4}{|l|}{\textbf{Access to code}}\\
+Have some or all code available&12&38\%&\\
+Some or all code in principle available on request&8&25\%&\\
+No code available&12&38\%&\\\hline
+\multicolumn{4}{|l|}{{\textbf{Evidence of basic good software engineering practice}}}\\
+{Evidence program designed rigorously}&{0}&{0\%}&\\
+{Evidence source code properly tested}&{0}&{0\%}&\\
+{Other methods, e.g., independent coding methods}&{1}&{3\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
+Substantial code documentation and comments&2&6\%&\\
+Comments explain some code intent&3&9\%&\\
+Procedural comments (e.g., author, date, copyright)&10&31\%&\\
+No usable comments&17&53\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Repository use}}\\
+Code repository (e.g., GitHub) --- 1 was empty&10&31\%&\\
+Data repository (e.g., Dryad or GitHub)&9&28\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
+Papers published in journals with code policies&26&81\%&\\
+Clear breaches of code policy (if any) & 11&42\%&($N=26$)\\
+\hline\end{tabular}
+% end expanding generated-summary-table.tex
 \end{center}
@@
 
-\input generated-clone-date.tex
+% expand generated-clone-date.tex
+% date generated by running generated-allGitRepos.sh downloading Git repos
+\def\clonedate {31 January 2022}
+\def\cloneyear {2022}
+\def\clonemonth {01}
+% end expanding generated-clone-date.tex
 \begin{center}
@@
 
-\input generated-repos.tex
+% expand generated-repos.tex
+\begin{tabular}{|l|r|rrr@{\hskip .75ex}l|}\hline
+&\multicolumn{1}{c|}{\bf PDF paper}&\multicolumn{4}{c|}{\bf Repository code \& data}\\ \cline{2-6}
+\multicolumn{1}{|c|}{\bf Github repository}&\multicolumn{1}{c|}{\bf Number}&\multicolumn{1}{c}{\bf Number}&\multicolumn{1}{c}{\bf Code}&\multicolumn{2}{c|}{\bf Data}\\
+\multicolumn{1}{|c|}{\bf and paper citation}&\multicolumn{1}{c|}{\bf of pages}&\multicolumn{1}{c}{\bf of files}&\multicolumn{1}{c}{\bf kLOC}&\multicolumn{2}{c|}{\bf bytes} \\ \hline\hline
+\reponame{AI-CDSS-Cardiovascular-Silo}{AI-CDSS-Cardiovascular-Silo} & \csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname & \commarise{206} & 143 & 64&Mb \\
+\reponame{blast-ct}{blast-ct} & \csname pagelength-blast-ct\endcsname & \commarise{44} & 2 & 87&Mb \\
+\reponame{covid-sim}{covid-sim} & \csname pagelength-covid-sim\endcsname & \commarise{229} & 25 & 734&Mb \\
+\reponame{lactModel}{lactModel} & \csname pagelength-lactModel\endcsname & \commarise{20} & 2 & 165&kb \\
+\reponame{LRM}{LRM} & \csname pagelength-LRM\endcsname & \commarise{125} & 8 & 2&Mb \\
+\reponame{manifold-ga}{manifold-ga} & \csname pagelength-manifold-ga\endcsname & \commarise{11} & 1 & \multicolumn{2}{c|}{---} \\
+\reponame{MetricSelectionFramework}{MetricSelectionFramework} & \csname pagelength-MetricSelectionFramework\endcsname & \commarise{44} & 4 & 236&kb \\
+\reponame{PENet}{PENet} & \csname pagelength-PENet\endcsname & \commarise{115} & 8 & 3&Mb \\
+\reponame{philter-ucsf}{philter-ucsf} & \csname pagelength-philter-ucsf\endcsname & \commarise{1987} & 13 & 32&Mb \\
+\reponame{PostoperativeOutcomes\_RiskNet}{PostoperativeOutcomes.RiskNet} & \csname pagelength-PostoperativeOutcomes.RiskNet\endcsname & \commarise{1} & --- & \multicolumn{2}{c|}{---} \\
+\reponame{SiameseChange}{SiameseChange} & \csname pagelength-SiameseChange\endcsname & \commarise{5} & 1 & 1&kb \\
+\hline \multicolumn{1}{|r|}{{\bf Average} ($N=11$)}&\makeAverage{\the\gitPages}{11}&253& 19& 84&Mb\\
+\hline \end{tabular}
+% end expanding generated-repos.tex
 \vskip 1ex
@@
 \advance \pubdelayinmonths by \clonemonth
-\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
+\caption{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
 Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting.
@@
 \subsection{Methodology}
-A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.
+A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected from the leading journals \journalBreakdown.
 
@@
 The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.
-
 The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.
 
-The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.
-
-%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).
+The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers,\footnote{Using Google Scholar it is estimated that over 40,000 papers meeting the title criteria were published in the month of July 2020.} but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice generally; (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful; and (\emph{c\/})
+the survey is not intended to be a formal, systematic sample of scientific research in general, but is intended to be sufficient to dispel the possibility that the issues described above earlier in this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they might have just happened to be software engineering outliers); (\emph{d\/}) finally, the code/data policies of the journals condoned at the time of the sample and continue to condone in 2022 poor practice (see \supplement\ section  \ref{supplementary-journal-policies-section}).
 
@@
 The sample selection criteria necessarily identified scientific research with software engineering contributions.
+
 No evidence of verification and validation was seen.
@@
 There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer --- who writes high quality code to start with!\ --- would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
+No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
 (See the \supplement\ for definitions and a list of standard methods.)
@@
 
-Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \input generated-clone-date.tex
+Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on % expand generated-clone-date.tex
+% date generated by running generated-allGitRepos.sh downloading Git repos
+\def\clonedate {31 January 2022}
+\def\cloneyear {2022}
+\def\clonemonth {01}
+% end expanding generated-clone-date.tex
 while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could not been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
@@
 The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \input generated-data.js-errors.tex
+JavaScript code, conveniently in the same file as the JSON data, checks (with % expand generated-data.js-errors.tex
+      30
+% end expanding generated-data.js-errors.tex
 possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset results reliably in this paper and its \supplement\ (e.g., table \ref{table-summary} in this paper) \emph{directly\/} from the data.
@@
 % \bibliographystyle{plos2015} - already provided in plos-header.tex
-\bibliography{paper-seb-main.bib}
+% expand bibliography paper-seb-main.bbl
+\begin{thebibliography}{10}
+
+\bibitem{abelson}
+Abelson RP.
+\newblock Statistics as Principled Argument.
+\newblock Lawrence Erlbaum Associates; 1995.
+
+\bibitem{fixit}
+Thimbleby H.
+\newblock Fix IT: How to see and solve the problems of digital healthcare.
+\newblock Oxford University Press; 2021.
+
+\bibitem{example-stats}
+Richards D, Enrique A, Eilert N, Franklin M, Palacios J, Duffy D, et~al.
+\newblock A pragmatic randomized waitlist-controlled effectiveness and
+  cost-effectiveness trial of digital interventions for depression and anxiety.
+\newblock Nature Digital Medicine. 2020;3(85).
+\newblock doi:{10.1038/s41746-020-0293-8}.
+
+\bibitem{Speigelhalter}
+Speigelhalter D.
+\newblock The Art of Statistics: Learning from Data.
+\newblock Pelican Books; 2019.
+
+\bibitem{Ben}
+Shneiderman B.
+\newblock Opinion: {The} dangers of faulty, biased, or malicious algorithms
+  requires independent oversight.
+\newblock Proceedings National Academy of Sciences. 2016;113(48):13538--13540.
+\newblock doi:{10.1073/pnas.1618211113}.
+
+\bibitem{se-bias}
+Friedman B, Nissenbaum H.
+\newblock Bias in Computer Systems.
+\newblock ACM Transactions on Information Systems. 1996;14(3):330--347.
+\newblock doi:{10.1145/230538.230561}.
+
+\bibitem{whitty}
+Whitty CJM.
+\newblock What makes an academic paper useful for health policy?
+\newblock BMC Medicine. 2015;13:301.
+\newblock doi:{10.1186/s12916-015-0544-8}.
+
+\bibitem{assurance-case}
+Habli I, Alexander R, Hawkins R, Sujan M, McDermid J, Picardi C, et~al.
+\newblock Enhancing {COVID-19} decision making by creating an assurance case
+  for epidemiological models.
+\newblock BMJ Health {\&} Care Informatics. 2020;27(e100165):1--5.
+\newblock doi:{10.1136/bmjhci-2020-100165}.
+
+\bibitem{assessing-quality}
+Kelly D, Sanders R.
+\newblock Assessing the Quality of Scientific Software.
+\newblock First International Workshop on Software Engineering For
+  Computational Science and Engineering {\cite{assessing-quality-conference}};
+  2008.
+\newblock Available from:
+  \url{http://citeseerx.ist.psu.edu/viewdoc/summary?doi=10.1.1.526.5076}.
+
+\bibitem{critiques}
+Sayburn A.
+\newblock Covid-19: {Experts} question analysis suggesting half {UK} population
+  has been infected.
+\newblock BMJ. 2020;368:m1216.
+\newblock doi:{10.1136/bmj.m1216}.
+
+\bibitem{diagnosis-reviews}
+Wynants L, Van~Calster B, Bonten MMJ, Collins GS, Debray TPA, De~Vos M, et~al.
+\newblock Prediction models for diagnosis and prognosis of covid-19 infection:
+  {Systematic} review and critical appraisal.
+\newblock BMJ. 2020;369(m1328).
+\newblock doi:{10.1136/bmj.m1328}.
+
+\bibitem{knuth}
+Knuth DE.
+\newblock The Art of Computer Programming (Seminumerical algorithms). vol.~2.
+\newblock 3rd ed. Addison-Wesley; 1998.
+
+\bibitem{science-review}
+Heesterbeek H, Anderson RM, Andreasen V, Andreasen V, Bansal S, Angelis DD,
+  et~al.
+\newblock Modeling infectious disease dynamics in the complex landscape of
+  global health.
+\newblock Science. 2015;347(6227):265--270.
+\newblock doi:{10.1126/science.aaa4339}.
+
+\bibitem{tripod}
+Moons KG, Altman DG, Reitsma JB, Ioannidis JP, Macaskill P, Steyerberg EW,
+  et~al.
+\newblock Transparent Reporting of a multivariable prediction model for
+  Individual Prognosis or Diagnosis ({TRIPOD}): {Explanation} and elaboration.
+\newblock Annals of Internal Medicine. 2015;162(1):W1--73.
+\newblock doi:{10.7326/M14-0698}.
+
+\bibitem{nature-summary}
+Adam D.
+\newblock Modelling the pandemic: {The} simulations driving the world's
+  response to {COVID}-19.
+\newblock Nature. 2020;580:316--318.
+\newblock doi:{10.1038/d41586-020-01003-6}.
+
+\bibitem{ICmodel}
+Ferguson NM, Laydon D, Nedjati-Gilani G, Imai N, Ainslie K, Baguelin M, et~al..
+  Impact of non-pharmaceutical interventions ({NPI}s) to reduce {COVID}-19
+  mortality and healthcare demand; 16 March 2020.
+\newblock Available from:
+  \url{www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf}.
+
+\bibitem{avianFluModel}
+Ferguson NM, Cummings DAT, Fraser C, Cajka JC, Cooley PC, Burke DS.
+\newblock Strategies for mitigating an influenza pandemic.
+\newblock Nature. 2005;437:209--214.
+\newblock doi:{10.1038/nature04017}.
+
+\bibitem{originalICmodel}
+Ferguson NM, Cummings DAT, Fraser C, Cajka JC, Cooley PC, Burke DS.
+\newblock Strategies for mitigating an influenza pandemic.
+\newblock Nature. 2006;442:448--452.
+\newblock doi:{10.1038/nature04795}.
+
+\bibitem{tweet}
+Ferguson N. Tweet; 22 March 2020.
+\newblock Available from:
+  \url{twitter.com/neil_ferguson/status/1241835454707699713}.
+
+\bibitem{ferguson-interview}
+Leake J.
+\newblock {Neil Ferguson} interview: {No} 10's infection guru recruits game
+  developers to build coronavirus pandemic model.
+\newblock The Sunday Times. 29 March 2020;.
+
+\bibitem{basic-reproducibilty}
+Hinsen K.
+\newblock Software Development for Reproducible Research.
+\newblock Computing in Science {\&} Engineering. 2013;15(4):60--63.
+\newblock doi:{10.1109/MCSE.2013.91}.
+
+\bibitem{thumbs-up}
+Chawla DS.
+\newblock Critiqued coronavirus simulation gets thumbs up from code-checking
+  efforts.
+\newblock Nature. 8 June 2020;582:323--324.
+
+\bibitem{codecheck}
+Scheuber A, {van Elsland} SL. Codecheck confirms reproducibility of {COVID-19}
+  model results; 1 June 2020.
+\newblock Available from:
+  \url{www.imperial.ac.uk/news/197875/codecheck-confirms-reproducibility-covid-19-model-results}.
+
+\bibitem{NVP}
+{Hatton} L, {Roberts} A.
+\newblock How accurate is scientific software?
+\newblock IEEE Transactions on Software Engineering. 1994;20(10):785--797.
+\newblock doi:{10.1109/32.328993}.
+
+\bibitem{nvp-ferguson}
+Halloran ME, Ferguson NM, Eubank S, Longini IM, Cummings DAT, Lewis B, et~al.
+\newblock Modeling targeted layered containment of an influenza pandemic in the
+  United States.
+\newblock Proceedings of the National Academy of Sciences.
+  2008;105(12):4639--4644.
+\newblock doi:{10.1073/pnas.0706849105}.
+
+\bibitem{refactoring}
+Ferguson N. Tweet; 22 March 2020.
+\newblock Available from:
+  \url{twitter.com/neil_ferguson/status/1241835456947519492}.
+
+\bibitem{bad-code}
+Richards D, Boudnik K.
+\newblock {Neil} {Ferguson's} {Imperial} model could be the most devastating
+  software mistake of all time.
+\newblock The Telegraph. 16 May 2020;{\hskip -.25em}.
+
+\bibitem{pseudo}
+Zlojutro A, Rey D, Gardner L.
+\newblock A decision-support framework to optimize border control for global
+  outbreak mitigation.
+\newblock Nature Scientific Reports. 2019;9(2216).
+\newblock doi:{10.1038/s41598-019-38665-w}.
+
+\bibitem{relit}
+Thimbleby H, Williams D.
+\newblock A tool for publishing reproducible algorithms {\&} {A} reproducible,
+  elegant algorithm for sequential experiments.
+\newblock Science of Computer Programming. 2018;156:45--67.
+\newblock doi:{10.1016/j.scico.2017.12.010}.
+
+\bibitem{nature-review}
+Perkel JM.
+\newblock How to fix your scientific coding errors.
+\newblock Nature. 2022;602:172--173.
+\newblock doi:{10.1038/d41586-022-00217-0}.
+
+\bibitem{psychological-modeling}
+Guest O, Martin AE.
+\newblock How Computational Modeling Can Force Theory Building in Psychological
+  Science.
+\newblock Perspectives on Psychological Science. 2021;16(4):789--802.
+\newblock doi:{10.1177/1745691620970585}.
+
+\bibitem{actr}
+{ACT-R Research Group}. ACT-R; 2022.
+\newblock Available from: \url{act-r.psy.cmu.edu/software}.
+
+\bibitem{ourReview}
+Ladkin PB, Littlewood B, Thimbleby H, Thomas M.
+\newblock The {Law Commission} presumption concerning the dependability of
+  computer evidence.
+\newblock Digital Evidence and Electronic Signature Law Review. 2020;17.
+\newblock doi:{10.14296/deeslr.v17i0.5143 (NB See
+  \url{journals.sas.ac.uk/deeslr/article/view/5143} until the DOI is
+  resolved.)}.
+
+\bibitem{hamming}
+Hamming RW.
+\newblock Numerical Methods for Scientists and Engineers.
+\newblock Dover Publications Inc.; 1987.
+
+\bibitem{unfinished}
+Kissler SM, Tedijanto C, Goldstein E, Kissler SM, Tedijanto C, Goldstein E,
+  et~al.
+\newblock Projecting the transmission dynamics of {SARS-CoV-2} through the
+  postpandemic period.
+\newblock Science. 2020;doi:{10.1126/science.abb5793}.
+
+\bibitem{lancet-unfinished}
+Verity, R, Okell, C L, Dorigatti, Winskill IP, et~al.
+\newblock Estimates of the severity of coronavirus disease 2019: {A}
+  model-based analysis.
+\newblock Lancet. 2020;doi:{10.1016/S1473-3099(20)30243-7}.
+
+\bibitem{hoare}
+Hoare CAR.
+\newblock An axiomatic basis for computer programming.
+\newblock Communications of the ACM. 1969;12(10):576--580.
+\newblock doi:{10.1145/363235.363259}.
+
+\bibitem{notebooks}
+Schnell S.
+\newblock Ten Simple Rules for a Computational Biologist's Laboratory Notebook.
+\newblock PLoS Computational Biology. 2015;11(9):e1004385.
+\newblock doi:{10.1371/journal.pcbi.1004385}.
+
+\bibitem{popper}
+Popper KR.
+\newblock Conjectures and Refutations: {The} Growth of Scientific Knowledge.
+\newblock 2nd ed. Routledge; 2002.
+
+\bibitem{machine-learning-reproducibility}
+Thimbleby H.
+\newblock Give your computer's {IQ} a boost --- {{\emph{{Journal of Machine
+  Learning Research}}}}.
+\newblock Times Higher Education Supplement. 9 May, 2004;.
+
+\bibitem{flu-model}
+Sander B, Nizam A, {Garrison Jr} LP, Postma MJ, Halloran ME, {Longini Jr } IM.
+\newblock Economic evaluation of influenza pandemic mitigation strategies in
+  the us using a stochastic microsimulation transmission model.
+\newblock Value Health. 2009;12(2):226--233.
+\newblock doi:{10.1111/j.1524-4733.2008.00437.x}.
+
+\bibitem{open-source}
+Fomel S.
+\newblock Reproducible Research as a Community Effort: Lessons from the
+  {Madagascar Project}.
+\newblock Computing in Science {\&} Engineering. 2015;17(1):20--26.
+\newblock doi:{10.1109/MCSE.2014.94}.
+
+\bibitem{stability}
+Salama M, Bahsoon R, Lago P.
+\newblock Stability in Software Engineering: Survey of the State-of-the-Art and
+  Research Directions.
+\newblock IEEE Transactions on Software Engineering. 2021;47(7):1468--1510.
+\newblock doi:{10.1109/TSE.2019.2925616}.
+
+\bibitem{cbc}
+Woodcock JCP, Larsen PG, Bicarregui JC, Fitzgerald JS.
+\newblock Formal methods: {Practice} and experience.
+\newblock ACM Computing Surveys. 2009;41(4).
+\newblock doi:{10.1145/1592434.1592436}.
+
+\bibitem{ethics-code}
+ACM.
+\newblock Code of Ethics and Professional Conduct.
+\newblock ACM; 2020.
+\newblock Available from: \url{www.acm.org/code-of-ethics}.
+
+\bibitem{acm-artifacts}
+ACM.
+\newblock Artifact Review and Badging --- Current. vol. Artifact Review and
+  Badging Version 1.1.
+\newblock ACM; 2020.
+\newblock Available from:
+  \url{www.acm.org/publications/policies/artifact-review-and-badging-current}.
+
+\bibitem{redmill}
+Redmill F.
+\newblock Understanding the Use, Misuse and Abuse of Safety Integrity Levels.
+\newblock In: Lessons in System Safety, Eighth Safety-critical Systems
+  Symposium; 2000.Available from:
+  \url{homepages.cs.ncl.ac.uk/felix.redmill/publications/1%20SILs.pdf}.
+
+\bibitem{iec61508}
+IEC 61508:2010 CMV Commented version, Functional safety of
+  electrical/electronic/programmable electronic safety-related systems; 2010.
+\newblock Available from: \url{webstore.iec.ch/publication/22273}.
+
+\bibitem{ABCs-SE}
+Stol KJ, Fitzgerald B.
+\newblock The ABC of Software Engineering Research.
+\newblock ACM Transactions on Software Engingeering and Methodology.
+  2018;27(3).
+\newblock doi:{10.1145/3241743}.
+
+\bibitem{lancet-retracted}
+Mehra MR, Desai SS, Ruschitzka F, Patel AN.
+\newblock {RETRACTED}: {Hydroxychloroquine} or chloroquine with or without a
+  macrolide for treatment of {COVID}-19: {A} multinational registry analysis.
+\newblock The Lancet. 2020; p. 1--10.
+\newblock doi:{10.1016/S0140-6736(20)31180-6}.
+
+\bibitem{science-lancet1}
+Servick K, Enserink M.
+\newblock A mysterious company's coronavirus papers in top medical journals may
+  be unraveling.
+\newblock Science. 2020;doi:{10.1126/science.abd1337}.
+
+\bibitem{science-lancet2}
+Servick K.
+\newblock {COVID}-19 data scandal prompts tweaks to elite journal's review
+  process.
+\newblock Science. 2020;doi:{10.1126/science.abe8656}.
+
+\bibitem{lancet-learning}
+{The Editors}.
+\newblock Learning from a retraction.
+\newblock The Lancet. 2020;396:799.
+\newblock doi:{10.1016/S0140-6736(20)31958-9}.
+
+\bibitem{jvs1}
+Hardouin S, Cheng TW, Mitchell EL, Raulli SJ, Jones DW, Siracuse JJ, et~al.
+\newblock Prevalence of unprofessional social media content among young
+  vascular surgeons.
+\newblock Journal of Vascular Surgery. 2020;72(2):667--671.
+\newblock doi:{10.1016/j.jvs.2019.10.069}.
+
+\bibitem{jvs2}
+Baumann J.
+\newblock \#{MedBikini} Backlash Exposes Research Ethics Boards' Digital Gaps.
+\newblock Bloomberg Law. 29 July, 2020;.
+
+\bibitem{jvs3}
+{The Editors {(of \emph{Journal of Vascular Surgery\/})}}.
+\newblock {Editors' Statement Regarding ``Prevalence of unprofessional social
+  media content among young vascular surgeons''}.
+\newblock FaceBook. 2020;.
+
+\bibitem{numerals}
+Thimbleby H, Cairns P.
+\newblock Interactive numerals.
+\newblock Royal Society Open Science. 2017;4(4):160903.
+\newblock doi:{10.1098/rsos.160903}.
+
+\bibitem{fda}
+Zhang Y, Masci P, Jones P, Thimbleby H.
+\newblock User Interface Software Errors in Medical Devices.
+\newblock Biomedical Instrumentation {\&} Technology. 2019;53(3):182--194.
+\newblock doi:{10.2345/0899-8205-53.3.182}.
+
+\bibitem{excel-fiasco}
+Thimbleby H.
+\newblock The problem isn't {Excel}, it's unprofessional software engineering.
+\newblock BMJ. 2020;371(m4181).
+\newblock doi:{10.1136/bmj.m4181}.
+
+\bibitem{science-delays}
+Wadman M.
+\newblock Could too much time between doses drive the coronavirus to outwit
+  vaccines?
+\newblock Science. 2021;doi:{10.1126/science.abg5655}.
+
+\bibitem{reproducibility}
+Benureau FCY, Rougier NP.
+\newblock Re-run, Repeat, Reproduce, Reuse, Replicate: {Transforming} Code into
+  Scientific.
+\newblock Frontiers in Neuroinformatics. 2017;11(69).
+\newblock doi:{10.3389/fninf.2017.00069}.
+
+\bibitem{parliamentary-evidence}
+{House of Commons Science and Technology Committee}. The {UK} response to
+  covid-19: {Use} of scientific advice; 16 December 2020.
+\newblock Available from:
+  \url{committees.parliament.uk/publications/4165/documents/41300/default}.
+
+\bibitem{my-parliamentary-evidence}
+Thimbleby H.
+\newblock Written Evidence Submitted by {Harold Thimbleby} to The {UK} response
+  to covid-19: {Use} of scientific advice.
+\newblock C190005. {House of Commons Science and Technology Committee}
+  {\cite{parliamentary-evidence}}; 29 April 2020.
+\newblock Available from:
+  \url{committees.parliament.uk/work/91/default/publications/written-evidence/?SearchTerm=thimbleby}.
+
+\bibitem{assessing-quality-conference}
+Carver JC.
+\newblock First International Workshop on Software Engineering for
+  Computational Science {\&} Engineering.
+\newblock Computing in Science {\&} Engineering. 2009;11(2):7--11.
+\newblock doi:{10.1109/MCSE.2009.30}.
+
+\end{thebibliography}
+
+% end expanding bibliography paper-seb-main.bbl
 

### 2022-02-28 expanded-paper-seb-supplementary-material.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
 
-\input paper-seb-macros.tex
+% expand paper-seb-macros.tex
+% These macros allow the paper and the supplementary material to share a single sequence of bibliography citations
+
+\DeclareUrlCommand\doi{\def\UrlLeft{{\textrm{DOI}}~}}
+\DeclareUrlCommand\url{\def\UrlLeft{{\textrm{URL}}~}}
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
+\def\mytitle{Computational science: \hbox{A (fixable) failure of software~engineering}}
+%Achieving professional software engineering \hbox{in scientific research}}
+
+\def \citeeg#1{(e.g., \cite{#1})}
+
+\def\inputifexists#1#2{\IfFileExists{#1}{\input{#1}}{\typeout{No file #1. #2}}}
+
+% calculate percentages
+\newcount \t
+\newcount \tu
+\def\pc#1#2{\t = #1%
+\multiply \t by 100000%
+\divide \t by #2% ?? 0 times percentage
+\tu = \t
+\divide \t by 1000% percentage, but no idea of remainder
+\multiply \t by 1000% 1000 times percentage, with 00 as bottom digits
+\advance \tu by -\t% bottom two digits
+\divide \t by 1000\relax%
+% round to nearest integer
+% if ending in > .5 round up
+% if ending in exactly .5 round towards nearest even number
+\ifnum \tu > 500 % round up
+	\advance \t by 1 
+\else
+	\ifnum \tu = 500 
+		\ifodd \t % it's odd, so round up to even number 
+			\advance \t by 1
+		\else % leave rounded to even number 	
+		\fi
+	\fi
+\fi
+\the\t\%}
+
+% insert commas into large numbers
+\def\commarise#1{%[#1]
+	\t=#1
+    \divide \t by 1000
+    \ifnum \t > 0
+    	\the\t,%
+    	\tu=#1
+    	\multiply \t by 1000
+    	\advance \tu by -\t
+		\ifnum \tu < 100 0\fi
+		\ifnum \tu < 10 0\fi
+		\the\tu
+    \else
+    	#1%
+    \fi
+}
+
+% simple macro to use a number register to pluralise (add an s) to words....
+% \plural{\t}{fred} => \the\t\ freds if plural, or => one fred if singular
+\def\plural#1#2{\ifnum #1=1 
+	one #2%
+\else
+	\the#1\ #2s%
+\fi}
+
+% initialisBibliography <title> <starting number> <introductory text>
+\newcount\bibciten \bibciten=0
+\def\initialiseBibliography#1#2#3{%
+	\global\def\refname{#1}
+    \global\bibciten=#2
+    \global\def\startBibliography{#3}
+}
+
+\def\bibskip{\vskip 1ex} % the gap between bib items generated by data.js
+
+% read in any constants defined from the JSON data
+% expand generated-constants.tex
+\def\journalBreakdown{\emph{Lancet Digital Health\/} ($N=6$), \emph{Nature Digital Medicine\/} ($N=12$) and \emph{Royal Society Open Science\/} ($N=14$)}
+\def\tabularJournalBreakdown{&\hbox to 3em {\hfill 6}\hskip 1em \emph{Lancet Digital Health}\\&\hbox to 3em {\hfill 12}\hskip 1em \emph{Nature Digital Medicine}\\&\hbox to 3em {\hfill 14}\hskip 1em \emph{Royal Society Open Science}\\}
+\global\newcount \dataN \global\dataN=32
+\global\newcount \countAuthors \global\countAuthors=264
+\global\newcount \countHasBreach \global\countHasBreach=11
+\global\newcount \countHasPolicy \global\countHasPolicy=26
+\global\newcount \countUsesVersionControlRepository \global\countUsesVersionControlRepository=10
+\global\newcount \counthasDataRepository \global\counthasDataRepository=9
+\global\newcount \countNoCodeInRepo \global\countNoCodeInRepo=1
+\global\newcount \numberOfJournals \global\numberOfJournals=3
+\global\newcount \countCodetested \global\countCodetested=0
+\global\newcount \hasDevelopedRigorously \global\hasDevelopedRigorously=0
+% end expanding generated-constants.tex
+% expand generated-info-for-main.tex
+\newlabel{supplementary-material-makefiles}{{2}{11}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{on-code-data-publication}{{3}{14}{XXX}{YYYY}{ZZZZ}}
+\newlabel{supplementary-Speigelhalter-section}{{4}{17}{DUMMYA}{DUMMYB}{DUMMYC}}
+\def\MaxMainPaperCitationNumber{64}
+\newlabel{supplementary-journal-policies-section}{{5}{25}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementary-summary-table}{{5}{30}{DUMMY1}{DUMMY2}{DUMMY3}}
+\expandafter\def \csname cite-MetricSelectionFramework\endcsname{[97]}
+\expandafter\def \csname cite-PENet\endcsname{[101]}
+\expandafter\def \csname cite-PostoperativeOutcomes.RiskNet\endcsname{[103]}
+\expandafter\def \csname cite-philter-ucsf\endcsname{[104]}
+\expandafter\def \csname cite-AI-CDSS-Cardiovascular-Silo\endcsname{[105]}
+\expandafter\def \csname cite-SiameseChange\endcsname{[107]}
+\expandafter\def \csname cite-dryad.1g1jwstrw\endcsname{[108]}
+\expandafter\def \csname cite-rsos.192210\endcsname{[109]}
+\expandafter\def \csname cite-ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ\endcsname{[110]}
+\expandafter\def \csname cite-rsos.200566\endcsname{[111]}
+\expandafter\def \csname cite-?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c\endcsname{[112]}
+\expandafter\def \csname cite-dryad.vx0k6djnr\endcsname{[115]}
+\expandafter\def \csname cite-lactModel\endcsname{[116]}
+\expandafter\def \csname cite-LRM\endcsname{[118]}
+\expandafter\def \csname cite-1\endcsname{[119]}
+\expandafter\def \csname cite-manifold-ga\endcsname{[124]}
+\expandafter\def \csname cite-blast-ct\endcsname{[126]}
+% end expanding generated-info-for-main.tex
+
+\def\supplement{Supplemental Material}
+% end expanding paper-seb-macros.tex
 
@@
       {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
+      % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
       \newcount\mainNumberOfReferences \mainNumberOfReferences=0
       \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
+      % expand paper-seb-main.aux
+\relax 
+\providecommand\hyper@newdestlabel[2]{}
+\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
+\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
+\global\let\oldcontentsline\contentsline
+\gdef\contentsline#1#2#3#4{\oldcontentsline{#1}{#2}{#3}}
+\global\let\oldnewlabel\newlabel
+\gdef\newlabel#1#2{\newlabelxx{#1}#2}
+\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
+\AtEndDocument{\ifx\hyper@anchor\@undefined
+\let\contentsline\oldcontentsline
+\let\newlabel\oldnewlabel
+\fi}
+\fi}
+\global\let\hyper@last\relax 
+\gdef\HyperFirstAtBeginDocument#1{#1}
+\providecommand*\HyPL@Entry[1]{}
+\bibstyle{plos2015}
+\citation{abelson}
+\HyPL@Entry{0<</S/D>>}
+\citation{fixit}
+\citation{example-stats}
+\citation{Speigelhalter}
+\@writefile{toc}{\contentsline {section}{\numberline {1}Introduction}{2}{section.1}\protected@file@percent }
+\newlabel{problems}{{1}{2}{Introduction}{section.1}{}}
+\citation{Ben,se-bias}
+\citation{whitty}
+\citation{assurance-case}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(1.a)}The role of code in science and scientific publication}{4}{subsection.1.1}\protected@file@percent }
+\citation{assessing-quality}
+\citation{critiques,diagnosis-reviews}
+\citation{knuth}
+\citation{science-review}
+\citation{diagnosis-reviews}
+\citation{tripod}
+\citation{diagnosis-reviews}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(1.b)}Bugs, code and programming}{5}{subsection.1.2}\protected@file@percent }
+\newlabel{knowledge}{{(1.b)}{5}{Bugs, code and programming}{subsection.1.2}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {2}State of the art in pandemic modeling}{5}{section.2}\protected@file@percent }
+\citation{nature-summary,ICmodel}
+\citation{avianFluModel,originalICmodel}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{tweet}
+\citation{ferguson-interview}
+\citation{basic-reproducibilty}
+\citation{thumbs-up}
+\citation{nature-summary,ICmodel}
+\citation{codecheck,thumbs-up}
+\citation{NVP}
+\citation{nvp-ferguson}
+\citation{refactoring}
+\citation{bad-code}
+\citation{pseudo}
+\citation{relit}
+\citation{nature-review}
+\citation{nature-review}
+\citation{psychological-modeling}
+\citation{psychological-modeling}
+\citation{actr}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(2.a)}Science beyond epidemiology}{7}{subsection.2.1}\protected@file@percent }
+\citation{psychological-modeling}
+\citation{Ben,nature-review,se-bias}
+\citation{ourReview}
+\citation{hamming}
+\citation{NVP}
+\citation{unfinished}
+\citation{unfinished,lancet-unfinished}
+\citation{hoare}
+\citation{notebooks,popper}
+\citation{machine-learning-reproducibility}
+\citation{relit}
+\citation{ICmodel}
+\citation{ICmodel}
+\citation{ICmodel}
+\citation{flu-model}
+\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Overview of peer-reviewed paper sample.\relax }}{9}{table.caption.4}\protected@file@percent }
+\providecommand*\caption@xref[2]{\@setref\relax\@undefined{#1}}
+\newlabel{table-overview}{{1}{9}{Overview of peer-reviewed paper sample.\relax }{table.caption.4}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {3}A pilot survey of peer-reviewed research relying on code}{9}{section.3}\protected@file@percent }
+\newlabel{survey-section}{{3}{9}{A pilot survey of peer-reviewed research relying on code}{section.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(3.a)}Methodology}{9}{subsection.3.1}\protected@file@percent }
+\@writefile{lot}{\contentsline {table}{\numberline {2}{\ignorespaces Summary of survey results.\relax }}{10}{table.caption.5}\protected@file@percent }
+\newlabel{table-summary}{{2}{10}{Summary of survey results.\relax }{table.caption.5}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {3}{\ignorespaces Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt  {covid-sim} \cite  {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the Supplemental Material), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+Paper [103]\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, 21\ months had already elapsed since the paper referring to the code was published.\relax }}{11}{table.caption.6}\protected@file@percent }
+\newlabel{table-repo-summary}{{3}{11}{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt {covid-sim} \cite {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement ), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname \ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the \pubdelayinmonths \ months had already elapsed since the paper referring to the code was published.\relax }{table.caption.6}{}}
+\citation{open-source}
+\citation{stability}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(3.b)}Summary of results}{12}{subsection.3.2}\protected@file@percent }
+\citation{notebooks}
+\citation{cbc}
+\@writefile{toc}{\contentsline {section}{\numberline {4}Discussion}{13}{section.4}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {(4.a)}A call to action}{13}{subsection.4.1}\protected@file@percent }
+\newlabel{summary}{{(4.a)}{13}{A call to action}{subsection.4.1}{}}
+\citation{ethics-code}
+\citation{acm-artifacts}
+\citation{redmill,iec61508}
+\citation{basic-reproducibilty,open-source}
+\citation{ABCs-SE}
+\citation{lancet-retracted}
+\citation{science-lancet1,science-lancet2}
+\citation{lancet-learning}
+\@writefile{toc}{\contentsline {subsection}{\numberline {(4.b)}Suggestions for further work}{15}{subsection.4.2}\protected@file@percent }
+\newlabel{jvs-policy}{{(4.b)}{15}{Suggestions for further work}{subsection.4.2}{}}
+\citation{jvs1}
+\citation{jvs2,jvs3}
+\citation{jvs1}
+\citation{numerals}
+\citation{fda}
+\citation{tweet,ferguson-interview}
+\citation{excel-fiasco}
+\citation{science-delays}
+\citation{whitty}
+\@writefile{toc}{\contentsline {section}{\numberline {5}Conclusions}{16}{section.5}\protected@file@percent }
+\citation{fixit}
+\citation{reproducibility,relit,popper}
+\citation{parliamentary-evidence,my-parliamentary-evidence}
+\bibdata{paper-seb-main.bib}
+\bibcite{abelson}{1}
+\bibcite{fixit}{2}
+\bibcite{example-stats}{3}
+\bibcite{Speigelhalter}{4}
+\bibcite{Ben}{5}
+\bibcite{se-bias}{6}
+\bibcite{whitty}{7}
+\bibcite{assurance-case}{8}
+\bibcite{assessing-quality}{9}
+\citation{assessing-quality-conference}
+\bibcite{critiques}{10}
+\bibcite{diagnosis-reviews}{11}
+\bibcite{knuth}{12}
+\bibcite{science-review}{13}
+\bibcite{tripod}{14}
+\bibcite{nature-summary}{15}
+\bibcite{ICmodel}{16}
+\bibcite{avianFluModel}{17}
+\bibcite{originalICmodel}{18}
+\bibcite{tweet}{19}
+\bibcite{ferguson-interview}{20}
+\bibcite{basic-reproducibilty}{21}
+\bibcite{thumbs-up}{22}
+\bibcite{codecheck}{23}
+\bibcite{NVP}{24}
+\bibcite{nvp-ferguson}{25}
+\bibcite{refactoring}{26}
+\bibcite{bad-code}{27}
+\bibcite{pseudo}{28}
+\bibcite{relit}{29}
+\bibcite{nature-review}{30}
+\bibcite{psychological-modeling}{31}
+\bibcite{actr}{32}
+\bibcite{ourReview}{33}
+\bibcite{hamming}{34}
+\bibcite{unfinished}{35}
+\bibcite{lancet-unfinished}{36}
+\bibcite{hoare}{37}
+\bibcite{notebooks}{38}
+\bibcite{popper}{39}
+\bibcite{machine-learning-reproducibility}{40}
+\bibcite{flu-model}{41}
+\bibcite{open-source}{42}
+\bibcite{stability}{43}
+\bibcite{cbc}{44}
+\bibcite{ethics-code}{45}
+\bibcite{acm-artifacts}{46}
+\bibcite{redmill}{47}
+\bibcite{iec61508}{48}
+\bibcite{ABCs-SE}{49}
+\bibcite{lancet-retracted}{50}
+\bibcite{science-lancet1}{51}
+\bibcite{science-lancet2}{52}
+\bibcite{lancet-learning}{53}
+\bibcite{jvs1}{54}
+\bibcite{jvs2}{55}
+\bibcite{jvs3}{56}
+\bibcite{numerals}{57}
+\bibcite{fda}{58}
+\bibcite{excel-fiasco}{59}
+\bibcite{science-delays}{60}
+\bibcite{reproducibility}{61}
+\bibcite{parliamentary-evidence}{62}
+\bibcite{my-parliamentary-evidence}{63}
+\citation{parliamentary-evidence}
+\bibcite{assessing-quality-conference}{64}
+\newlabel{LastPage}{{}{22}{}{page.22}{}}
+\xdef\lastpage@lastpage{22}
+\xdef\lastpage@lastpageHy{22}
+\gdef \@abspage@last{22}
+% end expanding paper-seb-main.aux
       }
@@
 
-%\newpage
-
 \makeatletter
@@
 
-%\renewcommand{\thesubsection}{\alph{subsection}}
 \def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
@@
 \def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
+\noindent
+% expand generated-supplementary.toc
+\contents{1 Further issues for Software Engineering Boards (SEBs)}{3}
+\subcontents{a\ {Brief definition}}{3}
+\subcontents{b\ Relationships of SEBs to Ethics Boards}{4}
+\subcontents{c\ {SEBs are necessary but not sufficient}}{5}
+\contents{2 Software engineering best practice}{5}
+\subcontents{a\ Introduction and standard references}{5}
+\subcontents{b\ Essential components of best practice}{6}
+\subcontents{\hbox to \howwide {\hfill (0)} Requirements}{6}
+\subcontents{\hbox to \howwide {\hfill (1)} {Formal methods}}{6}
+\subcontents{\hbox to \howwide {\hfill (2)} Defensive programming}{8}
+\subcontents{\hbox to \howwide {\hfill (3)} {Using dependable programming languages}}{8}
+\subcontents{\hbox to \howwide {\hfill (4)} Open source and version control}{8}
+\subcontents{\hbox to \howwide {\hfill (5)} Rigorous testing}{10}
+\subcontents{\hbox to \howwide {\hfill (6)} {Good documentation and record keeping}}{10}
+\subcontents{\hbox to \howwide {\hfill (7)} Usability}{11}
+\subcontents{\hbox to \howwide {\hfill (8)} Reusing quality solutions}{12}
+\subcontents{\hbox to \howwide {\hfill (9)} Simplicity}{12}
+\subcontents{\hbox to \howwide {\hfill (10)} {Compliance with standards}}{13}
+\subcontents{\hbox to \howwide {\hfill (11)} Effective multidisciplinary teamwork}{13}
+\subcontents{\hbox to \howwide {\hfill (12)} {Continuous Professional Development (CPD)}}{13}
+\subcontents{\hbox to \howwide {\hfill (13)} {Security and other factors}}{14}
+\contents{3 Code, data and publication}{14}
+\contents{4 {The Speigelhalter trustworhiness questions}}{17}
+\subcontents{a\ {How trustworthy are the numbers?}}{17}
+\subcontents{b\ {How trustworthy is the source?}}{18}
+\subcontents{c\ {How trustworthy is the interpretation?}}{18}
+\contents{5 {Summary of pilot survey}}{21}
+\subcontents{a\ {Assessment criteria and methods}}{21}
+\subcontents{b\ {Detecting and defending against error}}{24}
+\subcontents{c\ {Code policies of sampled journals}}{25}
+\subcontents{d\ {Assessments and criteria}}{26}
+\subcontents{e\ {Summary of assessments}}{30}
+\subcontents{f\ {References for sampled papers}}{31}
+% end expanding generated-supplementary.toc
 }
+
 \newwrite\contentsfile
@@
 \end{enumerate}
+
 There are other ideas to help make SEBs work, but it is clear they are part of the solution.
@@
 
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
+\vbox{\noindent
+      % expand generated-minicontents.tex
+\nnInContents{0}{Requirements}
+\nnInContents{1}{{Formal methods}}
+\nnInContents{2}{Defensive programming}
+\nnInContents{3}{{Using dependable programming languages}}
+\nnInContents{4}{Open source and version control}
+\nnInContents{5}{Rigorous testing}
+\nnInContents{6}{{Good documentation and record keeping}}
+\nnInContents{7}{Usability}
+\nnInContents{8}{Reusing quality solutions}
+\nnInContents{9}{Simplicity}
+\nnInContents{10}{{Compliance with standards}}
+\nnInContents{11}{Effective multidisciplinary teamwork}
+\nnInContents{12}{{Continuous Professional Development (CPD)}}
+\nnInContents{13}{{Security and other factors}}
+% end expanding generated-minicontents.tex
       }
@@
 
-\bibliography{paper-seb-supplementary-material.bib}
+% expand bibliography paper-seb-supplementary-material.bbl
+\begin{thebibliography}{10}
+
+\bibitem{vancouver}
+{Vancouver Group}.
+\newblock Uniform requirements for manuscripts submitted to biomedical
+  journals.
+\newblock JAMA. 1997;277(11):927--934.
+\newblock doi:{10.1001/jama.1997.03540350077040}.
+
+\bibitem{cosmos}
+Stepney S, Polack FAC, Alden K, Andrews PS, Bown JL, Droop A, et~al.
+\newblock Engineering Simulations as Scientific Instruments: A Pattern
+  Language.
+\newblock Springer; 2018.
+
+\bibitem{ai-ethics}
+Misselhorn C.
+\newblock Artificial Morality. {Concepts}, Issues and Challenges.
+\newblock Social Science and Public Policy. 2018;55:161--169.
+\newblock doi:{10.1007/s12115-018-0229-y}.
+
+\bibitem{sommerville}
+Sommerville I.
+\newblock Software Engineering.
+\newblock 10th ed. Pearson; 2015.
+
+\bibitem{knight}
+Knight J.
+\newblock Fundamentals of Dependable Computing for Software Engineers.
+\newblock CRC Press; 2012.
+
+\bibitem{wpf}
+Zeller A.
+\newblock Why Programs Fail: A Guide to Systematic Debugging.
+\newblock Morgan Kaufmann; 2006.
+
+\bibitem{humphrey}
+Humphrey WS.
+\newblock PSP: A Self-Improvement Process for Software Engineers.
+\newblock Addison Wesley; 2005.
+
+\bibitem{managers}
+Humphrey WS.
+\newblock Winning with Software: An Executive Strategy.
+\newblock Addison-Wesley Professional; 2001.
+
+\bibitem{heedless}
+Thimbleby H.
+\newblock Heedless Programming: Ignoring Detectable Error is a Widespread
+  Hazard.
+\newblock Software --- Practice {\&} Experience. 2012;42(11):1393--1407.
+\newblock doi:{10.1002/spe.1141}.
+
+\bibitem{plos}
+Chao DL, Halloran ME, Obenchain VJ, {Longini Jr} IM.
+\newblock FluTE, a Publicly Available Stochastic Influenza Epidemic Simulation
+  Model.
+\newblock PLOS Computational Biology. 2010;6(1):e1000656.
+\newblock doi:{10.1371/journal.pcbi.1000656}.
+
+\bibitem{excel1}
+Abir M, Nelson C, Chan EW, Al-Ibrahim H, Cutter C, Patel K, et~al.
+\newblock RAND Critical Care Surge Response Tool: An Excel-Based Model for
+  Helping Hospitals Respond to the COVID-19 Crisis.
+\newblock RAND Corporation; 2020.
+\newblock Available from: \url{www.rand.org/pubs/tools/TLA164-1.html}.
+
+\bibitem{excel2}
+Alvarez NM, Gonzalez-Gonzalez E, Trujillo-de Santiago G.
+\newblock Modeling {COVID-19} epidemics in an Excel spreadsheet:
+  {Democratizing} the access to first-hand accurate predictions of epidemic
+  outbreaks.
+\newblock MedRxiv. 2020;doi:{10.1101/2020.03.23.20041590}.
+
+\bibitem{npm}
+Schlueter IZ.
+\newblock Blog: kik, left-pad, and npm.
+\newblock NPM Blog; 23 March 2016.
+\newblock Available from:
+  \url{blog.npmjs.org/post/141577284765/kik-left-pad-and-npm}.
+
+\bibitem{LP}
+Knuth DE.
+\newblock Literate programming. vol.~27.
+\newblock Center for the Study of Language and Information Publication Lecture
+  Notes; 1992.
+
+\bibitem{hci1}
+Shneiderman B, Plaisant C, Cohen M, {\emph{et al}}.
+\newblock Designing the User Interface: Strategies for Effective Human-Computer
+  Interaction.
+\newblock 6th ed. Pearson; 2016.
+\newblock Available from: \url{www.cs.umd.edu/hcil/DTUI6}.
+
+\bibitem{hci2}
+Thimbleby H.
+\newblock Press On: Principles of Interaction Programming.
+\newblock MIT Press; 2007.
+
+\bibitem{flexibility}
+Hanson C, Sussman GJ.
+\newblock Software design for flexibility: How to avoid programming yourself
+  into a corner.
+\newblock MIT Press; 2021.
+
+\bibitem{iso}
+{ISO/IEC JTC 1/SC 7 Software and systems engineering Committees}.
+\newblock Software engineering --- Guidelines for the application of ISO
+  9001:2015 to computer software.
+\newblock International Organization for Standardization (ISO); 2015.
+\newblock Available from: \url{www.iso.org/standard/74348.html}.
+
+\bibitem{178C}
+{RTCA Committee SC-205}.
+\newblock DO-178C --- Software Considerations in Airborne Systems and Equipment
+  Certification.
+\newblock RTCA; 2011.
+\newblock Available from: \url{my.rtca.org/NC__Product?id=a1B36000001IcmqEAC}.
+
+\bibitem{peerReview}
+Baum T, {Le\ss mann} H, Schneider K.
+\newblock The Choice of Code Review Process: A Survey on the State of the
+  Practice.
+\newblock In: Lecture Notes in Computer Science. vol. 10611 of Product-Focused
+  Software Process Improvement: 18th International Conference; 2017. p.
+  111--127.
+
+\bibitem{ssi-report}
+Brett A, Croucher M, Haines R, Hettrick S, Hetherington J, Stillwell M, et~al.
+\newblock State of the Nation Report for Research Software Engineers.
+\newblock Research Software Engineer Network; 2017.
+\newblock Available from: \url{zenodo.org/record/495360#.Xyfuyi2ZOCM}.
+
+\bibitem{ssi-url}
+{Software Sustainability Institute}. Web site; 2020.
+\newblock Available from: \url{software.ac.uk}.
+
+\bibitem{security-engineering}
+Anderson R.
+\newblock Security Engineering.
+\newblock 3rd ed. Wiley; 2020.
+
+\bibitem{cyber-cacm}
+Shostack A, Zurko ME.
+\newblock Secure Development Tools and Techniques Need More Research That Will
+  Increase Their Impact and Effectiveness in Practice.
+\newblock Communications of the ACM. 2020;63(5):39--41.
+\newblock doi:{10.1145/3386908}.
+
+\bibitem{Schneier}
+Schneier B.
+\newblock Click Here To Kill Everybody --- Security and Survival in a
+  Hyper-connected World.
+\newblock W. W. Norton {\&} Company, Inc; 2018.
+
+\bibitem{electronic-evidence}
+Mason S, Seng D.
+\newblock Electronic Evidence.
+\newblock 4th ed. Humanities Digital Library; 2017.
+
+\bibitem{viruses}
+Thimbleby H, Anderson SO, Cairns P.
+\newblock A Framework for Modelling Trojans and Computer Virus Infection.
+\newblock Computer Journal. 1999;41(7):444--458.
+\newblock doi:{10.1093/comjnl/41.7.444}.
+
+\bibitem{strachey}
+Strachey C.
+\newblock Fundamental Concepts in Programming Languages.
+\newblock Higher-Order and Symbolic Computation. 2000;13:11--49.
+\newblock doi:{10.1023/A:1010000313106}.
+
+\bibitem{henderson}
+Henderson P.
+\newblock Functional Programming: Application and Implementation.
+\newblock Computer Science. Prentice-Hall International; 1980.
+
+\bibitem{RMarkdown}
+Xie Y, Allaire JJ, Grolemund G.
+\newblock R Markdown: The Definitive Guide.
+\newblock Chapman {\&} Hall/CRC; 2020.
+
+\bibitem{RS-policy}
+Hurst P. The Royal Society -- Data sharing and mining; 2022.
+
+\end{thebibliography}
+
+% end expanding bibliography paper-seb-supplementary-material.bbl
 
@@
 
-\input generated-example-data.tex
+% expand generated-example-data.tex
+\def\countFields{30}
+\noindent\texttt{
+\hspace*{2em}\{\hspace*{2em}\newline
+\hspace*{4em} accessed: "14 July 2020",\newline
+\hspace*{4em} doubleChecked: "17 January 2021",\newline
+\hspace*{4em} authors: "Callahan A, Steinberg E, Fries JA, Gomba \ldots,\newline
+\hspace*{4em} year: 2020,\newline
+\hspace*{4em} title: "Estimating the efficacy of symptom-based \ldots,\newline
+\hspace*{4em} volume: 3,\newline
+\hspace*{4em} number: 95,\newline
+\hspace*{4em} journal: "Nature Digital Medicine",\newline
+\hspace*{4em} doi: "10.1038/s41746-020-0300-0",\newline
+\hspace*{4em} dataComment: "On request",\newline
+\hspace*{4em} hasCodeInPrinciple: 1,\newline
+\hspace*{4em} codeComment: "``Code is available upon request from th \ldots,\newline
+\hspace*{4em} pages: 3\newline
+\hspace*{2em}\}
+}
+% end expanding generated-example-data.tex
 
@@
 
+\newdimen\tabwidth
+\tabwidth=\textwidth
+\advance \tabwidth by -4em
+\def\specifyurl{\vskip 3mm
+\begingroup\raggedleft\scriptsize
+} 
+\def\specifyurletc#1{\endgroup\\
+\hbox{ }\hfill\begin{tabular}{|p{\tabwidth}@{}}
+#1
+\end{tabular}\vskip .3cm
+}
+
 {\sf
@@
 Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
+As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.}
+
+\specifyurl
+\url{royalsociety.org/journals/authors/author-guidelines/#data}
+\specifyurletc{Accessed 29 July 2020; the policy has been revised (undated, but accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
+The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
+
+\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
+{\sf --- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
+[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.}
+
+\specifyurl
+\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
+\specifyurletc{Accessed 29 July 2020; since updated to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
+Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.'' (accessed 2 February 2022).
 }
 
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
+\noindent\textbf{\emph{Lancet Digital Health\/} author guidelines}\\
+Journal has detailed data policies, but no code policy.
+
+\specifyurl
 \url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
+\specifyurletc{Accessed 29 July 2020. Still no code policy (accessed 2 February 2022).}
 
 \noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
+Journal has detailed data policies, but no code policy.
 While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
 
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
+\vskip 3mm \noindent {\sf --- The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
+[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.}
+\specifyurl
 \url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
+\specifyurletc{Accessed 29 July 2020. Policy unchanged when accessed 2 February 2022.}
 
@@
 Legend:\\
-\input generated-legend.tex
+% expand generated-legend.tex
+\begin{tabular}{lp{4.5in}}
+\flagStyle{$\sf P_c$}&Journal has a code policy (see section~\ref{supplementary-journal-policies-section})\\
+\flagStyle{$\sf P_{{\mbox{\scriptsize c-breach}}}$}&Paper breaches journal code policy (see section~\ref{supplementary-journal-policies-section})\\
+\flagStyle{$\sf R_c$}&Paper uses a code repository (e.g., GitHub)\\
+\flagStyle{$\sf R_{\mbox{\scriptsize c-empty}}$}&Code repository contains no code\\
+\flagStyle{$\sf R_d$}&Paper uses a data repository (e.g., Dryad, Figshare, GitHub)\\
+\flagStyle{$\sf S_{\mbox{\scriptsize NONE}}$}&No code available at all (note: code is not expected for standard models, systems or statistical methods)\\
+\flagStyle{$\sf S_p$}&Paper says source code is available in principle\\
+\flagStyle{$\sf S_{+}$}&Paper or URL provides source code\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize rigorous}}}$}&Evidence that source code was developed rigorously\\
+\flagStyle{$\sf S_{\mbox{\scriptsize tested}}$}&Evidence that source code has been run with a clean build and tested\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize otherSE}}}$}&Other evidence of good practice; see details in summary table\\
+\flagStyle{$\sf C_0$}&Code has no non-trivial comments\\
+\flagStyle{$\sf C_1$}&Code only has trivial comments (e.g., copyright)\\
+\flagStyle{$\sf C_2$}&Helpful comments explaining code intent, rather than rephrasing the code\\
+\flagStyle{$\sf C_{+}$}&Code has substantial, useful comments and documentation\\
+\end{tabular}
+% end expanding generated-legend.tex
 
@@
 \sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
+% expand generated-assessments.tex
+\citenum{ref-1} & On request\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
+Documented overview, but only trivial comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''``''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request''\flagStyle{ } & Poor commenting, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team''\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
+However, the data are available from the corresponding author on reasonable request''\flagStyle{ } & Empty GitHub repository: ``Code coming soon\ldots'' it says\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
+Data from UCSF are not available at this time as they have not been legally certified as being De-Identified, however, this process is underway and the data may be available by the time of publication by contacting the authors.
+Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply''\flagStyle{ } & Basic documentation, very little comment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request''\flagStyle{ } & Trivial comments, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available''\flagStyle{ } & ``We used only free and open-source software'' some of which is unspecified\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{paper-usesRMarkdown} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
+\citenum{ref-14} & Data directly written into program code\flagStyle{ } & Basic Matlab with routine comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-15} & Data available on Dryad plus publicly available data from the 1000 genomes project.
+Currently (apparently) for private view\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
+Paper describes using two contrasting methods to help confirm correctness\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-16} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-17} & On request\flagStyle{ } & R lightly commented\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ethics-paper} & No data required\flagStyle{ } & Unrunnable incomplete code fragment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-19} & Data embedded in PDF\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-20} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-21} & Partial data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including manual\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
+\citenum{ref-22} & No data required\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-23} & Data cited, not all available\flagStyle{ } & Trivial documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{example-numerical-error} & On Figshare\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
+Helpful features to make usable for third parties\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-25} & Data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-26} & Data on various web sites\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-27} & Data on request\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-28} & Data on request\flagStyle{ } & Python scripts can be requested\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-29} & Unspecified location on large website requiring registration\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-30} & Available to researchers who meet criteria for access to confidential data\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-31} & Data access conditional on approved study proposal\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-32} & Unspecified locations on several large websites\flagStyle{ } & Python used and apparently GitHub, but --- an oversight?
+--- no code is available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+% end expanding generated-assessments.tex
 \end{longtable}
@@
 \begin{center}
-\input generated-summary-table.tex
+% expand generated-summary-table.tex
+\begin{tabular}{|rrrc|}\hline
+Number of papers sampled relying on code&32&100\%&\\\hline\hline
+\multicolumn{4}{|l|}{\textbf{Access to code}}\\
+Have some or all code available&12&38\%&\\
+Some or all code in principle available on request&8&25\%&\\
+No code available&12&38\%&\\\hline
+\multicolumn{4}{|l|}{{\textbf{Evidence of basic good software engineering practice}}}\\
+{Evidence program designed rigorously}&{0}&{0\%}&\\
+{Evidence source code properly tested}&{0}&{0\%}&\\
+{Other methods, e.g., independent coding methods}&{1}&{3\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
+Substantial code documentation and comments&2&6\%&\\
+Comments explain some code intent&3&9\%&\\
+Procedural comments (e.g., author, date, copyright)&10&31\%&\\
+No usable comments&17&53\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Repository use}}\\
+Code repository (e.g., GitHub) --- 1 was empty&10&31\%&\\
+Data repository (e.g., Dryad or GitHub)&9&28\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
+Papers published in journals with code policies&26&81\%&\\
+Clear breaches of code policy (if any) & 11&42\%&($N=26$)\\
+\hline\end{tabular}
+% end expanding generated-summary-table.tex
 \end{center}
@@
 
-\input generated-supplementary-references.tex
+% expand generated-supplementary-references.tex
+\vbox{\bibitem{ref-1}
+Callahan A,  Steinberg E,  Fries JA,  Gombar S,  Patel B,  Corbin CK and  Shah NH, ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp, 2020.
+DOI \texttt{10.1038/s41746-020-0300-0}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{MetricSelectionFramework}\bibitem{ref-2}
+Kanzler CM,  Rinderknecht MD,  Schwarz A,  Lamers I,  Gagnon C,  Held JPO,  Feys P,  Luft AR,  Gassert R and  Lambercy O, ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp, 2020.
+DOI \texttt{10.1038/s41746-020-0286-7} {Code \url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-3}
+Roy A,  Nikolitch K,  McGinn R,  Jinah S,  Klement W and  Kaminsky ZA, ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp, 2020.
+DOI \texttt{10.1038/s41746-020-0287-6}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-4}
+Levine DM,  Co Z,  Newmark LP,  Groisser AR,  Holmgren AJ,  Haas JA and  Bates DW, ``Design and testing of a mobile health application rating tool,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp, 2020.
+DOI \texttt{10.1038/s41746-020-0268-9}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-5}
+Kannampallil T,  Smyth JM,  Jones S,  Payne PRO and  Ma J, ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp, 2020.
+DOI \texttt{10.1038/s41746-020-0278-7}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{PENet}\bibitem{ref-6}
+Huang S,  Kothari T,  Banerjee I,   Chute C,  Ball RL,  Borus N,  Huang A,  Patel BN,  Rajpurkar P,  Irvin J,  Dunnmon J,   Bledsoe J,  Shpanskaya K,  Dhaliwal A,  Zamanian R,  Ng AY and  Lungren MP, ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0266-y} {Code \url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-7}
+Dhruva SS,  Ross JS,  Akar JG,   Caldwell B,  Childers K,  Chow W,  Ciaccio L,  Coplan P,  Dong J,  Dykhoff HJ,  Johnston S,  Kellogg T,  Long C,  Noseworthy PA,  Roberts K,  Saha A,  Yoo A and  Shah ND, ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
+Hofer IS,  Lee C,  Gabel E,  Baldi P and  Cannesson M, ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp, 2020.
+DOI \texttt{10.1038/s41746-020-0248-0} {Code \url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
+Norgeot B,  Muenzen K,  Peterson TA,  Fan X,  Glicksberg BS,  Schenk G,  Rutenberg E,  Oskotsky B,  Sirota M,  Yazdany J,  Schmajuk G,  Ludwig D,  Goldstein T and  Butte AJ, ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0258-y} {Code \url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
+Choi D,  Park JJ,  Ali T and  Lee S, ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp, 2020.
+DOI \texttt{10.1038/s41746-020-0261-3} {Code \url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-11}
+Hilton CB,  Milinovich A,  Felix C,  Vakharia N,   Crone T,   Donovan C,  Proctor A and  Nazha A, ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
+Li MD,  Chang K,  Bearce B,  Chang BY,  Huang AJ,  Campbell JP,  Brown JM,  Singh P,  Hoebel KV,   Erdo{\u g}mu\c{s} D,  Ioannidis S,  Palmer W,  Chiang MF and  Kalpathy-Cramer J, ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0255-1} {Code \url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
+Hoffman JI,  Nagel R,  Litzke V,  Wells DA and  Amos W, ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp, 2020.
+DOI \texttt{10.1098/rsos.200419} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
+Gr\"onquist P,  Panchadcharam P,  Wood D,  Menges A,  R\"uggeberg M and  Wittel FK, ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp, 2020.
+DOI \texttt{10.1098/rsos.192210} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{ref-15}
+Amos W, ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp, 2020.
+DOI \texttt{10.1098/rsos.191900} {Code \url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
+Gordon M,  Viganola D,  Bishop M,  Chen Y,  Dreber A,  Goldfedder B,  Holzmeister F,  Johannesson M,  Liu Y,  Twardy C,  Wang J and  Pfeiffer T, ``Are replication rates the same across academic fields?
+Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp, 2020.
+DOI \texttt{10.1098/rsos.200566} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
+Evans D and  Field AP, ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp, 2020.
+DOI \texttt{10.1098/rsos.200422} {Code \url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ethics-paper}
+Beale N,  Battey H,  Davison AC and  MacKay RS, ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp, 2020.
+DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-19}
+Cherevko AA,  Gologush TS,  Petrenko IA,  Ostapenko VV and  Panarin VA, ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp, 2020.
+DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
+Soczawa-Stronczyk AA and  Bocian M, ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp, 2020.
+DOI \texttt{10.1098/rsos.200622} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{lactModel}\bibitem{ref-21}
+Duruz S,  Vajana E,  Burren A,  Flury C and  Joost S, ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp, 2020.
+DOI \texttt{10.1098/rsos.200638} {Code \url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-22}
+de Azevedo EZD,  Dantas DV and  Daura-Jorge FG, ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp, 2020.
+DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{LRM}\bibitem{ref-23}
+Abdolhosseini-Qomi AM,  Jafari SH,  Taghizadeh A,  Yazdani N,  Asadpour M and  Rahgozar M, ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp, 2020.
+DOI \texttt{10.1098/rsos.191928} {Code \url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{1}\bibitem{example-numerical-error}
+Webster J and  Amos M, ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp, 2020.
+DOI \texttt{10.1098/rsos.200307} {Code \url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-25}
+Zhu Y-l,  Wang C-J,  Gao F,  Xiao Z-x,  Zhao P-l and  Wang J-y, ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp, 2020.
+DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-26}
+Yu B,  Scott CJ,  Xue X,  Yue X and  Dou X, ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp, 2020.
+DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-27}
+Joon-myoung K,  Younghoon C,  Ki-Hyun J,  Soohyun C,  Kyung-Hee K,  Seung D B,  Soomin J,  Jinsik P and  Byung-Hee O, ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-28}
+Zhu H,  Cheng C,  Yin H,  Li X,  Zuo P,  Ding J,  Lin F,  Wang J,  Zhou B,  Li Y,  Hu S,  Xiong Y,  Wang B,  Wan G,  Yang X and  Yuan Y, ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
+Fung R,  Villar J,  Dashti A,  Ismail LC,  Staines-Urias E,  Ohuma EO,  Salomon LJ,  Victora CG,  Barros FC,  Lambert A,  Carvalho M,  Jaffer Y A,  Noble JA,  Gravett MG,  Purwar M,  Pang R,  Bertino E,  Munim S,  Min AM,  McGready R,  Norris SA,  Bhutta ZA,  Kennedy SH,  Papageorghiou AT and  Ourmazd A, ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30131-X} {Code \url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-30}
+Sabanayagam C,  Xu D,  Ting DSW,  Nusinovici S,  Banu R,  Hamzah H,  Lim C,  Tham Y-C,  Cheung CY,  Tai ES,  Wang XY,  Jonas JB,   Cheng C-Y,  Lee ML,  Hsu W and  Wong TY, ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30063-7}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{blast-ct}\bibitem{ref-31}
+Monteiro M,  Newcombe VF,  Mathieu F,  Adatia K,  Kamnitsas K,  Ferrante E,  Das T,  Whitehouse D,  Rueckert D,  Menon DK and  Glocker B, ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30085-6} {Code \url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-32}
+Liu K-L,  Wu T,  Chen P-T,  Tsai Y M,  Roth H,  Wu M-S,  Liao W-C and  Wang W, ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30078-9}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
+
+% end expanding generated-supplementary-references.tex
 

### 2022-02-28 generated-constants.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
-\renewcommand\thesubsection{\thesection.\alph{subsection}}
-%\usepackage{hyperref}
-%\usepackage{refcount}
-\usepackage{url}
-\usepackage{doi}
-\usepackage{longtable}
-\usepackage{color}
-\usepackage{amsmath,amssymb}
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\input paper-seb-macros.tex
-
-\makeatletter
-      \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
-      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
-      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
-      }
-\makeatother
-
-\begin{document}
-
-\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}
-
-\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}
-
-%\orcid{0000-0003-2222-4243}
-
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
-\def\href#1{#1}
-
-\maketitle
-
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
-
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
-
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
-
-% start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
-
-\setcounter{section}{0}
-\def\n#1{{\sf\bfseries ({#1})}}
-
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
-
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
-
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
-
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
-
-%\newpage
-
-\makeatletter
-\let \bslash = \@backslashchar
-\makeatother
-
-%\renewcommand{\thesubsection}{\alph{subsection}}
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
-
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
-}
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
-
-\newpage
-\mysection{Further issues for Software Engineering Boards (SEBs)}
-
-\highlightmysubsection{Brief definition}
-Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
-
-Further details of the SEB proposal is in the main paper.
-Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:
-
-\begin{enumerate}\raggedright
-\item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
-
-\item
-When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}.
-Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
-How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
-
-\item 
-SEBs require policies on professional issues such as membership, transparency, and accountability.
-
-\item
-There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).
-
-\item
-Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs.
-For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?
-
-\item
-In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling.
-More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.
-
-\item
-SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
-
-\item
-Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large.
-Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
-\end{enumerate}
-There are other ideas to help make SEBs work, but it is clear they are part of the solution.
-We must not let perfection be the enemy of the good.
-SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.
-
-\mysubsection{Relationships of SEBs to Ethics Boards}
-
-\begin{enumerate}\raggedright
-\item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
-SEBs may also provide mentoring and training.
-
-\item
-Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
-
-\item 
-Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious).
-On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own.
-SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is.
-As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.
-
-\item
-SEBs should have a two-way collaboration with Ethics Boards.
-
-\begin{itemize}
-\item SEBs have to deal with ethical concerns, and how they may be implemented in code.
-One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.
-
-\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
-\end{itemize}
-
-\item
-Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
-\end{enumerate}
-
-\changemysubsection{SEBs are necessary but not sufficient}
-The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
-
-SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results.
-As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
-It needs doing.
-
-\mysection{Software engineering best practice}
-\label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
-
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
-
-The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible.
-Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer.
-Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}.
-Further suggestions for background reading can be found throughout this section.
-
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
-It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
-On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
-
-In practice, requirements and expectations change.
-Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
-In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
-
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
-
-{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy.
-To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe.
-Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}
-
-{The type and use of mathematics used in software engineering is formal methods.
-Not using formal methods ensures the resulting code is unsafe and unreliable.
-Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}
-
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
-Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
-Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.
-
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
-
-Many programming languages and programming environments have integrated features that support formal methods.
-For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@.
-In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
-Hoare is also a very good writer.
-
-Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way.
-Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy).
-Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do.
-Unfortunately, because formal methods are mathematical, few programmers have experience of using them.
-Fortunately tools are widely available to help use formal methods very effectively.
-
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
-Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
-The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code).
-Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.
-
-A special case of defensive programming appropriate for pandemic modeling is mixing methods.
-Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
-
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
-Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
-The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do.
-C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
-SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
-It is appreciated that the models may change and be adapted as new data and insights become available.
-Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable.
-Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used.
-Version control is supported by many tools (such as Git, Subversion, etc).
-
-If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
-Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
-
-{If results cannot be reproduced, has anything reliable been contributed?
-When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
-Reproducing the same results relying on the same codebase tells you little.
-The more independent reproductions of results the greater the evidence for belief in the implications.}
-
-Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes.
-Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.
-
-Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community.
-Open source is not a panacea, however; it raises many trade-offs.
-Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
-
-Open source raises important licensing and management questions to ensure the quality of contributions.
-A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository.
-This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.
-
-Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate.
-Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.
-
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
-In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
-Rigorous testing includes methods like fault injection.
-Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs.
-Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).
-
-It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again.
-For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).
-
-It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested.
-If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.
-
-There are many other important testing methods \cite{sommerville,knight,NVP}.
-
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
-Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
-
-{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories.
-One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
-}
-
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
-This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
-For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have?
-In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it.
-How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist.
-An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}
-
-There are many tools to help manage documentation (Javadoc, Doxygen, \ldots).
-Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}.
-Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.
-
-Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature).
-All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work.
-In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.
-
-A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
-
-Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project.
-Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.
-
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
-\globalLabel{supplementary-material-makefiles}
-Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
-Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team.
-Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.
-
-{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described.
-When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible.
-Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine.
-Makefiles typically also provide demo and test runs and other helpful features.
-Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}
-
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
-Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
-Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
-Indeed, reuse, scrutiny, and consensus are the foundations of good science.
-
-Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.
-
-A special case of reuse is to use software tools to help with software development.
-The tools (if appropriately chosen) have been carefully developed and widely tested.
-Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.
-
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
-When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
-This way lies over-fitting, a problem familiar from statistics (and machine learning).
-Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident.
-Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct.
-A simpler program is easier to prove correct, easier to program, and easier to debug.
-A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened.
-Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.
-
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
-
-To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}.
-However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate.
-Publications should then cite the  standards to which their computer models comply.
-
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
-
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
-
-As this long list illustrates, Software Engineering is a complex and wide-ranging subject.
-Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc).
-Multidisciplinary teamwork is essential.
-
-Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs.
-Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand).
-The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.
-
-Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.
-
-Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software.
-A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be.
-In short, users have to be brought into the software team too.
-
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
-Continuous Professional Development (CPD) is essential.
-
-Ironically, the more organized CPD the more likely the content itself will lag behind.
-There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees.
-Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.
-
-The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.
-
-Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
-
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
-Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
-
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
-In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
-
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
-The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
-
-This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
-
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
-
-A fragment of program code might say
-
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
-
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
-The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
-The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
-
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
-
-It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
-Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
-
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
-Data is then run on the virtual machine as program.
-Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions.
-Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program.
-There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal.
-In short, any distinctions between code and data are impossible to maintain.}
-
-In the present paper, we knowingly built on this blur between data and code.
-However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
-
-The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language.
-\LaTeX\ not only has text (as you are reading right now) but it also has code.
-For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted.
-When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters?
-The distinctions are all a bit moot.
-In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs.
-Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.
-
-The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above).
-As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data.
-A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer.
-In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.
-
-{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood).
-The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}
-
-Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.
-
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
-
-\def\questionsection#1{\changemysubsection{#1}}
-\def\question#1{\item \emph{#1\/}}
-\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}
-
-\questionsection{How trustworthy are the numbers?}
-\newcounter{resumeCounter}
-
-\begin{enumerate}
-\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.
-
-\sequestion{How rigorously has the software engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
-
-\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it.
-Were different methods used and compared, or was all confidence put into a single implementation?
-What internal consistency checks does the implementation have?
-Were invariants and assertions defined and checked?
-}
-
-\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.
-
-\sequestion{How are the claims presented that give us confidence in the code that they are based on?
-Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
-
-\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.
-
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the source?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed.
-Ask yourself, `Why does this source want me to hear this story?'
-
-\sequestion{The source of many science stories is the output of running some code.
-How reliable is this code?
-What evidence is there that the code was well-engineered so its reliability can be trusted?}
-
-\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.
-
-\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}
-
-\question{What am I not being told?} This is perhaps the most important question of all.
-Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.
-
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
-So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
-
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.
-
-\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}
-
-\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.
-
-\sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
-
-\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.
-
-\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}
-
-\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
-\end{enumerate}
-
-\newcount\temporary \temporary=\mainNumberOfReferences
-\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
-\advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
-
-\makeatletter    
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
-\makeatother
-
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
-\label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
-A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence and editorials were excluded.
-The sample represents what the editorial and the broader peer review community considers to be good practice.
-
-The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution.
-Indeed, all sampled papers used code in their research.
-Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable.
-It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).
-
-Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed.
-Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code.
-Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.
-
-There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
-
-The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted.
-Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey.
-What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized.
-Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
-
-As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.
-
-The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):
-
-\input generated-example-data.tex
-
-The data was written out by hand, after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
-(A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
-
-A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
-Hence, running the data generates the core human-readable information used in this paper.
-
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
-
-\begin{quote}
-\journalBreakdown 
-\end{quote}
-
---- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.
-
-An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:
-
-\begin{quote}
-\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
-\end{quote}
-
-it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data.
-The phrase above is now in fact written in \LaTeX\ in the paper as follows:
-
-\begin{quote}\tt\small
-Code repositories were used by \\
-\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
-\end{quote}
-
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
-
-Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
-
-The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-\highlightmysubsection{Detecting and defending against error}
-\label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
-% check it works for other numbers 
-%\countUsesVersionControlRepository=296
-
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
-\newcount\t
-\t=\countUsesVersionControlRepository
-\divide \t by 10
-\newcount\fraction
-\fraction=\countUsesVersionControlRepository
-\newcount\tent \tent=\t \multiply \tent by 10
-\advance\fraction by -\tent
-\the\t.\the\fraction,
-\t=\countUsesVersionControlRepository
-\advance \t by -1
-\the\t,
-\t=\countUsesVersionControlRepository
-\advance \t by 1
-\the\t, or
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
-
-\begin{itemize}\raggedright
-\item
-The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
-\divide \t by 10
-\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t) \cite{fixit}.
-
-\item
-The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
-\end{itemize}
-
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
-
-\newcount\min \min=5
-\newcount\max \max=20
-%\countUsesVersionControlRepository=4
-\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
-\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
-\ifnum \countUsesVersionControlRepository>\max
-	\else 
-		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
-		\fi
-\fi
-
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
-
-\highlightmysubsection{Assessments and criteria}
-
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
-
-Legend:\\
-\input generated-legend.tex
-
-\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
-\end{longtable}
-}
-
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
-
-\begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
-\raggedright
-
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
-
-\end{thebibliography}
-
-\immediate\closeout\infofile
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
-\end{document}
+\def\journalBreakdown{\emph{Lancet Digital Health\/} ($N=6$), \emph{Nature Digital Medicine\/} ($N=12$) and \emph{Royal Society Open Science\/} ($N=14$)}
+\def\tabularJournalBreakdown{&\hbox to 3em {\hfill 6}\hskip 1em \emph{Lancet Digital Health}\\&\hbox to 3em {\hfill 12}\hskip 1em \emph{Nature Digital Medicine}\\&\hbox to 3em {\hfill 14}\hskip 1em \emph{Royal Society Open Science}\\}
+\global\newcount \dataN \global\dataN=32
+\global\newcount \countAuthors \global\countAuthors=264
+\global\newcount \countHasBreach \global\countHasBreach=11
+\global\newcount \countHasPolicy \global\countHasPolicy=26
+\global\newcount \countUsesVersionControlRepository \global\countUsesVersionControlRepository=10
+\global\newcount \counthasDataRepository \global\counthasDataRepository=9
+\global\newcount \countNoCodeInRepo \global\countNoCodeInRepo=1
+\global\newcount \numberOfJournals \global\numberOfJournals=3
+\global\newcount \countCodetested \global\countCodetested=0
+\global\newcount \hasDevelopedRigorously \global\hasDevelopedRigorously=0

### 2022-02-28 generated-example-data.tex
All files updated for PLOS Feb 28 2022 submission

\def\countFields{32}
\noindent\texttt{
\hspace*{\codeIndent}\{\hspace*{\codeIndent}\newline
\hspace*{2\codeIndent} accessed: "14 July 2020",\newline
\hspace*{2\codeIndent} doubleChecked: "17 January 2021",\newline
\hspace*{2\codeIndent} authors: "Callahan A, Steinberg E, Fries JA, Gomba \ldots,\newline
\hspace*{2\codeIndent} year: 2020,\newline
\hspace*{2\codeIndent} title: "Estimating the efficacy of symptom-based \ldots,\newline
\hspace*{2\codeIndent} volume: 3,\newline
\hspace*{2\codeIndent} number: 95,\newline
\hspace*{2\codeIndent} journal: "Nature Digital Medicine",\newline
\hspace*{2\codeIndent} doi: "10.1038/s41746-020-0300-0",\newline
\hspace*{2\codeIndent} dataComment: "On request",\newline
\hspace*{2\codeIndent} hasCodeInPrinciple: 1,\newline
\hspace*{2\codeIndent} codeComment: "``Code is available upon request from th \ldots,\newline
\hspace*{2\codeIndent} pages: 3\newline
\hspace*{\codeIndent}\}
}

### 2022-02-28 generated-info-for-main.tex
All files updated for PLOS Feb 28 2022 submission

\newlabel{supplementary-material-makefiles}{{2}{40}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{on-code-data-publication}{{3}{43}{XXX}{YYYY}{ZZZZ}}
\newlabel{supplementary-Speigelhalter-section}{{4}{46}{DUMMYA}{DUMMYB}{DUMMYC}}
\def\MaxMainPaperCitationNumber{79}
\newlabel{supplementary-journal-policies-section}{{5}{55}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{supplementary-summary-table}{{5}{60}{DUMMY1}{DUMMY2}{DUMMY3}}
\expandafter\def \csname cite-MetricSelectionFramework\endcsname{[113]}
\expandafter\def \csname cite-PENet\endcsname{[117]}
\expandafter\def \csname cite-PostoperativeOutcomes.RiskNet\endcsname{[119]}
\expandafter\def \csname cite-philter-ucsf\endcsname{[120]}
\expandafter\def \csname cite-AI-CDSS-Cardiovascular-Silo\endcsname{[121]}
\expandafter\def \csname cite-SiameseChange\endcsname{[123]}
\expandafter\def \csname cite-dryad.1g1jwstrw\endcsname{[124]}
\expandafter\def \csname cite-rsos.192210\endcsname{[125]}
\expandafter\def \csname cite-ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ\endcsname{[126]}
\expandafter\def \csname cite-rsos.200566\endcsname{[127]}
\expandafter\def \csname cite-?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c\endcsname{[128]}
\expandafter\def \csname cite-dryad.vx0k6djnr\endcsname{[131]}
\expandafter\def \csname cite-lactModel\endcsname{[132]}
\expandafter\def \csname cite-LRM\endcsname{[134]}
\expandafter\def \csname cite-1\endcsname{[135]}
\expandafter\def \csname cite-manifold-ga\endcsname{[140]}
\expandafter\def \csname cite-blast-ct\endcsname{[142]}

### 2022-02-28 generated-repos.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-main.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\input plos-header.tex
-
-\renewcommand\thesubsection{(\thesection.\alph{subsection})}
-\usepackage{hyperref}
-
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\makeatletter
-    \def\do@url@hyp{\do/}
-    \usepackage{moreverb,url}
-    %    \renewcommand{\thefootnote}{\arabic{footnote}}
-\makeatother
-
-\input paper-seb-macros.tex
-
-\input generated-page-lengths.tex
-
-\begin{document}
-
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Harold Thimbleby
-\\
-\bigskip
-See Change Fellow in Digital Health
-\\
-\bigskip
-
-% Current address notes
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
-&\texttt{harold@thimbleby.net}\\
-&ORCID 0000-0003-2222-4243
-\end{tabular}
-
-%\orcid{0000-0003-2222-4243}
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-\noindent
-\emph{Background:} Computer code underpins modern science.
-While scientific models are routinely peer reviewed, the algorithms and the quality of code implementing them often avoid scrutiny; therefore the details of many scientific conclusions cannot be rigorously justified.
-
-\noindent\emph{Problem:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit.
-Additionally, both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects.
-Code is fallible, so any interpretation that relies on code is therefore fallible, and if the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized.
-In turn, scientific claims cannot be properly scrutinized.
-
-\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.
-
-\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.
-
-\noindent\emph{Solutions:} Code can be improved using software engineering. This paper argues for specific solutions:
-
-\newdimen \mywidth \mywidth=\textwidth
-\advance \mywidth by -1em
-\parshape 2 0em \textwidth 1.5em \mywidth 
-\noindent 
-\hbox to 1.5em{ 1.\hfill}Professional software engineers can help and should be involved, particularly in critical research;
-
-\parshape 2 0em \textwidth 1.5em \mywidth 
-\noindent 
-\hbox to 1.5em{ 2.\hfill}``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;
-
-\parshape 2 0em \textwidth 1.5em \mywidth 
-\noindent 
-\hbox to 1.5em{ 3.\hfill}Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent software engineers.
-
-\noindent
-The \supplement\ provides a summary of professional software engineering best practice relevant to scientific research and publication.
-
-%\end{abstract}
-
-\vfill\begin{quote}
-\hfill``Criticism is the mother of methodology.''
-
-\hfill Robert P Abelson, \emph{Statistics as Principled Argument\/} \cite{abelson}\end{quote}
-\vfill
-\paragraph*{Keywords:} 
-Computational science; Software Engineering; Reproducibility; Scrutiny
-
-\newpage\section*{Author summary}
-
-%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
-Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales.
-His research focuses on human error and computer system design, particularly for healthcare.
-%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
-Harold won the British Computer Society Wilkes Medal.
-He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder.
-His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.
-See his web site, \texttt{www.harold.thimbleby.net}, for more details.
-\linenumbers
-
-%\maketitle
-
-\section{Introduction}\label{problems}
-Computational methods are comparatively new, powerful, and very flexible tools for scientists.
-Computational methods rely on programming, but unfortunately poor programming is very easy use to apparently get results with; such results are likely to be misleading.
-The role of computational methods can be compared to statistics, another useful scientific tool.
-Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics.
-Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.
-
-{I}{n good science}, all statistics, methods and results have to be reported very carefully and in precise detail. For example, a statistical claim might be summarized as follows:
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\sf\hskip -\wd0\box0 %Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
-\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 ~($\sf 95\% \mbox{\sf ~CI~} 0.78, 4.27, p = 0.0009$).
-Between group effect size $\sf d = 0.55 ~(95\% \mbox{\sf ~CI~} 0.32, 0.77)$.'' \cite{example-stats}\end{quote}
-
-This typical wording briefly summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
-It is a \emph{lingua franca}.
-It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics.
-
-Scientists write like this --- and conferences and journals require it --- because statistical claims need to be properly accountable and documented in a clear way.
-Speigelhalter \cite{Speigelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Speigelhalter-section}).
-Results should not be uncritically accepted just because they are claimed.
-
-The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted.
-Scientists work closely with competent, often specialist, statisticians who engage with the research from experiment design through to analysis and publication.
-Further, it is assumed that statistics will be peer reviewed, and that review will often lead to improvement.
-
-Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement.
-Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians.
-Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.
-
-Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them will be generated by computer.
-Indeed, computers are now very widely used, not just to calculate statistics, but to run the models (the term is defined below), do the data sampling and processing, and even to operate the sensors that generate the data that is analyzed.
-Computers are used to run human-participant surveys, such as web-based surveys.
-The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers.
-Computers even help with the word processing and typesetting of research.
-
-In short, computers, data, and computer code are central to modern science.
-However, using code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
-Software code, like statistics, is also subject to unintentional bias \cite{se-bias}.
-All these issues are non-trivial concerns requiring technical expertise to manage well.
-
-A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.
-
-Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims.
-Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.
-
-%When relevant details of models are omitted (or are not described in recognisable ways), it is impossible to properly scrutinise claims.
-Describing a model used, perhaps mathematically, or providing the code (e.g., to download) that implements it is as inadequate as just providing raw data without details of the statistics analysis supporting the claim.
-
-We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
-An interesting overlap with statistical and software engineering scrutiny is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)\@.'' yet \emph{how\/} those analyses might have been performed is not discussed.
-The problem is that admitting using SAS 9.2 or any other named system does not really help scrutiny, as such systems can do anything with the data.
-A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is reliable.
-
-It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements the models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases.
-Being able to understand (at least in principle) the exact code used in implementing a model is critical to having confidence in the results that rely on it.
-Unfortunately code is rarely considered a substantial part of the science to which it contributes.
-
-This paper reviews a selection of papers in leading international journals, and finds that both papers and journal policies take code for granted.
-This paper then argues that, just as is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.
-Specifically, code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny.
-Developing justifiably reliable code is the concern of the field of \emph{software engineering}, which will be discussed further below, as well as more substantially in this paper's \supplement).
-
-This paper shows that unreliable computational dependencies in science are widespread.
-Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction.
-In short, the quality of much modern science seems to be undermined because the code it relies on is rarely of adequate quality \emph{for the uses to which it is put}.
-
-This paper explores the extent of these software engineering problems in published science.
-The paper additionally suggests some ways forward.
-The \supplement\ is an integral part of the suggested solutions.
-In particular, the \supplement\ section~\ref{supplementary-Speigelhalter-section} summarizes Speigelhater's uncontroversial statistics probes and draws explicit analogies for the critical assessment of the quality of scientific code.
-
-\subsection{The role of code in science and scientific publication}
-%Yet without code, models could not be run.
-
-For the purposes of this paper, models map theory and parameters to describe phenomena, typically to make predictions or to test and refine the models.
-With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.
-
-Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive.
-The chance that a thousand lines of hand-written code is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability.
-How to do this well is the concern of software engineering, and is discussed throughout this paper.
-
-While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are (certainly in all cases reviewed later in this paper) they are long, complex, inscrutable and (our survey shows) lack adequate documentation.
-Therefore the models are very likely to be unreliable \emph{in principle}.
-If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if has not been clearly documented?
-Without documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?
-
-Without substantial documentation it is impossible to scrutinize code properly.
-Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications.
-What does it mean?
-What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation?
-Does the code mean what was intended?
-What are the assumptions on \texttt{y} and do they hold invariantly?
-Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown.
-A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of overflow).
-
-Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code.
-Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim.
-If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed.
-It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.
-
-Contrary to the views expressed in the present paper, some authors have asserted that code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality}.
-On the contrary, if code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways.
-Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.
-
-Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes.
-The question is whether it matters, \emph{when\/} it matters, and when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work.
-
-%This paper explores the role of code in scientific research.
-Initially focusing on examples from pandemic modelling (because of its relevance to serious matters of public health) this paper shows that scientific culture, including editorial processes, have not adapted to the increasingly dominant role of code and managing the risks of its fallibility.
-The paper then suggests how to improve, including building mutual engagement across the science and software engineering communities.
-
-\subsection{Bugs, code and programming}\label{knowledge}
-Critiques of data and model assumptions are very common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned.
-Yet data and program are formally equivalent (see further discussion in \supplement, section \ref{on-code-data-publication}).
-Program code has as great an affect on results as the data.
-Code is harder to scrutinize, which means errors in code can have more subtle effects on results.
-
-It should be noted that almost all code contains ``magic numbers'' --- that is, data masquerading as code.
-This common practice ensures that published data is very rarely all of the data because it omits the magic numbers.
-This emphasizes the need for repositories to require the inclusion of code so all data is actually available.
-
-Bugs can be understood as discrepancies between what code is intended to do and what it actually does.
-Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered.
-Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination.
-None of the code examined for this paper includes either.
-
-If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
-Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things.
-For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the sense of producing incorrect results --- that is, what was wanted was wrong.
-Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.
-
-\iffalse Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, unqualified programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering.
-Dunning and Kruger go on to say,
-
-\begin{quote}\sf\raggedright\setbox0=\hbox{``}\hskip -\wd0\copy0 People usually choose what they think is the most reasonable and optimal option \hbox{[\hskip 2pt\ldots]} The failure to recognize that one has performed poorly will instead leave one to assume that one has performed well; as a result, the incompetent will tend to grossly overestimate their skills and abilities.
-\hbox{[\hskip 2pt\ldots]} Not only do these people reach erroneous conclusions and make unfortunate choices, but their incompetence robs them of the metacognitive ability to realize it.''\end{quote}
-
-Unlike many skills (skating, brain surgery, \ldots) programming, typical of much engineering, is one where errors can go unnoticed for long periods of time --- things seem to work nicely right up to the moment they fail.
-The worse programmers are, the more trivial bugs they tend to make, but trivial bugs are easy to find so, ironically, being a poor programmer \emph{increases\/} one's self-assessment because debugging seems very productive.
-It is easy for poor programmers and their associates to believe they are better than they actually are.
-%This is fertile ground for the better-than-average bias \cite{dunning-kruger}.
-
-It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}.
-We \emph{all\/} make mistakes we are unaware of.
-If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.
-
-{Everyone is subject to Human Factors (including the author of the present paper, e.g., as discussed in \cite{enigma}): for instance, the standard cognitive bias of confirmation bias encourages us to look for bugs when code fails to do what is expected and then debug it to produce better results, but if code generates expected results not to bother to debug it further.
-This of course tends to make code increasingly conform to prior expectations, whether or not those expectations are scientifically justified.
-Typically, there was no prior specification of the code, so the code should be right, especially after all the debugging to make it ``correct''!
-Thus coding routinely suffers from HARKing (Hypothesizing After the Results are Known \cite{harking}), a methodological trap widely recognized in statistics.}
-
-Computers themselves are also a part of the problem.
-Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable.
-Programs can be written so that it is not possible to determine what they do or how they do it (whether deliberate obfuscation or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
-The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
-\fi
-
-\section{State of the art in pandemic modeling}
-A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no conscious role for code or computation as such.
-It appears that the new insights, advances, rigor, and problems that computers bring to research are not considered relevant.
-
-A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
-(TRIPOD guidelines ignore code completely.) The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted.
-It should be noted that flowcharts, which the review did consider, are programs intended for direct human use.
-Flowcharts, too, should be designed as carefully as code, for exactly the same reason: it is hard to program reliably.
-
-A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} for H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling.
-The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.
-
-A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\sf\hskip -\wd0\box0 
-For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
-'' \cite{ferguson-interview}\end{quote}
-
-This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
-The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
-(Typical $h$ values vary by discipline.)} of 93.
-Ferguson should be well aware of the standards of coding used in at least his own field.
-This comment, quoted above, is therefore likely to be representative of the standards of the field as a whole.
-
-Ferguson's admission is tantamount to saying that the published scientific findings are not reproducible.\footnote{A constructive discussion of software engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.} This is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
-If this C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation.
-It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified over fifteen years without becoming incoherent.
-If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
-In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
-
-The programming language C which was used is not a dependable language; to develop reliable code in C requires professional tools and skills.
-Moreover, C code is not portable, which limits making it available for other scientists to use safely (C notoriously gets different results with different compliers, libraries, or hardware).
-The \supplement\ discusses these issues further.
-
-Ferguson, author of the code, says of the criticisms of his code, ``However, none of the criticisms of the code affects the mathematics or science of the simulation'' \cite{thumbs-up}.
-Indeed.
-The problem the current paper is raising is that the theoretical epidemiology may be fine, but if it is not mapped into code that correctly implements the models, then the program's output cannot be relied on without independent evidence.
-In science, this is the normal requirement of \emph{reproducibility}.
-
-The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed the code could be run again and produced comparable results (compared, apparently, to an Excel spreadsheet!).
-That can be achieved at this low level of sophistication is hardly surprising, regardless of the quality of the code.
-There was no scientific insight that merits the use of the word ``reproduction.'' If reproducibility is to be a useful scientific criterion, an \emph{independently\/} developed model needs to produce equivalent results (called $N$ version programming, a standard software engineering practice \cite{NVP}) like public health surely requires --- as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.
-Meanwhile, it is a shame that using software for scientific models has enabled the bar to reproducibility, as understood by journals such as \emph{Nature}, to be lowered to a mechanical level that is only sufficient to detect some forms of dishonesty, as opposed to methodological limitations, which is the point of reproducing work.
-
-Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).
-It is
-\newcount\u
-\newcount\v
-\newcount\w
-\newcount\frac
-\input generated-covidsimSummary.tex
-kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
-Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done.
-If nothing else, as the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models.
-It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place.
-The revised code has now been published, and it has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
-
-Some epidemiology papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
-Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications.
-Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it is arguably even worse than not publishing code at all.
-Pseudo-code is not precise enough to help anyone scrutinize a model; copying pseudo-code introduces bugs.
-An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}.
-{The \supplement\ provides further discussion of reproducibility.}
-
-\subsection{Science beyond epidemiology}
-Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims.
-People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place.
-The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even downright flaws.
-
-Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}.
-World-class software can attain maybe 1 bug per 1,000 lines of code.
-Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new insights into their models.
-In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- can be subject to numerical errors \cite{hamming}.
-It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion).
-Such bugs undermine confidence in model results.
-
-Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what researchers have done, but merely making code available (for instance, providing it in their \supplement\ with papers, putting it in repositories, or using open source) is not sufficient.
-It is noteworthy that some papers disclosed that they had access to special hardware.
-
-Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available.
-While some \citeeg{unfinished,lancet-unfinished} do make what they call ``documented'' code available, they provide no more than superficial comments.
-This is \emph{not\/} documentation as properly understood.
-Such comments do not explain code, explain contracts, nor explain algorithms.
-Some readers of the present paper may not recognize these technical software terms; contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming {(see the \supplement\ for a checklist of many relevant, professional software engineering concepts and resources)}.
-
-If full code is made available, it may be technically ``reproducible,'' but the scientific point is to be able to understand and challenge, potentially refute, the findings; to do that, much more is required than merely being able to run the code \cite{notebooks,popper}.
-
-Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper {and indeed in computer science research itself \cite{machine-learning-reproducibility}}) is inscrutable.
-Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do.
-Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.
-
-Making outline code (including pseudo-code) available without proper documentation and without discussing its limitations is unethical: it encourages others to reproduce and build on poor work.
-
-\makeatletter
-\long\def\@makecaption#1#2{%
-  \vskip\abovecaptionskip
-  \sbox\@tempboxa{#1: #2}%
-  \ifdim \wd\@tempboxa >\hsize
-    \textbf{#1}: #2\par
-  \else
-    \global \@minipagefalse
-    \hb@xt@\hsize{\hfil\box\@tempboxa\hfil}%
-  \fi
-  \vskip\belowcaptionskip}
-\makeatother
-
-\begin{table*}
-\begin{center}
-\begin{tabular}{|rl|} \hline
-\the\numberOfJournals&Journals\\
-\the\dataN&Papers:\\
-\tabularJournalBreakdown
-\the\countAuthors&Published authors\\
-\the\totalPages&Published journal pages\\
-July 2020&Sample month\\ \hline
-\end{tabular}
-\end{center}
-
-%\newcount \total
-%\total=\dataN
-%\multiply \total by 4 % three reviewers, one editor
-%\advance \total by \countAuthors
-%\the\total\
-%leading scientists assuming authors, 1 editor, and 3 reviewers per paper, but not counting acknowledged colleagues et al, covering  (plus appendices and supporting material) in total.
-%
-\caption{Overview of peer-reviewed paper sample.}
-\label{table-overview}
-\end{table*}
-
-\begin{table*}
-\begin{center}\normalsize
-\input generated-summary-table.tex
-%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
-\end{center}
-
-\caption{Summary of survey results.}
-\label{table-summary}
-\end{table*}
-
-\begin{table*}[t]
-
-\def\reponame#1#2{{\tt #1} \csname cite-#2\endcsname}
-
-\expandafter\def\csname pagelength-covid-sim\endcsname{20}
-\expandafter\def\csname cite-covid-sim\endcsname{\cite{ICmodel}}
-
-% divide #1 by #2
-\newcount \uu
-\newcount \vv
-\def\makeAverage#1#2{%
-	\uu=#1
-	\vv=#2
-	\divide \uu by \vv
-	\the\uu
-}
-
-\input generated-clone-date.tex
-\begin{center}
-
-\expandafter \advance \gitPages by \csname pagelength-covid-sim\endcsname
-
-\input generated-repos.tex
-\vskip 1ex
-\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
-\\
-\small Repository clones downloaded and automatically summarized \clonedate
-\end{center}
-\newcount \pubdelayinmonths
-\pubdelayinmonths=\cloneyear
-\advance \pubdelayinmonths by -2020
-\multiply \pubdelayinmonths by 12
-\advance \pubdelayinmonths by -4 % published in April
-\advance \pubdelayinmonths by \clonemonth
-\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
-Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting.
-Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
-Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the\pubdelayinmonths\ months had already elapsed since the paper referring to the code was published.}
-\label{table-repo-summary}
-\end{table*}
-
-\section{A pilot survey of peer-reviewed research relying on code}
-\label{survey-section}
-The problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
-For instance, a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model} provides no details of its code.
-Its \supplement, which might have provided code, no longer exists.
-
-Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized.
-We therefore undertook a survey of recent papers in leading peer reviewed journals.
-
-\subsection{Methodology}
-A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.
-
-The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal.
-All papers sampled are Open Access, although for some papers some or all of the associated data has restricted access.
-Table \ref{table-overview} is an overview of the sample.
-
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence, and editorials were excluded.
-The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.
-
-The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.
-
-The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.
-
-%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).
-
-The \the\dataN\ papers cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install.
-Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment was made whether the code provided actually reproduced a paper's specific claims.
-Indeed, if we trust the papers that their code was actually run and provides the results as reported, then merely running their code (when provided in full) merely checks the paper/code consistency, but will not assess the quality or reliability of the code.
-
-%Assessing code quality has been a lively topic in software engineering for decades, and has international standards (ISO/IEC 9126, updated to ISO 25000).
-Here, code is assessed\footnote{The author of this paper has been assessing undergraduate and postgraduate computer science student code since the 1970s, using double marking and moderation.} using the n criteria of \cite{software-quality} as a marking scheme with scores out of 100/n.
-
-The full methodology, data and analysis, as well as many software engineering and Human Factors techniques to help improve reliability and reproducibility used in the preparation of the present paper, are provided in the \supplement, which is also available from a repository (see details at the end of this paper).
-
-%\emph{It is important that this survey is not taken as an evaluation of any specific paper: individual measurements are inevitably noisy.
-Instead, the aim is to form a view of scientific publishing generally.
-Mean scores, as summarized in table \ref{table-summary} and \ref{table-quality}, are more reliable measurements than the individual paper score breakdowns provided in the \supplement.}
-
-\subsection{Summary of results}
-The sample selection criteria necessarily identified scientific research with software engineering contributions.
-No evidence of verification and validation was seen.
-There was only one example of basic software engineering methods, namely independent coding.
-There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer --- who writes high quality code to start with!\ --- would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
-(See the \supplement\ for definitions and a list of standard methods.)
-
-Much code depended on substantial manual intervention to compile it.
-All code (where actually provided) was sufficiently complex that, if it was supposed to be used or scrutinized, required more substantial documentation than was provided.
-
-On the whole, on the basis of the sample evidence, scientists do not make their code available, and rarely provide adequate documentation (see table \ref{table-summary}).
-
-It is noteworthy that the papers themselves, typically a few published pages, are very small compared to the code they rely on; the papers represent much less authoring effort than the code and data that they rely on for their claims and correctness (see table \ref{table-repo-summary}).
-With one minor exception (which may have been accidental), no papers reported anything on any software engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
-
-None of the papers used any specific software engineering methods, such as open source \cite{open-source} and other standard methodologies provided in the \supplement, to help manage their processes and help improve quality.
-Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reason this paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}
-
-\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
-In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach.
-The implication of these results is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies, or they are technically unable to assess them --- and, if so, they also fail to say they are unable to assess them.
-This lack of expertise is consistent with the limited awareness of software engineering best practice that is manifest in the published papers (and resources) themselves.
-
-Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \input generated-clone-date.tex
-while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could not been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
-
-Sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
-
-Overall, there was no evidence that any code had been developed carefully, {let alone by using recognized professional software engineering methods}. In particular, 
-\ifnum \countCodetested=0
-no papers
-\else only \plural{\countCodetested}{paper}
-\fi in the survey provide any claims or evidence of effective testing, for instance with evidence that tests were run on clean builds.
-{While it may sound unrealistic to ask for evidence on software quality in a paper written for another field of science, the need is no less than the need for standard levels of rigor in statistics reporting, as discussed in the opening of this paper.}
-
-Data repositories (the Dryad Digital Repository, Figshare or similar) were used by \plural{\counthasDataRepository}{paper} to provide structured access to their data.
-Unlike GitHub, which is a general purpose repository, Dryad has scientifically-informed guidelines on handling data, and all papers that used Dryad provided more than just their raw data --- they provided a little, sometimes substantial, documentation for their data.
-At the time of writing, Dryad is not helpful for managing code --- its model appears to be founded on the requirement that once published papers must refer to exactly the data they used, so further refinements on the data (or code) are taboo, even with version control.
-
-Key findings from the survey are summarized in tables \ref{table-summary} and \ref{table-repo-summary}, and are discussed in greater detail in the \supplement.
-
-\section{Discussion}
-Effective access to quality code is not routine in science.
-Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
-The evidence suggests that, generally, some but not all manually developed code is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no clear evidence (over the survey period) that any published code was maintained using the repositories.
-
-The \supplement\ provides a summary of the analysis and full citations for all papers in the sample.
-
-\subsection{A call to action}\label{summary}
-Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide.
-Given the urgency of rigorously understanding COVID-19, any epidemic for that matter, it is essential that epidemiologists engage professional software engineers to help develop reliable laboratory methodologies.
-For instance, code lab books can be generated and signed easily.
-
-Software used for informing public health policy, medical research or other medical applications is \emph{critical software}.
-Professional critical software development, as used in aviation and the nuclear power industry, is (put briefly) based on \emph{correct by construction}: \cite{cbc} effectively, design it right first time, {supported by numerous rigorous techniques, such as formal methods, to manage error.
-(See extensive discussion in this paper's \supplement.)}\ Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable.
-Conversely, not following these practices undermines the rigor of the science.
-
-An analogous situation arises in ethics.
-There are some sorts of research that are ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
-Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems.
-National funders, and others, therefore require Ethics Boards to formally review ethical quality.
-Medical journals will not publish research that has not undergone appropriate ethical review.
-
-Analogously, and supplementing Ethics Boards, Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality software engineering.
-Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to epidemic modeling papers --- and in due course, all scientific papers --- being required to include Software Engineering Board clearance statements as appropriate.
-{Software Engineers have a code of ethics that applies to \emph{their\/} work in epidemic modeling \cite{ethics-code}.}
-
-{The present paper did not explore data, because in almost all cases the code and data were explained so poorly and archived so haphazardly it would be impossible to know whether the author's intentions were being followed.\footnote{{For the present paper, all the code, data, analysis and documents are available for download in a single zip file.}} Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
-Ideally, the authors should test a clean deployed build of their code and save the results.
-Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims.
-This requirement is no more unreasonable than requesting good statistics, as discussed in the opening of the paper.
-And the solution is the same: relevant experts --- whether statisticians or software engineers --- need to be routinely available and engaged with the science.
-SEBs would be a straight forward way of helping achieve this.}
-
-There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university.
-Active, professional software engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art.
-There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.
-
-Open Source generally improves the quality of software.
-SEBs will take account of the fact that open source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers.
-Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis {and other quality control methods for which the lead scientist and software engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).}\ {Software engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer reviewed publications \cite{acm-artifacts}.}\ {See this paper's \supplement\ for further suggestions.}
-
-A potential argument against SEBs is that they may become onerous, onerous to run and onerous to comply with their requirements.
-A more mature view is that SEBs need their processes to be adaptable and proportionate.
-If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation.
-Hence SEBs processes are likely to follow a risk analysis, perhaps starting with a simple checklist.
-{There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar.
-Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.}
-
-A professional organization, such as the UK Royal Academy of Engineering ideally working in collaboration with other national international bodies such as IFIP, should be asked to develop and support a framework for SEBs.
-SEBs could be quickly established to provide direct access to mature software engineering expertise for both researchers and for journals seeking competent peer reviewers.
-In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations.
-Given the urgency, this paper recommends that \emph{ad hoc\/} SEBs should be established for this purpose.
-
-SEBs are a new suggestion, providing a supportive, collaborative process.
-Methodological suggestions already made in the literature include open source and specific software engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}.
-While \cite{ABCs-SE} provides an insightful framework to conceptualize approaches and compare their merits, there is clearly scope for much further research to provide an evidence base to motivate and assess appropriate interventions to help scientists do more rigorous and effective software engineering to support their research and publishing.
-These and further issues are discussed at greater length in the \supplement.
-
-\subsection{Suggestions for further work}
-
-Although this study of software engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
-However, the insights appear to be general.
-
-Further work to extend the scope of the survey beyond the basic requirements of the present paper is of course worthwhile, but  the following cases (listed in the next few paragraphs) suggest that the problem is widespread.
-We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor.
-
-\label{jvs-policy}
-\begin{enumerate}\raggedright
-\item
-The journal \emph{The Lancet\/} published and then subsequently retracted a paper on using hydroxychloroquine as a treatment for COVID \cite{lancet-retracted}.
-The paper was found to rely on fraudulent data \cite{science-lancet1,science-lancet2}.
-\emph{The Lancet\/} subsequently tightened its data policies \cite{lancet-learning}, for instance to require that more than one author must have directly accessed and verified the data reported in the manuscript.
-Curiously, the original (now retracted) paper declares
-\begin{quote}\sf
-\setbox0=\hbox{``}\hskip -\wd0\copy0 \ldots\ all authors participated in critical revision of the manuscript for important intellectual content.
-MRM and ANP supervised the study.
-All authors approved the final manuscript and were responsible for the decision to submit for publication.'' \end{quote}
-which seems to suggest that several original authors of the paper would have been happy to make the new declarations --- and, of course, if there is fraud (as was established in this case) it seems likely that authors who make the new declarations of accessing and verifying data are unlikely to make reliable declarations.
-
-\hskip 1em\emph{The Lancet\/} still has no code publication policy (see the \supplement), and for more than one author to have ``direct access'' to the data they are very likely to access the data through the same code.
-If the code is faulty or fraudulent, an additional author's confirmation of the data is insufficient, and there is at least as much reason for code to be fraudulent (not least because code is much harder to scrutinize than data).
-Code needs more than one author to check it, and ideally reviewers independent of the authors so they do not share the same assumptions and systems (for instance shared libraries, let alone potential collusion in fraud).
-
-\item 
-In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms.
-Notably, the paper fails to provide access to its anonymized data (with or without qualification), and fails to define the data anonymization algorithm, and also fails to even mention the code that it developed and used to perform its study.
-The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see \supplement\ section \ref{supplementary-journal-policies-section}).
-Ironically, the retracted article \cite{jvs1} is still online (as of August 2020) with no reference to any editorial statement to the effect that it has been retracted, despite this being trivial --- and necessary --- to achieve in the widely-accessed online medium.
-
-\item
-Medical research often aims to establish a formula to define a clinical parameter (such as body mass index, BMI) or to specify an optimal drug dose or other intervention for treatment.
-These formulas, for which there is conventional clinical evidence, are often used as the basis for computer code that provides advice or even directly controls interventions.
-Unfortunately a simple formula as may be published in a medical paper is \emph{never\/} sufficient to specify code to implement it safely.
-For example, clinical papers do not need to evaluate or manage user error when operating apps, and therefore the statistical results of the research will be idealistic compared to the outcomes using an app under real conditions --- which is what the clinical research is supposedly for.
-A widespread bug (and its fix) that is often overlooked is discussed in \cite{numerals}; the paper includes an example of a popular clinical calculator (based on published clinical research) that calculated nonsense, and potentially dangerous, results.
-The paper \cite{fda} summarizes evidence that such bugs, ignored by the clinical research literature, are commonplace in medical systems and devices.
-\end{enumerate}
-
-\section{Conclusions}
-We need to improve the quality of software engineering that supports science.
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed in shows that current coding practice makes for poor science.
-In the current pandemic, scientific modeling, such as track and trace \cite{excel-fiasco}, balancing COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.
-
-The challenge to scientific research is to manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices.
-Computer code should be explicit, accessible (well-structured, etc), and properly documented.
-Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}.
-Professional software methodologies should not be ignored.
-
-Unfortunately, while programming is easy, and is often taken for granted, programming \emph{well\/} is very difficult \cite{fixit}.
-We know from software research than ordinary programming is very buggy and unreliable.
-Simply put, without quality code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
-Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion:  computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper}.
-
-Software Engineering Boards (as introduced in this paper) are a straight forward, constructive and practical way to support and improve computer-based science.
-This paper's \supplement\ summarizes the relevant professional software engineering practice that Software Engineering Boards would use, including discussing how and why software engineering helps improve code reliability, dependability, and quality.
-
-\section*{Supporting information}
-\newcount\csrefcount \csrefcount=0
-\def\csref{\global\advance\csrefcount by 1}
-\def\ethics#1{\paragraph*{Ethics}#1}
-\def\ack#1{\paragraph*{Acknowledgments}#1}
-\def\dataaccess#1{\paragraph*{Data and code access}#1}
-\def\aucontribute#1{\paragraph*{Author contribution}#1}
-\def\competing#1{\paragraph*{Competing interests}#1}
-\def\funding#1{\paragraph*{Funding}#1}
-
-\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson provided very helpful comments.}
-%\the\csrefcount\ computer science reviewers.
-
-\ethics{This article presents research with ethical considerations but does not fall not within the usual scope of ethics policies.}
-
-\dataaccess{There is a full discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}.
-All material is available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards} (which has been tested in a clean build, etc).
-The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \input generated-data.js-errors.tex
-possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset results reliably in this paper and its \supplement\ (e.g., table \ref{table-summary} in this paper) \emph{directly\/} from the data.
-In addition, a standard CSV file is generated in case this is more convenient to use, for instance to browse in Excel or other spreadsheet application.
-All data, analysis, and details of the methods used can be found in the \supplement.}
-
-\aucontribute{Harold Thimbleby is the sole author.
-An preliminary outline of this paper, containing no supplementary material or data, was submitted to the UK Parliamentary Science and Technology Select Committee's inquiry into UK Science, Research and Technology Capability and Influence in Global Disease Outbreaks, under reference LAS905222, 7 April, 2020.
-The evidence, which was not peer reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper.
-It is available to the public \cite{parliamentary-evidence,my-parliamentary-evidence}.}
-
-\competing{The author declares no competing interests.}
-
-\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], and by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186].
-The funders had no involvement in the research or in this paper.}
-
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-{\raggedright
-\initialiseBibliography{References}{1}{}
-% \bibliographystyle{plos2015} - already provided in plos-header.tex
-\bibliography{paper-seb-main.bib}
-
-% Note that the supplementary material calculates how many references there are,
-% because it reads in this .tex file's .aux file so that it can cite the references 
-% above with the same number sequence. 
-%
-% It simple redefines \bibitem to work it out while it
-% defines all the citations from the bibliography above.
-%
-% The same technique allows the supplementary material to have 2 more bibliographies,
-% with consecutive numbering.
-}
-
-\end{document}
-
-%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
-%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}
-
-%\def \allInOne {don't include header in supplemental material}
-%input paper-seb-supplementary-material.tex
-
-\label{LastPage}
-\end{document}
+\begin{tabular}{|l|r|rrr@{\hskip .75ex}l|}\hline
+&\multicolumn{1}{c|}{\bf PDF paper}&\multicolumn{4}{c|}{\bf Repository code \& data}\\ \cline{2-6}
+\multicolumn{1}{|c|}{\bf Github repository}&\multicolumn{1}{c|}{\bf Number}&\multicolumn{1}{c}{\bf Number}&\multicolumn{1}{c}{\bf Code}&\multicolumn{2}{c|}{\bf Data}\\
+\multicolumn{1}{|c|}{\bf and paper citation}&\multicolumn{1}{c|}{\bf of pages}&\multicolumn{1}{c}{\bf of files}&\multicolumn{1}{c}{\bf kLOC}&\multicolumn{2}{c|}{\bf bytes} \\ \hline\hline
+\reponame{AI-CDSS-Cardiovascular-Silo}{AI-CDSS-Cardiovascular-Silo} & \csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname & \commarise{206} & 143 & 64&Mb \\
+\reponame{blast-ct}{blast-ct} & \csname pagelength-blast-ct\endcsname & \commarise{44} & 2 & 87&Mb \\
+\reponame{covid-sim}{covid-sim} & \csname pagelength-covid-sim\endcsname & \commarise{229} & 25 & 734&Mb \\
+\reponame{lactModel}{lactModel} & \csname pagelength-lactModel\endcsname & \commarise{20} & 2 & 165&kb \\
+\reponame{LRM}{LRM} & \csname pagelength-LRM\endcsname & \commarise{125} & 8 & 2&Mb \\
+\reponame{manifold-ga}{manifold-ga} & \csname pagelength-manifold-ga\endcsname & \commarise{11} & 1 & \multicolumn{2}{c|}{---} \\
+\reponame{MetricSelectionFramework}{MetricSelectionFramework} & \csname pagelength-MetricSelectionFramework\endcsname & \commarise{44} & 4 & 236&kb \\
+\reponame{PENet}{PENet} & \csname pagelength-PENet\endcsname & \commarise{115} & 8 & 3&Mb \\
+\reponame{philter-ucsf}{philter-ucsf} & \csname pagelength-philter-ucsf\endcsname & \commarise{1987} & 13 & 32&Mb \\
+\reponame{PostoperativeOutcomes\_RiskNet}{PostoperativeOutcomes.RiskNet} & \csname pagelength-PostoperativeOutcomes.RiskNet\endcsname & \commarise{1} & --- & \multicolumn{2}{c|}{---} \\
+\reponame{SiameseChange}{SiameseChange} & \csname pagelength-SiameseChange\endcsname & \commarise{5} & 1 & 1&kb \\
+\hline \multicolumn{1}{|r|}{{\bf Average} ($N=11$)}&\makeAverage{\the\gitPages}{11}&253& 19& 84&Mb\\
+\hline \end{tabular}

### 2022-02-28 generated-summary-table.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
-\renewcommand\thesubsection{\thesection.\alph{subsection}}
-%\usepackage{hyperref}
-%\usepackage{refcount}
-\usepackage{url}
-\usepackage{doi}
-\usepackage{longtable}
-\usepackage{color}
-\usepackage{amsmath,amssymb}
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\input paper-seb-macros.tex
-
-\makeatletter
-      \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
-      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
-      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
-      }
-\makeatother
-
-\begin{document}
-
-\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}
-
-\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}
-
-%\orcid{0000-0003-2222-4243}
-
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
-\def\href#1{#1}
-
-\maketitle
-
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
-
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
-
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
-
-% start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
-
-\setcounter{section}{0}
-\def\n#1{{\sf\bfseries ({#1})}}
-
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
-
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
-
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
-
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
-
-%\newpage
-
-\makeatletter
-\let \bslash = \@backslashchar
-\makeatother
-
-%\renewcommand{\thesubsection}{\alph{subsection}}
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
-
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
-}
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
-
-\newpage
-\mysection{Further issues for Software Engineering Boards (SEBs)}
-
-\highlightmysubsection{Brief definition}
-Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
-
-Further details of the SEB proposal is in the main paper.
-Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:
-
-\begin{enumerate}\raggedright
-\item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
-
-\item
-When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}.
-Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
-How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
-
-\item 
-SEBs require policies on professional issues such as membership, transparency, and accountability.
-
-\item
-There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).
-
-\item
-Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs.
-For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?
-
-\item
-In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling.
-More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.
-
-\item
-SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
-
-\item
-Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large.
-Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
-\end{enumerate}
-There are other ideas to help make SEBs work, but it is clear they are part of the solution.
-We must not let perfection be the enemy of the good.
-SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.
-
-\mysubsection{Relationships of SEBs to Ethics Boards}
-
-\begin{enumerate}\raggedright
-\item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
-SEBs may also provide mentoring and training.
-
-\item
-Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
-
-\item 
-Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious).
-On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own.
-SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is.
-As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.
-
-\item
-SEBs should have a two-way collaboration with Ethics Boards.
-
-\begin{itemize}
-\item SEBs have to deal with ethical concerns, and how they may be implemented in code.
-One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.
-
-\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
-\end{itemize}
-
-\item
-Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
-\end{enumerate}
-
-\changemysubsection{SEBs are necessary but not sufficient}
-The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
-
-SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results.
-As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
-It needs doing.
-
-\mysection{Software engineering best practice}
-\label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
-
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
-
-The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible.
-Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer.
-Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}.
-Further suggestions for background reading can be found throughout this section.
-
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
-It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
-On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
-
-In practice, requirements and expectations change.
-Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
-In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
-
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
-
-{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy.
-To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe.
-Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}
-
-{The type and use of mathematics used in software engineering is formal methods.
-Not using formal methods ensures the resulting code is unsafe and unreliable.
-Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}
-
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
-Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
-Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.
-
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
-
-Many programming languages and programming environments have integrated features that support formal methods.
-For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@.
-In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
-Hoare is also a very good writer.
-
-Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way.
-Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy).
-Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do.
-Unfortunately, because formal methods are mathematical, few programmers have experience of using them.
-Fortunately tools are widely available to help use formal methods very effectively.
-
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
-Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
-The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code).
-Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.
-
-A special case of defensive programming appropriate for pandemic modeling is mixing methods.
-Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
-
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
-Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
-The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do.
-C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
-SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
-It is appreciated that the models may change and be adapted as new data and insights become available.
-Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable.
-Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used.
-Version control is supported by many tools (such as Git, Subversion, etc).
-
-If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
-Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
-
-{If results cannot be reproduced, has anything reliable been contributed?
-When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
-Reproducing the same results relying on the same codebase tells you little.
-The more independent reproductions of results the greater the evidence for belief in the implications.}
-
-Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes.
-Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.
-
-Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community.
-Open source is not a panacea, however; it raises many trade-offs.
-Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
-
-Open source raises important licensing and management questions to ensure the quality of contributions.
-A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository.
-This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.
-
-Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate.
-Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.
-
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
-In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
-Rigorous testing includes methods like fault injection.
-Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs.
-Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).
-
-It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again.
-For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).
-
-It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested.
-If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.
-
-There are many other important testing methods \cite{sommerville,knight,NVP}.
-
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
-Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
-
-{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories.
-One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
-}
-
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
-This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
-For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have?
-In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it.
-How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist.
-An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}
-
-There are many tools to help manage documentation (Javadoc, Doxygen, \ldots).
-Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}.
-Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.
-
-Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature).
-All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work.
-In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.
-
-A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
-
-Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project.
-Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.
-
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
-\globalLabel{supplementary-material-makefiles}
-Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
-Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team.
-Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.
-
-{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described.
-When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible.
-Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine.
-Makefiles typically also provide demo and test runs and other helpful features.
-Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}
-
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
-Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
-Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
-Indeed, reuse, scrutiny, and consensus are the foundations of good science.
-
-Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.
-
-A special case of reuse is to use software tools to help with software development.
-The tools (if appropriately chosen) have been carefully developed and widely tested.
-Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.
-
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
-When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
-This way lies over-fitting, a problem familiar from statistics (and machine learning).
-Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident.
-Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct.
-A simpler program is easier to prove correct, easier to program, and easier to debug.
-A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened.
-Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.
-
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
-
-To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}.
-However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate.
-Publications should then cite the  standards to which their computer models comply.
-
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
-
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
-
-As this long list illustrates, Software Engineering is a complex and wide-ranging subject.
-Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc).
-Multidisciplinary teamwork is essential.
-
-Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs.
-Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand).
-The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.
-
-Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.
-
-Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software.
-A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be.
-In short, users have to be brought into the software team too.
-
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
-Continuous Professional Development (CPD) is essential.
-
-Ironically, the more organized CPD the more likely the content itself will lag behind.
-There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees.
-Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.
-
-The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.
-
-Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
-
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
-Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
-
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
-In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
-
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
-The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
-
-This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
-
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
-
-A fragment of program code might say
-
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
-
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
-The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
-The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
-
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
-
-It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
-Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
-
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
-Data is then run on the virtual machine as program.
-Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions.
-Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program.
-There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal.
-In short, any distinctions between code and data are impossible to maintain.}
-
-In the present paper, we knowingly built on this blur between data and code.
-However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
-
-The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language.
-\LaTeX\ not only has text (as you are reading right now) but it also has code.
-For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted.
-When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters?
-The distinctions are all a bit moot.
-In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs.
-Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.
-
-The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above).
-As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data.
-A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer.
-In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.
-
-{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood).
-The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}
-
-Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.
-
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
-
-\def\questionsection#1{\changemysubsection{#1}}
-\def\question#1{\item \emph{#1\/}}
-\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}
-
-\questionsection{How trustworthy are the numbers?}
-\newcounter{resumeCounter}
-
-\begin{enumerate}
-\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.
-
-\sequestion{How rigorously has the software engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
-
-\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it.
-Were different methods used and compared, or was all confidence put into a single implementation?
-What internal consistency checks does the implementation have?
-Were invariants and assertions defined and checked?
-}
-
-\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.
-
-\sequestion{How are the claims presented that give us confidence in the code that they are based on?
-Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
-
-\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.
-
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the source?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed.
-Ask yourself, `Why does this source want me to hear this story?'
-
-\sequestion{The source of many science stories is the output of running some code.
-How reliable is this code?
-What evidence is there that the code was well-engineered so its reliability can be trusted?}
-
-\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.
-
-\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}
-
-\question{What am I not being told?} This is perhaps the most important question of all.
-Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.
-
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
-So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
-
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.
-
-\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}
-
-\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.
-
-\sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
-
-\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.
-
-\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}
-
-\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
-\end{enumerate}
-
-\newcount\temporary \temporary=\mainNumberOfReferences
-\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
-\advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
-
-\makeatletter    
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
-\makeatother
-
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
-\label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
-A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence and editorials were excluded.
-The sample represents what the editorial and the broader peer review community considers to be good practice.
-
-The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution.
-Indeed, all sampled papers used code in their research.
-Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable.
-It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).
-
-Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed.
-Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code.
-Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.
-
-There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
-
-The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted.
-Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey.
-What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized.
-Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
-
-As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.
-
-The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):
-
-\input generated-example-data.tex
-
-The data was written out by hand, after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
-(A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
-
-A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
-Hence, running the data generates the core human-readable information used in this paper.
-
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
-
-\begin{quote}
-\journalBreakdown 
-\end{quote}
-
---- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.
-
-An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:
-
-\begin{quote}
-\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
-\end{quote}
-
-it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data.
-The phrase above is now in fact written in \LaTeX\ in the paper as follows:
-
-\begin{quote}\tt\small
-Code repositories were used by \\
-\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
-\end{quote}
-
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
-
-Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
-
-The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-\highlightmysubsection{Detecting and defending against error}
-\label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
-% check it works for other numbers 
-%\countUsesVersionControlRepository=296
-
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
-\newcount\t
-\t=\countUsesVersionControlRepository
-\divide \t by 10
-\newcount\fraction
-\fraction=\countUsesVersionControlRepository
-\newcount\tent \tent=\t \multiply \tent by 10
-\advance\fraction by -\tent
-\the\t.\the\fraction,
-\t=\countUsesVersionControlRepository
-\advance \t by -1
-\the\t,
-\t=\countUsesVersionControlRepository
-\advance \t by 1
-\the\t, or
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
-
-\begin{itemize}\raggedright
-\item
-The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
-\divide \t by 10
-\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t) \cite{fixit}.
-
-\item
-The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
-\end{itemize}
-
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
-
-\newcount\min \min=5
-\newcount\max \max=20
-%\countUsesVersionControlRepository=4
-\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
-\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
-\ifnum \countUsesVersionControlRepository>\max
-	\else 
-		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
-		\fi
-\fi
-
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
-
-\highlightmysubsection{Assessments and criteria}
-
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
-
-Legend:\\
-\input generated-legend.tex
-
-\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
-\end{longtable}
-}
-
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
-
-\begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
-\raggedright
-
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
-
-\end{thebibliography}
-
-\immediate\closeout\infofile
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
-\end{document}
+\begin{tabular}{|rrrc|}\hline
+Number of papers sampled relying on code&32&100\%&\\\hline\hline
+\multicolumn{4}{|l|}{\textbf{Access to code}}\\
+Have some or all code available&12&38\%&\\
+Some or all code in principle available on request&8&25\%&\\
+No code available&12&38\%&\\\hline
+\multicolumn{4}{|l|}{{\textbf{Evidence of basic good software engineering practice}}}\\
+Evidence program designed rigorously&{0}&{0\%}&\\
+Evidence source code properly tested&{0}&{0\%}&\\
+\begin{change}Evidence of any tool-based development\end{change}&{\begin{change}0\end{change}}&{\begin{change}0\%\end{change}}&\\
+\begin{change}Team or open source based development\end{change}&{\begin{change}0\end{change}}&{\begin{change}0\%\end{change}}&\\
+Other methods, e.g., independent coding methods&{1}&{3\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
+Substantial code documentation and comments&2&6\%&\\
+Comments explain some code intent&3&9\%&\\
+Procedural comments (e.g., author, date, copyright)&10&31\%&\\
+No usable comments&17&53\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Repository use}}\\
+Code repository (e.g., GitHub) --- 1 was empty&10&31\%&\\
+Data repository (e.g., Dryad or GitHub)&9&28\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
+Papers published in journals with code policies&26&81\%&\\
+Clear breaches of code policy (if any) & 11&42\%&($N=26$)\\
+\hline\end{tabular}

### 2022-02-28 generated-supplementary-references.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
-\renewcommand\thesubsection{\thesection.\alph{subsection}}
-%\usepackage{hyperref}
-%\usepackage{refcount}
-\usepackage{url}
-\usepackage{doi}
-\usepackage{longtable}
-\usepackage{color}
-\usepackage{amsmath,amssymb}
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
+\vbox{\bibitem{ref-1}
+Callahan A,  Steinberg E,  Fries JA,  Gombar S,  Patel B,  Corbin CK and  Shah NH, ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp, 2020.
+DOI \texttt{10.1038/s41746-020-0300-0}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\input paper-seb-macros.tex
+\vbox{\maprepo{MetricSelectionFramework}\bibitem{ref-2}
+Kanzler CM,  Rinderknecht MD,  Schwarz A,  Lamers I,  Gagnon C,  Held JPO,  Feys P,  Luft AR,  Gassert R and  Lambercy O, ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp, 2020.
+DOI \texttt{10.1038/s41746-020-0286-7} {Code \url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\makeatletter
-      \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
-      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
-      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
-      }
-\makeatother
+\vbox{\bibitem{ref-3}
+Roy A,  Nikolitch K,  McGinn R,  Jinah S,  Klement W and  Kaminsky ZA, ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp, 2020.
+DOI \texttt{10.1038/s41746-020-0287-6}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\begin{document}
+\vbox{\bibitem{ref-4}
+Levine DM,  Co Z,  Newmark LP,  Groisser AR,  Holmgren AJ,  Haas JA and  Bates DW, ``Design and testing of a mobile health application rating tool,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp, 2020.
+DOI \texttt{10.1038/s41746-020-0268-9}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}
+\vbox{\bibitem{ref-5}
+Kannampallil T,  Smyth JM,  Jones S,  Payne PRO and  Ma J, ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp, 2020.
+DOI \texttt{10.1038/s41746-020-0278-7}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}
+\vbox{\maprepo{PENet}\bibitem{ref-6}
+Huang S,  Kothari T,  Banerjee I,   Chute C,  Ball RL,  Borus N,  Huang A,  Patel BN,  Rajpurkar P,  Irvin J,  Dunnmon J,   Bledsoe J,  Shpanskaya K,  Dhaliwal A,  Zamanian R,  Ng AY and  Lungren MP, ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0266-y} {Code \url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-%\orcid{0000-0003-2222-4243}
+\vbox{\bibitem{ref-7}
+Dhruva SS,  Ross JS,  Akar JG,   Caldwell B,  Childers K,  Chow W,  Ciaccio L,  Coplan P,  Dong J,  Dykhoff HJ,  Johnston S,  Kellogg T,  Long C,  Noseworthy PA,  Roberts K,  Saha A,  Yoo A and  Shah ND, ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
+\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
+Hofer IS,  Lee C,  Gabel E,  Baldi P and  Cannesson M, ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp, 2020.
+DOI \texttt{10.1038/s41746-020-0248-0} {Code \url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\def\href#1{#1}
+\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
+Norgeot B,  Muenzen K,  Peterson TA,  Fan X,  Glicksberg BS,  Schenk G,  Rutenberg E,  Oskotsky B,  Sirota M,  Yazdany J,  Schmajuk G,  Ludwig D,  Goldstein T and  Butte AJ, ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0258-y} {Code \url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\maketitle
+\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
+Choi D,  Park JJ,  Ali T and  Lee S, ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp, 2020.
+DOI \texttt{10.1038/s41746-020-0261-3} {Code \url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
+\vbox{\bibitem{ref-11}
+Hilton CB,  Milinovich A,  Felix C,  Vakharia N,   Crone T,   Donovan C,  Proctor A and  Nazha A, ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
+\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
+Li MD,  Chang K,  Bearce B,  Chang BY,  Huang AJ,  Campbell JP,  Brown JM,  Singh P,  Hoebel KV,   Erdo{\u g}mu\c{s} D,  Ioannidis S,  Palmer W,  Chiang MF and  Kalpathy-Cramer J, ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0255-1} {Code \url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip
 
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
+\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
+Hoffman JI,  Nagel R,  Litzke V,  Wells DA and  Amos W, ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp, 2020.
+DOI \texttt{10.1098/rsos.200419} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
+\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
+Gr\"onquist P,  Panchadcharam P,  Wood D,  Menges A,  R\"uggeberg M and  Wittel FK, ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp, 2020.
+DOI \texttt{10.1098/rsos.192210} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-% start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
+\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{onlyPaperWithChecks}
+Amos W, ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp, 2020.
+DOI \texttt{10.1098/rsos.191900} {Code \url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\setcounter{section}{0}
-\def\n#1{{\sf\bfseries ({#1})}}
+\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
+Gordon M,  Viganola D,  Bishop M,  Chen Y,  Dreber A,  Goldfedder B,  Holzmeister F,  Johannesson M,  Liu Y,  Twardy C,  Wang J and  Pfeiffer T, ``Are replication rates the same across academic fields?
+Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp, 2020.
+DOI \texttt{10.1098/rsos.200566} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
+\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
+Evans D and  Field AP, ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp, 2020.
+DOI \texttt{10.1098/rsos.200422} {Code \url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
+\vbox{\bibitem{ethics-paper}
+Beale N,  Battey H,  Davison AC and  MacKay RS, ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp, 2020.
+DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
+\vbox{\bibitem{ref-19}
+Cherevko AA,  Gologush TS,  Petrenko IA,  Ostapenko VV and  Panarin VA, ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp, 2020.
+DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
+\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
+Soczawa-Stronczyk AA and  Bocian M, ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp, 2020.
+DOI \texttt{10.1098/rsos.200622} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-%\newpage
+\vbox{\maprepo{lactModel}\bibitem{ref-21}
+Duruz S,  Vajana E,  Burren A,  Flury C and  Joost S, ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp, 2020.
+DOI \texttt{10.1098/rsos.200638} {Code \url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\makeatletter
-\let \bslash = \@backslashchar
-\makeatother
+\vbox{\bibitem{ref-22}
+de Azevedo EZD,  Dantas DV and  Daura-Jorge FG, ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp, 2020.
+DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-%\renewcommand{\thesubsection}{\alph{subsection}}
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
+\vbox{\maprepo{LRM}\bibitem{ref-23}
+Abdolhosseini-Qomi AM,  Jafari SH,  Taghizadeh A,  Yazdani N,  Asadpour M and  Rahgozar M, ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp, 2020.
+DOI \texttt{10.1098/rsos.191928} {Code \url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
-}
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
+\vbox{\maprepo{1}\bibitem{example-numerical-error}
+Webster J and  Amos M, ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp, 2020.
+DOI \texttt{10.1098/rsos.200307} {Code \url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\newpage
-\mysection{Further issues for Software Engineering Boards (SEBs)}
+\vbox{\bibitem{ref-25}
+Zhu Y-l,  Wang C-J,  Gao F,  Xiao Z-x,  Zhao P-l and  Wang J-y, ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp, 2020.
+DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\highlightmysubsection{Brief definition}
-Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
+\vbox{\bibitem{ref-26}
+Yu B,  Scott CJ,  Xue X,  Yue X and  Dou X, ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp, 2020.
+DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-Further details of the SEB proposal is in the main paper.
-Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:
+\vbox{\bibitem{ref-27}
+Joon-myoung K,  Younghoon C,  Ki-Hyun J,  Soohyun C,  Kyung-Hee K,  Seung D B,  Soomin J,  Jinsik P and  Byung-Hee O, ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\begin{enumerate}\raggedright
-\item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
+\vbox{\bibitem{ref-28}
+Zhu H,  Cheng C,  Yin H,  Li X,  Zuo P,  Ding J,  Lin F,  Wang J,  Zhou B,  Li Y,  Hu S,  Xiong Y,  Wang B,  Wan G,  Yang X and  Yuan Y, ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\item
-When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}.
-Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
-How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
+\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
+Fung R,  Villar J,  Dashti A,  Ismail LC,  Staines-Urias E,  Ohuma EO,  Salomon LJ,  Victora CG,  Barros FC,  Lambert A,  Carvalho M,  Jaffer Y A,  Noble JA,  Gravett MG,  Purwar M,  Pang R,  Bertino E,  Munim S,  Min AM,  McGready R,  Norris SA,  Bhutta ZA,  Kennedy SH,  Papageorghiou AT and  Ourmazd A, ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30131-X} {Code \url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\item 
-SEBs require policies on professional issues such as membership, transparency, and accountability.
+\vbox{\bibitem{ref-30}
+Sabanayagam C,  Xu D,  Ting DSW,  Nusinovici S,  Banu R,  Hamzah H,  Lim C,  Tham Y-C,  Cheung CY,  Tai ES,  Wang XY,  Jonas JB,   Cheng C-Y,  Lee ML,  Hsu W and  Wong TY, ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30063-7}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
 
-\item
-There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).
+\vbox{\maprepo{blast-ct}\bibitem{ref-31}
+Monteiro M,  Newcombe VF,  Mathieu F,  Adatia K,  Kamnitsas K,  Ferrante E,  Das T,  Whitehouse D,  Rueckert D,  Menon DK and  Glocker B, ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30085-6} {Code \url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
 
-\item
-Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs.
-For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?
-
-\item
-In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling.
-More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.
-
-\item
-SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
-
-\item
-Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large.
-Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
-\end{enumerate}
-There are other ideas to help make SEBs work, but it is clear they are part of the solution.
-We must not let perfection be the enemy of the good.
-SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.
-
-\mysubsection{Relationships of SEBs to Ethics Boards}
-
-\begin{enumerate}\raggedright
-\item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
-SEBs may also provide mentoring and training.
-
-\item
-Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
-
-\item 
-Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious).
-On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own.
-SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is.
-As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.
-
-\item
-SEBs should have a two-way collaboration with Ethics Boards.
-
-\begin{itemize}
-\item SEBs have to deal with ethical concerns, and how they may be implemented in code.
-One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.
-
-\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
-\end{itemize}
-
-\item
-Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
-\end{enumerate}
-
-\changemysubsection{SEBs are necessary but not sufficient}
-The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
-
-SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results.
-As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
-It needs doing.
-
-\mysection{Software engineering best practice}
-\label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
-
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
-
-The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible.
-Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer.
-Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}.
-Further suggestions for background reading can be found throughout this section.
-
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
-It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
-On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
-
-In practice, requirements and expectations change.
-Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
-In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
-
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
-
-{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy.
-To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe.
-Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}
-
-{The type and use of mathematics used in software engineering is formal methods.
-Not using formal methods ensures the resulting code is unsafe and unreliable.
-Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}
-
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
-Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
-Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.
-
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
-
-Many programming languages and programming environments have integrated features that support formal methods.
-For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@.
-In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
-Hoare is also a very good writer.
-
-Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way.
-Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy).
-Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do.
-Unfortunately, because formal methods are mathematical, few programmers have experience of using them.
-Fortunately tools are widely available to help use formal methods very effectively.
-
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
-Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
-The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code).
-Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.
-
-A special case of defensive programming appropriate for pandemic modeling is mixing methods.
-Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
-
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
-Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
-The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do.
-C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
-SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
-It is appreciated that the models may change and be adapted as new data and insights become available.
-Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable.
-Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used.
-Version control is supported by many tools (such as Git, Subversion, etc).
-
-If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
-Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
-
-{If results cannot be reproduced, has anything reliable been contributed?
-When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
-Reproducing the same results relying on the same codebase tells you little.
-The more independent reproductions of results the greater the evidence for belief in the implications.}
-
-Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes.
-Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.
-
-Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community.
-Open source is not a panacea, however; it raises many trade-offs.
-Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
-
-Open source raises important licensing and management questions to ensure the quality of contributions.
-A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository.
-This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.
-
-Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate.
-Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.
-
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
-In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
-Rigorous testing includes methods like fault injection.
-Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs.
-Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).
-
-It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again.
-For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).
-
-It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested.
-If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.
-
-There are many other important testing methods \cite{sommerville,knight,NVP}.
-
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
-Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
-
-{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories.
-One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
-}
-
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
-This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
-For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have?
-In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it.
-How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist.
-An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}
-
-There are many tools to help manage documentation (Javadoc, Doxygen, \ldots).
-Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}.
-Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.
-
-Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature).
-All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work.
-In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.
-
-A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
-
-Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project.
-Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.
-
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
-\globalLabel{supplementary-material-makefiles}
-Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
-Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team.
-Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.
-
-{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described.
-When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible.
-Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine.
-Makefiles typically also provide demo and test runs and other helpful features.
-Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}
-
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
-Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
-Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
-Indeed, reuse, scrutiny, and consensus are the foundations of good science.
-
-Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.
-
-A special case of reuse is to use software tools to help with software development.
-The tools (if appropriately chosen) have been carefully developed and widely tested.
-Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.
-
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
-When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
-This way lies over-fitting, a problem familiar from statistics (and machine learning).
-Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident.
-Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct.
-A simpler program is easier to prove correct, easier to program, and easier to debug.
-A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened.
-Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.
-
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
-
-To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}.
-However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate.
-Publications should then cite the  standards to which their computer models comply.
-
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
-
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
-
-As this long list illustrates, Software Engineering is a complex and wide-ranging subject.
-Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc).
-Multidisciplinary teamwork is essential.
-
-Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs.
-Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand).
-The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.
-
-Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.
-
-Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software.
-A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be.
-In short, users have to be brought into the software team too.
-
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
-Continuous Professional Development (CPD) is essential.
-
-Ironically, the more organized CPD the more likely the content itself will lag behind.
-There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees.
-Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.
-
-The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.
-
-Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
-
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
-Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
-
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
-In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
-
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
-The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
-
-This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
-
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
-
-A fragment of program code might say
-
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
-
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
-The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
-The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
-
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
-
-It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
-Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
-
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
-Data is then run on the virtual machine as program.
-Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions.
-Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program.
-There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal.
-In short, any distinctions between code and data are impossible to maintain.}
-
-In the present paper, we knowingly built on this blur between data and code.
-However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
-
-The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language.
-\LaTeX\ not only has text (as you are reading right now) but it also has code.
-For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted.
-When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters?
-The distinctions are all a bit moot.
-In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs.
-Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.
-
-The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above).
-As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data.
-A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer.
-In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.
-
-{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood).
-The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}
-
-Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.
-
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
-
-\def\questionsection#1{\changemysubsection{#1}}
-\def\question#1{\item \emph{#1\/}}
-\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}
-
-\questionsection{How trustworthy are the numbers?}
-\newcounter{resumeCounter}
-
-\begin{enumerate}
-\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.
-
-\sequestion{How rigorously has the software engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
-
-\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it.
-Were different methods used and compared, or was all confidence put into a single implementation?
-What internal consistency checks does the implementation have?
-Were invariants and assertions defined and checked?
-}
-
-\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.
-
-\sequestion{How are the claims presented that give us confidence in the code that they are based on?
-Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
-
-\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.
-
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the source?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed.
-Ask yourself, `Why does this source want me to hear this story?'
-
-\sequestion{The source of many science stories is the output of running some code.
-How reliable is this code?
-What evidence is there that the code was well-engineered so its reliability can be trusted?}
-
-\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.
-
-\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}
-
-\question{What am I not being told?} This is perhaps the most important question of all.
-Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.
-
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
-So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
-
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.
-
-\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}
-
-\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.
-
-\sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
-
-\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.
-
-\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}
-
-\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
-\end{enumerate}
-
-\newcount\temporary \temporary=\mainNumberOfReferences
-\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
-\advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
-
-\makeatletter    
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
-\makeatother
-
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
-\label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
-A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence and editorials were excluded.
-The sample represents what the editorial and the broader peer review community considers to be good practice.
-
-The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution.
-Indeed, all sampled papers used code in their research.
-Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable.
-It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).
-
-Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed.
-Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code.
-Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.
-
-There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
-
-The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted.
-Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey.
-What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized.
-Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
-
-As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.
-
-The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):
-
-\input generated-example-data.tex
-
-The data was written out by hand, after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
-(A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
-
-A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
-Hence, running the data generates the core human-readable information used in this paper.
-
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
-
-\begin{quote}
-\journalBreakdown 
-\end{quote}
-
---- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.
-
-An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:
-
-\begin{quote}
-\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
-\end{quote}
-
-it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data.
-The phrase above is now in fact written in \LaTeX\ in the paper as follows:
-
-\begin{quote}\tt\small
-Code repositories were used by \\
-\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
-\end{quote}
-
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
-
-Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
-
-The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-\highlightmysubsection{Detecting and defending against error}
-\label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
-% check it works for other numbers 
-%\countUsesVersionControlRepository=296
-
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
-\newcount\t
-\t=\countUsesVersionControlRepository
-\divide \t by 10
-\newcount\fraction
-\fraction=\countUsesVersionControlRepository
-\newcount\tent \tent=\t \multiply \tent by 10
-\advance\fraction by -\tent
-\the\t.\the\fraction,
-\t=\countUsesVersionControlRepository
-\advance \t by -1
-\the\t,
-\t=\countUsesVersionControlRepository
-\advance \t by 1
-\the\t, or
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
-
-\begin{itemize}\raggedright
-\item
-The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
-\divide \t by 10
-\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t) \cite{fixit}.
-
-\item
-The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
-\end{itemize}
-
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
-
-\newcount\min \min=5
-\newcount\max \max=20
-%\countUsesVersionControlRepository=4
-\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
-\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
-\ifnum \countUsesVersionControlRepository>\max
-	\else 
-		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
-		\fi
-\fi
-
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
-
-\highlightmysubsection{Assessments and criteria}
-
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
-
-Legend:\\
-\input generated-legend.tex
-
-\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
-\end{longtable}
-}
-
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
-
-\begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
-\raggedright
-
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
-
-\end{thebibliography}
-
-\immediate\closeout\infofile
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
-\end{document}
+\vbox{\bibitem{ref-32}
+Liu K-L,  Wu T,  Chen P-T,  Tsai Y M,  Roth H,  Wu M-S,  Liao W-C and  Wang W, ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30078-9}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip

### 2022-02-28 paper-seb-main.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-main.tex
@@
 \input plos-header.tex
-
-\renewcommand\thesubsection{(\thesection.\alph{subsection})}
-\usepackage{hyperref}
-
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\makeatletter
-    \def\do@url@hyp{\do/}
-    \usepackage{moreverb,url}
-    %    \renewcommand{\thefootnote}{\arabic{footnote}}
-\makeatother
 
@@
 \begin{document}
+\linenumbers
 
@@
 \begin{flushleft}
+\begin{change}
 {\Large
@@
 }
+\end{change}
 \newline
@@
 \bigskip
-
-% Current address notes
 
@@
 
-\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.
+\begin{change}\noindent\emph{Methodology:} This paper critiques the quality of software engineering practice, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions, and undertakes a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
+\end{change}
 
-\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.
+\noindent\emph{Results:} Scientists rarely assure the quality of code they rely on, rarely make complete code available, and rarely provide adequate documentation to understand or use their code reliably.
+\begin{change}Journals do not always have a code policy, and many have a relaxed approach.\end{change}
 
@@
 See his web site, \texttt{www.harold.thimbleby.net}, for more details.
-\linenumbers
-
-%\maketitle
 
@@
 We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
-An interesting overlap with statistical and software engineering scrutiny is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)\@.'' yet \emph{how\/} those analyses might have been performed is not discussed.
-The problem is that admitting using SAS 9.2 or any other named system does not really help scrutiny, as such systems can do anything with the data.
-A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is reliable.
+An interesting overlap with statistical and software engineering sloppiness is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA),'' yet \emph{how\/} those analyses might have been performed is not discussed.
+The problem is that admitting using SAS 9.2 or any other named system does not help scrutiny, as such systems can do anything with the data.
+A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is appropriate and suffciently reliable.
 
@@
 
-Contrary to the views expressed in the present paper, some authors have asserted that code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality}.
-On the contrary, if code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways.
+Contrary to the views expressed in the present paper, some authors have asserted that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality} --- code can also be used to analyze and critique scientific theories directly.
+If code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways.
 Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.
@@
 
+\begin{change}
+\subsection{The normal scientific emphasis on data}
+Data has been at the center of science, certainly since the earliest days of astronomy collecting planetary and other information.
+Today it is widely recognized that lack of accessible and usable data that has already been collected limits the progress of science.
+Low quality data and poor access to data causes reproducibility problems, an increasingly recognized problem --- in 2015 it was estimated that \$28B a year is spent on preclinical research alone that is not reproducible \cite{preclinical-reproducibility}.
+
+There are many current activities to proceduralize and standardize the more effective curation and use of data, such as the FAIR principles (Findable, Accessible, Interoperable and Reusable --- both for machines and for people) for scientific data management and stewardship \cite{fair}, and in the development of journal and national funder policies.
+For example, the 2022 update to the US National Institutes of Health data (not code) policies \cite{nih-policy} is described as a ``seismic mandate'' by \emph{Nature\/} \cite{nih-nature} in its attempt to improve reproducibility and open science.
+
+On the whole, these cost estimates and initiatives under-play or ignore the role of software as a critical part of at least some scientific endeavor: code has become the new laboratory for almost all science.
+The role of code specifically in modeling is discussed throughout this paper; without bespoke code, proposed models (unless very abstract) cannot make a quantifiable contribution to the literature.
+More generally, much data is embedded in code, and in the limit code and data are indistinguishable (see \supplement).
+Furthermore, code has additional problems of versions and compatibility beyond those of data, for example suitable compilers to run old code may no longer be available, and --- worse --- programming systems may silently produce different results when used on different computers.
+
+In general, without proper management of code --- for example to detect and report version control differences --- sharing code may even be counter-productive.
+(The data and code shared with the present paper includes cryptographic checksums; if somebody reproducing the work described here does not obtain the same checksums, then there are problems that need investigation before relying on the reproducibility of the data.)
+
+\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
+It is a fallacy that programming is easy \cite{fixit}. More correctly, toy programming is easy, but real programming is very difficult.
+
+An analogy helps.
+Building houses is very easy --- indeed, many of us have built toy Lego houses.
+Obviously, though, a Lego house is not a \emph{real\/} house.
+It is not large enough or strong enough for safe human habitation!
+This is obvious because we can see Lego houses.
+In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also invisible.
+Thus the ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed.
+It is like sincerely appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations.
+
+Certainly, building is much easier and faster when such technical details are ignored, as anyone who has experienced a cowboy build may attest.
+
+Unlike building (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a very new discipline, and the problems of poor programming are not widely appreciated.
+Relevant professional standards are not enforced.
+Problems for the reliability of science arise when legitimate doodling with software drifts into claiming scientific results that do not have the reliable engineering structures underpinning them, let alone the properly developed and documented archived code, to justify them.
+In many countries, you cannot even start to build without first having plans approved, but who writes plans for software?
+
+Since programming appears to be so easy, developing code has low status.
+The developers of code are rarely acknowledged in scientific papers.
+The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the technical contributors to it.
+While this view prevails, the vicious cycle is that the low status means software development is done casually, which reinforces the low status.
+
+In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code}, because code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.
+
+Scientists have been collecting and analyzing data for millennia, and curating data is taken seriously as a part of normal science and peer reviewed publication.
+Journal policies widely require appropriate discussion of data (much like they require appropriate discussion of statistics).
+Journals often require archiving data in standard formats so it can be accessed for further scientific work, such as reproduction.
+
+Unfortunately, very few scientists and even fewer journals realize that data and code are theoretically and in practice indistinguishable (see \supplement).
+Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling --- and the reasons for those policies --- should also apply at least as strictly to code.
+
+\subsection{The central role of code is ignored}\label{central-role-of-code}
+It is important that experiments and analysis, such as statistical analysis, are performed reliably and ethically.
+There are many protocols and journal policies that enforce good practice, for example a journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature.
+Yet PRISMA, like many policies, ignores the central role of computers, and ignores the software engineering principles that assure computation is reliable and reliably reported.
+
+There are many more styles of scientific investigation than systematic reviews, of course, but as PRISMA has had considerable development and is widely required by journals, it will serve as an example of overlooking the benefits of code in doing science.
+
+PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which sounds reasonable enough.
+PRISMA covers the review process carefully.
+For example, the authors should report the number of papers they included in their review.
+Perhaps $N=1093$.
+This number is then written into their draft paper, likely in several places.
+As the authors reads and revises their paper, submit it and respond to peer reviewers, it is likely that the number of papers included in the survey changes, or other details may change too.
+The authors now have a maintenance problem: where are the numbers that have changed, and what should they now be changed to?
+Doing a search-and-replace, whether automated or by hand, is fraught with difficulties.
+What happens if $1093$ is used for some other purposes as well?
+What happens if some of the $1093$ values were written as $1,093$ and are not noticed?
+Then there are the Human Factors, where slips and errors will happen in this process anyway.
+Similar iterative revision cycles happen with any paper, not just with systematic reviews.
+Typos, slips during cut-and-paste, and other errors are common.
+They are problems across all science.
+
+If a computer program is involved in the process (it generally is for a systematic review) then the value of $N$ could easily be stored in a file where the paper typesetting process can access it.
+For example, if \LaTeX\ is the system of choice, the analysis could generate, say,
+
+\begin{center}\texttt{$\backslash$newcommand\{$\backslash$N-papers-reviewed\}\{1093\}}\end{center}
+
+so that when and wherever the author writes \texttt{$\backslash$N-papers-reviewed} in their paper's text, the typeset paper says 1093 or whatever the correct value happens to be at the time.
+If so, then \emph{whenever\/} the survey is updated, the value cited in the paper is \emph{correctly\/} updated without any further intervention from the author.
+
+In general, not just numbers, but any data, text, graphs or tables, etc, from the analysis can be reliably inserted into a paper automatically.
+
+A special case of this process is the automatic handling of bibliographies and citations: what used to be a tedious manual process is now very commonly automated and for exactly the same reasons, that is, simultaneously improving ease of use and reliability.
+Note that automatic bibliographies insert both free text and citation numbers: data values are arbitrary.
+Moreover, automatic bibliography systems also provide many useful validation checks on the data, for instance that a journal article citation actually specifies a journal name.
+
+Increasingly, many analysis systems have themselves introduced features to automate keeping papers up to date and for reliably referring to correct values (what Software Engineers would call maintaining invariants with the data).
+R~(e.g., using RMarkdown) and Mathematica are popular examples of such general purpose analysis systems that provide powerful features to help authors more easily produce reliable papers.
+In particular, they make it completely trivial to read and edit a paper, revise the analysis, redraw tables and diagrams etc, and obtain a correctly revised paper with no further manual work.
+Crucially, this encourages the author to explore the analysis and competing analyses more than usual, so their confidence in the final analysis and quoted results is much higher than usual.
+Furthermore, RMarkdown and Mathematica also have powerful mechanisms to create slide shows, and therefore the guarantees of automatic correctness transfer to presentations about the research as well.
+
+An important consequence of using a code-based automated process is that it removes barriers to revising and improving papers.
+Before such methods were used, it was tedious and risked introducing new errors if one undertook to update research and papers.
+When the process is automated, however, there are no such barriers.
+
+Of course errors may still occur, but there is an important and dramatic improvement.
+Each time a paper is processed, the \emph{same\/} code is re-used.
+So each revision and proof reading of the paper helps detect those errors, and once any error is detected and corrected it will not arise again.
+Moreover, since the same potential error, if any, is repeatedly proof read and checked (it occurs in each iteration of the paper, and perhaps in many places in each version of the paper), this reduces the chances that the error is missed.
+This is a very different situation to the manual processing of data, where every time a paper is edited or revised, the data may be corrupted accidentally in new ways and in new places.
+
+The present paper performed a \ifnum \dataN<100 \textbf{small}\fi\ pilot survey of \textbf{\plural{\dataN}{paper}} sampled from \textbf{\plural{\numberOfJournals}{journal}}, \textbf{\the\counthasDataRepository} of which used code repositories.
+The repositories were most recently accessed on \textbf{\clonedate} for automatic analysis for generating the summary data reported in this paper.
+All those facts, highlighted in bold here but also used throughout this paper and in the \supplement, are ensured correct by the author using an automatic process (discussed in full in the \supplement).
+One should note, as emphasized above, that as these specific values as used above also appear in many places and in many different contexts in the present paper, this significantly reduces the Human Factors problems of noticing errors.\footnote{\begin{change}For example, the author types 1093, but makes a transposition slip and writes 1039.
+Since the author thinks they know what they wrote, when they proof read 1039 they misread it (this is a very common problem with misspellings and grammatical ambiguities).
+These errors (transposition and misreading, etc) do not happen consistently each time, so the more often 1039 occurs in the paper \emph{if generated automatically}, the more likely the errors will be addressed.
+See \cite{enigma} for an actual example.\end{change}}
+
+PRISMA says nothing about how to ensure the final results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations.
+Ironically, many journal publishing policies (including \emph{PLOS ONE}\/) forbid using such techniques (see the \supplement\ for further discussion).
+Again, such rules reinforce the fallacy that code is trivial and unimportant.
+\end{change}
+
 \subsection{Bugs, code and programming}\label{knowledge}
@@
 This common practice ensures that published data is very rarely all of the data because it omits the magic numbers.
-This emphasizes the need for repositories to require the inclusion of code so all data is actually available.
+This emphasizes the need for repositories to require the inclusion of code so all data, \begin{change}including that embedded in the code\end{change}, is actually available.
 
-Bugs can be understood as discrepancies between what code is intended to do and what it actually does.
+Bugs can be understood as discrepancies between what code is ought to do and what it actually does.
 Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered.
 Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination.
-None of the code examined for this paper includes either.
+None of the code examined for this paper includes any such techniques.
 
 If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
-Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things.
-For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the sense of producing incorrect results --- that is, what was wanted was wrong.
+Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time).
+For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the wider sense of producing incorrect results --- that is, what was intended was wrong.
+
+Random numbers are widely used in computational science, e.g., for simulation or for randomizing experiments.
 Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.
 
-\iffalse Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, unqualified programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering.
+\begin{change}
+\subsection{Human Factors of programming}
+People would generally not make coding errors if there were aware they were making errors.
+Unfortunately programming is a very demanding activity, which causes tunnel vision (also known as loss of situational awareness).
+The consequence is programmers focus on ``the'' problem as it appears in the code, and are unaware they are not considering wider issues.
+The correctness, generality, and usability of a program are therefore often unintentionally sacrificed to making code work at all.
+
+Confirmation bias is a standard Human Factors problem \cite{fixit} (which scientists have many techniques to avoid), which encourages us to perform tests that show our programs work.
+Instead, we should be rigorously testing ways in which programs can fail.
+This is exactly the same issue pointed out by Popper \cite{popper}: scientists should experiment to find reasons why hypotheses are false, and indeed use simple hypotheses that are testable.
+Software is really no more than a collection of sophisticated hypotheses.
+
+Standard Human Factors mitigations for these problems include team working, with appropriate precautions to manage authority gradients (where the Human Factors oversights of the leader influence the team).
+Many computerized mitigations are also available --- strong typing, code analyzers, formal methods, and so on.
+See the \supplement.
+
+Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering.
 Dunning and Kruger go on to say,
@@
 If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.
+Moreover, if we do not work in a diverse team, nobody will ever point this out, so the potential problems it causes will never be addressed.
 
@@
 Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable.
-Programs can be written so that it is not possible to determine what they do or how they do it (whether deliberate obfuscation or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
+Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
 The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
-\fi
+\end{change}
 
@@
 \input generated-covidsimSummary.tex
-kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
+kLOC (thousands of lines of code) written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
 Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done.
@@
 
-\subsection{Science beyond epidemiology}
-Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims.
+\section{Science beyond epidemiology}
+\label{section-science-beyond-epidemiology}
+\begin{change}
+Epidemiology has a high profile because of the current COVID pandemic.
+There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact.
+Climate change modeling is one such example.
+
+A short 2022 summary of typical problems of software engineering impacting science generally appears in \emph{Nature} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration.
+The paper \cite{nature-review} unfortunately makes some misleading comments about the simplicity of software engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''
+
+Guest and Martin \cite{psychological-modeling} in another 2022 paper promote the use of computational modeling, arguing that through writing code, one debugs scientific thinking.
+Psychology, Guest and Martin's focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}.
+In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived.
+Computational models can even be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks.
+It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely (e.g., ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973 and is now a 120 kLOC Common LISP and Python system \cite{actr}; and of course any paper using ACT-R would require additional code on top of the basic framework).
+
+The psychology paper \cite{psychological-modeling} demonstrates building an example computational model from scratch to illustrate their own framework of computational science.
+In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not explored.
+Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.
+
+The Python code is presented without derivation or discussion, as if software engineering is trivial.
+The program listed in the paper certainly runs without obvious problems (ignoring typographical errors due to the journal's publishers), but ironically the Python does not implement the mathematical specification explicitly provided for it,\footnote{\begin{change}More precisely: the program has a bug, and/or the specification given is wrong or too abstract.\end{change}} thus unintentionally undermining the argument of the paper.
+
+One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses.
+Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct (at least when finished).
+Otherwise, computational modeling with inadequate software engineering will very likely introduce errors into scientific thinking.
+\end{change}
+
+Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims.
 People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place.
@@
 \input generated-summary-table.tex
-%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
 \end{center}
@@
 
-\input generated-clone-date.tex
 \begin{center}
@@
 \end{center}
-\newcount \pubdelayinmonths
-\pubdelayinmonths=\cloneyear
-\advance \pubdelayinmonths by -2020
-\multiply \pubdelayinmonths by 12
-\advance \pubdelayinmonths by -4 % published in April
-\advance \pubdelayinmonths by \clonemonth
-\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
+
+\caption{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
 Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting.
 Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
-Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the\pubdelayinmonths\ months had already elapsed since the paper referring to the code was published.}
+At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.}
 \label{table-repo-summary}
@@
 
-\section{A pilot survey of peer-reviewed research relying on code}
+\subsection{A pilot survey of peer-reviewed research relying on code}
 \label{survey-section}
@@
 Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized.
-We therefore undertook a survey of recent papers in leading peer reviewed journals.
+\begin{change}We therefore undertook a pilot survey of papers in leading peer reviewed journals.\end{change}
 
-\subsection{Methodology}
-A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.
+\subsubsection{Methodology}
+\begin{change}
+There are many ways to do good science.
+Normal science routinely involves either performing new research and reporting it in the literature, or performing reviews --- such as systematic reviews --- of the existing literature to form a consensus view of an existing body of research.
+Typically, research offers new theories or results, points out previous errors, or uses statistical or qualitative methods to reduce the variability in previously reported results.
+In any case, the intention is to provide reliable knowledge and to bolster consensus \cite{ziman}.
+
+In the present case, we have identified limitations in the methodology of earlier scientific contributions, throwing serious doubt on their soundness and reliability; furthermore, we identified critical misunderstandings in the literature citing those papers, including misunderstanding the nature of code reproducibility.
+(Shortly, we will present some ways forward, in section \ref{section-discussion}.)
+
+It is possible, if not likely (e.g., given the report in \emph{Nature\/}, cited above \cite{nature-review}), that such limitations are more widespread than the specific examples considered so far, which are anyway limited primarily to epidemiology, and to COVID research performed under extreme pandemic pressure.
+We now therefore undertake a small, manageable stratified sample of peer reviewed literature to explore whether the issues are more widespread.
+\end{change}
+
+A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected from the leading journals \journalBreakdown.
 
 The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal.
-All papers sampled are Open Access, although for some papers some or all of the associated data has restricted access.
+All papers sampled are Open Access, although for some papers some or all of the associated data has \begin{change}no or\end{change} restricted access, \begin{change}in some cases despite the relevant journal policies on code\end{change}.
 Table \ref{table-overview} is an overview of the sample.
 
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
+Papers were selected from the journals' July 2020 \begin{change}then\end{change} new online listings where the paper's title implied that code had been used in the research.
 Commentary, correspondence, and editorials were excluded.
 The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.
+The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on and assumed the quality of code used in their research.
 
-The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.
+\begin{change}
+This convenience sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers,\footnote{\begin{change}Using Google Scholar it is estimated that over 40,000 papers meeting the title criteria were published in the month of July 2020.\end{change}} but~\ldots
 
-The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.
+\begin{enumerate}
+\item the selected journals are leading peer-reviewed scientific journals that set the standards for scientific publishing practice generally (although the sample shows that code policies are not always enforced);
 
-%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).
+\item as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been productively more insightful (this view is consistent with the multi-disciplinary reports in \cite{nature-review}, mentioned in section \ref{section-science-beyond-epidemiology});
 
-The \the\dataN\ papers cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install.
-Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment was made whether the code provided actually reproduced a paper's specific claims.
-Indeed, if we trust the papers that their code was actually run and provides the results as reported, then merely running their code (when provided in full) merely checks the paper/code consistency, but will not assess the quality or reliability of the code.
+\item the survey is not intended to be a formal, systematic sample of scientific research in general, but is intended to be sufficient to dispel the possibility that the issues described above earlier in this paper are isolated practice unique to a few papers in epidemiology, perhaps an idiosyncrasy of a few authors in a particular field, or perhaps due to an initial chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they might have just happened to be software engineering outliers);
 
-%Assessing code quality has been a lively topic in software engineering for decades, and has international standards (ISO/IEC 9126, updated to ISO 25000).
-Here, code is assessed\footnote{The author of this paper has been assessing undergraduate and postgraduate computer science student code since the 1970s, using double marking and moderation.} using the n criteria of \cite{software-quality} as a marking scheme with scores out of 100/n.
+\item the code/data policies of the \plural{\numberOfJournals}{journal} condoned at the time of the sample \emph{and continue to condone\/} poor practice at the time of writing the present paper (\monthname{\month} \the\year) --- for specific details and further explanation of the problems, see \supplement\ section \ref{supplementary-journal-policies-section};
 
-The full methodology, data and analysis, as well as many software engineering and Human Factors techniques to help improve reliability and reproducibility used in the preparation of the present paper, are provided in the \supplement, which is also available from a repository (see details at the end of this paper).
+\item
+the fact that the specifically identified problems are elementary errors in software engineering (see the discussion in section \ref{section-discussion}) suggests more sophisticated analysis is not required;
+
+\item 
+finally, the present paper's \LaTeX\ source, as well as all documented code and data, are available from a repository, which provides a convenient framework for easily refining or developing the research as may be desired (see details at the end of this paper).
+\end{enumerate}
+\end{change}
+
+The \the\dataN\ papers surveyed cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install.
+Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment \begin{change}could realistically be\end{change} made whether the code provided actually reproduced a paper's specific claims.
+Indeed, if we trust the papers that their code was actually run and provides the results as reported, then running their code (when provided in full) would merely check the paper/code consistency but will not assess the quality or reliability of the code.
+\begin{change}Indeed, in most scientific papers there are layers of expert scientific work, interpretation and abstraction, lying between the computational models and the report in the paper.\end{change}
+
+The present paper's full methodology, data and analysis, as well as many \begin{change}rigorous\end{change} software engineering and Human Factors techniques to help improve reliability and reproducibility used in its preparation, are provided in the \supplement.
 
@@
 
-\subsection{Summary of results}
+\subsubsection{Summary of results}
+\label{summary-of-results}
 The sample selection criteria necessarily identified scientific research with software engineering contributions.
+
 No evidence of verification and validation was seen.
-There was only one example of basic software engineering methods, namely independent coding.
+There was only one example of very basic software engineering methods, namely independent coding, and even then the independent code used for testing was not uploaded to the paper's code repository, so the independent testing is not available for reviewers or readers of the paper.
+(See \supplement\ for more details.)
+
 There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer --- who writes high quality code to start with!\ --- would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
-(See the \supplement\ for definitions and a list of standard methods.)
+No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
+(See the \supplement\ for a list of standard methods.)
 
-Much code depended on substantial manual intervention to compile it.
-All code (where actually provided) was sufficiently complex that, if it was supposed to be used or scrutinized, required more substantial documentation than was provided.
+Much code depended on specific software versions, specific libraries, and substantial manual intervention to compile it.
+All code (where actually provided) was sufficiently complex that, if it was to be used or scrutinized, required more substantial documentation than was provided.
 
-On the whole, on the basis of the sample evidence, scientists do not make their code available, and rarely provide adequate documentation (see table \ref{table-summary}).
+On the whole, on the basis of the sample evidence, scientists do not make their code \emph{usably\/} available, and rarely provide adequate documentation (see table \ref{table-summary}).
 
-It is noteworthy that the papers themselves, typically a few published pages, are very small compared to the code they rely on; the papers represent much less authoring effort than the code and data that they rely on for their claims and correctness (see table \ref{table-repo-summary}).
-With one minor exception (which may have been accidental), no papers reported anything on any software engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
+With the one minor exception, no papers reported anything on any software engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
+The papers themselves, typically only a few published pages, are very brief compared to the substantial code they rely on (see table \ref{table-repo-summary}).
 
-None of the papers used any specific software engineering methods, such as open source \cite{open-source} and other standard methodologies provided in the \supplement, to help manage their processes and help improve quality.
+With the one exception, none of the papers used any specific software engineering methods, such as open source \cite{open-source} or other standard methodologies provided in the \supplement, to help manage their processes and help improve quality.
 Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reason this paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}
 
-\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
+\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual publishing practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
 In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach.
-The implication of these results is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies, or they are technically unable to assess them --- and, if so, they also fail to say they are unable to assess them.
+The implication is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies --- they (or their editors and reviewers) are technically unable to assess them.
 This lack of expertise is consistent with the limited awareness of software engineering best practice that is manifest in the published papers (and resources) themselves.
 
-Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \input generated-clone-date.tex
-while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could not been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
+Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \clonedate\
+while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could never have been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
 
-Sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
+The sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
 
@@
 \section{Discussion}
-Effective access to quality code is not routine in science.
+\label{section-discussion}
+\begin{change}
+Effective use or access to quality code is evidently not routine in science. 
+\end{change} 
 Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
@@
 
+\begin{change}
+\subsection{Action must be interdisciplinary}
+\setcounter{footnote}{1}
+Code is only part of science, and only one critical factor in the wider reproducibility crisis: SEBs must work with --- and be engaged by --- other reproducibility initiatives.
+
+Relying on Software Engineering Boards (SEBs) \emph{alone\/} would continue one of the current besetting problems about the role of code.
+The conventional view is that scientists do the hard work compared to the ``easy'' coding work\footnote{The programming is easy fallacy is refuted in sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}.} so they just need to tell programmers what to do.
+This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, which he explores at some length, is that we need to spend more effort in working out what computers should do (i.e., do the science better) and then just tell programmers to do \emph{that}.
+
+On the contrary, competent software engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and often that logic needs refining or optimizing.
+In other words, the software engineers can bring important insights back into the science, hence improving or changing the questions and assumptions the science relies on.
+This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but that is ill-conditioned, there is no good answer.'' Ideally, then, it is not just a sequential process of science $\rightarrow$ code $\rightarrow$ results, but an iterative cycle of collaboration and growing mutual understanding.
+
+In short, the way SEBs work and are used is crucial to their success.
+Software engineers can help improve the science, so it is not just a matter of asking a SEB  whether some coding practices (like documentation) are satisfactory, but whether the SEB has insights into the science itself too.
+Most effectively, this requires interdisciplinary working practices (science plus software engineering) with mutual respect for their contributing expertise.
+
+\subsection{SEBs could be enforced}
+It is routine for funders and journals to require appropriate ethical processes and ethical statements, typically specifying data security and confidentiality, appropriate handling of vulnerable participants, consent, and so on.
+It would be easy for funders and journals to require equivalent types of statements on software engineering.
+
+\end{change}
+
 \subsection{Suggestions for further work}
-
 Although this study of software engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
@@
 We need to improve the quality of software engineering that supports science.
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed in shows that current coding practice makes for poor science.
-In the current pandemic, scientific modeling, such as track and trace \cite{excel-fiasco}, balancing COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.
+While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed shows that current coding practice makes for poor science.
+In a pandemic, scientific modeling, such as epidemiological modeling, track and trace \cite{excel-fiasco}, modeling COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.
 
@@
 We know from software research than ordinary programming is very buggy and unreliable.
-Simply put, without quality code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
+Without adequately specified and documented code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
 Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion:  computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper}.
@@
 \def\csref{\global\advance\csrefcount by 1}
-\def\ethics#1{\paragraph*{Ethics}#1}
-\def\ack#1{\paragraph*{Acknowledgments}#1}
-\def\dataaccess#1{\paragraph*{Data and code access}#1}
-\def\aucontribute#1{\paragraph*{Author contribution}#1}
-\def\competing#1{\paragraph*{Competing interests}#1}
-\def\funding#1{\paragraph*{Funding}#1}
+\long\def\ethics#1{\paragraph*{Ethics}#1}
+\long\def\ack#1{\paragraph*{Acknowledgments}#1}
+\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
+\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
+\long\def\competing#1{\paragraph*{Competing interests}#1}
+\long\def\funding#1{\paragraph*{Funding}#1}
 
@@
 
-\ethics{This article presents research with ethical considerations but does not fall not within the usual scope of ethics policies.}
+\ethics{This article presents research with ethical considerations but does not fall within the usual scope of ethics policies.}
 
 \dataaccess{There is a full discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}.
+The \supplement\ also presents all raw data in tabular form.
 All material is available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards} (which has been tested in a clean build, etc).
 The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \input generated-data.js-errors.tex
-possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset results reliably in this paper and its \supplement\ (e.g., table \ref{table-summary} in this paper) \emph{directly\/} from the data.
-In addition, a standard CSV file is generated in case this is more convenient to use, for instance to browse in Excel or other spreadsheet application.
-All data, analysis, and details of the methods used can be found in the \supplement.}
+JavaScript code, conveniently in the same file as the JSON data, checks (with \the\JSONerrorCount\ possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset all results reliably in this paper and in its \supplement\ \emph{directly\/} from the automatic data analysis (e.g., see table \ref{table-summary} in this paper).
+\begin{change}The \supplement\ describes how \emph{PLOS ONE\/} and other journal publishing policies unfortunately do not support such automatic approaches, undermining the benefits of reliable reproducibility.\end{change}
+
+In addition, a standard CSV file is generated in case this is more convenient to use, for instance to browse in Excel or other spreadsheet application.}
 
@@
 
+%\label{LastPage}
+%save Latex register    {linenumber}
+\makeatletter
+\immediate\write\@auxout{\string\def\string\MainPaperLastLine{\arabic{linenumber}}}
+\makeatother
+\end{document}
+
+\def \allInOne {define this so we don't include header in supplemental material}
+\input paper-seb-supplementary-material.tex
 \end{document}
@@
 
-%\def \allInOne {don't include header in supplemental material}
-%input paper-seb-supplementary-material.tex
-
-\label{LastPage}
 \end{document}

### 2022-02-28 paper-seb-supplementary-material.tex
All files updated for PLOS Feb 28 2022 submission

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
+\iftrue
     \PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
+    \documentclass[10pt,letterpaper]{article}
+
+    \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+    % Use adjustwidth environment to exceed column width (see example table in text)
+    \usepackage{changepage}
+
+    % line numbers
+    \usepackage[right]{lineno}
+    \linenumbers
+
     \renewcommand\thesubsection{\thesection.\alph{subsection}}
@@
     \usepackage{amsmath,amssymb}
+    \usepackage{fancyhdr}
     \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
@@
     \input paper-seb-macros.tex
+\fi
 
@@
       {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
+      % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
       \newcount\mainNumberOfReferences \mainNumberOfReferences=0
@@
       \input paper-seb-main.aux 
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
+      \resetlinenumber[\MainPaperLastLine]
       }
+
 \makeatother
@@
 
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
 \def\href#1{#1}
@@
 \maketitle
+\pagestyle{fancy}
+
+% pageref gives us a hyperlink which looks unsightly
+\makeatletter
+\def\rememberpage{\immediate\write\@auxout{\bslash gdef\bslash SupplementalLastPage{\thepage}}}
+\makeatother
+
+\rfoot{\supplement\ \thepage/\SupplementalLastPage}
+\lfoot{\today}
+\cfoot{}
+\lhead{}
+\rhead{}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\thispagestyle{empty}
 
@@
 % as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
+\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
 }
@@
 % start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
+\setcounterpageref{page}{LastPage}
+\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
 
@@
 \def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
+\noindent
+\input generated-supplementary.toc
 }
+
 \newwrite\contentsfile
@@
 
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
+\vbox{\noindent
+      \input generated-minicontents.tex
       }
@@
 Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
+
+\begin{change}Interestingly, the only paper reviewed that claimed to do any independent testing \cite{onlyPaperWithChecks} failed to include any testing in its data or code repository, so the testing itself --- the essential quality assurance of the code --- is not open to scrutiny (e.g., the code and the ``independent'' code are likely to contain common code, data, and common bugs).\end{change}
 
@@
 
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
+\begin{change}
+All computer systems are in principle equivalent to Turing Machines, and Turing Machines make no distinction between program and data.
+It is possible to define Turing Machines that do separate program code and data, but as soon as a Universal Turing Machine is constructed, its data \emph{is\/} code.
+Indeed, Universal Turing Machines are a theoretical abstraction of virtual machines, which are used widely in practical computing.
+Java, for instance, runs in a virtual machine, so any Java program code (and any data it uses) is in fact merely \emph{all\/} data to the Java virtual machine.
+At another extreme, $\lambda$-calculus is purely program source code, yet $\lambda$-calculus is equivalent to Turing Machine computation.
+Therefore, even the ``pure'' programs of $\lambda$-calculus also represent data.
+\end{change}
+
+These elementary theoretical considerations underly an important practical fact: there is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
+
+There is no code/data distinction one can imagine that cannot easily, even accidentally, be circumvented.
 In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
@@
 This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
+Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs and irreproducibility are the inevitable (and confusing) consequence.
 
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
+The main paper points out that data is often embedded in code using ``magic numbers.'' Let's now explain how.
 
-A fragment of program code might say
+A simple fragment of program code might say
 
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
+\newcount \magicNumber \magicNumber = 324
+\begin{center}\texttt{x = \the\magicNumber+sin(theta*pi/180);}\end{center}
 
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
+This is clearly all source code, but the number \the\magicNumber\ above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
 The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
+Some programming languages like Mathematica treat $\pi$ as an exact mathematical constant (e.g., Mathematica calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
+\url{www.fhgr.ch/en/specialist-areas/applied-future-technologies/davis-centre/pi-challenge}}
+
 The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
 
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
+Unfortunately, there is data that is extremely easy to overlook (and therefore very hard to manage) because it is embedded in arbitrary ways in code.
+You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of the $\pi$ in the expression, you assume \texttt{theta} is degrees and \texttt{sin} is taking radians as its parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
+Confusingly, even if it calculates sines, it is generally a different function when the code is run on a different computer producing numbers that are not exactly the same.
 
 It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
+
+\begin{change}
+\subsection{When magic numbers become magic code}
+Data often controls the flow of code. For example, data summarizing patients may include their gender, but the program processes males and females differently. Then data becomes code.
+
+Arbitrary numbers appearing in code are obviously magic numbers, but code often conceals the magic numbers of data by ``programming them away'' during the coding process.
+
+For example, the magic number \the\magicNumber\ was explicit in the line of code shown above, but if somewhere else the program says
+
+\begin{center}\texttt{if evenQ(\the\magicNumber) then A; else B;}\end{center}
+
+many programmers would optimize this to \texttt{A}, because they know the condition is true because of their assumptions.
+This now seems to be a more efficient program because it has avoided a test (which a modern complier would have optimized away anyway).
+Unfortunately, the previously explicit dependency of the code on the magic number \the\magicNumber\ has completely disappeared.
+
+Obviously this example seems trivial, but it illustrates that programmers do some of their programming while writing code, and many assumptions disappear completely and have no representation in the final code.
+More complex code will have many facts hard wired into the code --- so in fact the code contains data.
 Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
 
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
+This is one reason bugs --- effectively incorrect assumptions --- are so hard to find, because they have no concrete form in the final program.
+
+\subsection{When data is code}
+{Many computer programs blur the simplistic code/data distinctions deliberately, to create virtual machines.
 Data is then run on the virtual machine as program.
@@
 
+AI and Machine Learning are further examples of exploiting data as code.
+Typically a program learns from a training set of data, and then processes future data differently depending on what it has learned.
+In other words, the original data becomes a model which is now code.
+
+\subsection{Exploiting code as data for more reliable science}
+\end{change}\noindent
 In the present paper, we knowingly built on this blur between data and code.
@@
 What evidence is there that the code was well-engineered so its reliability can be trusted?}
+
+\begin{change}\sequestion{What evidence is there of rigorous (e.g., code review and tool-based) independent methods being used to manage coding bias?}\end{change}
 
@@
 
+\begin{change}
+Although the pilot survey is not a systematic review, following Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) \cite{prisma} it is good practice to disclose details of the reviewers.
+In the present case, the selected papers were assessed by the author of the present paper.
+The study was not blinded.
+This is, of course, a limitation of the study.
+However, the reviewer is a full professor of computer science, who has taught and assessed computer software since the 1970s, using moderated and peer-reviewed processes for undergraduate and postgraduate computing degrees, and is well aware assessing code quality has been a lively topic in software engineering for decades (there is now international standard ISO/IEC 9126, updated to ISO 25000).
+The author has written legal documents analyzing software for criminal cases involving faulty software.
+The author has approximately
+\newcount \papers \papers=340 % peer reviewed
+\advance \papers by 188 %invited
+\the\papers\ published papers in computer science.
+
+The evaluations performed for the present paper are at a trivial level where sources of bias should have a negligible effect, particularly given that the overall conclusion is consistent across both the diverse sample and the computational science-based papers cited in the paper that did not form part of the selected sample.
+As said in the body the paper: \emph{the fact that the specifically identified problems are elementary errors in software engineering (see the discussion in \emph{[main paper]\/} section \ref{section-discussion}) suggests more sophisticated analysis is not required}.
+
+In any case, as stated in the main paper, the full dataset and analysis code is available at \begin{center}\url{github.com/haroldthimbleby/Software-Enginering-Boards}\end{center} and the reviewed papers are (unless retracted) still available online for independent assessment.
+%The data is also provided in full later in this \supplement.
+\end{change}
+
 There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
+Different programming languages also develop ``cultures'' and tool-based systems that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
+For scientific code, however, the explanatory role is critical, and this is what was assessed in the present survey.
+It is notable that no such tool-based approach to code or documentation was used in any code reviewed.
 
@@
 Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
+It is more like ``can we see the critical pages from your lab book so we understand what you did?''\ rather than ``can we have a free run of your laboratory, even though we don't understand the details of the science?''
 
@@
 
+\newdimen \codeIndent
+\codeIndent = 1em
 \input generated-example-data.tex
 
-The data was written out by hand, after reading and reviewing each paper in the survey.
+The data was entered by hand (as JSON terms), after reading and reviewing each paper in the survey.
 In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
@@
 
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
+where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required.
+
+\begin{change}For each of the \plural{\dataVariableCount}{variable} used in the paper, JavaScript generates a \LaTeX\ header file that declares and assigns a calculated value.
+The header file is then included in the paper using \LaTeX's standard \texttt{\textbackslash input} command.
+Here is an example of one such automatic definition:
+
+\begin{verse}\tt
+\textbackslash newcount \textbackslash dataVariableCount \\
+\textbackslash dataVariableCount = \the\dataVariableCount
+\end{verse}
+
+so the named value is then available for the author to use when the paper is typeset.
+The tables generated for the present paper also include data that did not require names for explicit use elsewhere in the \LaTeX\ files, so these numbers are not counted in the specific \the\dataVariableCount\ variable count.
+\end{change}
 
 Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
+For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary}, as used in the main paper.
 
@@
 \label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
 % check it works for other numbers 
@@
 
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
+Normally, when we write a number like \the\countUsesVersionControlRepository\ in a paper, especially longer or more complex numbers, we will later proof read them as ``the numbers we intended to write'' --- as remembering what we meant is easier than reading the details.
+Unfortunately, a sentence would likely seem to make as much sense when a number has been erroneously typed as, say,
 \newcount\t
@@
 \multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
+\the\t\ --- we hardly bother to pay attention because we \begin{change}think we\end{change} know what we are reading; at least we know what we meant to write.
+Worse, the more often we proof read a document, the more we remember, so the better we know what we think we said, and the more casual our proof reading becomes.
+It is very hard to spot all of our own typos.
 
@@
 
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
+All the discussion and examples above were generated automatically, and have been checked correct for other correct values than \the\countUsesVersionControlRepository.
 This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
+For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
+\begin{change}The approach turns a possibly-hard-to-spot \emph{single\/} error into \emph{multiple\/} errors spread throughout the paper into different contexts, thus increasing the chances of noticing the error.
 
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
+It must be emphasized that an automatically-guaranteed number that is supposed to be the same appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
+As the number is proof read, the different contexts encourage it to be read more carefully, and in different ways.
+\end{change}
+
+If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their values) then further checks would have been made to help detect and avoid errors.
+\LaTeX\ itself makes it very easy to check that numbers fall within reasonable ranges, or to have any other required safety properties.
+For the present paper, a potential problem is \begin{change}if\end{change} the paper is mistakenly typeset \emph{before\/} the latest JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense \begin{change}(e.g., from a debugging run of \texttt{data.js}).
+Variables might be checked as follows:\end{change}.
 
@@
 			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
+								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits set for this paper.}
 		\fi
@@
 
+\begin{change}
+\highlightmysubsection{Defending against system problems}
+Code can become obsolete as programming languages develop and compilers are improved.
+Typically, compilers first warn that code is ``deprecated'' and then later versions reject the old code.
+Furthermore, when code is run on different computers, different operating systems, and with different compilers, it is common to obtain different results.
+Data, too, is subject to the same problems, but data standards and formats are far more stable than code standards, so ``data rot'' is less of a risk (but no less a problem when it occurs) than ``software rot.''
+
+Additionally, errors can be the result of human slips, such as accidentally deleting a line of code or a line of data in a spreadsheet.
+Such corruption errors are hard to detect unless specific steps are taken to ensure the integrity of code and data \cite{excel-fiasco}.
+Checksums are the simplest way to detect such errors, but during active research more refined techniques might be used in addition, for example checking that the number of rows of data in a spreadsheet monotonically increases.
+In the present paper, the JSON data is more structured than a spreadsheet matrix, and a number (as it happens, \the\JSONerrorCount) of other consistency checks are imposed on the data.
+
+To protect against version, portability and other problems, the Github repository for the present paper includes a check on software versions and a checksum check for all possibly affected files, including the data file.
+This does not solve the problem, but it ensures anyone developing or reproducing the paper's work will at least be forewarned of potential version or portability problems.
+The Github repository itself can be used to restore files that have been corrupted.
+
+\highlightmysubsection{Problems of restrictive journal policies}
+\label{restrictive-policies}
+Automatically generated variables are used throughout the paper and this \supplement.
+As usual, \LaTeX\ detects any spelling errors in the use of variables, thus helping protect the paper against typos that could otherwise mislead the paper's readers.
+Conveniently, \LaTeX\ also supports sophisticated calculations itself \cite{latex-calculus}, so the typeset paper can use any variable values in further calculations without going back to modify the data source file (in the present case, \texttt{data.js}).
+In practice this enables the author to avoid copying-and-pasting values from a data source or calculator, and then overlooking keeping them up to date with changes to the data or formula required.
+
+For example, the caption of table \ref{table-repo-summary} in the main paper calculates its ``\plural{\pubdelayinmonths}{month}'' figure from the generated variables recording the repository date of cloning used to provide the data to construct the table.
+The number of months will of course be correctly updated if the paper's repository \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ is subsequently checked again:
+
+\begin{quote}
+At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.\end{quote}
+
+Of course, the data generation process itself checks that this surprising statement remains valid, and provides a warning if the wording may need revising.
+
+Unfortunately, although using generated variables and analyses from a paper's data is a very simple technique to help make published papers more reliable, some journals and preprint servers (such as \emph{IEEE Transactions on Software Engineering}, \emph{PLOS ONE}, and \emph{arXiv\/}) do not permit papers to be submitted using \LaTeX\ source code that uses the standard \texttt{\textbackslash input}, \texttt{\textbackslash bibliography}, and other related commands.
+Typically they also do not support running any data collection or analysis either (which the present paper does when it clones repositories).
+
+Another program (\texttt{expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
+Of course, the expanded version now contains all variables as fixed constants, so the submitted paper is misleading and useless to other researchers if the data is modified --- the effort to ensure all published numbers are automatically correct is defeated.
+Such restrictive publishing policies undermine reproducibility.
+\end{change}
+
 \highlightmysubsection{Code policies of sampled journals}
@@
 
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
+It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{detecting-against-error} and \ref{restrictive-policies}.
 In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
 \\
+
+\newdimen\tabwidth
+\tabwidth=\textwidth
+\advance \tabwidth by -4em
+\def\specifyurl{\vskip 3mm
+\begingroup\raggedleft\scriptsize
+} 
+\def\specifyurletc#1{\endgroup\\
+\hbox{ }\hfill\begin{change}\begin{tabular}{|p{\tabwidth}@{}}
+#1
+\end{tabular}\end{change}\vskip .3cm
+}
 
@@
 Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
+As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.}
 
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
+\specifyurl
+\url{royalsociety.org/journals/authors/author-guidelines/#data}
+\specifyurletc{Accessed 29 July 2020; the policy has been revised (undated, but accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
+The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
+
+\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
+{\sf --- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
+[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.}
+
+\specifyurl
+\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
+\specifyurletc{Accessed 29 July 2020; since updated to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
+Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.'' (accessed 2 February 2022).
 }
 
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
+\noindent\textbf{\emph{Lancet Digital Health\/} author guidelines}\\
+Journal has detailed data policies, but no code policy.
 
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
+\specifyurl
 \url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
+\specifyurletc{Accessed 29 July 2020. Still no code policy (accessed 2 February 2022).}
 
 \noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
+Journal has detailed data policies, but no code policy.
 While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
 
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
+\vskip 3mm \noindent {\sf --- The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
+[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.}
+\specifyurl
 \url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
+\specifyurletc{Accessed 29 July 2020. Policy unchanged when accessed 2 February 2022.}
 
-\highlightmysubsection{Assessments and criteria}
+\highlightmysubsection{Sample assessments and criteria}
 
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
+\def\flagStyle#1{\textcolor{blue}{\sf #1}}
 
+\noindent
+Assessment flags are \flagStyle{highlighted in color} to be clearer in the following tables.\\
+
+\noindent
 Legend:\\
@@
 This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
+See section \ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
 
@@
 \immediate\closeout\minicontentsfile
+\rememberpage
 \end{document}

### 2022-02-28 data.js

@@ -66,6 +66,14 @@ var descriptionsOfFields = {
         documentation: "Evidence that source code was developed rigorously",
         flag: "$\\sf S_{{\\mbox{\\scriptsize rigorous}}}$"
     },
+    hasToolBasedDevelopment: {
+       documentation: "Evidence of any tool-based development",
+        flag: "$\\sf S_{{\\mbox{\\scriptsize tools}}}$"
+    },
+    hasOpenSourceDevelopment: {
+       documentation: "Team or open source development",
+        flag: "$\\sf S_{{\\mbox{\\scriptsize open source}}}$"
+    },
     hasOtherTechniques: {
         documentation: "Other evidence of good practice; see details in summary table",
         flag: "$\\sf S_{{\\mbox{\\scriptsize otherSE}}}$"
@@ -110,7 +118,7 @@ var descriptionsOfFields = {
         documentation: "Number of pages"
     },
     reference: {
-        // reference number is automatically generated, but you can override it if you want to 
+        // unique reference number is automatically generated, but you can override it if you want to 
         // refer to a particular paper in the Latex papers
         documentation: "Internal Latex citation identifier (needs to be unique, which is checked)"
     },
@@ -127,7 +135,7 @@ var descriptionsOfFields = {
         essential: 1
     },
     error: {
-        documentation: "Set to an error string if required"
+        documentation: "Set to an error string if needed"
     }
 };
 
@@ -392,9 +400,10 @@ var data = [{
         hasCodeInPrinciple: 1,
         hasOtherTechniques: 1,
         hasGoodComment: 1,
-        codeComment: "Code available for private view, though some code available with minor comments. Paper describes using two contrasting methods to help confirm correctness",
[…]

### 2022-02-28 generated-allGitRepos.sh

@@ -1,0 +1,45 @@
+rm -rf "git-MetricSelectionFramework"
+git clone "http://github.com/ChristophKanzler/MetricSelectionFramework" "git-MetricSelectionFramework"
+
+rm -rf "git-PENet"
+git clone "http://github.com/marshuang80/PENet" "git-PENet"
+
+rm -rf "git-PostoperativeOutcomes_RiskNet"
+git clone "http://github.com/cklee219/PostoperativeOutcomes_RiskNet" "git-PostoperativeOutcomes_RiskNet"
+
+rm -rf "git-philter-ucsf"
+git clone "http://github.com/BCHSI/philter-ucsf" "git-philter-ucsf"
+
+rm -rf "git-AI-CDSS-Cardiovascular-Silo"
+git clone "http://github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo" "git-AI-CDSS-Cardiovascular-Silo"
+
+rm -rf "git-SiameseChange"
+git clone "http://github.com/QTIM-Lab/SiameseChange" "git-SiameseChange"
+
+rm -rf "git-lactModel"
+git clone "http://github.com/SolangeD/lactModel" "git-lactModel"
+
+rm -rf "git-LRM"
+git clone "http://github.com/UT-NSG/LRM" "git-LRM"
+
+rm -rf "git-manifold-ga"
+git clone "http://github.com/ki-analysis/manifold-ga" "git-manifold-ga"
+
+rm -rf "git-blast-ct"
+git clone "http://github.com/biomedia-mira/blast-ct" "git-blast-ct"
+
+( echo "% date generated by running generated-allGitRepos.sh downloading Git repos";
+echo \\\\def\\\\clonedate {\\\\ignorespaces `date "+%e %B %Y"`}
+echo \\\\def\\\\cloneyear {`date "+%Y"`}
+echo \\\\def\\\\clonemonth {`date "+%m"`}
+) > ../generated-clone-date.tex
+
+(echo % shell-script calculated metavariables; echo "\\\newcount \\\JSONerrorCount";echo "\\\newcount \\\dataVariableCount") > ../generated-metadata.tex
+
+# how many errors can data.js find?
+echo "\\\JSONerrorCount =" `grep "error(" ../data.js|wc -l` >> ../generated-metadata.tex
+
+# how many variables does data.js define?
+echo "\\\dataVariableCount =" >> ../generated-metadata.tex
[…]

### 2022-02-28 generated-open-doi.sh

@@ -1,0 +1,32 @@
+open "http://www.doi.org/10.1038/s41746-020-0300-0"
+open "http://www.doi.org/10.1038/s41746-020-0286-7"
+open "http://www.doi.org/10.1038/s41746-020-0287-6"
+open "http://www.doi.org/10.1038/s41746-020-0268-9"
+open "http://www.doi.org/10.1038/s41746-020-0278-7"
+open "http://www.doi.org/10.1038/s41746-020-0266-y"
+open "http://www.doi.org/10.1038/s41746-020-0265-z"
+open "http://www.doi.org/10.1038/s41746-020-0248-0"
+open "http://www.doi.org/10.1038/s41746-020-0258-y"
+open "http://www.doi.org/10.1038/s41746-020-0261-3"
+open "http://www.doi.org/10.1038/s41746-020-0249-z"
+open "http://www.doi.org/10.1038/s41746-020-0255-1"
+open "http://www.doi.org/10.1098/rsos.200419"
+open "http://www.doi.org/10.1098/rsos.192210"
+open "http://www.doi.org/10.1098/rsos.191900"
+open "http://www.doi.org/10.1098/rsos.200566"
+open "http://www.doi.org/10.1098/rsos.200422"
+open "http://www.doi.org/10.1098/rsos.200462"
+open "http://www.doi.org/10.1098/rsos.191992"
+open "http://www.doi.org/10.1098/rsos.200622"
+open "http://www.doi.org/10.1098/rsos.200638"
+open "http://www.doi.org/10.1098/rsos.200621"
+open "http://www.doi.org/10.1098/rsos.191928"
+open "http://www.doi.org/10.1098/rsos.200307"
+open "http://www.doi.org/10.1098/rsos.191653"
+open "http://www.doi.org/10.1098/rsos.200320"
+open "http://www.doi.org/10.1016/S2589-7500(20)30108-4"
+open "http://www.doi.org/10.1016/S2589-7500(20)30107-2"
+open "http://www.doi.org/10.1016/S2589-7500(20)30131-X"
+open "http://www.doi.org/10.1016/S2589-7500(20)30063-7"
+open "http://www.doi.org/10.1016/S2589-7500(20)30085-6"
+open "http://www.doi.org/10.1016/S2589-7500(20)30078-9"

### 2023-01-31 generated/info-for-main.tex
Updating to January 2023 submission to Computer Journal

\newlabel{supplementary-material-makefiles}{{11}{38}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{on-code-data-publication}{{12}{41}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{RAP-diagrams}{{12}{44}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{excelFigure}{{2}{44}{AAA}{BBB}{CCC}}
\newlabel{supplementary-Spiegelhalter-section}{{13}{44}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{survey-section}{{14}{47}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{supplementaryTableRepoSummary}{{4}{49}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{supplementary-journal-policies-section}{{14}{52}{DUMMYA}{DUMMYB}{DUMMYC}}
\newlabel{restrictive-policies}{{14}{56}{DUMMYA}{DUMMYB}{DUMMYC}}
\def\MaxMainPaperCitationNumber{121}

### 2023-01-31 generated/legend.tex
Updating to January 2023 submission to Computer Journal

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
-\renewcommand\thesubsection{\thesection.\alph{subsection}}
-%\usepackage{hyperref}
-%\usepackage{refcount}
-\usepackage{url}
-\usepackage{doi}
-\usepackage{longtable}
-\usepackage{color}
-\usepackage{amsmath,amssymb}
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\input paper-seb-macros.tex
-
-\makeatletter
-      \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
-      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
-      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
-      }
-\makeatother
-
-\begin{document}
-
-\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}
-
-\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}
-
-%\orcid{0000-0003-2222-4243}
-
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
-\def\href#1{#1}
-
-\maketitle
-
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
-
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
-
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
-
-% start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
-
-\setcounter{section}{0}
-\def\n#1{{\sf\bfseries ({#1})}}
-
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
-
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
-
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
-
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
-
-%\newpage
-
-\makeatletter
-\let \bslash = \@backslashchar
-\makeatother
-
-%\renewcommand{\thesubsection}{\alph{subsection}}
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
-
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
-}
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
-
-\newpage
-\mysection{Further issues for Software Engineering Boards (SEBs)}
-
-\highlightmysubsection{Brief definition}
-Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
-
-Further details of the SEB proposal is in the main paper.
-Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:
-
-\begin{enumerate}\raggedright
-\item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
-
-\item
-When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}.
-Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
-How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
-
-\item 
-SEBs require policies on professional issues such as membership, transparency, and accountability.
-
-\item
-There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).
-
-\item
-Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs.
-For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?
-
-\item
-In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling.
-More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.
-
-\item
-SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
-
-\item
-Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large.
-Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
-\end{enumerate}
-There are other ideas to help make SEBs work, but it is clear they are part of the solution.
-We must not let perfection be the enemy of the good.
-SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.
-
-\mysubsection{Relationships of SEBs to Ethics Boards}
-
-\begin{enumerate}\raggedright
-\item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
-SEBs may also provide mentoring and training.
-
-\item
-Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
-
-\item 
-Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious).
-On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own.
-SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is.
-As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.
-
-\item
-SEBs should have a two-way collaboration with Ethics Boards.
-
-\begin{itemize}
-\item SEBs have to deal with ethical concerns, and how they may be implemented in code.
-One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.
-
-\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
-\end{itemize}
-
-\item
-Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
-\end{enumerate}
-
-\changemysubsection{SEBs are necessary but not sufficient}
-The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
-
-SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results.
-As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
-It needs doing.
-
-\mysection{Software engineering best practice}
-\label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
-
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
-
-The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible.
-Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer.
-Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}.
-Further suggestions for background reading can be found throughout this section.
-
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
-It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
-On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
-
-In practice, requirements and expectations change.
-Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
-In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
-
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
-
-{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy.
-To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe.
-Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}
-
-{The type and use of mathematics used in software engineering is formal methods.
-Not using formal methods ensures the resulting code is unsafe and unreliable.
-Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}
-
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
-Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
-Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.
-
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
-
-Many programming languages and programming environments have integrated features that support formal methods.
-For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@.
-In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
-Hoare is also a very good writer.
-
-Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way.
-Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy).
-Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do.
-Unfortunately, because formal methods are mathematical, few programmers have experience of using them.
-Fortunately tools are widely available to help use formal methods very effectively.
-
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
-Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
-The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code).
-Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.
-
-A special case of defensive programming appropriate for pandemic modeling is mixing methods.
-Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
-
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
-Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
-The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do.
-C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
-SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
-It is appreciated that the models may change and be adapted as new data and insights become available.
-Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable.
-Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used.
-Version control is supported by many tools (such as Git, Subversion, etc).
-
-If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
-Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
-
-{If results cannot be reproduced, has anything reliable been contributed?
-When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
-Reproducing the same results relying on the same codebase tells you little.
-The more independent reproductions of results the greater the evidence for belief in the implications.}
-
-Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes.
-Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.
-
-Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community.
-Open source is not a panacea, however; it raises many trade-offs.
-Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
-
-Open source raises important licensing and management questions to ensure the quality of contributions.
-A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository.
-This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.
-
-Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate.
-Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.
-
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
-In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
-Rigorous testing includes methods like fault injection.
-Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs.
-Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).
-
-It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again.
-For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).
-
-It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested.
-If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.
-
-There are many other important testing methods \cite{sommerville,knight,NVP}.
-
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
-Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
-
-{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories.
-One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
-}
-
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
-This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
-For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have?
-In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it.
-How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist.
-An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}
-
-There are many tools to help manage documentation (Javadoc, Doxygen, \ldots).
-Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}.
-Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.
-
-Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature).
-All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work.
-In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.
-
-A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
-
-Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project.
-Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.
-
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
-\globalLabel{supplementary-material-makefiles}
-Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
-Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team.
-Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.
-
-{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described.
-When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible.
-Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine.
-Makefiles typically also provide demo and test runs and other helpful features.
-Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}
-
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
-Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
-Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
-Indeed, reuse, scrutiny, and consensus are the foundations of good science.
-
-Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.
-
-A special case of reuse is to use software tools to help with software development.
-The tools (if appropriately chosen) have been carefully developed and widely tested.
-Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.
-
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
-When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
-This way lies over-fitting, a problem familiar from statistics (and machine learning).
-Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident.
-Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct.
-A simpler program is easier to prove correct, easier to program, and easier to debug.
-A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened.
-Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.
-
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
-
-To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}.
-However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate.
-Publications should then cite the  standards to which their computer models comply.
-
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
-
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
-
-As this long list illustrates, Software Engineering is a complex and wide-ranging subject.
-Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc).
-Multidisciplinary teamwork is essential.
-
-Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs.
-Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand).
-The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.
-
-Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.
-
-Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software.
-A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be.
-In short, users have to be brought into the software team too.
-
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
-Continuous Professional Development (CPD) is essential.
-
-Ironically, the more organized CPD the more likely the content itself will lag behind.
-There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees.
-Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.
-
-The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.
-
-Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
-
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
-Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
-
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
-In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
-
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
-The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
-
-This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
-
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
-
-A fragment of program code might say
-
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
-
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
-The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
-The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
-
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
-
-It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
-Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
-
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
-Data is then run on the virtual machine as program.
-Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions.
-Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program.
-There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal.
-In short, any distinctions between code and data are impossible to maintain.}
-
-In the present paper, we knowingly built on this blur between data and code.
-However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
-
-The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language.
-\LaTeX\ not only has text (as you are reading right now) but it also has code.
-For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted.
-When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters?
-The distinctions are all a bit moot.
-In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs.
-Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.
-
-The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above).
-As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data.
-A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer.
-In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.
-
-{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood).
-The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}
-
-Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.
-
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
-
-\def\questionsection#1{\changemysubsection{#1}}
-\def\question#1{\item \emph{#1\/}}
-\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}
-
-\questionsection{How trustworthy are the numbers?}
-\newcounter{resumeCounter}
-
-\begin{enumerate}
-\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.
-
-\sequestion{How rigorously has the software engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
-
-\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it.
-Were different methods used and compared, or was all confidence put into a single implementation?
-What internal consistency checks does the implementation have?
-Were invariants and assertions defined and checked?
-}
-
-\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.
-
-\sequestion{How are the claims presented that give us confidence in the code that they are based on?
-Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
-
-\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.
-
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the source?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed.
-Ask yourself, `Why does this source want me to hear this story?'
-
-\sequestion{The source of many science stories is the output of running some code.
-How reliable is this code?
-What evidence is there that the code was well-engineered so its reliability can be trusted?}
-
-\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.
-
-\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}
-
-\question{What am I not being told?} This is perhaps the most important question of all.
-Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.
-
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
-So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
-
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.
-
-\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}
-
-\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.
-
-\sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
-
-\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.
-
-\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}
-
-\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
-\end{enumerate}
-
-\newcount\temporary \temporary=\mainNumberOfReferences
-\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
-\advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
-
-\makeatletter    
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
-\makeatother
-
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
-\label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
-A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence and editorials were excluded.
-The sample represents what the editorial and the broader peer review community considers to be good practice.
-
-The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution.
-Indeed, all sampled papers used code in their research.
-Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable.
-It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).
-
-Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed.
-Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code.
-Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.
-
-There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
-
-The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted.
-Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey.
-What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized.
-Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
-
-As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.
-
-The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):
-
-\input generated-example-data.tex
-
-The data was written out by hand, after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
-(A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
-
-A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
-Hence, running the data generates the core human-readable information used in this paper.
-
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
-
-\begin{quote}
-\journalBreakdown 
-\end{quote}
-
---- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.
-
-An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:
-
-\begin{quote}
-\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
-\end{quote}
-
-it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data.
-The phrase above is now in fact written in \LaTeX\ in the paper as follows:
-
-\begin{quote}\tt\small
-Code repositories were used by \\
-\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
-\end{quote}
-
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
-
-Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
-
-The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-\highlightmysubsection{Detecting and defending against error}
-\label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
-% check it works for other numbers 
-%\countUsesVersionControlRepository=296
-
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
-\newcount\t
-\t=\countUsesVersionControlRepository
-\divide \t by 10
-\newcount\fraction
-\fraction=\countUsesVersionControlRepository
-\newcount\tent \tent=\t \multiply \tent by 10
-\advance\fraction by -\tent
-\the\t.\the\fraction,
-\t=\countUsesVersionControlRepository
-\advance \t by -1
-\the\t,
-\t=\countUsesVersionControlRepository
-\advance \t by 1
-\the\t, or
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
-
-\begin{itemize}\raggedright
-\item
-The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
-\divide \t by 10
-\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t) \cite{fixit}.
-
-\item
-The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
-\end{itemize}
-
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
-
-\newcount\min \min=5
-\newcount\max \max=20
-%\countUsesVersionControlRepository=4
-\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
-\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
-\ifnum \countUsesVersionControlRepository>\max
-	\else 
-		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
-		\fi
-\fi
-
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
-
-\highlightmysubsection{Assessments and criteria}
-
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
-
-Legend:\\
-\input generated-legend.tex
-
-\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
-\end{longtable}
-}
-
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
-
-\begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
-\raggedright
-
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
-
-\end{thebibliography}
-
-\immediate\closeout\infofile
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
-\end{document}
+\begin{tabular}{lp{4.5in}}
+\flagStyle{$\sf P_c$}&Journal has a code policy (see section~\ref{supplementary-journal-policies-section})\\
+\flagStyle{$\sf P_{{\mbox{\scriptsize c-breach}}}$}&Paper breaches journal code policy (see section~\ref{supplementary-journal-policies-section})\\
+\flagStyle{$\sf R_c$}&Paper uses a code repository (e.g., GitHub)\\
+\flagStyle{$\sf R_{\mbox{\scriptsize c-empty}}$}&Code repository contains no code\\
+\flagStyle{$\sf R_d$}&Paper uses a data repository (e.g., Dryad, Figshare, GitHub)\\
+\flagStyle{$\sf S_{\mbox{\scriptsize NONE}}$}&No code available at all (note: code is not expected for standard models, systems or statistical methods)\\
+\flagStyle{$\sf S_p$}&Paper says source code is available in principle\\
+\flagStyle{$\sf S_{+}$}&Paper or URL provides source code\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize rigorous}}}$}&Evidence that source code was developed rigorously\\
+\flagStyle{$\sf S_{\mbox{\scriptsize tested}}$}&Evidence that source code has been run with a clean build and tested\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize tools}}}$}&Evidence of any tool-based development\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize open source}}}$}&Team or open source development\\
+\flagStyle{$\sf S_{{\mbox{\scriptsize otherSE}}}$}&Other evidence of good practice; see details in summary table\\
+\flagStyle{$\sf C_0$}&Code has no non-trivial comments\\
+\flagStyle{$\sf C_1$}&Code only has trivial comments (e.g., copyright)\\
+\flagStyle{$\sf C_2$}&Helpful comments explaining code intent, rather than rephrasing the code\\
+\flagStyle{$\sf C_{+}$}&Code has substantial, useful comments and documentation\\
+\end{tabular}

### 2023-01-31 generated/make-help.tex
Updating to January 2023 submission to Computer Journal

% 1 - make all
% 2 - make archive
% 3 - make check-same
% 4 - make check-update
% 5 - make check-versions
% 6 - make data
% 7 - make expand
% 8 - make help
% 9 - make help-brief
% 10 - make mathematica
% 11 - make one-file
% 12 - make pdf
% 13 - make push
% 14 - make readme
% 15 - make really-tidyup
% 16 - make tidyup
% 17 - make zip
% 18 - make zip-data
{\sf\begin{tabular}{rp{4.5in}}
\texttt{make all}&   Analyze the data, then typeset the main PDF files (\texttt{paper-seb-main.pdf} and \texttt{paper-seb-supplementary-material.pdf}).\\
\multicolumn{1}{l@{\vdots}}{}&\\
\texttt{make data}&   Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc. In particular, this runs \texttt{node data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them. Note that downloading all the repositories in a reasonable time needs decent internet bandwidth.\\
\multicolumn{1}{l@{\vdots}}{}&\\
\texttt{make help}&   Explain how to use \texttt{make}, and list all available options for using \texttt{make}.\\
\multicolumn{1}{l@{\vdots}}{}&\\
\texttt{make one-file}&   Make a single PDF file paper-seb.pdf (i.e., paper + appendix) all in one.\\
\multicolumn{1}{l@{\vdots}}{}&\\
\texttt{make tidyup}&   Remove all easily generated files, except the main PDFs and the large Git repositories needed for the pilot survey.\\
\end{tabular}}

### 2023-01-31 generated/stand-alone-over-fitting-code.tex
Updating to January 2023 submission to Computer Journal

\def\numbertoname#1{#1}
\def\remembernamevalue#1#2{}
\newcount\countRepos
\newcount\countMissingCode
\newcount\countCommentedOut
\newcount\countNotCommentedOut
\documentclass[11pt]{article}
\usepackage{geometry} 
\geometry{a4paper}  
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{url}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert \#1 `dirname #1`/`basename #1 .tif`.png}
\def\supplement{Supplement}
\begin{document}
\section{Section generated by \emph{Mathematica\/} taken from full paper}\subsection{The siren call of over-fitting code}\label{over-fit}
%Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that. We do the science, and we are in the world, so there are challenging meta-theoretical issues as well as practical experimental and resourcing issues. Karl Popper is one of many philosophers of science to develop principles and theories about science \cite{popper-three-worlds,popper-conjectures-refutations}.

%Popper was concerned with the conventional sciences, where the world, Nature, is assumed fixed and in principle knowable. Writing just six years after Popper, Herbert Simon introduced the ``sciences of the artificial'' \cite{sciences-artificial}, which instead focus on understanding artificial worlds, most typically computer programs. Programs effectively create worlds, and the sciences of the artificial seek to understand and improve those worlds. Crucially, there is no philosophical or other reason why artificial worlds should be based on principles, let alone be simple or elegant. Conventional science's \ae sthetics need not apply.

%Science to date has been inevitably driven by human \ae sthetics: that is, simple, elegant, understandable, beautiful, brief, exciting, intelligible \ldots\ all are human properties that are applied to doing, funding, defining, and selecting ``good'' science.

\noindent
Poor code undermines scientific quality: poor code can generate plausible results from \emph{any\/} data, including even fictional science. A temptation is that %artificial world \cite{sciences-artificial}, built on
adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena. \def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
\def\figureStarDetails{*}
\begin{figure\figureStarDetails}[t]
{ \newdimen \imageWidth 
  \imageWidth=\textwidth
  \imageWidthCalculation
  \begin{center}\includegraphics[width=\imageWidth]{generated/mathematicaplot.pdf}\end{center}}
\caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them. Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model. The over-fitted polynomial model fits the sample \emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
\label{fig-overfit}
\end{figure\figureStarDetails}
In modeling terms, superficially successful computer programs tend to \emph{over-fit\/} phenomena \cite{over-fit}. Instead of using code to test the models, we tinker with the code adapting it closer and closer to our prejudices. The code then apparently confirms our science, since we fitted it to our understanding.

In general, an \emph{over-fitted model\/} fits a set of data closely, but contains more parameters than can be justified by the science. An over-fitted model fails to reliably predict results beyond the scope of the data it has been fitted to. Over-fitting is a well-known problem, but the point is that when code is used, over-fitting is done unconsciously by programmers adjusting the code --- its parameters, its structure, its embedded data, the calculations it performs --- that specifies the model. ``A model over-fits if it is  more complex than another model that fits equally well'' \cite{over-fit}, is a criterion that describes almost every program! Programmers without the discipline and experience to manage the unlimited adaptability of code debug, alter, and extend their code to make it do what they think it should do. This becomes a vicious circle as the idea of what the science is becomes driven by the code. Rather than debugging code by improving its fit to the actual science, it gets debugged and extended to fit the expectations.

The problems of over-fitting data may be visualized using a Real$\rightarrow$Real function of one variable (figure \ref{fig-overfit}). The code that generates an over-fitted curve seems to work very well: in the example shown, the over-fitted curve fits the sampled data exactly; indeed, the code used here will fit any new data exactly as well. But the code has a negligible ability to predict new data or to describe theories of the data, which is the point of modeling. The fact that over-fitted code seems to work well is deceptive.

Looking specifically at the data plotted in figure \ref{fig-overfit} if, for the sake of argument, we assume the error in the data is normally distributed then the values the over-fitted code generates outside the range of the sample are improbable. For example, the basic linear model predicts$\hat{y}(0)=0.5$, versus the extreme value predicted by the over-fitting code, $\hat{y}(0)=-41.8$,  even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.

While over-fitting data is a well-known problem, the point for this paper is that \emph{code\/} itself can easily be over-fitted. Code can of course be over-fitted in more complex ways than can be illustrated with elementary polynomials, as here. Code over-fitting is much harder to recognize because there may be no simple graph plot, like figure \ref{fig-overfit}, to highlight the problems. Furthermore, almost all code is far more intricate than the two trivial polynomials used to illustrated figure \ref{fig-overfit}.

Unfortunately, code in published science \emph{is\/} often over-fitted, and over-fitted in a way that is very hard to scrutinize. For example, in epidemiology (which is considered in section \ref{section-pandemic-modeling}) it is routine to use very complicated, large dynamical models parameterized with numerous social, cultural, health, demographic and geographical data. The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases. Indeed, many of the programs in the survey used comment to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countRepos\ papers in the pilot survey that reported use of code repositories (covering \input ../generated/kloc.tex
\totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code. The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\remembernamevalue{numbernamesexamplefootnote}{\thefootnote}}

Furthermore, scientific support code is rarely documented well enough to know what it should have been doing, which should be answered by a specification. With no clear specification and documentation, the code can be arbitrarily hacked to get any convenient results, since no particular specification for it has been defined that it should adhere to. Thus we risk doing and promoting substandard science because we --- the scientists and the publication process --- are not managing the unlimited adaptability and complexity of code that science has come to rely~on. This is over-fitting of the worst kind --- in conventional over-fitting one can at least hope to see that the fitting is over-parameterized for the data, but in code over-fitting the code and specification are not visible, therefore not adequately scrutinized, and --- worse --- the ``data'' the code over-fits includes the entire conceptual contribution of the paper.

Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
\vskip 1ex
\noindent 
{\newdimen\boxwidth
\boxwidth = \columnwidth
\advance\boxwidth by - 1em
\begin{tabular}{@{}l@{}}
\framebox[\columnwidth][r]{\hfill\parbox{\boxwidth}{This section (\ref{over-fit}) of this paper, including figure \ref{fig-overfit}, was generated as a computational paper in a \emph{Mathematica\/} notebook, then exported to \LaTeX\ to include here. The \supplement\ provides a more complete explanation, as well as the data and details of the fitted polynomials (also generated automatically). This paper's repository includes the notebook.

%\hskip 1em Generated text and data presented elsewhere in this paper were created by a \texttt{make} file running shell scripts and JavaScript code analyzing repositories, JSON survey data, etc. Results were exported to files of \LaTeX\ markup and definitions, then imported into this paper and expanded where needed.
}\hfill}\end{tabular}}
\section{Data and polynomials}
\begin{center}\sf\begin{tabular}{|c|r|r|r|r|r|}\hline
$x={}$&1.00&1.10&2.20&2.60&4.50\\
$y={}$&1.50&2.70&4.90&5.70&8.20\\
\hline\end{tabular}
\end{center}

Showing both polynomials with coefficients to 2 decimal places, the linear least squares model to fit this data is:

$$
\hat{y}=0.49+1.80x
$$

and the Lagrange polynomial model (an exact fit to the 5 data points) is

$$
\hat{y}=-41.79+85.46x-56.30x^2+15.65x^3-1.51x^4
$$

The name of the relevant \emph{Mathematica\/} notebook file where all data and code for this section (and for generating the paper's section \ref{over-fit} and its figure \ref{fig-overfit}) can be found in \texttt{programs/over-fitting-section.nb}, which is included in the Git repository for the paper.\end{document}

### 2023-01-31 generated/summary-table.tex
Updating to January 2023 submission to Computer Journal

\begin{tabular}{|rrrc|}\hline
Number of papers sampled relying on code&32&100\%&\\\hline\hline
\multicolumn{4}{|l|}{\textbf{Access to code}}\\
Some or all code available&12&38\%&\\
Some or all code in principle available on request&8&25\%&\\
Requested code actually made available (within 2 years 7 months${}^{\star}$)&0&0\%& \\\hline
 \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
Evidence program designed rigorously&{0}&{0\%}&\\
Evidence source code properly tested&{0}&{0\%}&\\
Evidence of any tool-based development&{0}&{0\%}&\\
Team or open source based development&{0}&{0\%}&\\
Other methods, e.g., independent coding methods&{1}&{3\%}&\\\hline
\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
Substantial code documentation and comments&2&6\%&\\
Comments explain some code intent&3&9\%&\\
Procedural comments (e.g., author, date, copyright)&10&31\%&\\
No usable comments&17&53\%&\\\hline
\multicolumn{4}{|l|}{\textbf{Repository use}}\\
Used code repository (e.g., GitHub)&9&28\%&\\
Used data repository (e.g., Dryad or GitHub)&9&28\%&\\
Empty repository&1&3\%&\\\hline
\multicolumn{4}{|l|}{{\textbf{Evidence of documented processes}}}\\
Evidence of RAP/\RAPstar\ or any other principles in use to support scrutiny&{0}&{0\%}&\\\hline
\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
Papers published in journals with code policies&26&81\%&\\
Clear breaches of journal code policy (if any) & 11&42\%&(\emph{N}~=~26)\\
\hline\end{tabular}\\
\vskip 1ex ${}^{\star}$Time of 2 years 7 months is wait between code request and date of generating this table.

### 2023-01-31 programs/data.js

@@ -1,0 +1,1401 @@
+// Harold Thimbleby, 2020-2021
+// JSON data and Javascript code to process it
+
+'use strict';
+
+// documentation for fields in the JSON data
+// you don't need to use all fields in the data (flags default to unset)
+// first we start with definitions of all the data fields
+var descriptionsOfFields = {
+    accessed: {
+        documentation: "Date when accessed",
+        essential: 1
+    },
+    doubleChecked: {
+        documentation: "Date when double checked",
+        essential: 1
+    },
+    authors: {
+        // surname initials comma-separated list
+        // code can be modified to generate author lists in other formats as required by journal
+        // search for function fixAuthors(s) to see what we do:
+        //     list comma-separated authors, final one with 'and' but then say et al if more than maxAuthors authors
+        documentation: "Authors",
+        essential: 1
+    },
+    codeComment: {
+        documentation: "Evaluator's notes about code",
+        essential: 1
+    },
+    codeURL: {
+        documentation: "URL (typically GitHub) for code repository"
+    },
+    codeCommentedOut: {
+        documentation: "Has a repo, and executable code has been commented out"
+    },
+    dataComment: {
+        documentation: "Evaluator's notes about data",
+        essential: 1
+    },
+    doi: {
+        documentation: "DOI",
+        essential: 1
+    },
+    error: {
+        documentation: "Set to an error string if needed"
+    },
+    hasBreach: {
+        documentation: "Paper breaches journal code policy (see section~\\ref{supplementary-journal-policies-section})",
+        flag: "$\\sf P_{{\\mbox{\\scriptsize c-breach}}}$" // how a flag will be abbreviated in the paper text
+    },
+    hasCodeInPrinciple: {
+        documentation: "Paper says source code is available in principle",
+        flag: "$\\sf S_p$"
+    },
+    hasCodePolicy: { // flag value set automatically from journal name
+        documentation: "Journal has a code policy (see section~\\ref{supplementary-journal-policies-section})",
+        flag: "$\\sf P_c$"
+    },
+    hasCodeRepo: {
+        documentation: "Paper uses a code repository (e.g., GitHub)",
+        flag: "$\\sf R_c$"
+    },
+    hasCodeTested: {
+        documentation: "Evidence that source code has been run with a clean build and tested",
[…]

### 2023-01-31 programs/expand.js

@@ -1,0 +1,67 @@
+// Harold Thimbleby, 2020-2022
+// Javascript code to expand TeX \input commands
+
+var TeXFiles = ["paper-seb-main.tex", "paper-seb-supplementary-material.tex"];
+
+'use strict';
+
+var fs = require('fs');
+
+function saveFile(fileName, string) {
+    fs.open(fileName, 'w', function (err, fd) {
+        if (err) {
+            if (err.code === 'EEXIST') {
+                console.error(fileName + " can't be over-written");
+                return;
+            }
+            console.error("** Some error trying to write " + fileName);
+            throw err;
+        }
+        fs.writeFileSync(fd, string, function (err) {
+            if (err) {
+                return console.log("** Saving " + fileName + " got error: " + err);
+            }
+        })
+    });
+}
+
+function getFile(file) {
+	file = file.trim();
+    try {
+        return fs.readFileSync(file, {encoding:'utf8', flag:'r'});
+    } catch (err) {
+        console.log("** Can't read file " + file + " - " + err);
+        return "\\error couldn't read " + file + "\n";
+    }
+}
+
+var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}");
+
+var bufferExpand, bufferInsert, auxfile;
+
+for (var i = 0; i < TeXFiles.length; i++) {
+    console.log("Recursively expand all \\input and \\bibliography macro calls in " + TeXFiles[i]);
+    auxfile = TeXFiles[i].replace(/.tex/, ".aux");
+    console.log("        \\input " + auxfile + " (automatically pre-inserted between \\makeatletter and \\makeatother)");
+    bufferExpand = "\\makeatletter\n"+getFile(auxfile) + "\n\\makeatother\n";
+    bufferExpand = bufferExpand + getFile(TeXFiles[i]); // puts in buffer
+    var replaced;
+    do {
+        replaced = false;
+        bufferExpand = bufferExpand.replace(re, function (match, filename) {
+            replaced = true;
+            console.log("        \\input " + filename);
+            bufferInsert = getFile(filename);
+            return "% expand " + filename + "\n" + bufferInsert + "% end expanding " + filename + "\n";
+        });
+        bufferExpand = bufferExpand.replace(rebib, function (match, filename) {
+        	replaced = true;
[…]

### 2023-01-31 programs/make-metadata.sh

@@ -1,0 +1,11 @@
+echo % shell script "(make-metadata.sh)" generated LaTeX variable values counting errors detected by data.js > generated/metadata.tex
+
+# how many errors can data.js find?
+echo \\\\newcount \\\\JSONerrorCount >> generated/metadata.tex
+echo \\\\JSONerrorCount = `grep "error(" data.js|wc -l` >> generated/metadata.tex
+
+# put count of variables (number registers plus macro definitions) to \JSONvariableCount
+echo \\\\newcount \\\\dataVariableCount >> generated/metadata.tex
+echo \\\\dataVariableCount = `egrep "(newcount)|(\\\\\\\\def)" generated-*|wc -l` >> generated/metadata.tex
+
+

### 2023-01-31 programs/obsolete javascript data fitting models/fittingData.js

@@ -1,0 +1,176 @@
+/*
+ Author: Harold Thimbleby, 2022
+ 
+ Given an array of (x,y) pairs as data, this simple program finds the best fit linear model (using 
+ least squares) and an exact polynomial model (a Lagrange polynomial) to fit the data.
+ 
+ The simple algorithms used are taken straight off Wikipedia:
+ 
+ 	Least Squares: https://en.wikipedia.org/wiki/Simple_linear_regression
+	Lagrange polynomial: https://en.wikipedia.org/wiki/Lagrange_polynomial
+	
+Both algorithms generate a vector of polynomial coefficients, and the function polynomialToString converts the coefficients to a string representation of the polynomial in x. Thus, for example, LeastSquares(data) generates the numerical vector [a,b] which represents the linear polynomial a+bx.
+
+ */
+
+var data = [{
+    x: 2.2,
+    y: 4.9
+}, {
+    x: 4.7,
+    y: 8.2
+}, {
+    x: 2.6,
+    y: 5.7
+}, {
+    x: 1.,
+    y: 2.7
+}, {
+    x: 0.9,
+    y: 1.5
+}];
+
+/* Typical shell command (using node to run the Javascript) and resulting output:
+
+    % node p.js
+    Given data:
+    [
+      { x: 2.2, y: 4.9 },
+      { x: 4.7, y: 8.2 },
+      { x: 2.6, y: 5.7 },
+      { x: 1, y: 2.7 },
+      { x: 0.9, y: 1.5 }
+    ]
+    Model coefficients to 4 significant figures:
+    
+      Linear least squares fit: y =  0.8232 +1.656 x
+      Linear evaluated at 5:  9.1032
+      Linear evaluated at 5 in LaTeX:
+        \def\lineary{$y(5)= 9.1 $}
+    
+      Lagrange polynomial:      y = -32.04 +69.63 x -46.72 x^2 +13.08 x^3 -1.257 x^4
+      Lagrange evaluated at 5:  -2.5149999999999864
+      Lagrange evaluated at 5 in LaTeX:
+        \def\exacty{$y(5)= -2.5 $}
+    
+    
+    ---
+    
+    In Mathematica format:
+    
+    data = {{2.2,4.9},{4.7,8.2},{2.6,5.7},{1,2.7},{0.9,1.5}};    
+    linear[x_] :=  0.8232 +1.656 x;
[…]

### 2023-01-31 paper-seb-main.tex
Fixing git messages

--- 2022-01-27 paper-seb-main.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\input plos-header.tex
+\documentclass{comjnl}
+\usepackage{ragged2e}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
-\renewcommand\thesubsection{(\thesection.\alph{subsection})}
-\usepackage{hyperref}
-
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\makeatletter
-    \def\do@url@hyp{\do/}
-    \usepackage{moreverb,url}
-    %    \renewcommand{\thefootnote}{\arabic{footnote}}
-\makeatother
+%----
+  \setlength\textheight{220mm}
+%----
 
 \input paper-seb-macros.tex
+\input generated/info-for-main.tex
+\input generated/page-lengths.tex
 
-\input generated-page-lengths.tex
+\long\def\ignore#1{}
+
+\title{\mytitle} 
+\author{Harold Thimbleby}
+\shortauthors{Harold Thimbleby}
+\affiliation{See Change Fellow in Digital Health, Swansea University}
+\email{harold@thimbleby.net}
+
+%\subject{Computational science (bioinformatics,~medical~computing,~etc)}
+
+\keywords{Computational Science; Software Engineering; Reproducibility; Scientific scrutiny; Reproducible Analytic Pipeline (RAP \& \RAPstar\hskip -1pt)}
+
+\received{30 April 2022}
+%\revised{--}
+
+%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
+%&\texttt{harold@thimbleby.net}\\
+%&ORCID 0000-0003-2222-4243
+%\end{tabular}
+
+%\orcid{0000-0003-2222-4243}
 
@@
 
-\vspace*{0.2in}
+\begin{abstract}
+\justifying
+%\color{red}
+%This is a placeholder for the revisions that are almost complete.
+I have uploaded it now in case the computer system would lose all record of the submission --- I have emailed to ask for a short extension, but with the offices shut, there is no chance of any negotiation.
+Thanks for your understanding.
+%\color{black}\vskip .2cm
 
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Harold Thimbleby
-\\
-\bigskip
-See Change Fellow in Digital Health
-\\
-\bigskip
+\noindent
+Scientists rarely assure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny.
+Even if available, scientists rarely provide adequate documentation to understand or use their code reliably.
+When code used in science avoids adequate scrutiny published science will be unreliable, because results --- data, graphs, images, inferences, etc --- generated by the code will themselves be unreliable.
 
-% Current address notes
+\hskip 1em This paper justifies and proposes ways to improve research and science relying on code:
 
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
-&\texttt{harold@thimbleby.net}\\
-&ORCID 0000-0003-2222-4243
-\end{tabular}
-
-%\orcid{0000-0003-2222-4243}
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
-\noindent
-\emph{Background:} Computer code underpins modern science.
-While scientific models are routinely peer reviewed, the algorithms and the quality of code implementing them often avoid scrutiny; therefore the details of many scientific conclusions cannot be rigorously justified.
-
-\noindent\emph{Problem:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit.
-Additionally, both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects.
-Code is fallible, so any interpretation that relies on code is therefore fallible, and if the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized.
+%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available.
+Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects.
+Code is fallible, so any interpretation that relies on code is also fallible.
+When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized.
 In turn, scientific claims cannot be properly scrutinized.
 
-\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.
+%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions.
+%The paper reports a pilot survey of peer-reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
 
-\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.
+%Code can be improved using Software Engineering. This paper argues for specific solutions:
 
-\noindent\emph{Solutions:} Code can be improved using software engineering. This paper argues for specific solutions:
+%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.
 
+%\noindent\emph{Solutions:}
+
+\raggedright
 \newdimen \mywidth \mywidth=\textwidth
-\advance \mywidth by -1em
-\parshape 2 0em \textwidth 1.5em \mywidth 
+\advance \mywidth by -3em
+
+\def\indented#1{\vskip 1ex
+\parshape 2 0em \textwidth 2em \mywidth 
 \noindent 
-\hbox to 1.5em{ 1.\hfill}Professional software engineers can help and should be involved, particularly in critical research;
+\hbox to 2em{ #1.\hfill}\ignorespaces}
 
-\parshape 2 0em \textwidth 1.5em \mywidth 
+\indented 1
+Professional Software Engineers can help, particularly in critical fields such as public health, climate change, and energy.
+
+\indented 3
+``Software Engineering Boards,'' analogous to Ethics or Institutional Review Boards, should be instigated and used.
+
+\indented 4
+The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp.
+\RAPstar\ (or comparable interventions) could be supported and encouraged in journal, conference, and funding body policies.
+
+\vskip 1ex 
+\justifying
 \noindent
-\hbox to 1.5em{ 2.\hfill}``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;
+The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input.
+The \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ processes.
 
-\parshape 2 0em \textwidth 1.5em \mywidth 
-\noindent 
-\hbox to 1.5em{ 3.\hfill}Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent software engineers.
+%\begin{quote}
+\def\seprule{\vskip 2ex\hrule\vskip 2ex}
+\seprule
+\raggedright
+\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
+\zq Science is what we understand well enough to explain to a computer.''\\
+\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\
 
-\noindent
-The \supplement\ provides a summary of professional software engineering best practice relevant to scientific research and publication.
+\zq I have to write to discover what I am doing.''\\
+\hfill {\rm Flannery O'Connor, quoted in \emph{Write for your life\/} \cite{write-for-life}}
 
-%\end{abstract}
+\zq Criticism is the mother of methodology.''\\
+\hfill {\rm Robert P. Abelson in \emph{Statistics as Principled Argument\/} \cite{abelson}}
 
-\vfill\begin{quote}
-\hfill``Criticism is the mother of methodology.''
+\zq \hangindent \wd0 From its earliest times, science has operated by being open and transparent about methods and evidence, regardless of which technology has been in vogue.''\\
+\hfill {\rm Editorial in \emph{Nature\/} \cite{nature-chatgpt}}
 
-\hfill Robert P Abelson, \emph{Statistics as Principled Argument\/} \cite{abelson}\end{quote}
-\vfill
-\paragraph*{Keywords:} 
-Computational science; Software Engineering; Reproducibility; Scrutiny
+\seprule
+%\end{quote}
+\end{abstract}
+\maketitle
 
-\newpage\section*{Author summary}
-
-%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
-Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales.
+%\section*{Author summary}
+%
+%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
+%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales.
 His research focuses on human error and computer system design, particularly for healthcare.
-%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
-Harold won the British Computer Society Wilkes Medal.
+%%In addition to over 340 peer-reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
+%Harold won the British Computer Society Wilkes Medal.
 He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder.
@@
 See his web site, \texttt{www.harold.thimbleby.net}, for more details.
-\linenumbers
 
-%\maketitle
+%\noindent\emph{This paper evidences widespread problems arising from scientific papers relying on inadequately defined computer and inaccessible code, which undermines peer-review and replication, and undermines scientific quality generally.
+The risks vary, depending on how critical a paper's claims may be: ranging from superficial for opinion papers, to serious for papers in domains such as medicine, public health, or policy.
+This paper contrasts the situation with the mature approach to using statistics, and goes on to make proposals to help address the problem.
+The proposals are feasible, and presented in the hope of motivating their (or better solutions) being implemented, tested, and refined.}
 
-\section{Introduction}\label{problems}
-Computational methods are comparatively new, powerful, and very flexible tools for scientists.
-Computational methods rely on programming, but unfortunately poor programming is very easy use to apparently get results with; such results are likely to be misleading.
-The role of computational methods can be compared to statistics, another useful scientific tool.
-Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics.
-Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.
+\section{Introduction}
+Unreliable, often unstated and unexplored, code and computational dependencies (including using AI or ML) in science are widespread.
+Furthermore, code is rarely published or made accessible in a useful form, nor is it professionally scrutinized.
+Too often, therefore, code and results claimed do not contribute to reliable science, and do not support replication or reproduction on which future science can be firmly based.
 
-{I}{n good science}, all statistics, methods and results have to be reported very carefully and in precise detail. For example, a statistical claim might be summarized as follows:
+\begin{table}
+\vskip 1.2ex % align top of columns!
+\tableofcontents
+\vskip 3mm \hrule
+\end{table}
 
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\sf\hskip -\wd0\box0 %Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
-\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 ~($\sf 95\% \mbox{\sf ~CI~} 0.78, 4.27, p = 0.0009$).
-Between group effect size $\sf d = 0.55 ~(95\% \mbox{\sf ~CI~} 0.32, 0.77)$.'' \cite{example-stats}\end{quote}
-
-This typical wording briefly summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
-It is a \emph{lingua franca}.
-It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics.
-
-Scientists write like this --- and conferences and journals require it --- because statistical claims need to be properly accountable and documented in a clear way.
-Speigelhalter \cite{Speigelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Speigelhalter-section}).
-Results should not be uncritically accepted just because they are claimed.
-
-The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted.
-Scientists work closely with competent, often specialist, statisticians who engage with the research from experiment design through to analysis and publication.
-Further, it is assumed that statistics will be peer reviewed, and that review will often lead to improvement.
-
-Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement.
-Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians.
-Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.
-
-Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them will be generated by computer.
-Indeed, computers are now very widely used, not just to calculate statistics, but to run the models (the term is defined below), do the data sampling and processing, and even to operate the sensors that generate the data that is analyzed.
-Computers are used to run human-participant surveys, such as web-based surveys.
-The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers.
-Computers even help with the word processing and typesetting of research.
-
-In short, computers, data, and computer code are central to modern science.
-However, using code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
-Software code, like statistics, is also subject to unintentional bias \cite{se-bias}.
-All these issues are non-trivial concerns requiring technical expertise to manage well.
-
-A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.
-
-Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims.
-Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.
-
-%When relevant details of models are omitted (or are not described in recognisable ways), it is impossible to properly scrutinise claims.
-Describing a model used, perhaps mathematically, or providing the code (e.g., to download) that implements it is as inadequate as just providing raw data without details of the statistics analysis supporting the claim.
-
-We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
-An interesting overlap with statistical and software engineering scrutiny is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)\@.'' yet \emph{how\/} those analyses might have been performed is not discussed.
-The problem is that admitting using SAS 9.2 or any other named system does not really help scrutiny, as such systems can do anything with the data.
-A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is reliable.
-
-It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements the models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases.
-Being able to understand (at least in principle) the exact code used in implementing a model is critical to having confidence in the results that rely on it.
-Unfortunately code is rarely considered a substantial part of the science to which it contributes.
-
-This paper reviews a selection of papers in leading international journals, and finds that both papers and journal policies take code for granted.
-This paper then argues that, just as is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.
-Specifically, code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny.
-Developing justifiably reliable code is the concern of the field of \emph{software engineering}, which will be discussed further below, as well as more substantially in this paper's \supplement).
-
-This paper shows that unreliable computational dependencies in science are widespread.
-Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction.
-In short, the quality of much modern science seems to be undermined because the code it relies on is rarely of adequate quality \emph{for the uses to which it is put}.
-
-This paper explores the extent of these software engineering problems in published science.
-The paper additionally suggests some ways forward.
-The \supplement\ is an integral part of the suggested solutions.
-In particular, the \supplement\ section~\ref{supplementary-Speigelhalter-section} summarizes Speigelhater's uncontroversial statistics probes and draws explicit analogies for the critical assessment of the quality of scientific code.
-
-\subsection{The role of code in science and scientific publication}
-%Yet without code, models could not be run.
-
-For the purposes of this paper, models map theory and parameters to describe phenomena, typically to make predictions or to test and refine the models.
-With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.
-
-Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive.
-The chance that a thousand lines of hand-written code is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability.
-How to do this well is the concern of software engineering, and is discussed throughout this paper.
-
-While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are (certainly in all cases reviewed later in this paper) they are long, complex, inscrutable and (our survey shows) lack adequate documentation.
-Therefore the models are very likely to be unreliable \emph{in principle}.
-If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if has not been clearly documented?
-Without documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?
-
-Without substantial documentation it is impossible to scrutinize code properly.
-Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications.
-What does it mean?
-What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation?
-Does the code mean what was intended?
-What are the assumptions on \texttt{y} and do they hold invariantly?
-Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown.
-A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of overflow).
-
-Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code.
-Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim.
-If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed.
-It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.
-
-Contrary to the views expressed in the present paper, some authors have asserted that code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality}.
-On the contrary, if code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways.
-Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.
-
-Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes.
-The question is whether it matters, \emph{when\/} it matters, and when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work.
-
-%This paper explores the role of code in scientific research.
-Initially focusing on examples from pandemic modelling (because of its relevance to serious matters of public health) this paper shows that scientific culture, including editorial processes, have not adapted to the increasingly dominant role of code and managing the risks of its fallibility.
-The paper then suggests how to improve, including building mutual engagement across the science and software engineering communities.
-
-\subsection{Bugs, code and programming}\label{knowledge}
-Critiques of data and model assumptions are very common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned.
-Yet data and program are formally equivalent (see further discussion in \supplement, section \ref{on-code-data-publication}).
-Program code has as great an affect on results as the data.
-Code is harder to scrutinize, which means errors in code can have more subtle effects on results.
-
-It should be noted that almost all code contains ``magic numbers'' --- that is, data masquerading as code.
-This common practice ensures that published data is very rarely all of the data because it omits the magic numbers.
-This emphasizes the need for repositories to require the inclusion of code so all data is actually available.
-
-Bugs can be understood as discrepancies between what code is intended to do and what it actually does.
-Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered.
-Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination.
-None of the code examined for this paper includes either.
-
-If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
-Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things.
-For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the sense of producing incorrect results --- that is, what was wanted was wrong.
-Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.
-
-\iffalse Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, unqualified programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering.
-Dunning and Kruger go on to say,
-
-\begin{quote}\sf\raggedright\setbox0=\hbox{``}\hskip -\wd0\copy0 People usually choose what they think is the most reasonable and optimal option \hbox{[\hskip 2pt\ldots]} The failure to recognize that one has performed poorly will instead leave one to assume that one has performed well; as a result, the incompetent will tend to grossly overestimate their skills and abilities.
-\hbox{[\hskip 2pt\ldots]} Not only do these people reach erroneous conclusions and make unfortunate choices, but their incompetence robs them of the metacognitive ability to realize it.''\end{quote}
-
-Unlike many skills (skating, brain surgery, \ldots) programming, typical of much engineering, is one where errors can go unnoticed for long periods of time --- things seem to work nicely right up to the moment they fail.
-The worse programmers are, the more trivial bugs they tend to make, but trivial bugs are easy to find so, ironically, being a poor programmer \emph{increases\/} one's self-assessment because debugging seems very productive.
-It is easy for poor programmers and their associates to believe they are better than they actually are.
-%This is fertile ground for the better-than-average bias \cite{dunning-kruger}.
-
-It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}.
-We \emph{all\/} make mistakes we are unaware of.
-If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.
-
-{Everyone is subject to Human Factors (including the author of the present paper, e.g., as discussed in \cite{enigma}): for instance, the standard cognitive bias of confirmation bias encourages us to look for bugs when code fails to do what is expected and then debug it to produce better results, but if code generates expected results not to bother to debug it further.
-This of course tends to make code increasingly conform to prior expectations, whether or not those expectations are scientifically justified.
-Typically, there was no prior specification of the code, so the code should be right, especially after all the debugging to make it ``correct''!
-Thus coding routinely suffers from HARKing (Hypothesizing After the Results are Known \cite{harking}), a methodological trap widely recognized in statistics.}
-
-Computers themselves are also a part of the problem.
-Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable.
-Programs can be written so that it is not possible to determine what they do or how they do it (whether deliberate obfuscation or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
-The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
-\fi
-
-\section{State of the art in pandemic modeling}
-A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no conscious role for code or computation as such.
-It appears that the new insights, advances, rigor, and problems that computers bring to research are not considered relevant.
-
-A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
-(TRIPOD guidelines ignore code completely.) The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted.
-It should be noted that flowcharts, which the review did consider, are programs intended for direct human use.
-Flowcharts, too, should be designed as carefully as code, for exactly the same reason: it is hard to program reliably.
-
-A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} for H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling.
-The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.
-
-A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\sf\hskip -\wd0\box0 
-For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
-'' \cite{ferguson-interview}\end{quote}
-
-This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
-The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
-(Typical $h$ values vary by discipline.)} of 93.
-Ferguson should be well aware of the standards of coding used in at least his own field.
-This comment, quoted above, is therefore likely to be representative of the standards of the field as a whole.
-
-Ferguson's admission is tantamount to saying that the published scientific findings are not reproducible.\footnote{A constructive discussion of software engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.} This is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
-If this C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation.
-It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified over fifteen years without becoming incoherent.
-If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
-In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
-
-The programming language C which was used is not a dependable language; to develop reliable code in C requires professional tools and skills.
-Moreover, C code is not portable, which limits making it available for other scientists to use safely (C notoriously gets different results with different compliers, libraries, or hardware).
-The \supplement\ discusses these issues further.
-
-Ferguson, author of the code, says of the criticisms of his code, ``However, none of the criticisms of the code affects the mathematics or science of the simulation'' \cite{thumbs-up}.
-Indeed.
-The problem the current paper is raising is that the theoretical epidemiology may be fine, but if it is not mapped into code that correctly implements the models, then the program's output cannot be relied on without independent evidence.
-In science, this is the normal requirement of \emph{reproducibility}.
-
-The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed the code could be run again and produced comparable results (compared, apparently, to an Excel spreadsheet!).
-That can be achieved at this low level of sophistication is hardly surprising, regardless of the quality of the code.
-There was no scientific insight that merits the use of the word ``reproduction.'' If reproducibility is to be a useful scientific criterion, an \emph{independently\/} developed model needs to produce equivalent results (called $N$ version programming, a standard software engineering practice \cite{NVP}) like public health surely requires --- as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.
-Meanwhile, it is a shame that using software for scientific models has enabled the bar to reproducibility, as understood by journals such as \emph{Nature}, to be lowered to a mechanical level that is only sufficient to detect some forms of dishonesty, as opposed to methodological limitations, which is the point of reproducing work.
-
-Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).
-It is
-\newcount\u
-\newcount\v
-\newcount\w
-\newcount\frac
-\input generated-covidsimSummary.tex
-kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
-Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done.
-If nothing else, as the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models.
-It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place.
-The revised code has now been published, and it has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
-
-Some epidemiology papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
-Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications.
-Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it is arguably even worse than not publishing code at all.
-Pseudo-code is not precise enough to help anyone scrutinize a model; copying pseudo-code introduces bugs.
-An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}.
-{The \supplement\ provides further discussion of reproducibility.}
-
-\subsection{Science beyond epidemiology}
-Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims.
-People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place.
-The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even downright flaws.
-
-Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}.
-World-class software can attain maybe 1 bug per 1,000 lines of code.
-Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new insights into their models.
-In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- can be subject to numerical errors \cite{hamming}.
-It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion).
-Such bugs undermine confidence in model results.
-
-Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what researchers have done, but merely making code available (for instance, providing it in their \supplement\ with papers, putting it in repositories, or using open source) is not sufficient.
-It is noteworthy that some papers disclosed that they had access to special hardware.
-
-Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available.
-While some \citeeg{unfinished,lancet-unfinished} do make what they call ``documented'' code available, they provide no more than superficial comments.
-This is \emph{not\/} documentation as properly understood.
-Such comments do not explain code, explain contracts, nor explain algorithms.
-Some readers of the present paper may not recognize these technical software terms; contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming {(see the \supplement\ for a checklist of many relevant, professional software engineering concepts and resources)}.
-
-If full code is made available, it may be technically ``reproducible,'' but the scientific point is to be able to understand and challenge, potentially refute, the findings; to do that, much more is required than merely being able to run the code \cite{notebooks,popper}.
-
-Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper {and indeed in computer science research itself \cite{machine-learning-reproducibility}}) is inscrutable.
-Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do.
-Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.
-
-Making outline code (including pseudo-code) available without proper documentation and without discussing its limitations is unethical: it encourages others to reproduce and build on poor work.
-
-\makeatletter
-\long\def\@makecaption#1#2{%
-  \vskip\abovecaptionskip
-  \sbox\@tempboxa{#1: #2}%
-  \ifdim \wd\@tempboxa >\hsize
-    \textbf{#1}: #2\par
-  \else
-    \global \@minipagefalse
-    \hb@xt@\hsize{\hfil\box\@tempboxa\hfil}%
-  \fi
-  \vskip\belowcaptionskip}
-\makeatother
+This paper makes an explicit analogy between the use of statistics, which has clear standards for reliable use and presentation, and the use of code.
+Like statistics, code is often relied on to support key results in scientific publications, yet code is generally informal, inaccessible and incorrect.
+Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.
 
 \begin{table*}
-\begin{center}
+\begin{center}\sf\normalsize
 \begin{tabular}{|rl|} \hline
@@
 \end{center}
-
-%\newcount \total
-%\total=\dataN
-%\multiply \total by 4 % three reviewers, one editor
-%\advance \total by \countAuthors
-%\the\total\
-%leading scientists assuming authors, 1 editor, and 3 reviewers per paper, but not counting acknowledged colleagues et al, covering  (plus appendices and supporting material) in total.
-%
-\caption{Overview of peer-reviewed paper sample.}
+\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}.
+Survey methodology and data is provided in the \supplement.
+(The survey does not include the motivating papers \cite{ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref{section-pandemic-modeling}.)}
 \label{table-overview}
@@
 \begin{table*}
-\begin{center}\normalsize
-\input generated-summary-table.tex
-%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
+\begin{center}\sf\normalsize
+\input generated/summary-table.tex
 \end{center}
-
-\caption{Summary of survey results.}
+\caption{Breakdown of pilot survey of peer-reviewed science papers relying on code. }
 \label{table-summary}
@@
 
-\begin{table*}[t]
-
-\def\reponame#1#2{{\tt #1} \csname cite-#2\endcsname}
-
-\expandafter\def\csname pagelength-covid-sim\endcsname{20}
-\expandafter\def\csname cite-covid-sim\endcsname{\cite{ICmodel}}
-
-% divide #1 by #2
-\newcount \uu
-\newcount \vv
-\def\makeAverage#1#2{%
-	\uu=#1
-	\vv=#2
-	\divide \uu by \vv
-	\the\uu
-}
-
-\input generated-clone-date.tex
-\begin{center}
-
-\expandafter \advance \gitPages by \csname pagelength-covid-sim\endcsname
-
-\input generated-repos.tex
-\vskip 1ex
-\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
-\\
-\small Repository clones downloaded and automatically summarized \clonedate
-\end{center}
-\newcount \pubdelayinmonths
-\pubdelayinmonths=\cloneyear
-\advance \pubdelayinmonths by -2020
-\multiply \pubdelayinmonths by 12
-\advance \pubdelayinmonths by -4 % published in April
-\advance \pubdelayinmonths by \clonemonth
-\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
-Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting.
-Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
-Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the\pubdelayinmonths\ months had already elapsed since the paper referring to the code was published.}
-\label{table-repo-summary}
+\begin{table*}
+\newdimen\quarterwidth
+\newdimen\separation
+\separation=3mm
+\quarterwidth=\textwidth
+\advance\quarterwidth by -4\separation
+\divide\quarterwidth by 4
+\noindent{\sf\begin{tabular}{@{}c|c|c|c@{}}
+\bfseries Level 0&
+\bfseries Level 1&
+\bfseries Level 2&
+\bfseries Level 3 \\ \hline
+\multicolumn{1}{@{}p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
+\multicolumn{1}{|p{\quarterwidth}}{Article states whether code is available and, if so, where to access them.}&
+\multicolumn{1}{|p{\quarterwidth}}{Code must be posted to a trusted repository. Exceptions must be identified at article submission.}&
+\multicolumn{1}{|p{\quarterwidth}@{}}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.}
+\end{tabular}}
+\caption{The TOP committee's recommended levels for journal article code transparency.
+Level 0 is provided for a comparison that does not meet any TOP requirements.
+Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref{reproducibility-concerns}.}
+\label{TOPtable}
 \end{table*}
 
-\section{A pilot survey of peer-reviewed research relying on code}
-\label{survey-section}
-The problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
-For instance, a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model} provides no details of its code.
-Its \supplement, which might have provided code, no longer exists.
+This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research, because of its significance for informing and driving public health responses to the Covid pandemic.
+A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see the tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey.
+In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below).
+Only one paper mentioned any Software Engineering methods, albeit without technical details.
 
-Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized.
-We therefore undertook a survey of recent papers in leading peer reviewed journals.
+In the survey sample 81\% of papers were published in journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies.
+One paper declared it had accessible code, but the relevant repository was and still remains empty.
+The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience}.
 
-\subsection{Methodology}
-A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.
+This paper therefore argues that code should be developed and discussed in a professional, rigorous, and supportive environment that facilitates quality science with clear presentation and appropriately rigorous scrutiny of code.
+Its main contribution is to suggest straightforward ways to enable this.
+The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and inform raising the profile and constructively debating the issues more widely.
 
-The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal.
-All papers sampled are Open Access, although for some papers some or all of the associated data has restricted access.
-Table \ref{table-overview} is an overview of the sample.
+An extensive online \supplement\ appendix to this paper provides additional resources, including brief details of many Software Engineering practices relevant to supporting quality science.
+The supplement will be of particular interest to research software engineers supporting non-software-specialist scientists.
 
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence, and editorials were excluded.
-The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.
+\section{Background}
+The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences.
+There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature.
+For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.
 
-The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.
+Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
+Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, following international quality standards.
 
-The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.
+Computers are a unique, new technology, far more flexible and challenging than thermometers.
+Understanding computers and integrating them into science is likely to be harder than the development of modern thermometry.
+Computation not only expands science's paradigms and supports new discoveries (particularly with AI), but it also \emph{does\/} new science --- almost all modern laboratory instrumentation, including thermometers, is heavily computerized.
+Pushing the boundaries of science, then, involves pushing the boundaries of computer science.
+The synergy runs deep: for instance, while particle physics relies on powerful supercomputers, quantum physics itself is developing more powerful quantum computing.
 
-%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).
+``Computational science'' has come to mean a particular style of science based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational in this sense.
+Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ In all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications.
+All areas of science are being profoundly computerized.
+Furthermore, developments in computer science themselves drive science such that earlier science is even becoming obsolete as the computer technology moves on \cite{form}.
 
-The \the\dataN\ papers cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install.
-Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment was made whether the code provided actually reproduced a paper's specific claims.
-Indeed, if we trust the papers that their code was actually run and provides the results as reported, then merely running their code (when provided in full) merely checks the paper/code consistency, but will not assess the quality or reliability of the code.
+\subsection{Related code quality concerns}\label{related-work}
+Publishing high quality computer code has been strongly advocated since the earliest times, such as the \emph{Communications of the ACM\/} in its first issue in its first volume published in 1958, where it outlined its algorithm publication policy.
+The new policy was illustrated with a square root algorithm \cite{acm-algorithms}.
+However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the concern of the present paper.
+Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code.
 
-%Assessing code quality has been a lively topic in software engineering for decades, and has international standards (ISO/IEC 9126, updated to ISO 25000).
-Here, code is assessed\footnote{The author of this paper has been assessing undergraduate and postgraduate computer science student code since the 1970s, using double marking and moderation.} using the n criteria of \cite{software-quality} as a marking scheme with scores out of 100/n.
+In almost all published science, the code it relies on is taken for granted, just as in routine chemistry the quality of the glassware is not at issue.
+While chemists are trained in reliable methodologies, so taking quality glassware for granted is reasonable.
+Code is a newer innovation.
+Reliable code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}.
+Inevitably, because of these reasons taken collectively, much published science depends on unreliable code that is not explicitly discussed, was not peer-reviewed, and is not open to scrutiny, reproduction, or reuse by the scientific community.
 
-The full methodology, data and analysis, as well as many software engineering and Human Factors techniques to help improve reliability and reproducibility used in the preparation of the present paper, are provided in the \supplement, which is also available from a repository (see details at the end of this paper).
+A study of 863,878 Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully.
+Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of 9,000 research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences.
+They found a comparable result that 74\% of the code files analyzed failed.
+These results are consistent with this paper's findings, as summarized in tables \ref{table-overview} \& \ref{table-summary}.
 
-%\emph{It is important that this survey is not taken as an evaluation of any specific paper: individual measurements are inevitably noisy.
-Instead, the aim is to form a view of scientific publishing generally.
-Mean scores, as summarized in table \ref{table-summary} and \ref{table-quality}, are more reliable measurements than the individual paper score breakdowns provided in the \supplement.}
+The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' More generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse has already done so.
+These ideas may be compared to the present paper's proposal of Software Engineering Boards (SEBs), as discussed in detail in section \ref{summary}, supported by more specific suggestions in the \supplement.
 
-\subsection{Summary of results}
-The sample selection criteria necessarily identified scientific research with software engineering contributions.
-No evidence of verification and validation was seen.
-There was only one example of basic software engineering methods, namely independent coding.
-There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer --- who writes high quality code to start with!\ --- would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
-(See the \supplement\ for definitions and a list of standard methods.)
+These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis,reproducible-manifesto}.
+Concern has led to a new journal \emph{ReScience~C\/} to explore and encourage the explicit replication of previously published research \cite{rescience}; furthermore, \emph{ReScience~C\/} promotes new and open-source implementations in order to ensure that the original research is reproducible.
+However, most of its published concern focuses on data rather than code.
 
-Much code depended on substantial manual intervention to compile it.
-All code (where actually provided) was sufficiently complex that, if it was supposed to be used or scrutinized, required more substantial documentation than was provided.
+To start to address code quality issues, the Transparency and Openness Promotion Committee first met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}.
+TOP covers citation standards, replication standards, and code standards amongst others.
+TOP recommends levels of compliance to their recommendations, where level 0 does not meet the standard, and levels 1 to 3 are increasingly stringent.
+The TOP levels for code are shown in table \ref{TOPtable}.
+The TOP standards  continue to develop, and are now maintained on a wiki at \url{osf.io/9f6gx/wiki/Guidelines} \cite{TOP-wiki}.
+Note that TOP aligns with but is a stricter approach than the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which is critiqued in section \ref{critique-fair}.
 
-On the whole, on the basis of the sample evidence, scientists do not make their code available, and rarely provide adequate documentation (see table \ref{table-summary}).
+As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (section \ref{section-vicious-cycles}).
+The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths.
+The Society of Research Software Engineering (\url{https://society-rse.org}) has been established to further promote research software engineer interests.
 
-It is noteworthy that the papers themselves, typically a few published pages, are very small compared to the code they rely on; the papers represent much less authoring effort than the code and data that they rely on for their claims and correctness (see table \ref{table-repo-summary}).
-With one minor exception (which may have been accidental), no papers reported anything on any software engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
+There is no shortage of computational tools available to help address the problems.
+However, it should be noted that such tools are not a panacea, as the study \cite{jupyter-study} cited above makes clear.
+This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be important.
 
-None of the papers used any specific software engineering methods, such as open source \cite{open-source} and other standard methodologies provided in the \supplement, to help manage their processes and help improve quality.
-Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reason this paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}
+In addition to unintentional problems with code quality and reproducibility, scientific misconduct occurs when the outcome is intentional.
+While pure plagiarism, which is misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have realistic properties, the results are likely to be destructive.
+The Wakefield MMR fraud linking vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted, and has been extraordinarily destructive \cite{MMR-fraud}.
 
-\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
-In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach.
-The implication of these results is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies, or they are technically unable to assess them --- and, if so, they also fail to say they are unable to assess them.
-This lack of expertise is consistent with the limited awareness of software engineering best practice that is manifest in the published papers (and resources) themselves.
+A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists at least once fabricated, falsified, or modified results, and over a third undertook other questionable research practices \cite{misconduct-issues}.
+The meta-analysis qualifies the figures carefully, but these are alarming rates with any qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The author's survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}
 
-Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \input generated-clone-date.tex
-while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could not been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
+While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using the same methods as will improve research reproducibility, as discussed throughout the present paper.
 
-Sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
+\subsection{Computable papers}\label{computable-papers}
+``Electronic lab notebooks'' (ELNs) \cite{elns}, emphasize computer tools that specifically support laboratory notebook authoring and editing.
+In contrast, ``computable papers'' aim to support the scientific paper authoring process: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.
 
-Overall, there was no evidence that any code had been developed carefully, {let alone by using recognized professional software engineering methods}. In particular, 
-\ifnum \countCodetested=0
-no papers
-\else only \plural{\countCodetested}{paper}
-\fi in the survey provide any claims or evidence of effective testing, for instance with evidence that tests were run on clean builds.
-{While it may sound unrealistic to ask for evidence on software quality in a paper written for another field of science, the need is no less than the need for standard levels of rigor in statistics reporting, as discussed in the opening of this paper.}
+If, for example, HTML is being used to prepare a computational paper, it could include Javascript code like \texttt{<script> document.write(responses.total) </script>}, which would insert the result of running that Javascript code into the paper, such as the number into a sentence in an empirical paper, like ``We collected data and obtained 754 responses.''
 
-Data repositories (the Dryad Digital Repository, Figshare or similar) were used by \plural{\counthasDataRepository}{paper} to provide structured access to their data.
-Unlike GitHub, which is a general purpose repository, Dryad has scientifically-informed guidelines on handling data, and all papers that used Dryad provided more than just their raw data --- they provided a little, sometimes substantial, documentation for their data.
-At the time of writing, Dryad is not helpful for managing code --- its model appears to be founded on the requirement that once published papers must refer to exactly the data they used, so further refinements on the data (or code) are taboo, even with version control.
+Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with basic calculations (e.g., here we calculate $10!=
+\def\factorial#1{\ffactorial{#1}1}
+\newcount\t
+\def\ffactorial#1#2{%
+\ifnum#1=0 #2\else
+\ffactorial{\the\numexpr#1-1\relax}{\the\numexpr#1*#2\relax}%
+\fi}
+\factorial{10}$ by writing \LaTeX\ code \emph{in\/} this paper).
+More practically, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper.
 
-Key findings from the survey are summarized in tables \ref{table-summary} and \ref{table-repo-summary}, and are discussed in greater detail in the \supplement.
+\newdimen\boxwidth
+\boxwidth = \columnwidth
+\advance\boxwidth by - 1em
+\noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors.
+The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}
 
-\section{Discussion}
-Effective access to quality code is not routine in science.
+In the ``old days'' such numbers (\the\dataN, \the\countAuthors, etc) quoted above would have been manually worked out, then eyeballed and typed up by the authors.
+Instead, in this paper, those numbers (and many more) were computed automatically.
+Furthermore, they update automatically when the data or calculations change.
+They were inserted into the text of this paper automatically, and are auditable back to their sources.
+
+The idea is easy to implement.
+Code can generate text like \texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}} (where the \the\countAuthors\ is the calculated number) save that line of text to a file, which is then input into the paper where it is typeset in the normal way.
+Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the correct value is at the time.
+Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11} \ldots\ 1,000 etc, as done (for example) in this paper's footnote~\numbernamesexamplefootnote.
+
+Rather than use quirky languages like \LaTeX\ computable papers more often build on general-purpose programming languages.
+Python and R, for instance, have powerful features for processing data, plotting graphs, doing statistics, and more.
+
+The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct --- \emph{with no further work from the authors}.
+Indeed, if an author corrects a mistake, the correction will apply automatically to all future edits.
+
+However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown.
+Many authors therefore prefer a more structured approach imposed by tools designed for the purpose.
+A taut review is Perkel \cite{perkel-review}, but here four tools,  WEB, \emph{Mathematica}, Jupyter, and knitr, will serve to sample the variety of approaches that are available:
+
+\begin{itemize}\raggedright
+
+\item
+WEB is the earliest tool reviewed here.
+WEB was developed by Donald Knuth in 1984 \cite{LP,LP-book} as a batch (non-interactive) tool to support his then radical new concept of \emph{literate programming}.
+The idea was to facilitate programmers write literate documentation for their code.
+WEB combines a sequential documentation file with code that can be presented in any order, thus overcoming the problem that the best explanation of a program is not necessarily written in the same order as the code it explains.
+The original WEB allowed Pascal programs to be documented in \TeX, but many variants of WEB have since been developed that are more flexible in the systems they support.
+
+\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code.
+
+\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications \emph{using\/} the code, inserting output, such as graphs, generated by running it.
+
+\item
+\emph{Mathematica\/} is one of the earliest fully interactive notebook tools for computational papers.
+Notebooks were developed by Theodore Gray in 1988 \cite{notebook-patent}.
+Notebooks consist of a collection of ``cells.'' Some notebook cells are labelled as text or as section headings, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output.
+The new cell can be numbers, tables, mathematics, a plot, an image, or even more text --- the arbitrary output of running code, in fact.
+Moreover, a notebook itself is a \emph{Mathematica\/} expression, so it --- the paper itself --- can be analyzed or manipulated by code in any way.
+In particular, the entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.
+
+\hskip 1em A \emph{Mathematica\/} notebook can be published directly as a paper, but some code might be distracting.
+Typically the author therefore hides some or all code cells that are irrelevant to the narrative of the paper, but which nonetheless were required to generate the results presented.
+
+\hskip 1em The user manual for \emph{Mathematica\/} itself was written as a \emph{Mathematica\/} notebook, which ensured all its examples actually worked --- and probably helped ensure the correctness of the \emph{Mathematica\/} code used behind the scenes (see section \ref{paper-as-lab}).
+The book is now the largest example of software documentation in existence: in its latest edition it runs to over 10,000 pages.
+
+\item
+Jupyter was developed by Fernando P{\'e}rez and Brian Granger \cite{jupyter-history}, and takes a similar approach to \emph{Mathematica\/}, but is open-source and not closely integrated with any particular programming language, as \emph{Mathematica\/} is.
+Jupyter can be installed on a PC or run over the web.
+
+\hskip 1em Jupyter is both an authoring tool and a framework on which to build other tools: thus Google's colaboratory is built on top of Jupyter, using it as a foundation but making stylistic changes, including providing free computational resources.
+
+\hskip 1em Many extensive examples of using Jupyter notebooks (and other good practice, such as using repositories) to support large scale science projects can be found at the Gravitational Wave Open Science Center at \url{www.gw-openscience.org}.
+
+\item
+knitr \cite{knitr} is a powerful culmination of a variety of tools, Pweave, Sweave, and ideas from literate programming.
+Knitr combines a markdown document with R code, and is a more powerful approach than the analogous, and perhaps more familiar, HTML+Javascript example shown above to motivate this section.
+\end{itemize}
+
+\noindent
+Thimbleby \cite{spec-led-design} is a 1999 example of a peer-reviewed paper (about user interface design) written as a \emph{Mathematica\/} notebook, which makes the point that a distinctive feature of its methodology is that the \emph{Mathematica\/} notebook creates a fully inspectable and replicable process.
+The notebook is available on the author's web site; it can be checked by others, or easily extended or repurposed to support new research --- and it still works
+{\newcount \howlong \howlong = \year
+\advance \howlong by -1999
+\the\howlong}
+years later.
+
+There are tools to make using code more convenient and more reproducible, as this section briefly reviewed, but they are rarely used or used haphazardly, as the next section shows.
+
+\subsection{RAP: Reproducible Analytical Pipelines}\label{RAP-section}
+Writing a paper typically starts in a word processor (such as Microsoft Word), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on.
+The process will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.
+
+Table \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then edited into a paper.
+
+\begin{table*}
+{\sf
+\begin{center}
+\def\drop#1{\setbox0=\hbox{\lower .5em\hbox{#1}}%
+\ht0=0em
+\dp0=0em
+\copy0}
+\def\lowarrow{\drop{$\rightarrow$}}
+
+\def\bigBracket{$
+	\setbox0=\hbox{\lower 1.2ex\hbox{$\left.\vbox to 4em{\vfill}\right\}$}}
+	\ht0=0em \dp0=0em 
+	~\copy0
+$}
+
+\begin{tabular}{|lcl@{}cl@{\hskip 2em}cl|} \hline
+\drop{\bfseries Data sources}&\lowarrow&\bfseries \drop{Analysis} & \lowarrow&\bfseries Select results& \lowarrow& \bfseries Submit for\\
+\bfseries &&\bfseries & &\bfseries for write up & & \bfseries publication \\ \hline \hline 
+&&&&&&\\
+Experiments&& Hand calculations &&&&\\
+Standard data&& Packages  && Copy \& paste &&\\
+Search engines&& SPSS etc & \bigBracket & and edit data && Final paper\\
+Literature &&  Graphics packages  && (text, images, graphs, etc)&&\\
+Sensors && Specially-written code && into paper &&\\ 
+\vdots && \vdots &&&& \\ &&&&&&\\ \hline
+\end{tabular}\end{center}}\caption{A simplified schematic of the publication pipeline.
+For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
+The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline processes so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
+Additional schematics are provided in section \ref{RAP-diagrams} in the \supplement.}
+\label{fig-pipeline}
+\end{table*}
+
+\begin{table*}
+{\sf
+\newcount \enum
+\enum=1
+\renewcommand{\theenumi}{%\ifnum \enum<10 \hphantom{0}\fi
+\the\enum
+\global\advance \enum by 1}
+\begin{enumerate}
+\item
+Peer-review used to ensure the process is reproducible and to identify improvements 
+\item
+No or minimal manual interference;
+for example copy-paste, point-click and drag-drop steps replaced using computer code that can be inspected by others 
+\item
+Open-source programming languages so that processes do not rely on proprietary software and can be reproduced by others
+\item
+Version control software, such as Git, to guarantee an audit trail of changes made to code 
+\item
+Publication of code, whenever possible, on code hosting platforms such as GitHub to improve transparency 
+\item
+Well-commented code and embedded documentation to ensure the process can be understood and used by others 
+\item
+Embedding of existing quality assurance practices in code, following guidance set by recognized organizations 
+\end{enumerate}
+\emph{Adopting RAP principles is not necessarily about incorporating all of the above: implementing just some of these principles will generate valuable improvements.}
+}\caption{A minimum standard of RAP, based on the UK Statistics Authority summary \cite{RAPstandard}.}
+\label{figure-RAP-summary}
+\end{table*}
+
+For clarity, the schematic pipeline in table \ref{fig-pipeline} omits many steps in the creative scientific process.
+Furthermore, each step is iterated and modified as the research progresses, and, indeed, as referees require revision.
+The point is that in typical scientific practice each step in the table is largely or entirely manual, typically selecting and copying output from the previous phase, and then pasting and editing the results into the next.
+The pipeline of data $\rightarrow\cdots\rightarrow$ paper is then iterated by hand as the various components are refined and improved until the authors are happy with the final paper.
+
+As problems are found in a paper, the data, calculations and code are debugged, refactored, and refined.
+The process is rarely systematic, and even less likely to be documented --- after all, the atomic steps seem to be innocuous copy and paste actions.
+The final paper and the ideas it embodies are what matters.
+
+The insight of the reproducible analytic pipelines (RAP) proponents is that every time any step in the pipeline is performed it could have been automated \cite{rap,goldacre,turing-way}.
+If automated, it could then be repeated reliably --- unlike a manual cut and paste which is error-prone every time it is done.
+It can be repeated if any experimental data, literature, or other knowledge changes, and the paper's analysis brought up to date with ease.
+In particular, any other researcher, whether part of the authorship team or a later reader of the paper, can reproduce it reliably if the RAP process is made available.
+Table~\ref{figure-RAP-summary} provides a brief summary of RAP principles.
+
+For example, if the paper in question is a systematic review, it could be kept current by automatically re-running the programmed atomic actions that it was built with.
+Indeed, this ability is one of the original motivations of RAP, where Government agencies could easily generate up to date reports on request without having to repeat all the manual work and risk making procedural errors doing so.
+Furthermore, every time they update a publication, the RAP pipeline itself is reviewed and improved, so the quality of the reproduced work improves --- unlike in a non-RAP process where new errors are generally introduced every time the work is revisited.
+
+RAP not only helps develop reproducible science, and improve the quality of the science as the authors debug and refine their methodology, it also provides a precise audit trail that can be used to protect against fraud, as discussed in \cite{MMR-fraud}, and can in principle be largely automated and perform checks much faster and more efficiently than conventional \emph{post hoc\/} investigations.
+
+RAP embodies Donald Knuth's comment, also quoted after this paper's abstract,
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0
+Science is what we understand well enough to explain to a computer'' }
+\\ \hfill from the foreword to $A=B$ \cite{a=b}
+\end{quote}
+
+The corollary is that if we are doing arbitrary cut and paste that has not been programmed into a computer, then we are not doing good science.
+Science is in principle an algorithmic process, and therefore, as Knuth says, if we understand well enough what we are doing in science, we can explain it as code, as RAP, for a computer to run and rerun.
+
+\section{The statistics/code analogy}\label{statistics-computation-analogy}
+The central role of computational methods in science may be fruitfully compared to statistics, an established scientific tool.
+
+Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics.
+Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.
+
+In good science, all statistics, methods and results are reported very carefully and in precise detail \cite{apa-statistics,Mariusz-reporting,science-statistics-recommendations}, generally following strict journal or disciplinary guidelines.
+A statistical claim in a paper might be summarized as follows:
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
+[\ldots] Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.0009)$.
+Between group effect size $d = 0.55$ $(95\%$ CI $0.32,$ $0.77)$.''} \cite{example-stats}
+\end{quote}
+
+This wording formally summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
+It is a \emph{lingua franca}.
+It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results.
+Moreover, behind any such brief paragraph is a substantial, rigorous, and appropriate statistical analysis.
+
+Scientists write like this and conferences and journals require it because statistical claims need to be properly accountable and documented in a clear way.
+The journal \emph{Science}, for example, in its many explicit and quite technical statistics requirements requires
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 Adjustments made to alpha levels (e.g., Bonferroni correction) or other procedure used to account for multiple testing (e.g., false discovery rate control) should be reported.''} \cite{science-statistics-recommendations}
+\end{quote}
+
+Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Spiegelhalter-section}).
+Results should not be uncritically accepted just because they are claimed.
+The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted; in fact, there is widespread concern about the poor quality of statistics in science \cite{paul-hci,pnas-stats}.
+While it is assumed that statistics should be peer-reviewed, and that review will often lead to improvement, these critical papers show that reviewers and editors are often failing to pick up on poor statistics.
+
+Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement.
+Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians.
+Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.
+
+Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them, will be generated by computer.
+Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it.
+Many papers now explore the contribution of AI and ML to their fields.
+The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers.
+Computers even help with the word processing and typesetting of the research.
+
+In short, computers, data, and computer code are central to modern science, not just to the explicitly computational sciences.
+Some AI work is uncovering biases and ethical issues that were previously unrecognized, so computational sciences are not just routinely contributing to existing science but extending its reach and improving its quality.
+
+However, using any code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
+Software code, like statistics, is subject to unintentional bias \cite{Ben,se-bias}.
+All these issues are non-trivial concerns requiring technical expertise to manage well.
+As with statistics, good answers to such ``technical'' issues makes the science that relies on them better.
+
+A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model was reliably transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to, and possibly even no link to the code at all.
+
+Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims.
+Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly planned and reported code are widespread.
+
+We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
+
+An interesting overlap with statistical and Software Engineering sloppiness concerns the many papers that  disclose as part of their methodology that they used a particular software package, for example
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA).''} \cite{sas9.2}
+\end{quote}
+
+but without giving any further details.
+The paper cited above describes one of its multivariate analyses as ``multiple correspondence analysis (MCA) and ascendant hierarchical clustering'' with no specific details --- reproducing the work would be non-trivial since the methodology would have to be reconstructed from scratch, even assuming the data was made available by the authors as the paper gives no details of the data used or how to obtain it.
+
+The problem is that the common practice of declaring using a named computational system (such as SAS in this case) does not help scrutiny in the least, as such systems can do almost anything.
+\emph{How\/} those analyses might have been performed is not discussed, and one assumes it follows that the analyses could therefore not have been properly reviewed for basic scientific competence during the publication process.
+
+A reviewer, if nobody else, needs to actually examine the code used and its documentation to assess whether the analysis presented in the paper is appropriate and sufficiently reliable.
+Furthermore, if the analysis in this case actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible as it relies on idiosyncrasies in SAS version 9.2.
+Of course, an author can disclose the idiosyncratic dependencies; while this seems to be an onerous obligation, conversely it is arguable that if an author is unaware of dependencies, then their science relying on them is equally unreliable.
+
+It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, as well as to uncover unknown dependencies.
+Indeed, the code may be poorly written (as this paper shows it is), so the results derived from the code simply may not be reliable.
+
+\label{section-vicious-cycles}
+In normal scientific reporting (outside of teaching and assessing science) details of methodology are routinely glossed.
+A chemist does not say they cleaned their glassware.
+One might argue, then, that scientists need not discuss their code in detail because they know how to program and their code is correct.
+This argument is mistaken.
+Code is rarely considered a valuable part of the science to which it contributes, which creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.
+
+%\color{blue}
+\input generated/over-fitting-code-section.tex
+%\color{black}
+
+\section{The conventional role of code}
+
+Models map theory and parameters to describe phenomena, typically to make predictions, or to test and refine the theory supporting the models.
+With the possible exception of theoretical research, all but the simplest models require computers to use; indeed even theoretical mathematics is now routinely performed by computer.
+
+Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive.
+The chance that a thousand lines of code is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability.
+How to do this is the concern of Software Engineering.
+
+While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are they are long, complex, inscrutable and (as our survey shows) lack adequate documentation.
+Therefore the models are very likely to be unreliable \emph{in principle}.
+
+If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand their historical thinking well enough (say, just a few weeks or months later) to maintain it correctly if it has not been clearly documented?
+As a scientist pursues a research career building on their previous work, how can they be certain their work is reliable, and not merely converging to their  prejudices?
+Without proper documentation, including a reasoned case to assure that the approach taken is appropriate \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?
+
+Without substantial documentation it is impossible to scrutinize code properly.
+Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications.
+What does it mean?
+What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation?
+Does the code mean what was intended?
+What are the assumptions on \texttt{k}, \texttt{x}, and \texttt{y}, and do they hold invariantly?
+Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown.
+A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of under- or overflow).
+
+Without explicit links to the relevant models (typically mathematics), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code.
+Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim.
+If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed.
+It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.
+
+Some authors assert that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarizing data or properties of the data \cite{assessing-quality}.
+In reality, if code is inadequate, any so-called ``insights'' will be flawed, and flawed in unknown ways.
+Indeed, none of the papers sampled (described in \supplement\ section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
+
+Clearly, like statistics, coding can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between the extremes.
+The question is whether it matters, \emph{when\/} it matters, and, if so, when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work?
+
+\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
+It is a misconception that programming is easy and even children can do it \cite{fixit}. More correctly, toy programming is easy, but mature programming is very difficult.
+
+An analogy helps justify this key point.
+Building houses is very easy --- indeed, many of us have built toy Lego houses.
+Obviously, though, a Lego house is not a \emph{real\/} house.
+It is not large enough or strong enough for safe human habitation!
+This point is obvious because we can see Lego houses, and everyone is familiar with building-block play.
+Its real-world engineering limitations are too obvious to need stating.
+
+In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also made invisible.
+The ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed, and because --- like Lego --- toy programs can look inspiring but be unreliable, difficult to use, even dangerous.
+
+Saying programming is easy is like appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations.
+These are professional engineering issues that Lego builders ignore.
+Certainly, even real building is much easier and faster when the technical details are ignored, as anyone who has experienced a cowboy builder can attest.
+
+Unlike building houses (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture.
+Professional standards, even when they exist, are not enforced.
+
+Problems for the reliability of science arise when doodling and tweaking software drifts into claiming scientific results that do not have reliable engineering processes or structures underpinning them (let alone the properly developed and documented accessible code) to justify them.
+
+In many countries, there are laws that require all but the very simplest building structures to be formally approved from plans and inspected as they are built, but who writes plans for software, who inspects scientific models while they are being coded?
+Yet the consequences of building a shoddy garage have negligible impact compared to the consequences of writing poor code that informs national public health policies or climate change interventions.
+
+\subsection{The low status of coding}
+Since programming appears to be so easy, developing code has a correspondingly low status in scientific practice (and more widely).
+Developers of code are rarely acknowledged in scientific papers.
+The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the contributors to it.
+Because it is apparently easy, there is no need to work hard to make it correct.
+Because of the ease of over-fitting (section \ref{over-fit}), code ``works well'' with little skill or effort.
+While such mistaken views prevail, the vicious cycle is that the low status means software development is casualized, which reinforces the low status.
+
+Almost all scientific papers \emph{routinely\/} describe their experimental method, their data handling, and provide an overview of their analytic (usually statistical) methods.
+If they are theoretical papers, they will describe their mathematical models and data that is used to run or test their models.
+Outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods were used avoid or control for experimental error.
+
+%An example is the documentation at the code repository \cite{BayesWaveCode}, which requires three high-profile \emph{{Physics Review D}} papers \cite{BayesWave1,BayesWave2,BayesWave3} be cited to acknowledge any use, but only one of them, \cite{BayesWave3}, mentions the code, and even then it fails to acknowledge the contributions of its 10 programmers (despite 6 being co-authors).
+The paper does cite the repository URL, but unlike the papers it cites conventionally, it omits all the code's authors.
+Acknowledgement does not seem to matter, even to the people who did the work.
+
+In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code,TOP}, because  code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.
+Curiously, the importance of \emph{correct\/} code is rarely emphasized.
+
+Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement).
+Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.
+
+\subsection{The critical role of code is often ignored}\label{central-role-of-code}\label{PRISMA-statement}
+Because statistics, like code, is so readily susceptible to uncontrolled bias and error, there are many protocols and journal policies that enforce best practice, for example journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature.
+Yet PRISMA, like many such policies, ignores the critical role of code, and ignores the Software Engineering principles that assure code that research relies on is reliable and reliably reported.
+
+PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which is reasonable enough.
+PRISMA covers the review process carefully.
+For example, the authors should report the number of papers they included in their review.
+Perhaps $N=2000$.
+This number is then written into their draft paper, perhaps in several places.
+As the authors read and revise their paper, submit it, and respond to peer-reviewers, it is very likely that the number of papers in the survey changes, or other numbers and details may change.
+
+The authors now have a maintenance problem: where are the numbers that have changed, and what should they be changed to?
+Doing a search-and-replace, whether automated or by hand, is fraught with difficulties.
+What happens if $2000$ is used for some other purposes as well?
+What happens if some of the $2000$ values are written as $2,000$ or as $2000.0$?
+What happens if some $2000$ are year dates and are changed incorrectly?
+Then there are the Human Factors: slips and errors will happen in this process anyway \cite{engima}.
+Typos, slips during cut-and-paste, and other errors are common.
+Similar iterative revision cycles happen with any paper, not just with systematic reviews.
+PRISMA rightly deals with a wide range of methodological problems, but ignores the methodological problems (such as the example earlier in this paragraph) that using code na\"\i vely raises.
+
+The irony is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations.
+Such rule schemes reinforce the fallacy that code is trivial and unimportant.
+PRISMA consciously warns about  methodological issues, but ignores that poorly managed code raises comparable methodological issues that undermine the validity of the research it supports.
+
+\subsection{Bugs, code and programming}\label{knowledge}
+Critiques of data and model assumptions are increasingly common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned.
+Yet data and program are formally equivalent (see \supplement, section \ref{on-code-data-publication}).
+Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and probably useless.
+Code, however, is harder to scrutinize, which means that errors in code have subtle, often unnoticed effects on results.
+
+Almost all code contains ``magic numbers'' --- that is, data masquerading as code.
+This common practice ensures that published data is very rarely all of the data because it omits the magic numbers embedded in the code.
+Such issues emphasize the need for repositories to require the inclusion of code so all data, including that embedded in the code, is actually available.
+
+Bugs can be understood as discrepancies between what code ought to do and what it actually does.
+Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered.
+Contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination, or, better, failure to compile.
+None of the science surveyed in this paper includes any such basic techniques.
+
+Random numbers are widely used in computational science (and in many of the papers surveyed), for simulation or for randomizing experiments.
+Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of bugs nai\"\i ve bugs \cite{knuth}.
+
+If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
+Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time).
+For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted, but are bugs in the wider sense of producing incorrect results (e.g., ill-conditioning); that is, what was intended was~wrong.
+
+\subsection{Long-term problems of unreliable code}
+Scientists explore and extend the boundaries of rigorous knowledge.
+Put briefly, the purpose of scientific experiments is to vary details to either test and specify the boundaries of theories, or to discover new phenomena that then lead to theory revision.
+
+If poor code, or poorly documented code, is made available with scientific papers, the code is a natural place to start replicating and varying experimental conditions,  including both data or code.
+However, if the starting point is not accurately known, whether due to bugs, obscure code, or because of poor documentation, then experimental variations will have an unknown effect.
+Theory will then be driven by artifacts of the code, not genuine phenomena.
+
+In section \ref{section-pandemic-modeling}, below, an example is documented of a research code development process of at least 15 years' duration where the code was admitted to be completely undocumented, leaving details in just one author's head.
+None of the various related papers describe any controls over the drift of the science, or how independent researchers building on it might have been able to build with confidence rather than merely reproducing the same errors.
+
+Since the code in question was substantial and non-trivial, it is very unlikely that any constructive reproduction occurred outside the original laboratory and mindset; indeed, section \ref{reproducibility-concerns} describes how ``reproduction'' became trivialized because of community pressure to confirm the insights of this particular research.
+
+Trying to constructively refute aspects of this research in the Popperian sense \cite{popper-conjectures-refutations} would have been impossible.
+For example, had the relevant papers published critical code invariants then scientists building on the research could have explored whether those invariants remained valid and, if so, under what assumptions.
+In fact, invariants are the theories of code, and deserve as high a prominence in published  science as the domain theories the code itself is supporting investigating.
+
+%Since the symmetries in the scientific domains should be reflected in corresponding symmetries in well-designed code, this raises new possibilities of Formal Methods approaches (e.g., using automatic theorem provers) supporting advances in general science.
+
+\section{State of the art}% \\ in computational science}
+
+\subsection{Case study: Pandemic modeling}
+\label{section-pandemic-modeling}
+A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modeling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer-reviewers, there is no role for code or computation as such.
+It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.
+
+A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
+The review \cite{diagnosis-reviews} itself ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed.
+Indeed, TRIPOD guidelines ignore code completely.
+
+It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for human use.
+Flowcharts, too, should be designed as carefully as code, for exactly the same reasons.
+
+A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand, when it did not model air travel or other factors required for later western COVID-19 modeling.
+The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.
+
+A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 
+For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
+''} \cite{ferguson-interview}\end{quote}
+
+This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
+The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
+(Typical $h$ values vary by discipline.)} of 93.
+Ferguson should be well aware of the standards of coding used in his own field.
+Arguably, then, this comment quoted above is representative of the culture of the field.
+
+Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}
+
+Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
+
+If Ferguson's C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation.
+It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified and repurposed over fifteen years without becoming incoherent.
+
+If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
+In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
+
+The programming language C which was used \cite{tweet} is, like many popular programming languages, not a dependable language; to develop reliable code in C requires professional tools and skills.
+Some of the code was written in a na\"\i ve style (e.g., writing \texttt{*(a + i)} instead of \texttt{a[i]}, and with obscure numerical goto statements like \texttt{if(l == 0) goto S150}), and with C code that was translated simplistically from FORTRAN and Pascal code, from references dating from the 1970s and 1980s \citeeg{forsyth,Normal-Distributions}.
+
+Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, and hardware.
+In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C should be used: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}.
+Alternatively, a high integrity programming language, unrelated to C, such as SPARK Ada \cite{spark}, or modern languages (many related to the ``ML family'') like OCaml, F*, Haskell \cite{haskell}.
+These languages have learning curves; however their key benefit is that correct programs are far more likely and are \emph{much\/} faster to write.
+(The \supplement\ discusses these issues further.)
+
+Ferguson, author of the code, says of the criticisms of his code,
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 
+However, none of the criticisms of the code affects the mathematics or science of the simulation''} \cite{thumbs-up}
+\end{quote}
+
+This claim is implausible.
+
+The original work on theoretical epidemiology may be fine if it does not use any of his code, but if the science is not supported by code that correctly implements the models, then the program's output cannot be relied on without independent evidence.
+Over the fifteen plus years the code was in development the science it informs will have developed too, as will the relevant data; it is not clear how they will have remained in alignment.
+
+\label{discuss-covid-sim}
+Typically, models will be developed iteratively as their results are improved to better fit a scientist's goals --- but this, especially when it is done by tinkering, as here --- risks making the code arbitrarily fit the goals (that is, over-fitting; see section \ref{over-fit}), rather than to objectively elucidate the science.
+
+In fact, the Ferguson code, \texttt{covid-sim}, is very large at \covidsimkLOC\ kLOC (thousands of lines of code),\footnote{Ferguson's \texttt{covid-sim} system is composed of \covidsimfiles\ files, and uses \covidsimdata\ of data.
+It is now rewritten from C into C++ with Python, R, sh, YAML/JSON, etc.
+For more details, see \supplement.} so it is implausible that the ``mathematics or science'' has been correctly implemented in it.
+Ferguson's \emph{reported\/} science unlikely to be reliable.
+
+\subsection{Concerns with reproducibility}
+\label{reproducibility-concerns}
+Getting science right, which now, in turn, depends on correct code, is a normal requirement of \emph{reproducibility}.
+
+The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed that the code could be run again and produced comparable results (compared using an Excel spreadsheet).
+This weak test would pass provided the runs gave the same invalid answers --- it is not usefully a stronger test than just checking that the code compiles.
+
+Running code to obtain results claimed in a paper is a very weak test, and anyway one that should have been checked automatically during paper preparation and submission.
+If you do not know what you are reproducing, as is the case in this \emph{Nature\/} paper, there is little scientific value in doing so.
+
+Unfortunately, the terms reproducibility, replicability, and repeatability, have similar meanings in English but have been used in different specific technical ways by different authors.
+In \cite{codecheck,thumbs-up} the reproduction amounted to just re-running the original code.
+It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper; if the original code runs this confirms a basic level of access for the wider scientific community.
+But a more realistic criterion than basic reproduction in this sense is whether an \emph{independently\/} developed model developed from the same paper(s) specifications produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.
+However, much stronger scrutiny of code than ``reproduction'' is required to answer essential questions including:
+
+\newcount \enum
+\enum=1
+\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
+\the\enum
+\global\advance \enum by 1}
+\begin{enumerate}\raggedright
+\item Is the code valid: does it do what the paper claims?\footnote{Of course, the underlying science may be wrong to, so it is useful to distinguish \emph{internal validity\/} and \emph{external validity}.
+Internal validity occurs if the code does what the paper claims; external validity occurs if the code represents correct science (which the paper may have interpreted incorrectly.}
+\item Do other scientists, including reviewers and the authors, understand the code?
+\item Does the code implement the methods described in the paper?
+\item Has the code been over-fitted or tweaked to support specific claims in the paper?'
+\item Is there a definitive version of code?
+\item Is the code controlled and signed?
+\item What limitations does the code have?
+\item Was the code developed to any standard, and does it comply to that standard?
+\item How does the code protect against data, coding, and human error?
+\item Was the code tested adequately?
+\item Does the code depend on arbitrary parameters, data, or code to over-fit to obtain the published results?
+\item Is the code documented adequately, so we know what it is trying to do, and how?
+\item \ldots\ and so forth --- questions are numbered for reference. 
+\end{enumerate}
+
+All such questions also apply to specifications, documentation, assurance cases, test procedures, and other essential documents, not just to code.
+In turn, the levels of scrutiny demanded should be guided by explicit claims in the paper \cite{essence-of-software} --- for example, a pilot study requires weaker assurance than code that is developed concerning nuclear power, driverless vehicles, public health, etc.
+
+The questions in the list above are certainly hard to answer for all but the briefest code, but corresponding levels of quality assurance are demanded for other methodologies \cite{notebooks,popper-conjectures-refutations,tripod,prisma,nih-policy,nih-nature}, such as data preparation and statistics to support claims in peer-reviewed science.
+
+Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now open-source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).} Documenting code in hindsight, even if done rigorously, may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was originally intended to have done.
+As the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models; it will be different.
+
+It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
+The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
+
+Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
+Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications.
+Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it instead of making actual code available is worse than not publishing code at all (see \cite{chinese}).
+Pseudo-code is too vague to help anyone scrutinize a model; moreover, pseudo-code may mask over-fitting in code used that is not explicit in the pseudo-code.
+
+An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}.
+
+The \supplement\ provides further discussion of reproducibility.
+
+\subsection{Beyond pandemic modeling}
+\label{section-science-beyond-pandemic-modeling}
+
+Epidemiology has a high profile because of the COVID pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
+But other examples that were not rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code.
+Its supplementary material, which might have provided code, no longer exists.
+
+There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact.
+Climate change modeling is one such example that will have an impact long beyond the COVID pandemic.
+
+A short 2022 summary of typical problems of Software Engineering impacting science appears in \emph{Nature\/} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration.
+The paper \cite{nature-review}  makes some misleading comments about the simplicity of Software Engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''
+
+Guest and Martin promote the use of computational modeling \cite{psychological-modeling}, arguing that through writing code, one debugs scientific thinking.
+Psychology, their focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}.
+In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived.
+Computational models can be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks.
+It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely!
+
+For instance, ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973, and is now a 120 kLOC Common LISP and Python system \cite{actr}.
+Furthermore, any paper using ACT-R would require additional code on top of the basic ACT-R framework.
+
+The psychology paper \cite{psychological-modeling} presents an example computational model from scratch to illustrate a framework of computational science.
+In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not modeled.
+Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.
+
+The Python code is presented without derivation; Software Engineering is ignored.
+The program listed in the paper certainly runs without obvious problems (ignoring some typographical errors due to the journal's publishers), but ironically the Python does \emph{not\/} implement the mathematical specification explicitly provided for it, thus undermining the argument of the~paper.
+
+One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses.
+Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct.
+Otherwise, code based on inadequate Software Engineering will introduce errors into scientific thinking.
+
+Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims.
+People reading the code will not know how to obtain results because they do not know exactly what was intended in the first place.
+The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or flaws.
+
+Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}.
+World-class software can attain maybe 1 bug per 1,000 lines of code.
+Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new ``insights'' into their ideas, unable to distinguish genuine insights from artifacts of bugs.
+In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- is subject to numerical errors \cite{hamming}.
+It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion).
+Such bugs undermine confidence in model results.
+
+Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what the researchers have done and whether that corresponds closely to what they are reporting.
+
+Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available.
+While some \citeeg{unfinished,lancet-unfinished} make what they call ``documented'' code available, they provide no more than superficial comments.
+This is \emph{not\/} documentation as properly understood.
+Such comments do not explain code, explain contracts, nor explain algorithms.
+Contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming.
+
+Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper, and indeed in computer science research, e.g., \cite{machine-learning-reproducibility}) is inscrutable.
+Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do.
+Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.
+
+\label{section-discussion}
+\subsection{The narrow emphasis on data}
+\label{critique-fair}
+Data has been at the center of science, certainly since the earliest days of astronomy collecting planetary and other information.
+Today it is widely recognized that lack of accessible and usable data that has already been collected limits the progress of science.
+Low quality data and poor access to data causes reproducibility problems, an increasingly recognized problem --- in 2015 it was estimated that \$28 billion a year is spent on preclinical research that is not reproducible \cite{preclinical-reproducibility}.
+
+Curating data is taken seriously as a part of normal science and peer-reviewed publication.
+Journal policies widely require appropriate discussion of data, much as they require appropriate discussion of statistics.
+Journals often require archiving data in standard formats so it can be accessed for reproduction in further scientific work.
+
+There are many current activities to proceduralize and standardize the more effective curation and use of data, such as the FAIR principles (Findable, Accessible, Interoperable and Reusable) for scientific data management and stewardship \cite{fair,fair-principles}, and in the development of journal and national funder policies.
+For example, the 2022 update to the US National Institutes of Health data policies \cite{nih-policy} is described as a ``seismic mandate'' by \emph{Nature\/} \cite{nih-nature} in its attempt to improve reproducibility and open science yet they ignored code.
+
+These cost estimates and initiatives under-play the role of code as a critical component despite its becoming the new laboratory for almost all science.
+The role of code specifically in modeling is discussed throughout this paper; without bespoke code, proposed models (unless intended to be abstract) cannot make a quantifiable contribution to the literature.
+Code has additional problems of versions and compatibility beyond those of data, for example suitable compilers to run old code may no longer be available, and  programming systems may produce different results when used on different computers.
+
+In general, without proper management of code --- for example to record, detect and report version control differences --- sharing code may even be counter-productive.\footnote{The data and code shared with the present paper includes cryptographic checksums; if somebody reproducing the work described here does not obtain the same checksums at least when they start their work, then there are problems that need investigation before relying on the reproducibility of the data.}
+
 Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
-The evidence suggests that, generally, some but not all manually developed code is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no clear evidence (over the survey period) that any published code was maintained using the repositories.
+The evidence (see \supplement) suggests that, generally, some but rarely all develop code that is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no evidence (before, during, or after the date of the survey sample) that any published code was prepared or maintained \emph{using\/} repositories.
+This is consistent with the finished code being uploaded to a repository just for the purposes of satisfying publishing requirements, but not using one earlier probably because they did not understand the benefits of doing so.
 
-The \supplement\ provides a summary of the analysis and full citations for all papers in the sample.
+Using a standard repository for lodging a paper's supporting code would help other scientists access the code easily, but not using the repository for developing and maintaining the code means the author of the paper misses out on many helpful features of repositories, such as version control, open-source development and review, actions and other approaches for automating development, sharing workload, and so on.
 
-\subsection{A call to action}\label{summary}
+In summary, there is a lop-sided emphasis on data in science.
+In fact, data is useless without code, and code must be used to analyze data and can manipulate it in arbitrary ways --- some ways, like renormalization and other transformations, deliberately, and some accidentally caused by bugs and unwittingly misunderstanding programming techniques.
+Often code is used to extrapolate data, so the code itself effectively generates more data, or the code eliminates outliers so it effectively deletes data.
+Data is routinely formatted in standard ways (else code could not process it), but code is generally very architecture- and version-specific, so --- unless properly managed --- code goes obsolete faster than data.
+And so on.
+In short: the integrity of code and its availability to scrutiny is in fact both harder and more important than the usual requirements put on data.
+
+\section{Extending RAP to \RAPstar}\label{RAPstar-section}
+Code is usually seen as an independent set of files that are run to generate results, typically to be copied into a paper.
+But we can go further: a paper can itself embed code or become code \citeeg{relit,Executable-Paper}.
+This supports the generalization of RAP to form \RAPstarp.
+Essentially, \RAPstar\ is the recognition that computing is not just about programming computers (which results in RAP) but is about applying computational thinking \cite{computational-thinking-wing,computational-thinking} that supports and constructively analyzes any process, and in particular human and scientific processes that may not normally be computer-based at all.
+
+Once processes in the pipeline are automated, this means that there is code to run the steps again.
+Once there is code, it can be managed in a version control system.
+A version control system then provides an audit trail for free, as well as many advantages such as being able to backtrack to an earlier version to review earlier edits.
+Importantly, code can also perform sanity checks on the process --- a very simple example is automatic bibliography systems that check that journal names, DOIs are correct, and so forth.
+(They also allow the bibliographic data to be pooled and curated with other scientists, which improves its scope and quality.)
+
+GitHub is a tool that provides \emph{actions\/}, which are named specifications to run RAP's analytic pipelines, or workflows in GitHub's terms.
+GitHub happens to specify actions in the language YAML, which, being a textual notation, in turn means that all the helpful features of GitHub --- open-source, version control, and so on --- can be applied to the pipelined processes as well.
+Research pipelines can thus be made explicit, documented, shared and improved with open collaboration.
+
+Almost the entire scientific process can be automated (even with its interaction with the world automated by sensors and robots).
+There are many ways to do this; for example, \emph{Mathematica\/} makes the analysis of the data and the calculations and the paper ``the same thing'' in its integrated notebook user interface.
+The many alternatives include R Markdown, an approach based in the open mathematical system R \cite{RMarkdown}; a system, Lepton \cite{lepton}, which allows a \LaTeX\ document to execute and include arbitrary code, and language-independent notebook systems like Jupyter; and so on.
+There are also variations of literate programming, notably relit \cite{relit}, that enable papers themselves to generate the code they rely on and may describe, which therefore ensures the code run to support papers is exactly the code the author ran to support the claims.
+
+In all such systems, running the computational paper creates the publication.
+Indeed, every time the paper is run, the authors are likely to check the results after they have been re-computed, so the RAP process actively helps reduce errors.
+
+Here is the insight: the code, just like the paper is text, so the code \emph{itself\/} can fully form part of the pipeline, made reproducible and benefit from all the RAP benefits.
+To date, this critical point has been overlooked.
+Since using RAP to improve code, rather than just the pure scientific pipeline, has not been mentioned previously, we call it \RAPstar\ to make it clear this is a new and important generalization.
+\RAPstar\ helps improve code quality, for the same reasons RAP improves the quality of the paper and the science.
+
+Software engineers have many tools for automatic code development (such as Unix's \texttt{make}) but the idea that these tools can be used to integrate and help automate code authoring as well as its documentation \emph{and\/} paper authoring is radical.
+Amongst other things, it means that the \emph{entire\/} research and development process of the paper \emph{including\/} all its underlying code can be reproduced and reused by others.
+The present paper is a modest example of \RAPstarp; more details are given in the \supplement.
+
+Note that as \RAPstar\ objectifies how science is done to a standard sufficient to enable a computer to run it.
+\RAPstar\ enables all of the methodologies of Software Engineering to be brought to bear \emph{on the science itself}.
+\RAPstar\ means the normally tacit, manual, and undocumented processes of science become explicit code.
+Code can then be scrutinized, optimized, and ensured correct by standard Software Engineering practice --- thus \RAPstar\ does not just automate science for reproduction, it makes the automation explicit so the doing of science itself can be reasoned about --- not just mentally but supported by sophisticated tools, such as theorem proving and AI\@.
+Science will be improved.
+
+\section{The paper as a scientific laboratory}\label{paper-as-lab}
+The conventional view of science is that experiments are done \emph{then\/} written up. 
+%RAP and \RAPstar\ support and enhance this process with computational tools; in particular, the archival results of research are not just the reports and publications, but include the executable processes --- the pipelines in RAP terms --- of the research.
+However, it is more productive to think of the paper itself as an active laboratory, not just as a record of finished work.
+Viewing the paper itself as a laboratory encourages authors to copy and adopt laboratory best practice (such as keeping records, as RAP and \RAPstar\ suggest) into the preparation of the paper itself; it also encourages authors to see writing as a scientifically --- not just expressive --- creative act, and not just as the final summary of a period of scientific creativity that is not written down.
+In short, seen as a laboratory, writing the paper is no longer a reactive write-up of finished work, but it is an active part of doing the science itself.
+Writing a paper thus explores the space of scientific possibility as constructively as working in the field or on a bench: the paper is a laboratory.
+
+With a computational paper, authors can literally experiment \emph{in\/} the paper, exploring the effectiveness of ideas and explanations.
+Furthermore, they can experiment with hypotheses: for example, they can make a clear claim that at that point in the lifecycle of the paper is but a wish rather than an established fact --- sketching a direction they plan to go in and develop supporting arguments and evidence for.
+So they then do the experiments or calculations (including consulting other scientists and peer reviewers) to establish a justification and other details.
+Typically, the evidence they generate or the criticism they receive may not be quite what they expected, so they then revise the claim to be correct, or change it to be a more realistic claim still as yet to be fully substantiated, or they could delete it if it just turns out to be a mess, or they could develop some altogether much better approach to the work as a result of the exploration.
+
+Typically, many ideas in the paper will be linked to data from conventional experiments.
+For example a computational paper might calculate that the statistical power of an experiment is too low (there is an unacceptable risk of committing Type II errors), so the authors will decide to improve the experiments.
+
+If there is code in the paper, every time it is typeset, the code will be run.
+Therefore the authors of the paper proof-reading the results of the code will have opportunities to try to debug and improve the code.
+Again, the paper itself is acts like a laboratory, helping the authors refine the science.
+
+Note that systems capable of handling computational papers (including \LaTeX) can create conditional documents: for example, there could be a flag \texttt{publish}.
+If \texttt{publish} is false, the author can see all their private work and thinking, including all their experimental thinking and workings; but if \texttt{publish} is true, the paper would be typeset as for a submitted paper with the detailed working concealed to ensure a clean and concise presentation.
+In principle, also submitting to the journal with \texttt{publish} false (hence with data and workings visible) could satisfy journals' code and data requirements.
+Of course, when there is a large body of data or code, they need not all be an explicit in the computational paper: they would be made available separately in the usual way.
+
+The more we view the paper as a proactive scientific laboratory, the more we gain from the \RAPstar\ perspective, and the more science gains from improved, conceptually broadened, reliable, and reproducible science.
+The more we will want to engage mature Software Engineering standards too, because the quality and creativity of future science relies on them.
+
+\section{Call to action}\label{summary}
 Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide.
-Given the urgency of rigorously understanding COVID-19, any epidemic for that matter, it is essential that epidemiologists engage professional software engineers to help develop reliable laboratory methodologies.
-For instance, code lab books can be generated and signed easily.
 
-Software used for informing public health policy, medical research or other medical applications is \emph{critical software}.
-Professional critical software development, as used in aviation and the nuclear power industry, is (put briefly) based on \emph{correct by construction}: \cite{cbc} effectively, design it right first time, {supported by numerous rigorous techniques, such as formal methods, to manage error.
-(See extensive discussion in this paper's \supplement.)}\ Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable.
-Conversely, not following these practices undermines the rigor of the science.
+Inadequate code can be extremely problematic.
+Incorrect results might be used for supporting science, modeling COVID-19, informing public health policy, medical research, or used in other \emph{critical software}.
+Professional critical software development, as used in critical industries such as aviation and the nuclear power, is (put briefly) based on \emph{correct by construction\/} \cite{cbc}, effectively: design it right first time, supported by numerous rigorous techniques, such as Formal Methods, to manage error.
+Not coincidentally, such are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable, as required for supporting quality science.
+Conversely, not following these practices undermines the rigor of science.
 
 An analogous situation arises in ethics.
-There are some sorts of research that are ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
+
 Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems.
+Planned research may be ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
 National funders, and others, therefore require Ethics Boards to formally review ethical quality.
-Medical journals will not publish research that has not undergone appropriate ethical review.
+Medical journals will not publish research that has not undergone appropriate formal ethical review.
 
-Analogously, and supplementing Ethics Boards, Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality software engineering.
-Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to epidemic modeling papers --- and in due course, all scientific papers --- being required to include Software Engineering Board clearance statements as appropriate.
-{Software Engineers have a code of ethics that applies to \emph{their\/} work in epidemic modeling \cite{ethics-code}.}
+Analogously, and supplementing Ethics Boards, it is proposed that Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality Software Engineering to support research and publication processes.
+Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to  all scientific papers and funded research being required to include Software Engineering Board support and clearance statements as appropriate.
+Note that Software Engineers themselves have a code of ethics that applies to \emph{their\/} work \cite{ethics-code}.
 
-{The present paper did not explore data, because in almost all cases the code and data were explained so poorly and archived so haphazardly it would be impossible to know whether the author's intentions were being followed.\footnote{{For the present paper, all the code, data, analysis and documents are available for download in a single zip file.}} Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
+Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
 Ideally, the authors should test a clean deployed build of their code and save the results.
 Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims.
-This requirement is no more unreasonable than requesting good statistics, as discussed in the opening of the paper.
-And the solution is the same: relevant experts --- whether statisticians or software engineers --- need to be routinely available and engaged with the science.
-SEBs would be a straight forward way of helping achieve this.}
+This requirement is no more unreasonable than requesting good statistics, as discussed earlier.
+And the solution is the same: relevant experts --- whether statisticians or Software Engineers --- need to be routinely available and engaged with the science.
+Software Engineering Boards would be a straight forward way of helping achieve this.
 
 There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university.
-Active, professional software engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art.
+Active, professional Software Engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art.
 There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.
 
-Open Source generally improves the quality of software.
-SEBs will take account of the fact that open source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers.
-Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis {and other quality control methods for which the lead scientist and software engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).}\ {Software engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer reviewed publications \cite{acm-artifacts}.}\ {See this paper's \supplement\ for further suggestions.}
+Open-source generally improves the quality of software.
+SEBs will take account of the fact that open-source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers.
+Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis and other quality control methods for which the lead scientist and Software Engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).
+Software Engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer-reviewed publications \cite{acm-artifacts}.
 
-A potential argument against SEBs is that they may become onerous, onerous to run and onerous to comply with their requirements.
-A more mature view is that SEBs need their processes to be adaptable and proportionate.
+A potential argument against SEBs is that they may become onerous, onerous to run, and onerous to comply with their requirements.
+A more mature view is that SEBs need their processes to be adaptable and proportionate; indeed, few people consider Ethics Boards to be disproportionately onerous.
 If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation.
 Hence SEBs processes are likely to follow a risk analysis, perhaps starting with a simple checklist.
-{There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar.
-Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.}
+There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar.
+Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.
 
 A professional organization, such as the UK Royal Academy of Engineering ideally working in collaboration with other national international bodies such as IFIP, should be asked to develop and support a framework for SEBs.
-SEBs could be quickly established to provide direct access to mature software engineering expertise for both researchers and for journals seeking competent peer reviewers.
+SEBs could be quickly established to provide direct access to mature Software Engineering expertise for both researchers and for journals seeking competent peer-reviewers.
 In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations.
@@
 SEBs are a new suggestion, providing a supportive, collaborative process.
-Methodological suggestions already made in the literature include open source and specific software engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}.
-While \cite{ABCs-SE} provides an insightful framework to conceptualize approaches and compare their merits, there is clearly scope for much further research to provide an evidence base to motivate and assess appropriate interventions to help scientists do more rigorous and effective software engineering to support their research and publishing.
-These and further issues are discussed at greater length in the \supplement.
+They meet Tony Hoare's comments about the value of rigorous management of procedures \cite{hoare-unreasonable}, and widen them to non-programmer scientists.
+Methodological suggestions already made in the literature include open-source and specific Software Engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}.
+Reference \cite{ABCs-SE} provides an conceptual framework.
+However, there is scope for further research to provide an evidence base to motivate and assess appropriate interventions (such as those proposed in this paper) to help scientists do more rigorous and effective Software Engineering to support their research and publishing.
 
-\subsection{Suggestions for further work}
+An analogous proposal to SEBs has been made for Methods Review Boards \cite{mrb}, to help scientists ensure the methods they use are appropriate for addressing their research questions.
+The Methods Review Boards was motivated by an Ethics Board member noticing that often experimental methodologies are inadequate, and will waste time that cannot be corrected until the flaws are spotted too late during peer-review.
+The paper \cite{mrb} raises many of the same trade-offs that SEBs also face; indeed one would hope that Methods Review Boards would include Software Engineers or have SEBs as sub-boards --- as this paper has argued, Software Engineering is now a key methodology of science.
+As with SEBs, the goal is not to gatekeep, but to improve.
 
-Although this study of software engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
+\subsection{Action must be interdisciplinary}
+\setcounter{footnote}{1}
+Code is only part of science, and only one critical factor in the wider reproducibility crisis: SEBs must work with --- and be engaged by --- other reproducibility initiatives, such as TOP and Methods Review Boards.
+
+Relying on SEBs \emph{alone\/} would continue one of the current besetting problems about the role of code.
+The conventional view is that scientists do the hard work compared to the ``easy'' coding work (sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}) so they just need to tell programmers what to do.
+This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, which he explores at some length, is that we need to spend more effort in working out what computers should do (i.e., do the science better) and then \emph{just\/} tell programmers to do \emph{that}.
+
+On the contrary, competent Software Engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and how that needs refining or optimizing.
+In other words, the Software Engineers can bring important insights into the science, hence improving or changing the questions and assumptions the science relies on.
+This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but it's ill-conditioned, there is no good answer to that question.'' Ideally, then, it is not  a simple sequential process
+
+\begin{center}\sf\fbox{science specifies} $\rightarrow$ \fbox{code up} $\rightarrow$ \fbox{get results}, 
+\end{center}
+
+\noindent but an iterative cycle of collaboration and growing understanding, informed by Software Engineering best practice (via SEBs), and implemented using papers as laboratories.
+
+In short, the way SEBs work and are used will be crucial to the success of the science they support.
+Software engineers can help improve the science, so it is not just a matter of asking a SEB whether some coding practices (like documentation) are satisfactory, but whether the SEB has insights into the science itself too.
+Most effectively, this requires interdisciplinary working practices (science plus Software Engineering) with mutual respect for their contributing expertise.
+
+\subsection{Methodological statements}
+Many science journals, conferences, workshops, videos, books, etc, and funders require explicit statements how the authors have conformed to appropriate methodological standards covering issues such as conflicts of interest, ethics, data access, consent, authoring, funders and other acknowledgements of support, and so forth.
+Conformance to PRISMA, discussed above in section \ref{PRISMA-statement}, is one such methodological statement; like many journals, the high-profile journal \emph{PLOS ONE\/} has a list of similar methodologies it expects authors to comply with as appropriate.
+
+It would be easy for journals and funders to require equivalent types of statements on the quality of code, that is on the quality of its Software Engineering.
+
+Studies of data access statements show that they are unreliable: some authors withdraw papers when journals request access statements \cite{no-raw-data} (indicating that journals that do not make an explicit request are likely publishing papers that will not provide access), and some authors do not respond to access requests \cite{data-access}.
+How we help scientists who do not want to provide data or code access is one problem, but the serious issue for science is how to ensure any statements made are accurate, and any access provided is actually helpful (e.g., well-defined, versioned, etc) to support useful reproduction.
+
+Journals and funders should provide support for hosting data and code (and any other relevant data, such as qualitative data, video, etc), and the review process must check that authors actually provide the material as they claim in the methodological and access statements.
+Conversely, scientists should be able to access funding to ensure data and code access, and, as appropriate, funding for on-going maintenance of the databases, which will typically require funding beyond the end of the normal funding period.
+
+Intellectual Property (IP) is an increasing concern, both for author scientists and their sponsors who want royalties, and for other scientists wishing to freely build on the published science.
+Particularly concerning access to code, IP is potentially and often is in conflict with scientific openness.
+Methodological statements should be made concerning any IP associated with code, and to what extent this interferes with open access to code.
+More routine discussion should include raising known system dependencies, such as operating system, compiler, or special hardware dependencies; it is also appropriate to mention standards conformance, such as to IEEE Floating-Point Arithmetic (IEEE~754).
+
+Journal policies could start to explicitly encourage computationally reproducible science using RAP and \RAPstar\ techniques.
+That is, the research's methodology itself may be a mixture of data and code.
+As this paper's \supplement\ shows in section \ref{restrictive-policies}, many journals (e.g., \emph{PLOS ONE\/} and, ironically, \emph{IEEE Transactions on Software Engineering\/}) and repositories have policies that make RAP much harder or just counter-productive at the last step.
+
+Methodological statements should be required that make clear what access rights are available for RAP or \RAPstar\ material, as it is much more likely to raise IP issues that normal disclosures.
+In particular, if the authors plan on publishing a series of papers based on the same methodologies, the RAP/\RAPstar\ access might be provided in a later paper or held under escrow by the journal or funding body.
+
+Journals and funders often require data and code access statements, but as this paper has made clear, code is complex and it is rarely easy to understand and scrutinize even with access to substantial documentation (which is unusual).
+It is therefore recommended that journals and funders require \emph{assurance arguments} \cite{assurance-case}, a familiar technique from the safety assurance domain.
+Assurance arguments provide a concise, high-level argument that the system does what is claimed.
+Assurance arguments can be more or less detailed, and more or less formal in their approach; we envisage referees would have views on the level of detail and formality required for any specific contribution.
+
+Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access statements apply to all information, regardless of any arbitrary classification of it as code or data (or documentation, assurance case, etc).
+
+\subsection{Benefits beyond science}\label{benefits-beyond-science}
+Science increasingly recognizes the key supporting roles of code and computation, but many fields do not recognize computation as such, as a skilled discipline, and therefore they are missing out on the leverage that comes with recognizing computation as a first class player in their  activities.
+
+Healthcare is supported by computers, yet medical research papers remain traditional and do not refer to code.
+Yet clinical practice relies on computer analysis, so inevitably must use code unrelated to the code developed in research, which is unlikely to be developed to Software Engineering standards.
+Conversely, the critical issues (including patient safety) that assume code is more reliable than that needed for scientific research do not get evaluated by researchers.
+The gap is wide.
+The problems and missed opportunities of under-valued and poorly-managed code are ubiquitous in healthcare \cite{fixit}.
+Lasting solutions might well be initiated through SEBs or equivalent.
+
+Numerous problems in finance have been facilitated if not precipitated by computational na\" \i vety.
+JP Morgan Chase (JPM) lost over \$6 billion in a credit derivatives trade \cite{jpm}, in a costly parody of bad science --- compare the discussion here with figure \ref{excelFigure}, in a series of figures in the \supplement, which illustrates the same problem as it presents in many scientific papers.
+
+As reported in \cite{jpm}, the traders did not understand the trades, did not monitor them, doubled down when results were poor, and did not communicate the extent of their losses.
+They were using manual coding methods; as the report says:
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0
+[\ldots] the model operated through a series of Excel spreadsheets, which had to be completed manually, \textbf{by a process of copying and pasting data from one spreadsheet to another}.
+[Our emphasis.]
+
+\hskip -\wd0\box0
+[\ldots] this individual immediately made certain adjustments to formulas in the spreadsheets he used.
+These changes, which were not subject to an appropriate vetting process, inadvertently introduced two calculation errors
+
+\hskip -\wd0\box0
+[\ldots] after subtracting the old rate from the new rate, the spreadsheet divided by their sum instead of their average, as the modeler had intended.''
+
+etc}
+\cite{jpm-report}
+
+\end{quote}
+
+The report does not detail how the Excel spreadsheets were specified or coded, seemingly as unawqre of Software Engineering as the traders.
+However, one infers from the brief discussion of data handling (which is easier to automate) that it was all an unconsciously incompetent process.
+
+Reviewers in JPM failed to scrutinize not just the coding, but the trades informed by the code.
+They passed on optimistic reports.
+Then there was a merry-go-round of blame: ``the information communicated to the Risk Policy Committee \ldots\ did not suggest any significant problems \ldots\  there was no robust debate with the right facts at the right level about the portfolio risk.'' UK and US governments are now investigating fraud.
+Again, lasting solutions might well be initiated through SEBs or equivalent.
+
+Without taking the lessons of improving Software Engineering to other fields, including improving and broadening the recognition and career paths for developers, there will continue to be an unfortunate and unnecessary disconnect between competent coding and actual practice.
+Everything, from healthcare to finance --- not just science --- will suffer because the critical contributions of dependable code, quality Software Engineering, and competent Computational Thinking are not recognized, understood, valued, or required.
+
+\ignore{\subsection{Suggestions for further work}
+Although this study of Software Engineering in scientific research (specifically, in peer-reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
 However, the insights appear to be general.
@@
 We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor.
+}
 
-\label{jvs-policy}
+\section{Conclusions}
+In a pandemic, epidemiological modeling (discussed in this paper), track and trace \cite{excel-fiasco}, modeling mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, and so on, drive public policy and have a direct impact, whether positively or negatively, on the quality of life.
+
+While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID modeling, the wider evidence reviewed here suggests that scientific coding practice is inadequate in every field.
+As science becomes more and more reliant on computers, we need to correspondingly improve the quality of code, the quality of code policies, the quality of Software Engineering, and the quality of all scientists' understanding of computation and how to manage its unlimited complexity.
+
+The main challenges to mature computationally-realistic science are:
+
+\enum=1
 \begin{enumerate}\raggedright
 \item 
-The journal \emph{The Lancet\/} published and then subsequently retracted a paper on using hydroxychloroquine as a treatment for COVID \cite{lancet-retracted}.
-The paper was found to rely on fraudulent data \cite{science-lancet1,science-lancet2}.
-\emph{The Lancet\/} subsequently tightened its data policies \cite{lancet-learning}, for instance to require that more than one author must have directly accessed and verified the data reported in the manuscript.
-Curiously, the original (now retracted) paper declares
-\begin{quote}\sf
-\setbox0=\hbox{``}\hskip -\wd0\copy0 \ldots\ all authors participated in critical revision of the manuscript for important intellectual content.
-MRM and ANP supervised the study.
-All authors approved the final manuscript and were responsible for the decision to submit for publication.'' \end{quote}
-which seems to suggest that several original authors of the paper would have been happy to make the new declarations --- and, of course, if there is fraud (as was established in this case) it seems likely that authors who make the new declarations of accessing and verifying data are unlikely to make reliable declarations.
-
-\hskip 1em\emph{The Lancet\/} still has no code publication policy (see the \supplement), and for more than one author to have ``direct access'' to the data they are very likely to access the data through the same code.
-If the code is faulty or fraudulent, an additional author's confirmation of the data is insufficient, and there is at least as much reason for code to be fraudulent (not least because code is much harder to scrutinize than data).
-Code needs more than one author to check it, and ideally reviewers independent of the authors so they do not share the same assumptions and systems (for instance shared libraries, let alone potential collusion in fraud).
-
-\item 
-In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms.
-Notably, the paper fails to provide access to its anonymized data (with or without qualification), and fails to define the data anonymization algorithm, and also fails to even mention the code that it developed and used to perform its study.
-The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see \supplement\ section \ref{supplementary-journal-policies-section}).
-Ironically, the retracted article \cite{jvs1} is still online (as of August 2020) with no reference to any editorial statement to the effect that it has been retracted, despite this being trivial --- and necessary --- to achieve in the widely-accessed online medium.
-
-\item
-Medical research often aims to establish a formula to define a clinical parameter (such as body mass index, BMI) or to specify an optimal drug dose or other intervention for treatment.
-These formulas, for which there is conventional clinical evidence, are often used as the basis for computer code that provides advice or even directly controls interventions.
-Unfortunately a simple formula as may be published in a medical paper is \emph{never\/} sufficient to specify code to implement it safely.
-For example, clinical papers do not need to evaluate or manage user error when operating apps, and therefore the statistical results of the research will be idealistic compared to the outcomes using an app under real conditions --- which is what the clinical research is supposedly for.
-A widespread bug (and its fix) that is often overlooked is discussed in \cite{numerals}; the paper includes an example of a popular clinical calculator (based on published clinical research) that calculated nonsense, and potentially dangerous, results.
-The paper \cite{fda} summarizes evidence that such bugs, ignored by the clinical research literature, are commonplace in medical systems and devices.
-\end{enumerate}
-
-\section{Conclusions}
-We need to improve the quality of software engineering that supports science.
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed in shows that current coding practice makes for poor science.
-In the current pandemic, scientific modeling, such as track and trace \cite{excel-fiasco}, balancing COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.
-
-The challenge to scientific research is to manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices.
-Computer code should be explicit, accessible (well-structured, etc), and properly documented.
+To manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices that research and publications rely on.
+Computer code should be explicit, accessible (well-structured, etc), and adequately documented.
 Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}.
@@
 
-Unfortunately, while programming is easy, and is often taken for granted, programming \emph{well\/} is very difficult \cite{fixit}.
-We know from software research than ordinary programming is very buggy and unreliable.
-Simply put, without quality code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
-Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion:  computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper}.
+\item 
+To use computation to help make scientific processes explicit, so that they can be reproduced, scrutinized and improved.
+RAP is an increasingly popular way to help do some of this, but as this paper points out, RAP should be generalized to \RAPstar\ to help the computational parts of science as well, leading to a virtuous circle.
 
-Software Engineering Boards (as introduced in this paper) are a straight forward, constructive and practical way to support and improve computer-based science.
-This paper's \supplement\ summarizes the relevant professional software engineering practice that Software Engineering Boards would use, including discussing how and why software engineering helps improve code reliability, dependability, and quality.
+\item
+To support and develop the scientific community in the professional use of computation.
+
+\item
+To find effective ways to promote professional software engineers being recognized and participating  fully in scientific research, like professional statisticians routinely support quality research.
+\end{enumerate}
+
+While programming seems easy, and is often taken for granted and done casually, programming \emph{well\/} is very difficult \cite{fixit}.
+We know from software research that ordinary programming is very buggy and unreliable.
+Without adequately specified and documented code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
+Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion: computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not magically make it more reliable \cite{reproducibility,relit,popper-conjectures-refutations}.
+
+Software Engineering Boards, as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve computer-based science.
+
+This paper's \supplement\ summarizes relevant Software Engineering good practice that Software Engineering Boards would draw on, including discussing how and why Software Engineering helps improve code reliability, dependability, and quality.
+
+\makeatletter
+\immediate\write\@auxout{\string\def\string\continueSectionNumbers{\arabic{section}}}
+\makeatother
 
@@
 \def\csref{\global\advance\csrefcount by 1}
-\def\ethics#1{\paragraph*{Ethics}#1}
-\def\ack#1{\paragraph*{Acknowledgments}#1}
-\def\dataaccess#1{\paragraph*{Data and code access}#1}
-\def\aucontribute#1{\paragraph*{Author contribution}#1}
-\def\competing#1{\paragraph*{Competing interests}#1}
-\def\funding#1{\paragraph*{Funding}#1}
+\long\def\ethics#1{\paragraph*{Ethics}#1}
+\long\def\ack#1{\paragraph*{Acknowledgments}#1}
+\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
+\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
+\long\def\competing#1{\paragraph*{Competing interests}#1}
+\long\def\funding#1{\paragraph*{Funding}#1}
 
-\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson provided very helpful comments.}
+\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Konrad Hinsen, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, \csref Isaac Thimbleby, \csref Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson.
+The author also thanks the anonymous \emph{Computer Journal\/} referees who also contributed to the quality of this paper.}
 %\the\csrefcount\ computer science reviewers.
 
-\ethics{This article presents research with ethical considerations but does not fall not within the usual scope of ethics policies.}
+%\ethics{This article presents research with ethical considerations but does not fall within the usual scope of ethics policies.}
 
-\dataaccess{There is a full discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}.
-All material is available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards} (which has been tested in a clean build, etc).
+\dataaccess{There is an extended discussion of the methodology of this paper in the \supplement, section \ref{on-code-data-publication}.
+The \supplement\ also presents all raw data in tabular form.
+All material is also available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards},\footnote{\color{red}This is a temporary URL before meeting publication repository requirements for accepted papers.\color{black}} which has been tested in a clean build, etc.
+
 The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \input generated-data.js-errors.tex
-possible types of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset results reliably in this paper and its \supplement\ (e.g., table \ref{table-summary} in this paper) \emph{directly\/} from the data.
-In addition, a standard CSV file is generated in case this is more convenient to use, for instance to browse in Excel or other spreadsheet application.
-All data, analysis, and details of the methods used can be found in the \supplement.}
+JavaScript code, conveniently in the same file as the JSON data, checks (with \the\JSONerrorCount\ possible classes of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset all results reliably in this paper and in its \supplement\ \emph{directly\/} from the automatic data analysis.
+
+In addition, a standard CSV file is generated from the JSON in case this is more convenient, for instance to browse directly as a spreadsheet or to import easily into other programs.}
 
@@
 An preliminary outline of this paper, containing no supplementary material or data, was submitted to the UK Parliamentary Science and Technology Select Committee's inquiry into UK Science, Research and Technology Capability and Influence in Global Disease Outbreaks, under reference LAS905222, 7 April, 2020.
-The evidence, which was not peer reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper.
+The evidence, which was not peer-reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper.
 It is available to the public \cite{parliamentary-evidence,my-parliamentary-evidence}.}
@@
 
-\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], and by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186].
+\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186], and by Assuring Autonomy International Programme, Assuring Safe AI in Ambulance Service Triage.
 The funders had no involvement in the research or in this paper.}
 
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
+%\nolinenumbers
 
 {\raggedright
+
+\bibliographystyle{revisedCompJ}
+\def\urlprefix{}
 \initialiseBibliography{References}{1}{}
-% \bibliographystyle{plos2015} - already provided in plos-header.tex
+
+\newcount \bibitemn \bibitemn = 0
+\newcount \bibnewcolumn
+
+\bibnewcolumn = 112 % bibentry number of last entry on (final) left hand column
+% Hint: set it to a very large number, so the final column is as unbalanced as it gets
+% then, if it's worth fixing the final page, lower the number to some citation on the 
+% lower part of the left hand column; this citation will then become the last in that
+% column. Hopefully the righthand column will be of equal length!
+% PS Don't forget to check there is nothing on the following page!
+
+% if you can't get a decent column balance, tinker with the inter-citation spacing, below
+% it currently looks like "\vskip .2ex plus 1ex minus 1ex"
+% you'll be surprised what adjusting it by only 0.1ex or less achieves after 100 citations!
+
+\let\oldbibitem=\bibitem
+
+% calculate bib entry width
+\newdimen \bibEntryWidth
+\bibEntryWidth = \columnwidth
+\advance \bibEntryWidth by -\columnsep
+
+% I haven't figured out how to make the \labelwidth etc variable - it'd be nice
+% to have it suitable for numbers up to 9, then for numbers up to 99, then up to 999
+% depending on the largest bibitem number width in the column. As it is, it's set
+% to the width of 999 for all bibitems even though it takes a while for bibitems to get 
+% up to 3 figures. Sigh
+
+\def\bibitem#1{%
+\ifnum \bibitemn = \bibnewcolumn \egroup\newpage\bgroup \fi
+\ifnum \bibitemn = 0 \else \egroup \fi
+\advance \bibitemn by 1
+\vskip .3ex plus 1ex minus 1ex 
+\leavevmode \vbox\bgroup \parshape 1 0in \bibEntryWidth 
+\oldbibitem{#1}}
+
+\def\endofbibliography{ \egroup \vfill}
+
+% the class comjnl has a minor formatting bug in the setting of labelwidth; fixed below:
+\makeatletter
+\def\thebibliography#1{\section*{References}\small\list
+ {[\arabic{enumi}]}{\labelsep 0.5em%
+\hyphenpenalty=50\itemsep=-1pt
+ \settowidth\labelwidth{[0#1]}% previously too small, so added an extra digit (0)
+ \leftmargin\labelwidth
+\advance\leftmargin\labelsep
+ \usecounter{enumi}}
+ \def\newblock{\hskip .14em plus .38em minus -.1em}
+ \sloppy
+ \sfcode`\.=1000\relax}
+\let\endthebibliography=\endlist
+
+\def\@cite#1#2{[{#1\if@tempswa , #2\fi}]}
+\def\@biblabel#1{[#1]}
+\makeatother
+
 \bibliography{paper-seb-main.bib}
@@
 % because it reads in this .tex file's .aux file so that it can cite the references 
-% above with the same number sequence. 
+% using the same number sequence. 
 %
-% It simple redefines \bibitem to work it out while it
+% It simply redefines \bibitem to work it out while it
 % defines all the citations from the bibliography above.
@@
 
+%\label{LastPage}
+%save Latex register    {linenumber}
+%\makeatletter
+%\immediate\write\@auxout{\string\def\string\MainPaperLastLine{\arabic{linenumber}}}
+%\makeatother
+\end{document}
+
+\def \allInOne {define this so we don't include header in supplemental material}
+\input paper-seb-supplementary-material.tex
 \end{document}

### 2023-01-31 paper-seb-supplementary-material.tex
Fixing git messages

@@
     \PassOptionsToPackage{hyphens}{url}
-    \documentclass[10pt,letterpaper]{article}
-
-    \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-    % Use adjustwidth environment to exceed column width (see example table in text)
-    \usepackage{changepage}
-
-    % line numbers
-    \usepackage[right]{lineno}
-    \linenumbers
-
-    \renewcommand\thesubsection{\thesection.\alph{subsection}}
-    %\usepackage{hyperref}
-    %\usepackage{refcount}
+    \documentclass[10pt,a4paper]{article}
+    \usepackage[text={6.5in,10in},centering]{geometry}    
     \usepackage{url}
@@
     \usepackage{fancyhdr}
+    \usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{epstopdf}
     \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
     \def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}    
-
     \input paper-seb-macros.tex
@@
       \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
+      \let\oldbibcite=\bibcite
       % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
@@
       \input paper-seb-main.aux 
-      \resetlinenumber[\MainPaperLastLine]
-      }
-
+      %\resetlinenumber[\MainPaperLastLine]
+	  \let\bibcite=\oldbibcite
 \makeatother
+
+\input generated/page-lengths.tex
 
@@
 
-% pageref gives us a hyperlink which looks unsightly
+% pageref gives us a hyperlink which looks unsightly, so we do it this way...
 \makeatletter
@@
 % this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
+\newwrite\infofileForMain
+\immediate\openout\infofileForMain=generated/info-for-main.tex
+
+%\makeatletter
+%% saves name-value pairs in both the supplementary material .aux file and the info for the main file
+%\def\remembernamevalue#1#2{%
+%	\immediate\write\@auxout{\@backslashchar gdef\@backslashchar #1{#2}}%
+%	\immediate\write\infofileForMain{\@backslashchar gdef\@backslashchar #1{#2}}%
+%}
+%\makeatother
+
+\input generated/info-for-supplement.tex
+\newwrite\infofileForSupplement
+\immediate\openout\infofileForSupplement=generated/info-for-supplement.tex
 
@@
 \def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
+\typeout{Global globalLabel: #1}%
+\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
 % as well as in the current document....
 \immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
+}
+\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
+\typeout{Global globalTableLabel: #1}%
+\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
+% as well as in the current document....
+\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
+}
+\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
+\typeout{Global globalLabelValue: #1 := #2}%
+\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
 }
@@
 
-\setcounter{section}{0}
+%\setcounter{section}{0}
+\setcounter{section}{\continueSectionNumbers}
 \def\n#1{{\sf\bfseries ({#1})}}
@@
 \def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
+\def\changemysection#1{\vskip 2ex\section{{#1}}\sectioncontents{{#1}}}
 \def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
@@
 \def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent
-\input generated-supplementary.toc
+\noindent\small
+\input generated/supplementary.toc
 }
@@
 \newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
+\immediate\openout\contentsfile=generated/supplementary.toc
 
 \newpage
+\begin{center}
+\emph{Pagination, section, and citation numbering continues from the main paper}
+\end{center}
 \mysection{Further issues for Software Engineering Boards (SEBs)}
@@
 \item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
+Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at Software Engineering.
 
@@
 Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
+Often Software Engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
+Often the Software Engineers will be solving entirely new problems and contributing to the research.
 How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
+How Software Engineers are appropriately credited and cited for their contributions also needs exploring.
 
@@
 SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
+For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of Software Engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
 
@@
 \item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
+Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about Software Engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
 SEBs may also provide mentoring and training.
@@
 Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
+During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional Software Engineering.
 
@@
 Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
+It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional Software Engineers to avoid this problem.
 \end{enumerate}
@@
 As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
+In other words, improving the professionalization of Software Engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
+SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional Software Engineering, but this is a topic the present paper has not addressed.
 It needs doing.
@@
 
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
+This \supplement\ provides more explanations and justification for following standard Software Engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
+
+The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to Software Engineering in science \citeeg{cosmos}.
+Written and maintained by a team of experts, a substantial and wide-ranging reference is the Software Engineering Body of Knowledge (SWEBOK) \cite{swebok}, recognized as International Standards Organization Technical Report 19759.
+
+The Turing Institute has an excellent open resource \cite{turing-way}, though it emphasizes RAP for handling data and authoring papers rather than for programming reliably.
 
@@
 \vbox{\noindent
-      \input generated-minicontents.tex
+      \input generated/minicontents.tex
       }
@@
 \newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
+\immediate\openout\minicontentsfile=generated/minicontents.tex
 
@@
 On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
+The first step of Software Engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
 
@@
 Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
+Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice Software Engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
 In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
+This specification, carefully reviewed, is then used as the basis for a more rigorous Software Engineering process that implements a more reliable version of the system.
 
@@
 
-{The type and use of mathematics used in software engineering is formal methods.
+{The type and use of mathematics used in Software Engineering is called \textbf{formal methods}.
 Not using formal methods ensures the resulting code is unsafe and unreliable.
@@
 
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
+It is important to be clear that formal methods is a spectrum, from doing it as rigorously and comprehensively as possible using state of the art methods, to applying a basic ``formal methods mindset'' that any competent programmer could do.
+
+\begin{itemize}\raggedright\item An interesting paper explores a formal approach to coding differential equations \cite{differential-equations}, but it makes it clear that their approach is beyond most programmers.
+Such formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK \cite{spark}, and \emph{many\/} others).
 Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
@@
 
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
+\item At the other end of the spectrum, a formal methods mindset means being clear about basic mathematical properties of code, such as by using assertions, proving invariants in loops, and so on.
+This approach should be feasible --- and perhaps required --- for all scientific programmers.
+\end{itemize}
+
+Crucially, computer tools are available to catch common human errors that we are all prone to.
+Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort \cite{misra}.
 
@@
 In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
+
+Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of Software Engineering has advanced and become more specialized and sophisticated over the decades since.
 Hoare is also a very good writer.
@@
 
-\begin{change}Interestingly, the only paper reviewed that claimed to do any independent testing \cite{onlyPaperWithChecks} failed to include any testing in its data or code repository, so the testing itself --- the essential quality assurance of the code --- is not open to scrutiny (e.g., the code and the ``independent'' code are likely to contain common code, data, and common bugs).\end{change}
+Interestingly, the only paper reviewed that claimed to do any independent testing \cite{onlyPaperWithChecks} failed to include any testing in its data or code repository, so the testing itself --- the essential quality assurance of the code --- is not open to scrutiny (e.g., the code and the ``independent'' code are likely to contain common code, data, and common bugs).
 
@@
 C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
+(This is not the place for a review of bad programming languages, for which see \cite{heedless}, but Excel is even worse than C\@.)
+
+In particular, C is not a portable language, which means C code will work differently on different types of computer and operating system.
+
+SPARK Ada is a popular example of a much more appropriate high integrity programming language to use \cite{spark}.
 SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
+
+Other high integrity languages include  OCaml, F*, and Haskell; reference \cite{haskell} is an excellent introduction to Haskell, and introduces the wider issues of reliable programming in such languages.
 
@@
 When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
+Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using \emph{Mathematica\/} --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
 Reproducing the same results relying on the same codebase tells you little.
@@
 Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
+Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and as there are.are serious dependability problems with Excel (illustrated particularly in section \ref{benefits-beyond-science} of the main paper), open source projects competently implemented (e.g., avoiding or carefully managing the use of Excel) would help these people enormously.
 
@@
 
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
+For purely scientific purposes, perhaps the most important form of documentation is internal documentation: how to understand how and why the code works.
 This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
@@
 Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
+The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understand what useful guarantees these basic properties ensure).
+
+Note that reusing code assumes the originators of the code followed good Software Engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good Software Engineering practice, it too can be shared and further improved as it gets more exposure.
 Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
@@
 
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
+Note that medical device regulation, which has its own standards, is lagging behind professional Software Engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
 
@@
 
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
+\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in Software Engineering is continually evolving}{Continuous Professional Development (CPD)}
+
+As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in Software Engineering is also rapidly evolving.
 Continuous Professional Development (CPD) is essential.
@@
 Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
+{In Software Engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on Software Engineering standards in other fields.}
 
@@
 
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-\begin{change}
+For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example, one of the surveyed paper's code \cite{example-numerical-error} uses literal numbers at far too high a precision for the chosen language to be able to represent correctly (conformant implementations use IEEE 754 double precision 64-bit floating point).
+Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.
+The error belies misunderstandings in programming which may have wider effects, such as consequences of relying on the precision being higher than it is.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
+
+{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for Software Engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
+
+\highlightsec{14}{Human Factors are the foundation of everything}{Software is a human activity}
+
+Software is a human activity, and humans are fallible.
+Even the Software Engineering methodologies to developer better software are themselves human constructs, and are therefore subject to the same fallibilities.
+
+People would generally not make software errors if there were aware they were making errors.
+Unfortunately programming is a very demanding activity, which causes tunnel vision (also known as loss of situational awareness).
+Humans have limited cognitive capacity, and programming (especially programming in a competitive environment, like science) drives programmers to use as much of their cognitive skills for the task in hand.
+The consequence is programmers focus on ``the'' problem as it appears in the code, and inevitably become unaware they are not considering wider issues.
+The correctness, generality, ethics, and usability of a program are therefore often unintentionally sacrificed to making code work at all.
+
+Confirmation bias is a standard Human Factors problem \cite{fixit}, which encourages us to perform tests that show our programs work.
+Instead, we should be rigorously testing ways in which programs can fail as well.
+This is exactly the same issue pointed out by Popper \cite{popper-conjectures-refutations}: scientists should experiment to find reasons why hypotheses are false, and indeed use simple hypotheses that are testable.
+Software is really no more than a collection of sophisticated hypotheses, and Computer Science is a science of the artificial \cite{simon-artificial}.
+
+Standard Human Factors mitigations for such problems include team working, with appropriate precautions to manage authority gradients (where the Human Factors oversights of the leader influence the team).
+Many computerized mitigations are also available --- strong typing, code analyzers, formal methods, and so on, as described in this section of the \supplement.
+
+Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic Software Engineering.
+Dunning and Kruger go on to say,
+
+\begin{quote}\sf\raggedright\setbox0=\hbox{``}\hskip -\wd0\copy0 People usually choose what they think is the most reasonable and optimal option \hbox{[\hskip 2pt\ldots]} The failure to recognize that one has performed poorly will instead leave one to assume that one has performed well; as a result, the incompetent will tend to grossly overestimate their skills and abilities.
+\hbox{[\hskip 2pt\ldots]} Not only do these people reach erroneous conclusions and make unfortunate choices, but their incompetence robs them of the metacognitive ability to realize it.''\end{quote}
+
+Unlike many skills (skating, brain surgery, \ldots) programming, typical of much engineering, is one where errors can go unnoticed for long periods of time --- things seem to work nicely right up to the moment they fail.
+The worse programmers are, the more trivial bugs they tend to make, but trivial bugs are easy to find so, ironically, being a poor programmer \emph{increases\/} one's self-assessment because debugging seems very productive.
+It is easy for poor programmers and their associates to believe they are better than they actually are, fertile ground for the better-than-average bias \cite{dunning-kruger}.
+
+It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}.
+We \emph{all\/} make mistakes we are unaware of.
+If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.
+Moreover, if we do not work in a diverse team, nobody will ever point this out, so the potential problems it causes will never be addressed.
+
+{Everyone is subject to Human Factors (including the author of the present paper, e.g., as discussed in \cite{enigma}): for instance, the standard cognitive bias of confirmation bias encourages us to look for bugs when code fails to do what is expected and then debug it to produce better results, but if code generates expected results not to bother to debug it further.
+This of course tends to make code increasingly conform to prior expectations, whether or not those expectations are scientifically justified.
+Typically, there was no prior specification of the code, so the code should be right, especially after all the debugging to make it ``correct''!
+Thus coding routinely suffers from HARKing (Hypothesizing After the Results are Known \cite{harking}), a methodological trap widely recognized in statistics.}
+
+Computers themselves are also a part of the problem.
+Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable.
+Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
+The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
+
+\mysection{Code, data, and publication}
+\globalLabel{on-code-data-publication}
+
 All computer systems are in principle equivalent to Turing Machines, and Turing Machines make no distinction between program and data.
@@
 Therefore, even the ``pure'' programs of $\lambda$-calculus also represent data.
-\end{change}
 
@@
 The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Some programming languages like Mathematica treat $\pi$ as an exact mathematical constant (e.g., Mathematica calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
+Some programming languages like \emph{Mathematica\/} treat $\pi$ as an exact mathematical constant (e.g., \emph{Mathematica\/} calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
 \url{www.fhgr.ch/en/specialist-areas/applied-future-technologies/davis-centre/pi-challenge}}
@@
 
-\begin{change}
-\subsection{When magic numbers become magic code}
+\mysubsection{When magic numbers become magic code}
 Data often controls the flow of code. For example, data summarizing patients may include their gender, but the program processes males and females differently. Then data becomes code.
@@
 
-\subsection{When data is code}
+\mysubsection{When data is code}
 {Many computer programs blur the simplistic code/data distinctions deliberately, to create virtual machines.
@@
 
-\subsection{Exploiting code as data for more reliable science}
-\end{change}\noindent
-In the present paper, we knowingly built on this blur between data and code.
+\mysubsection{Exploiting code as data for more reliable science}
+In the present paper, we knowingly built on this blur between data and code, a special case of \RAPstarp.
 However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
@@
 
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
+\mysubsection{When data is text: Exploiting code for reliable publication}
+
+Section \ref{over-fit} of the main paper looks like part of an ordinary paper, but it (including the figure and calculations) was data generated by a \emph{Mathematica\/} program.
+
+Most programs are code plus comment, and their data comes from some external source or sources.
+In \emph{Mathematica}, programs are represented as ``notebooks,'' which can be structured like reports or papers.
+They have sections, which allow program code, data and program output to be arbitrarily mixed in a single file.
+
+For the purposes of the present paper, a notebook was created with a new type of data, \LaTeX\ text.
+The \LaTeX\ text can be any mix of text written by the author or material generated by running \emph{Mathematica}.
+A final step of running the \emph{Mathematica\/} notebook is to collect all of the \LaTeX\ material, whether written by hand or generated, and save it to a normal text file for \LaTeX\ to typeset.
+
+The idea is very simple, but very effective.
+The ``code'' for the paper includes the \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
+In fact, the notebook is written as a self-contained report or paper, as \emph{Mathematica\/} notebooks generally are, and it thoroughly documents how it works.
+
+For readers of this paper who do not have access to \emph{Mathematica\/} to run the notebook, a PDF of the notebook is included to show how it works.
+Note that the method can be applied in any programming language, but \emph{Mathematica\/} makes the interleaving of paper text and calculations (of arbitrary complexity) very easy --- in conventional programming environments the paper text would be separate (e.g., as data) and it would be much harder to keep the text and code in synchronization.
+
+More of such techniques for improving reliability are discussed in section \ref{data-methods-subsection}, where they are applied to accurately and reliably reporting the survey reported in the main paper.
+
+\mysubsection{Data and polynomials used in the paper}
+The data and polynomials used in the main paper's section \ref{over-fit}, and illustrated in the paper's figure \ref{fig-overfit}, is presented below, as generated in \LaTeX\ by the same \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
+Of course, in the paper itself, the data was not necessary to make the point, but if there is any need to replicate it or otherwise scrutinize the argument in the paper --- as there would be for more complex arguments in typical papers --- the data used and results are shown here.
+In a similar way, data and computed results could be presented and made available in other scientific papers.
+
+\input generated/over-fitting-data.tex
+
+\highlightmysubsection{Comparing conventional and RAP approaches}
+\globalLabel{RAP-diagrams}
+The similarities and differences between the conventional copy-and-paste approach to filling in data and diagrams in publications, the improved systematic RAP process, and using notebooks (such as \emph{Mathematica\/} or Jupyter), are illustrated in the sequence of schematics of figures \ref{fig:basic}, \ref{fig:excel}, \ref{fig:rap}, and \ref{fig:notebook}.
+
+\begin{figure} 
+   \centering
+   \includegraphics[width=5in]{generated/basic.pdf} 
+   \caption{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
+Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code.}
+   \label{fig:basic}
+\end{figure}
+\begin{figure} 
+   \centering
+   \includegraphics[width=5in]{generated/excel.pdf} 
+   \caption{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
+The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
+Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
+In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication.
+}
+   \globalLabelValue{excelFigure}{\thefigure}
+   \label{fig:excel}
+\end{figure}
+\begin{figure} 
+   \centering
+   \includegraphics[width=5in]{generated/import.pdf} 
+   \caption{Improving over the normal approach (figure \ref{fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+In the present paper, this was done by generating text files of \LaTeX\ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX\ names for all of the code-generated values.
+Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
+Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen.}
+   \label{fig:rap}
+\end{figure}
+\begin{figure}[t]
+   \centering
+   \includegraphics[width=5in]{generated/notebook.pdf} 
+   \caption{In a notebook system, such as \emph{Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
+The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
+Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
+In well-engineered code, the notebook will directly contain full documentation.
+Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc).}
+   \label{fig:notebook}
+\end{figure}
+
+\changemysection{The Spiegelhalter trustworthiness questions}
+\globalLabel{supplementary-Spiegelhalter-section}
+
+David Spiegelhalter is concerned how statistics is often misused and misunderstood.
+In his \emph{The Art of Statistics\/} \cite{Spiegelhalter} Spiegelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- and the claims need to be properly accountable.
+
+Spiegelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
+Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or publications relying on code --- analogues are suggested in \textbf{bold} below.
+
+What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask these questions is as risky as dismissing statistical scrutiny.
 
@@
 
-\sequestion{How rigorously has the software engineering been done?
+\sequestion{How rigorously has the Software Engineering been done?
 Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
@@
 
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
+\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate Software Engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
 \setcounter{resumeCounter}{\value{enumi}}
@@
 
-\begin{change}\sequestion{What evidence is there of rigorous (e.g., code review and tool-based) independent methods being used to manage coding bias?}\end{change}
+\sequestion{What evidence is there of rigorous (e.g., code review and tool-based) independent methods being used to manage coding bias?}
 
@@
 
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
+\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code, and just assume it is running correctly.
 So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
@@
 \sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
+The Software Engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and Software Engineering that justify the appropriateness of the chosen software implementation?
 The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
@@
 \advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
+\def\MaxMainPaperCitationNumber{\the\temporary}
+
+\highlightmysection{A pilot survey of computational science}
+\globalLabel{survey-section}
+
+The main paper was motivated by selected epidemiological papers and their problematic computational models (which are discussed in the paper).
+Are the issues illustrated by these case studies typical of science more broadly?
+We undertook, then, some selected studies of papers in different fields, then undertook a randomized, stratified pilot survey covering several leading peer reviewed journals.
+The point was not to establish the frequency of problems, so much as to sign whether the problematic case study was exceptional or typical.
+
+%There are many ways to do good science.
+Normal science routinely involves either performing new research and reporting it in the literature, or performing reviews --- such as systematic reviews --- of the existing literature to form a consensus view of an existing body of research.
+Typically, research offers new theories or results, points out previous errors, or uses statistical or qualitative methods to reduce the variability in previously reported results.
+In any case, the intention is to provide reliable knowledge and to bolster consensus \cite{ziman}.
+
+%In the present case, we have identified limitations in the methodology of certain scientific contributions, throwing doubt on their soundness and reliability; furthermore, we identified critical misunderstandings in the literature citing those papers, including misunderstanding the nature of code reproducibility.
+
+%It is possible, if not likely (e.g., given the report in \emph{Nature\/}, cited above \cite{nature-review}), that such limitations are more widespread than the specific examples considered so far, which are anyway limited primarily to epidemiology, and to COVID research performed under extreme pandemic pressure.
+We now therefore undertake a small, manageable stratified sample of peer reviewed literature to explore whether the issues are more widespread.
+
+The pilot study follows the RAP methodology.
+The data and code for this paper's pilot survey (and all other analysis used in the paper and in this \supplement) are available on GitHub.
+All raw data is converted into \LaTeX\ so that the analysis can be typeset directly in the paper; this \supplement\ also contains a complete tabular presentation of the data.
+
+Moreover, this paper itself follows the more general \RAPstar\ methodology.
+For example, as standard practice, a Unix makefile is used to make it easy to analyze and generate all data, typeset the paper, and more.
+Table \ref{table:options} shows the options provided.
+
+\begin{table}[t]
+\begin{center}
+\input generated/make-help.tex
+\end{center}
+\caption{Conforming to the \RAPstar\ methodology, the abbreviated summary above was generated automatically, by using \texttt{make data}.
+%(Saying \texttt{make help} would generate a full summary on the command line.)
+}
+\label{table:options}
+\end{table}
+
+The \RAPstar\ approach cannot enforce the truth of such as summary, but it very significantly increases the chances that the summary is correct and up to date.
+For example, if refactoring leads to an option being deleted, then it will also disappear \emph{with no further work\/} from the table above.
+Also, since the summary was proof-read at the same time as proof-reading this \supplement, following \RAPstar\ also increases the chances that any errors or functionality omissions or issues in the makefile have been detected and corrected.
+
+\highlightmysubsection{Selected journal case studies}
+
+The journal \emph{The Lancet\/} published and then subsequently retracted a paper on using hydroxychloroquine as a treatment for COVID \cite{lancet-retracted}.
+The paper was found to rely on fraudulent data \cite{science-lancet1,science-lancet2}.
+\emph{The Lancet\/} subsequently tightened its data policies \cite{lancet-learning}, for instance to require that more than one author must have directly accessed and verified the data reported in the manuscript.
+Curiously, the original (now retracted) paper declares
+\begin{quote}\sf
+\setbox0=\hbox{``}\hskip -\wd0\copy0 \ldots\ all authors participated in critical revision of the manuscript for important intellectual content.
+MRM and ANP supervised the study.
+All authors approved the final manuscript and were responsible for the decision to submit for publication.'' \end{quote}
+
+which seems to suggest that several original authors of the paper would have been happy to make the new declarations --- and, of course, if there is fraud (as was established in this case) it seems likely that authors who make the new declarations of accessing and verifying data are unlikely to make reliable declarations.
+
+\emph{The Lancet\/} still has no code publication policy, and for more than one author to have ``direct access'' to the data they are very likely to access the data through the same code.
+If the code is faulty or fraudulent, an additional author's confirmation of the data is insufficient, and there is at least as much reason for code to be fraudulent (not least because code is much harder to scrutinize than data).
+Code needs more than one author to check it, and ideally reviewers independent of the authors so they do not share the same assumptions and systems (for instance shared libraries, let alone potential collusion in fraud).
+
+In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms.
+Notably, the paper fails to provide access to its anonymized data (with or without qualification), and fails to define the data anonymization algorithm, and also fails to even mention the code that it developed and used to perform its study.
+The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see  section \ref{supplementary-journal-policies-section}).
+Ironically, the retracted article \cite{jvs1} is still online (as of August 2020) with no reference to any editorial statement to the effect that it has been retracted, despite this being trivial --- and necessary --- to achieve in the widely-accessed online medium.
+
+Medical research often aims to establish a formula to define a clinical parameter (such as body mass index, BMI) or to specify an optimal drug dose or other intervention for treatment.
+These formulas, for which there is conventional clinical evidence, are often used as the basis for computer code that provides advice or even directly controls interventions.
+Unfortunately a simple formula as may be published in a medical paper is \emph{never\/} sufficient to specify code to implement it safely.
+For example, clinical papers do not need to evaluate or manage user error when operating apps, and therefore the statistical results of the research will be idealistic compared to the outcomes using an app under real conditions --- which is what the clinical research is supposedly for.
+A widespread bug (and its fix) that is often overlooked is discussed in \cite{numerals}; the paper includes an example of a popular clinical calculator (based on published clinical research) that calculated nonsense, and potentially dangerous, results.
+The paper \cite{fda} summarizes evidence that such bugs, ignored by the clinical research literature, are commonplace in medical systems and devices.
+
+\highlightmysubsection{Pilot paper sample}
+A sample of \plural{\dataN}{recent paper} covering a broad range of science were sampled from the leading journals \journalBreakdown.
 
 \makeatletter
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
+\long\def\@makecaption#1#2{%
+  \vskip\abovecaptionskip
+  \sbox\@tempboxa{#1: #2}%
+  \ifdim \wd\@tempboxa >\hsize
+    \textbf{#1}: #2\par
+  \else
+    \global \@minipagefalse
+    \hb@xt@\hsize{\hfil\box\@tempboxa\hfil}%
+  \fi
+  \vskip\belowcaptionskip}
 \makeatother
 
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
+\begin{table*}[t]
+\begin{center}
+\begin{tabular}{|rl|} \hline
+\the\numberOfJournals&Journals\\
+\the\dataN&Papers:\\
+\tabularJournalBreakdown
+\the\countAuthors&Published authors\\
+\the\totalPages&Published journal pages\\
+July 2020&Sample month\\ \hline
+\end{tabular}
+\end{center}
+
+%\newcount \total
+%\total=\dataN
+%\multiply \total by 4 % three reviewers, one editor
+%\advance \total by \countAuthors
+%\the\total\
+%leading scientists assuming authors, 1 editor, and 3 reviewers per paper, but not counting acknowledged colleagues et al, covering  (plus appendices and supporting material) in total.
+%
+\caption{Overview of the peer-reviewed paper sample.}
+\label{supplement-table-overview}
+\end{table*}
+
+\begin{table*}[t]
+\begin{center}\normalsize
+\input generated/summary-table.tex
+\end{center}
+
+\caption{Summary of survey results.}
+\label{supplement-table-summary}
+\end{table*}
+
+\begin{table*}[t]
+
+\def\reponame#1#2{{\tt #1} \csname cite-#2\endcsname}
+
+\expandafter\def\csname pagelength-covid-sim\endcsname{20}
+\expandafter\def\csname cite-covid-sim\endcsname{\cite{ICmodel}}
+
+% divide #1 by #2
+\newcount \uu
+\newcount \vv
+\newcount \ww
+\def\makeAverage#1#2{%
+	\uu=#1
+	\vv=#2
+	\multiply \uu by 1000
+	\divide \uu by \vv
+	% uu is now 1000 times the answer we want
+	\ww = \uu
+	\divide \uu by 1000
+	\multiply \uu by 1000
+	\advance \ww by -\uu
+	\divide \uu by 1000
+	%[ #1 / #2 = \the\uu .\the\ww\ ]
+	\ifnum \ww > 499 \advance \uu by 1 \fi
+	\the\uu
+}
+
+\begin{center}
+
+\expandafter \advance \gitPages by \csname pagelength-covid-sim\endcsname
+
+\input generated/repos.tex
+\vskip 1ex
+\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
+\\
+\small Repository clones downloaded and automatically summarized \clonedate
+\end{center}
+
+\caption{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.}
+\label{table-repo-summary}\globalTableLabel{supplementaryTableRepoSummary}
+\end{table*}
+
+The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal.
+All papers sampled are Open Access, although for some papers some or all of the associated data has no or restricted access, in some cases despite the relevant journal policies on code.
+Table \ref{supplement-table-overview} is an overview of the sample.
+
+Papers were selected from the journals' July 2020 then new online listings where the paper's title implied that code had been used in the research.
+Commentary, correspondence, and editorials were excluded.
+The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.
+The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on and assumed the quality of code used in their research.
+
+This convenience sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers,\footnote{Using Google Scholar it is estimated that over 40,000 papers meeting the title criteria were published in the month of July 2020.} but~\ldots
+
+\begin{enumerate}
+\item the selected journals are leading peer-reviewed scientific journals that set the standards for scientific publishing practice generally (although the sample shows that code policies are not always enforced);
+
+\item as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been productively more insightful (this view is consistent with the multi-disciplinary reports in \cite{nature-review}, mentioned in section \ref{section-science-beyond-pandemic-modeling});
+
+\item the survey is not intended to be a formal, systematic sample of scientific research in general, but is intended to be sufficient to dispel the possibility that the issues described above earlier in this paper are isolated practice unique to a few papers in epidemiology, perhaps an idiosyncrasy of a few authors in a particular field, or perhaps due to an initial chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they might have just happened to be Software Engineering outliers);
+
+\item the code/data policies of the \plural{\numberOfJournals}{journal} condoned at the time of the sample \emph{and continue to condone\/} poor practice at the time of writing the present paper (\monthname{\month} \the\year) --- for specific details and further explanation of the problems, see \supplement\ section \ref{supplementary-journal-policies-section};
+
+\item
+the fact that the specifically identified problems are elementary errors in Software Engineering (see the discussion in section \ref{section-discussion}) suggests more sophisticated analysis is not required;
+
+\item 
+finally, the present paper's \LaTeX\ source, as well as all documented code and data, are available from a repository, which provides a convenient framework for easily refining or developing the research as may be desired (see details at the end of this paper).
+\end{enumerate}
+
+The \the\dataN\ papers surveyed cover a range of specialities, and it is unlikely that non-specialists can properly assess the code from the point of view of the specialism, not least because many of the papers sampled require specialist code libraries (and in the right combinations of versions) to be run that not everyone will have or be able to install.
+Code quality was therefore assessed by reading it --- due to the paper authors' complex and/or narrative interpretation of data, code, data and hardware/operating system dependencies, no assessment could realistically be made whether the code provided actually reproduced a paper's specific claims.
+Indeed, if we trust the papers that their code was actually run and provides the results as reported, then running their code (when provided in full) would merely check the paper/code consistency but will not assess the quality or reliability of the code.
+Indeed, in most scientific papers there are layers of expert scientific work, interpretation and abstraction, lying between the computational models and the report in the paper.
+
+\highlightmysubsection{Summary of results}
+\label{summary-of-results}
+The sample selection criteria necessarily identified scientific research with Software Engineering contributions.
+
+No evidence of verification and validation was seen.
+There was only one example of very basic Software Engineering methods, namely independent coding, and even then the independent code used for testing was not uploaded to the paper's code repository, so the independent testing is not available for reviewers or readers of the paper.
+(See \supplement\ for more details.)
+
+There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
+No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
+(See the \supplement\ for a list of standard methods.)
+
+Much code depended on specific software versions, specific libraries, and substantial manual intervention to compile it.
+All code (where actually provided) was sufficiently complex that, if it was to be used or scrutinized, required more substantial documentation than was provided.
+
+On the whole, on the basis of the sample evidence, scientists do not make their code \emph{usably\/} available, and rarely provide adequate documentation (see table \ref{supplement-table-summary}).
+
+With the one minor exception, no papers reported anything on any Software Engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
+The papers themselves, typically only a few published pages, are very brief compared to the substantial code they rely on (see table \ref{table-repo-summary}).
+
+With the one exception, none of the papers used any specific Software Engineering methods, such as open source \cite{open-source} or other standard methodologies provided in this \supplement, to help manage their processes and help improve quality.
+Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reasons the present paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}
+
+\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual publishing practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
+In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach.
+The implication is that the fields, and the editorial expertise of leading journals, misunderstand and dismiss code policies --- they (or their editors and reviewers) are technically unable to assess them.
+This lack of expertise is consistent with the limited awareness of Software Engineering best practice that is manifest in the published papers (and resources) themselves.
+
+Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \clonedate\
+while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could never have been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
+
+The sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
+
+Overall, there was no evidence that any code had been developed carefully, let alone by using recognized professional Software Engineering methods. In particular, 
+\ifnum \countCodetested=0
+no papers
+\else only \plural{\countCodetested}{paper}
+\fi in the survey provide any claims or evidence of effective testing, for instance with evidence that tests were run on clean builds.
+{While it may sound unrealistic to ask for evidence on software quality in a paper written for another field of science, the need is no less than the need for standard levels of rigor in statistics reporting, as discussed in the opening of this paper.}
+
+Data repositories (the Dryad Digital Repository, Figshare or similar) were used by \plural{\counthasDataRepository}{paper} to provide structured access to their data.
+Unlike GitHub, which is a general purpose repository, Dryad has scientifically-informed guidelines on handling data, and all papers that used Dryad provided more than just their raw data --- they provided a little, sometimes substantial, documentation for their data.
+At the time of writing, Dryad is not helpful for managing code --- its model appears to be founded on the requirement that once published papers must refer to exactly the data they used, so further refinements on the data (or code) are taboo, even with version control.
+
+\highlightmysubsection{Current code policies of sampled journals}
+\globalLabel{supplementary-journal-policies-section}
+
+It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{detecting-against-error} and \ref{restrictive-policies}.
+In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
+
+Note that journal policies were first accessed on 29 July 2020, close after the period covered by the pilot survey, thus presumably fairly closely reflecting the policies in use for the papers in the survey on the dates when they were each submitted.
+Unfortunately for the survey, the journals do not make clear what exact policies were applied to each paper when the papers were submitted (on the other hand, the survey shows that policies are not rigorously enforced).
+\\
+
+\newdimen\tabwidth
+\tabwidth=\textwidth
+\advance \tabwidth by -4em
+\def\specifyurl{\vskip 3mm
+\begingroup\raggedleft\scriptsize
+} 
+\def\specifyurletc#1{\endgroup\\
+\hbox{ }\hfill\begin{tabular}{|p{\tabwidth}@{}}
+#1
+\end{tabular}\vskip .3cm
+}
+
+{\sf
+\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
+--- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
+These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
+[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
+Data and code should be deposited in a form that will allow maximum reuse.
+As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.}
+
+\specifyurl
+\url{royalsociety.org/journals/authors/author-guidelines/#data}
+\specifyurletc{Since first accessed 29 July 2020, the policy has been revised (undated, accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
+The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
+
+\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
+{\sf --- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
+[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.}
+
+\specifyurl
+\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
+\specifyurletc{Accessed 29 July 2020; since updated (accessed 2 February 2022) to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
+Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.''
+}
+
+\noindent\textbf{\emph{Lancet Digital Health\/} author guidelines}\\
+Journal has detailed data policies, but no code policy.
+
+\specifyurl
+\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
+\specifyurletc{Accessed 29 July 2020. Still no code policy when accessed 2 February 2022.}
+
+\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
+The \emph{Journal of Vascular Surgery\/} has detailed data policies, but no code policy.
+While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant:
+
+\vskip 3mm \noindent {\sf --- The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
+[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.}
+\specifyurl
+\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
+\specifyurletc{Accessed 29 July 2020. Policy unchanged when accessed 2 February 2022.}
+
 \label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
+\highlightmysubsection{Assessment criteria and methods}
+\label{data-methods-subsection}
 A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
@@
 
-\begin{change}
 Although the pilot survey is not a systematic review, following Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) \cite{prisma} it is good practice to disclose details of the reviewers.
@@
 This is, of course, a limitation of the study.
-However, the reviewer is a full professor of computer science, who has taught and assessed computer software since the 1970s, using moderated and peer-reviewed processes for undergraduate and postgraduate computing degrees, and is well aware assessing code quality has been a lively topic in software engineering for decades (there is now international standard ISO/IEC 9126, updated to ISO 25000).
+However, the reviewer is a full professor of computer science, who has taught and assessed computer software since the 1970s, using moderated and peer-reviewed processes for undergraduate and postgraduate computing degrees, and is well aware assessing code quality has been a lively topic in Software Engineering for decades (there is now international standard ISO/IEC 9126, updated to ISO 25000).
 The author has written legal documents analyzing software for criminal cases involving faulty software.
@@
 The evaluations performed for the present paper are at a trivial level where sources of bias should have a negligible effect, particularly given that the overall conclusion is consistent across both the diverse sample and the computational science-based papers cited in the paper that did not form part of the selected sample.
-As said in the body the paper: \emph{the fact that the specifically identified problems are elementary errors in software engineering (see the discussion in \emph{[main paper]\/} section \ref{section-discussion}) suggests more sophisticated analysis is not required}.
+As said in the body the paper: \emph{the fact that the specifically identified problems are elementary errors in Software Engineering (see the discussion in \emph{[main paper]\/} section \ref{section-discussion}) suggests more sophisticated analysis is not required}.
 
@@
 %The data is also provided in full later in this \supplement.
-\end{change}
 
@@
 \codeIndent = 1em
-\input generated-example-data.tex
+\vskip 1ex
+\vbox{%
+  \input generated/example-data.tex
+}
+\vskip 1ex
 
 The data was entered by hand (as JSON terms), after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
+In total there are \the\countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
 (A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
@@
 
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
+The JavaScript data+program generates files from the JSON; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.
+
+For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
 
@@
 
-\begin{change}For each of the \plural{\dataVariableCount}{variable} used in the paper, JavaScript generates a \LaTeX\ header file that declares and assigns a calculated value.
-The header file is then included in the paper using \LaTeX's standard \texttt{\textbackslash input} command.
+Each of the \plural{\dataVariableCount}{variable} used in the paper were defined in automatically-written \LaTeX\ header files that declare them and assigns appropriate values.
+The header files are included in the paper using \LaTeX's standard \texttt{\textbackslash input} command.
 Here is an example of one such automatic definition:
@@
 so the named value is then available for the author to use when the paper is typeset.
-The tables generated for the present paper also include data that did not require names for explicit use elsewhere in the \LaTeX\ files, so these numbers are not counted in the specific \the\dataVariableCount\ variable count.
-\end{change}
 
 Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary}, as used in the main paper.
+For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (as it happens, using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary} for this \supplement.
 
 The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
+
+% This commented-out sentiment is repeated later...
+% Finally, because of their approach to automatic typesetting, some journals require that no files are explicitly \texttt{input} into \LaTeX, so a simple program is provided in the GitHub repository to recursively expand all \texttt{input} commands before submission.
+If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.
+It is curious that the journals cannot do this for themselves.
 
@@
 \multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we \begin{change}think we\end{change} know what we are reading; at least we know what we meant to write.
+\the\t\ --- we hardly bother to pay attention because we think we know what we are reading; at least we know what we meant to write.
 Worse, the more often we proof read a document, the more we remember, so the better we know what we think we said, and the more casual our proof reading becomes.
@@
 For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-\begin{change}The approach turns a possibly-hard-to-spot \emph{single\/} error into \emph{multiple\/} errors spread throughout the paper into different contexts, thus increasing the chances of noticing the error.
+The approach turns a possibly-hard-to-spot \emph{single\/} error into \emph{multiple\/} errors spread throughout the paper into different contexts, thus increasing the chances of noticing the error.
 
@@
 As the number is proof read, the different contexts encourage it to be read more carefully, and in different ways.
-\end{change}
 
@@
 \LaTeX\ itself makes it very easy to check that numbers fall within reasonable ranges, or to have any other required safety properties.
-For the present paper, a potential problem is \begin{change}if\end{change} the paper is mistakenly typeset \emph{before\/} the latest JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense \begin{change}(e.g., from a debugging run of \texttt{data.js}).
-Variables might be checked as follows:\end{change}.
+For the present paper, a potential problem is if the paper is mistakenly typeset \emph{before\/} the latest JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense (e.g., from a debugging run of \texttt{data.js}).
+Variables might be checked as follows:
 
@@
 The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\begin{change}
+Note that failing a sanity check indicates a problem that needs to be fixed, but passing a sanity check does not prove a paper correct, but the more sanity checks that are passed (and the harsher the checks) the more confidence we can have that the data has been processed correctly.
+Of course, when formal methods are employed in the software development process, the confidence in correctness can be very high.
+
 \highlightmysubsection{Defending against system problems}
@@
 
-To protect against version, portability and other problems, the Github repository for the present paper includes a check on software versions and a checksum check for all possibly affected files, including the data file.
+To protect against version, portability and other problems, the GitHub repository for the present paper includes a check on software versions and a checksum check for all possibly affected files, including the data file.
 This does not solve the problem, but it ensures anyone developing or reproducing the paper's work will at least be forewarned of potential version or portability problems.
-The Github repository itself can be used to restore files that have been corrupted.
+The GitHub repository itself can be used to restore files that have been corrupted.
 
 \highlightmysubsection{Problems of restrictive journal policies}
-\label{restrictive-policies}
+\globalLabel{restrictive-policies}
 Automatically generated variables are used throughout the paper and this \supplement.
@@
 Typically they also do not support running any data collection or analysis either (which the present paper does when it clones repositories).
-
-Another program (\texttt{expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
+These policies undermine the drive towards RAP and \RAPstarp.
+
+Another program (\texttt{programs/expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
 Of course, the expanded version now contains all variables as fixed constants, so the submitted paper is misleading and useless to other researchers if the data is modified --- the effort to ensure all published numbers are automatically correct is defeated.
 Such restrictive publishing policies undermine reproducibility.
-\end{change}
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{detecting-against-error} and \ref{restrictive-policies}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-\newdimen\tabwidth
-\tabwidth=\textwidth
-\advance \tabwidth by -4em
-\def\specifyurl{\vskip 3mm
-\begingroup\raggedleft\scriptsize
-} 
-\def\specifyurletc#1{\endgroup\\
-\hbox{ }\hfill\begin{change}\begin{tabular}{|p{\tabwidth}@{}}
-#1
-\end{tabular}\end{change}\vskip .3cm
-}
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.}
-
-\specifyurl
-\url{royalsociety.org/journals/authors/author-guidelines/#data}
-\specifyurletc{Accessed 29 July 2020; the policy has been revised (undated, but accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
-The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
-{\sf --- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.}
-
-\specifyurl
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
-\specifyurletc{Accessed 29 July 2020; since updated to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
-Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.'' (accessed 2 February 2022).
-}
-
-\noindent\textbf{\emph{Lancet Digital Health\/} author guidelines}\\
-Journal has detailed data policies, but no code policy.
-
-\specifyurl
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\specifyurletc{Accessed 29 July 2020. Still no code policy (accessed 2 February 2022).}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
-Journal has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-\vskip 3mm \noindent {\sf --- The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.}
-\specifyurl
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\specifyurletc{Accessed 29 July 2020. Policy unchanged when accessed 2 February 2022.}
-
-\highlightmysubsection{Sample assessments and criteria}
+
+\highlightmysubsection{Sample assessment and scoring}
 
@@
 \noindent
-Legend:\\
-\input generated-legend.tex
-
-\newpage
+\vbox{\noindent Legend:\vskip 1ex
+\input generated/legend.tex
+}
+%\newpage
 
@@
 \def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
+\begin{longtable}{@{}cp{2.8in}p{2.8in}@{}}
 \sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
+\input generated/assessments.tex
 \end{longtable}
@@
 
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
+\makeatletter    
+    \renewcommand{\bibitem}[2][DEFAULT]{%
+    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
+        \parshape 2 0em \columnwidth \labelwidth \shortercol
+        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
+        \global\advance\bibciten by 1
+        \ignorespaces 
+    }
+\makeatother
+
+\bibliographystyle{revisedCompJ}
+
+\global \newdimen \shortercol 
+\renewenvironment{thebibliography}[1]{%
+	\section*{\uppercase{\refname}}
+    \startBibliography
+    \sectioncontents{\unnumberedrefname}
+    \raggedright
+    \parindent=0em
+    \setbox0=\hbox{\hskip 1ex[#1]}
+    \labelwidth=\wd0
+    \advance \labelwidth by 1ex
+    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
+}{}
+
+\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper; the following references are exclusively cited in this \supplement.
+\vskip 2ex}
+
+\immediate\write\infofileForMain{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
+
+{\newcount \bibitemn \bibitemn = 0
+\let\oldbibitem=\bibitem
+\def\bibitem#1{\ifnum \bibitemn = 0 \else \egroup\vskip 1.2ex plus 1ex \fi
+\advance \bibitemn by 1
+\vbox \bgroup \oldbibitem{#1} }
+\def\endofbibliography{ \egroup }
+  \bibliography{paper-seb-supplementary-material.bib}
+}
+
+\initialiseBibliography{References for surveyed papers}{\the\bibciten}{}
 
 \begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
 \raggedright
 
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
+\def\maprepo#1{\immediate\write\infofileForSupplement{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
+
+\input generated/supplementary-references.tex
 
@@
 
-\immediate\closeout\infofile
+\immediate\closeout\infofileForMain
 \immediate\closeout\contentsfile
 \immediate\closeout\minicontentsfile
+\immediate\closeout\infofileForSupplement
 \rememberpage

### 2023-01-31 data.js

@@ -1,1336 +1,0 @@
-// Harold Thimbleby, 2020-2021
-// JSON data and Javascript code to process it
-
-'use strict';
-
-// documentation for fields in the JSON data
-// you don't need to use all fields in the data (flags default to unset)
-// first we start with definitions of all the data fields
-var descriptionsOfFields = {
-    accessed: {
-        documentation: "Date when accessed",
-        essential: 1
-    },
-    doubleChecked: {
-        documentation: "Date when double checked",
-        essential: 1
-    },
-    authors: {
-        // surname initials comma-separated list
-        // code can be modified to generate author lists in other formats as required by journal
-        // search for function fixAuthors(s) to see what we do:
-        //     list comma-separated authors, final one with 'and' but then say et al if more than maxAuthors authors
-        documentation: "Authors",
-        essential: 1
-    },
-    codeComment: {
-        documentation: "Evaluator's notes about code",
-        essential: 1
-    },
-    dataComment: {
-        documentation: "Evaluator's notes about data",
-        essential: 1
-    },
-    doi: {
-        documentation: "DOI"
-    },
-    hasBreach: {
-        documentation: "Paper breaches journal code policy (see section~\\ref{supplementary-journal-policies-section})",
-        flag: "$\\sf P_{{\\mbox{\\scriptsize c-breach}}}$" // how a flag will be abbreviated in the paper text
-    },
-    hasCodeInPrinciple: {
-        documentation: "Paper says source code is available in principle",
-        flag: "$\\sf S_p$"
-    },
-    hasCodePolicy: { // flag value set automatically from journal name
-        documentation: "Journal has a code policy (see section~\\ref{supplementary-journal-policies-section})",
-        flag: "$\\sf P_c$"
-    },
-    hasDirectCode: {
-        documentation: "Paper or URL provides source code",
-        flag: "$\\sf S_{+}$"
-    },
-    codeURL: {
-        documentation: "URL (typically GitHub) for code repository"
-    },
-    hasDataRepo: {
-        documentation: "Paper uses a data repository (e.g., Dryad, Figshare, GitHub)",
-        flag: "$\\sf R_d$",
-        dataFlag: 1 // this flag is for the data comment (defaults to code flag)
-    },
-    hasCodeRepo: {
-        documentation: "Paper uses a code repository (e.g., GitHub)",
-        flag: "$\\sf R_c$"
-    },
[…]

### 2023-01-31 expand.js

@@ -1,63 +1,0 @@
-// Harold Thimbleby, 2020-2022
-// Javascript code to expand TeX \input commands
-
-var TeXFiles = ["paper-seb-main.tex", "paper-seb-supplementary-material.tex"];
-
-'use strict';
-
-var fs = require('fs');
-
-function saveFile(fileName, string) {
-    fs.open(fileName, 'w', function (err, fd) {
-        if (err) {
-            if (err.code === 'EEXIST') {
-                console.error(fileName + " can't be over-written");
-                return;
-            }
-            console.error("** Some error trying to write " + fileName);
-            throw err;
-        }
-        fs.writeFileSync(fd, string, function (err) {
-            if (err) {
-                return console.log("** Saving " + fileName + " got error: " + err);
-            }
-        })
-    });
-}
-
-function getFile(file) {
-    try {
-        return fs.readFileSync(file, 'utf8');
-    } catch (err) {
-        console.log("** Can't read file " + file + " - " + err);
-        return "\\error couldn't read " + file + "\n";
-    }
-}
-
-var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}");
-
-var bufferExpand, bufferInsert;
-
-for (var i = 0; i < TeXFiles.length; i++) {
-    console.log("Recursively expand all \\input and \\bibliography macro calls in " + TeXFiles[i]);
-    bufferExpand = getFile(TeXFiles[i]); // puts in buffer
-    var replaced;
-    do {
-        replaced = false;
-        bufferExpand = bufferExpand.replace(re, function (match, filename) {
-            replaced = true;
-            console.log("        \\input " + filename);
-            bufferInsert = getFile(filename);
-            return "% expand " + filename + "\n" + bufferInsert + "% end expanding " + filename + "\n";
-        });
-        bufferExpand = bufferExpand.replace(rebib, function (match, filename) {
-        	replaced = true;
-            console.log("      * \\bibliography{"+filename+".bib} replaced with " + filename+".bbl");
-            bufferInsert = getFile(filename+".bbl");
-            return "% expand bibliography " + filename + ".bbl\n" + bufferInsert + "\n% end expanding bibliography " + filename + ".bbl\n";
-        });
-    }
-    while (replaced);
-    saveFile("expanded-" + TeXFiles[i], bufferExpand);
-    console.log("  ->  Saved as: " + "expanded-" + TeXFiles[i] + "\n");
-}

### 2023-01-31 generated-allGitRepos.sh

@@ -1,45 +1,0 @@
-rm -rf "git-MetricSelectionFramework"
-git clone "http://github.com/ChristophKanzler/MetricSelectionFramework" "git-MetricSelectionFramework"
-
-rm -rf "git-PENet"
-git clone "http://github.com/marshuang80/PENet" "git-PENet"
-
-rm -rf "git-PostoperativeOutcomes_RiskNet"
-git clone "http://github.com/cklee219/PostoperativeOutcomes_RiskNet" "git-PostoperativeOutcomes_RiskNet"
-
-rm -rf "git-philter-ucsf"
-git clone "http://github.com/BCHSI/philter-ucsf" "git-philter-ucsf"
-
-rm -rf "git-AI-CDSS-Cardiovascular-Silo"
-git clone "http://github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo" "git-AI-CDSS-Cardiovascular-Silo"
-
-rm -rf "git-SiameseChange"
-git clone "http://github.com/QTIM-Lab/SiameseChange" "git-SiameseChange"
-
-rm -rf "git-lactModel"
-git clone "http://github.com/SolangeD/lactModel" "git-lactModel"
-
-rm -rf "git-LRM"
-git clone "http://github.com/UT-NSG/LRM" "git-LRM"
-
-rm -rf "git-manifold-ga"
-git clone "http://github.com/ki-analysis/manifold-ga" "git-manifold-ga"
-
-rm -rf "git-blast-ct"
-git clone "http://github.com/biomedia-mira/blast-ct" "git-blast-ct"
-
-( echo "% date generated by running generated-allGitRepos.sh downloading Git repos";
-echo \\\\def\\\\clonedate {\\\\ignorespaces `date "+%e %B %Y"`}
-echo \\\\def\\\\cloneyear {`date "+%Y"`}
-echo \\\\def\\\\clonemonth {`date "+%m"`}
-) > ../generated-clone-date.tex
-
-(echo % shell-script calculated metavariables; echo "\\\newcount \\\JSONerrorCount";echo "\\\newcount \\\dataVariableCount") > ../generated-metadata.tex
-
-# how many errors can data.js find?
-echo "\\\JSONerrorCount =" `grep "error(" ../data.js|wc -l` >> ../generated-metadata.tex
-
-# how many variables does data.js define?
-echo "\\\dataVariableCount =" >> ../generated-metadata.tex
[…]

### 2023-01-31 generated-open-doi.sh

@@ -1,32 +1,0 @@
-open "http://www.doi.org/10.1038/s41746-020-0300-0"
-open "http://www.doi.org/10.1038/s41746-020-0286-7"
-open "http://www.doi.org/10.1038/s41746-020-0287-6"
-open "http://www.doi.org/10.1038/s41746-020-0268-9"
-open "http://www.doi.org/10.1038/s41746-020-0278-7"
-open "http://www.doi.org/10.1038/s41746-020-0266-y"
-open "http://www.doi.org/10.1038/s41746-020-0265-z"
-open "http://www.doi.org/10.1038/s41746-020-0248-0"
-open "http://www.doi.org/10.1038/s41746-020-0258-y"
-open "http://www.doi.org/10.1038/s41746-020-0261-3"
-open "http://www.doi.org/10.1038/s41746-020-0249-z"
-open "http://www.doi.org/10.1038/s41746-020-0255-1"
-open "http://www.doi.org/10.1098/rsos.200419"
-open "http://www.doi.org/10.1098/rsos.192210"
-open "http://www.doi.org/10.1098/rsos.191900"
-open "http://www.doi.org/10.1098/rsos.200566"
-open "http://www.doi.org/10.1098/rsos.200422"
-open "http://www.doi.org/10.1098/rsos.200462"
-open "http://www.doi.org/10.1098/rsos.191992"
-open "http://www.doi.org/10.1098/rsos.200622"
-open "http://www.doi.org/10.1098/rsos.200638"
-open "http://www.doi.org/10.1098/rsos.200621"
-open "http://www.doi.org/10.1098/rsos.191928"
-open "http://www.doi.org/10.1098/rsos.200307"
-open "http://www.doi.org/10.1098/rsos.191653"
-open "http://www.doi.org/10.1098/rsos.200320"
-open "http://www.doi.org/10.1016/S2589-7500(20)30108-4"
-open "http://www.doi.org/10.1016/S2589-7500(20)30107-2"
-open "http://www.doi.org/10.1016/S2589-7500(20)30131-X"
-open "http://www.doi.org/10.1016/S2589-7500(20)30063-7"
-open "http://www.doi.org/10.1016/S2589-7500(20)30085-6"
-open "http://www.doi.org/10.1016/S2589-7500(20)30078-9"

### 2023-01-31 response-to-referees/paper-seb-main (original).tex
More

\documentclass{comjnl}
\usepackage{ragged2e}
\usepackage{amsmath}
\usepackage{amssymb}

%----
  \setlength\textheight{220mm}
%----

\input paper-seb-macros.tex
\input generated-info-for-main.tex
\input generated-page-lengths.tex

\long\def\ignore#1{}

%\linenumbers

\title{\mytitle} 
\author{Harold Thimbleby}
\shortauthors{Harold Thimbleby}
\affiliation{See Change Fellow in Digital Health, Swansea University}
\email{harold@thimbleby.net}

%\subject{Computational science (bioinformatics,~medical~computing,~etc)}

\keywords{Computational Science; Software Engineering; Reproducibility; Scientific scrutiny; Reproducible Analytic Pathway\raise 1ex\hbox{$\star$} (\italicRAPstar\hskip -1pt)}

\received{30 April 2022}
%\revised{--}

%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
%&\texttt{harold@thimbleby.net}\\
%&ORCID 0000-0003-2222-4243
%\end{tabular}

%\orcid{0000-0003-2222-4243}

\begin{document}

\begin{abstract}
\justifying
\noindent
While data, analysis, and the results of running scientific models (including AI and ML) are described in peer reviewed papers, the algorithms and code generating insights and results often avoid adequate scrutiny. If code is not correct, then the contributions of published papers relying on unchecked code will be suspect.

\hskip 1em This paper shows that scientists rarely assure the structure and quality of code they rely on, rarely make full code available for wider use or scrutiny, and rarely provide adequate documentation to understand or use their code reliably. This paper therefore justifies and proposes ways to mature the computational sciences:

%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available. Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is also fallible. When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions. 
%The paper reports a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.

%Code can be improved using Software Engineering. This paper argues for specific solutions:

%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.

%\noindent\emph{Solutions:}

\raggedright
\newdimen \mywidth \mywidth=\textwidth
\advance \mywidth by -3em

\def\indented#1{\vskip 1ex
\parshape 2 0em \textwidth 2em \mywidth 
\noindent 
\hbox to 2em{ #1.\hfill}\ignorespaces}

\indented 1
Professional Software Engineers can help and should be involved, particularly in critical research such as public health, climate, etc;

\indented 2
``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;

\indented 3
Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent Software Engineers;

\indented 4
\begin{change}The Reproducible Analytic Pathway (RAP) methodology should be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces and terms \RAPstarp. Furthermore \RAPstar\ should be supported and encouraged in journal, conference, and funding body policies.
\end{change}

\vskip 1ex 
\noindent
The \supplement\ for this paper provides a summary of professional Software Engineering best practice relevant to scientific research and publication. It also includes suggestions for \RAPstar\ processes, and a pilot survey of code quality in leading peer-reviewed journals that corroborates the concerns of the paper.

\begin{quote}\def\seprule{\vskip 2ex\hrule\vskip 2ex}
\seprule

\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
\zq Science is what we understand well enough to explain to a computer.''\\
\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\

\zq I have to write to discover what I am doing.''\\
\hfill {\rm Flannery O'Connor, quoted in \emph{Write for your life\/} \cite{write-for-life}}

\zq Criticism is the mother of methodology.''\\
\hfill {\rm Robert P. Abelson in \emph{Statistics as Principled Argument\/} \cite{abelson}}

\seprule
\end{quote}
\end{abstract}
\maketitle

%\section*{Author summary}
%
%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
%Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.

\section{Introduction}
\begin{change}
Unreliable, often unstated and unexplored, computational dependencies in science are widespread. Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction. In short, the quality of much computational science (including AI and ML) is undermined because the code it relies on is rarely of adequate quality for the uses to which it is put.

This paper explores the nature and extent of these Software Engineering problems in published science. The paper suggests ways to improve the reliability of computational sciences.

Code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny. Developing justifiably reliable code is the concern of the field of Software Engineering, which is discussed further below, as well as substantially in the paper's \supplement, which is an integral part of the suggested solutions and literature survey.

This paper makes an explicit analogy between reliable computation and reliable statistics, a field that has adopted clear standards for reliable use and presentation. The \supplement\ (section~\ref{supplementary-Spiegelhalter-section}) summarizes Spiegelhater's uncontroversial statistics probes and  --- just as it is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.

\begin{table*}
\begin{center}\normalsize
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}
\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}.}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\normalsize
\input generated-summary-table.tex
\end{center}
\caption{Summary of a pilot survey of computational science papers. Full details of the survey, all data, and methodology are in the \supplement.}
\label{table-summary}
\end{table*}
\end{change}

The body of this paper reviews some specific epidemiological computational science and its avoidable problems. The \supplement\ further describes a broad pilot survey of papers covering a wide range of computational science published in leading journals, confirming that the issues raised in the body of the paper are widespread and not specific to science that may have been rushed because of relevance to the COVID-19 pandemic. The quality of the supporting code in both epidemiological science and in the survey was very weak (see a summary in Tables \ref{table-overview} \& \ref{table-summary}). No papers provided any evidence their code was adequately tested or rigorously developed; none used methodologies like RAP or \RAPstarp, and only one paper discussed any relevant Software Engineering methods (independent coding). Although in the sample 81\% of papers were published in journals that had code policies (which themselves are weak), 42\% of surveyed papers in those journals breached their code policies. One paper declared it had accessible code, but the relevant repository was and remains empty. Such results about code are comparable to surveys of data and data access \citeeg{machine-learning-reproducibility,no-raw-data,data-access}, and form part of the wider reproducibility crisis \citeeg{reproducibility-crisis}.

\begin{change}
\section{Background}
The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences. There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature. For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in. Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}. %For example, if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate. 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theory development and technical maturation, to having robust, off-the-shelf instruments that measure temperature in reliable, internationally standardized units.

Far more flexible and challenging than thermometers, computers are a unique, new technology. Understanding computers and integrating them into science is likely to take longer than intrumentalizing temperature did. Computation not only expands science's paradigms and supports new discoveries, but it can also \emph{do\/} new science --- and with AI, robotics, and other techniques, computation can do both real and speculative investigations autonomously, and better than humans.

``Computational science'' has come to mean a particular style of science, based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational. Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ in all fields of science, computation is used at every step, from note taking, sound and photographic recording, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting finished publications.

All areas of science are now being profoundly computerized, and we need to do it well.

\subsection{The statistics/computation analogy}\label{statistics-computation-analogy}
The central role of computational methods in science may be fruitfully compared to statistics, an established scientific tool --- and also, of course, like computer science, a substantial field of research in its own right.

Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics. Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians. 
\end{change}

In good science, all statistics, methods and results are reported very carefully and in precise detail \cite{apa-statistics,Mariusz-reporting}, generally following strict journal or disciplinary guidelines. For example, a statistical claim might be summarized as follows:

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.0009)$. Between group effect size $d = 0.55$ $(95\%$ CI $0.32,$ $0.77)$.'' \cite{example-stats}
\end{quote}

This typical wording formally summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize. It is a \emph{lingua franca}. It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results. Moreover, behind any such brief paragraph is a substantial, rigorous, and appropriate statistical analysis.

Scientists write like this and conferences and journals require it because statistical claims need to be properly accountable and documented in a clear way. Authors, too, must avoid the easy temptation of being unjustifiably over-confident with their own results.

Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Spiegelhalter-section}). Results should not be uncritically accepted just because they are claimed. The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted; in fact, there is widespread concern about the poor quality of statistics in science \cite{paul-hci,pnas-stats}. While it is assumed that statistics should be peer reviewed, and that review will often lead to improvement, these critical papers show that reviewers and editors are often failing to pick up on poor statistics.

Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement. Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians. Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.

Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them, will be generated by computer. Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it. Many papers now explore the contribution of AI and ML to their fields. The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers. Computers even help with the word processing and typesetting of the research.

In short, computers, data, and computer code are central to modern science, not just to the explicitly computational sciences. Some AI work is uncovering biases and ethical issues that were previously unrecognized, so computational sciences are not just routinely contributing to existing science but extending its reach and improving its quality.

However, using any code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.  Software code, like statistics, is also subject to unintentional bias \cite{Ben,se-bias}.  All these issues are non-trivial concerns requiring technical expertise to manage well. As with statistics, good answers to such ``technical'' issues makes the science that relies on them better.

A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.

Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims. Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.

We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.

An interesting overlap with statistical and Software Engineering sloppiness concerns the many papers that  disclose as part of their methodology that they used a particular package, for example

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)'' \cite{sas9.2}
\end{quote}

but without giving any further details. The paper cited above describes one of its multivariate analyses as ``multiple correspondence analysis (MCA) and ascendant hierarchical clustering'' with no specific details --- reproducing the work would be non-trivial since the methodology would have to be reconstructed from scratch, even assuming the data was made available by the authors as the paper gives no details of the data used or how to obtain it. The problem is that the common practice of declaring using a named computational system (such as SAS in this case) does not help scrutiny, as such systems can do almost anything. \emph{How\/} those analyses might have been performed is not discussed, and one assumes it follows that the analyses could therefore not have been reviewed for basic scientific competence during the publication process.

A reviewer, if nobody else, needs to actually examine the statistical code used and its documentation to assess whether the analysis presented in the paper is appropriate and sufficiently reliable. Furthermore, if the analysis actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible if it relies on idiosyncrasies in SAS\@. Of course, an author can disclose the idiosyncratic dependencies on, say, version 9.2; while this seems to be an onerous obligation, conversely it is arguable that if an author is unaware of code dependencies, then their science relying on them is equally unreliable.

It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, and to uncover unknown dependencies. Indeed, the code may be poorly written (as this paper shows is very likely), so the results derived from the code simply may not be reliable. Being able to understand (at least in principle --- e.g., relying on reviewers to have checked it) the exact code used in implementing a model is critical to having confidence in the results that rely on it.

In normal scientific reporting (outside of teaching and assessing science) details of methodology are routinely glossed. A chemist does not say they cleaned their glassware. One might argue, then, that scientists need not discuss their code in any detail because they know how to program and their code is correct. This argument is mistaken, as this paper shows. Unfortunately, code is rarely considered a valuable part of the science to which it contributes and this creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.

\subsection{The role of code \\ in science and scientific publication}
%Yet without code, models could not be run.

Models map theory and parameters to describe phenomena, typically to make predictions, or to test and refine the theory supporting the models. With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.

Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive. The chance that a thousand lines of code, hand-written or otherwise, is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability. How to do this well is the concern of Software Engineering, and is discussed throughout this paper and in the \supplement.

While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are they are long, complex, inscrutable and (our survey shows) lack adequate documentation. Therefore the models are very likely to be unreliable \emph{in principle}.

If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if it has not been clearly documented? As a scientist pursues a research career building on their previous work, how can they be certain their work is reliable, and not merely converging to their  prejudices? Without proper documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?

Without substantial documentation it is impossible to scrutinize code properly. Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications. What does it mean? What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation? Does the code mean what was intended? What are the assumptions on \texttt{k}, \texttt{x}, and \texttt{y}, and do they hold invariantly? Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown. A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of under- or overflow).

Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code. Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim. If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed. It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.

Contrary to the views expressed in the present paper, some authors assert that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarizing data or properties of the data \cite{assessing-quality}. If code is inadequate, the so-called ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways. Indeed, none of the papers sampled (described in section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.

Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes. The question is whether it matters, \emph{when\/} it matters, and, if so, when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work?

\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
It is a misconception that programming is easy and even children can do it \cite{fixit}. More correctly, toy programming is easy, but real programming is very difficult.

An analogy helps. Building houses is very easy --- indeed, many of us have built toy Lego houses. Obviously, though, a Lego house is not a \emph{real\/} house. It is not large enough or strong enough for safe human habitation! This point is obvious because we can see Lego houses, and everyone is familiar with building-block play; its real-world engineering limitations are too obvious to need stating.

In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also invisible. The ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed, and because --- like Lego --- toy programs can look good but be unreliable, difficult to use and even dangerous. Thinking programming is easy is like appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations. These are professional engineering issues that Lego builders ignore. Certainly, even real building is much easier and faster when the technical details are ignored, as anyone who has experienced a cowboy builder can attest.

Unlike building (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a very new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture. Relevant professional standards are not enforced. Problems for the reliability of science arise when legitimate doodling and tweaking software drifts into claiming scientific results that do not have reliable engineering structures underpinning them, let alone the properly developed and documented archived code, to justify them. In many countries, there are laws that require all but the very simplest structures to be formally approved from plans and inspected as they are built, but who writes plans for software, who inspects scientific models as they care coded? Yet the consequences of building a shoddy garage have negligible impact compared to the consequences of writing shoddy code that informs national public health policies or climate change interventions.

Quite different, more formal, reasoning confirms the misconception about programming. Lambda calculus ($\lambda$-calculus) is one of the very simplest of programming notations \cite{lambda}; it is just made up of strings of the four symbols $\lambda.()$ plus other symbols that have no pre-defined meanings. It is hard to think of a programming language that could look simpler, although of course anything can be written out in binary as a ``trivial'' sequence of \texttt{1}s and \texttt{0}s, but at least lambda calculus has a visible structure.

Lambda calculus is powerful enough to do what any computer program can do; it is Turing Complete.

Now consider this brief lambda expression: $\lambda t. (\lambda x. t (x x)) (\lambda x. t (x x))$. This expression  (the paradoxical combinator) allows us to easily define recursive functions \emph{without\/} any recursion. Such ``simple'' lambda calculus expressions are clearly profound. Indeed, lambda calculus has a place in the history of logic because it was the source of the first undecidable logic problems (the Halting Problem being perhaps the most famous example). Lambda calculus thus proves that simple programs are \emph{not\/} simple; just because children can write programs does not mean anybody, including the children, can understand them. Indeed, ``programming is easy'' is contradicted by the Halting Problem, which shows that at least one trivial question about a program, whether it halts, is in general unanswerable.

Since programming appears to be so easy, developing code has low status in scientific practice. Developers of code are rarely acknowledged in scientific papers. The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the technical contributors to it. Because it is easy, there is no need to work hard to make it correct --- anyway, it obviously works. While such views prevail, the vicious cycle is that the low status means software development is done casually, which reinforces the low status.

In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code}, because code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.

Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement). Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.

\subsection{The central role of code is ignored}\label{central-role-of-code}
Almost all scientific papers \emph{routinely\/} describe their experimental method, their data handling, and provide an overview of their analytic (usually statistical) methods. If they are theoretical papers, they will describe their mathematical models and any data that is used to confirm their models. Outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods avoid or control for experimental error. The relative silence on code is curious as science now depends on correct code as much as it ever depended on controlled experiments. While scientists avoid or manage contaminated samples, they ought to avoid or manage code bugs which have the same confounding impact.

It is important that experiments and analysis are performed reliably and ethically. Because statistics, like code, is so readily susceptible to uncontrolled bias and error, there are many protocols and journal policies that enforce best practice, for example journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature. Yet PRISMA, like many such policies, ignores the critical role of code, and ignores the Software Engineering principles that assure computation is reliable and reliably reported.

PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which is reasonable enough. PRISMA covers the review process carefully. For example, the authors should report the number of papers they included in their review. Perhaps $N=2000$. This number is then written into their draft paper, likely in several places. As the authors read and revise their paper, submit it, and respond to peer reviewers, it is very likely that the number of papers included in the survey changes, or other numbers and details may change. The authors now have a maintenance problem: where are the numbers that have changed, and what should they be changed to? Doing a search-and-replace, whether automated or by hand, is fraught with difficulties. What happens if $2000$ is used for some other purposes as well? What happens if some of the $2000$ values are written as $2,000$ or $2000.0$ and are not noticed? What happens if some $2000$ are year dates and are changed incorrectly? Then there are the Human Factors:  slips and errors will happen in this process anyway. Typos, slips during cut-and-paste, and other errors are common. Similar iterative revision cycles happen with any paper, not just with systematic reviews. Avoidable errors are problems across all of science.

If a computer model is involved in the process (as it generally would be for a systematic review) then the value of $N$ in this example could very easily be stored in a file where the paper typesetting process can access it. For example, if \LaTeX\ is the system of choice, the analysis could generate, say,

\begin{center}\texttt{$\backslash$newcommand\{$\backslash$N-papers-reviewed\}\{2000\}}\end{center}

so that when and wherever the authors write \texttt{$\backslash$N-papers-reviewed} in their paper's text, the typeset paper says 2000 or whatever the correct value happens to be at the time. If so, then \emph{whenever\/} the survey is updated, the value cited in the paper is automatically and \emph{correctly\/} updated without any further intervention from the author.

In general, not just PRISMA and not just numbers, but any data, text, graphs or tables, etc, can be reliably inserted into a paper automatically.

PRISMA says nothing about how to ensure the final results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations. Such rules reinforce the fallacy that code is trivial and unimportant.

These issues with PRISMA may sound trivial, but they are a symptom of the relaxed attitude to code that undermines the reliability of all published computational science.

\subsection{Bugs, code and programming}\label{knowledge}
Critiques of data and model assumptions are increasingly common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned. Yet data and program are formally equivalent (see \supplement, section \ref{on-code-data-publication}). Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and probably useless. Code, however, is harder to scrutinize, which means that errors in code have subtle, often unnoticed effects on results.

Almost all code contains ``magic numbers'' --- that is, data masquerading as code. This common practice ensures that published data is very rarely all of the data because it omits the magic numbers embedded in the code. Data is often ``hard coded'' in programs with no explicit representation. Such issues emphasize the need for repositories to require the inclusion of code so all data, including that embedded in the code, is actually available.

Bugs can be understood as discrepancies between what code ought to do and what it actually does. Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered. Contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination, or, better, failure to compile. None of the computational science surveyed or cited in this paper includes any such basic techniques.

If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs. Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time). For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted, but are bugs in the wider sense of producing incorrect results (e.g., ill-conditioning) that is, what was intended was wrong.

Random numbers are widely used in computational science, for simulation or for randomizing experiments. Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of bugs nai\"\i ve bugs \cite{knuth}.

\begin{change}
\subsection{The long-term problems of unreliable code}
Scientists explore and extend the boundaries of rigorous knowledge. Put briefly, the purpose of scientific experiments is to vary details to either test and specify the boundaries of theories, or to discover new phenomena that lead to theory revision.

If poor code, or poorly documented code, is made available with scientific papers, the code (or the absence of code) is a natural place to start replicating and varying experimental conditions,  including both data or code. However, if the starting point is not accurately known, whether due to bugs, obscure code, or because of poor documentation, then experimental variations will have an unknown effect. Theory will then be driven by artifacts of the code, not genuine phenomena.

In section \ref{section-pandemic-modeling}, below, an example is documented of a research code development process of at least 15 years' duration where the code was admitted to be completely undocumented, leaving details in just one author's head. None of the various related papers describe any controls over the drift of the science, or how independent researchers building on it might have been able to build with confidence rather than merely reproducing the same errors (if any).

Since the code in question was substantial and non-trivial, it is very unlikely that any constructive reproduction occurred outside the original laboratory and mindset; indeed, section \ref{reproducibility-concerns} describes how ``reproduction'' became trivialized because of community pressure to confirm the insights of this particular research.

Trying to constructively refute aspects of this research in the Popperian sense \cite{popper-conjectures-refutations} would have been impossible. For example, had the relevant papers published critical code invariants then scientists building on the research could have explored whether those invariants remained valid and, if so, under what assumptions. In fact, invariants are the theories of code, and deserve as high a prominence in published computational science as the domain theories the computational modeling itself is supporting investigating.

Since the symmetries in the scientific domains should be reflected in corresponding symmetries in well-designed code, this raises new possibilities of Formal Methods approaches (e.g., using automatic theorem provers) supporting advances in general science.
\end{change}

\section{State of the art \\ in computational modeling}
\label{section-pandemic-modeling}
A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no role for code or computation as such. It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.

A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}. The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed. Indeed, TRIPOD guidelines ignore code completely.

It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for direct human use. Flowcharts, too, should be designed as carefully as code, for exactly the same reason as it is hard to program reliably --- flowcharts can present arbitrarily complex algorithms and are often sketchy and vague on details that are critical to reliable use.

A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling. The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.

A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 
For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
'' \cite{ferguson-interview}\end{quote}

This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times. The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022. (Typical $h$ values vary by discipline.)} of 93. Ferguson should be well aware of the standards of coding used in at least his own field. This comment, quoted above, is arguably representative of the standards of the field as a whole.

Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}

Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.

If Ferguson's C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation. It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified and repurposed over fifteen years without becoming incoherent, and someone recognizing that documenting it would help make it easier to maintain.

If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results. In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.

The programming language C which was used \cite{tweet} is not a dependable language; to develop reliable code in C requires professional tools and skills. Some of the code is written in a na\"\i ve style (e.g., writing \texttt{*(a + i)} instead of \texttt{a[i]}, and with obscure numerical goto statements like \texttt{if(l == 0) goto S150}), transliterating pseudo-code, FORTRAN and Pascal code published in the 1970s and 1980s \citeeg{forsyth,Normal-Distributions}, with the risks that entails for introducing typos. %The code makes unexplained use of types such as float, double, int, 32 bit int, ...

Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, or hardware. In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is used, which manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}. (The \supplement\ discusses these issues further.)

Ferguson, author of the code, says of the criticisms of his code,

\begin{quote}\raggedright
\setbox0=\hbox{``}
\hskip -\wd0\box0 
However, none of the criticisms of the code affects the mathematics or science of the simulation'' \cite{thumbs-up}
\end{quote}

This claim is implausible. The original work on theoretical epidemiology may be fine if it does not use any of his code, but if the science is not supported by code that correctly implements the models, then the program's output cannot be relied on without independent evidence. Over the 15+ years the code was in development, the science running it informs will have drifted just as the code it relies on has.

Typically, models will be developed iteratively as their results are improved to better fit a scientist's goals --- but this, especially when it is done by tinkering, as here --- risks making the code arbitrarily fit the goals (comparable to over-fitting; see section \ref{over-fit}), rather than to objectively elucidate the science. In fact, the Ferguson computational model is very large,\footnote{Ferguson's \texttt{covid-sim} system is \covidsimkLOC\ kLOC (thousands of lines of code), composed of \covidsimfiles\ files, and uses \covidsimdata\ of data. It is now rewritten from C into C++ with Python, R, sh, YML/JSON, etc. For more details, see \supplement.} so it is implausible that ``mathematics or science'' has been correctly implemented in them. Therefore Ferguson's \emph{reported\/} science cannot be reliable. Getting the science right, which depends on correct code, is a normal requirement of \emph{reproducibility}.

\subsection{Concerns with reproducibility}
\label{reproducibility-concerns}
The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed the code could be run again and produced comparable results (compared using an Excel spreadsheet). Given that code was available, reproducing results with such a low level of sophistication is hardly surprising, as doing so does not depend on the validity of the code. That running code generates the results claimed in a paper is a very weak test, and one that should ideally have been established automatically at paper submission. \emph{If you do not know what you are reproducing, there is no scientific value in doing so.}

Unfortunately, the terms reproducibility, replicability, and repeatability, have similar meanings in English and have been used in different ways by different authors. In \cite{codecheck,thumbs-up} the reproduction amounted to re-running the original code. It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper, and if the original code runs this confirms a basic level of access for the wider scientific community.

A stronger criterion than basic reproduction is whether an \emph{independently\/} developed model developed from the same paper(s) produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues. However, much stronger scrutiny of code is required to answer essential questions including:

\newcount \enum
\enum=1
\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
\the\enum
\global\advance \enum by 1}
\begin{enumerate}\raggedright
\item Do other scientists, including reviewers and the authors, understand the code?
\item Does the code implement the methods described in the paper?
\item Has the code been over-fitted or tweaked to support specific claims in the paper?'
\item Is there a definitive version of code?
\item Is the code controlled and signed?
\item What limitations does the code have?
\item Was the code developed to any standard, and does it comply to that standard?
\item How does the code protect against data, coding, and human error?
\item Was the code tested adequately?
\item \ldots\ and so forth --- questions are numbered for reference, not because any particular number or range of questions is complete. 
\end{enumerate}

All such questions also apply to specifications, documentation, assurance cases, test procedures, and other essential documents, not just to code. In turn, the levels of scrutiny demanded should be guided by explicit conceptual claims in the paper \cite{essence-of-software} --- for example, a pilot study requires weaker assurance than code that is developed concerning nuclear power, driverless vehicles, public health, etc.

The questions in the list above are certainly hard to answer for all but the shortest systems, but corresponding levels of quality assurance are demanded for other methodologies \cite{notebooks,popper-conjectures-refutations,tripod,prisma,nih-policy,nih-nature}, such as data preparation and statistics to support claims in peer reviewed science.

Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).} Documenting code in hindsight, even if done rigorously, may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was originally intended to have done. As the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models; it will be different. It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost. The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.

Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming. Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications. Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it alone is even worse than not publishing code at all (see \cite{chinese}). Pseudo-code is not precise enough to help anyone scrutinize a model; copying pseudo-code introduces bugs.

An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}. The \supplement\ provides further discussion of reproducibility.

\subsection{Computational science \\ beyond pandemic modeling}
\label{section-science-beyond-pandemic-modeling}

Epidemiology has a high profile because of the COVID pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. Other examples that were no obviously rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code. Its \supplement, which might have provided code, no longer exists.

There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact. Climate change modeling is one such example that will have an impact long beyond the COVID pandemic.

A short 2022 summary of typical problems of Software Engineering impacting science appears in \emph{Nature\/} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration. The paper \cite{nature-review}  makes some misleading comments about the simplicity of Software Engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''

Guest and Martin in a 2022 paper \cite{psychological-modeling} promote the use of computational modeling, arguing that through writing code, one debugs scientific thinking. Psychology, their focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}. In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived. Computational models can be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks. It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely: for instance, ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973 and is now a 120 kLOC Common LISP and Python system \cite{actr}. Any paper using ACT-R would require additional code on top of the basic ACT-R framework.

The psychology paper \cite{psychological-modeling} presents an example computational model from scratch to illustrate a framework of computational science. In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not modeled. Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.

The Python code is presented without derivation; Software Engineering is ignored. The program listed in the paper certainly runs without obvious problems (ignoring some typographical errors due to the journal's publishers), but ironically the Python does not implement the mathematical specification explicitly provided for it, thus unintentionally undermining the argument of the paper.

One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses. Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct (at least when finished). Otherwise, computational modeling with inadequate Software Engineering will introduce errors into scientific thinking.

Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims. People reading the code will not know how to obtain results because they do not know exactly what was intended in the first place. The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even  flaws.

Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}. World-class software can attain maybe 1 bug per 1,000 lines of code. Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new insights into their models. In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- is subject to numerical errors \cite{hamming}. It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion). Such bugs undermine confidence in model results.

Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what researchers have done, but merely making code available (for instance, providing it in their \supplement\ with papers, putting it in repositories, or using open source) is not sufficient, for example running models may require access to special hardware.

Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available. While some \citeeg{unfinished,lancet-unfinished} make what they call ``documented'' code available, they provide no more than superficial comments. This is \emph{not\/} documentation as properly understood. Such comments do not explain code, explain contracts, nor explain algorithms. Contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming {(see the \supplement\ for a checklist of many relevant, professional Software Engineering concepts and resources)}.

Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper, and indeed in computer science research, e.g., \cite{machine-learning-reproducibility}) is inscrutable. Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do. Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.

Making outline code (including pseudo-code) available without proper documentation and without discussing its limitations is unethical: it encourages others to reproduce and build on poor work.

\section{Improving \\ the computational sciences}
\label{section-discussion}
\subsection{The emphasis on data}
Data has been at the center of science, certainly since the earliest days of astronomy collecting planetary and other information. Today it is widely recognized that lack of accessible and usable data that has already been collected limits the progress of science. Low quality data and poor access to data causes reproducibility problems, an increasingly recognized problem --- in 2015 it was estimated that \$28B a year is spent on preclinical research alone that is not reproducible \cite{preclinical-reproducibility}.

Curating data is taken seriously as a part of normal science and peer reviewed publication. Journal policies widely require appropriate discussion of data, much like they require appropriate discussion of statistics. Journals often require archiving data in standard formats so it can be accessed for reproduction in further scientific work.

There are many current activities to proceduralize and standardize the more effective curation and use of data, such as the FAIR principles (Findable, Accessible, Interoperable and Reusable) for scientific data management and stewardship \cite{fair,fair-principles}, and in the development of journal and national funder policies. For example, the 2022 update to the US National Institutes of Health data policies \cite{nih-policy} is described as a ``seismic mandate'' by \emph{Nature\/} \cite{nih-nature} in its attempt to improve reproducibility and open science. They did not include code.

These cost estimates and initiatives under-play the role of code as a critical component despite its becoming the new laboratory for almost all science. The role of code specifically in modeling is discussed throughout this paper; without bespoke code, proposed models (unless very abstract) cannot make a quantifiable contribution to the literature. More generally, much data is embedded in code, and in the limit code and data are indistinguishable (see \supplement). Code has additional problems of versions and compatibility beyond those of data, for example suitable compilers to run old code may no longer be available, and --- worse --- programming systems may silently produce different results when used on different computers.

In general, without proper management of code --- for example to detect and report version control differences --- sharing code may even be counter-productive.\footnote{The data and code shared with the present paper includes cryptographic checksums; if somebody reproducing the work described here does not obtain the same checksums at least when they start their work, then there are problems that need investigation before relying on the reproducibility of the data.}

Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers. The evidence (see \supplement) suggests that, generally, some but rarely all develop code that is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no evidence (before, during, or after the date of the survey sample) that any published code was maintained using repositories. This is consistent with finished code being uploaded to a repository just for the purposes of satisfying publishing requirements, but not being maintained in a repository because the authors found it advantageous to do so. Using a standard repository for lodging a paper's supporting code helps other scientists access the code easily, but not using the repository for developing and maintaining the code means the author of the paper misses out on many helpful features of repositories, such as version control, open source development and review, actions and other approaches for automating \RAPstarp\ (see below, section \ref{RAP-section}), and so on --- depending on the repository features available.

\begin{change}
\subsection{\RAPstar s: Generalized reproducible \\ analytical pipelines}\label{RAP-section}
Writing a paper typically starts in a word processor (such as Microsoft Word), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on. The process will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.

The simplified diagram in figure \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then edited into the paper.

\begin{figure*}[t]
\begin{center}
\def\drop#1{\setbox0=\hbox{\lower .5em\hbox{#1}}%
\ht0=0em
\dp0=0em
\copy0}
\def\lowarrow{\drop{$\rightarrow$}}

\def\bigBracket{$
	\setbox0=\hbox{\lower 1.2ex\hbox{$\left.\vbox to 4em{\vfill}\right\}$}}
	\ht0=0em \dp0=0em 
	~\copy0
$}
\begin{change}
\begin{tabular}{|lcl@{}cl@{\hskip 2em}cl|} \hline
\drop{\bf Data sources}&\lowarrow&\bf \drop{Analysis} & \lowarrow&\bf Select results& \lowarrow& \bf Submit for\\
\bf &&\bf & &\bf for write up & & \bf publication \\ \hline \hline 
&&&&&&\\
Experiments&& Hand calculations &&&&\\
Standard data&& Packages  && Copy \& paste &&\\
Search engines&& SPSS etc & \bigBracket & and edit data && Final paper\\
Literature &&  Graphics packages  && (text, images, graphs, etc)&&\\
Sensors && Specially-written code && into paper &&\\ 
\vdots && \vdots &&&& \\ &&&&&&\\ \hline
\end{tabular}\end{change}
\end{center}\caption{\begin{change}A simplified schematic of the publication pipeline. For clarity, the pipeline has been linearized; in general, there will be much cyclic iteration and refinement. The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline processes so that they can be run automatically, and hence reproduce the results that underpin the final paper. The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.\end{change}}
\label{fig-pipeline}
\end{figure*}

For simplicity, the schematic pipeline in figure \ref{fig-pipeline} omits showing many standard steps in the creative scientific process: each step is iterated and modified as the research progresses, and as referees require revisions. The point illustrated, however, is that in typical scientific practice each arrowed step in the diagram leading to a published paper is largely or entirely manual, typically selecting and copying output from the previous phase, and then pasting the results into the next. The pipeline steps of data $\rightarrow\cdots\rightarrow$ paper is then repeatedly run by hand as the various components are refined and improved until the authors are happy with the paper.

Different data is selected; calculations and analyses are modified; programs are debugged. As problems are detected in the paper, the data, calculations and programs are reviewed and refined. The process is rarely systematic, and even less likely to be documented --- after all, the atomic steps are trivial copy and paste actions. The final paper and the ideas it embodies are all that matters.

The basic insight of the reproducible analytic pipelines (RAP) proponents is that every time any atomic step in the pipeline process is performed it could have been automated \cite{rap,goldacre,turing-way}. If automated, it could then be repeated reliably --- unlike a manual cut and paste which is error-prone (in different ways!)\ every time it is done. In particular, when a process is automated, any other researcher, whether part of the authorship team or a later reader of the paper, can reproduce it reliably. It can be repeated if any experimental data, literature, or other knowledge changes, and the paper's analysis brought up to date with ease.

For example, if the paper in question is a systematic review, in principle it could be kept current by automatically re-running the programmed atomic actions that it was built with in the first place. Indeed, this ability is one of the original motivations of RAP: Government agencies can produce up to date reports on request without having to repeat all the manual work and risk making procedural errors in doing so. Each time they do so, the RAP pipeline is reviewed and improved, so the quality of the reproduced work improves --- unlike in a non-RAP process where new errors are generally introduced.

RAP embodies Donald Knuth's comment,

\begin{quote}\raggedright
{\setbox0=\hbox{``}
\hskip -\wd0\box0
Science is what we understand well enough to explain to a computer'' }
\\ \hfill from the foreword to $A=B$ \cite{a=b}
\end{quote}

The corollary is that if we are doing cut and paste that is arbitrary and cannot be programmed into a computer, then we are not doing science. Science is an algorithmic process, and therefore, as Knuth says, if we understand well enough what we are doing in science, we can explain it as programs, as code, for a computer to automate. That is RAP in a nutshell.

For example, in the present paper, we analyzed \the\dataN\ papers with \the\countAuthors\ authors, and one of the papers used a program that was composed out of \covidsimfiles\ files and had over \covidsimkLOC\ thousand lines of code. In the ``old days'' these numbers would have been manually worked out, then read and typed up by one or more of the authors. This is a potential source of error. As the analysis is extended the numbers may well change, and the authors would have to stay aware that a number like \the\dataN\ will need checking and updating over the period the paper is being written. It is an error-prone process, and has to be regularly repeated. Worse, other researchers may have no idea how the numbers were generated --- the paper is not fully reproducible. (More generally than simple numbers, as illustrated in this paragraph, papers may also have tables, diagrams, plots, and other types of result generated during the research.) Instead, in this paper, all those numbers (and many more) were computed automatically and change automatically if the data changes, and they were then inserted into the text of this typeset paper automatically. The figures are very probably correct.

Once processes in the pipeline are automated, this means that there is code to can run those steps again. Once there is code, it can be managed in a version control system. A version control system then provides an audit trail for free, as well as many advantages such as being able to backtrack to an earlier version to undo now-unwanted edits. Importantly, the automatic code can perform sanity checks on the process --- a very simple example is automatic bibliography systems that check that journal names, DOIs are correct and that references are correctly numbered, and so forth. They also allow the bibliographic data to be pooled and curated with other scientists, which improves its scope and quality.

Many systems provide tools to do this. GitHub (which is mentioned throughout this paper) provides \emph{actions\/}, which are named specifications that run workflows. GitHub happens to specify actions in the language Yaml, which, being a textual notation, in turn means that all the helpful features of GitHub --- open source, version control, etc --- can be applied to these pipelined processes as well. Helpful pipelines can be documented, shared and improved with open collaboration.

In the limit, almost the entire scientific process can be automated (and its interaction with the world automated with robots). There are many ways to do this; for example, the mathematical programming system \emph{Mathematica\/} makes the analysis of the data and the calculations and the paper ``the same thing'' in its integrated notebook user interface --- which behaves a bit like Microsoft Word, except that formulas can be evaluated and plotted, etc, with ease. Alternatives include R Markdown, an approach based in the open mathematical system R; many variations of literate programming \cite{relit}; a system, Lepton  \cite{lepton}, that allows a \LaTeX\ document to execute and include arbitrary code, and language-independent notebook systems like JupyterLab; and so on.

In all such systems, rerunning calculations re-creates the paper. Once \emph{Mathematica\/} (or which other system is used) has been set up, there is no repeated tinkering and error-prone copy and paste. Indeed, every time the system is run, the authors are likely to double-check the results after they have been re-computed --- so the RAP process actively helps reduce errors. The benefits and ideas of \RAPstar\ are starting to be more widely recognized, if not coherently integrated under that name. For example, the Executable Paper Grand Challenge Workshop \cite{Executable-Paper} explored the benefits of running scientific papers \emph{as\/} programs.

But a final paper is not the only product scientific authors create and publish. In computational modeling, they also often create the code and data that generates results for their paper, and both are often made available. Here is a critical insight: the code, just like the paper, contains text (source code and documentation) and data (e.g., constants) which have been copied and pasted from elsewhere. In more sophisticated RAPs, then, the coding process itself is fully part of the pipeline.

Software engineers have many tools for automatic code development (such as Unix's \texttt{make}) but the idea that these tools can be used to integrate and help automate code authoring as well as its documentation \emph{and\/} paper authoring is radical. Amongst other things, it means that the \emph{entire\/} research and development process of the paper (as well as all its underlying code) can be reproduced by others. Since this view has not been emphasized previously, we shall call it \RAPstar\ --- and we hope people will ask what that term means. The present paper is a modest example of \RAPstarp, more details of which are described in the \supplement.

Note that as \RAPstar\ objectifies how science is done to a standard sufficient to enable a computer to run, it enables all of the methodologies of Software Engineering to be brought to bear. \RAPstar\ means the normally tacit, manual, and undocumented processes of science have become explicit code. Code can  be scrutinized, optimized, and ensured correct by standard Software Engineering practice --- thus \RAPstar\ does not just automate science for reproduction, it makes the automation explicit so doing science can be reasoned about --- mentally and automatically with sophisticated tools, theorem proving and AI and so on --- and hence science will be improved.
\end{change}

\section{A call to action}\label{summary}
Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide.

Software used for informing public health policy, medical research or other medical applications is \emph{critical software}. Professional critical software development, as used in aviation and the nuclear power industry, is (put briefly) based on \emph{correct by construction}: \cite{cbc} effectively, design it right first time, supported by numerous rigorous techniques, such as Formal Methods, to manage error. (See extensive discussion in this paper's \supplement.) Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable. Conversely, not following these practices undermines the rigor of the science.

An analogous situation arises in ethics. There are some sorts of research that are ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work. Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems. National funders, and others, therefore require Ethics Boards to formally review ethical quality. Medical journals will not publish research that has not undergone appropriate ethical review.

Analogously, and supplementing Ethics Boards, Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality Software Engineering. Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to epidemic modeling papers --- and in due course, all scientific papers --- being required to include Software Engineering Board clearance statements as appropriate. {Software Engineers have a code of ethics that applies to \emph{their\/} work in epidemic modeling \cite{ethics-code}.}

{The present paper did not explore data, because in almost all cases the code and data were explained so poorly and archived so haphazardly it would be impossible to know whether the author's intentions were being followed.\footnote{{For the present paper, all the code, data, analysis and documents are available for download in a single zip file.}} Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data. Ideally, the authors should test a clean deployed build of their code and save the results. Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims. This requirement is no more unreasonable than requesting good statistics, as discussed in the opening of the paper. And the solution is the same: relevant experts --- whether statisticians or Software Engineers --- need to be routinely available and engaged with the science. SEBs would be a straight forward way of helping achieve this.}

There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university. Active, professional Software Engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art. There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.

Open Source generally improves the quality of software. SEBs will take account of the fact that open source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers. Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis {and other quality control methods for which the lead scientist and Software Engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).}\ {Software Engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer reviewed publications \cite{acm-artifacts}.}\ {See this paper's \supplement\ for further suggestions.}

A potential argument against SEBs is that they may become onerous, onerous to run and onerous to comply with their requirements. A more mature view is that SEBs need their processes to be adaptable and proportionate. If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation. Hence SEBs processes are likely to follow a risk analysis, perhaps starting with a simple checklist. {There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar. Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.}

A professional organization, such as the UK Royal Academy of Engineering ideally working in collaboration with other national international bodies such as IFIP, should be asked to develop and support a framework for SEBs. SEBs could be quickly established to provide direct access to mature Software Engineering expertise for both researchers and for journals seeking competent peer reviewers. In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations. Given the urgency, this paper recommends that \emph{ad hoc\/} SEBs should be established for this purpose.

SEBs are a new suggestion, providing a supportive, collaborative process. Methodological suggestions already made in the literature include open source and specific Software Engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}. While \cite{ABCs-SE} provides an insightful framework to conceptualize approaches and compare their merits, there is clearly scope for much further research to provide an evidence base to motivate and assess appropriate interventions to help scientists do more rigorous and effective Software Engineering to support their research and publishing. These and further issues are discussed at greater length in the \supplement.

\subsection{Action must be mutual \\ and interdisciplinary}
\setcounter{footnote}{1}
Code is only part of science, and only one critical factor in the wider reproducibility crisis: SEBs must work with --- and be engaged by --- other reproducibility initiatives.

Relying on Software Engineering Boards (SEBs) \emph{alone\/} would continue one of the current besetting problems about the role of code. The conventional view is that scientists do the hard work compared to the ``easy'' coding work (see sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}) so they just need to tell programmers what to do. This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, which he explores at some length, is that we need to spend more effort in working out what computers should do (i.e., do the science better) and then just tell programmers to do \emph{that}.

On the contrary, competent Software Engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and often that logic needs refining or optimizing. In other words, the Software Engineers can bring important insights back into the science, hence improving or changing the questions and assumptions the science relies on. This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but that is ill-conditioned, there is no good answer.'' Ideally, then, it is not just a sequential process of science $\rightarrow$ code $\rightarrow$ results, but an iterative cycle of collaboration and growing mutual understanding.

In short, the way SEBs work and are used is crucial to their success. Software engineers can help improve the science, so it is not just a matter of asking a SEB  whether some coding practices (like documentation) are satisfactory, but whether the SEB has insights into the science itself too. Most effectively, this requires interdisciplinary working practices (science plus Software Engineering) with mutual respect for their contributing expertise.

\subsection{Methodological statements}
Most journals (using the word ``journals'' as a shorthand for all publishing outlets, conferences, workshops, videos, books, etc) and funders require explicit statements how the authors have conformed to appropriate methodological standards covering issues, such as conflicts of interest, ethics, data access, consent, and so forth. It would be easy for journals and funders to require equivalent types of statements on Software Engineering.

Studies of data access statements show that they are unreliable: some authors withdraw papers when journals request access statements \cite{no-raw-data} (indicating that journals that do not make an explicit request are likely publishing papers that will not provide access), and some authors do not respond to access requests \cite{data-access}. How we help scientists who do not want to provide data or code access is one problem, but the serious issue for science is how to ensure any statements made are accurate, and any access provided is actually helpful (e.g., well-defined, versioned, etc) to support useful reproduction. Journals and funders should provide support for hosting data and code (and any other relevant data, such as qualitative data, video, etc), and the review process must check that authors actually provide the material as they claim in the methodological and access statements. Conversely, scientists should be able to access funding to ensure data and code access, and, as appropriate, funding for on-going maintenance of the databases, which will typically require funding beyond the end of the normal funding period.

Intellectual Property (IP) is an increasing concern, both for author scientists and their sponsors who want royalties, and for other scientists wishing to freely build on the published science. Particularly concerning access to code, IP is potentially and often is in conflict with scientific openness. Methodological statements should be made concerning any IP associated with code, and to what extent this interferes with open access to code. More routine discussion should include raising known system dependencies, such as operating system, compiler, or special hardware dependencies; it is also appropriate to mention standards conformance, such as to IEEE Floating-Point Arithmetic (IEEE~754).

Journal policies could start to explicitly encourage computationally reproducible science using RAP and \RAPstar\ techniques. That is, the research's methodology itself may be a mixture of data and code. As this paper's \supplement\ shows in section \ref{restrictive-policies}, many journals (e.g., \emph{PLOS ONE\/} and, ironically, \emph{IEEE Transactions on Software Engineering\/}) and repositories have policies that make RAP harder. Methodological statements should be required that make clear what access rights are available for RAP or \RAPstar\ material, as it is much more likely to raise IP issues that normal disclosures. In particular, if the authors plan on publishing a series of papers based on the same methodologies, the RAP/\RAPstar\ access might be provided in a later paper or held under escrow by the journal (or funding body).

Journals and funders often require data and code access statements, but as this paper has made clear, code is complex and it is rarely easy to understand and scrutinize even with access to substantial documentation (which is unusual). It is therefore recommended that journals and funders require \emph{assurance arguments} \cite{assurance-case}, a familiar technique particularly from the safety assurance domain. Assurance arguments provide a concise, high-level argument that the system does what is claimed. Assurance arguments can be more or less detailed, and more or less formal in their approach; we envisage referees would have views on the level of detail and formality required for any specific contribution.

Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access statements apply to all information, regardless of any arbitrary classification of it as code or data (or documentation, assurance case, etc).

\begin{change}
\subsection{The paper as a scientific laboratory}
RAP (see section \ref{RAP-section} and figure \ref{fig-pipeline} ) recognizes that much of doing science is, to the extent that it is reproducible, a proceduralized, algorithmically-framed process: many scientific processes, most obviously in computational science, can be automated and reproduced at will (at least when they are done to high enough Software Engineering standards). \RAPstar\ further recognizes that the codification of science is itself a scientific object. That is, the code recording the scientific pipeline is not just scaffolding to support a scientific project, but is in itself a scientific product. It can be criticized, debugged, optimized, refactored, generalized into a virtual machine --- code --- that can do any related type of science. Written well, the code is an explanation of how particular science is done.

In other words, as a paper is developed the computational contributions do not just generate results, but they objectify the scientific processes that have generated the results: they make doing science a first class object that can be seen, scrutinized and thought about --- it is code, after all --- generalized and applied more widely. In computer science terms, the paper (paper, data, code, etc) becomes a first-class object; that is, it has all the rights and operations as available to any other object. In scientific terms, then, it can be controlled, measured, stored, mixed, split, searched, analyzed, written-up, and so forth as an object of proper scientific enquiry. This is trivially true in the field of Computer Science, since (for many papers) the object of study \emph{is\/} a computer program --- the only debate is how much of that code is literally in the paper (or downloadable from it) or just talked about in the paper.

Code thus becomes a tool reifying scientific insights, which therefore (if of adequate quality) itself merits publication, so others can reproduce, generalize and progress the science confidently themselves. The scientific paper turns from being a rhetorical record of some scientific investigation into also embodying the algorithms that performed the science. \RAPstar\ makes the algorithms explicit. The paper becomes the laboratory where the code is developed, tested and reasoned about. As the algorithms in code are explicit, science gains leverage and will advance more rapidly.

One of the many exciting opportunities is transforming how scientists code, how they turn computation from mere support into a keystone of thinking about their activities.

Today, computational science papers typically have some mathematical equations to state and communicate their theories, but the results are generated by embarrassingly complex imperative programs that defy scrutiny, typically written in languages like C, Python, or Java. Increasingly, though, computational science is building on programming languages that can directly support science. In a language like \emph{Mathematica}, a scientist can program directly in (say) systems of differential equations, and there need be no difference at all between how ideas in the paper and the ideas in the code are expressed. In fact, the paper has become the scientist's laboratory, the place where they do research.

The more we view the paper as a scientific laboratory, the more we gain from the \RAPstar\ perspective and the more science will gain from improved, conceptually broadened computational science. The more we will want to mature Software Engineering standards too, because the quality of future science depends on them.

\subsection{Benefits beyond computational science}
Computational sciences recognize the key role of computation (if not sufficiently emphasizing the role of code), but many fields do not recognize computation as such and therefore they are missing out on the leverage that comes with recognizing computation as a first class player in their scientific activities.

One example will suffice. A lot of healthcare is supported by computers, yet medical papers remain traditional and only very exceptionally refer to or include code. However, much clinical practice relies on computer analysis, but inevitably it has to use code unrelated to the code developed by the scientists doing the primary research, as their code is neither peer reviewed nor published. Code is professionally invisible, has no prestige, and there is therefore no investment in studying or improving it. Conversely, the challenging critical issues (including patient safety) that require clinical code to be more reliable than scientific tinkering do not get evaluated by researchers. Certainly, the issues, if noticed,  would stimulate further science --- sexism and racism in healthcare algorithms being a case in point \cite{fixit}.

Without taking the lessons of improved computational science to other fields, like medicine, there will continue to be an unfortunate and unnecessary disconnect between research and practice, and one where both science and practice suffer because code is not recognized as a contribution to science \cite{fixit}.

\input models-data-fitting/generated-modelsSection.tex

%\subsection{Why improving \\ computational sciences matters}\label{over-fit}
%Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that. We do the science, and we are in the world, so there are challenging meta-theoretical issues as well as practical experimental and resourcing issues. Karl Popper is one of many philosophers of science to develop principles and theories about science \cite{popper-three-worlds,popper-conjectures-refutations}. Arguably first noticed by Herbert Simon \cite{sciences-artificial}, the ``sciences of the artificial'' now raise new issues, as does more recent non-reviewed mass publication of results (including publication of results based on non-reviewed computational models), for instance in archival systems such as \url{arxiv.org}
%
%Science to date has been inevitably driven by human \ae sthetics: that is, simple, elegant, understandable, beautiful, brief, exciting, intelligible, are all human properties that are applied to doing, funding, defining, and selecting ``good'' science.
%
%Computation is starting to undermine such human-oriented criteria, for what are a good criteria of a simple theory when a computer model can generate a human-understandable interpretations of \emph{anything\/} including arbitrarily complex, buggy, even fictional theories? The danger is that superficially successful computer programs may unintentionally over-fit phenomena, so apparently confirm our understanding of them. The code we use, being inscrutable, may come to model our desires better than model the data. The classic examples of over-fitting are visually-obvious complex Real$\rightarrow$Real functions of one variable (e.g., as visualized in a simple example shown in figure \ref{fig-overfit}), but when the models are implemented as huge computer programs with gigabytes or more of parameters, there is no comprehensible and accurate representation for them. In general, we do not (sometimes, cannot) see their complexity and arbitrariness,\footnote{\input generated-fittingModels-data.tex
%These models were written without relying on mathematical packages so the code is explicit, hence making it more reasonable to make a direct comparison. Indeed, the code for models looks superficially similar, emphasizing that (\emph{i\/}) simple variations, changes or mistakes in code can have dramatic effects on results, and  (\emph{ii\/}) visualizing the results of running code --- as shown in figure \ref{fig-overfit} --- can help identify bugs (if any).} and attempts to improve (``refactor'') them would only introduce further obscurities.
%
%Until we improve our standards for the computational sciences --- this paper argues by starting by intentionally engaging competent Software Engineering (and also updating the under-pinning philosophy of science, perhaps beginning with a Popper World~0) --- we risk doing and promoting bad science because we are not managing the unleashed complexity of the codes that our science relies on.
%
%\begin{figure}[t]
%{ \begin{change}  \centering
%\newdimen \tw \tw=\textwidth \advance \tw by -\columnsep \divide \tw by 2
%   \includegraphics[width=\tw]{plot.pdf} 
%\end{change} }  
%\input generated-modelsAt5.tex
%\caption{%
%Much computational science is concerned with finding plausible multi-dimensional models that fit data. Shown here is notional experimental 2D data (the dots), together with a linear regression and an exact polynomial model. The over-fitted polynomial model fits the sample exactly, but since the experimental data is presumably subject to error (indicated by the confidence interval, itself estimated from another model) the linear model would generally be considered a better description of the experimental results. If we assume that the error in measuring $y$ is normally distributed (or similar), the extreme values the over-fitted model predicts would be considered unlikely; for example, \exacty\ predicted by the exact model versus \lineary\ predicted by the linear model.}
%\label{fig-overfit}
%\end{figure}

\ignore{\subsection{Suggestions for further work}
Although this study of Software Engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope. However, the insights appear to be general.

Further work to extend the scope of the survey beyond the basic requirements of the present paper is of course worthwhile, but the following cases (listed in the next few paragraphs) suggest that the problem is widespread. We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor. }
\end{change}

\section{Conclusions}
We need to improve the quality of code and Software Engineering that supports science. While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed shows that current coding practice makes for poor science in many fields. In a pandemic, scientific modeling, such as epidemiological modeling, track and trace \cite{excel-fiasco}, modeling COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life. Unfortunately, Software Engineering good practice, to help \emph{do\/} science has been absent to date.

The main challenges to mature computational scientific research are:

\enum=1
\begin{enumerate}\raggedright
\item 
To manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices that papers rely on. Computer code should be explicit, accessible (well-structured, etc), and properly documented. Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}. Professional software methodologies should not be ignored.

\item 
To use computation to help make scientific processes explicit, so that they can be reproduced, scrutinized and improved. RAP is an increasingly popular way to help do some of this, but as this paper points out, RAP should be generalized to \RAPstar\ to help the computational parts of science as well, leading to a virtuous circle.

\item
To support and develop the scientific community in the professional use of computation.
\end{enumerate}

While programming seems easy, and is often taken for granted and done casually, programming \emph{well\/} is very difficult \cite{fixit}. We know from software research than ordinary programming is very buggy and unreliable. Without adequately specified and documented code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect. Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion: computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper-conjectures-refutations}.

Software Engineering Boards (as introduced in this paper) are a straight forward, constructive and practical way to support and improve computer-based science. This paper's \supplement\ summarizes the relevant professional Software Engineering practice that Software Engineering Boards would use, including discussing how and why Software Engineering helps improve code reliability, dependability, and quality.

\makeatletter
\immediate\write\@auxout{\string\def\string\continueSectionNumbers{\arabic{section}}}
\makeatother

\section*{Supporting information}
\newcount\csrefcount \csrefcount=0
\def\csref{\global\advance\csrefcount by 1}
\long\def\ethics#1{\paragraph*{Ethics}#1}
\long\def\ack#1{\paragraph*{Acknowledgments}#1}
\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
\long\def\competing#1{\paragraph*{Competing interests}#1}
\long\def\funding#1{\paragraph*{Funding}#1}

\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson provided very helpful comments.}
%\the\csrefcount\ computer science reviewers.

%\ethics{This article presents research with ethical considerations but does not fall within the usual scope of ethics policies.}

\dataaccess{There is an extended discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}. The \supplement\ also presents all raw data in tabular form. All material is also available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards},\footnote{\color{red}This is a temporary URL before meeting publication repository requirements for accepted papers.\color{black}} which has been tested in a clean build, etc.

The data is encoded in JSON\@. JavaScript code, conveniently in the same file as the JSON data, checks (with \the\JSONerrorCount\ possible classes of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset all results reliably in this paper and in its \supplement\ \emph{directly\/} from the automatic data analysis.

In addition, a standard CSV file is generated from the JSON in case this is more convenient, for instance to browse directly in Excel.}

\aucontribute{Harold Thimbleby is the sole author. An preliminary outline of this paper, containing no supplementary material or data, was submitted to the UK Parliamentary Science and Technology Select Committee's inquiry into UK Science, Research and Technology Capability and Influence in Global Disease Outbreaks, under reference LAS905222, 7 April, 2020. The evidence, which was not peer reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper. It is available to the public \cite{parliamentary-evidence,my-parliamentary-evidence}.}

\competing{The author declares no competing interests.}

\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186], and by Assuring Autonomy International Programme, Assuring Safe AI in Ambulance Service Triage. The funders had no involvement in the research or in this paper.}

%\nolinenumbers

{\raggedright

\bibliographystyle{mysiam}
\def\urlprefix{}
\initialiseBibliography{References}{1}{}

\newcount \bibitemn \bibitemn = 0
\let\oldbibitem=\bibitem
\def\bibitem#1{\ifnum \bibitemn = 0 \else \egroup\vskip .7ex plus 1ex \fi
\advance \bibitemn by 1
\vbox \bgroup \oldbibitem{#1} }
\def\endofbibliography{ \egroup \vfill}
\bibliography{paper-seb-main.bib}

% Note that the supplementary material calculates how many references there are,
% because it reads in this .tex file's .aux file so that it can cite the references 
% above with the same number sequence. 
%
% It simple redefines \bibitem to work it out while it
% defines all the citations from the bibliography above.
%
% The same technique allows the supplementary material to have 2 more bibliographies,
% with consecutive numbering.
}

%\label{LastPage}
%save Latex register    {linenumber}
%\makeatletter
%\immediate\write\@auxout{\string\def\string\MainPaperLastLine{\arabic{linenumber}}}
%\makeatother
\end{document}

\def \allInOne {define this so we don't include header in supplemental material}
\input paper-seb-supplementary-material.tex
\end{document}

%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}

\end{document}

### 2023-01-31 response-to-referees/paper-seb-supplementary-material (original).tex
More

--- 2023-01-31 paper-seb-supplementary-material.tex
@@
     \usepackage{fancyhdr}
-    \usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
     \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
@@
 
-\input generated/page-lengths.tex
+\input generated-page-lengths.tex
 
@@
 
-% pageref gives us a hyperlink which looks unsightly, so we do it this way...
+% pageref gives us a hyperlink which looks unsightly
 \makeatletter
@@
 \newwrite\infofileForMain
-\immediate\openout\infofileForMain=generated/info-for-main.tex
-
-%\makeatletter
-%% saves name-value pairs in both the supplementary material .aux file and the info for the main file
-%\def\remembernamevalue#1#2{%
-%	\immediate\write\@auxout{\@backslashchar gdef\@backslashchar #1{#2}}%
-%	\immediate\write\infofileForMain{\@backslashchar gdef\@backslashchar #1{#2}}%
-%}
-%\makeatother
-
-\input generated/info-for-supplement.tex
+\immediate\openout\infofileForMain=generated-info-for-main.tex
+
+\input generated-info-for-supplement.tex
 \newwrite\infofileForSupplement
-\immediate\openout\infofileForSupplement=generated/info-for-supplement.tex
+\immediate\openout\infofileForSupplement=generated-info-for-supplement.tex
 
@@
 \def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalLabel: #1}%
+\typeout{Global xref: #1}%
 \immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
@@
 \immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalTableLabel: #1}%
-\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalLabelValue: #1 := #2}%
-\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
 }
@@
 \noindent\small
-\input generated/supplementary.toc
+\input generated-supplementary.toc
 }
@@
 \newwrite\contentsfile
-\immediate\openout\contentsfile=generated/supplementary.toc
+\immediate\openout\contentsfile=generated-supplementary.toc
 
@@
 \vbox{\noindent
-      \input generated/minicontents.tex
+      \input generated-minicontents.tex
       }
@@
 \newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated/minicontents.tex
+\immediate\openout\minicontentsfile=generated-minicontents.tex
 
@@
 
-{The type and use of mathematics used in Software Engineering is called \textbf{formal methods}.
+{The type and use of mathematics used in Software Engineering is formal methods.
 Not using formal methods ensures the resulting code is unsafe and unreliable.
@@
 
-It is important to be clear that formal methods is a spectrum, from doing it as rigorously and comprehensively as possible using state of the art methods, to applying a basic ``formal methods mindset'' that any competent programmer could do.
-
-\begin{itemize}\raggedright\item An interesting paper explores a formal approach to coding differential equations \cite{differential-equations}, but it makes it clear that their approach is beyond most programmers.
-Such formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK \cite{spark}, and \emph{many\/} others).
+Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
 Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
@@
 
-\item At the other end of the spectrum, a formal methods mindset means being clear about basic mathematical properties of code, such as by using assertions, proving invariants in loops, and so on.
-This approach should be feasible --- and perhaps required --- for all scientific programmers.
-\end{itemize}
-
-Crucially, computer tools are available to catch common human errors that we are all prone to.
+Crucially, tools are designed to catch common human errors that we are all prone to.
 Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort \cite{misra}.
@@
 In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-
 Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of Software Engineering has advanced and become more specialized and sophisticated over the decades since.
@@
 C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review of bad programming languages, for which see \cite{heedless}, but Excel is even worse than C\@.)
-
-In particular, C is not a portable language, which means C code will work differently on different types of computer and operating system.
-
-SPARK Ada is a popular example of a much more appropriate high integrity programming language to use \cite{spark}.
+(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
+SPARK Ada is one example of a much more appropriate programming language to use.
 SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-Other high integrity languages include  OCaml, F*, and Haskell; reference \cite{haskell} is an excellent introduction to Haskell, and introduces the wider issues of reliable programming in such languages.
 
@@
 When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using \emph{Mathematica\/} --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
+Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
 Reproducing the same results relying on the same codebase tells you little.
@@
 Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and as there are.are serious dependability problems with Excel (illustrated particularly in section \ref{benefits-beyond-science} of the main paper), open source projects competently implemented (e.g., avoiding or carefully managing the use of Excel) would help these people enormously.
+Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
 
@@
 
-\mysection{Code, data, and publication}
+\mysection{Code, data and publication}
 \globalLabel{on-code-data-publication}
@@
 The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Some programming languages like \emph{Mathematica\/} treat $\pi$ as an exact mathematical constant (e.g., \emph{Mathematica\/} calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
+Some programming languages like Mathematica treat $\pi$ as an exact mathematical constant (e.g., Mathematica calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
 \url{www.fhgr.ch/en/specialist-areas/applied-future-technologies/davis-centre/pi-challenge}}
@@
 
-Section \ref{over-fit} of the main paper looks like part of an ordinary paper, but it (including the figure and calculations) was data generated by a \emph{Mathematica\/} program.
+Section \ref{over-fit} of the main paper looks like part of an ordinary paper but in fact it was generated by a \emph{Mathematica\/} program.
+The section looks like typical text from a paper, but it was data generated by a program.
 
@@
 The idea is very simple, but very effective.
-The ``code'' for the paper includes the \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
+The ``code'' for this paper includes the \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
 In fact, the notebook is written as a self-contained report or paper, as \emph{Mathematica\/} notebooks generally are, and it thoroughly documents how it works.
@@
 
-\mysubsection{Data and polynomials used in the paper}
-The data and polynomials used in the main paper's section \ref{over-fit}, and illustrated in the paper's figure \ref{fig-overfit}, is presented below, as generated in \LaTeX\ by the same \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
-Of course, in the paper itself, the data was not necessary to make the point, but if there is any need to replicate it or otherwise scrutinize the argument in the paper --- as there would be for more complex arguments in typical papers --- the data used and results are shown here.
-In a similar way, data and computed results could be presented and made available in other scientific papers.
-
-\input generated/over-fitting-data.tex
-
-\highlightmysubsection{Comparing conventional and RAP approaches}
-\globalLabel{RAP-diagrams}
-The similarities and differences between the conventional copy-and-paste approach to filling in data and diagrams in publications, the improved systematic RAP process, and using notebooks (such as \emph{Mathematica\/} or Jupyter), are illustrated in the sequence of schematics of figures \ref{fig:basic}, \ref{fig:excel}, \ref{fig:rap}, and \ref{fig:notebook}.
-
-\begin{figure} 
-   \centering
-   \includegraphics[width=5in]{generated/basic.pdf} 
-   \caption{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
-Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code.}
-   \label{fig:basic}
-\end{figure}
-\begin{figure} 
-   \centering
-   \includegraphics[width=5in]{generated/excel.pdf} 
-   \caption{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
-The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
-Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
-In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication.
-}
-   \globalLabelValue{excelFigure}{\thefigure}
-   \label{fig:excel}
-\end{figure}
-\begin{figure} 
-   \centering
-   \includegraphics[width=5in]{generated/import.pdf} 
-   \caption{Improving over the normal approach (figure \ref{fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
-In the present paper, this was done by generating text files of \LaTeX\ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX\ names for all of the code-generated values.
-Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
-Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen.}
-   \label{fig:rap}
-\end{figure}
-\begin{figure}[t]
-   \centering
-   \includegraphics[width=5in]{generated/notebook.pdf} 
-   \caption{In a notebook system, such as \emph{Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
-The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
-Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
-In well-engineered code, the notebook will directly contain full documentation.
-Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc).}
-   \label{fig:notebook}
-\end{figure}
-
 \changemysection{The Spiegelhalter trustworthiness questions}
@@
 David Spiegelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Spiegelhalter} Spiegelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- and the claims need to be properly accountable.
-
+In his \emph{The Art of Statistics\/} \cite{Spiegelhalter} Spiegelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
 Spiegelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or publications relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask these questions is as risky as dismissing statistical scrutiny.
+Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
+
+What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
 
@@
 
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code, and just assume it is running correctly.
+\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
 So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
@@
 
-The main paper was motivated by selected epidemiological papers and their problematic computational models (which are discussed in the paper).
-Are the issues illustrated by these case studies typical of science more broadly?
-We undertook, then, some selected studies of papers in different fields, then undertook a randomized, stratified pilot survey covering several leading peer reviewed journals.
-The point was not to establish the frequency of problems, so much as to sign whether the problematic case study was exceptional or typical.
+The main paper focuses on selected epidemiological papers and their computational models.
+Are the issues illustrated by these case studies typical of computational science more broadly?
+We undertook, first, some selected studies of papers in different fields, then undertook a randomized, stratified pilot survey covering several leading peer reviewed journals.
 
@@
 The pilot study follows the RAP methodology.
-The data and code for this paper's pilot survey (and all other analysis used in the paper and in this \supplement) are available on GitHub.
+The data and code for this paper's pilot survey (and all other analysis used in the paper and in this \supplement) are available on Github.
 All raw data is converted into \LaTeX\ so that the analysis can be typeset directly in the paper; this \supplement\ also contains a complete tabular presentation of the data.
 
-Moreover, this paper itself follows the more general \RAPstar\ methodology.
+Moreover, this paper follows the more general \RAPstar\ methodology.
 For example, as standard practice, a Unix makefile is used to make it easy to analyze and generate all data, typeset the paper, and more.
-Table \ref{table:options} shows the options provided.
-
-\begin{table}[t]
-\begin{center}
-\input generated/make-help.tex
+The following options are provided:
+
+\vskip 1ex{\small \begin{center}
+\input generated.make-help.tex
 \end{center}
-\caption{Conforming to the \RAPstar\ methodology, the abbreviated summary above was generated automatically, by using \texttt{make data}.
-%(Saying \texttt{make help} would generate a full summary on the command line.)
-}
-\label{table:options}
-\end{table}
-
-The \RAPstar\ approach cannot enforce the truth of such as summary, but it very significantly increases the chances that the summary is correct and up to date.
+}\vskip 1ex
+
+Conforming to the \RAPstar\ methodology, the summary above (including all the \LaTeX\ typesetting details) was generated automatically (in fact, by just saying ``\texttt{make data}'', which also generated all the other data).
+However, when \texttt{make help} or \texttt{make help-brief} are run, they generate plain text summaries without any of the \LaTeX\ typesetting details used here (e.g., \LaTeX\ will appear simply as \texttt{Latex}).
+
+The \RAPstar\ approach cannot enforce the truth of the summary, but it very significantly increases the chances that the option summary is correct and up to date.
 For example, if refactoring leads to an option being deleted, then it will also disappear \emph{with no further work\/} from the table above.
@@
 %
-\caption{Overview of the peer-reviewed paper sample.}
+\caption{Overview of peer-reviewed paper sample.}
 \label{supplement-table-overview}
@@
 \begin{center}\normalsize
-\input generated/summary-table.tex
+\input generated-summary-table.tex
 \end{center}
@@
 \newcount \vv
-\newcount \ww
 \def\makeAverage#1#2{%
@@
 	\vv=#2
-	\multiply \uu by 1000
 	\divide \uu by \vv
-	% uu is now 1000 times the answer we want
-	\ww = \uu
-	\divide \uu by 1000
-	\multiply \uu by 1000
-	\advance \ww by -\uu
-	\divide \uu by 1000
-	%[ #1 / #2 = \the\uu .\the\ww\ ]
-	\ifnum \ww > 499 \advance \uu by 1 \fi
 	\the\uu
@@
 
-\input generated/repos.tex
+\input generated-repos.tex
 \vskip 1ex
@@
 At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.}
-\label{table-repo-summary}\globalTableLabel{supplementaryTableRepoSummary}
+\label{table-repo-summary}
 \end{table*}
@@
 
-With the one exception, none of the papers used any specific Software Engineering methods, such as open source \cite{open-source} or other standard methodologies provided in this \supplement, to help manage their processes and help improve quality.
+With the one exception, none of the papers used any specific Software Engineering methods, such as open source \cite{open-source} or other standard methodologies provided in the \supplement, to help manage their processes and help improve quality.
 Although software stability \cite{stability} is a relatively new concept, understood as methodologies, such as portability, to provide long-term value of software, it is curious that none of the papers made any attempt at stability (however understood) despite the irony that all the papers were published in archival journals.\footnote{Reasons the present paper does not directly assess the quality of software in the surveyed papers include: many papers did not provide complete software; it was not possible to find correct versions of all software systems to run the models; also, no papers provided adequate test suites so that correct operation of software could be confirmed objectively.}
@@
 
-Overall, there was no evidence that any code had been developed carefully, let alone by using recognized professional Software Engineering methods. In particular, 
+Overall, there was no evidence that any code had been developed carefully, {let alone by using recognized professional Software Engineering methods}. In particular, 
 \ifnum \countCodetested=0
@@
 In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-
-Note that journal policies were first accessed on 29 July 2020, close after the period covered by the pilot survey, thus presumably fairly closely reflecting the policies in use for the papers in the survey on the dates when they were each submitted.
-Unfortunately for the survey, the journals do not make clear what exact policies were applied to each paper when the papers were submitted (on the other hand, the survey shows that policies are not rigorously enforced).
 \\
@@
 \url{royalsociety.org/journals/authors/author-guidelines/#data}
-\specifyurletc{Since first accessed 29 July 2020, the policy has been revised (undated, accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
+\specifyurletc{Accessed 29 July 2020; the policy has been revised (undated, but accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
 The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
@@
 \url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
-\specifyurletc{Accessed 29 July 2020; since updated (accessed 2 February 2022) to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
-Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.''
+\specifyurletc{Accessed 29 July 2020; since updated to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
+Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.'' (accessed 2 February 2022).
 }
@@
 \url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\specifyurletc{Accessed 29 July 2020. Still no code policy when accessed 2 February 2022.}
+\specifyurletc{Accessed 29 July 2020. Still no code policy (accessed 2 February 2022).}
 
 \noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
-The \emph{Journal of Vascular Surgery\/} has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant:
+Journal has detailed data policies, but no code policy. While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant:
 
@@
 \vbox{%
-  \input generated/example-data.tex
+  \input generated-example-data.tex
 }
@@
 
-The JavaScript data+program generates files from the JSON; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.
-
-For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
+The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
+If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
 
@@
 Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (as it happens, using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary} for this \supplement.
+For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary}, as used in the main paper.
 
 The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-% This commented-out sentiment is repeated later...
-% Finally, because of their approach to automatic typesetting, some journals require that no files are explicitly \texttt{input} into \LaTeX, so a simple program is provided in the GitHub repository to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.
-It is curious that the journals cannot do this for themselves.
 
@@
 The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-Note that failing a sanity check indicates a problem that needs to be fixed, but passing a sanity check does not prove a paper correct, but the more sanity checks that are passed (and the harsher the checks) the more confidence we can have that the data has been processed correctly.
-Of course, when formal methods are employed in the software development process, the confidence in correctness can be very high.
 
@@
 
-To protect against version, portability and other problems, the GitHub repository for the present paper includes a check on software versions and a checksum check for all possibly affected files, including the data file.
+To protect against version, portability and other problems, the Github repository for the present paper includes a check on software versions and a checksum check for all possibly affected files, including the data file.
 This does not solve the problem, but it ensures anyone developing or reproducing the paper's work will at least be forewarned of potential version or portability problems.
-The GitHub repository itself can be used to restore files that have been corrupted.
+The Github repository itself can be used to restore files that have been corrupted.
 
@@
 
-Another program (\texttt{programs/expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
+Another program (\texttt{expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
 Of course, the expanded version now contains all variables as fixed constants, so the submitted paper is misleading and useless to other researchers if the data is modified --- the effort to ensure all published numbers are automatically correct is defeated.
@@
 \vbox{\noindent Legend:\vskip 1ex
-\input generated/legend.tex
+\input generated-legend.tex
 }
@@
 \sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated/assessments.tex
+\input generated-assessments.tex
 \end{longtable}
@@
 
-\bibliographystyle{revisedCompJ}
+\bibliographystyle{mysiam}
 
@@
 
-\input generated/supplementary-references.tex
+\input generated-supplementary-references.tex
 

### 2023-02-20 response-to-referees/OLD deleted lambda notes.tex
Fantastic corrections noticed by Konrad Hinsen

--- 2023-01-31 response-to-referees/paper-seb-main (original).tex
@@
-\documentclass{comjnl}
-\usepackage{ragged2e}
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-%----
-  \setlength\textheight{220mm}
-%----
-
-\input paper-seb-macros.tex
-\input generated-info-for-main.tex
-\input generated-page-lengths.tex
-
-\long\def\ignore#1{}
-
-%\linenumbers
-
-\title{\mytitle} 
-\author{Harold Thimbleby}
-\shortauthors{Harold Thimbleby}
-\affiliation{See Change Fellow in Digital Health, Swansea University}
-\email{harold@thimbleby.net}
-
-%\subject{Computational science (bioinformatics,~medical~computing,~etc)}
-
-\keywords{Computational Science; Software Engineering; Reproducibility; Scientific scrutiny; Reproducible Analytic Pathway\raise 1ex\hbox{$\star$} (\italicRAPstar\hskip -1pt)}
-
-\received{30 April 2022}
-%\revised{--}
-
-%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
-%&\texttt{harold@thimbleby.net}\\
-%&ORCID 0000-0003-2222-4243
-%\end{tabular}
-
-%\orcid{0000-0003-2222-4243}
-
-\begin{document}
-
-\begin{abstract}
-\justifying
-\noindent
-While data, analysis, and the results of running scientific models (including AI and ML) are described in peer reviewed papers, the algorithms and code generating insights and results often avoid adequate scrutiny.
-If code is not correct, then the contributions of published papers relying on unchecked code will be suspect.
-
-\hskip 1em This paper shows that scientists rarely assure the structure and quality of code they rely on, rarely make full code available for wider use or scrutiny, and rarely provide adequate documentation to understand or use their code reliably.
-This paper therefore justifies and proposes ways to mature the computational sciences:
-
-%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available.
-Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects.
-Code is fallible, so any interpretation that relies on code is also fallible.
-When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized.
-In turn, scientific claims cannot be properly scrutinized.
-
-%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions.
-%The paper reports a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
-
-%Code can be improved using Software Engineering. This paper argues for specific solutions:
-
-%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.
-
-%\noindent\emph{Solutions:}
-
-\raggedright
-\newdimen \mywidth \mywidth=\textwidth
-\advance \mywidth by -3em
-
-\def\indented#1{\vskip 1ex
-\parshape 2 0em \textwidth 2em \mywidth 
-\noindent 
-\hbox to 2em{ #1.\hfill}\ignorespaces}
-
-\indented 1
-Professional Software Engineers can help and should be involved, particularly in critical research such as public health, climate, etc;
-
-\indented 2
-``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;
-
-\indented 3
-Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent Software Engineers;
-
-\indented 4
-\begin{change}The Reproducible Analytic Pathway (RAP) methodology should be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces and terms \RAPstarp.
-Furthermore \RAPstar\ should be supported and encouraged in journal, conference, and funding body policies.
-\end{change}
-
-\vskip 1ex 
-\noindent
-The \supplement\ for this paper provides a summary of professional Software Engineering best practice relevant to scientific research and publication.
-It also includes suggestions for \RAPstar\ processes, and a pilot survey of code quality in leading peer-reviewed journals that corroborates the concerns of the paper.
-
-\begin{quote}\def\seprule{\vskip 2ex\hrule\vskip 2ex}
-\seprule
-
-\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
-\zq Science is what we understand well enough to explain to a computer.''\\
-\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\
-
-\zq I have to write to discover what I am doing.''\\
-\hfill {\rm Flannery O'Connor, quoted in \emph{Write for your life\/} \cite{write-for-life}}
-
-\zq Criticism is the mother of methodology.''\\
-\hfill {\rm Robert P. Abelson in \emph{Statistics as Principled Argument\/} \cite{abelson}}
-
-\seprule
-\end{quote}
-\end{abstract}
-\maketitle
-
-%\section*{Author summary}
-%
-%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
-%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales.
-His research focuses on human error and computer system design, particularly for healthcare.
-%%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
-%Harold won the British Computer Society Wilkes Medal.
-He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder.
-His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.
-See his web site, \texttt{www.harold.thimbleby.net}, for more details.
-
-\section{Introduction}
-\begin{change}
-Unreliable, often unstated and unexplored, computational dependencies in science are widespread.
-Furthermore, code is rarely published in any useful form or professionally scrutinized, and therefore the code itself does not contribute to furthering reliable science, for instance through replication or reproduction.
-In short, the quality of much computational science (including AI and ML) is undermined because the code it relies on is rarely of adequate quality for the uses to which it is put.
-
-This paper explores the nature and extent of these Software Engineering problems in published science. The paper suggests ways to improve the reliability of computational sciences.
-
-Code should be developed and discussed in a sufficiently professional, rigorous, and recognizable way that is able to support clear presentation and scrutiny.
-Developing justifiably reliable code is the concern of the field of Software Engineering, which is discussed further below, as well as substantially in the paper's \supplement, which is an integral part of the suggested solutions and literature survey.
-
-This paper makes an explicit analogy between reliable computation and reliable statistics, a field that has adopted clear standards for reliable use and presentation.
-The \supplement\ (section~\ref{supplementary-Spiegelhalter-section}) summarizes Spiegelhater's uncontroversial statistics probes and  --- just as it is routine for statistics, code and results from code (and the data it is run on) need to be discussed and presented in a way that properly assure belief in any claims derived from using them.
-
-\begin{table*}
-\begin{center}\normalsize
-\begin{tabular}{|rl|} \hline
-\the\numberOfJournals&Journals\\
-\the\dataN&Papers:\\
-\tabularJournalBreakdown
-\the\countAuthors&Published authors\\
-\the\totalPages&Published journal pages\\
-July 2020&Sample month\\ \hline
-\end{tabular}
-\end{center}
-\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}.}
-\label{table-overview}
-\end{table*}
-
-\begin{table*}
-\begin{center}\normalsize
-\input generated-summary-table.tex
-\end{center}
-\caption{Summary of a pilot survey of computational science papers. Full details of the survey, all data, and methodology are in the \supplement.}
-\label{table-summary}
-\end{table*}
-\end{change}
-
-The body of this paper reviews some specific epidemiological computational science and its avoidable problems.
-The \supplement\ further describes a broad pilot survey of papers covering a wide range of computational science published in leading journals, confirming that the issues raised in the body of the paper are widespread and not specific to science that may have been rushed because of relevance to the COVID-19 pandemic.
-The quality of the supporting code in both epidemiological science and in the survey was very weak (see a summary in Tables \ref{table-overview} \& \ref{table-summary}).
-No papers provided any evidence their code was adequately tested or rigorously developed; none used methodologies like RAP or \RAPstarp, and only one paper discussed any relevant Software Engineering methods (independent coding).
-Although in the sample 81\% of papers were published in journals that had code policies (which themselves are weak), 42\% of surveyed papers in those journals breached their code policies.
-One paper declared it had accessible code, but the relevant repository was and remains empty.
-Such results about code are comparable to surveys of data and data access \citeeg{machine-learning-reproducibility,no-raw-data,data-access}, and form part of the wider reproducibility crisis \citeeg{reproducibility-crisis}.
-
-\begin{change}
-\section{Background}
-The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences.
-There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature.
-For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.
-Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}.
-%For example, if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.
-Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theory development and technical maturation, to having robust, off-the-shelf instruments that measure temperature in reliable, internationally standardized units.
-
-Far more flexible and challenging than thermometers, computers are a unique, new technology.
-Understanding computers and integrating them into science is likely to take longer than intrumentalizing temperature did.
-Computation not only expands science's paradigms and supports new discoveries, but it can also \emph{do\/} new science --- and with AI, robotics, and other techniques, computation can do both real and speculative investigations autonomously, and better than humans.
-
-``Computational science'' has come to mean a particular style of science, based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational.
-Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ in all fields of science, computation is used at every step, from note taking, sound and photographic recording, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting finished publications.
-
-All areas of science are now being profoundly computerized, and we need to do it well.
-
-\subsection{The statistics/computation analogy}\label{statistics-computation-analogy}
-The central role of computational methods in science may be fruitfully compared to statistics, an established scientific tool --- and also, of course, like computer science, a substantial field of research in its own right.
-
-Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics.
-Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.
-\end{change}
-
-In good science, all statistics, methods and results are reported very carefully and in precise detail \cite{apa-statistics,Mariusz-reporting}, generally following strict journal or disciplinary guidelines.
-For example, a statistical claim might be summarized as follows:
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\hskip -\wd0\box0 Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
-\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.0009)$.
-Between group effect size $d = 0.55$ $(95\%$ CI $0.32,$ $0.77)$.'' \cite{example-stats}
-\end{quote}
-
-This typical wording formally summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
-It is a \emph{lingua franca}.
-It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results.
-Moreover, behind any such brief paragraph is a substantial, rigorous, and appropriate statistical analysis.
-
-Scientists write like this and conferences and journals require it because statistical claims need to be properly accountable and documented in a clear way.
-Authors, too, must avoid the easy temptation of being unjustifiably over-confident with their own results.
-
-Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Spiegelhalter-section}).
-Results should not be uncritically accepted just because they are claimed.
-The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted; in fact, there is widespread concern about the poor quality of statistics in science \cite{paul-hci,pnas-stats}.
-While it is assumed that statistics should be peer reviewed, and that review will often lead to improvement, these critical papers show that reviewers and editors are often failing to pick up on poor statistics.
-
-Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement.
-Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians.
-Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.
-
-Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them, will be generated by computer.
-Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it.
-Many papers now explore the contribution of AI and ML to their fields.
-The data --- including the databases and bibliographic sources --- and code to analyze it is all stored and manipulated on computers.
-Computers even help with the word processing and typesetting of the research.
-
-In short, computers, data, and computer code are central to modern science, not just to the explicitly computational sciences.
-Some AI work is uncovering biases and ethical issues that were previously unrecognized, so computational sciences are not just routinely contributing to existing science but extending its reach and improving its quality.
-
-However, using any code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
-Software code, like statistics, is also subject to unintentional bias \cite{Ben,se-bias}.
-All these issues are non-trivial concerns requiring technical expertise to manage well.
-As with statistics, good answers to such ``technical'' issues makes the science that relies on them better.
-
-A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.
-
-Failure to properly document and explain computer code undermines the scientific value of the models and the results they generate, in the same way as failure to properly articulate statistics undermines the value of any scientific claims.
-Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly reported code are widespread.
-
-We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained.
-
-An interesting overlap with statistical and Software Engineering sloppiness concerns the many papers that  disclose as part of their methodology that they used a particular package, for example
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\hskip -\wd0\box0 Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)'' \cite{sas9.2}
-\end{quote}
-
-but without giving any further details.
-The paper cited above describes one of its multivariate analyses as ``multiple correspondence analysis (MCA) and ascendant hierarchical clustering'' with no specific details --- reproducing the work would be non-trivial since the methodology would have to be reconstructed from scratch, even assuming the data was made available by the authors as the paper gives no details of the data used or how to obtain it.
-The problem is that the common practice of declaring using a named computational system (such as SAS in this case) does not help scrutiny, as such systems can do almost anything.
-\emph{How\/} those analyses might have been performed is not discussed, and one assumes it follows that the analyses could therefore not have been reviewed for basic scientific competence during the publication process.
-
-A reviewer, if nobody else, needs to actually examine the statistical code used and its documentation to assess whether the analysis presented in the paper is appropriate and sufficiently reliable.
-Furthermore, if the analysis actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible if it relies on idiosyncrasies in SAS\@.
-Of course, an author can disclose the idiosyncratic dependencies on, say, version 9.2; while this seems to be an onerous obligation, conversely it is arguable that if an author is unaware of code dependencies, then their science relying on them is equally unreliable.
-
-It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, and to uncover unknown dependencies.
-Indeed, the code may be poorly written (as this paper shows is very likely), so the results derived from the code simply may not be reliable.
-Being able to understand (at least in principle --- e.g., relying on reviewers to have checked it) the exact code used in implementing a model is critical to having confidence in the results that rely on it.
-
-In normal scientific reporting (outside of teaching and assessing science) details of methodology are routinely glossed.
-A chemist does not say they cleaned their glassware.
-One might argue, then, that scientists need not discuss their code in any detail because they know how to program and their code is correct.
-This argument is mistaken, as this paper shows.
-Unfortunately, code is rarely considered a valuable part of the science to which it contributes and this creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.
-
-\subsection{The role of code \\ in science and scientific publication}
-%Yet without code, models could not be run.
-
-Models map theory and parameters to describe phenomena, typically to make predictions, or to test and refine the theory supporting the models.
-With the possible exception of theoretical research, all but the simplest models require computers to evaluate; indeed even theoretical mathematics is now routinely performed by computer systems.
-
-Whereas the mathematical form of a model may be concise and readily explained, even a basic computational representation of a model can easily run to thousands of lines of code, and its parameters --- its data --- may also be extensive.
-The chance that a thousand lines of code, hand-written or otherwise, is error free is negligible, and therefore good practice demands that checks and constraints should be applied to improve its reliability.
-How to do this well is the concern of Software Engineering, and is discussed throughout this paper and in the \supplement.
-
-While scientific research may rely on relatively easily-scrutinized mathematical models, or models that seem in principle easy to mathematize, the models that are run on computers to obtain the results published are sometimes not disclosed, and even when they are they are long, complex, inscrutable and (our survey shows) lack adequate documentation.
-Therefore the models are very likely to be unreliable \emph{in principle}.
-
-If code is not well-documented, this is not only a problem for reviewers and scientists reading the research to understand the intention of the code, but it also causes problems for the original researchers themselves: how can they understand its thinking well enough (e.g., a few weeks or months later) to maintain it correctly if it has not been clearly documented?
-As a scientist pursues a research career building on their previous work, how can they be certain their work is reliable, and not merely converging to their  prejudices?
-Without proper documentation, including a reasoned case to assure that the approach taken is sound \cite{assurance-case}, how do researchers, let alone reviewers, know exactly what they are doing?
-
-Without substantial documentation it is impossible to scrutinize code properly.
-Consider just the single line ``\texttt{y = k*exp(x)}'' where there can be \emph{no\/} concept of its correctness \emph{unless\/} there is also an explicitly stated relation between the code and the mathematical specifications.
-What does it mean?
-What does \texttt{k} mean --- is it a basic constant or the result of some previous complex calculation?
-Does the code mean what was intended?
-What are the assumptions on \texttt{k}, \texttt{x}, and \texttt{y}, and do they hold invariantly?
-Moreover, as code generally consists of thousands of such lines, with numerous inter-dependencies, plus calling on many complex libraries of support code, it is inevitable that the \emph{collective\/} meaning will be unknown.
-A good programer would (in the example here) at least check that \texttt{k} and \texttt{x} are in range and that \texttt{k*exp} was behaving as expected (e.g., in case of under- or overflow).
-
-Without explicit links to the relevant models (typically mathematics, depending on the claims), it is impossible to reason whether any code is correct, and in turn it is impossible to scientifically scrutinize results obtained from using the code.
-Not providing code and documentation, providing partial code, or providing code without the associated reasoning is analogous to claiming ``statistical results are significant'' without any discussion of the relevant methods and statistical details that justify making such a claim.
-If such an unjustified pseudo-statistical claim was made in a scientific paper, a reviewer would be justified in asking whether a competent experiment had even been performed.
-It would be generous to ask the author to provide the missing details so the paper could be better reviewed on resubmission.
-
-Contrary to the views expressed in the present paper, some authors assert that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarizing data or properties of the data \cite{assessing-quality}.
-If code is inadequate, the so-called ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways.
-Indeed, none of the papers sampled (described in section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.
-
-Clearly, like statistics, programming (coding) can be done poorly and reported poorly, or it can be done well and reported well --- and any mix between these extremes.
-The question is whether it matters, \emph{when\/} it matters, and, if so, when it does, \emph{what\/} can be done to \emph{appropriately\/} help improve the quality of code (and discussions about the code) in scientific work?
-
-\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
-It is a misconception that programming is easy and even children can do it \cite{fixit}. More correctly, toy programming is easy, but real programming is very difficult.
-
-An analogy helps.
-Building houses is very easy --- indeed, many of us have built toy Lego houses.
-Obviously, though, a Lego house is not a \emph{real\/} house.
-It is not large enough or strong enough for safe human habitation!
-This point is obvious because we can see Lego houses, and everyone is familiar with building-block play; its real-world engineering limitations are too obvious to need stating.
-
-In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also invisible.
-The ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed, and because --- like Lego --- toy programs can look good but be unreliable, difficult to use and even dangerous.
-Thinking programming is easy is like appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations.
-These are professional engineering issues that Lego builders ignore.
-Certainly, even real building is much easier and faster when the technical details are ignored, as anyone who has experienced a cowboy builder can attest.
-
-Unlike building (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a very new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture.
-Relevant professional standards are not enforced.
-Problems for the reliability of science arise when legitimate doodling and tweaking software drifts into claiming scientific results that do not have reliable engineering structures underpinning them, let alone the properly developed and documented archived code, to justify them.
-In many countries, there are laws that require all but the very simplest structures to be formally approved from plans and inspected as they are built, but who writes plans for software, who inspects scientific models as they care coded?
-Yet the consequences of building a shoddy garage have negligible impact compared to the consequences of writing shoddy code that informs national public health policies or climate change interventions.
-
-Quite different, more formal, reasoning confirms the misconception about programming.
+\color{red}Quite different, more formal, reasoning confirms the misconception about programming.
 Lambda calculus ($\lambda$-calculus) is one of the very simplest of programming notations \cite{lambda}; it is just made up of strings of the four symbols $\lambda.()$ plus other symbols that have no pre-defined meanings.
@@
 Such ``simple'' lambda calculus expressions are clearly profound.
-Indeed, lambda calculus has a place in the history of logic because it was the source of the first undecidable logic problems (the Halting Problem being perhaps the most famous example).
+Indeed, lambda calculus has a place in the history of logic because it was the source of the first undecidable logic problems (the Halting Problem being perhaps a famous example).
 Lambda calculus thus proves that simple programs are \emph{not\/} simple; just because children can write programs does not mean anybody, including the children, can understand them.
 Indeed, ``programming is easy'' is contradicted by the Halting Problem, which shows that at least one trivial question about a program, whether it halts, is in general unanswerable.
-
-Since programming appears to be so easy, developing code has low status in scientific practice.
-Developers of code are rarely acknowledged in scientific papers.
-The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the technical contributors to it.
-Because it is easy, there is no need to work hard to make it correct --- anyway, it obviously works.
-While such views prevail, the vicious cycle is that the low status means software development is done casually, which reinforces the low status.
-
-In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code}, because code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.
-
-Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement).
-Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.
-
-\subsection{The central role of code is ignored}\label{central-role-of-code}
-Almost all scientific papers \emph{routinely\/} describe their experimental method, their data handling, and provide an overview of their analytic (usually statistical) methods.
-If they are theoretical papers, they will describe their mathematical models and any data that is used to confirm their models.
-Outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods avoid or control for experimental error.
-The relative silence on code is curious as science now depends on correct code as much as it ever depended on controlled experiments.
-While scientists avoid or manage contaminated samples, they ought to avoid or manage code bugs which have the same confounding impact.
-
-It is important that experiments and analysis are performed reliably and ethically.
-Because statistics, like code, is so readily susceptible to uncontrolled bias and error, there are many protocols and journal policies that enforce best practice, for example journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature.
-Yet PRISMA, like many such policies, ignores the critical role of code, and ignores the Software Engineering principles that assure computation is reliable and reliably reported.
-
-PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which is reasonable enough.
-PRISMA covers the review process carefully.
-For example, the authors should report the number of papers they included in their review.
-Perhaps $N=2000$.
-This number is then written into their draft paper, likely in several places.
-As the authors read and revise their paper, submit it, and respond to peer reviewers, it is very likely that the number of papers included in the survey changes, or other numbers and details may change.
-The authors now have a maintenance problem: where are the numbers that have changed, and what should they be changed to?
-Doing a search-and-replace, whether automated or by hand, is fraught with difficulties.
-What happens if $2000$ is used for some other purposes as well?
-What happens if some of the $2000$ values are written as $2,000$ or $2000.0$ and are not noticed?
-What happens if some $2000$ are year dates and are changed incorrectly?
-Then there are the Human Factors:  slips and errors will happen in this process anyway.
-Typos, slips during cut-and-paste, and other errors are common.
-Similar iterative revision cycles happen with any paper, not just with systematic reviews.
-Avoidable errors are problems across all of science.
-
-If a computer model is involved in the process (as it generally would be for a systematic review) then the value of $N$ in this example could very easily be stored in a file where the paper typesetting process can access it.
-For example, if \LaTeX\ is the system of choice, the analysis could generate, say,
-
-\begin{center}\texttt{$\backslash$newcommand\{$\backslash$N-papers-reviewed\}\{2000\}}\end{center}
-
-so that when and wherever the authors write \texttt{$\backslash$N-papers-reviewed} in their paper's text, the typeset paper says 2000 or whatever the correct value happens to be at the time.
-If so, then \emph{whenever\/} the survey is updated, the value cited in the paper is automatically and \emph{correctly\/} updated without any further intervention from the author.
-
-In general, not just PRISMA and not just numbers, but any data, text, graphs or tables, etc, can be reliably inserted into a paper automatically.
-
-PRISMA says nothing about how to ensure the final results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations.
-Such rules reinforce the fallacy that code is trivial and unimportant.
-
-These issues with PRISMA may sound trivial, but they are a symptom of the relaxed attitude to code that undermines the reliability of all published computational science.
-
-\subsection{Bugs, code and programming}\label{knowledge}
-Critiques of data and model assumptions are increasingly common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned.
-Yet data and program are formally equivalent (see \supplement, section \ref{on-code-data-publication}).
-Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and probably useless.
-Code, however, is harder to scrutinize, which means that errors in code have subtle, often unnoticed effects on results.
-
-Almost all code contains ``magic numbers'' --- that is, data masquerading as code.
-This common practice ensures that published data is very rarely all of the data because it omits the magic numbers embedded in the code.
-Data is often ``hard coded'' in programs with no explicit representation.
-Such issues emphasize the need for repositories to require the inclusion of code so all data, including that embedded in the code, is actually available.
-
-Bugs can be understood as discrepancies between what code ought to do and what it actually does.
-Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered.
-Contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination, or, better, failure to compile.
-None of the computational science surveyed or cited in this paper includes any such basic techniques.
-
-If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
-Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time).
-For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted, but are bugs in the wider sense of producing incorrect results (e.g., ill-conditioning) that is, what was intended was wrong.
-
-Random numbers are widely used in computational science, for simulation or for randomizing experiments.
-Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of bugs nai\"\i ve bugs \cite{knuth}.
-
-\begin{change}
-\subsection{The long-term problems of unreliable code}
-Scientists explore and extend the boundaries of rigorous knowledge.
-Put briefly, the purpose of scientific experiments is to vary details to either test and specify the boundaries of theories, or to discover new phenomena that lead to theory revision.
-
-If poor code, or poorly documented code, is made available with scientific papers, the code (or the absence of code) is a natural place to start replicating and varying experimental conditions,  including both data or code.
-However, if the starting point is not accurately known, whether due to bugs, obscure code, or because of poor documentation, then experimental variations will have an unknown effect.
-Theory will then be driven by artifacts of the code, not genuine phenomena.
-
-In section \ref{section-pandemic-modeling}, below, an example is documented of a research code development process of at least 15 years' duration where the code was admitted to be completely undocumented, leaving details in just one author's head.
-None of the various related papers describe any controls over the drift of the science, or how independent researchers building on it might have been able to build with confidence rather than merely reproducing the same errors (if any).
-
-Since the code in question was substantial and non-trivial, it is very unlikely that any constructive reproduction occurred outside the original laboratory and mindset; indeed, section \ref{reproducibility-concerns} describes how ``reproduction'' became trivialized because of community pressure to confirm the insights of this particular research.
-
-Trying to constructively refute aspects of this research in the Popperian sense \cite{popper-conjectures-refutations} would have been impossible.
-For example, had the relevant papers published critical code invariants then scientists building on the research could have explored whether those invariants remained valid and, if so, under what assumptions.
-In fact, invariants are the theories of code, and deserve as high a prominence in published computational science as the domain theories the computational modeling itself is supporting investigating.
-
-Since the symmetries in the scientific domains should be reflected in corresponding symmetries in well-designed code, this raises new possibilities of Formal Methods approaches (e.g., using automatic theorem provers) supporting advances in general science.
-\end{change}
-
-\section{State of the art \\ in computational modeling}
-\label{section-pandemic-modeling}
-A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modelling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer reviewers, there is no role for code or computation as such.
-It appears that the new insights, advances, rigor, and problems that computers bring to research were not considered relevant.
-
-A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
-The review \cite{diagnosis-reviews} ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed.
-Indeed, TRIPOD guidelines ignore code completely.
-
-It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for direct human use.
-Flowcharts, too, should be designed as carefully as code, for exactly the same reason as it is hard to program reliably --- flowcharts can present arbitrarily complex algorithms and are often sketchy and vague on details that are critical to reliable use.
-
-A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand; it did not model air travel or other factors required for later western COVID-19 modeling.
-The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.
-
-A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\hskip -\wd0\box0 
-For me the code is not a mess, but it's all in my head, completely undocumented. Nobody would be able to use it~\ldots
-'' \cite{ferguson-interview}\end{quote}
-
-This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
-The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
-(Typical $h$ values vary by discipline.)} of 93.
-Ferguson should be well aware of the standards of coding used in at least his own field.
-This comment, quoted above, is arguably representative of the standards of the field as a whole.
-
-Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}
-
-Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
-
-If Ferguson's C code had been made available for review, the reviewers would not have known how to evaluate it without the relevant documentation.
-It is, in fact, hard to imagine how a large undocumented program could have been repeatedly modified and repurposed over fifteen years without becoming incoherent, and someone recognizing that documenting it would help make it easier to maintain.
-
-If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
-In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
-
-The programming language C which was used \cite{tweet} is not a dependable language; to develop reliable code in C requires professional tools and skills.
-Some of the code is written in a na\"\i ve style (e.g., writing \texttt{*(a + i)} instead of \texttt{a[i]}, and with obscure numerical goto statements like \texttt{if(l == 0) goto S150}), transliterating pseudo-code, FORTRAN and Pascal code published in the 1970s and 1980s \citeeg{forsyth,Normal-Distributions}, with the risks that entails for introducing typos.
-%The code makes unexplained use of types such as float, double, int, 32 bit int, ...
-
-Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, or hardware.
-In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is used, which manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}.
-(The \supplement\ discusses these issues further.)
-
-Ferguson, author of the code, says of the criticisms of his code,
-
-\begin{quote}\raggedright
-\setbox0=\hbox{``}
-\hskip -\wd0\box0 
-However, none of the criticisms of the code affects the mathematics or science of the simulation'' \cite{thumbs-up}
-\end{quote}
-
-This claim is implausible.
-The original work on theoretical epidemiology may be fine if it does not use any of his code, but if the science is not supported by code that correctly implements the models, then the program's output cannot be relied on without independent evidence.
-Over the 15+ years the code was in development, the science running it informs will have drifted just as the code it relies on has.
-
-Typically, models will be developed iteratively as their results are improved to better fit a scientist's goals --- but this, especially when it is done by tinkering, as here --- risks making the code arbitrarily fit the goals (comparable to over-fitting; see section \ref{over-fit}), rather than to objectively elucidate the science.
-In fact, the Ferguson computational model is very large,\footnote{Ferguson's \texttt{covid-sim} system is \covidsimkLOC\ kLOC (thousands of lines of code), composed of \covidsimfiles\ files, and uses \covidsimdata\ of data.
-It is now rewritten from C into C++ with Python, R, sh, YML/JSON, etc.
-For more details, see \supplement.} so it is implausible that ``mathematics or science'' has been correctly implemented in them.
-Therefore Ferguson's \emph{reported\/} science cannot be reliable.
-Getting the science right, which depends on correct code, is a normal requirement of \emph{reproducibility}.
-
-\subsection{Concerns with reproducibility}
-\label{reproducibility-concerns}
-The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed the code could be run again and produced comparable results (compared using an Excel spreadsheet).
-Given that code was available, reproducing results with such a low level of sophistication is hardly surprising, as doing so does not depend on the validity of the code.
-That running code generates the results claimed in a paper is a very weak test, and one that should ideally have been established automatically at paper submission.
-\emph{If you do not know what you are reproducing, there is no scientific value in doing so.}
-
-Unfortunately, the terms reproducibility, replicability, and repeatability, have similar meanings in English and have been used in different ways by different authors.
-In \cite{codecheck,thumbs-up} the reproduction amounted to re-running the original code.
-It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper, and if the original code runs this confirms a basic level of access for the wider scientific community.
-
-A stronger criterion than basic reproduction is whether an \emph{independently\/} developed model developed from the same paper(s) produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.
-However, much stronger scrutiny of code is required to answer essential questions including:
-
-\newcount \enum
-\enum=1
-\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
-\the\enum
-\global\advance \enum by 1}
-\begin{enumerate}\raggedright
-\item Do other scientists, including reviewers and the authors, understand the code?
-\item Does the code implement the methods described in the paper?
-\item Has the code been over-fitted or tweaked to support specific claims in the paper?'
-\item Is there a definitive version of code?
-\item Is the code controlled and signed?
-\item What limitations does the code have?
-\item Was the code developed to any standard, and does it comply to that standard?
-\item How does the code protect against data, coding, and human error?
-\item Was the code tested adequately?
-\item \ldots\ and so forth --- questions are numbered for reference, not because any particular number or range of questions is complete. 
-\end{enumerate}
-
-All such questions also apply to specifications, documentation, assurance cases, test procedures, and other essential documents, not just to code.
-In turn, the levels of scrutiny demanded should be guided by explicit conceptual claims in the paper \cite{essence-of-software} --- for example, a pilot study requires weaker assurance than code that is developed concerning nuclear power, driverless vehicles, public health, etc.
-
-The questions in the list above are certainly hard to answer for all but the shortest systems, but corresponding levels of quality assurance are demanded for other methodologies \cite{notebooks,popper-conjectures-refutations,tripod,prisma,nih-policy,nih-nature}, such as data preparation and statistics to support claims in peer reviewed science.
-
-Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).} Documenting code in hindsight, even if done rigorously, may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was originally intended to have done.
-As the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models; it will be different.
-It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
-The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
-
-Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
-Pseudo-code looks deceptively like real code that might be copied to try to reproduce it, but pseudo-code introduces invisible and unknown simplifications.
-Pseudo-code, properly used, can give a helpful impression of the overall approach of an algorithm, certainly, but pseudo-code alone is not a surrogate for code: using it alone is even worse than not publishing code at all (see \cite{chinese}).
-Pseudo-code is not precise enough to help anyone scrutinize a model; copying pseudo-code introduces bugs.
-
-An extensive criticism of pseudo-code, and discussion of tools for reliable publication of code can be found elsewhere \cite{relit}. The \supplement\ provides further discussion of reproducibility.
-
-\subsection{Computational science \\ beyond pandemic modeling}
-\label{section-science-beyond-pandemic-modeling}
-
-Epidemiology has a high profile because of the COVID pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
-Other examples that were no obviously rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code.
-Its \supplement, which might have provided code, no longer exists.
-
-There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact.
-Climate change modeling is one such example that will have an impact long beyond the COVID pandemic.
-
-A short 2022 summary of typical problems of Software Engineering impacting science appears in \emph{Nature\/} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration.
-The paper \cite{nature-review}  makes some misleading comments about the simplicity of Software Engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''
-
-Guest and Martin in a 2022 paper \cite{psychological-modeling} promote the use of computational modeling, arguing that through writing code, one debugs scientific thinking.
-Psychology, their focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}.
-In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived.
-Computational models can be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks.
-It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely: for instance, ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973 and is now a 120 kLOC Common LISP and Python system \cite{actr}.
-Any paper using ACT-R would require additional code on top of the basic ACT-R framework.
-
-The psychology paper \cite{psychological-modeling} presents an example computational model from scratch to illustrate a framework of computational science.
-In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not modeled.
-Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.
-
-The Python code is presented without derivation; Software Engineering is ignored.
-The program listed in the paper certainly runs without obvious problems (ignoring some typographical errors due to the journal's publishers), but ironically the Python does not implement the mathematical specification explicitly provided for it, thus unintentionally undermining the argument of the paper.
-
-One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses.
-Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct (at least when finished).
-Otherwise, computational modeling with inadequate Software Engineering will introduce errors into scientific thinking.
-
-Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims.
-People reading the code will not know how to obtain results because they do not know exactly what was intended in the first place.
-The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even  flaws.
-
-Even good quality code has, on average, a defect every 100 lines --- and {such a low} rate is only achieved by experienced industrial software developers \cite{ourReview}.
-World-class software can attain maybe 1 bug per 1,000 lines of code.
-Code developed for experimental research purposes will have higher rates of bugs than professional industrial software, because the code is less well-defined and evolves as the researchers gain new insights into their models.
-In addition, and perhaps more widely recognized, code --- especially but not exclusively mathematical code --- is subject to numerical errors \cite{hamming}.
-It is therefore inevitable that typical modeling code has many bugs (reference \cite{NVP} is a slightly-dated but very insightful discussion).
-Such bugs undermine confidence in model results.
-
-Only if there is access to the \emph{actual\/} code and data (in the specific version that was used for preparing the paper) does anyone know what researchers have done, but merely making code available (for instance, providing it in their \supplement\ with papers, putting it in repositories, or using open source) is not sufficient, for example running models may require access to special hardware.
-
-Some COVID-19 papers \citeeg{unfinished} make unfinished, incomplete code available.
-While some \citeeg{unfinished,lancet-unfinished} make what they call ``documented'' code available, they provide no more than superficial comments.
-This is \emph{not\/} documentation as properly understood.
-Such comments do not explain code, explain contracts, nor explain algorithms.
-Contracts, for instance, originated in work in the 1960s \cite{hoare}, and are now well-established practice in reliable programming {(see the \supplement\ for a checklist of many relevant, professional Software Engineering concepts and resources)}.
-
-Even if a computer can run it, badly-written code (as found in \emph{all\/} the research reviewed in the present paper, and indeed in computer science research, e.g., \cite{machine-learning-reproducibility}) is inscrutable.
-Only if there is access to \emph{adequate\/} documentation can anyone know what the researchers \emph{intended\/} to do.
-Without all three (code, data, adequate documentation), there are dangers that a paper simplifies or exaggerates the results reported, and that omissions, bugs and errors in the code or data, generally unnoticed by the paper's authors and reviewers, will have affected the results they report \cite{relit}.
-
-Making outline code (including pseudo-code) available without proper documentation and without discussing its limitations is unethical: it encourages others to reproduce and build on poor work.
-
-\section{Improving \\ the computational sciences}
-\label{section-discussion}
-\subsection{The emphasis on data}
-Data has been at the center of science, certainly since the earliest days of astronomy collecting planetary and other information.
-Today it is widely recognized that lack of accessible and usable data that has already been collected limits the progress of science.
-Low quality data and poor access to data causes reproducibility problems, an increasingly recognized problem --- in 2015 it was estimated that \$28B a year is spent on preclinical research alone that is not reproducible \cite{preclinical-reproducibility}.
-
-Curating data is taken seriously as a part of normal science and peer reviewed publication.
-Journal policies widely require appropriate discussion of data, much like they require appropriate discussion of statistics.
-Journals often require archiving data in standard formats so it can be accessed for reproduction in further scientific work.
-
-There are many current activities to proceduralize and standardize the more effective curation and use of data, such as the FAIR principles (Findable, Accessible, Interoperable and Reusable) for scientific data management and stewardship \cite{fair,fair-principles}, and in the development of journal and national funder policies.
-For example, the 2022 update to the US National Institutes of Health data policies \cite{nih-policy} is described as a ``seismic mandate'' by \emph{Nature\/} \cite{nih-nature} in its attempt to improve reproducibility and open science.
-They did not include code.
-
-These cost estimates and initiatives under-play the role of code as a critical component despite its becoming the new laboratory for almost all science.
-The role of code specifically in modeling is discussed throughout this paper; without bespoke code, proposed models (unless very abstract) cannot make a quantifiable contribution to the literature.
-More generally, much data is embedded in code, and in the limit code and data are indistinguishable (see \supplement).
-Code has additional problems of versions and compatibility beyond those of data, for example suitable compilers to run old code may no longer be available, and --- worse --- programming systems may silently produce different results when used on different computers.
-
-In general, without proper management of code --- for example to detect and report version control differences --- sharing code may even be counter-productive.\footnote{The data and code shared with the present paper includes cryptographic checksums; if somebody reproducing the work described here does not obtain the same checksums at least when they start their work, then there are problems that need investigation before relying on the reproducibility of the data.}
-
-Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
-The evidence (see \supplement) suggests that, generally, some but rarely all develop code that is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no evidence (before, during, or after the date of the survey sample) that any published code was maintained using repositories.
-This is consistent with finished code being uploaded to a repository just for the purposes of satisfying publishing requirements, but not being maintained in a repository because the authors found it advantageous to do so.
-Using a standard repository for lodging a paper's supporting code helps other scientists access the code easily, but not using the repository for developing and maintaining the code means the author of the paper misses out on many helpful features of repositories, such as version control, open source development and review, actions and other approaches for automating \RAPstarp\ (see below, section \ref{RAP-section}), and so on --- depending on the repository features available.
-
-\begin{change}
-\subsection{\RAPstar s: Generalized reproducible \\ analytical pipelines}\label{RAP-section}
-Writing a paper typically starts in a word processor (such as Microsoft Word), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on.
-The process will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.
-
-The simplified diagram in figure \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then edited into the paper.
-
-\begin{figure*}[t]
-\begin{center}
-\def\drop#1{\setbox0=\hbox{\lower .5em\hbox{#1}}%
-\ht0=0em
-\dp0=0em
-\copy0}
-\def\lowarrow{\drop{$\rightarrow$}}
-
-\def\bigBracket{$
-	\setbox0=\hbox{\lower 1.2ex\hbox{$\left.\vbox to 4em{\vfill}\right\}$}}
-	\ht0=0em \dp0=0em 
-	~\copy0
-$}
-\begin{change}
-\begin{tabular}{|lcl@{}cl@{\hskip 2em}cl|} \hline
-\drop{\bf Data sources}&\lowarrow&\bf \drop{Analysis} & \lowarrow&\bf Select results& \lowarrow& \bf Submit for\\
-\bf &&\bf & &\bf for write up & & \bf publication \\ \hline \hline 
-&&&&&&\\
-Experiments&& Hand calculations &&&&\\
-Standard data&& Packages  && Copy \& paste &&\\
-Search engines&& SPSS etc & \bigBracket & and edit data && Final paper\\
-Literature &&  Graphics packages  && (text, images, graphs, etc)&&\\
-Sensors && Specially-written code && into paper &&\\ 
-\vdots && \vdots &&&& \\ &&&&&&\\ \hline
-\end{tabular}\end{change}
-\end{center}\caption{\begin{change}A simplified schematic of the publication pipeline.
-For clarity, the pipeline has been linearized; in general, there will be much cyclic iteration and refinement.
-The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline processes so that they can be run automatically, and hence reproduce the results that underpin the final paper.
-The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.\end{change}}
-\label{fig-pipeline}
-\end{figure*}
-
-For simplicity, the schematic pipeline in figure \ref{fig-pipeline} omits showing many standard steps in the creative scientific process: each step is iterated and modified as the research progresses, and as referees require revisions.
-The point illustrated, however, is that in typical scientific practice each arrowed step in the diagram leading to a published paper is largely or entirely manual, typically selecting and copying output from the previous phase, and then pasting the results into the next.
-The pipeline steps of data $\rightarrow\cdots\rightarrow$ paper is then repeatedly run by hand as the various components are refined and improved until the authors are happy with the paper.
-
-Different data is selected; calculations and analyses are modified; programs are debugged.
-As problems are detected in the paper, the data, calculations and programs are reviewed and refined.
-The process is rarely systematic, and even less likely to be documented --- after all, the atomic steps are trivial copy and paste actions.
-The final paper and the ideas it embodies are all that matters.
-
-The basic insight of the reproducible analytic pipelines (RAP) proponents is that every time any atomic step in the pipeline process is performed it could have been automated \cite{rap,goldacre,turing-way}.
-If automated, it could then be repeated reliably --- unlike a manual cut and paste which is error-prone (in different ways!)\ every time it is done.
-In particular, when a process is automated, any other researcher, whether part of the authorship team or a later reader of the paper, can reproduce it reliably.
-It can be repeated if any experimental data, literature, or other knowledge changes, and the paper's analysis brought up to date with ease.
-
-For example, if the paper in question is a systematic review, in principle it could be kept current by automatically re-running the programmed atomic actions that it was built with in the first place.
-Indeed, this ability is one of the original motivations of RAP: Government agencies can produce up to date reports on request without having to repeat all the manual work and risk making procedural errors in doing so.
-Each time they do so, the RAP pipeline is reviewed and improved, so the quality of the reproduced work improves --- unlike in a non-RAP process where new errors are generally introduced.
-
-RAP embodies Donald Knuth's comment,
-
-\begin{quote}\raggedright
-{\setbox0=\hbox{``}
-\hskip -\wd0\box0
-Science is what we understand well enough to explain to a computer'' }
-\\ \hfill from the foreword to $A=B$ \cite{a=b}
-\end{quote}
-
-The corollary is that if we are doing cut and paste that is arbitrary and cannot be programmed into a computer, then we are not doing science.
-Science is an algorithmic process, and therefore, as Knuth says, if we understand well enough what we are doing in science, we can explain it as programs, as code, for a computer to automate.
-That is RAP in a nutshell.
-
-For example, in the present paper, we analyzed \the\dataN\ papers with \the\countAuthors\ authors, and one of the papers used a program that was composed out of \covidsimfiles\ files and had over \covidsimkLOC\ thousand lines of code.
-In the ``old days'' these numbers would have been manually worked out, then read and typed up by one or more of the authors.
-This is a potential source of error.
-As the analysis is extended the numbers may well change, and the authors would have to stay aware that a number like \the\dataN\ will need checking and updating over the period the paper is being written.
-It is an error-prone process, and has to be regularly repeated.
-Worse, other researchers may have no idea how the numbers were generated --- the paper is not fully reproducible.
-(More generally than simple numbers, as illustrated in this paragraph, papers may also have tables, diagrams, plots, and other types of result generated during the research.) Instead, in this paper, all those numbers (and many more) were computed automatically and change automatically if the data changes, and they were then inserted into the text of this typeset paper automatically.
-The figures are very probably correct.
-
-Once processes in the pipeline are automated, this means that there is code to can run those steps again.
-Once there is code, it can be managed in a version control system.
-A version control system then provides an audit trail for free, as well as many advantages such as being able to backtrack to an earlier version to undo now-unwanted edits.
-Importantly, the automatic code can perform sanity checks on the process --- a very simple example is automatic bibliography systems that check that journal names, DOIs are correct and that references are correctly numbered, and so forth.
-They also allow the bibliographic data to be pooled and curated with other scientists, which improves its scope and quality.
-
-Many systems provide tools to do this.
-GitHub (which is mentioned throughout this paper) provides \emph{actions\/}, which are named specifications that run workflows.
-GitHub happens to specify actions in the language Yaml, which, being a textual notation, in turn means that all the helpful features of GitHub --- open source, version control, etc --- can be applied to these pipelined processes as well.
-Helpful pipelines can be documented, shared and improved with open collaboration.
-
-In the limit, almost the entire scientific process can be automated (and its interaction with the world automated with robots).
-There are many ways to do this; for example, the mathematical programming system \emph{Mathematica\/} makes the analysis of the data and the calculations and the paper ``the same thing'' in its integrated notebook user interface --- which behaves a bit like Microsoft Word, except that formulas can be evaluated and plotted, etc, with ease.
-Alternatives include R Markdown, an approach based in the open mathematical system R; many variations of literate programming \cite{relit}; a system, Lepton  \cite{lepton}, that allows a \LaTeX\ document to execute and include arbitrary code, and language-independent notebook systems like JupyterLab; and so on.
-
-In all such systems, rerunning calculations re-creates the paper.
-Once \emph{Mathematica\/} (or which other system is used) has been set up, there is no repeated tinkering and error-prone copy and paste.
-Indeed, every time the system is run, the authors are likely to double-check the results after they have been re-computed --- so the RAP process actively helps reduce errors.
-The benefits and ideas of \RAPstar\ are starting to be more widely recognized, if not coherently integrated under that name.
-For example, the Executable Paper Grand Challenge Workshop \cite{Executable-Paper} explored the benefits of running scientific papers \emph{as\/} programs.
-
-But a final paper is not the only product scientific authors create and publish.
-In computational modeling, they also often create the code and data that generates results for their paper, and both are often made available.
-Here is a critical insight: the code, just like the paper, contains text (source code and documentation) and data (e.g., constants) which have been copied and pasted from elsewhere.
-In more sophisticated RAPs, then, the coding process itself is fully part of the pipeline.
-
-Software engineers have many tools for automatic code development (such as Unix's \texttt{make}) but the idea that these tools can be used to integrate and help automate code authoring as well as its documentation \emph{and\/} paper authoring is radical.
-Amongst other things, it means that the \emph{entire\/} research and development process of the paper (as well as all its underlying code) can be reproduced by others.
-Since this view has not been emphasized previously, we shall call it \RAPstar\ --- and we hope people will ask what that term means.
-The present paper is a modest example of \RAPstarp, more details of which are described in the \supplement.
-
-Note that as \RAPstar\ objectifies how science is done to a standard sufficient to enable a computer to run, it enables all of the methodologies of Software Engineering to be brought to bear.
-\RAPstar\ means the normally tacit, manual, and undocumented processes of science have become explicit code.
-Code can  be scrutinized, optimized, and ensured correct by standard Software Engineering practice --- thus \RAPstar\ does not just automate science for reproduction, it makes the automation explicit so doing science can be reasoned about --- mentally and automatically with sophisticated tools, theorem proving and AI and so on --- and hence science will be improved.
-\end{change}
-
-\section{A call to action}\label{summary}
-Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide.
-
-Software used for informing public health policy, medical research or other medical applications is \emph{critical software}.
-Professional critical software development, as used in aviation and the nuclear power industry, is (put briefly) based on \emph{correct by construction}: \cite{cbc} effectively, design it right first time, supported by numerous rigorous techniques, such as Formal Methods, to manage error.
-(See extensive discussion in this paper's \supplement.) Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable.
-Conversely, not following these practices undermines the rigor of the science.
-
-An analogous situation arises in ethics.
-There are some sorts of research that are ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
-Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems.
-National funders, and others, therefore require Ethics Boards to formally review ethical quality.
-Medical journals will not publish research that has not undergone appropriate ethical review.
-
-Analogously, and supplementing Ethics Boards, Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality Software Engineering.
-Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to epidemic modeling papers --- and in due course, all scientific papers --- being required to include Software Engineering Board clearance statements as appropriate.
-{Software Engineers have a code of ethics that applies to \emph{their\/} work in epidemic modeling \cite{ethics-code}.}
-
-{The present paper did not explore data, because in almost all cases the code and data were explained so poorly and archived so haphazardly it would be impossible to know whether the author's intentions were being followed.\footnote{{For the present paper, all the code, data, analysis and documents are available for download in a single zip file.}} Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
-Ideally, the authors should test a clean deployed build of their code and save the results.
-Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims.
-This requirement is no more unreasonable than requesting good statistics, as discussed in the opening of the paper.
-And the solution is the same: relevant experts --- whether statisticians or Software Engineers --- need to be routinely available and engaged with the science.
-SEBs would be a straight forward way of helping achieve this.}
-
-There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university.
-Active, professional Software Engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art.
-There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.
-
-Open Source generally improves the quality of software.
-SEBs will take account of the fact that open source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers.
-Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis {and other quality control methods for which the lead scientist and Software Engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).}\ {Software Engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer reviewed publications \cite{acm-artifacts}.}\ {See this paper's \supplement\ for further suggestions.}
-
-A potential argument against SEBs is that they may become onerous, onerous to run and onerous to comply with their requirements.
-A more mature view is that SEBs need their processes to be adaptable and proportionate.
-If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation.
-Hence SEBs processes are likely to follow a risk analysis, perhaps starting with a simple checklist.
-{There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar.
-Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.}
-
-A professional organization, such as the UK Royal Academy of Engineering ideally working in collaboration with other national international bodies such as IFIP, should be asked to develop and support a framework for SEBs.
-SEBs could be quickly established to provide direct access to mature Software Engineering expertise for both researchers and for journals seeking competent peer reviewers.
-In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations.
-Given the urgency, this paper recommends that \emph{ad hoc\/} SEBs should be established for this purpose.
-
-SEBs are a new suggestion, providing a supportive, collaborative process.
-Methodological suggestions already made in the literature include open source and specific Software Engineering methodologies to improve reproducibility \cite{basic-reproducibilty,open-source}.
-While \cite{ABCs-SE} provides an insightful framework to conceptualize approaches and compare their merits, there is clearly scope for much further research to provide an evidence base to motivate and assess appropriate interventions to help scientists do more rigorous and effective Software Engineering to support their research and publishing.
-These and further issues are discussed at greater length in the \supplement.
-
-\subsection{Action must be mutual \\ and interdisciplinary}
-\setcounter{footnote}{1}
-Code is only part of science, and only one critical factor in the wider reproducibility crisis: SEBs must work with --- and be engaged by --- other reproducibility initiatives.
-
-Relying on Software Engineering Boards (SEBs) \emph{alone\/} would continue one of the current besetting problems about the role of code.
-The conventional view is that scientists do the hard work compared to the ``easy'' coding work (see sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}) so they just need to tell programmers what to do.
-This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, which he explores at some length, is that we need to spend more effort in working out what computers should do (i.e., do the science better) and then just tell programmers to do \emph{that}.
-
-On the contrary, competent Software Engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and often that logic needs refining or optimizing.
-In other words, the Software Engineers can bring important insights back into the science, hence improving or changing the questions and assumptions the science relies on.
-This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but that is ill-conditioned, there is no good answer.'' Ideally, then, it is not just a sequential process of science $\rightarrow$ code $\rightarrow$ results, but an iterative cycle of collaboration and growing mutual understanding.
-
-In short, the way SEBs work and are used is crucial to their success.
-Software engineers can help improve the science, so it is not just a matter of asking a SEB  whether some coding practices (like documentation) are satisfactory, but whether the SEB has insights into the science itself too.
-Most effectively, this requires interdisciplinary working practices (science plus Software Engineering) with mutual respect for their contributing expertise.
-
-\subsection{Methodological statements}
-Most journals (using the word ``journals'' as a shorthand for all publishing outlets, conferences, workshops, videos, books, etc) and funders require explicit statements how the authors have conformed to appropriate methodological standards covering issues, such as conflicts of interest, ethics, data access, consent, and so forth.
-It would be easy for journals and funders to require equivalent types of statements on Software Engineering.
-
-Studies of data access statements show that they are unreliable: some authors withdraw papers when journals request access statements \cite{no-raw-data} (indicating that journals that do not make an explicit request are likely publishing papers that will not provide access), and some authors do not respond to access requests \cite{data-access}.
-How we help scientists who do not want to provide data or code access is one problem, but the serious issue for science is how to ensure any statements made are accurate, and any access provided is actually helpful (e.g., well-defined, versioned, etc) to support useful reproduction.
-Journals and funders should provide support for hosting data and code (and any other relevant data, such as qualitative data, video, etc), and the review process must check that authors actually provide the material as they claim in the methodological and access statements.
-Conversely, scientists should be able to access funding to ensure data and code access, and, as appropriate, funding for on-going maintenance of the databases, which will typically require funding beyond the end of the normal funding period.
-
-Intellectual Property (IP) is an increasing concern, both for author scientists and their sponsors who want royalties, and for other scientists wishing to freely build on the published science.
-Particularly concerning access to code, IP is potentially and often is in conflict with scientific openness.
-Methodological statements should be made concerning any IP associated with code, and to what extent this interferes with open access to code.
-More routine discussion should include raising known system dependencies, such as operating system, compiler, or special hardware dependencies; it is also appropriate to mention standards conformance, such as to IEEE Floating-Point Arithmetic (IEEE~754).
-
-Journal policies could start to explicitly encourage computationally reproducible science using RAP and \RAPstar\ techniques.
-That is, the research's methodology itself may be a mixture of data and code.
-As this paper's \supplement\ shows in section \ref{restrictive-policies}, many journals (e.g., \emph{PLOS ONE\/} and, ironically, \emph{IEEE Transactions on Software Engineering\/}) and repositories have policies that make RAP harder.
-Methodological statements should be required that make clear what access rights are available for RAP or \RAPstar\ material, as it is much more likely to raise IP issues that normal disclosures.
-In particular, if the authors plan on publishing a series of papers based on the same methodologies, the RAP/\RAPstar\ access might be provided in a later paper or held under escrow by the journal (or funding body).
-
-Journals and funders often require data and code access statements, but as this paper has made clear, code is complex and it is rarely easy to understand and scrutinize even with access to substantial documentation (which is unusual).
-It is therefore recommended that journals and funders require \emph{assurance arguments} \cite{assurance-case}, a familiar technique particularly from the safety assurance domain.
-Assurance arguments provide a concise, high-level argument that the system does what is claimed.
-Assurance arguments can be more or less detailed, and more or less formal in their approach; we envisage referees would have views on the level of detail and formality required for any specific contribution.
-
-Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access statements apply to all information, regardless of any arbitrary classification of it as code or data (or documentation, assurance case, etc).
-
-\begin{change}
-\subsection{The paper as a scientific laboratory}
-RAP (see section \ref{RAP-section} and figure \ref{fig-pipeline} ) recognizes that much of doing science is, to the extent that it is reproducible, a proceduralized, algorithmically-framed process: many scientific processes, most obviously in computational science, can be automated and reproduced at will (at least when they are done to high enough Software Engineering standards).
-\RAPstar\ further recognizes that the codification of science is itself a scientific object.
-That is, the code recording the scientific pipeline is not just scaffolding to support a scientific project, but is in itself a scientific product.
-It can be criticized, debugged, optimized, refactored, generalized into a virtual machine --- code --- that can do any related type of science.
-Written well, the code is an explanation of how particular science is done.
-
-In other words, as a paper is developed the computational contributions do not just generate results, but they objectify the scientific processes that have generated the results: they make doing science a first class object that can be seen, scrutinized and thought about --- it is code, after all --- generalized and applied more widely.
-In computer science terms, the paper (paper, data, code, etc) becomes a first-class object; that is, it has all the rights and operations as available to any other object.
-In scientific terms, then, it can be controlled, measured, stored, mixed, split, searched, analyzed, written-up, and so forth as an object of proper scientific enquiry.
-This is trivially true in the field of Computer Science, since (for many papers) the object of study \emph{is\/} a computer program --- the only debate is how much of that code is literally in the paper (or downloadable from it) or just talked about in the paper.
-
-Code thus becomes a tool reifying scientific insights, which therefore (if of adequate quality) itself merits publication, so others can reproduce, generalize and progress the science confidently themselves.
-The scientific paper turns from being a rhetorical record of some scientific investigation into also embodying the algorithms that performed the science.
-\RAPstar\ makes the algorithms explicit.
-The paper becomes the laboratory where the code is developed, tested and reasoned about.
-As the algorithms in code are explicit, science gains leverage and will advance more rapidly.
-
-One of the many exciting opportunities is transforming how scientists code, how they turn computation from mere support into a keystone of thinking about their activities.
-
-Today, computational science papers typically have some mathematical equations to state and communicate their theories, but the results are generated by embarrassingly complex imperative programs that defy scrutiny, typically written in languages like C, Python, or Java.
-Increasingly, though, computational science is building on programming languages that can directly support science.
-In a language like \emph{Mathematica}, a scientist can program directly in (say) systems of differential equations, and there need be no difference at all between how ideas in the paper and the ideas in the code are expressed.
-In fact, the paper has become the scientist's laboratory, the place where they do research.
-
-The more we view the paper as a scientific laboratory, the more we gain from the \RAPstar\ perspective and the more science will gain from improved, conceptually broadened computational science.
-The more we will want to mature Software Engineering standards too, because the quality of future science depends on them.
-
-\subsection{Benefits beyond computational science}
-Computational sciences recognize the key role of computation (if not sufficiently emphasizing the role of code), but many fields do not recognize computation as such and therefore they are missing out on the leverage that comes with recognizing computation as a first class player in their scientific activities.
-
-One example will suffice.
-A lot of healthcare is supported by computers, yet medical papers remain traditional and only very exceptionally refer to or include code.
-However, much clinical practice relies on computer analysis, but inevitably it has to use code unrelated to the code developed by the scientists doing the primary research, as their code is neither peer reviewed nor published.
-Code is professionally invisible, has no prestige, and there is therefore no investment in studying or improving it.
-Conversely, the challenging critical issues (including patient safety) that require clinical code to be more reliable than scientific tinkering do not get evaluated by researchers.
-Certainly, the issues, if noticed,  would stimulate further science --- sexism and racism in healthcare algorithms being a case in point \cite{fixit}.
-
-Without taking the lessons of improved computational science to other fields, like medicine, there will continue to be an unfortunate and unnecessary disconnect between research and practice, and one where both science and practice suffer because code is not recognized as a contribution to science \cite{fixit}.
-
-\input models-data-fitting/generated-modelsSection.tex
-
-%\subsection{Why improving \\ computational sciences matters}\label{over-fit}
-%Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that.
-We do the science, and we are in the world, so there are challenging meta-theoretical issues as well as practical experimental and resourcing issues.
-Karl Popper is one of many philosophers of science to develop principles and theories about science \cite{popper-three-worlds,popper-conjectures-refutations}.
-Arguably first noticed by Herbert Simon \cite{sciences-artificial}, the ``sciences of the artificial'' now raise new issues, as does more recent non-reviewed mass publication of results (including publication of results based on non-reviewed computational models), for instance in archival systems such as \url{arxiv.org}
-%
-%Science to date has been inevitably driven by human \ae sthetics: that is, simple, elegant, understandable, beautiful, brief, exciting, intelligible, are all human properties that are applied to doing, funding, defining, and selecting ``good'' science.
-%
-%Computation is starting to undermine such human-oriented criteria, for what are a good criteria of a simple theory when a computer model can generate a human-understandable interpretations of \emph{anything\/} including arbitrarily complex, buggy, even fictional theories?
-The danger is that superficially successful computer programs may unintentionally over-fit phenomena, so apparently confirm our understanding of them.
-The code we use, being inscrutable, may come to model our desires better than model the data.
-The classic examples of over-fitting are visually-obvious complex Real$\rightarrow$Real functions of one variable (e.g., as visualized in a simple example shown in figure \ref{fig-overfit}), but when the models are implemented as huge computer programs with gigabytes or more of parameters, there is no comprehensible and accurate representation for them.
-In general, we do not (sometimes, cannot) see their complexity and arbitrariness,\footnote{\input generated-fittingModels-data.tex
-%These models were written without relying on mathematical packages so the code is explicit, hence making it more reasonable to make a direct comparison.
-Indeed, the code for models looks superficially similar, emphasizing that (\emph{i\/}) simple variations, changes or mistakes in code can have dramatic effects on results, and  (\emph{ii\/}) visualizing the results of running code --- as shown in figure \ref{fig-overfit} --- can help identify bugs (if any).} and attempts to improve (``refactor'') them would only introduce further obscurities.
-%
-%Until we improve our standards for the computational sciences --- this paper argues by starting by intentionally engaging competent Software Engineering (and also updating the under-pinning philosophy of science, perhaps beginning with a Popper World~0) --- we risk doing and promoting bad science because we are not managing the unleashed complexity of the codes that our science relies on.
-%
-%\begin{figure}[t]
-%{ \begin{change}  \centering
-%\newdimen \tw \tw=\textwidth \advance \tw by -\columnsep \divide \tw by 2
-%   \includegraphics[width=\tw]{plot.pdf} 
-%\end{change} }  
-%\input generated-modelsAt5.tex
-%\caption{%
-%Much computational science is concerned with finding plausible multi-dimensional models that fit data.
-Shown here is notional experimental 2D data (the dots), together with a linear regression and an exact polynomial model.
-The over-fitted polynomial model fits the sample exactly, but since the experimental data is presumably subject to error (indicated by the confidence interval, itself estimated from another model) the linear model would generally be considered a better description of the experimental results.
-If we assume that the error in measuring $y$ is normally distributed (or similar), the extreme values the over-fitted model predicts would be considered unlikely; for example, \exacty\ predicted by the exact model versus \lineary\ predicted by the linear model.}
-%\label{fig-overfit}
-%\end{figure}
-
-\ignore{\subsection{Suggestions for further work}
-Although this study of Software Engineering in scientific research (specifically, in peer reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
-However, the insights appear to be general.
-
-Further work to extend the scope of the survey beyond the basic requirements of the present paper is of course worthwhile, but the following cases (listed in the next few paragraphs) suggest that the problem is widespread.
-We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor.
-}
-\end{change}
-
-\section{Conclusions}
-We need to improve the quality of code and Software Engineering that supports science.
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview}, the wider evidence reviewed shows that current coding practice makes for poor science in many fields.
-In a pandemic, scientific modeling, such as epidemiological modeling, track and trace \cite{excel-fiasco}, modeling COVID mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, etc, drive public policy and have a direct impact on quality of life.
-Unfortunately, Software Engineering good practice, to help \emph{do\/} science has been absent to date.
-
-The main challenges to mature computational scientific research are:
-
-\enum=1
-\begin{enumerate}\raggedright
-\item 
-To manage software development to reduce the unnoticed and unknown impacts of bugs and poor programming practices that papers rely on.
-Computer code should be explicit, accessible (well-structured, etc), and properly documented.
-Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}.
-Professional software methodologies should not be ignored.
-
-\item 
-To use computation to help make scientific processes explicit, so that they can be reproduced, scrutinized and improved.
-RAP is an increasingly popular way to help do some of this, but as this paper points out, RAP should be generalized to \RAPstar\ to help the computational parts of science as well, leading to a virtuous circle.
-
-\item
-To support and develop the scientific community in the professional use of computation.
-\end{enumerate}
-
-While programming seems easy, and is often taken for granted and done casually, programming \emph{well\/} is very difficult \cite{fixit}.
-We know from software research than ordinary programming is very buggy and unreliable.
-Without adequately specified and documented code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
-Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion: computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not make it more reliable \cite{reproducibility,relit,popper-conjectures-refutations}.
-
-Software Engineering Boards (as introduced in this paper) are a straight forward, constructive and practical way to support and improve computer-based science.
-This paper's \supplement\ summarizes the relevant professional Software Engineering practice that Software Engineering Boards would use, including discussing how and why Software Engineering helps improve code reliability, dependability, and quality.
-
-\makeatletter
-\immediate\write\@auxout{\string\def\string\continueSectionNumbers{\arabic{section}}}
-\makeatother
-
-\section*{Supporting information}
-\newcount\csrefcount \csrefcount=0
-\def\csref{\global\advance\csrefcount by 1}
-\long\def\ethics#1{\paragraph*{Ethics}#1}
-\long\def\ack#1{\paragraph*{Acknowledgments}#1}
-\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
-\long\def\aucontribute#1{\paragraph*{Author contribution}#1}
-\long\def\competing#1{\paragraph*{Competing interests}#1}
-\long\def\funding#1{\paragraph*{Funding}#1}
-
-\ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson provided very helpful comments.}
-%\the\csrefcount\ computer science reviewers.
-
-%\ethics{This article presents research with ethical considerations but does not fall within the usual scope of ethics policies.}
-
-\dataaccess{There is an extended discussion of the methodology of this paper and its benefits in the \supplement, section \ref{on-code-data-publication}.
-The \supplement\ also presents all raw data in tabular form.
-All material is also available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards},\footnote{\color{red}This is a temporary URL before meeting publication repository requirements for accepted papers.\color{black}} which has been tested in a clean build, etc.
-
-The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \the\JSONerrorCount\ possible classes of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset all results reliably in this paper and in its \supplement\ \emph{directly\/} from the automatic data analysis.
-
-In addition, a standard CSV file is generated from the JSON in case this is more convenient, for instance to browse directly in Excel.}
-
-\aucontribute{Harold Thimbleby is the sole author.
-An preliminary outline of this paper, containing no supplementary material or data, was submitted to the UK Parliamentary Science and Technology Select Committee's inquiry into UK Science, Research and Technology Capability and Influence in Global Disease Outbreaks, under reference LAS905222, 7 April, 2020.
-The evidence, which was not peer reviewed and is only available after an explicit search, briefly summarizes the case for Software Engineering Boards, but without the detailed analysis and case studies of the literature, etc, that are in the present paper.
-It is available to the public \cite{parliamentary-evidence,my-parliamentary-evidence}.}
-
-\competing{The author declares no competing interests.}
-
-\funding{This work was jointly supported by See Change (M\&RA-P), Scotland (an anonymous funder), by the Engineering and Physical Sciences Research Council [grant EP/M022722/1], by the Royal Academy of Engineering through the Engineering~X Pandemic Preparedness Programme [grant EXPP2021\textbackslash 1\textbackslash 186], and by Assuring Autonomy International Programme, Assuring Safe AI in Ambulance Service Triage.
-The funders had no involvement in the research or in this paper.}
-
-%\nolinenumbers
-
-{\raggedright
-
-\bibliographystyle{mysiam}
-\def\urlprefix{}
-\initialiseBibliography{References}{1}{}
-
-\newcount \bibitemn \bibitemn = 0
-\let\oldbibitem=\bibitem
-\def\bibitem#1{\ifnum \bibitemn = 0 \else \egroup\vskip .7ex plus 1ex \fi
-\advance \bibitemn by 1
-\vbox \bgroup \oldbibitem{#1} }
-\def\endofbibliography{ \egroup \vfill}
-\bibliography{paper-seb-main.bib}
-
-% Note that the supplementary material calculates how many references there are,
-% because it reads in this .tex file's .aux file so that it can cite the references 
-% above with the same number sequence. 
-%
-% It simple redefines \bibitem to work it out while it
-% defines all the citations from the bibliography above.
-%
-% The same technique allows the supplementary material to have 2 more bibliographies,
-% with consecutive numbering.
-}
-
-%\label{LastPage}
-%save Latex register    {linenumber}
-%\makeatletter
-%\immediate\write\@auxout{\string\def\string\MainPaperLastLine{\arabic{linenumber}}}
-%\makeatother
-\end{document}
-
-\def \allInOne {define this so we don't include header in supplemental material}
-\input paper-seb-supplementary-material.tex
-\end{document}
-
-%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
-%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}
-
-\end{document}
+\color{black}

### 2023-02-20 response-to-referees/response 30 Jan 2022.tex
Fantastic corrections noticed by Konrad Hinsen

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}\usepackage{url}\usepackage{xcolor}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcount\noten
\noten=1
\long\def\note#1{

\color{red}{\begin{tabular}{r|p{4.7in}}
\hphantom{\hskip .4in}\ifnum \noten<10 \hphantom{0}\fi{\sf \the\noten}& #1
\end{tabular}}\color{black}
\advance\noten by 1}

\title{COMPJ-2022-05-0325: Response to referees}
\author{Harold Thimbleby}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}

\noindent 
\color{red}
%\begin{quote}\emph{I have sent an email requesting a short extension please --- I've finished revising the main paper, but not yet finished the Supplemental Material (it has been Christmas and New Year~\ldots). I should finish early next week. I uploaded these notes on my revisions in case the computer system destroys the old submission if I do not meet the deadline.}\end{quote}

\noindent Thank you for this opportunity to respond to reviewers' comments. Almost all of my responses, detailed below, have resulted in helpful improvements to the paper.

I have written my responses to the reviewers' comments in red. If it helps in referring to them in future correspondence, I have numbered my responses.

In addition to responding to reviewers' comments, I have also edited the paper to improve its clarity, structure, and where possible to reduce its length. The writing is now tighter, and in particular the reviewers' repeated concern about ``opinion'' has been addressed by appropriate rewriting --- mostly replacing ``shoulds'' with ``coulds'' and other ways of making clear the what was an unqualified opinion is a suggestion for further work.

\color{black}

\vskip .5cm\hrule height 2pt \vskip .5cm

\noindent
Dear Professor Thimbleby,

COMPJ-2022-05-0325 -- Maturing the computational sciences

I am writing to you with regard to your paper COMPJ-2022-05-0325, which you submitted to The Computer Journal. It has now been examined by referees and by a member of the editorial board.

In view of their recommendations, I would appreciate it if you would revise your paper, carefully taking into account all the comments, and resubmit it by 03-Jan-2023 via the Manuscript Central website at \url{http://mc.manuscriptcentral.com/compj}

Please also make sure a DATA AVAILABILITY STATEMENT is included in the manuscript. Please check\\\url{https://academic.oup.com/comjnl/pages/Manuscript_Preparation_Submission} and \url{https://academic.oup.com/journals/pages/authors/preparing_your_manuscript/research-data-policy#data2}

To submit your revision, please go to *** PLEASE NOTE: This is a two-step process. After clicking on the link, you will be directed to a webpage to confirm. ***

\url{https://mc.manuscriptcentral.com/compj?URL_MASK=}\\\url{ae7f1f867b624f72b0d553807c8d9146}

Please also submit a list of the changes you have made in response to each of the referees' comments. Your paper will then be re-examined and a decision will be made about whether it may be published.

Thank you very much.

Best regards,

Deputy Editor

The Computer Journal

\newpage

\section{REVIEWERS' COMMENTS:}

\subsection{Reviewer: 1}

The paper has a misleading title -- I think it would be better if it focussed on software and people -- or maybe on improving reliability. \note{Thanks to this concern, I have improved and changed the title to ``Improving science that uses code.''}

Having a vision for the future is good, but it needs to be matched with a set of practical changes to get there. I don't see that here, just a vision of an alternative world.

I think this paper is too long and too complex.  \note{As well as structural improvements, I have also added a short table of contents, which makes the structure of the much paper much clearer than before, also helping address this reviewer's concern. In fact, I used the new table of contents to guide some streamlining throughout the paper!}

Some of it is opinion, some of it is observation, and some of it is advertising. And these parts are interspersed in the paper.

In table 2, GitHub is listed as both a code repository and a data repository, which is confusing.  It might be better to talk about the different meanings of repository, such as in the git sense vs in the library/archive sense. \note{GitHub can be used as a repository for anything (such as \emph{The Turing Way: A Handbook for Reproducible Data Science}, which is a massive project with 419 authors collaborating. I have therefore not changed table 2.}

There is likely a correlation between good software engineering practices and good code quality, but I am less sure that there is a causation, or that good quality code cannot be written without good software engineering practices.

Re ``A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to. -- this reminds me of \url{https://doi.org/10.12688/f1000research.25561.1} \note{I had thought of citing this f1000 paper, but it reports on a meeting duplicating the earlier work of TOP, which I do cite; furthermore this f1000 paper  just makes a list of questions --- unlike the TOP papers which have concrete proposals.}

There is likely a difference between how statistical practices are described concisely in a paper (page 3 line 54) and how software can described, which is probably better done through the software itself than through text about the software. \note{Why? Just as a statistical transformation might be described by name in wording like ``Bonferroni adjusted estimated mean difference~\ldots'' as my paper says, an algorithm can be described by precise wording like ``we use Knuth's (1977) generalization of Dijkstra's (1959) algorithm for directed hypergraphs~\ldots''}

Re the comments on Ferguson in the intro to Section 3, this is correct, but it's not the end of the story. An effort was made to improve the code, which I believe was successful. \note{This effort was (and is) cited and discussed in the paper.}

I don't think this changes the argument of this paper, but the paper should acknowledge that we do know how to fix these things, we just don't do it until we have to (due to limited resources, no culture that requires it, incentives, etc.) \note{I disagree. The Ferguson case involved fixing a particular case of very bad code after the event of its unravelling and Ferguson's admissions (as cited); it is not fixing the problem the paper addresses, which is fixing the persistent widespread use of bad code in science.}

\note{The reviewer says ``know how to fix these things, we just don't do it until we have to'' so we agree the issue is to create a route from the problem to a solution; I think my paper provides evidence that the problem is real and widespread, and lays out many ways to start doing this.}

I don't agree that C is not a dependable language, nor that it is not portable. The issues raised here are not unique to C, but are common to almost all languages in usage in computational science. \note{The whole industry-standard MISRA C approach is based on the fact that the community recognizes that C is not dependable, and MISRA and it's work is already cited. The reason for mentioning C in particular is that it was the language chosen by Ferguson, as explained in the case study; however I agree the unreliability issues are common across many languages, and I now make this point clearer.}

These statement are philosophical in this context. \note{Hmm. Only if you believe that C is dependable, a view which I argue against!}

The RAP* discussion seems too detailed for this paper, and the supplementary material is too long -- this could be published separately (or as a pre-print) and referenced here. \note{The length of the RAP discussion in the paper main body seems right to me, and is supported by Reviewer 2's comments. (Please  note that I changed my term RAP* to RAP+ as I feel this much better indicates that RAP+ is an extension of the original RAP than RAP*.)}

\note{I agree that the supplemental material is long, but it will be online so I am not sure that this is a problem. The  supplemental material is divided into several topics, and includes a table of contents, so readers can navigate it easily.}

Re Software Engineering Boards, what about other boards for other types of work: perhaps we should have data collection boards, statistical review boards, reproducibility boards, etc.  But I don't think creating new collections of gatekeepers is going to be either practical or useful.  
\note{I hadn't thought of it as gatekeeping, thanks for raising the concern. I've now added a paragraph and a citation to an article in \emph{Nature\/} to explain that the idea is \emph{not\/} gatekeeping.}

Better standards and better education make sense to me, but more enforcers do not. \note{There is no sense, and no sense expressed in the paper, in which Software Engineering Boards are gatekeepers or enforcers. The paper makes an explicit analogy with Ethics Boards, which are widely accepted and work well. I don't see any problems, then, nor any ``thin end of a wedge'' problem the reviewer characterizes here.}

I don't think the comment about methodological statements at the start of 5.b is correct. Most venues in which I publish do not require all or even many of these statements from authors. \note{I changed ``most journals'' to ``Many science journals'' (although I know of none that do not require some such statements); in addition, I have cross referenced to the paper's discussion of PRISMA, which itself is a methodology that, if used, authors are expected to declare and comply with. PLOS ONE is an example of a major journal that makes a list of similar methodologies and when the journal expects compliance; I have now made a note of this point in the paper.}

Much of this section and 5.c also seem to follow the advertising of RAP* with statements that not only is it good, but it should be required. This is again opinion, and not justified. Would this help? maybe. Is it the best solution? I have no idea. \note{I disagree that the benefits of RAP and RAP* are opinion; I provide reasoned arguments, and examples of the problems RAP and RAP* avoid.}

\note{I note that RAP and RAP+ (its new name) are very good solutions, and I am of course explicitly drawing on the existing very favorable literature on RAP\@. There may be better solutions available in the future, but until then they are the best known solutions. Furthermore, ``the best is the enemy of the good'' --- it is routine to publish a good idea, as this paper proposes, and the scientific community can then seek ways to refine the idea. The point is, a good idea has to be published first.}

The second paragraph in 5.e is not at all correct in my opinion. I think this is an unrealistic view of science. \note{I briefly considered citing Einstein, who commented on this point, and others in my support, but I've deleted the comments in the paper, as the paper's core message is not in fact affected either way.}

The acknowledgements statement isn't grammatically correct. \note{Thanks for spotting this! Problem fixed.}

After reading the paper, the only thing I feel that I have learned is that the author has strong opinions about how bad things are and how the world should work that would be better. But I'm not convinced that this is the correct analysis and solution, or even if it is, how we would get to this future world from today's world in practice. \note{The new Background section I have added in my revision makes it clear that my opinions are closely aligned to, and often summarize, substantial work by others.}

\subsection{Reviewer: 2}

This manuscript on the state of computational science today is unusual in several respects. It is mostly an opinion paper, advocating for more rigor in the use of computational methods in scientific research. 
\note{As mentioned in my comments to Reviewer 1, my opinions are rigorously justified, and traceable back to peer reviewed literature. It is of course an opinion that scientific research should be more rigorous, but I am unaware of significant dissent to this opinion. (I admire Paul Feyerabend, e.g., his \emph{Against Method}, but the paper does not need to get into the philosophy of science.)}

In doing so, it provides a rather exhaustive list of arguments, to the point of almost being a PhD-level course on the subject. \note{I am aware --- and I hope --- that scientists, not just computer scientists and software engineers, who can benefit from my analysis will read the paper. I therefore felt it important to write clearly, and to provide as much background and direct information as possible. In particular, the Supplement provides a lot of useful material and pointers to further reading.}

Finally, it contributes an analysis of code-and-data sharing in the literature to the arguments, making it in part a research paper. The writing style is much more colloquial than a typical scientific article, making the manuscript lengthy but also very readable. Whether or not these particularities are acceptable in The Computer Journal is a matter of editorial policy, which is not for me to judge. 
\note{Thank you for recognizing the readability of my paper. I tried hard to make the paper accessible, as I hope many scientists, not just computer scientists/non-programmers, will read the paper.}

My following review mostly treats the manuscript as an opinion paper.

Opinion papers being a matter of opinion, I'll do my best to judge the pertinence of the argument independently of whether I agre with them. But I'll state my prejudices right away: I do agree with most of the observations and conclusions in this paper. \note{Thanks!}

The most important omission in my opinion is the lack of context. Many past and present efforts are aligned with the author's objectives and have made and sometimes implemented similar propositions, but the manuscript does not mention them. \note{I did, but the mentions were scattered around in the paper. I have now introduced a focused Background section (with additional references to substantiate a more powerful argument) early in the paper collecting everything together to address this helpful point.}

The publication of scientific software in source code form has been advocated as early as 1969 by K.V. Roberts (\url{https://doi.org/10.1016/0010-4655(69)90011-3}), using some of the same arguments that are made in the present manuscript. The next call for publishing software and data I am aware of is by Jon Claerbout in 1992 (\url{https://doi.org/10.1190/1.1822162}). \note{The reviewer makes some pertinent points here, and I have therefore added material to the Background section to provide an overview of the relevant literature. In particular, I note that concern for publishing code goes back even further than the reviewer mentions; I cite work as far back as 1958, in fact to \emph{CACM\/}'s first issue of its very first volume.}

Since then, the pace at which similar calls for more rigorous treatment of software and computation have been made has accelerated steadily. More importantly, there are many initiatives that implement parts of the author's recommendations, and it would be more helpful to discuss how their impact and utility can be improved than to pretend they don't exist and propose new initiatives to be started from scratch. \note{Yes, it was an oversight that there was no stand-alone, clear section in the paper discussing prior work: it was scattered around in various citations. I have therefore now added a new section, Background.}

Some examples:

\begin{itemize}\raggedright\item The Software Sustainibility [sic] Institute (briefly mentioned in the supplementary material) and CODECHECK in the UK, and CASCaD in France, resemble the author's SEBs in many ways. More generally, there is an ongoing movement in many countries for establishing the ``research software engineer'' as a recognized role in scientific research, which shows that the need for better software engineering is increasingly recognized. \note{Rather than cite such an almost unlimited list of initiatives, I discuss some large studies and the key TOP initiative.}
\item  New on-line journals such as The Journal of Open Source Software or ReScience practice reviewing of scientific code at various levels (and find that this is not as easy to do as it may seem). \note{I hadn't heard of \emph{ReScience C\/} (and \emph{ReScience X\/}) --- thanks. I now cite them (in the new Related work section) and what they are achieving.}
\item  Various courses, e.g. by Software Carpentry and Data Carpentry, and also several on-line courses, teach what the author calls RAP (a name I have never seen before) to an audience of mostly PhD students. \note{The training courses I am aware of (including all those I searched for and found on the internet) are very low level introductions to coding, and really inadequate for coding to support reliable peer reviewed research.}
\end{itemize}

I find rather few completely new arguments and proposals in the present manuscript, but it has the merit of being exhaustive, detailed, and combining advocacy with practical advice.

In the following, I will concentrate on the new aspects.

An important original contribution is the analogy between the introduction of computation and statistics into mainstream research. It deserves publication on its own because of its pertinence. The analogy is very fitting, and helps both in understanding the inadequacy of today's scientific code management and in improving the situation. Moreover, the rigorification of statistics has well advanced to the point that most scientists now realize its benefits. That's a good motivation to launch a similar campaign for more rigor in computation. \note{Thanks for these encouraging comments. The important point that scientists now recognize the value of rigorous statistics motivates considering rigor in computation is now mentioned in the introduction (section 1).}

Another original contribution of the proposal of Software Engineering Boards, but I have to say that I don't find the arguments very convincing. I do agree about the necessity of introducing software engineering practices into computational research, and to check publications for respecting best practices in the field. But the implication of software engineers can take many forms (as part of research teams, as experts in peer review, \ldots) and I don't find the arguments for the specific institution ``SEB'' very compelling. The analogy with ethics boards is weak, as the numerous differences listed by the author suggest. \note{The point of the analogy is that ethics boards \emph{do\/} work and are recognized as available --- if not required --- in almost all research institutions. Ethics Boards are routine, and are not onerous as this reviewer suggests. I added a few extra references here, including an analogous proposal for methodology boards --- in a recent \emph{Nature\/} paper, which also explores their pros and cons.}

A more general comment about the author's discussions of software engineering in science is the lack of any reference to the particularities of scientific software, compared to software in business or industrial contexts, which is the focus of most software engineers today. And even within the category of scientific software, there are vast differences. An epidemiology simulation (software as a reliable tool) meant to inform public policies must be treated differently from a prototype implementation of a new data analysis technique (software as an executable scientific method). Both need more rigor, but not in the same place. \note{I do think this is an obvious point. For example, scientific coding is obviously different to game coding, e-commerce, and so on. Discussing the varieties of code seems to me to be beyond the scope of the paper.}

As an example, the author heavily insists (in the supplementary material) on the use of formal methods. They are indeed the appropriate tools in situations where the conformance of an efficient implementation with a known-to-be-good specification is the bottleneck in establishing correctness. But often it is the specification itself that is incomplete, informal, or incorrect. A common situation in computational research is a rapid development cycle in which scientific questions, specifications, and implementations are frequently updated. It is in my opinion an open question if formal methods are helpful in this setting. They add rigor to one step of the cycle, but at such a high cost (with the current state of the art at least) that other steps risk being neglected. Illustration: a piece of code formally proven correct to a rather deep level: \url{https://doi.org/10.1016/j.camwa.2014.06.004}. Impressive work, but also a Herculean effort, for a single relatively small algorithm. \note{Indeed! I note that the paper this reviewer cites is 25 tight technical pages, and that its authors admit that their approach is beyond many coders. I have therefore now clarified (throughout) that a formal methods \emph{mindset\/} is required, not a total formal proof of code as in this paper. For example, as now explained particularly in the Supplement, you could require explicit invariants, assertions and other formal ideas in code: this is a formal mindset in action, but avoids ``the best is the enemy of the good'' the reviewer seems to fear.}

This illustrates a more general problem with the author's recommendations (in the supplementary material): there is no clear path for getting from today's research processes to more rigorous processes that include all of the author's ideas. Implementing all of them is such a huge effort that at best it will suffer the same fate as climate change policies: everybody agrees it should happen, but the amount of work requires is so paralyzing that nothing happens at all. A grand view, such as outlined in this manuscript, needs to be complemented by a list of small initial steps that most scientists recognize as achievable. \note{I think here the reviewer has fallen  for the perfect-is-the-enemy-of-the-good interpretation! I hope my rewording throughout dispels this problem.}

Another interesting original contribution is the RAP* method. \note{I've renamed it RAP+, as this is more suggestive of an extension to RAP itself, which the name RAP* wasn't.}

It is not completely original in the sense that many people have practiced variants of it in the past, but I haven't seen it named before, nor discussed to the extent presented in the manuscript. The author's analysis of the literature also serves as a case study for RAP*. The discussion of RAP* is a welcome transition from ``how to save research from computers'' (the main topic of the manuscript) to ``how to improve research using computers''. I hope to see more work in this direction in the future. \note{Thanks for these comments. Of course, RAP* is on the route towards what I envisage, and therefore these welcome comments contradict the reviewer's  worries expressed in their previous paragraph!}

Finally, in case the editors of The Computer Journal decide that this lengthy opinion-paper-plus-course does not fit their editorial policies, a few suggestions for shorter articles extracted from this material that I think would be received favorably by the scientific community:

\note{I put all the helpful ``course'' material as reference in the Supplement, which will be online rather than part of the paper itself.}

\note{I am not really clear why the reviewer focuses apparently negatively on the paper being opinion. Surely, Tony Hoare's paper on quicksort was just opinion, and wasn't even the best way of sorting under many assumptions --- but that doesn't stop it being a brilliant and now classic paper. So far as I can see, all my opinions are argued, justified or readily traceable back to the literature. I'd be happy to fix any specific unsupported opinion of concern but the review is not specific.}

\begin{enumerate}\raggedright\item
A much shorter and focused opinion paper on the topic: ``We made statistics in research more rigorous, now let's do the same for the use of software''.
\note{This is a good idea, but of course computation can do a lot more than statistics, so I don't think such a paper, properly done, would be much different from my current offering --- especially as there seems to be very little literature making the same points one could thus build on much more concisely.}
\item  A review of software engineering techniques, explaining their expected benefits for computational science. The main audience would be research software engineers. \note{I put all of this material in the Supplement, which I imagine will be online so its length is not material.}
\item  A tutorial on the RAP* method, using the data analysis from the present manuscript as a case study. \note{This is already provided in the supplement.}
\end{enumerate}

Konrad Hinsen, CNRS, France
(Note to editor: I have decided to sign all my reviews. Please don't remove my signature.)

\note{I very much appreciate your disclosure. Thanks for all your work. Are you happy being named in the paper's Acknowledgements?}

\subsection{Reviewer: 3}

David Spiegelhalter's name is misspelt ``Speigelhalter'' throughout. The first time it is misspelt ``Speigelhater''. \note{Oh no! Thanks for this correction. Spelling now fixed throughout.}

\section{EDITORIAL BOARD RECOMMENDATION:}

Associate Editor

Comments to the Author:

Please address the comments raised by the reviewers, and list what changes you have or haven't made with some justifications.

\vskip 1cm\hrule

\section{REFERENCES:}

General. Please construct references in the following style:

\begin{enumerate}\raggedright\item The references should appear in the text in numerical order, as should the reference list. \note{Now done.}
\item  Where there are more than 10 contributing authors please list only the first author and use et al. \note{Now done.}
\item  The following format for references should be followed: \note{Now done.}
\end{enumerate}

Journal article:
Galton, A. (1992) Logic as a formal method. Comp. J., 35, 431-440.
(author name, year, title of paper, title of journal, volume number, page numbers)

Book:
Hogger, C. (1990) Essentials of Logic Programming. Clarendon Press, Oxford.
(author name, year, title of book, publisher, publisher location)

Chapter:
Harel, D. (1984) Dynamic logic. In Gabbay, D. and Guenther, F. (eds), Handbook of Philosophical Logic. D. Reidel, Dordrecht.
(author name, year, title of chapter, editors names, title of book, publisher, publisher location)

Conference:
Crochemore, M. and Verin, R. (1997) Direct construction of compact directed acyclic word graphs. Proceedings of CPM 97, Cambridge, MA, 12-14 August, pp.~192-211. Springer-Verlag, Berlin.
(author name, year, title of conference, place of conference, date of conference, page numbers in proceedings, publisher, publisher location)

Technical documents:
ITU-T. Z.500 (1997) Framework on formal methods in conformance testing. International Telecommunications Union, Geneva, Switzerland.
(document number, year, document title, publisher, publisher location)

Could you please check your references carefully (in particular, proceedings: location of conference, date of conference, publisher name and location\ldots) and make sure that they comply with the above format?

\color{red}
\section{SUMMARY OF CHANGES}
In addition to the changes in response to the reviewers' specific comments as listed above, I have uploaded a \texttt{changes.pdf} file which is an automatically generated difference listing of the changes I have made. Text in blue is new; text in red is deleted.

The diff process I used (using \texttt{latexdiff}) doesn't read included files and has other irritating bugs, so, for instance, changes to data (including changes in tables) are not shown, and there are various typesetting errors. Regardless of the details, it shows how almost all of the paper's text has been substantially improved, usefully including showing deleted text that I or the referees considered unnecessary.

The added table of contents I hope shows the reviewers the revised paper is now much clearer and more clearly structured; I think the table of contents will be helpful for journal readers too, but it does not need to be included in the final paper if it is not considered to fall within the journal's accepted style.

The supplementary material has minor corrections and simplifications throughout, but no substantial changes comparable to those made in the main paper.

\end{document}

### 2023-02-20 generated/allGitRepos.sh

@@ -1,36 +1,0 @@
-# run this shell script in directory models
-rm -rf "git-MetricSelectionFramework"
-git clone "http://github.com/ChristophKanzler/MetricSelectionFramework" "git-MetricSelectionFramework"
-
-rm -rf "git-PENet"
-git clone "http://github.com/marshuang80/PENet" "git-PENet"
-
-rm -rf "git-PostoperativeOutcomes_RiskNet"
-git clone "http://github.com/cklee219/PostoperativeOutcomes_RiskNet" "git-PostoperativeOutcomes_RiskNet"
-
-rm -rf "git-philter-ucsf"
-git clone "http://github.com/BCHSI/philter-ucsf" "git-philter-ucsf"
-
-rm -rf "git-AI-CDSS-Cardiovascular-Silo"
-git clone "http://github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo" "git-AI-CDSS-Cardiovascular-Silo"
-
-rm -rf "git-SiameseChange"
-git clone "http://github.com/QTIM-Lab/SiameseChange" "git-SiameseChange"
-
-rm -rf "git-lactModel"
-git clone "http://github.com/SolangeD/lactModel" "git-lactModel"
-
-rm -rf "git-LRM"
-git clone "http://github.com/UT-NSG/LRM" "git-LRM"
-
-rm -rf "git-manifold-ga"
-git clone "http://github.com/ki-analysis/manifold-ga" "git-manifold-ga"
-
-rm -rf "git-blast-ct"
-git clone "http://github.com/biomedia-mira/blast-ct" "git-blast-ct"
-
-( echo "% date generated by running generated/allGitRepos.sh downloading Git repos";
-echo \\\\def\\\\clonedate {\\\\ignorespaces `date "+%e %B %Y"`}
-echo \\\\def\\\\cloneyear {`date "+%Y"`}
-echo \\\\def\\\\clonemonth {`date "+%m"`}
-) > ../generated/clone-date.tex

### 2023-02-20 programs/data.js

@@ -1065,7 +1065,7 @@ for (var i = 0; i < data.length; i++) {
     for (var f in descriptionsOfFields)
         s += csvise(data[i][f], f == lastflag);
 }
-saveFile("generated/data.csv", s, "All data from data.js expressed in CSV for convenience\n");
+saveFile("generated/data.csv", s, "All data from data.js expressed in CSV for convenience");
 
 s = "# run this shell script in directory models\n";
 var spagelengths = "";
@@ -1093,7 +1093,7 @@ s += "( echo \"% date generated by running generated/allGitRepos.sh downloading 
     "echo \\\\\\\\def\\\\\\\\clonemonth {`date \"+%m\"`}\n" +
     ") > ../generated/clone-date.tex\n";
 
-saveFile("generated/allGitRepos.sh", s, "Shell script to download all available " + n + " GitHub repositories \n         generated/clone-date.tex - Running the download script also saves the download dates of GitHub clones")
+saveFile("generated/allGitRepos.sh", s, "Shell script to download all available " + n + " GitHub repositories \n       generated/clone-date.tex - Running the download script also saves the download dates of GitHub clones")
     //console.log(spagelengths);
 spagelengths += "\\newcount \\gitPages \\gitPages=" + gitpagelengths + "\n";
 spagelengths += "\\newcount \\totalPages \\totalPages=" + totalpagelengths + "\n";

### 2023-02-20 programs/make-metadata.sh

@@ -2,10 +2,10 @@ echo % shell script "(make-metadata.sh)" generated LaTeX variable values countin
 
 # how many errors can data.js find?
 echo \\\\newcount \\\\JSONerrorCount >> generated/metadata.tex
-echo \\\\JSONerrorCount = `grep "error(" data.js|wc -l` >> generated/metadata.tex
+echo \\\\JSONerrorCount = `grep "error(" programs/data.js|wc -l` >> generated/metadata.tex
 
 # put count of variables (number registers plus macro definitions) to \JSONvariableCount
 echo \\\\newcount \\\\dataVariableCount >> generated/metadata.tex
-echo \\\\dataVariableCount = `egrep "(newcount)|(\\\\\\\\def)" generated-*|wc -l` >> generated/metadata.tex
+echo \\\\dataVariableCount = `egrep "(newcount)|(\\\\\\\\def)" generated/*|wc -l` >> generated/metadata.tex
 
 

### 2023-02-20 generated/assessments.tex
Various corrections inspired by Konrad Hinsen

--- 2022-01-27 paper-seb-supplementary-material.tex
@@
-\PassOptionsToPackage{hyphens}{url}
-\documentclass{article}
-\renewcommand\thesubsection{\thesection.\alph{subsection}}
-%\usepackage{hyperref}
-%\usepackage{refcount}
-\usepackage{url}
-\usepackage{doi}
-\usepackage{longtable}
-\usepackage{color}
-\usepackage{amsmath,amssymb}
-\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
-\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
-
-\input paper-seb-macros.tex
-
-\makeatletter
-      \renewcommand{\thefootnote}{\arabic{footnote}}
-      {\let\oldbibcite=\bibcite
-      % \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
-      \newcount\mainNumberOfReferences \mainNumberOfReferences=0
-      \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux
-      %\inputifexists{paper-seb-main.aux}{You need to run Latex on paper-seb-main.tex because we need its aux file}
-      }
-\makeatother
-
-\begin{document}
-
-\title{\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex}
-
-\author{Harold Thimbleby, \texttt{harold@thimbleby.net}}
-
-%\orcid{0000-0003-2222-4243}
-
-% To make various LaTeX processors do the right thing with page size.
-\def\pprw{8.5in}
-\def\pprh{11in}
-\special{papersize=\pprw,\pprh}
-\setlength{\paperwidth}{\pprw}
-\setlength{\paperheight}{\pprh}
-\setlength{\pdfpagewidth}{\pprw}
-\setlength{\pdfpageheight}{\pprh}
-
-\def\href#1{#1}
-
-\maketitle
-
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofile
-\immediate\openout\infofile=generated-info-for-main.tex
-
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
-
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\immediate\write\infofile{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\string\newlabel{#1}{{\arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
-
-% start supplementary material's page counters to continue on from the paper's page numbering
-%\setcounterpageref{page}{LastPage}
-%\addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
-
-\setcounter{section}{0}
-\def\n#1{{\sf\bfseries ({#1})}}
-
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
-
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
-
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
-
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
-
-%\newpage
-
-\makeatletter
-\let \bslash = \@backslashchar
-\makeatother
-
-%\renewcommand{\thesubsection}{\alph{subsection}}
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
-
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\input generated-supplementary.toc
-%\inputifexists{generated-supplementary.toc}{you need to Latex supplementary-material.tex again}
-}
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated-supplementary.toc
-
-\newpage
-\mysection{Further issues for Software Engineering Boards (SEBs)}
-
-\highlightmysubsection{Brief definition}
-Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
-
-Further details of the SEB proposal is in the main paper.
-Here we raise further issues for SEBs (additional to those covered in the main paper's introduction to SEBs), potential limitations and possible responses that can be addressed over time:
-
-\begin{enumerate}\raggedright
-\item 
-Until there are national qualifications, nobody --- certainly nobody without professional training in software --- really knows just how bad (or good) they are at software engineering.
-
-\item
-When code is taken seriously, concerns may be raised on programmers' contributions to research, intellectual property rights, and co-authoring \cite{vancouver}.
-Software engineering is a hard, creative discipline, and getting epidemiological (and other scientific) models to work is generally a significant challenge, on a par with the setting up and exploring the mathematical models themselves.
-Often software engineers will need to explore boundary cases of models, and this typically involves hard technical mathematics \cite{hamming}.
-Often the software engineers will be solving entirely new problems and contributing to the research.
-How this is handled needs exploring.
-How software engineers are appropriately credited and cited for their contributions also needs exploring.
-
-\item 
-SEBs require policies on professional issues such as membership, transparency, and accountability.
-
-\item
-There should be a clear separation between the SEB members' activities as part of the Board, and their other activities, including professional advice, code development, or training (which is likely to be in demand from the same people who require formal approvals from the SEBs).
-
-\item
-Professional Engineering Bodies have a central role to play in professionalism, ranging from education and accreditation to providing professional structures and policies for SEBs.
-For example, should and if so how should the programming skills taught to computational scientists (epidemiologists, computational biologists, economists, computational chemists,~\ldots\@) be accredited?
-
-\item
-In the main paper, SEBs are viewed as a constructive contribution to good science, specifically helping improve the quality of epidemiological modeling.
-More generally, SEBs will have wider roles, for instance in overseeing software subject to medical device regulation \cite{fixit}.
-
-\item
-SEBs may fruitfully collaborate with other engineering disciplines to share and develop best practice.
-For example, engineers in other domains (e.g., civil engineers) routinely sign off projects, yet, on the other hand, they often overlook the quality of software engineering their projects implicitly rely on  --- for the same reasons as the scientific work discussed in this paper overlooks the dependence on quality software.
-
-\item
-Clearly, at least while this paper's concepts are tested and mature, SEBs will need to collaborate closely with research organizations, journals, and funding agencies in order to develop incremental developments to policies and processes that will be most effective, and which can be introduced most productively over time to the scientific community at large.
-Funding agencies may wish to support such strategic work, as they have previously funded one-off projects such as \cite{cosmos}.
-\end{enumerate}
-There are other ideas to help make SEBs work, but it is clear they are part of the solution.
-We must not let perfection be the enemy of the good.
-SEBs don't need to be perfect on day one, but they do need to get going in some shape or form to start making their vital contribution.
-
-\mysubsection{Relationships of SEBs to Ethics Boards}
-
-\begin{enumerate}\raggedright
-\item 
-Although SEBs may start with a checklist approach, like Ethics Boards generally do, it cannot be assumed that people approaching SEBs know enough about software engineering to perform adequate software assessments when there is any risk (as there is in public policy, medical apps, and so on).
-SEBs may also provide mentoring and training.
-
-\item
-Unlike Ethics Boards, which provide hands-off oversight, SEBs should provide professional advice, perhaps providing training or actually helping hands-on develop appropriately reliable software.
-During a pandemic SEBs would be very willing to do this, but in the long run it is not sustainable as voluntary labour, so all research, particularly medical research, should include support for professional software engineering.
-
-\item 
-Ethics Boards typically require researchers to fill in forms and provide details, which is a feasible approach as researchers know if they are doing experiments on children, for instance, so the forms are relatively easy to fill in (if often quite tedious).
-On the other hand, few healthcare and medical researchers understand software and programming, so they are \emph{not\/} able to fill in useful software forms on their own.
-SEBs need to know how well engineered the software really is, not how good its developers \emph{think\/} it is.
-As typical programs are enormous, SEBs are either going to need resources to evaluate programs, or they will need to supervise independent bodies that can do it for them.
-
-\item
-SEBs should have a two-way collaboration with Ethics Boards.
-
-\begin{itemize}
-\item SEBs have to deal with ethical concerns, and how they may be implemented in code.
-One of the papers \cite{ethics-paper} in the survey (discussed later in this \supplement) is a case in point, as is the growing cross-fertilization between AI and ethics \citeeg{ai-ethics}.
-
-\item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
-\end{itemize}
-
-\item
-Like some Ethics Boards, SEBs might become, or be perceived as becoming, onerous and heavy handed --- as if the Board is not interested in ethics but only in following a bureaucratic pathway.
-It seems essential, then, that SEBs have (and perhaps are chaired by) experienced, practicing, professional software engineers to avoid this problem.
-\end{enumerate}
-
-\changemysubsection{SEBs are necessary but not sufficient}
-The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
-
-SEBs address the problems identified at the laboratory end of doing science; they do not address the processes of review, editorial control, and action based on claimed results.
-As shown in the review of \plural{\dataN}{paper}, only some journals have code policies, and the policies are not enforced.
-In other words, improving the professionalization of software engineering has to proceed from doing science, which the paper covers, to the downstream issues of review and publication.
-SEBs may work with journals, funding agencies and even international standards agencies to improve broader awareness of professional software engineering, but this is a topic the present paper has not addressed.
-It needs doing.
-
-\mysection{Software engineering best practice}
-\label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
-
-This \supplement\ provides more explanations and justification for following standard software engineering practices that support reliable modeling, reliable research, and, most generally, reliable science.
-
-The reader is referred to standard textbooks for more information \citeeg{sommerville,knight}, as well as to specialized texts that are more specifically addressed to software engineering in science \citeeg{cosmos}.
-
-The book \emph{Why Programs Fail\/} \cite{wpf} is a very good practical guide to developing better code, and will be found very accessible.
-Humphrey \cite{humphrey} outlines a thorough discipline for anyone wanting to become a good programmer.
-Improvement is such an important activity, Humphrey has also published a book to persuade managers of the benefits \cite{managers}.
-Further suggestions for background reading can be found throughout this section.
-
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent\input generated-minicontents.tex
-%\inputifexists{generated-minicontents.tex}{You need to run Latex again on supplementary-material.tex to generate the minicontents file}
-}
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated-minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
-It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
-On the other hand, if code is intended for life-critical applications, then it is worth putting more engineering effort into it.
-The first step of software engineering, then, is to assess the requirements, specifically the reliability requirements of the code that is going to be produced.
-
-In practice, requirements and expectations change.
-Early experimental code, developed informally, may well be built on later to support models intended to inform public policy, for instance.
-Unfortunately, prototypes may impress project leaders who then want to rush into production software because, it seems, ``it obviously works.'' Fortunately, best practice software engineering can be adopted at any stage, particularly by using \emph{reverse engineering}.
-In reverse engineering, one carefully works out (generally partly automatically) what has already been implemented.
-This specification, carefully reviewed, is then used as the basis for a more rigorous software engineering process that implements a more reliable version of the system.
-
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
-
-{In the physical world, to do something as simple as design and build a barbecue, you would need to use elementary mathematics to calculate how many bricks to buy.
-To build something more substantial, such as block of flats, you would need to use structural engineering (with certified structural engineers) to ensure the building was safe.
-Although programming lends itself to mathematical analysis, it is surprising that few programmers use explicit mathematics at all in the design and implementation of software.}
-
-{The type and use of mathematics used in software engineering is formal methods.
-Not using formal methods ensures the resulting code is unsafe and unreliable.
-Of particular relevance to scientific modeling: there must be an explicit use of formal methods to ensure  mathematical models (such as differential equations) are correctly implemented in code (and to understand the any limitations of doing so).}
-
-Formal methods require sophisticated knowledge of logic \cite{cbc}, as well as practical knowledge of using appropriate formal methods tools (Alloy, HOL, PVS, SPARK, and \emph{many\/} others).
-Using the right tools is essential for reliable programming, because the tools do quickly and reliably what, done by hand, would be slow and error-prone.
-Standard tools cover verification, static analysis of code, version control, documentation, and so on --- this paper explains why some of these activities are essential for reliable programming below.
-
-Crucially, tools are designed to catch common human errors that we are all prone to.
-Many tools are designed to avoid common human errors arising \emph{in the first place\/}; notably, the MISRA C toolset simply stops the developer using the most error-prone features of normal C, and hence improves the quality of programming with little effort.
-
-Many programming languages and programming environments have integrated features that support formal methods.
-For example, Hoare's triples \cite{hoare} (and formal thinking based on similar ideas) are readily supported by assertions, as either provided explicitly in a programming language or through a simple API\@.
-In particular, assertions readily support contracts, an important rigorous way of programming: assertions allow the program, the programming language, or tools (as the case may be) to automatically (and hence rigorously) check essential details of the program.
-Hoare's original 1969 paper \cite{hoare} is very strongly recommended because it is a classic paper that has stood the test of time; in the 1960s it was leading research, but now it can be read as an excellent introduction, given how the field of software engineering has advanced and become more specialized and sophisticated over the decades since.
-Hoare is also a very good writer.
-
-Formal methods have the huge advantage that they ``think differently'' and therefore help uncover design problems and bugs that can be found in no other way.
-Because formal methods are logical, mathematical theories (safety properties, and so forth) can be expressed and checked (often automatically); this provides a very high degree of insight into a program's details, and hence supports fault tolerance (e.g., redundancy).
-Ultimately, formal methods provides good reasons to believe the quality of the final code --- that it does what it is supposed to do.
-Unfortunately, because formal methods are mathematical, few programmers have experience of using them.
-Fortunately tools are widely available to help use formal methods very effectively.
-
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
-Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
-The great advantage of defensive programming is that it detects, and may be able to recover from, bugs that have been missed earlier in the development process (such as typos in the code).
-Defensive programming requires professional training to be used effectively, for example it is not widely known that some choices of programming language make defensive programming unnecessarily hard \cite{heedless}.
-
-A special case of defensive programming appropriate for pandemic modeling is mixing methods.
-Do not rely on one programming method, but mix methods (e.g., different numerical methods) to use and compare multiple approaches to the modeling.
-
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
-Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
-The fewer constraints a language imposes, the easier it \emph{seems\/} to be to program in, but the lack of constraints means the language cannot provide the checks stricter languages do.
-C, for instance, which is one of the languages widely used for modeling \cite{tweet,plos}, is not a good choice for a reliable programming language --- it has many intrinsic weaknesses that are well-known to professionals, but which frequently trap inexperienced programmers.
-(This is not the place for a review \cite{heedless} but Excel is even worse in this regard.) In particular, C is not a portable language (unless extreme care is taken), which means models will work differently on different types of computer.
-SPARK Ada is one example of a much more appropriate programming language to use.
-SPARK Ada also has the advantage that most Ada programmers are better qualified than most C programmers.
-
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
-It is appreciated that the models may change and be adapted as new data and insights become available.
-Changing models makes it even harder to ensure that they are correct, and thus emphasizes the relevance of the core message this paper: we have to find ways to make computer models more reliable, inspectable, and verifiable.
-Version control keeps a record of what code was used when, and enables reconstruction of earlier versions of code that has been used.
-Version control is supported by many tools (such as Git, Subversion, etc).
-
-If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
-Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
-
-{If results cannot be reproduced, has anything reliable been contributed?
-When a modeling paper presents results from a model, it is important to reproduce those results without using the same code.
-Better still, research should be reproduced without sharing libraries or APIs (for example, results from a model using R might be reproduced using Mathematica --- this is a case of $N$ (where, in this case, $N=2$) version Programming \cite{NVP}).
-Reproducing the same results relying on the same codebase tells you little.
-The more independent reproductions of results the greater the evidence for belief in the implications.}
-
-Clearly, with the transformations a program from avian flu in Thailand \cite{avianFluModel} to COVID-19 in the United States and in Great Britain \cite{ICmodel} taking place over many years, version control would have been very helpful to keep proper track of the changes.
-Note that professional version control repositories also provide secure off-site back up, ensuring the long-term access to the code and documentation --- this would avoid loss of \supplement\ problems, as occurred in \cite{flu-model}.
-
-Most version control systems would, in addition, enable open source methods so the code could be shared --- and reviewed --- by a wider community.
-Open source is not a panacea, however; it raises many trade-offs.
-Particularly for world-wide concerns like pandemic modeling, it increases diversity in the software developers, and fosters a diverse scientific collaboration.
-Open source can raise people's standards --- some countries \cite{excel1,excel2} are using Excel models to manage COVID-19, and open source projects properly implemented would help these people enormously.
-
-Open source raises important licensing and management questions to ensure the quality of contributions.
-A salutary open source case is NPM, where lawyers from a company called Kik triggered Azer Ko\c{c}ulu, that is, a \emph{single\/} programmer, to remove all his code from a repository.
-This caused problems to many thousands of JavaScript programmers worldwide who could no longer compile anything --- ironically, including Kik itself \cite{npm}.
-
-Critically in the case of epidemic modeling, open source democratizes the model development and interpretation, and enables properly-informed public debate.
-Note that many (if not most) successful open source projects have had a closed team of highly dedicated and directly employed developers \cite{open-source}.
-
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
-In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
-Rigorous testing includes methods like fault injection.
-Here, the idea is that if testing finds no bugs, that may be because the testing is not rigorous enough rather than that the program actually has no bugs.
-Fault injection inserts random bugs, and then testing gives statistical insights into the number of bugs in a program (depending on how many deliberate bugs it successfully finds).
-
-It is very tempting to test code while it is being built, save some or all of the code on a repository, but forget to check that the code has not changed out of recognition of the earlier tests --- tests should be saved so that modified code can easily be tested again.
-For example, if a test reveals a bug, the bug should be fixed \emph{and\/} the test needs to be re-run to check the fix worked (and did not introduce other bugs previously eliminated).
-
-It is important that code is saved and then downloaded to a clean site, confirmed it is consistent, and a new build made (preferably by an independent tester), which is then re-tested.
-If this procedure (or equivalent) is not followed, there is no assurance that the code made available with the paper is complete and works reliably.
-
-There are many other important testing methods \cite{sommerville,knight,NVP}.
-
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
-Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
-
-{For critical projects, such as for pandemic modeling, all documentation (including software) should be formally controlled, typically digitally signed and backed up in secure repositories.
-One would also expect a structured assurance case to be made, both to help the authors understand and complete their own reasoning and to help reviewers scrutinize it \cite{assurance-case}.
-}
-
-For purely scientific purposes, perhaps the most important form of internal is internal documentation: how to understand how and why the code works.
-This is different from developer documentation, which is how to \emph{use\/} the code in other programs.
-For example, code for solving a differential equation needs explaining --- what method does it use, what assumptions does it have?
-In contrast, the developer documentation for differentiation would say things like it solves ordinary differential equations with parameters $e$ for the function $f$ with the independent variable $x$ in the interval $[u,v]$, or whatever, but \emph{how\/} it solves equations is of little interest to the developer who just needs to use it.
-How code works --- internal documentation --- is essential for the epidemiologist, or more generally any scientist.
-An example of a simple SIR epidemiological model's internal documentation can be found at \url{http://www.harold.thimbleby.net/sir}
-
-There are many tools to help manage documentation (Javadoc, Doxygen, \ldots).
-Literate programming is one very effective way of documenting code, and has been used for very large programming projects \cite{LP}.
-Literate programming has also been used directly to help publish clearer and more rigorous papers based on code \cite{relit} --- a paper that also includes a wider review of the issues.
-
-Documentation should be supplemented by details of algorithms and proofs of correctness (or references to appropriate literature).
-All the documentation needs to be available to enable others to correctly download, install and correctly use a program --- and to enable them, should they wish, to repurpose it reliably for their own work.
-In addition, documentation requires specifications and, in turn, \emph{their\/} documentation.
-
-A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
-
-Without proper record keeping, code becomes almost impossible to maintain if programmers leave the project.
-Note that computer tools can make record keeping, laboratory books etc, trivial --- if they are used.
-
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
-\globalLabel{supplementary-material-makefiles}
-Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
-Often the programmers developing code know it so well they misjudge how easy it will be for anyone else to use it --- this is a very serious problem for the lone programmer (possibly working in another country) supporting a research team.
-Usability is especially important when programs are to be used by other researchers and by non-programmers, including epidemiologists.
-
-{In publishing science, an important class of user includes the scientists and others who will use or replicate the work described.
-When code used in research is non-trivial, it is essential that the process of successfully downloading code and configuring it to run is made as usable as possible.
-Typically so-called makefiles are provided, which are shell scripts or apps that run on the target machine, establish its hardware and other features, then automatically configure and compile the code to work on that machine.
-Makefiles typically also provide demo and test runs and other helpful features.
-Other approaches to improve usability are zip files, so every relevant file can be conveniently downloaded in one step, and using standard repositories, such as GitHub which allow new forks to be made, and so on.}
-
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
-Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
-The more code that is reused, the more likely many people will have contributed to improving it --- for example, reusing a standard database package will provide Atomicity, Consistency, Isolation, and Durability (so-called ACID properties) without any further work (nor even needing to understanding what useful guarantees these basic properties ensure).
-
-Note that reusing code assumes the originators of the code followed good software engineering practice --- particularly including good documentation; equally, if the code being developed building on it follows good software engineering practice, it too can be shared and further improved as it gets more exposure.
-Its quality improves through having scrutiny by the wider community, and in successful cases, leading to consensus on the best methods.
-Indeed, reuse, scrutiny, and consensus are the foundations of good science.
-
-Anticipating reuse during program development is called \emph{flexibility}, where various programming techniques can greatly enhance the ease and reliability of reuse \cite{flexibility}.
-
-A special case of reuse is to use software tools to help with software development.
-The tools (if appropriately chosen) have been carefully developed and widely tested.
-Tools enable software developers to avoid or solve complex programming problems (including maintenance) repeatedly and with ease.
-
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
-When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
-This way lies over-fitting, a problem familiar from statistics (and machine learning).
-Programs can be made over-complex and they can then do anything; an over-complex program may seem correct by accident.
-Instead, the hallmarks of good science are that of parsimony and simplicity; if a simple program can do what is needed it is more likely to be correct.
-A simpler program is easier to prove correct, easier to program, and easier to debug.
-A special case of needing simplicity is when fixing bugs: instead of fixing bugs one at a time, one should be fixing the \emph{reasons\/} why the bugs have happened.
-Generally, when bugs are fixed, programmers should determine \emph{why\/} the bugs occurred, and thence repair the program more strategically.
-
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
-
-To ensure adherence to best practice and, importantly, to avoid being unaware of relevant methodologies, professional software development projects adopt and adhere to relevant standards, such as ISO/IEC/IEEE 90003:2018 \cite{iso}.
-However, for safety-critical models or models of national policy significance, much stronger standards such as aviation software standards, such as RTCA DO-178C/EUROCAE ED-12C \cite{178C}, commonly called DO-178C, will be more appropriate.
-Publications should then cite the  standards to which their computer models comply.
-
-Note that medical device regulation, which has its own standards, is lagging behind professional software engineering practice, and currently provides no useful guidance for critical software development \cite{fixit}.
-
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
-
-As this long list illustrates, Software Engineering is a complex and wide-ranging subject.
-Software engineering cannot be done effectively by individuals working alone (for instance, code review is impossible for individuals to perform effectively), even without considering the complexities of the domain the code is intended for (in the present case, including pandemic modeling, mathematical modeling, public health policy, etc).
-Multidisciplinary teamwork is essential.
-
-Modern software is complex, and no one person can have the skills to understand all relevant aspects of all but the most trivial of programs.
-Furthermore, programming is a cognitively demanding task, and causes loss of situational awareness (that is, cognitive ``overload'' making one unable to track requirements beyond those thought to be directly related to the specific task in hand).
-The main solution to both problems is teamwork, to bring fresh insights, different mindsets and skills to the task.
-
-Peer review of code is an essential teamwork practice in reliable program development: \cite{peerReview,knight} it is easy to make programming mistakes that one is unaware of, and an independent peer review process is required to help identify such unnoticed errors.
-
-Almost all software will be used by other people, and user interface design is the field concerned with developing usable and effective software.
-A fundamental component of user interface design is working with users and user testing: without engaging users, developers are very likely to introduce quirks that make systems less usable (often less safe) than they should be.
-In short, users have to be brought into the software team too.
-
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in software engineering is continually evolving}{Continuous Professional Development (CPD)}
-
-As computing technology continues to develop rapidly --- especially as new programming tools and systems are introduced --- best practice in software engineering is also rapidly evolving.
-Continuous Professional Development (CPD) is essential.
-
-Ironically, the more organized CPD the more likely the content itself will lag behind.
-There is an argument for two-way links between universities (and other research organizations), research science developers, including enabling developers to undertake part-time research degrees.
-Research degrees teach not just current best-practice but also how to stay abreast of the relevant technologies and literature as it develops.
-
-The UK's Software Sustainability Institute is one initiative that is making important contributions \cite{ssi-report,ssi-url}, and its web site will no doubt remain timely and up to date in a way that this paper cannot.
-
-Note that CPD is not just a matter of learning current best practice, but a continual process as best practice itself continually evolves.
-{In software engineering, a current (as of 2021) initiative concerns reproducible code artifacts and badging papers to clearly show the approaches they take \cite{acm-artifacts}, and this will in due course have a direct impact on software engineering standards in other fields.}
-
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
-Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
-
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example \cite{example-numerical-error} was noticed to use literal numbers at too high a precision for the chosen language, where conformant implementations use IEEE 754 double precision 64-bit floating point.
-Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.}} Hamming \cite{hamming} is considered a classic, but there is a huge choice available.
-
-{For reasons of space, the present paper does not discuss the issues raised by AI, nor the many very important, non-trivial social and professional concerns, which have complex implications for software engineering practice, such as managing programming teams, data ethics, privacy, legal liability \cite{Schneier}, or software as a matter in law, as in disputes over model results or disputes over ownership of code \cite{electronic-evidence}.}
-
-\mysection{Code, data and publication}
-\immediate\write\infofile{\bslash newlabel{on-code-data-publication}{{\arabic{section}}{\thepage}{XXX}{YYYY}{ZZZZ}}}
-
-There is no fundamental difference between code and data, and no distinction that is relevant for scientific publication purposes.
-There is no distinction one can imagine that cannot easily, even accidentally, be circumvented.
-In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
-
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
-The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
-
-This section's discussion may sound like arcane and irrelevant pedantry, but these issues are at the very foundations of Computer Science.\footnote{{Many of the foundational issues were explored thoroughly by Christopher Strachey and others in the 1960s; Strachey's classic lectures are reprinted in an accessible 2000 publication \cite{strachey}.
-Being originally a very old paper this classic introduction is much easier to read than many more recent discussions of the foundations of Computer Science.}} If we ignore or misunderstand these basic things --- or overlook them in policies and procedures --- bugs are the inevitable (and confusing) consequence.
-
-The main paper points out that data is often embedded in code as ``magic numbers.'' Let's now explain how.
-
-A fragment of program code might say
-
-\begin{center}\texttt{x = 324.9+sin(theta*pi/2);}\end{center}
-
-This is clearly all source code, but the number 324.9 above is likely to be some sort of relevant data, though it might be a physical constant whose value does not depend at all on \emph{this\/} experiment.
-The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Historically, even $\pi$ was definitely an empirical value, but today it is a mathematical constant --- except in experiments to determine its value empirically.
-The point is, the distinctions between data, program and even mathematical constants are purely a matter of perspective.
-
-Unfortunately, there is data that is extremely easy to overlook (and therefore is very hard to manage).
-You may assume that the function \texttt{sin}, as used in the calculation example above, is the standard trigonometric function for calculating sines (and because of $\pi$, you assume it is taking radians as the parameter type) but almost all programming languages allow \texttt{sin} to be any function whatsoever.
-Confusingly, it is generally a different function when the code is run on a different computer.
-
-It is impossible to tell.
-More complex code will often have many facts ``hard wired'' into the code --- so in fact the code contains data.
-Code can even read in formulas from data and compile them to perform further calculations, and so on.
-Equally, data can control the flow of code.
-For example, data summarizing patients may include their gender, but the program processes males and females differently.
-Then data starts becoming like code.
-
-{Many computer programs blur the distinctions deliberately, to create virtual machines.
-Data is then run on the virtual machine as program.
-Many programs provide standard features to do this, such as LISP's and JavaScript's \texttt{eval} functions.
-Henderson's book \cite{henderson} builds an elegant Pascal program to run \emph{any\/} LISP program as data, and then shows that the LISP program can run itself running other programs, so it is now its own code and \emph{its\/} data --- despite being purely data to the Pascal program.
-There are numerous advantages to doing this, including: the Pascal program is not just reading data, but structured data that must conform to the rules of LISP; the LISP running itself runs faster than the original Pascal running LISP, even though the Pascal virtual machine is still doing it in the recursive case; LISP is a much more powerful language than Pascal, so a virtual machine can be used to escape the barriers of a limited implementation such as Pascal.
-In short, any distinctions between code and data are impossible to maintain.}
-
-In the present paper, we knowingly built on this blur between data and code.
-However, what we did was not unusual except in our explicit and rigorous approach to managing and summarizing data reliably in the paper.
-
-The paper and its \supplement\ are typeset in \LaTeX, a popular typesetting language.
-\LaTeX\ not only has text (as you are reading right now) but it also has code.
-For example, ``\LaTeX'' was typeset by running the code for a macro called \verb|\LaTeX|, which then calculated how to position the letters as they are wanted.
-When $\pi$ was written above, the code that generated what you read actually said \verb|$\pi$| --- so is this data that just says $\pi$ or is it code that tells the computer to change character sets from Latin to Greek, and then uses \verb|\pi| as a program variable name to select a particular glyph from the data about typesetting Greek characters?
-The distinctions are all a bit moot.
-In other words, the publication itself is data to a \LaTeX\ program, and within that data it includes further programs.
-Indeed, \LaTeX\ is run on a virtual machine, in exactly the same way that Henderson's LISP is, and doing so provides the same advantages.
-
-The data for this paper's survey was itself originally written as literal text in \LaTeX: it meant that \LaTeX\ could process it to produce a typeset table (as in the \supplement\ above).
-As the extent of the data grew, it rapidly became apparent that \LaTeX\ is a poor choice to manage structured data.
-A simple JavaScript program was written to convert the \LaTeX\ data into JSON (which is much more readable than \LaTeX) and also generate CSV files that can be processed in standard office software such as Excel, which some readers may prefer.
-In fact, examining and comparing the same data in the contrasting formats, this typeset file, in JSON, and in Excel (reading the generated CSV) provided multiple different perspectives of the data that increased redundancy and confidence that the data was correct and correctly handled.
-
-{It is important to note that using such techniques is quite routine in science publication, though often pre-existing tools are used to streamline the process (and to ensure that it is more widely understood).
-The paper \cite{paper-usesRMarkdown}, for example, in addition to using a typesetting system for publication, also placed its code in a repository using R Markdown \cite{RMarkdown}, a programming environment based on R designed for generating and documenting lab books --- almost the polar opposite of \LaTeX, which is designed for publication but can be used for programming.}
-
-Finally note that what may look like magic numbers used throughout the present paper (such as the $\the\dataN$, as in ``\plural{\dataN}{paper} were evaluated'') are all in fact named, calculated and placed \emph{in situ\/} directly from computations performed on the JSON paper's data.
-
-\changemysection{The Speigelhalter trustworhiness questions}
-\globalLabel{supplementary-Speigelhalter-section}
-
-David Speigelhalter is concerned how statistics is often misused and misunderstood.
-In his \emph{The Art of Statistics\/} \cite{Speigelhalter} Speigelhalter brings together his advice for making reliable statistical claims: they need to be accessible, intelligible, assessable, and usable --- the claims need to be properly accountable.
-Speigelhalter proposes ten questions to ask when confronted with any claim based on statistical evidence.
-Some of his questions are quite general, and might be applied to any sort of scientific claims, but all have analogous questions that could be addressed to software code or papers relying on code --- analogues are suggested in \textbf{bold} below.
-
-What might seem like dauntingly technical software issues are no more demanding than the basic statistical issues that are regularly acceded to; failing to ask them is as risky as dismissing statistical scrutiny.
-
-\def\questionsection#1{\changemysubsection{#1}}
-\def\question#1{\item \emph{#1\/}}
-\def\sequestion#1{\begin{itemize}\raggedright\item[$\blacktriangleright$] \textbf{#1}\end{itemize}}
-
-\questionsection{How trustworthy are the numbers?}
-\newcounter{resumeCounter}
-
-\begin{enumerate}
-\question{How rigorously has the study been done?} For example, check for `internal validity,' appropriate design and wording of questions, pre-registration of the protocol, take a representative sample, using randomization, and making a fair comparison with a control group.
-
-\sequestion{How rigorously has the software engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
-
-\sequestion{``Internal validity'' assumes that there is evidence the programmers had uncertainty in the code's reliability and checked it.
-Were different methods used and compared, or was all confidence put into a single implementation?
-What internal consistency checks does the implementation have?
-Were invariants and assertions defined and checked?
-}
-
-\question{What is the statistical uncertainty/confidence in the findings?} Check margins of error, confidence intervals, statistical significance, multiple comparisons, systemic bias.
-
-\sequestion{How are the claims presented that give us confidence in the code that they are based on?
-Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
-
-\question{Is the summary appropriate?} Check appropriate use of averages, variability, relative and absolute risks.
-
-\sequestion{If the claims are exploratory, weaker standards of coding can be used; if the claims are a basis for critical decisions, then there should be evidence of using appropriate software engineering (such as defensive programming) to provide appropriate confidence in the results claimed.}
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the source?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How reliable is the source of the story?} Consider the possibility of a biased source with conflicts of interest, and check publication is independently peer-reviewed.
-Ask yourself, `Why does this source want me to hear this story?'
-
-\sequestion{The source of many science stories is the output of running some code.
-How reliable is this code?
-What evidence is there that the code was well-engineered so its reliability can be trusted?}
-
-\question{Is the story being spun?} Be aware of the use of framing, emotional appeal through quoting anecdotes about extreme cases, misleading graphs, exaggerated headlines, big-sounding numbers.
-
-\sequestion{Be wary of AI and ML which may have been trained by chance or specifically (if not deliberately) to get the results described.}
-
-\question{What am I not being told?} This is perhaps the most important question of all.
-Think about cherry-picked results, missing information that would conflict with the story, and lack of independent comment.
-
-\sequestion{Cherry picking with code is often unconscious and is very common: when running code produces the ``cherries'' for a paper it is tempting to stop testing the code and just assume it is running correctly.
-So, what evidence is there that the code was rigorously developed and cherry picking avoided?}
-
-\setcounter{resumeCounter}{\value{enumi}}
-\end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
-\begin{enumerate}
-\setcounter{enumi}{\value{resumeCounter}}
-\question{How does the claim fit with what else is known?} Consider the context, appropriate comparators, including historical data, and what other studies have shown, ideally in a meta-analysis.
-
-\sequestion{Is there any discussion of the code and how does it compare with other peer-reviewed publications using code used for similar purposes?}
-
-\question{What's the claimed explanation for what has been seen?} Vital issues are correlation v.\ causation, regression to the mean, inappropriate claim that a non-significant result means `no effect,' confounding attribution, prosecutor's fallacy.
-
-\sequestion{These are all good statistical questions.
-The software engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and software engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
-
-\question{How relevant to the story is the audience?} Think about generalizability, whether the people being studied are special case, has there been an extrapolation from mice to people.
-
-\sequestion{Generalizability is equivalent to is the code available, easy to understand and use for more general purposes --- including further work and checking the reproducibility of the claims being made?}
-
-\question{Is the claimed effect important?} Check whether the magnitude of the effect is practically significant, and be especially wary of claims of `increased risk.'
-\end{enumerate}
-
-\newcount\temporary \temporary=\mainNumberOfReferences
-\newcount\temporaryPlusOne \temporaryPlusOne=\temporary
-\advance \temporaryPlusOne by 1
-
-\bibliographystyle{plos2015}
-
-\global \newdimen \shortercol 
-\renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
-    \startBibliography
-    \raggedright
-    \parindent=0em
-    \setbox0=\hbox{\hskip 1ex[#1]}
-    \labelwidth=\wd0
-    \advance \labelwidth by 1ex
-    \shortercol=\columnwidth \advance \shortercol by -\labelwidth
-}{}
-
-\makeatletter    
-    \renewcommand{\bibitem}[2][DEFAULT]{%
-    	\if@filesw \immediate\write\@auxout{\string\bibcite{#2}{{{\the\bibciten}{}{{}}{{}}}}}\fi
-        \parshape 2 0em \columnwidth \labelwidth \shortercol
-        \leavevmode\hbox to \labelwidth{\hfill[\the\bibciten]\hskip 1ex}%
-        \global\advance\bibciten by 1
-        \ignorespaces 
-    }
-\makeatother
-
-\initialiseBibliography{Additional references for \supplement}{\temporaryPlusOne}{References numbered 1--\the\temporary\ appear in the reference list in the main paper.}
-
-\immediate\write\infofile{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
-\bibliography{paper-seb-supplementary-material.bib}
-
-\highlightmysection{Summary of pilot survey}
-\label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}\label{data-methods-subsection}
-A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
-Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research.
-Commentary, correspondence and editorials were excluded.
-The sample represents what the editorial and the broader peer review community considers to be good practice.
-
-The selection process will have certainly missed some papers that use code, but the criterion selects papers where the wording of the title indicates that the authors consider code to be a component of the scientific contribution.
-Indeed, all sampled papers used code in their research.
-Although there is unavoidable subjectivity in the paper evaluations and uncontrolled bias from using a single evaluator (the author of this paper), it is hoped that using a sample of \plural{\dataN}{paper} from \plural{\numberOfJournals}{diverse journal} is sufficient to randomize errors so that they largely cancel out, and the overall trends as discussed in this paper are reliable.
-It should be noted that, except where a paper provides a URL to a code repository, much code was disorganized so possibly not all code was reviewed because it was too hard to find (some emails to authors have not been responded to).
-
-Since almost every scientific paper relies on generic computer code (calculating statistics, plotting graphs, storing and manipulating data, accessing internet resources, etc), the baseline of papers using code was not assessed.
-Papers whose title indicated their contribution included or relied on bespoke code were selected, and all those clearly relied heavily on their own specifically developed code.
-Papers that may have relied on bespoke code but whose titles made no such implication were not assessed.
-
-There is considerable debate over what good commenting practice is, but this is because comments have many roles --- from helping students to get marks in assessments, asserting intellectual rights, reminding the developer of things to do, managing version control, to explaining the code to third parties.
-Different programming languages also develop ``cultures'' that encourage different approaches for comments (examples include R Markdown, Mathematica Notebooks, JavaDoc, Haskell's Haddock, and so on).
-For scientific code, however, the explanatory role is critical, and this is what was assessed in this survey.
-Note that comments used for explanation does not preclude any other uses.
-
-The completeness or executability of code was not assessed, although if code was obviously incomplete this was noted.
-Whether code runs as claimed is a matter of research integrity, which is beyond the scope of this survey.
-What is relevant to the study is whether the code is described in sufficient detail that the methods used can be scrutinized.
-Obviously being able to run the code will help, but clarity in documentation and comments is critical.
-It is more like ``can we see the critical pages from your lab book so we understand what you did?'' rather than ``can we have free run of your laboratory?''
-
-As an informal survey, intended to establish whether the issues in epidemic modeling were more widespread, and given the very poor level of documentation found in scientific code, it was not felt necessary to have independent or blind assessment.
-
-The data was recorded in JSON (JavaScript Object Notation), which is a simple standard data format. A typical entry in the data file looks like this (with long field values truncated for clarity):
-
-\input generated-example-data.tex
-
-The data was written out by hand, after reading and reviewing each paper in the survey.
-In total there are \countFields\ data fields available for documenting papers, but not all need be used for each paper; for example, the field \texttt{hasCodeTested} defaults to \texttt{false}, so it need not be set --- it is also an error to set it if another field asserts there is no code to evaluate!
-(A separate JSON data structure maps the data fields to English descriptions, along with default values if they are optional descriptors.)
-
-A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
-Hence, running the data generates the core human-readable information used in this paper.
-
-The JavaScript data+program generates files from the JSON, with all the definitions; these files were then included in both the main paper and in this \supplement, so when the paper or \supplement\ is typeset all tables and specific data items are typeset automatically, consistently and reliably by \LaTeX\@.\footnote{Because of its approach to automatic typesetting, this journal requires that no files are explicitly \texttt{input} into \LaTeX, so a simple JavaScript program is used to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.} For example, the register \texttt{\bslash dataN} is set to the value \the\dataN, which is the total number of papers assessed in the JSON data, and the macro \texttt{\bslash journalBreakdown} is defined directly from the data to be the following text (when typeset in \LaTeX):
-
-\begin{quote}
-\journalBreakdown 
-\end{quote}
-
---- which is the breakdown of the total $N=\the\dataN$ by journal name. The \emph{exact\/} same text was also used in the main paper.
-
-An interesting consequence of this automatic approach is that as the author found themselves starting to write text such as:
-
-\begin{quote}
-\tt Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} \ldots
-\end{quote}
-
-it motivated extending the JavaScript data processing so that \emph{all\/} specific quantities mentioned in the paper are traceable directly back to the JSON data.
-The phrase above is now in fact written in \LaTeX\ in the paper as follows:
-
-\begin{quote}\tt\small
-Code repositories were used by \\
-\bslash plural\{\bslash countUsesVersionControlRepository\}\{paper\} \ldots
-\end{quote}
-
-where {\tt\small \bslash plural} automatically writes a word (``paper'' in this case) in singular or plural form as required (in this case, the variable \texttt{countUsesVersionControlRepository} = \the\countUsesVersionControlRepository).
-When typeset the text above is generated with the relevant number or numbers inserted from the JSON data.
-
-Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured to generate table \ref{table-repo-summary}, as used in the main paper.
-
-The full JavaScript JSON data and processing code (including the makefile) is provided in this paper's repository, as described in the main paper.
-
-\highlightmysubsection{Detecting and defending against error}
-\label{detecting-against-error}
-It must be emphasized that an automatically-guaranteed same number appearing in multiple different contexts is an extremely effective way of defending against common Human Factors errors.
-
-% check it works for other numbers 
-%\countUsesVersionControlRepository=296
-
-Normally, when we write \the\countUsesVersionControlRepository\ in a paper, we proof read it as ``just a number'' and the sentence would likely seem to make as much sense if the number was 
-\newcount\t
-\t=\countUsesVersionControlRepository
-\divide \t by 10
-\newcount\fraction
-\fraction=\countUsesVersionControlRepository
-\newcount\tent \tent=\t \multiply \tent by 10
-\advance\fraction by -\tent
-\the\t.\the\fraction,
-\t=\countUsesVersionControlRepository
-\advance \t by -1
-\the\t,
-\t=\countUsesVersionControlRepository
-\advance \t by 1
-\the\t, or
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t\ --- we hardly bother to pay attention because we know what we are reading; at least we know what we meant to write. It is very hard to spot one's own typos.
-
-\begin{itemize}\raggedright
-\item
-The first and last errors above are examples of the very common error of ``out by ten'' (common partly because the correct number, \the\countUsesVersionControlRepository\ looks very similar to \t=\countUsesVersionControlRepository
-\divide \t by 10
-\the\t.\the\fraction, and \the\countUsesVersionControlRepository.0 also looks very similar to 
-\t=\countUsesVersionControlRepository
-\multiply \t by 10
-\the\t) \cite{fixit}.
-
-\item
-The middle two errors above are examples of the common error of ``out by one,'' or ``fence post errors'' frequently made by mixing up counting fences or the posts (there is usually one more post than fence panel) \cite{fixit}.
-\end{itemize}
-
-All the discussion and examples above were generated automatically, and have been checked  correct for other base values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the sentences say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
-
-If any of the numbers used in a paper were safety critical (e.g., lives directly depend on their correct values) then further checks would have been made to help detect and avoid errors.
-For example, \LaTeX\ makes it very easy to check that numbers fall within reasonable ranges, or have any other required safety properties.
-For the present paper, a potential problem is that the paper is typeset \emph{before\/} the JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense.
-
-\newcount\min \min=5
-\newcount\max \max=20
-%\countUsesVersionControlRepository=4
-\ifnum \countUsesVersionControlRepository>\max countUsesVersionControlRepository is far too large! \fi
-\ifnum \countUsesVersionControlRepository<\min countUsesVersionControlRepository is far too small! \fi
-\ifnum \countUsesVersionControlRepository>\max
-	\else 
-		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits for this paper.}
-		\fi
-\fi
-
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-
-\highlightmysubsection{Code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in section \ref{detecting-against-error}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-\\
-
-{\sf
-\noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
-These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
-[\ldots] As a minimum, sufficient information and data are required to allow others to replicate all study findings reported in the article.
-Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.
-
-\vskip 3mm
-{\raggedleft\scriptsize
-\url{royalsociety.org/journals/authors/author-guidelines/#data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
---- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.
-
-{\vskip 3mm\scriptsize
-\raggedleft
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data} \\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Lancet Digital Health\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\\
-accessed 29 July 2020.\par
-}
-
-\noindent\textbf{Extract from \emph{Journal of Vascular Surgery\/} author guidelines}\\
---- Has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant for section \ref{jvs-policy} in the main paper:
-
-The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.
-{\vskip 3mm
-\raggedleft\scriptsize
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\\
-accessed 29 July 2020.\par
-}
-}
-
-\highlightmysubsection{Assessments and criteria}
-
-\def\flagStyle#1{\textcolor{red}{\sf #1}}
-
-Legend:\\
-\input generated-legend.tex
-
-\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.25in}p{2.25in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated-assessments.tex
-\end{longtable}
-}
-
-\highlightmysubsection{Summary of assessments}
-
-\immediate\write\infofile{\bslash newlabel{supplementary-summary-table}{{\arabic{section}}{\thepage}{DUMMY1}{DUMMY2}{DUMMY3}}}
-
-\begin{center}
-\input generated-summary-table.tex
-\end{center}
-
-This is exactly the same table as table \ref{table-summary} from the main paper, reproduced here for convenience.
-See section \ref{data-methods-section}.\ref{data-methods-subsection} in this \supplement\ for details of the process that generated it.
-
-\stepcounter{subsection}
-\def\reftitle{References for sampled papers}
-\def\unnumberedrefname{{\reftitle}}
-\renewcommand\refname{\subsectioncounter\ {\unnumberedrefname}}
-
-\initialiseBibliography{{\alph{subsection} \reftitle}}{\the\bibciten}{}
-
-\begin{thebibliography}{999}
-\subsectioncontents{\unnumberedrefname}
-\raggedright
-
-% horrible hack -- we want to refer to surveyed papers in the main body, but only in the
-% figure summarising github repos, so we do it specially....
-
-\def\maprepo#1{\immediate\write\infofile{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
-
-\input generated-supplementary-references.tex
-
-\end{thebibliography}
-
-\immediate\closeout\infofile
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
-\end{document}
+\citenum{ref-1} & On request\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
+Documented overview, but only trivial comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''``''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request''\flagStyle{ } & Poor commenting, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team''\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
+However, the data are available from the corresponding author on reasonable request''\flagStyle{ } & Empty GitHub repository: ``Code coming soon\ldots'' it says\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
+Data from UCSF are not available at this time as they have not been legally certified as being De-Identified, however, this process is underway and the data may be available by the time of publication by contacting the authors.
+Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply''\flagStyle{ } & Basic documentation, very little comment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request''\flagStyle{ } & Trivial comments, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available''\flagStyle{ } & ``We used only free and open-source software'' some of which is unspecified\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{paper-usesRMarkdown} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
+\citenum{ref-14} & Data directly written into program code\flagStyle{ } & Basic Matlab with routine comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
+Currently (apparently) for private view\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
+Paper describes using two contrasting methods to help confirm correctness, ``As an additional check, I also coded the calculation of D based on a probabilistic approach, using genotype frequencies in each population to calculate the expected frequencies of each possible two-genotype combination (electronic supplementary material, table S1).
+Essentially identical results were obtained.'' but the contrasting method is not available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-16} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-17} & On request\flagStyle{ } & R lightly commented\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ethics-paper} & No data required\flagStyle{ } & Unrunnable incomplete code fragment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-19} & Data embedded in PDF\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-20} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-21} & Partial data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including manual\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
+\citenum{ref-22} & No data required\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-23} & Data cited, not all available\flagStyle{ } & Trivial documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{example-numerical-error} & On Figshare\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
+Helpful features to make usable for third parties\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-25} & Data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-26} & Data on various web sites\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-27} & Data on request\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-28} & Data on request\flagStyle{ } & Python scripts can be requested\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-29} & Unspecified location on large website requiring registration\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-30} & Available to researchers who meet criteria for access to confidential data\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-31} & Data access conditional on approved study proposal\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-32} & Unspecified locations on several large websites\flagStyle{ } & Python used and apparently GitHub, but --- an oversight?
+--- no code is available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\

### 2023-02-20 generated/make-help.tex
Various corrections inspired by Konrad Hinsen

@@
 \texttt{make data}&   Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc.
-In particular, this runs \texttt{node data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
+In particular, this runs \texttt{node programs/data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
 Note that downloading all the repositories in a reasonable time needs decent internet bandwidth.\\
@@
 \multicolumn{1}{l@{\vdots}}{}&\\
-\texttt{make tidyup}&   Remove all easily generated files, except the main PDFs and the large Git repositories needed for the pilot survey.\\
+\texttt{make tidyup}&   Tidyup before doing a git commit.
+Remove all easily generated files, and the large Git repositories needed for the pilot survey.
+Do not remove the main PDFs, or the \LaTeX\ data include files.\\
 \end{tabular}}

### 2023-02-20 generated/allGitRepos.sh

@@ -1,0 +1,36 @@
+# run this shell script in directory models
+rm -rf "git-MetricSelectionFramework"
+git clone "http://github.com/ChristophKanzler/MetricSelectionFramework" "git-MetricSelectionFramework"
+
+rm -rf "git-PENet"
+git clone "http://github.com/marshuang80/PENet" "git-PENet"
+
+rm -rf "git-PostoperativeOutcomes_RiskNet"
+git clone "http://github.com/cklee219/PostoperativeOutcomes_RiskNet" "git-PostoperativeOutcomes_RiskNet"
+
+rm -rf "git-philter-ucsf"
+git clone "http://github.com/BCHSI/philter-ucsf" "git-philter-ucsf"
+
+rm -rf "git-AI-CDSS-Cardiovascular-Silo"
+git clone "http://github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo" "git-AI-CDSS-Cardiovascular-Silo"
+
+rm -rf "git-SiameseChange"
+git clone "http://github.com/QTIM-Lab/SiameseChange" "git-SiameseChange"
+
+rm -rf "git-lactModel"
+git clone "http://github.com/SolangeD/lactModel" "git-lactModel"
+
+rm -rf "git-LRM"
+git clone "http://github.com/UT-NSG/LRM" "git-LRM"
+
+rm -rf "git-manifold-ga"
+git clone "http://github.com/ki-analysis/manifold-ga" "git-manifold-ga"
+
+rm -rf "git-blast-ct"
+git clone "http://github.com/biomedia-mira/blast-ct" "git-blast-ct"
+
+( echo "% date generated by running generated/allGitRepos.sh downloading Git repos";
+echo \\\\def\\\\clonedate {\\\\ignorespaces `date "+%e %B %Y"`}
+echo \\\\def\\\\cloneyear {`date "+%Y"`}
+echo \\\\def\\\\clonemonth {`date "+%m"`}
+) > ../generated/clone-date.tex

### 2023-02-20 generated/info-for-main.tex
Sorting out .aux files not being deleted on make tidyup

@@
-\newlabel{supplementary-material-makefiles}{{11}{38}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{on-code-data-publication}{{12}{41}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{RAP-diagrams}{{12}{44}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{excelFigure}{{2}{44}{AAA}{BBB}{CCC}}
-\newlabel{supplementary-Spiegelhalter-section}{{13}{44}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{survey-section}{{14}{47}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{supplementaryTableRepoSummary}{{4}{49}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{supplementary-journal-policies-section}{{14}{52}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{restrictive-policies}{{14}{56}{DUMMYA}{DUMMYB}{DUMMYC}}
-\def\MaxMainPaperCitationNumber{121}
+\newlabel{supplementary-material-makefiles}{{11}{34}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{on-code-data-publication}{{12}{37}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{RAP-diagrams}{{12}{40}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{excelFigure}{{2}{40}{AAA}{BBB}{CCC}}
+\newlabel{supplementary-Spiegelhalter-section}{{13}{40}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{survey-section}{{14}{43}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementaryTableRepoSummary}{{4}{45}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementary-journal-policies-section}{{14}{48}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{restrictive-policies}{{14}{52}{DUMMYA}{DUMMYB}{DUMMYC}}
+\def\MaxMainPaperCitationNumber{0}

### 2023-02-20 generated/over-fitting-code-section.tex
Sorting out .aux files not being deleted on make tidyup

--- 2023-01-31 generated/stand-alone-over-fitting-code.tex
@@
-\def\numbertoname#1{#1}
-\def\remembernamevalue#1#2{}
-\newcount\countRepos
-\newcount\countMissingCode
-\newcount\countCommentedOut
-\newcount\countNotCommentedOut
-\documentclass[11pt]{article}
-\usepackage{geometry} 
-\geometry{a4paper}  
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-\usepackage{url}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert \#1 `dirname #1`/`basename #1 .tif`.png}
-\def\supplement{Supplement}
-\begin{document}
-\section{Section generated by \emph{Mathematica\/} taken from full paper}\subsection{The siren call of over-fitting code}\label{over-fit}
+\subsection{The siren call of over-fitting code}\label{over-fit}
 %Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that.
@@
 \def\figureStarDetails{*}
+
 \begin{figure\figureStarDetails}[t]
@@
   \imageWidthCalculation
-  \begin{center}\includegraphics[width=\imageWidth]{generated/mathematicaplot.pdf}\end{center}}
+  \begin{center}\includegraphics[width=\imageWidth]{
+generated/mathematicaplot.pdf
+}\end{center}}
 \caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
@@
 \end{figure\figureStarDetails}
+
 In modeling terms, superficially successful computer programs tend to \emph{over-fit\/} phenomena \cite{over-fit}.
@@
 Looking specifically at the data plotted in figure \ref{fig-overfit} if, for the sake of argument, we assume the error in the data is normally distributed then the values the over-fitted code generates outside the range of the sample are improbable.
-For example, the basic linear model predicts$\hat{y}(0)=0.5$, versus the extreme value predicted by the over-fitting code, $\hat{y}(0)=-41.8$,  even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.
+For example, the basic linear model predicts
+$\hat{y}(0)=0.5$, 
+versus the extreme value predicted by the over-fitting code, 
+$\hat{y}(0)=-41.8$, 
+ even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.
 
@@
 The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases.
-Indeed, many of the programs in the survey used comment to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countRepos\ papers in the pilot survey that reported use of code repositories (covering \input ../generated/kloc.tex
+Indeed, many of the programs in the survey used comment to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countRepos\ papers in the pilot survey that reported use of code repositories (covering \input generated/kloc.tex
 \totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
@@
 Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
+
 \vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
@@
 }\hfill}\end{tabular}}
-\section{Data and polynomials}
-\begin{center}\sf\begin{tabular}{|c|r|r|r|r|r|}\hline
-$x={}$&1.00&1.10&2.20&2.60&4.50\\
-$y={}$&1.50&2.70&4.90&5.70&8.20\\
-\hline\end{tabular}
-\end{center}
-
-Showing both polynomials with coefficients to 2 decimal places, the linear least squares model to fit this data is:
-
-$$
-\hat{y}=0.49+1.80x
-$$
-
-and the Lagrange polynomial model (an exact fit to the 5 data points) is
-
-$$
-\hat{y}=-41.79+85.46x-56.30x^2+15.65x^3-1.51x^4
-$$
-
-The name of the relevant \emph{Mathematica\/} notebook file where all data and code for this section (and for generating the paper's section \ref{over-fit} and its figure \ref{fig-overfit}) can be found in \texttt{programs/over-fitting-section.nb}, which is included in the Git repository for the paper.\end{document}

### 2023-02-20 generated/over-fitting-data.tex
Sorting out .aux files not being deleted on make tidyup

--- 2023-01-31 generated/stand-alone-over-fitting-code.tex
@@
-\def\numbertoname#1{#1}
-\def\remembernamevalue#1#2{}
-\newcount\countRepos
-\newcount\countMissingCode
-\newcount\countCommentedOut
-\newcount\countNotCommentedOut
-\documentclass[11pt]{article}
-\usepackage{geometry} 
-\geometry{a4paper}  
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{epstopdf}
-\usepackage{url}
-\DeclareGraphicsRule{.tif}{png}{.png}{`convert \#1 `dirname #1`/`basename #1 .tif`.png}
-\def\supplement{Supplement}
-\begin{document}
-\section{Section generated by \emph{Mathematica\/} taken from full paper}\subsection{The siren call of over-fitting code}\label{over-fit}
-%Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that.
-We do the science, and we are in the world, so there are challenging meta-theoretical issues as well as practical experimental and resourcing issues.
-Karl Popper is one of many philosophers of science to develop principles and theories about science \cite{popper-three-worlds,popper-conjectures-refutations}.
-
-%Popper was concerned with the conventional sciences, where the world, Nature, is assumed fixed and in principle knowable.
-Writing just six years after Popper, Herbert Simon introduced the ``sciences of the artificial'' \cite{sciences-artificial}, which instead focus on understanding artificial worlds, most typically computer programs.
-Programs effectively create worlds, and the sciences of the artificial seek to understand and improve those worlds.
-Crucially, there is no philosophical or other reason why artificial worlds should be based on principles, let alone be simple or elegant.
-Conventional science's \ae sthetics need not apply.
-
-%Science to date has been inevitably driven by human \ae sthetics: that is, simple, elegant, understandable, beautiful, brief, exciting, intelligible \ldots\ all are human properties that are applied to doing, funding, defining, and selecting ``good'' science.
-
-\noindent
-Poor code undermines scientific quality: poor code can generate plausible results from \emph{any\/} data, including even fictional science.
-A temptation is that %artificial world \cite{sciences-artificial}, built on
-adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena.
-\def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
-\def\figureStarDetails{*}
-\begin{figure\figureStarDetails}[t]
-{ \newdimen \imageWidth 
-  \imageWidth=\textwidth
-  \imageWidthCalculation
-  \begin{center}\includegraphics[width=\imageWidth]{generated/mathematicaplot.pdf}\end{center}}
-\caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
-Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
-The over-fitted polynomial model fits the sample \emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
-\label{fig-overfit}
-\end{figure\figureStarDetails}
-In modeling terms, superficially successful computer programs tend to \emph{over-fit\/} phenomena \cite{over-fit}.
-Instead of using code to test the models, we tinker with the code adapting it closer and closer to our prejudices.
-The code then apparently confirms our science, since we fitted it to our understanding.
-
-In general, an \emph{over-fitted model\/} fits a set of data closely, but contains more parameters than can be justified by the science.
-An over-fitted model fails to reliably predict results beyond the scope of the data it has been fitted to.
-Over-fitting is a well-known problem, but the point is that when code is used, over-fitting is done unconsciously by programmers adjusting the code --- its parameters, its structure, its embedded data, the calculations it performs --- that specifies the model.
-``A model over-fits if it is  more complex than another model that fits equally well'' \cite{over-fit}, is a criterion that describes almost every program!
-Programmers without the discipline and experience to manage the unlimited adaptability of code debug, alter, and extend their code to make it do what they think it should do.
-This becomes a vicious circle as the idea of what the science is becomes driven by the code.
-Rather than debugging code by improving its fit to the actual science, it gets debugged and extended to fit the expectations.
-
-The problems of over-fitting data may be visualized using a Real$\rightarrow$Real function of one variable (figure \ref{fig-overfit}).
-The code that generates an over-fitted curve seems to work very well: in the example shown, the over-fitted curve fits the sampled data exactly; indeed, the code used here will fit any new data exactly as well.
-But the code has a negligible ability to predict new data or to describe theories of the data, which is the point of modeling.
-The fact that over-fitted code seems to work well is deceptive.
-
-Looking specifically at the data plotted in figure \ref{fig-overfit} if, for the sake of argument, we assume the error in the data is normally distributed then the values the over-fitted code generates outside the range of the sample are improbable.
-For example, the basic linear model predicts$\hat{y}(0)=0.5$, versus the extreme value predicted by the over-fitting code, $\hat{y}(0)=-41.8$,  even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.
-
-While over-fitting data is a well-known problem, the point for this paper is that \emph{code\/} itself can easily be over-fitted.
-Code can of course be over-fitted in more complex ways than can be illustrated with elementary polynomials, as here.
-Code over-fitting is much harder to recognize because there may be no simple graph plot, like figure \ref{fig-overfit}, to highlight the problems.
-Furthermore, almost all code is far more intricate than the two trivial polynomials used to illustrated figure \ref{fig-overfit}.
-
-Unfortunately, code in published science \emph{is\/} often over-fitted, and over-fitted in a way that is very hard to scrutinize.
-For example, in epidemiology (which is considered in section \ref{section-pandemic-modeling}) it is routine to use very complicated, large dynamical models parameterized with numerous social, cultural, health, demographic and geographical data.
-The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases.
-Indeed, many of the programs in the survey used comment to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countRepos\ papers in the pilot survey that reported use of code repositories (covering \input ../generated/kloc.tex
-\totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
-The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\remembernamevalue{numbernamesexamplefootnote}{\thefootnote}}
-
-Furthermore, scientific support code is rarely documented well enough to know what it should have been doing, which should be answered by a specification.
-With no clear specification and documentation, the code can be arbitrarily hacked to get any convenient results, since no particular specification for it has been defined that it should adhere to.
-Thus we risk doing and promoting substandard science because we --- the scientists and the publication process --- are not managing the unlimited adaptability and complexity of code that science has come to rely~on.
-This is over-fitting of the worst kind --- in conventional over-fitting one can at least hope to see that the fitting is over-parameterized for the data, but in code over-fitting the code and specification are not visible, therefore not adequately scrutinized, and --- worse --- the ``data'' the code over-fits includes the entire conceptual contribution of the paper.
-
-Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
-\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
-\vskip 1ex
-\noindent 
-{\newdimen\boxwidth
-\boxwidth = \columnwidth
-\advance\boxwidth by - 1em
-\begin{tabular}{@{}l@{}}
-\framebox[\columnwidth][r]{\hfill\parbox{\boxwidth}{This section (\ref{over-fit}) of this paper, including figure \ref{fig-overfit}, was generated as a computational paper in a \emph{Mathematica\/} notebook, then exported to \LaTeX\ to include here.
-The \supplement\ provides a more complete explanation, as well as the data and details of the fitted polynomials (also generated automatically).
-This paper's repository includes the notebook.
-
-%\hskip 1em Generated text and data presented elsewhere in this paper were created by a \texttt{make} file running shell scripts and JavaScript code analyzing repositories, JSON survey data, etc.
-Results were exported to files of \LaTeX\ markup and definitions, then imported into this paper and expanded where needed.
-}\hfill}\end{tabular}}
-\section{Data and polynomials}
 \begin{center}\sf\begin{tabular}{|c|r|r|r|r|r|}\hline
@@
 
-The name of the relevant \emph{Mathematica\/} notebook file where all data and code for this section (and for generating the paper's section \ref{over-fit} and its figure \ref{fig-overfit}) can be found in \texttt{programs/over-fitting-section.nb}, which is included in the Git repository for the paper.\end{document}
+The name of the relevant \emph{Mathematica\/} notebook file where all data and code for this section (and for generating the paper's section \ref{over-fit} and its figure \ref{fig-overfit}) can be found in \texttt{programs/over-fitting-section.nb}, which is included in the Git repository for the paper.

### 2023-02-27 generated/over-fitting-code-section.tex
Generalising image formats - PDF/JPG etc
Fixing mathematica script bugs

@@
   \begin{center}\includegraphics[width=\imageWidth]{
-generated/mathematicaplot.pdf
+generated/mathematicaplot.jpg
 }\end{center}}
@@
 Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
-The over-fitted polynomial model fits the sample \emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
+The over-fitted polynomial model fits the sample
+\emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
 \label{fig-overfit}

### 2023-02-27 generated/stand-alone-over-fitting-code.tex
Generalising image formats - PDF/JPG etc
Fixing mathematica script bugs

@@
   \imageWidthCalculation
-  \begin{center}\includegraphics[width=\imageWidth]{generated/mathematicaplot.pdf}\end{center}}
+  \begin{center}\includegraphics[width=\imageWidth]{generated/mathematicaplot.jpg}\end{center}}
 \caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
 Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
-The over-fitted polynomial model fits the sample \emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
+The over-fitted polynomial model fits the sample
+\emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
 \label{fig-overfit}

### 2023-02-27 generated/make-help.tex
Mainly getting mathematica+make to work together

@@
+% Edit makefile to change the selected make command choices, as follows:
 % 1 - make all
@@
 % 10 - make mathematica
-% 11 - make one-file
-% 12 - make pdf
-% 13 - make push
-% 14 - make readme
-% 15 - make really-tidyup
-% 16 - make tidyup
-% 17 - make zip
-% 18 - make zip-data
+% 11 - make mathematica-open
+% 12 - make one-file
+% 13 - make pdf
+% 14 - make push
+% 15 - make readme
+% 16 - make really-tidyup
+% 17 - make tidyup
+% 18 - make zip
+% 19 - make zip-data
+
 {\sf\begin{tabular}{rp{4.5in}}
-\texttt{make all}&   Analyze the data, then typeset the main PDF files (\texttt{paper-seb-main.pdf} and \texttt{paper-seb-supplementary-material.pdf}).\\
+
+\texttt{make all}&
+      Analyze the data, then typeset the main PDF files (\texttt{paper-seb-main.pdf} and \texttt{paper-seb-supplementary-material.pdf}).\\
    \multicolumn{1}{l@{\vdots}}{}&\\
-\texttt{make data}&   Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc.
-In particular, this runs \texttt{node programs/data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
+
+\texttt{make data}&
+      Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc.
+This \texttt{make} option runs \texttt{node programs/data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
 Note that downloading all the repositories in a reasonable time needs decent internet bandwidth.\\
    \multicolumn{1}{l@{\vdots}}{}&\\
-\texttt{make help}&   Explain how to use \texttt{make}, and list all available options for using \texttt{make}.\\
+
+\texttt{make help}&
+      Explain how to use \texttt{make}, and list all available options for using \texttt{make}.\\
    \multicolumn{1}{l@{\vdots}}{}&\\
-\texttt{make one-file}&   Make a single PDF file paper-seb.pdf (i.e., paper + appendix) all in one.\\
-\multicolumn{1}{l@{\vdots}}{}&\\
-\texttt{make tidyup}&   Tidyup before doing a git commit.
+
+\texttt{make tidyup}&
+      Tidyup before doing a Git commit.
 Remove all easily generated files, and the large Git repositories needed for the pilot survey.
-Do not remove the main PDFs, or the \LaTeX\ data include files.\\
+Do not remove the main PDFs, or the \LaTeX\ data include files.
+Do not remove the .aux files, as \LaTeX\ runs much more smoothly with them.\\
 \end{tabular}}

### 2023-02-27 generated/summary-table.tex
Mainly getting mathematica+make to work together

@@
 Some or all code in principle available on request&8&25\%&\\
-Requested code actually made available (within 2 years 7 months${}^{\star}$)&0&0\%& \\\hline
+Requested code actually made available (within 2 years 8 months${}^{\star}$)&0&0\%& \\\hline
  \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
@@
 \hline\end{tabular}\\
-\vskip 1ex ${}^{\star}$Time of 2 years 7 months is wait between code request and date of generating this table.
+\vskip 1ex ${}^{\star}$Time of 2 years 8 months is wait between code request and date of generating this table.

### 2023-02-27 paper-seb-supplementary-material.tex
Added some conditional Latex to handle temporarily underfined SupplementalLastPage macro
May fix PDF bounding box portability prolem
Generalising image formats - PDF/JPG etc
Mainly getting mathematica+make to work together

@@
     \usepackage{fancyhdr}
+
     \usepackage{graphicx}
+	\PassOptionsToPackage{dvipdfmx}{graphicx}
+
 	\usepackage{amssymb}
@@
 \makeatletter
+
 \def\rememberpage{\immediate\write\@auxout{\bslash gdef\bslash SupplementalLastPage{\thepage}}}
+
+% putting \typeout in the right footer will get repeated warnings (which I hope will be noticed!)
+% to complement the red text in the footer itself on every page
+\@ifundefined{SupplementalLastPage}{\rfoot{\supplement\ \thepage\ \textcolor{red}{\bf(Rerun to get number of pages)}\typeout{* Rerun to get correct number of pages in the running footer}}}{\rfoot{\supplement\ \thepage/\SupplementalLastPage}}
+
 \makeatother
 
-\rfoot{\supplement\ \thepage/\SupplementalLastPage}
+%\rfoot{\supplement\ \thepage/\SupplementalLastPage}
 \lfoot{\today}
@@
    \centering
-   \includegraphics[width=5in]{generated/basic.pdf} 
+   \includegraphics[width=5in]{generated/basic.\imageFileSuffix} 
    \caption{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
@@
    \centering
-   \includegraphics[width=5in]{generated/excel.pdf} 
+   \includegraphics[width=5in]{generated/excel.\imageFileSuffix} 
    \caption{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
@@
    \centering
-   \includegraphics[width=5in]{generated/import.pdf} 
+   \includegraphics[width=5in]{generated/import.\imageFileSuffix} 
    \caption{Improving over the normal approach (figure \ref{fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
@@
    \centering
-   \includegraphics[width=5in]{generated/notebook.pdf} 
+   \includegraphics[width=5in]{generated/notebook.\imageFileSuffix} 
    \caption{In a notebook system, such as \emph{Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
@@
 \caption{Conforming to the \RAPstar\ methodology, the abbreviated summary above was generated automatically, by using \texttt{make data}.
+A full list of \texttt{make} options is generated by \texttt{make} or \texttt{make help} when run on the Unix command line.
 %(Saying \texttt{make help} would generate a full summary on the command line.)

### 2023-02-27 programs/runMathematicaNotebook.sh

@@ -1,0 +1,56 @@
+#!/bin/sh
+#
+# @file runMathematicaNotebook.sh
+#
+# Extract code from a Mathematica notebook and run it using wolframscript
+#
+# Usage: runMathematicaNotebook.sh notebook.nb
+#
+# Based on: https://mathematica.stackexchange.com/a/31157
+# from https://gist.github.com/tueda/9281f0b894eef9cad09d7e25d48d3a7a 
+# Unlike the original on stackexchange, this shell script 
+#     - it runs a Mathematica notebook but does not update it
+#     - it runs the notebook for its side-effects, such as any Export[files,results]
+#
+
+WOLFRAMSCRIPT=wolframscript
+script=tmp-mathematica-script
+
+if [ $# != 1 ]; then
+  echo "Usage: $0 file.nb" >&2
+  exit 1
+fi
+
+cat <<'END' >$script
+Begin["shellScriptContext`"];
+outputCellPattern = Cell[_, "Output", ___];
+cellEval[Cell[b_BoxData, "Input", rest___]] :=
+  Cell[
+    CellGroupData[{
+      Cell[b, "Input", rest],
+      Cell[BoxData@ToBoxes@(ToExpression @@ b), "Output"]
+    }, Open]
+  ];
+
+cellEval[x : (_CellGroupData | _Cell | _Notebook)] := cellEval /@ x;
+cellEval[l_List] := cellEval /@ Cases[l, Except@outputCellPattern, {1}];
+cellEval[somethingElse_] := somethingElse;
+
+evaluateNotebook[notebook_] := cellEval@Get[notebook];
+
+nbFullFileName = FileNameJoin[{Directory[], $ScriptCommandLine[[1]]}];
+
+End[];
+
+shellScriptContext`evaluateNotebook[shellScriptContext`nbFullFileName];
+
+END
+
+echo Run Mathematica script to evaluate notebook $1 for its side-effects, like data or image Exports
+$WOLFRAMSCRIPT -f $script -script $1
+err=$?
+
+#rm $script
+
+exit $err
+

### 2023-02-27 programs/runMathematicaNotebook.sh
Tidied up programs and scripts

@@ -50,7 +50,8 @@ echo Run Mathematica script to evaluate notebook $1 for its side-effects, like d
 $WOLFRAMSCRIPT -f $script -script $1
 err=$?
 
-#rm $script
+echo Remove tmp file $script
+rm $script
 
 exit $err
 

### 2023-05-12 Untitled-2.tex
Revisions for Computer Journal

--- 2023-02-20 response-to-referees/response 30 Jan 2022.tex
@@
 \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{a4paper}                   % ... or a4paper or a5paper or ... 
+\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
 %\geometry{landscape}                % Activate for for rotated page geometry
@@
 \usepackage{amssymb}
-\usepackage{epstopdf}\usepackage{url}\usepackage{xcolor}
+\usepackage{epstopdf}
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newcount\noten
-\noten=1
-\long\def\note#1{
-
-\color{red}{\begin{tabular}{r|p{4.7in}}
-\hphantom{\hskip .4in}\ifnum \noten<10 \hphantom{0}\fi{\sf \the\noten}& #1
-\end{tabular}}\color{black}
-\advance\noten by 1}
-
-\title{COMPJ-2022-05-0325: Response to referees}
-\author{Harold Thimbleby}
+\input paper-seb-macros.tex
+\title{Brief Article}
+\author{The Author}
 %\date{}                                           % Activate to display a given date or no date
@@
 
-\noindent 
-\color{red}
-%\begin{quote}\emph{I have sent an email requesting a short extension please --- I've finished revising the main paper, but not yet finished the Supplemental Material (it has been Christmas and New Year~\ldots).
-I should finish early next week.
-I uploaded these notes on my revisions in case the computer system destroys the old submission if I do not meet the deadline.}\end{quote}
-
-\noindent Thank you for this opportunity to respond to reviewers' comments. Almost all of my responses, detailed below, have resulted in helpful improvements to the paper.
-
-I have written my responses to the reviewers' comments in red. If it helps in referring to them in future correspondence, I have numbered my responses.
-
-In addition to responding to reviewers' comments, I have also edited the paper to improve its clarity, structure, and where possible to reduce its length.
-The writing is now tighter, and in particular the reviewers' repeated concern about ``opinion'' has been addressed by appropriate rewriting --- mostly replacing ``shoulds'' with ``coulds'' and other ways of making clear the what was an unqualified opinion is a suggestion for further work.
-
-\color{black}
-
-\vskip .5cm\hrule height 2pt \vskip .5cm
-
-\noindent
-Dear Professor Thimbleby,
-
-COMPJ-2022-05-0325 -- Maturing the computational sciences
-
-I am writing to you with regard to your paper COMPJ-2022-05-0325, which you submitted to The Computer Journal. It has now been examined by referees and by a member of the editorial board.
-
-In view of their recommendations, I would appreciate it if you would revise your paper, carefully taking into account all the comments, and resubmit it by 03-Jan-2023 via the Manuscript Central website at \url{http://mc.manuscriptcentral.com/compj}
-
-Please also make sure a DATA AVAILABILITY STATEMENT is included in the manuscript.
-Please check\\\url{https://academic.oup.com/comjnl/pages/Manuscript_Preparation_Submission} and \url{https://academic.oup.com/journals/pages/authors/preparing_your_manuscript/research-data-policy#data2}
-
-To submit your revision, please go to *** PLEASE NOTE: This is a two-step process. After clicking on the link, you will be directed to a webpage to confirm. ***
-
-\url{https://mc.manuscriptcentral.com/compj?URL_MASK=}\\\url{ae7f1f867b624f72b0d553807c8d9146}
-
-Please also submit a list of the changes you have made in response to each of the referees' comments. Your paper will then be re-examined and a decision will be made about whether it may be published.
-
-Thank you very much.
-
-Best regards,
-
-Deputy Editor
-
-The Computer Journal
-
-\newpage
-
-\section{REVIEWERS' COMMENTS:}
-
-\subsection{Reviewer: 1}
-
-The paper has a misleading title -- I think it would be better if it focussed on software and people -- or maybe on improving reliability.
-\note{Thanks to this concern, I have improved and changed the title to ``Improving science that uses code.''}
-
-Having a vision for the future is good, but it needs to be matched with a set of practical changes to get there. I don't see that here, just a vision of an alternative world.
-
-I think this paper is too long and too complex.
-\note{As well as structural improvements, I have also added a short table of contents, which makes the structure of the much paper much clearer than before, also helping address this reviewer's concern.
-In fact, I used the new table of contents to guide some streamlining throughout the paper!}
-
-Some of it is opinion, some of it is observation, and some of it is advertising. And these parts are interspersed in the paper.
-
-In table 2, GitHub is listed as both a code repository and a data repository, which is confusing.
-It might be better to talk about the different meanings of repository, such as in the git sense vs in the library/archive sense.
-\note{GitHub can be used as a repository for anything (such as \emph{The Turing Way: A Handbook for Reproducible Data Science}, which is a massive project with 419 authors collaborating.
-I have therefore not changed table 2.}
-
-There is likely a correlation between good software engineering practices and good code quality, but I am less sure that there is a causation, or that good quality code cannot be written without good software engineering practices.
-
-Re ``A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model is transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to.
--- this reminds me of \url{https://doi.org/10.12688/f1000research.25561.1} \note{I had thought of citing this f1000 paper, but it reports on a meeting duplicating the earlier work of TOP, which I do cite; furthermore this f1000 paper  just makes a list of questions --- unlike the TOP papers which have concrete proposals.}
-
-There is likely a difference between how statistical practices are described concisely in a paper (page 3 line 54) and how software can described, which is probably better done through the software itself than through text about the software.
-\note{Why?
-Just as a statistical transformation might be described by name in wording like ``Bonferroni adjusted estimated mean difference~\ldots'' as my paper says, an algorithm can be described by precise wording like ``we use Knuth's (1977) generalization of Dijkstra's (1959) algorithm for directed hypergraphs~\ldots''}
-
-Re the comments on Ferguson in the intro to Section 3, this is correct, but it's not the end of the story.
-An effort was made to improve the code, which I believe was successful.
-\note{This effort was (and is) cited and discussed in the paper.}
-
-I don't think this changes the argument of this paper, but the paper should acknowledge that we do know how to fix these things, we just don't do it until we have to (due to limited resources, no culture that requires it, incentives, etc.) \note{I disagree.
-The Ferguson case involved fixing a particular case of very bad code after the event of its unravelling and Ferguson's admissions (as cited); it is not fixing the problem the paper addresses, which is fixing the persistent widespread use of bad code in science.}
-
-\note{The reviewer says ``know how to fix these things, we just don't do it until we have to'' so we agree the issue is to create a route from the problem to a solution; I think my paper provides evidence that the problem is real and widespread, and lays out many ways to start doing this.}
-
-I don't agree that C is not a dependable language, nor that it is not portable.
-The issues raised here are not unique to C, but are common to almost all languages in usage in computational science.
-\note{The whole industry-standard MISRA C approach is based on the fact that the community recognizes that C is not dependable, and MISRA and it's work is already cited.
-The reason for mentioning C in particular is that it was the language chosen by Ferguson, as explained in the case study; however I agree the unreliability issues are common across many languages, and I now make this point clearer.}
-
-These statement are philosophical in this context. \note{Hmm. Only if you believe that C is dependable, a view which I argue against!}
-
-The RAP* discussion seems too detailed for this paper, and the supplementary material is too long -- this could be published separately (or as a pre-print) and referenced here.
-\note{The length of the RAP discussion in the paper main body seems right to me, and is supported by Reviewer 2's comments.
-(Please  note that I changed my term RAP* to RAP+ as I feel this much better indicates that RAP+ is an extension of the original RAP than RAP*.)}
-
-\note{I agree that the supplemental material is long, but it will be online so I am not sure that this is a problem.
-The  supplemental material is divided into several topics, and includes a table of contents, so readers can navigate it easily.}
-
-Re Software Engineering Boards, what about other boards for other types of work: perhaps we should have data collection boards, statistical review boards, reproducibility boards, etc.
-But I don't think creating new collections of gatekeepers is going to be either practical or useful.
-\note{I hadn't thought of it as gatekeeping, thanks for raising the concern.
-I've now added a paragraph and a citation to an article in \emph{Nature\/} to explain that the idea is \emph{not\/} gatekeeping.}
-
-Better standards and better education make sense to me, but more enforcers do not.
-\note{There is no sense, and no sense expressed in the paper, in which Software Engineering Boards are gatekeepers or enforcers.
-The paper makes an explicit analogy with Ethics Boards, which are widely accepted and work well.
-I don't see any problems, then, nor any ``thin end of a wedge'' problem the reviewer characterizes here.}
-
-I don't think the comment about methodological statements at the start of 5.b is correct.
-Most venues in which I publish do not require all or even many of these statements from authors.
-\note{I changed ``most journals'' to ``Many science journals'' (although I know of none that do not require some such statements); in addition, I have cross referenced to the paper's discussion of PRISMA, which itself is a methodology that, if used, authors are expected to declare and comply with.
-PLOS ONE is an example of a major journal that makes a list of similar methodologies and when the journal expects compliance; I have now made a note of this point in the paper.}
-
-Much of this section and 5.c also seem to follow the advertising of RAP* with statements that not only is it good, but it should be required.
-This is again opinion, and not justified.
-Would this help?
-maybe.
-Is it the best solution?
-I have no idea.
-\note{I disagree that the benefits of RAP and RAP* are opinion; I provide reasoned arguments, and examples of the problems RAP and RAP* avoid.}
-
-\note{I note that RAP and RAP+ (its new name) are very good solutions, and I am of course explicitly drawing on the existing very favorable literature on RAP\@.
-There may be better solutions available in the future, but until then they are the best known solutions.
-Furthermore, ``the best is the enemy of the good'' --- it is routine to publish a good idea, as this paper proposes, and the scientific community can then seek ways to refine the idea.
-The point is, a good idea has to be published first.}
-
-The second paragraph in 5.e is not at all correct in my opinion.
-I think this is an unrealistic view of science.
-\note{I briefly considered citing Einstein, who commented on this point, and others in my support, but I've deleted the comments in the paper, as the paper's core message is not in fact affected either way.}
-
-The acknowledgements statement isn't grammatically correct. \note{Thanks for spotting this! Problem fixed.}
-
-After reading the paper, the only thing I feel that I have learned is that the author has strong opinions about how bad things are and how the world should work that would be better.
-But I'm not convinced that this is the correct analysis and solution, or even if it is, how we would get to this future world from today's world in practice.
-\note{The new Background section I have added in my revision makes it clear that my opinions are closely aligned to, and often summarize, substantial work by others.}
-
-\subsection{Reviewer: 2}
-
-This manuscript on the state of computational science today is unusual in several respects.
-It is mostly an opinion paper, advocating for more rigor in the use of computational methods in scientific research.
-\note{As mentioned in my comments to Reviewer 1, my opinions are rigorously justified, and traceable back to peer reviewed literature.
-It is of course an opinion that scientific research should be more rigorous, but I am unaware of significant dissent to this opinion.
-(I admire Paul Feyerabend, e.g., his \emph{Against Method}, but the paper does not need to get into the philosophy of science.)}
-
-In doing so, it provides a rather exhaustive list of arguments, to the point of almost being a PhD-level course on the subject.
-\note{I am aware --- and I hope --- that scientists, not just computer scientists and software engineers, who can benefit from my analysis will read the paper.
-I therefore felt it important to write clearly, and to provide as much background and direct information as possible.
-In particular, the Supplement provides a lot of useful material and pointers to further reading.}
-
-Finally, it contributes an analysis of code-and-data sharing in the literature to the arguments, making it in part a research paper.
-The writing style is much more colloquial than a typical scientific article, making the manuscript lengthy but also very readable.
-Whether or not these particularities are acceptable in The Computer Journal is a matter of editorial policy, which is not for me to judge.
-\note{Thank you for recognizing the readability of my paper. I tried hard to make the paper accessible, as I hope many scientists, not just computer scientists/non-programmers, will read the paper.}
-
-My following review mostly treats the manuscript as an opinion paper.
-
-Opinion papers being a matter of opinion, I'll do my best to judge the pertinence of the argument independently of whether I agre with them.
-But I'll state my prejudices right away: I do agree with most of the observations and conclusions in this paper.
-\note{Thanks!}
-
-The most important omission in my opinion is the lack of context.
-Many past and present efforts are aligned with the author's objectives and have made and sometimes implemented similar propositions, but the manuscript does not mention them.
-\note{I did, but the mentions were scattered around in the paper.
-I have now introduced a focused Background section (with additional references to substantiate a more powerful argument) early in the paper collecting everything together to address this helpful point.}
-
-The publication of scientific software in source code form has been advocated as early as 1969 by K.V.
-Roberts (\url{https://doi.org/10.1016/0010-4655(69)90011-3}), using some of the same arguments that are made in the present manuscript.
-The next call for publishing software and data I am aware of is by Jon Claerbout in 1992 (\url{https://doi.org/10.1190/1.1822162}).
-\note{The reviewer makes some pertinent points here, and I have therefore added material to the Background section to provide an overview of the relevant literature.
-In particular, I note that concern for publishing code goes back even further than the reviewer mentions; I cite work as far back as 1958, in fact to \emph{CACM\/}'s first issue of its very first volume.}
-
-Since then, the pace at which similar calls for more rigorous treatment of software and computation have been made has accelerated steadily.
-More importantly, there are many initiatives that implement parts of the author's recommendations, and it would be more helpful to discuss how their impact and utility can be improved than to pretend they don't exist and propose new initiatives to be started from scratch.
-\note{Yes, it was an oversight that there was no stand-alone, clear section in the paper discussing prior work: it was scattered around in various citations.
-I have therefore now added a new section, Background.}
-
-Some examples:
-
-\begin{itemize}\raggedright\item The Software Sustainibility [sic] Institute (briefly mentioned in the supplementary material) and CODECHECK in the UK, and CASCaD in France, resemble the author's SEBs in many ways.
-More generally, there is an ongoing movement in many countries for establishing the ``research software engineer'' as a recognized role in scientific research, which shows that the need for better software engineering is increasingly recognized.
-\note{Rather than cite such an almost unlimited list of initiatives, I discuss some large studies and the key TOP initiative.}
-\item  New on-line journals such as The Journal of Open Source Software or ReScience practice reviewing of scientific code at various levels (and find that this is not as easy to do as it may seem).
-\note{I hadn't heard of \emph{ReScience C\/} (and \emph{ReScience X\/}) --- thanks.
-I now cite them (in the new Related work section) and what they are achieving.}
-\item  Various courses, e.g.
-by Software Carpentry and Data Carpentry, and also several on-line courses, teach what the author calls RAP (a name I have never seen before) to an audience of mostly PhD students.
-\note{The training courses I am aware of (including all those I searched for and found on the internet) are very low level introductions to coding, and really inadequate for coding to support reliable peer reviewed research.}
-\end{itemize}
-
-I find rather few completely new arguments and proposals in the present manuscript, but it has the merit of being exhaustive, detailed, and combining advocacy with practical advice.
-
-In the following, I will concentrate on the new aspects.
-
-An important original contribution is the analogy between the introduction of computation and statistics into mainstream research.
-It deserves publication on its own because of its pertinence.
-The analogy is very fitting, and helps both in understanding the inadequacy of today's scientific code management and in improving the situation.
-Moreover, the rigorification of statistics has well advanced to the point that most scientists now realize its benefits.
-That's a good motivation to launch a similar campaign for more rigor in computation.
-\note{Thanks for these encouraging comments.
-The important point that scientists now recognize the value of rigorous statistics motivates considering rigor in computation is now mentioned in the introduction (section 1).}
-
-Another original contribution of the proposal of Software Engineering Boards, but I have to say that I don't find the arguments very convincing.
-I do agree about the necessity of introducing software engineering practices into computational research, and to check publications for respecting best practices in the field.
-But the implication of software engineers can take many forms (as part of research teams, as experts in peer review, \ldots) and I don't find the arguments for the specific institution ``SEB'' very compelling.
-The analogy with ethics boards is weak, as the numerous differences listed by the author suggest.
-\note{The point of the analogy is that ethics boards \emph{do\/} work and are recognized as available --- if not required --- in almost all research institutions.
-Ethics Boards are routine, and are not onerous as this reviewer suggests.
-I added a few extra references here, including an analogous proposal for methodology boards --- in a recent \emph{Nature\/} paper, which also explores their pros and cons.}
-
-A more general comment about the author's discussions of software engineering in science is the lack of any reference to the particularities of scientific software, compared to software in business or industrial contexts, which is the focus of most software engineers today.
-And even within the category of scientific software, there are vast differences.
-An epidemiology simulation (software as a reliable tool) meant to inform public policies must be treated differently from a prototype implementation of a new data analysis technique (software as an executable scientific method).
-Both need more rigor, but not in the same place.
-\note{I do think this is an obvious point.
-For example, scientific coding is obviously different to game coding, e-commerce, and so on.
-Discussing the varieties of code seems to me to be beyond the scope of the paper.}
-
-As an example, the author heavily insists (in the supplementary material) on the use of formal methods.
-They are indeed the appropriate tools in situations where the conformance of an efficient implementation with a known-to-be-good specification is the bottleneck in establishing correctness.
-But often it is the specification itself that is incomplete, informal, or incorrect.
-A common situation in computational research is a rapid development cycle in which scientific questions, specifications, and implementations are frequently updated.
-It is in my opinion an open question if formal methods are helpful in this setting.
-They add rigor to one step of the cycle, but at such a high cost (with the current state of the art at least) that other steps risk being neglected.
-Illustration: a piece of code formally proven correct to a rather deep level: \url{https://doi.org/10.1016/j.camwa.2014.06.004}.
-Impressive work, but also a Herculean effort, for a single relatively small algorithm.
-\note{Indeed!
-I note that the paper this reviewer cites is 25 tight technical pages, and that its authors admit that their approach is beyond many coders.
-I have therefore now clarified (throughout) that a formal methods \emph{mindset\/} is required, not a total formal proof of code as in this paper.
-For example, as now explained particularly in the Supplement, you could require explicit invariants, assertions and other formal ideas in code: this is a formal mindset in action, but avoids ``the best is the enemy of the good'' the reviewer seems to fear.}
-
-This illustrates a more general problem with the author's recommendations (in the supplementary material): there is no clear path for getting from today's research processes to more rigorous processes that include all of the author's ideas.
-Implementing all of them is such a huge effort that at best it will suffer the same fate as climate change policies: everybody agrees it should happen, but the amount of work requires is so paralyzing that nothing happens at all.
-A grand view, such as outlined in this manuscript, needs to be complemented by a list of small initial steps that most scientists recognize as achievable.
-\note{I think here the reviewer has fallen  for the perfect-is-the-enemy-of-the-good interpretation!
-I hope my rewording throughout dispels this problem.}
-
-Another interesting original contribution is the RAP* method. \note{I've renamed it RAP+, as this is more suggestive of an extension to RAP itself, which the name RAP* wasn't.}
-
-It is not completely original in the sense that many people have practiced variants of it in the past, but I haven't seen it named before, nor discussed to the extent presented in the manuscript.
-The author's analysis of the literature also serves as a case study for RAP*.
-The discussion of RAP* is a welcome transition from ``how to save research from computers'' (the main topic of the manuscript) to ``how to improve research using computers''.
-I hope to see more work in this direction in the future.
-\note{Thanks for these comments.
-Of course, RAP* is on the route towards what I envisage, and therefore these welcome comments contradict the reviewer's  worries expressed in their previous paragraph!}
-
-Finally, in case the editors of The Computer Journal decide that this lengthy opinion-paper-plus-course does not fit their editorial policies, a few suggestions for shorter articles extracted from this material that I think would be received favorably by the scientific community:
-
-\note{I put all the helpful ``course'' material as reference in the Supplement, which will be online rather than part of the paper itself.}
-
-\note{I am not really clear why the reviewer focuses apparently negatively on the paper being opinion.
-Surely, Tony Hoare's paper on quicksort was just opinion, and wasn't even the best way of sorting under many assumptions --- but that doesn't stop it being a brilliant and now classic paper.
-So far as I can see, all my opinions are argued, justified or readily traceable back to the literature.
-I'd be happy to fix any specific unsupported opinion of concern but the review is not specific.}
-
-\begin{enumerate}\raggedright\item
-A much shorter and focused opinion paper on the topic: ``We made statistics in research more rigorous, now let's do the same for the use of software''.
-\note{This is a good idea, but of course computation can do a lot more than statistics, so I don't think such a paper, properly done, would be much different from my current offering --- especially as there seems to be very little literature making the same points one could thus build on much more concisely.}
-\item  A review of software engineering techniques, explaining their expected benefits for computational science.
-The main audience would be research software engineers.
-\note{I put all of this material in the Supplement, which I imagine will be online so its length is not material.}
-\item  A tutorial on the RAP* method, using the data analysis from the present manuscript as a case study. \note{This is already provided in the supplement.}
-\end{enumerate}
-
-Konrad Hinsen, CNRS, France
-(Note to editor: I have decided to sign all my reviews. Please don't remove my signature.)
-
-\note{I very much appreciate your disclosure. Thanks for all your work. Are you happy being named in the paper's Acknowledgements?}
-
-\subsection{Reviewer: 3}
-
-David Spiegelhalter's name is misspelt ``Speigelhalter'' throughout. The first time it is misspelt ``Speigelhater''. \note{Oh no! Thanks for this correction. Spelling now fixed throughout.}
-
-\section{EDITORIAL BOARD RECOMMENDATION:}
-
-Associate Editor
-
-Comments to the Author:
-
-Please address the comments raised by the reviewers, and list what changes you have or haven't made with some justifications.
-
-\vskip 1cm\hrule
-
-\section{REFERENCES:}
-
-General. Please construct references in the following style:
-
-\begin{enumerate}\raggedright\item The references should appear in the text in numerical order, as should the reference list. \note{Now done.}
-\item  Where there are more than 10 contributing authors please list only the first author and use et al. \note{Now done.}
-\item  The following format for references should be followed: \note{Now done.}
-\end{enumerate}
-
-Journal article:
-Galton, A. (1992) Logic as a formal method. Comp. J., 35, 431-440.
-(author name, year, title of paper, title of journal, volume number, page numbers)
-
-Book:
-Hogger, C. (1990) Essentials of Logic Programming. Clarendon Press, Oxford.
-(author name, year, title of book, publisher, publisher location)
-
-Chapter:
-Harel, D. (1984) Dynamic logic. In Gabbay, D. and Guenther, F. (eds), Handbook of Philosophical Logic. D. Reidel, Dordrecht.
-(author name, year, title of chapter, editors names, title of book, publisher, publisher location)
-
-Conference:
-Crochemore, M. and Verin, R. (1997) Direct construction of compact directed acyclic word graphs. Proceedings of CPM 97, Cambridge, MA, 12-14 August, pp.~192-211. Springer-Verlag, Berlin.
-(author name, year, title of conference, place of conference, date of conference, page numbers in proceedings, publisher, publisher location)
-
-Technical documents:
-ITU-T. Z.500 (1997) Framework on formal methods in conformance testing. International Telecommunications Union, Geneva, Switzerland.
-(document number, year, document title, publisher, publisher location)
-
-Could you please check your references carefully (in particular, proceedings: location of conference, date of conference, publisher name and location\ldots) and make sure that they comply with the above format?
-
-\color{red}
-\section{SUMMARY OF CHANGES}
-In addition to the changes in response to the reviewers' specific comments as listed above, I have uploaded a \texttt{changes.pdf} file which is an automatically generated difference listing of the changes I have made.
-Text in blue is new; text in red is deleted.
-
-The diff process I used (using \texttt{latexdiff}) doesn't read included files and has other irritating bugs, so, for instance, changes to data (including changes in tables) are not shown, and there are various typesetting errors.
-Regardless of the details, it shows how almost all of the paper's text has been substantially improved, usefully including showing deleted text that I or the referees considered unnecessary.
-
-The added table of contents I hope shows the reviewers the revised paper is now much clearer and more clearly structured; I think the table of contents will be helpful for journal readers too, but it does not need to be included in the final paper if it is not considered to fall within the journal's accepted style.
-
-The supplementary material has minor corrections and simplifications throughout, but no substantial changes comparable to those made in the main paper.
+\begin{tabular}{cc}
+\commarise{123456789}
+&
+\digitBlock{123456789}
+\end{tabular}
 

### 2023-05-12 generated/assessments.tex
Revisions for Computer Journal

@@
-\citenum{ref-1} & On request\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
-Documented overview, but only trivial comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''``''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request''\flagStyle{ } & Poor commenting, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team''\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-1} & On request.\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations.''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
+Documented overview, but only trivial comments.
+Assertions in code only relate to UI.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}$\sf C_a$\hskip 3pt{}}\\
+\citenum{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available.''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available.''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors).\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request.''\flagStyle{ } & Poor commenting, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team.''\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
 \citenum{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
-However, the data are available from the corresponding author on reasonable request''\flagStyle{ } & Empty GitHub repository: ``Code coming soon\ldots'' it says\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+However, the data are available from the corresponding author on reasonable request.''\flagStyle{ } & Empty GitHub repository: ``Code coming soon \ldots'' paper says.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
 \citenum{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
 Data from UCSF are not available at this time as they have not been legally certified as being De-Identified, however, this process is underway and the data may be available by the time of publication by contacting the authors.
-Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply''\flagStyle{ } & Basic documentation, very little comment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request''\flagStyle{ } & Trivial comments, no documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available''\flagStyle{ } & ``We used only free and open-source software'' some of which is unspecified\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{paper-usesRMarkdown} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
-\citenum{ref-14} & Data directly written into program code\flagStyle{ } & Basic Matlab with routine comments\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply.''\flagStyle{ } & Basic documentation, very little comment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request.''\flagStyle{ } & Trivial comments, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available.''\flagStyle{ } & ``We used only free and open-source software'' --- some of which is unspecified.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments only.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{paper-usesRMarkdown} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\citenum{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
 \citenum{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
-Currently (apparently) for private view\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
+Currently (apparently) for private view.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
 Paper describes using two contrasting methods to help confirm correctness, ``As an additional check, I also coded the calculation of D based on a probabilistic approach, using genotype frequencies in each population to calculate the expected frequencies of each possible two-genotype combination (electronic supplementary material, table S1).
-Essentially identical results were obtained.'' but the contrasting method is not available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-16} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-17} & On request\flagStyle{ } & R lightly commented\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ethics-paper} & No data required\flagStyle{ } & Unrunnable incomplete code fragment\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-19} & Data embedded in PDF\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-20} & Data available on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-21} & Partial data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including manual\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_{+}$\hskip 3pt{}}\\
-\citenum{ref-22} & No data required\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-23} & Data cited, not all available\flagStyle{ } & Trivial documentation\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{example-numerical-error} & On Figshare\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
-Helpful features to make usable for third parties\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-25} & Data on Dryad\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-26} & Data on various web sites\flagStyle{ } & No code available\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-27} & Data on request\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-28} & Data on request\flagStyle{ } & Python scripts can be requested\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-29} & Unspecified location on large website requiring registration\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-30} & Available to researchers who meet criteria for access to confidential data\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-31} & Data access conditional on approved study proposal\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-32} & Unspecified locations on several large websites\flagStyle{ } & Python used and apparently GitHub, but --- an oversight?
---- no code is available\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+Essentially identical results were obtained.'' --- but the contrasting method is not available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-16} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-17} & On request.\flagStyle{ } & R, lightly commented.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ethics-paper} & No data required.\flagStyle{ } & Unrunnable incomplete code fragment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-19} & Data embedded in PDF.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-20} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\citenum{ref-21} & Partial data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including model's manual.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\citenum{ref-22} & No data required.\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-23} & Data cited, not all available.\flagStyle{ } & Trivial documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{example-numerical-error} & On Figshare.\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
+Helpful features to make usable for third parties.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-25} & Data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-26} & Data on various web sites.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-27} & Data on request.\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-28} & Data on request.\flagStyle{ } & Python scripts can be requested.\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
+\citenum{ref-29} & Unspecified location on large website requiring registration.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\citenum{ref-30} & Available to researchers who meet criteria for access to confidential data.\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\citenum{ref-31} & Data access conditional on approved study proposal.\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\citenum{ref-32} & Unspecified locations on several large websites.\flagStyle{ } & Python used and apparently on GitHub, but --- an oversight?
+--- no code is available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\

### 2023-05-12 generated/info-for-main.tex
Don't delete .bbl files
Revisions for Computer Journal

@@
-\newlabel{supplementary-material-makefiles}{{11}{34}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{on-code-data-publication}{{12}{37}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{RAP-diagrams}{{12}{40}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{excelFigure}{{2}{40}{AAA}{BBB}{CCC}}
-\newlabel{supplementary-Spiegelhalter-section}{{13}{40}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{survey-section}{{14}{43}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{supplementaryTableRepoSummary}{{4}{45}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{supplementary-journal-policies-section}{{14}{48}{DUMMYA}{DUMMYB}{DUMMYC}}
-\newlabel{restrictive-policies}{{14}{52}{DUMMYA}{DUMMYB}{DUMMYC}}
-\def\MaxMainPaperCitationNumber{0}
+\newlabel{supplementary-material-makefiles}{{9}{42}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{on-code-data-publication}{{10}{45}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{RAP-diagrams}{{10}{48}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{excelFigure}{{2}{48}{AAA}{BBB}{CCC}}
+\newlabel{supplementary-Spiegelhalter-section}{{11}{48}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{survey-section}{{12}{51}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementaryTableRepoSummary}{{4}{53}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{supplementary-journal-policies-section}{{12}{56}{DUMMYA}{DUMMYB}{DUMMYC}}
+\newlabel{restrictive-policies}{{12}{60}{DUMMYA}{DUMMYB}{DUMMYC}}
+\def\MaxMainPaperCitationNumber{126}

### 2023-05-12 generated/legend.tex
Revisions for Computer Journal

@@
-\begin{tabular}{lp{4.5in}}
-\flagStyle{$\sf P_c$}&Journal has a code policy (see section~\ref{supplementary-journal-policies-section})\\
-\flagStyle{$\sf P_{{\mbox{\scriptsize c-breach}}}$}&Paper breaches journal code policy (see section~\ref{supplementary-journal-policies-section})\\
-\flagStyle{$\sf R_c$}&Paper uses a code repository (e.g., GitHub)\\
-\flagStyle{$\sf R_{\mbox{\scriptsize c-empty}}$}&Code repository contains no code\\
-\flagStyle{$\sf R_d$}&Paper uses a data repository (e.g., Dryad, Figshare, GitHub)\\
-\flagStyle{$\sf S_{\mbox{\scriptsize NONE}}$}&No code available at all (note: code is not expected for standard models, systems or statistical methods)\\
-\flagStyle{$\sf S_p$}&Paper says source code is available in principle\\
-\flagStyle{$\sf S_{+}$}&Paper or URL provides source code\\
-\flagStyle{$\sf S_{{\mbox{\scriptsize rigorous}}}$}&Evidence that source code was developed rigorously\\
-\flagStyle{$\sf S_{\mbox{\scriptsize tested}}$}&Evidence that source code has been run with a clean build and tested\\
-\flagStyle{$\sf S_{{\mbox{\scriptsize tools}}}$}&Evidence of any tool-based development\\
-\flagStyle{$\sf S_{{\mbox{\scriptsize open source}}}$}&Team or open source development\\
-\flagStyle{$\sf S_{{\mbox{\scriptsize otherSE}}}$}&Other evidence of good practice; see details in summary table\\
-\flagStyle{$\sf C_0$}&Code has no non-trivial comments\\
-\flagStyle{$\sf C_1$}&Code only has trivial comments (e.g., copyright)\\
-\flagStyle{$\sf C_2$}&Helpful comments explaining code intent, rather than rephrasing the code\\
-\flagStyle{$\sf C_{+}$}&Code has substantial, useful comments and documentation\\
+\begin{tabular}{lrp{4.5in}}
+&\bfseries{Paper}&\\\bfseries{Flag}&\bfseries{count}&\bfseries{Meaning}\\ \hline 
+\flagStyle{$\sf P_c$}&26&Journal has a code policy (see section~\ref{supplementary-journal-policies-section})\\ 
+\flagStyle{$\sf P_{{\mbox{\scriptsize c-breach}}}$}&11&Paper breaches journal code policy (see section~\ref{supplementary-journal-policies-section})\\ 
+\flagStyle{$\sf R_c$}&10&Paper uses a code repository (e.g., GitHub)\\ 
+\flagStyle{$\sf R_{\mbox{\scriptsize c-empty}}$}&1&Code repository contains no code\\ 
+\flagStyle{$\sf R_d$}&9&Paper uses a data repository (e.g., Dryad, Figshare, GitHub)\\ 
+\flagStyle{$\sf S_{\mbox{\scriptsize NONE}}$}&12&No code available at all (note: code is not expected for standard models, systems or statistical methods)\\ 
+\flagStyle{$\sf S_p$}&8&Paper says source code is available in principle\\ 
+\flagStyle{$\sf S_{+}$}&12&Paper or URL provides source code\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize rigorous}}}$}&0&Evidence that source code was developed rigorously\\ 
+\flagStyle{$\sf S_{\mbox{\scriptsize tested}}$}&0&Evidence that source code has been run with a clean build and tested\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize tools}}}$}&0&Evidence of any tool-based development\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize open source}}}$}&0&Team or open source development\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize otherSE}}}$}&1&Other evidence of good practice; see details in summary table\\ 
+\flagStyle{$\sf C_0$}&10&Code only has comments unrelated to code intent (e.g., copyright)\\ 
+\flagStyle{$\sf C_1$}&1&Code only has trivial or obvious comments\\ 
+\flagStyle{$\sf C_2$}&3&Helpful comments explaining code intent, rather than rephrasing the code\\ 
+\flagStyle{$\sf C_3$}&2&Code has substantial, useful comments, and documentation\\ 
+\flagStyle{$\sf C_a$}&1&Code in repository uses assertions\\ 
+\flagStyle{$\sf C_c$}&0&Code in repository uses pre- or post-conditions or similar\\ 
 \end{tabular}

### 2023-05-12 generated/over-fitting-code-section.tex
Mainly getting mathematica+make to work together
Getting make and mathematica to run together properly
Tidied up programs and scripts
Revisions for Computer Journal

--- 2023-02-20 generated/over-fitting-code-section.tex
@@
 \noindent
-Poor code undermines scientific quality: poor code can generate plausible results from \emph{any\/} data, including even fictional science.
+Poor code undermines scientific quality: poor code can generate plausible and possibly misleading results from \emph{any\/} data, including even fictional science.
 A temptation is that %artificial world \cite{sciences-artificial}, built on
 adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena. 
+% wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed!
+\def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
+\def\figureStarDetails{*}
+
+% wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed because the text also ends up in notebook cells!
 \def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
@@
 In modeling terms, superficially successful computer programs tend to \emph{over-fit\/} phenomena \cite{over-fit}.
-Instead of using code to test the models, we tinker with the code adapting it closer and closer to our prejudices.
+Instead of using code to test the models, we tinker with the code adapting it to be closer and closer to our prejudices.
 The code then apparently confirms our science, since we fitted it to our understanding.
@@
 The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases.
-Indeed, many of the programs in the survey used comment to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countRepos\ papers in the pilot survey that reported use of code repositories (covering \input generated/kloc.tex
+Indeed, many of the programs in the survey used comments to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countUsesVersionControlRepository\ papers in the pilot survey that reported use of code repositories (covering \input generated/kloc.tex
 \totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
@@
 
-\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
+%\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
 \vskip 1ex

### 2023-05-12 generated/repos.tex
Various corrections inspired by Konrad Hinsen
Mainly getting mathematica+make to work together
Revisions for Computer Journal

--- 2022-02-28 generated-repos.tex
@@
 \multicolumn{1}{|c|}{\bf and paper citation}&\multicolumn{1}{c|}{\bf of pages}&\multicolumn{1}{c}{\bf of files}&\multicolumn{1}{c}{\bf kLOC}&\multicolumn{2}{c|}{\bf bytes} \\ \hline\hline
-\reponame{AI-CDSS-Cardiovascular-Silo}{AI-CDSS-Cardiovascular-Silo} & \csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname & \commarise{206} & 143 & 64&Mb \\
-\reponame{blast-ct}{blast-ct} & \csname pagelength-blast-ct\endcsname & \commarise{44} & 2 & 87&Mb \\
-\reponame{covid-sim}{covid-sim} & \csname pagelength-covid-sim\endcsname & \commarise{229} & 25 & 734&Mb \\
-\reponame{lactModel}{lactModel} & \csname pagelength-lactModel\endcsname & \commarise{20} & 2 & 165&kb \\
-\reponame{LRM}{LRM} & \csname pagelength-LRM\endcsname & \commarise{125} & 8 & 2&Mb \\
-\reponame{manifold-ga}{manifold-ga} & \csname pagelength-manifold-ga\endcsname & \commarise{11} & 1 & \multicolumn{2}{c|}{---} \\
-\reponame{MetricSelectionFramework}{MetricSelectionFramework} & \csname pagelength-MetricSelectionFramework\endcsname & \commarise{44} & 4 & 236&kb \\
-\reponame{PENet}{PENet} & \csname pagelength-PENet\endcsname & \commarise{115} & 8 & 3&Mb \\
-\reponame{philter-ucsf}{philter-ucsf} & \csname pagelength-philter-ucsf\endcsname & \commarise{1987} & 13 & 32&Mb \\
-\reponame{PostoperativeOutcomes\_RiskNet}{PostoperativeOutcomes.RiskNet} & \csname pagelength-PostoperativeOutcomes.RiskNet\endcsname & \commarise{1} & --- & \multicolumn{2}{c|}{---} \\
-\reponame{SiameseChange}{SiameseChange} & \csname pagelength-SiameseChange\endcsname & \commarise{5} & 1 & 1&kb \\
-\hline \multicolumn{1}{|r|}{{\bf Average} ($N=11$)}&\makeAverage{\the\gitPages}{11}&253& 19& 84&Mb\\
+\reponame{AI-CDSS-Cardiovascular-Silo}{AI-CDSS-Cardiovascular-Silo} & \csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname & \digitBlock{206} & 143 & 64&Mb \\
+\reponame{blast-ct}{blast-ct} & \csname pagelength-blast-ct\endcsname & \digitBlock{54} & 3 & 238&Mb \\
+\reponame{covid-sim}{covid-sim} & \csname pagelength-covid-sim\endcsname & \digitBlock{229} & 25 & 734&Mb \\
+\reponame{lactModel}{lactModel} & \csname pagelength-lactModel\endcsname & \digitBlock{20} & 2 & 165&kb \\
+\reponame{LRM}{LRM} & \csname pagelength-LRM\endcsname & \digitBlock{125} & 8 & 2&Mb \\
+\reponame{manifold-ga}{manifold-ga} & \csname pagelength-manifold-ga\endcsname & \digitBlock{11} & 1 & \multicolumn{2}{c|}{---} \\
+\reponame{MetricSelectionFramework}{MetricSelectionFramework} & \csname pagelength-MetricSelectionFramework\endcsname & \digitBlock{44} & 4 & 236&kb \\
+\reponame{PENet}{PENet} & \csname pagelength-PENet\endcsname & \digitBlock{117} & 8 & 4&Mb \\
+\reponame{philter-ucsf}{philter-ucsf} & \csname pagelength-philter-ucsf\endcsname & \digitBlock{1987} & 13 & 32&Mb \\
+\reponame{PostoperativeOutcomes\_RiskNet}{PostoperativeOutcomes.RiskNet} & \csname pagelength-PostoperativeOutcomes.RiskNet\endcsname & \digitBlock{1} & --- & \multicolumn{2}{c|}{---} \\
+\reponame{SiameseChange}{SiameseChange} & \csname pagelength-SiameseChange\endcsname & \digitBlock{5} & 1 & 1&kb \\
+% Numbers below are rounded, but more precise averages are 254.454545 -- 18824.000000 -- 97771458.363636 = 1075486042 / 11
+\hline \multicolumn{1}{|r|}{{\bf Average} ($N=11$)}&\makeAverage{\the\gitPages}{11}&254& 19& 98&Mb\\
 \hline \end{tabular}

### 2023-05-12 generated/saved-files.tex
Revisions for Computer Journal

allGitRepos.sh & Shell script generated to download all available 10 GitHub repositories\\
assessments.tex & Main summary table for Supplementary Material\\
clone-date.tex & Running the download script also saves the download dates of GitHub clones\\
constants.tex & Generated \LaTeX\  definitions and constants\\
data-check.html & Convenient list of data sources with DOIs in HTML (enables checking of data just by clicking on HTML links)\\
data.csv & All data from \texttt{data.js} expressed in CSV for convenience\\
example-data.tex & \LaTeX\  example of JSON data\\
flagData.nb & Flags expressed as Mathematica code (facilitates ML or other analysis)\\
legend.tex & Legend for the main assessment table\\
open-doi.sh & Shell script to conveniently open all DOIs\\
page-lengths.tex & Page lengths and totals for Git data, and total pages of papers\\
saved-files.tex & This table of files generated by \texttt{data.js}\\
summary-table.tex & Short summary table\\
supplementary-references.tex & Reference list for Supplementary Material\\

### 2023-05-12 generated/summary-table.tex
Revisions for Computer Journal

@@
 Some or all code in principle available on request&8&25\%&\\
-Requested code actually made available (within 2 years 8 months${}^{\star}$)&0&0\%& \\\hline
+Requested code actually made available (within 2 years 10 months${}^{\star}$)&0&0\%& \\\hline
  \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
@@
 \hline\end{tabular}\\
-\vskip 1ex ${}^{\star}$Time of 2 years 8 months is wait between code request and date of generating this table.
+\vskip 1ex ${}^{\star}$Time of 2 years 10 months is wait between code request and date of generating this table.

### 2023-05-12 generated/supplementary-references.tex
Revisions for Computer Journal

\vbox{\bibitem{ref-1}
{\sc A. Callahan, E. Steinberg, J. A. Fries, S. Gombar, B. Patel, C. K. Corbin, and N. H. Shah}, ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp, 2020. DOI \texttt{10.1038/s41746-020-0300-0}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{MetricSelectionFramework}\bibitem{ref-2}
{\sc C. M. Kanzler, M. D. Rinderknecht, A. Schwarz, I. Lamers, C. Gagnon, J. P. O. Held, P. Feys, A. R. Luft, R. Gassert, and O. Lambercy}, ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp, 2020. DOI \texttt{10.1038/s41746-020-0286-7} {Code \url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\bibitem{ref-3}
{\sc A. Roy, K. Nikolitch, R. McGinn, S. Jinah, W. Klement, and Z. A. Kaminsky}, ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp, 2020. DOI \texttt{10.1038/s41746-020-0287-6}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\bibitem{ref-4}
{\sc D. M. Levine, Z. Co, L. P. Newmark, A. R. Groisser, A. J. Holmgren, J. A. Haas, and D. W. Bates}, ``Design and testing of a mobile health application rating tool.,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp, 2020. DOI \texttt{10.1038/s41746-020-0268-9}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\bibitem{ref-5}
{\sc T. Kannampallil, J. M. Smyth, S. Jones, P. R. O. Payne, and J. Ma}, ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp, 2020. DOI \texttt{10.1038/s41746-020-0278-7}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{PENet}\bibitem{ref-6}
{\sc S. Huang, T. Kothari, I. Banerjee, C. Chute, R. L. Ball, N. Borus, A. Huang, B. N. Patel, P. Rajpurkar, J. Irvin, J. Dunnmon, J. Bledsoe, K. Shpanskaya, A. Dhaliwal, R. Zamanian, A. Y. Ng, and M. P. Lungren}, ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp, 2020. DOI \texttt{10.1038/s41746-020-0266-y} {Code \url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\bibitem{ref-7}
{\sc S. S. Dhruva, J. S. Ross, J. G. Akar, B. Caldwell, K. Childers, W. Chow, L. Ciaccio, P. Coplan, J. Dong, H. J. Dykhoff, S. Johnston, T. Kellogg, C. Long, P. A. Noseworthy, K. Roberts, A. Saha, A. Yoo, and N. D. Shah}, ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp, 2020. DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
{\sc I. S. Hofer, C. Lee, E. Gabel, P. Baldi, and M. Cannesson}, ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp, 2020. DOI \texttt{10.1038/s41746-020-0248-0} {Code \url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
{\sc B. Norgeot, K. Muenzen, T. A. Peterson, X. Fan, B. S. Glicksberg, G. Schenk, E. Rutenberg, B. Oskotsky, M. Sirota, J. Yazdany, G. Schmajuk, D. Ludwig, T. Goldstein, and A. J. Butte}, ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp, 2020. DOI \texttt{10.1038/s41746-020-0258-y} {Code \url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
{\sc D. Choi, J. J. Park, T. Ali, and S. Lee}, ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp, 2020. DOI \texttt{10.1038/s41746-020-0261-3} {Code \url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\bibitem{ref-11}
{\sc C. B. Hilton, A. Milinovich, C. Felix, N. Vakharia, T. Crone, C. Donovan, A. Proctor, and A. Nazha}, ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp, 2020. DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip

\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
{\sc M. D. Li, K. Chang, B. Bearce, B. Y. Chang, A. J. Huang, J. P. Campbell, J. M. Brown, P. Singh, K. V. Hoebel, D. Erdo{\u g}mu\c{s}, S. Ioannidis, W. Palmer, M. F. Chiang, and J. Kalpathy-Cramer}, ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp, 2020. DOI \texttt{10.1038/s41746-020-0255-1} {Code \url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip

\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
{\sc J. I. Hoffman, R. Nagel, V. Litzke, D. A. Wells, and W. Amos}, ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp, 2020. DOI \texttt{10.1098/rsos.200419} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
{\sc P. Gr\"onquist, P. Panchadcharam, D. Wood, A. Menges, M. R\"uggeberg, and F. K. Wittel}, ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp, 2020. DOI \texttt{10.1098/rsos.192210} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{onlyPaperWithChecks}
{\sc W. Amos}, ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp, 2020. DOI \texttt{10.1098/rsos.191900} {Code \url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
{\sc M. Gordon, D. Viganola, M. Bishop, Y. Chen, A. Dreber, B. Goldfedder, F. Holzmeister, M. Johannesson, Y. Liu, C. Twardy, J. Wang, and T. Pfeiffer}, ``Are replication rates the same across academic fields? Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp, 2020. DOI \texttt{10.1098/rsos.200566} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
{\sc D. Evans, and A. P. Field}, ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp, 2020. DOI \texttt{10.1098/rsos.200422} {Code \url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ethics-paper}
{\sc N. Beale, H. Battey, A. C. Davison, and R. S. MacKay}, ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp, 2020. DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-19}
{\sc A. A. Cherevko, T. S. Gologush, I. A. Petrenko, V. V. Ostapenko, and V. A. Panarin}, ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp, 2020. DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
{\sc A. A. Soczawa-Stronczyk, and M. Bocian}, ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp, 2020. DOI \texttt{10.1098/rsos.200622} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{lactModel}\bibitem{ref-21}
{\sc S. Duruz, E. Vajana, A. Burren, C. Flury, and S. Joost}, ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp, 2020. DOI \texttt{10.1098/rsos.200638} {Code \url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-22}
{\sc E. Z. D. de Azevedo, D. V. Dantas, and F. G. Daura-Jorge}, ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp, 2020. DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{LRM}\bibitem{ref-23}
{\sc A. M. Abdolhosseini-Qomi, S. H. Jafari, A. Taghizadeh, N. Yazdani, M. Asadpour, and M. Rahgozar}, ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp, 2020. DOI \texttt{10.1098/rsos.191928} {Code \url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{1}\bibitem{example-numerical-error}
{\sc J. Webster, and M. Amos}, ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp, 2020. DOI \texttt{10.1098/rsos.200307} {Code \url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-25}
{\sc Y-l. Zhu, C-J. Wang, F. Gao, Z-x. Xiao, P-l. Zhao, and J-y. Wang}, ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp, 2020. DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-26}
{\sc B. Yu, C. J. Scott, X. Xue, X. Yue, and X. Dou}, ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp, 2020. DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-27}
{\sc K. Joon-myoung, C. Younghoon, J. Ki-Hyun, C. Soohyun, K. Kyung-Hee, B. Seung D, J. Soomin, P. Jinsik, and O. Byung-Hee}, ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-28}
{\sc H. Zhu, C. Cheng, H. Yin, X. Li, P. Zuo, J. Ding, F. Lin, J. Wang, B. Zhou, Y. Li, S. Hu, Y. Xiong, B. Wang, G. Wan, X. Yang, and Y. Yuan}, ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
{\sc R. Fung, J. Villar, A. Dashti, L. C. Ismail, E. Staines-Urias, E. O. Ohuma, L. J. Salomon, C. G. Victora, F. C. Barros, A. Lambert, M. Carvalho, A. Jaffer Y, J. A. Noble, M. G. Gravett, M. Purwar, R. Pang, E. Bertino, S. Munim, A. M. Min, R. McGready, S. A. Norris, Z. A. Bhutta, S. H. Kennedy, A. T. Papageorghiou, and A. Ourmazd}, ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30131-X} {Code \url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\bibitem{ref-30}
{\sc C. Sabanayagam, D. Xu, D. S. W. Ting, S. Nusinovici, R. Banu, H. Hamzah, C. Lim, Y-C. Tham, C. Y. Cheung, E. S. Tai, X. Y. Wang, J. B. Jonas, C-Y. Cheng, M. L. Lee, W. Hsu, and T. Y. Wong}, ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30063-7}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip

\vbox{\maprepo{blast-ct}\bibitem{ref-31}
{\sc M. Monteiro, V. F. Newcombe, F. Mathieu, K. Adatia, K. Kamnitsas, E. Ferrante, T. Das, D. Whitehouse, D. Rueckert, D. K. Menon, and B. Glocker}, ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30085-6} {Code \url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip

\vbox{\bibitem{ref-32}
{\sc K-L. Liu, T. Wu, P-T. Chen, M. Tsai Y, H. Roth, M-S. Wu, W-C. Liao, and W. Wang}, ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp, 2020. DOI \texttt{10.1016/S2589-7500(20)30078-9}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip

### 2023-05-12 paper-seb-main.tex
Revisions for Computer Journal

@@
 \usepackage{amssymb}
+
+\setcounter{tocdepth}{3}
 
@@
 \noindent
-Scientists rarely assure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny.
+Scientists rarely rigorously ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny.
 Even if available, scientists rarely provide adequate documentation to understand or use their code reliably.
-When code used in science avoids adequate scrutiny published science will be unreliable, because results --- data, graphs, images, inferences, etc --- generated by the code will themselves be unreliable.
-
-\hskip 1em This paper justifies and proposes ways to improve research and science relying on code:
+When code used in science avoids adequate scrutiny, published science will be unreliable and unverifiable because results --- data, graphs, images, inferences, and so forth --- informed or generated by the code will themselves be unreliable and unverifiable.
+
+\hskip 1em This paper proposes and justifies  ways to improve research and science relying on code:
 
@@
 The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp.
-\RAPstar\ (or comparable interventions) could be supported and encouraged in journal, conference, and funding body policies.
+\RAPstar\ (or comparable interventions) could be supported and or even required in journal, conference, and funding body policies.
 
@@
 The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input.
-The \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ processes.
+The paper's \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ workflows.
 
@@
 Unreliable, often unstated and unexplored, code and computational dependencies (including using AI or ML) in science are widespread.
-Furthermore, code is rarely published or made accessible in a useful form, nor is it professionally scrutinized.
-Too often, therefore, code and results claimed do not contribute to reliable science, and do not support replication or reproduction on which future science can be firmly based.
+Furthermore, code is rarely published or made accessible in a usable form; it is generally too onerous or impossible to verify or scrutinize.
+Ironically, computers should be able to make reproducibility easier, yet too often code and results claimed do not contribute to reliable science, and do not support verification, replication, or reproduction on which future science can be firmly based.
 
@@
 Like statistics, code is often relied on to support key results in scientific publications, yet code is generally informal, inaccessible and incorrect.
-Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.
+Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on professional Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.
 
@@
 \end{tabular}
-\end{center}
+\end{center}\vskip -2ex
 \caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}.
@@
 \input generated/summary-table.tex
-\end{center}
+\end{center}\vskip -2ex
 \caption{Breakdown of pilot survey of peer-reviewed science papers relying on code. }
@@
 \newdimen\separation
-\separation=3mm
+\separation=4.5mm
 \quarterwidth=\textwidth
@@
 \divide\quarterwidth by 4
-\noindent{\sf\begin{tabular}{@{}c|c|c|c@{}}
+\begin{center}
+\noindent{\sf\begin{tabular}{|@{}c|c|c|c@{}|} \hline
 \bfseries Level 0&
@@
 \bfseries Level 3 \\ \hline
-\multicolumn{1}{@{}p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
+\multicolumn{1}{|p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
 \multicolumn{1}{|p{\quarterwidth}}{Article states whether code is available and, if so, where to access them.}&
 \multicolumn{1}{|p{\quarterwidth}}{Code must be posted to a trusted repository. Exceptions must be identified at article submission.}&
-\multicolumn{1}{|p{\quarterwidth}@{}}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.}
-\end{tabular}}
+\multicolumn{1}{|p{\quarterwidth}|}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.} \\ \hline
+\end{tabular}}\end{center}\vskip -2ex
 \caption{The TOP committee's recommended levels for journal article code transparency.
@@
 
-This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research, because of its significance for informing and driving public health responses to the Covid pandemic.
-A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see the tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey.
+This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the Covid pandemic.
+A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey.
 In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below).
-Only one paper mentioned any Software Engineering methods, albeit without technical details.
-
-In the survey sample 81\% of papers were published in journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies.
+Only one paper mentioned any Software Engineering methods, albeit simplistic and without technical details.
+
+In the survey sample 81\% of papers were published in leading journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies.
 One paper declared it had accessible code, but the relevant repository was and still remains empty.
-The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience}.
+The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience} discussed throughout this paper.
 
@@
 Its main contribution is to suggest straightforward ways to enable this.
-The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and inform raising the profile and constructively debating the issues more widely.
+The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and to inform raising the profile and constructively debating the issues more widely.
 
@@
 \section{Background}
-The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences.
-There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature.
+The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drove and expanded the sciences.
+There are fascinating periods when new ideas and science unified; for example, thermometers could not measure temperature in any meaningful way until the underlying science was mature.
 For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.
 
-Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
-Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, following international quality standards.
+Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant.
+Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
+Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, that follow international quality standards.
 
 Computers are a unique, new technology, far more flexible and challenging than thermometers.
-Understanding computers and integrating them into science is likely to be harder than the development of modern thermometry.
+Understanding computers and integrating them into science is far harder than the tortured development of modern thermometry.
 Computation not only expands science's paradigms and supports new discoveries (particularly with AI), but it also \emph{does\/} new science --- almost all modern laboratory instrumentation, including thermometers, is heavily computerized.
-Pushing the boundaries of science, then, involves pushing the boundaries of computer science.
+Relying on computer models has become routine.
+The dependency of modern science on computers is far more tangled and complex than the now-resolved dependence on reliable temperature measurement; computers affect how scientists in all disciplines \emph{think}.
+
+Pushing the boundaries of science, then, now involves pushing the boundaries of computer science.
 The synergy runs deep: for instance, while particle physics relies on powerful supercomputers, quantum physics itself is developing more powerful quantum computing.
@@
 ``Computational science'' has come to mean a particular style of science based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational in this sense.
-Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ In all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications.
+
+Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ in all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications.
 All areas of science are being profoundly computerized.
@@
 
-\subsection{Related code quality concerns}\label{related-work}
+\subsection{Code quality concerns}\label{related-work}
 Publishing high quality computer code has been strongly advocated since the earliest times, such as the \emph{Communications of the ACM\/} in its first issue in its first volume published in 1958, where it outlined its algorithm publication policy.
 The new policy was illustrated with a square root algorithm \cite{acm-algorithms}.
-However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the concern of the present paper.
-Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code.
+However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the particular concern of the present paper.
+Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code in general science.
 
@@
 While chemists are trained in reliable methodologies, so taking quality glassware for granted is reasonable.
-Code is a newer innovation.
-Reliable code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}.
+Code is a newer innovation, and quality code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}.
 Inevitably, because of these reasons taken collectively, much published science depends on unreliable code that is not explicitly discussed, was not peer-reviewed, and is not open to scrutiny, reproduction, or reuse by the scientific community.
-
-A study of 863,878 Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully.
-Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of 9,000 research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences.
+The situation with code is analogous to chemists using contaminated glassware with no awareness how its effects might be controlled or might affect results.
+
+Is it a problem?
+A study of \digitBlock{863878} Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully.
+Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of \digitBlock{9000} research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences.
 They found a comparable result that 74\% of the code files analyzed failed.
@@
 
-The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' More generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse has already done so.
+The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' It will not be obvious for most practicing scientists how to do this, so, more generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse (the source of the \cite{trisovic} data) has already done so.
 These ideas may be compared to the present paper's proposal of Software Engineering Boards (SEBs), as discussed in detail in section \ref{summary}, supported by more specific suggestions in the \supplement.
 
-These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis,reproducible-manifesto}.
-Concern has led to a new journal \emph{ReScience~C\/} to explore and encourage the explicit replication of previously published research \cite{rescience}; furthermore, \emph{ReScience~C\/} promotes new and open-source implementations in order to ensure that the original research is reproducible.
-However, most of its published concern focuses on data rather than code.
-
-To start to address code quality issues, the Transparency and Openness Promotion Committee first met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}.
+These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis}.
+Concern has led to new journals, \emph{Journal of Open Source Software\/} (JOSS) \cite{joss}, \emph{ReScience~C\/} \cite{rescience}, and  others, to explore and encourage the explicit replication of previously published research.
+
+To start to address code quality issues the Transparency and Openness Promotion Committee met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}.
 TOP covers citation standards, replication standards, and code standards amongst others.
@@
 The TOP standards continue to develop, and are now maintained on a wiki at \url{osf.io/9f6gx/wiki/Guidelines} \cite{TOP-wiki}.
-Note that TOP aligns with but is a stricter approach than the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which is critiqued in section \ref{critique-fair}.
-
-As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (section \ref{section-vicious-cycles}).
+TOP can be compared with the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which places greater emphasis on data rather than code; FAIR itself is critiqued in section \ref{critique-fair}.
+
+As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (see section \ref{section-vicious-cycles}).
 The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths.
@@
 However, it should be noted that such tools are not a panacea, as the study \cite{jupyter-study} cited above makes clear.
-This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be important.
-
-In addition to unintentional problems with code quality and reproducibility, scientific misconduct occurs when the outcome is intentional.
-While pure plagiarism, which is misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have realistic properties, the results are likely to be destructive.
-The Wakefield MMR fraud linking vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted, and has been extraordinarily destructive \cite{MMR-fraud}.
-
-A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists at least once fabricated, falsified, or modified results, and over a third undertook other questionable research practices \cite{misconduct-issues}.
-The meta-analysis qualifies the figures carefully, but these are alarming rates with any qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The author's survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}
-
-While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using the same methods as will improve research reproducibility, as discussed throughout the present paper.
+This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be critical.
+
+In addition to unintentional problems with code quality and reproducibility, actual scientific misconduct occurs when the outcome is intentional.
+While pure plagiarism, which is a form of misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have deceptive properties, the results are likely to be destructive.
+The Wakefield MMR fraud claiming to link vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted; this misconduct has been extraordinarily destructive \cite{MMR-fraud}.
+
+A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists have at least once fabricated, falsified, or modified results, and over a third have undertaken other questionable research practices \cite{misconduct-issues}.
+The meta-analysis qualifies the figures carefully, but these are alarming rates regardless of the qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The present paper's author's own survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}
+
+While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using exactly the same methods as will improve research reproducibility, as discussed throughout the present paper.
 
@@
 ``Electronic lab notebooks'' (ELNs) \cite{elns}, emphasize computer tools that specifically support laboratory notebook authoring and editing.
-In contrast, ``computable papers'' aim to support the scientific paper authoring process: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.
-
-If, for example, HTML is being used to prepare a computational paper, it could include Javascript code like \texttt{<script> document.write(responses.total) </script>}, which would insert the result of running that Javascript code into the paper, such as the number into a sentence in an empirical paper, like ``We collected data and obtained 754 responses.''
-
-Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with basic calculations (e.g., here we calculate $10!=
-\def\factorial#1{\ffactorial{#1}1}
+In contrast, so-called ``computable papers'' aim to support the scientific paper authoring workflow: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.
+
+If, for example, HTML is being used for a computational paper, the paper's text could include Javascript code like
+
+{\tt <script>\par \hskip 2em document.write(responses.total)\par </script>}
+
+This illustrative code would insert the result of running it into the paper, perhaps like ``We collected data and obtained 754 responses,'' where the 754 is the value of the variable \texttt{responses.total} when the paper was formatted.
+The point is that the number 754 (or whatever) is computed automatically from the data, so any changes or improvements to the data or the methodology analyzing it will translate into the published number being updated and inserted into the paper at the appropriate point.
+
+\def\factorial#1{\digitBlock{\ffactorial{#1}{1}}}
 \newcount\t
@@
 \fi}
-\factorial{10}$ by writing \LaTeX\ code \emph{in\/} this paper).
-More practically, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper.
-
-\newdimen\boxwidth
+
+Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with computable paper calculations. Here, as a simple example, we calculate that
+$10!$ is equal to \factorial{10} just by writing the \LaTeX\ code \verb|\factorial{10}| \emph{in\/} this paper to calculate arbitrary factorials and group the result into conventional blocks of three digits by inserting small spaces.
+Note that although \texttt{factorial} is not a standard \LaTeX\ function, it was readily defined by code also \emph{in\/} the present paper, so it can be easily modified by the author for other purposes as required.
+
+More practically for helping author complex scientific papers, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper.
+Here is a small example, where each generated fact (taken from the present paper's analysis) has been highlighted in \textbf{bold}:
+
+\vskip 1ex\newdimen\boxwidth
 \boxwidth = \columnwidth
@@
 \noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors.
-The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}
-
-In the ``old days'' such numbers (\the\dataN, \the\countAuthors, etc) quoted above would have been manually worked out, then eyeballed and typed up by the authors.
+The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}\vskip 1ex
+
+In the ``old days'' such numbers (\the\dataN; \the\countAuthors; etc) quoted above would have been manually worked out, eyeballed, typed up, and then (hopefully) double-checked by the authors.
 Instead, in this paper, those numbers (and many more) were computed automatically.
-Furthermore, they update automatically when the data or calculations change.
 They were inserted into the text of this paper automatically, and are auditable back to their sources.
-
-The idea is easy to implement.
-Code can generate text like \texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}} (where the \the\countAuthors\ is the calculated number) save that line of text to a file, which is then input into the paper where it is typeset in the normal way.
-Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the correct value is at the time.
-Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11} \ldots\ 1,000 etc, as done (for example) in this paper's footnote~\numbernamesexamplefootnote.
-
-Rather than use quirky languages like \LaTeX\ computable papers more often build on general-purpose programming languages.
-Python and R, for instance, have powerful features for processing data, plotting graphs, doing statistics, and more.
-
-The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct --- \emph{with no further work from the authors}.
-Indeed, if an author corrects a mistake, the correction will apply automatically to all future edits.
-
-However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown.
-Many authors therefore prefer a more structured approach imposed by tools designed for the purpose.
-A taut review is Perkel \cite{perkel-review}, but here four tools,  WEB, \emph{Mathematica}, Jupyter, and knitr, will serve to sample the variety of approaches that are available:
+Furthermore, they will update automatically --- with no further work from the author --- when the data or calculations change.
+
+The idea is easy to implement. Continuing using \LaTeX\ as the illustrative word processing system, code could generate text like
+
+\begin{center}\texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}}\end{center}
+
+where the \the\countAuthors\ is the calculated number, save that line of text to a file, which is then import it into the paper where it is typeset in the normal way.
+Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time.
+Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done (more realistically) in this paper's footnote~\numbernamesexamplefootnote.
+
+The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper also update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct \emph{with no further work from the authors}.
+Indeed, if an author corrects a mistake, the correction will apply automatically to all future revisions.
+
+However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown, which means authors may make unnoticed mistakes.
+Many authors therefore prefer a more structured approach imposed by tools designed for the purpose, where their structure helps impose uniformity and helps prevent and check for errors.
+
+A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes (in an appendix) a substantial and useful literature review.
+Here four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:
 
@@
 
-\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code.
-
-\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications \emph{using\/} the code, inserting output, such as graphs, generated by running it.
+\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code required for computer execution.
+
+\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications just \emph{using\/} code, inserting output, such as statistics or graphs, generated by running it.
 
@@
 Notebooks were developed by Theodore Gray in 1988 \cite{notebook-patent}.
-Notebooks consist of a collection of ``cells.'' Some notebook cells are labelled as text or as section headings, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output.
+Notebooks consist of a collection of ``cells,'' where cells can be labelled as text or as a section heading, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output.
 The new cell can be numbers, tables, mathematics, a plot, an image, or even more text --- the arbitrary output of running code, in fact.
 Moreover, a notebook itself is a \emph{Mathematica\/} expression, so it --- the paper itself --- can be analyzed or manipulated by code in any way.
-In particular, the entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.
-
-\hskip 1em A \emph{Mathematica\/} notebook can be published directly as a paper, but some code might be distracting.
-Typically the author therefore hides some or all code cells that are irrelevant to the narrative of the paper, but which nonetheless were required to generate the results presented.
+The entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.
+
+\hskip 1em A \emph{Mathematica\/} notebook can be published directly as a paper, but some code might be distracting for a publication.
+Typically the author therefore optionally hides some or all code cells that are irrelevant to the narrative of the paper, but which nonetheless were required to generate the results presented.
 
 \hskip 1em The user manual for \emph{Mathematica\/} itself was written as a \emph{Mathematica\/} notebook, which ensured all its examples actually worked --- and probably helped ensure the correctness of the \emph{Mathematica\/} code used behind the scenes (see section \ref{paper-as-lab}).
-The book is now the largest example of software documentation in existence: in its latest edition it runs to over 10,000 pages.
+The book is now the largest example of software documentation in existence: in its latest edition it runs to over \digitBlock{10000} pages.
 
 \item
-Jupyter was developed by Fernando P{\'e}rez and Brian Granger \cite{jupyter-history}, and takes a similar approach to \emph{Mathematica\/}, but is open-source and not closely integrated with any particular programming language, as \emph{Mathematica\/} is.
-Jupyter can be installed on a PC or run over the web.
+Jupyter was developed by Fernando P{\'e}rez and Brian Granger \cite{jupyter-history}, and takes a similar approach to \emph{Mathematica\/}, but Jupyter is open-source and not closely integrated with any particular programming language, as \emph{Mathematica\/} is.
+Jupyter can be installed on a local computer or run over the web.
 
@@
 
-\hskip 1em Many extensive examples of using Jupyter notebooks (and other good practice, such as using repositories) to support large scale science projects can be found at the Gravitational Wave Open Science Center at \url{www.gw-openscience.org}.
+\hskip 1em Jupyter is very popular and widely-known.
+Many extensive examples of using Jupyter notebooks (and other good practice, such as using repositories) to support large scale science projects can be found at the Gravitational Wave Open Science Center at \url{www.gw-openscience.org}.
 
@@
 \the\howlong}
-years later.
-
-There are tools to make using code more convenient and more reproducible, as this section briefly reviewed, but they are rarely used or used haphazardly, as the next section shows.
+years later.\footnote{\emph{Mathematica\/} is an example of a proprietary system: using it requires a paid-for license, which is a limitation on reproducibility; worse, in the long-run the system owners may go out of business and the code would potentially be unusable at any price.
+However, these limitations are also limitations that impact free software: versions may become obsolete, and the community may move on and stop maintaining old systems.
+The impact on reproducibility is much the same as with proprietary systems.
+The common solution is to code in whichever language is chosen in as portable a way as possible, to chose a system that uses a well-defined notation in an open representation (such as XML or ASCII), so that if the worst happens the old code, or at least the key parts of it, can be translated to run on a new system.}
+
+There are many tools to make using code more convenient and more reproducible, as this section briefly reviewed, but unfortunately they are rarely used or used haphazardly, as the next section shows.
 
 \subsection{RAP: Reproducible Analytical Pipelines}\label{RAP-section}
-Writing a paper typically starts in a word processor (such as Microsoft Word), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on.
-The process will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.
-
-Table \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then edited into a paper.
+Writing a paper typically starts with a word processor (such as Microsoft Word or \LaTeX), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on.
+This workflow will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.
+
+Table \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then collected and edited into a paper.
 
@@
 \begin{tabular}{|lcl@{}cl@{\hskip 2em}cl|} \hline
-\drop{\bfseries Data sources}&\lowarrow&\bfseries \drop{Analysis} & \lowarrow&\bfseries Select results& \lowarrow& \bfseries Submit for\\
+\drop{\bfseries Data sources}&\lowarrow&\bfseries \drop{Models and analysis} & \lowarrow&\bfseries Select results& \lowarrow& \bfseries Submit for\\
 \bfseries &&\bfseries & &\bfseries for write up & & \bfseries publication \\ \hline \hline 
@@
 Literature && Graphics packages  && (text, images, graphs, etc)&&\\
-Sensors && Specially-written code && into paper &&\\ 
+Sensors && Specially-written code, && into paper &&\\ 
+&& \hskip 1em C, Python, R, Mathematica, etc  && && \\
 \vdots && \vdots &&&& \\ &&&&&&\\ \hline
@@
 For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
-The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline processes so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline workflow so that they can be run automatically, and hence reproduce the results that underpin the final paper.
 The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
@@
 \global\advance \enum by 1}
-\begin{enumerate}
-\item
-Peer-review used to ensure the process is reproducible and to identify improvements 
-\item
-No or minimal manual interference;
+\begin{center}\begin{tabular}{|rp{6.5in}|}
+\hline 1&
+Peer-review is used to ensure the workflow followed is reproducible and to identify improvements 
+\\\hline 2&No or minimal manual interference;
 for example copy-paste, point-click and drag-drop steps replaced using computer code that can be inspected by others 
-\item
-Open-source programming languages so that processes do not rely on proprietary software and can be reproduced by others
-\item
-Version control software, such as Git, to guarantee an audit trail of changes made to code 
-\item
-Publication of code, whenever possible, on code hosting platforms such as GitHub to improve transparency 
-\item
-Well-commented code and embedded documentation to ensure the process can be understood and used by others 
-\item
-Embedding of existing quality assurance practices in code, following guidance set by recognized organizations 
-\end{enumerate}
-\emph{Adopting RAP principles is not necessarily about incorporating all of the above: implementing just some of these principles will generate valuable improvements.}
+\\\hline 3&Open-source programming languages so that processes do not rely on proprietary software and can be reproduced by others
+\\\hline 4&Version control software, such as Git, to guarantee an audit trail of changes made to code 
+\\\hline 5&Publication of code, whenever possible, on code hosting platforms such as GitHub to improve transparency 
+\\\hline 6&Well-commented code and embedded documentation to ensure the workflow can be understood and used by others 
+\\\hline 7&Embedding of existing quality assurance practices in code, following guidance set by recognized organizations 
+\\\hline 
+\multicolumn{2}{@{}c@{}}{\begin{minipage}{6.8in}\vskip 1ex
+Adopting RAP principles is not necessarily about incorporating all of the above: implementing just some of these principles will generate valuable improvements.\end{minipage}}
+\end{tabular}\end{center}\vskip -1ex
 }\caption{A minimum standard of RAP, based on the UK Statistics Authority summary \cite{RAPstandard}.}
@@
 
-For clarity, the schematic pipeline in table \ref{fig-pipeline} omits many steps in the creative scientific process.
+For clarity, the schematic pipeline in table \ref{fig-pipeline} omits many steps in the creative scientific workflow.
 Furthermore, each step is iterated and modified as the research progresses, and, indeed, as referees require revision.
 The point is that in typical scientific practice each step in the table is largely or entirely manual, typically selecting and copying output from the previous phase, and then pasting and editing the results into the next.
-The pipeline of data $\rightarrow\cdots\rightarrow$ paper is then iterated by hand as the various components are refined and improved until the authors are happy with the final paper.
+The pipeline of data $\rightarrow\cdots\rightarrow$ paper is then iterated by hand as the various components are refined and improved until the authors (and funders, editors, and referees) are happy with the final paper.
 
 As problems are found in a paper, the data, calculations and code are debugged, refactored, and refined.
-The process is rarely systematic, and even less likely to be documented --- after all, the atomic steps seem to be innocuous copy and paste actions.
+The workflow is rarely systematic, and even less likely to be documented --- after all, the atomic steps seem to be innocuous copy and paste actions.
 The final paper and the ideas it embodies are what matters.
@@
 The insight of the reproducible analytic pipelines (RAP) proponents is that every time any step in the pipeline is performed it could have been automated \cite{rap,goldacre,turing-way}.
-If automated, it could then be repeated reliably --- unlike a manual cut and paste which is error-prone every time it is done.
-It can be repeated if any experimental data, literature, or other knowledge changes, and the paper's analysis brought up to date with ease.
-In particular, any other researcher, whether part of the authorship team or a later reader of the paper, can reproduce it reliably if the RAP process is made available.
+If automated, it could then be repeated reliably --- unlike a manual cut and paste which is potentially different and certainly error-prone every time it is performed.
+If automated, any part of the workflow can be reliably repeated if any experimental data, literature, or other knowledge changes; the paper's analysis will brought up to date with ease.
+In particular, any other researcher, whether part of the authorship team or a later reader of the paper, can reproduce the paper and its results reliably provided that the RAP workflow is made available.
 Table~\ref{figure-RAP-summary} provides a brief summary of RAP principles.
@@
 For example, if the paper in question is a systematic review, it could be kept current by automatically re-running the programmed atomic actions that it was built with.
-Indeed, this ability is one of the original motivations of RAP, where Government agencies could easily generate up to date reports on request without having to repeat all the manual work and risk making procedural errors doing so.
-Furthermore, every time they update a publication, the RAP pipeline itself is reviewed and improved, so the quality of the reproduced work improves --- unlike in a non-RAP process where new errors are generally introduced every time the work is revisited.
-
-RAP not only helps develop reproducible science, and improve the quality of the science as the authors debug and refine their methodology, it also provides a precise audit trail that can be used to protect against fraud, as discussed in \cite{MMR-fraud}, and can in principle be largely automated and perform checks much faster and more efficiently than conventional \emph{post hoc\/} investigations.
+Indeed, this ability is one of the original motivations of RAP, so Government agencies could easily generate up to date reports on request without having to repeat all the manual work, and risk making procedural errors doing so.
+Furthermore, every time a publication is re-derived or updated, the RAP pipeline itself is reviewed and improved, so the quality of the reproduced work improves --- unlike in a non-RAP workflow where new errors are potentially introduced every time a work is revisited.
+
+RAP not only helps develop reproducible science, and improve the quality of the science as the authors debug and refine their methodology, it also provides a precise audit trail that can be used to protect against fraud, as discussed in \cite{MMR-fraud}.
+RAP can perform checks much faster and more efficiently than conventional \emph{post hoc\/} investigations.
 
@@
 
-The corollary is that if we are doing arbitrary cut and paste that has not been programmed into a computer, then we are not doing good science.
-Science is in principle an algorithmic process, and therefore, as Knuth says, if we understand well enough what we are doing in science, we can explain it as code, as RAP, for a computer to run and rerun.
+The corollary is that if we are doing arbitrary cut and paste that has not been programmed into a computer, then we are not doing good science; we are certainly not explaining what we are doing to the computer.
+Science is in principle an algorithmic process, and therefore, as Knuth says, if we understand well enough what we are doing in science, we can explain it as code, specifically in a RAP, for a computer to automatically run and rerun.
+
+\subsubsection{Potential code}
+Code is usually thought of as text written in a programming language, such as Python or C, but this ignores special cases of what can be called \emph{potential code\/}: processes that could be presented in code, but have not been and therefore are invisible.
+Such potential code cannot be reasoned about as rigorously as they would be had they been expressed explicitly in code.
+The loss of scrutiny, loss of the ability to reason rigorously, the loss of the  ability to review potential code are problems that RAP tries to address.
+Potential code includes:
+
+\begin{enumerate}\raggedright
+\item Critical algorithmic steps may never be codified.
+Writing a paper may involve creating an image and copying it into a word processed document.
+Indeed, every time the image is modified, the process must be repeated.
+In an important sense, the author is executing a computational process, effectively using code that has never been written down.
+There is a process here, but it is informal and may be run differently each time.
+However, it could have been coded and reproduced precisely every time it was needed.
+Moreover, if the process had been coded explicitly, it could be reasoned about, critiqued and improved.
+
+\item Compiling and running programs (for instance to generate results) is also a computational process that may not be recorded in code.
+The author, more explicitly than creating and using images, runs programs to get results --- but the process of running the programs may not be recorded.
+Typically, if the author notices that the processes are repetitious, they may develop a shell script to codify the repetitive process.
+
+\item Processes in writing a paper, configuring software and generating data may be codified, but the author discovers that their codified processes do not generate quite what they want.
+Rather than debugging the code, it is tempting to manually edit the final results.
+The author knows they could have coded things correctly, but this seems too tedious --- especially if the edits required seem minor.
+
+\item Another case of potential code arises when a paper has been completed, but referees or the publishers require some minor fixes.
+These minor fixes are easier to implement in the paper as simple textual corrections, rather than revisiting and updating the explicit code that informed or generated the paper (and updating the repositories and so forth).
+The RAP approach would require the entire workflow to be revised, not just the final presentation in the paper.
+
+\end{enumerate}
+
+In all cases, a computational process is involved that could have been explicitly coded, but defining the general case as program code seemed harder than an \emph{ad hoc\/} implicit process.
+The problem for reproducibility is that the final science depends on this potential code as much as on any explicit code, but it is nowhere recorded, and therefore reconstructing the science will be unreliable.
+
+\subsubsection{RAP as research}
+RAP itself is an object of research.
+For example, in reproducing the results of a paper published over a decade previously, \cite{Courtes} shows that RAP workflows --- which had not been considered or followed at the time --- can be semi-automatically reconstructed (along with software dependencies) using suitable tools, thus making the original paper and the experiment it depended on fully reproducible.
+Furthermore, the newly derived and now explicit pipelines were specified as pure functions, meaning that the workflow was fully-defined using nothing but explicit functions and explicit parameters, arguably a prerequisite for rigorous, deterministic reproducibility.
+
+While this reproduction of an old paper obtained important insights, deriving a RAP workflow after the fact cannot benefit the original scientific process.
+Instead, insights from this \emph{post hoc\/} reproduction methodology \cite{Courtes} can be used as an insightful basis to help advance approaches to RAP; that is, doing RAP explicitly (in, for example, the ways that \cite{Courtes} explores) while the science is being developed would provide powerful and constructive insights \emph{during\/} the scientific process, rather than later in hindsight.
 
 \section{The statistics/code analogy}\label{statistics-computation-analogy}
-The central role of computational methods in science may be fruitfully compared to statistics, an established scientific tool.
+The central role of computational methods in science may be fruitfully compared to using statistics, an established scientific tool.
 
 Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics.
-Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.
+Often scientists do not realize the limitations of their own statistical skills, particularly when developing new experiments, so careful scientists generally work closely with professional statisticians.
 
@@
 \hskip -\wd0\box0 Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
-[\ldots] Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.0009)$.
+[\ldots] Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 $( 95\%$ CI $0.78, 4.27, p = 0.000\hskip 0.3em 9)$.
 Between group effect size $d = 0.55$ $(95\%$ CI $0.32,$ $0.77)$.''} \cite{example-stats}
@@
 
-This wording formally summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
+This standard wording formally summarizes confidence intervals, $p$ levels, and so on, to present various statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize.
 It is a \emph{lingua franca}.
-It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results.
+It may look technical, but it is written in the standard, widely accepted form for summarizing statistics --- it is a clear, rigorous, and readily interpreted way to express uncertainty in results.
 Moreover, behind any such brief paragraph is a substantial, rigorous, and appropriate statistical analysis.
@@
 The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted; in fact, there is widespread concern about the poor quality of statistics in science \cite{paul-hci,pnas-stats}.
-While it is assumed that statistics should be peer-reviewed, and that review will often lead to improvement, these critical papers show that reviewers and editors are often failing to pick up on poor statistics.
+While it is assumed that statistics should be peer-reviewed, and that review will often lead to improvement, critical papers like \cite{paul-hci,pnas-stats} show that reviewers and editors are often failing to pick up on poor statistics.
 
 Scientists accept that statistics is a distinct, professional science, itself subject of research and continual improvement.
-Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are insufficient training for rigorous work in science --- their main value, perhaps, is to help scientists to understand the value of collaborating with specialist statisticians.
+Among other implications of the depth and progress of the field of statistics, undergraduate statistics options for general scientists are recognized as insufficient training for rigorous work in science --- their main value, arguably, is to help scientists to understand the value of collaborating with specialist statisticians.
 Collaboration with statisticians is particularly important when new types of work are undertaken, where the statistical pitfalls have not already been well-explored.
@@
 Except in the most trivial of cases, all numbers and graphs, along with the statistics underlying them, will be generated by computer.
-Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it.
+Indeed, computers are now very widely used, not just to calculate statistics, but to run the models, to do the data sampling and processing, to operate the sensors or surveys that generate the data, and to process it all.
 Many papers now explore the contribution of AI and ML to their fields.
@@
 However, using any code raises many critical questions: formats, backup, cyber-vulnerability, version control, integrity checking (e.g., managing human error), auditing, debugging and testing, and more.
+Is the code correct, and is it dependable enough to justify the claims the scientists would like to make?
 Software code, like statistics, is subject to unintentional bias \cite{Ben,se-bias}.
 All these issues are non-trivial concerns requiring technical expertise to manage well.
-As with statistics, good answers to such ``technical'' issues makes the science that relies on them better.
-
-A common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model was reliably transformed into code that generates the results the paper summarizes; if so, the code may have problems that cannot be identified as there is no specification to reference it to, and possibly even no link to the code at all.
+As with statistics, good answers to such ``technical'' issues makes the science that relies on them better; conversely, failing to properly address the questions makes the science suspect.
+
+For example, a common oversight in scientific papers is to present a model, such as a set of differential equations, but omit how that model was reliably transformed into the code that generates the results the paper summarizes.
+The code may have problems that cannot be identified as there is no specification to reference it to, and possibly even no link to the code at all.
 
@@
 Indeed, as few papers use code that is as well-understood and as well-researched as standard statistical procedures (such as Bonferroni corrections), the scientific problems of poorly planned and reported code are widespread.
+The terms ``invariant,'' ``pre-condition,'' ``post-condition'' are basic technical terms  used in reliable coding, yet none of these concepts appear in  in any of the code repositories referred to in this paper's survey.
+Only one project uses assertions, and then only for checking user interface input data rather than the correct operation of the paper's model (see \supplement\ data summary).
+These basic coding concepts are simpler than Bonferroni corrections.
 
@@
 
-but without giving any further details.
-The paper cited above describes one of its multivariate analyses as ``multiple correspondence analysis (MCA) and ascendant hierarchical clustering'' with no specific details --- reproducing the work would be non-trivial since the methodology would have to be reconstructed from scratch, even assuming the data was made available by the authors as the paper gives no details of the data used or how to obtain it.
-
-The problem is that the common practice of declaring using a named computational system (such as SAS in this case) does not help scrutiny in the least, as such systems can do almost anything.
-\emph{How\/} those analyses might have been performed is not discussed, and one assumes it follows that the analyses could therefore not have been properly reviewed for basic scientific competence during the publication process.
+but without giving more details.
+Besides, the authors have not made their code available so it is moot what system it runs on.
+The problem is that the common practice of declaring using a named system (such as SAS in this case) does not help scrutiny in the least, as such systems can do almost anything.
+\emph{How\/} those analyses might have been performed was not discussed, and one assumes it follows that the analyses could therefore not have been properly reviewed for scientific competence during the publication workflow.
 
 A reviewer, if nobody else, needs to actually examine the code used and its documentation to assess whether the analysis presented in the paper is appropriate and sufficiently reliable.
-Furthermore, if the analysis in this case actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible as it relies on idiosyncrasies in SAS version 9.2.
+Furthermore, if the analysis in this case actually depended on using SAS version 9.2, and not \emph{any\/} general purpose statistical system, then it is problematic because it is not reproducible as it depends on idiosyncrasies in SAS version 9.2.
 Of course, an author can disclose the idiosyncratic dependencies; while this seems to be an onerous obligation, conversely it is arguable that if an author is unaware of dependencies, then their science relying on them is equally unreliable.
 
-It is recognized that to make critical claims, models need to be run under varying assumptions \cite{whitty}, yet somehow it is easy to overlook that the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, as well as to uncover unknown dependencies.
-Indeed, the code may be poorly written (as this paper shows it is), so the results derived from the code simply may not be reliable.
+It is recognized that to make critical claims, models must be run under varying assumptions \cite{whitty}, yet somehow it is overlook edthat the code that implements those models also needs to be carefully tested under varying assumptions to uncover and fix bugs and biases, as well as to uncover unknown dependencies.
+Indeed, the code may be poorly written (as this paper shows it often is), so the results derived from the code simply may not be reliable.
 
@@
 This argument is mistaken.
-Code is rarely considered a valuable part of the science to which it contributes, which creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.
+Code is rarely considered a valuable part of the science to which it contributes (section \ref{lowCodeStatus}), which creates a vicious cycle of ignoring code, leading to ignoring the critical --- and non-trivial --- role of correct code in science.
 
@@
 Some authors assert that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarizing data or properties of the data \cite{assessing-quality}.
-In reality, if code is inadequate, any so-called ``insights'' will be flawed, and flawed in unknown ways.
+In reality, if code is inadequate, any so-called ``insights'' will be potentially flawed, and flawed in unknown ways.
 Indeed, none of the papers sampled (described in \supplement\ section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
@@
 \subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
-It is a misconception that programming is easy and even children can do it \cite{fixit}. More correctly, toy programming is easy, but mature programming is very difficult.
+It is a misconception that programming is easy and even children can do it \cite{fixit}.
+More correctly, toy programming is easy, but mature programming is very difficult.
+% (compare: ``brain surgery is very easy and anybody can do it, but it is very hard and hardly anyone can do it if you require the procedure to be successful'').
 
@@
 It is not large enough or strong enough for safe human habitation!
-This point is obvious because we can see Lego houses, and everyone is familiar with building-block play.
+This point is obvious because we can see Lego houses, and everyone is familiar with the limitations of building-block play.
 Its real-world engineering limitations are too obvious to need stating.
@@
 
-Unlike building houses (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture.
+Unlike building houses (the Code of Hammurabi dates from around 1755\hskip .2em{\sc bc}\footnote{The Code of Hammurabi says, ``(\S 233) If a builder constructs a house [and] does not make it according to specifications, and a wall then buckles, that builder shall make that wall sound using his own silver.'' \cite{Hammurabi}}), programming is a new discipline, and the problems of poor programming are not widely appreciated or embedded in our culture.
 Professional standards, even when they exist, are not enforced.
@@
 
-\subsection{The low status of coding}
+\subsection{The low status of coding}\label{lowCodeStatus}
 Since programming appears to be so easy, developing code has a correspondingly low status in scientific practice (and more widely).
@@
 If they are theoretical papers, they will describe their mathematical models and data that is used to run or test their models.
-Outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods were used avoid or control for experimental error.
+However, outside of pure computer science, scientific papers are almost entirely silent on the code they rely on and how it was developed --- in particular, how the code might have been protected from bugs, analogously to how appropriate experimental methods were used avoid or control for experimental error.
 
@@
 
-In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code,TOP}, because  code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.
-Curiously, the importance of \emph{correct\/} code is rarely emphasized.
-
-Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement).
+Since published papers rarely mention their code, new papers contributing to the literature do not write about their code either, so the low status of code persists.
+
+In reaction to this vicious cycle, there is a growing movement to use and cite code correctly \cite{cite-code,TOP}, because code \emph{is\/} important, particularly for the reproduction, testing, and extension of any scientific work.
+(Code also needs to be correct, not just cited correctly.)
+
+%Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement).
 Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.
@@
 Because statistics, like code, is so readily susceptible to uncontrolled bias and error, there are many protocols and journal policies that enforce best practice, for example journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature.
-Yet PRISMA, like many such policies, ignores the critical role of code, and ignores the Software Engineering principles that assure code that research relies on is reliable and reliably reported.
-
-PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which is reasonable enough.
-PRISMA covers the review process carefully.
-For example, the authors should report the number of papers they included in their review.
-Perhaps $N=2000$.
-This number is then written into their draft paper, perhaps in several places.
-As the authors read and revise their paper, submit it, and respond to peer-reviewers, it is very likely that the number of papers in the survey changes, or other numbers and details may change.
-
-The authors now have a maintenance problem: where are the numbers that have changed, and what should they be changed to?
+PRISMA is a leading and influential research protocol, and it serves to make the point that the critical role of code is often strategically ignored.
+
+PRISMA is not concerned with the reproducibility of the literature reviewed, nor the reproducibility of systematic reviews themselves.
+PRISMA not only ignores the role of code, it ignores the Software Engineering principles that assure code that research relies on is reliable and reliably reported.
+
+PRISMA covers the review workflow.
+For example, it states that authors should report the number of papers they included in their review.
+Perhaps $N=2\hskip 0.3em 000$.
+This number will then written be into the review, perhaps in several places.
+As the authors read and revise their paper and respond to peer-reviewers, it is likely that the number of papers in the survey will change; other numbers and details will certainly change.
+
+The authors now have a maintenance problem that PRISMA does not address: where are the numbers that have changed, and what should they be changed to?
 Doing a search-and-replace, whether automated or by hand, is fraught with difficulties.
 What happens if $2000$ is used for some other purposes as well?
-What happens if some of the $2000$ values are written as $2,000$ or as $2000.0$?
+What happens if some of the $2000$ values are written as $2,000$ or as $2\hskip 0.3em 000.0$?
 What happens if some $2000$ are year dates and are changed incorrectly?
-Then there are the Human Factors: slips and errors will happen in this process anyway \cite{engima}.
+Then there are the Human Factors: slips and errors will happen in this workflow anyway \cite{engima}.
 Typos, slips during cut-and-paste, and other errors are common.
 Similar iterative revision cycles happen with any paper, not just with systematic reviews.
-PRISMA rightly deals with a wide range of methodological problems, but ignores the methodological problems (such as the example earlier in this paragraph) that using code na\"\i vely raises.
-
-The irony is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations.
-Such rule schemes reinforce the fallacy that code is trivial and unimportant.
-PRISMA consciously warns about  methodological issues, but ignores that poorly managed code raises comparable methodological issues that undermine the validity of the research it supports.
+PRISMA ignores the methodological problems of using code (such as the example earlier in this paragraph).
+
+The irony is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented, despite this being one of PRISMA's explicit motivations.
+PRISMA explicitly warns about methodological issues, but ignores that poorly managed code raises methodological issues that will undermine the validity of research it supports.
+PRISMA reinforces the culture that code is trivial and incidental to research.
 
@@
 Critiques of data and model assumptions are increasingly common \cite{critiques,diagnosis-reviews} but program code is rarely mentioned.
-Yet data and program are formally equivalent (see \supplement, section \ref{on-code-data-publication}).
-Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and probably useless.
+Program code has as great an effect on results as the data; in fact, without code, the data would be uninterpreted and almost useless.
 Code, however, is harder to scrutinize, which means that errors in code have subtle, often unnoticed effects on results.
 
-Almost all code contains ``magic numbers'' --- that is, data masquerading as code.
-This common practice ensures that published data is very rarely all of the data because it omits the magic numbers embedded in the code.
+Program code contains data.
+Almost all code contains ``magic numbers'' --- that is, data masquerading as code (see \supplement, section \ref{on-code-data-publication}).
+This common practice ensures that published data is rarely all of the relevant data because it omits the magic numbers embedded in the code.
 Such issues emphasize the need for repositories to require the inclusion of code so all data, including that embedded in the code, is actually available.
+
+Conversely, much data contains code.
+Excel spreadsheets are often used to manage code, and almost all Excel spreadsheets contain macros and formul\ae\ --- code embedded in the data.
+JSON, the JavaScript Object Notation, is a structured data language, but as it is part of the JavaScript programming language there is no practical code/data distinction.
+Indeed, the present paper's data is in a file \texttt{data.js} that also contains the JavaScript code to analyze it and generate results (such as Table \ref{table-summary}) for presentation in this paper.
+
+Although convenience and convention treat data and code differently, ultimately, data and code are formally equivalent (see \supplement, section \ref{on-code-data-publication}) in the important sense that there is no pre-determined boundary between them or formal criteria to distinguish them: different scientific projects can draw their boundaries differently and as they see best fits their work.
+Indeed, all of the papers reviewed here where code was available include critical data in their code.
+It therefore makes no sense to have separate rules for data and code --- except in the most trivial cases \emph{both\/} are equally essential for verification, building on, and reproduction of work.
 
@@
 Random numbers are widely used in computational science (and in many of the papers surveyed), for simulation or for randomizing experiments.
-Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of bugs nai\"\i ve bugs \cite{knuth}.
+Misuse of random numbers (e.g., using standard libraries without testing them) is a very common cause of nai\"\i ve bugs \cite{knuth}.
 
 If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs.
-Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time).
-For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted, but are bugs in the wider sense of producing incorrect results (e.g., ill-conditioning); that is, what was intended was~wrong.
+Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit}.
+Intentional bugs occur in code that correctly does what was intended, but what was intended was itself faulty (students and inexperienced programmers regularly make intentional bugs).
+Intentional bugs frequently arise in numerical modeling, where using an inappropriate method can introduce errors that are not bugs in the sense of failing to correctly implement what was wanted, but are bugs in the sense that the wrong numerical method was chosen and inaccurate results are obtained; that is, what was intended was~wrong.
 
@@
 The programming language C which was used \cite{tweet} is, like many popular programming languages, not a dependable language; to develop reliable code in C requires professional tools and skills.
-Some of the code was written in a na\"\i ve style (e.g., writing \texttt{*(a + i)} instead of \texttt{a[i]}, and with obscure numerical goto statements like \texttt{if(l == 0) goto S150}), and with C code that was translated simplistically from FORTRAN and Pascal code, from references dating from the 1970s and 1980s \citeeg{forsyth,Normal-Distributions}.
-
-Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, and hardware.
-In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C should be used: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}.
-Alternatively, a high integrity programming language, unrelated to C, such as SPARK Ada \cite{spark}, or modern languages (many related to the ``ML family'') like OCaml, F*, Haskell \cite{haskell}.
-These languages have learning curves; however their key benefit is that correct programs are far more likely and are \emph{much\/} faster to write.
+Some of the code was written in a na\"\i ve style (e.g., writing \texttt{*(a + i)} instead of \texttt{a[i]}, and with obscure numerical goto statements like \texttt{if(l == 0) goto S150}), and with C code that was translated simplistically from FORTRAN and Pascal code, from references dating back to the 1970s and 1980s \citeeg{forsyth,Normal-Distributions}.
+
+Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, or hardware.
+In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is preferred: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}.
+Alternatively, a high integrity programming language, unrelated to C, such as SPARK Ada \cite{spark}, or modern languages (many related to the ``ML family'') like OCaml, F*, Haskell \cite{haskell} could be used.
+These languages have steeper learning curves; however their key benefit is that correct programs are far more likely and are \emph{much\/} faster to write.
 (The \supplement\ discusses these issues further.)
@@
 
-In fact, the Ferguson code, \texttt{covid-sim}, is very large at \covidsimkLOC\ kLOC (thousands of lines of code),\footnote{Ferguson's \texttt{covid-sim} system is composed of \covidsimfiles\ files, and uses \covidsimdata\ of data.
+In fact, the Ferguson code, \texttt{covid-sim}, is a very large program at \covidsimkLOC\ kLOC (thousands of lines of code),\footnote{Ferguson's \texttt{covid-sim} system is composed of \covidsimfiles\ files, and uses \covidsimdata\ of data.
 It is now rewritten from C into C++ with Python, R, sh, YAML/JSON, etc.
-For more details, see \supplement.} so it is implausible that the ``mathematics or science'' has been correctly implemented in it.
-Ferguson's \emph{reported\/} science unlikely to be reliable.
+For more details, see \supplement.} so it is implausible that the ``mathematics or science'' has been correctly implemented in it without error, particularly as there is no discussion of methodologies to code reliably.
+Ferguson's \emph{reported\/} science is consequently unlikely to be reliable.
 
@@
 
-Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now open-source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).} Documenting code in hindsight, even if done rigorously, may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was originally intended to have done.
+Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now opeve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
+The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
+n-source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021).} Documenting code in hindsight, even if done rigorously, may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was originally intended to have done.
 As the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models; it will be different.
 
-It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
-The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
-
+It is well-known that documenting code helps impro
 Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
@@
 \label{section-discussion}
-\subsection{The narrow emphasis on data}
+\subsection{The lop-sided emphasis on data}
 \label{critique-fair}
@@
 
-Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
+Using structured repositories that provide suggestions for and which encourage good practice (such as Dryad\footnote{Dryad \url{datadryad.org} curates raw, unprocessed data.
+At the time of writing, Dryad excludes code; however, it uses a separate organization, Zenodo \url{zenodo.org}, to host code and other relevant information.
+This arbitrary separation is unfortunate as it increases management problems, increases reproducibility problems, limits using RAP, and most seriously limits how scientists can structure their data and code to best suit their research (see section \ref{knowledge}).} and GitHub), and requiring their use, would be a lever to improve the quality and value of code and documentation in published papers.
 The evidence (see \supplement) suggests that, generally, some but rarely all develop code that is uploaded to a repository just before submitting the paper in order to ``go through the motions.'' In the surveyed papers there is no evidence (before, during, or after the date of the survey sample) that any published code was prepared or maintained \emph{using\/} repositories.
-This is consistent with the finished code being uploaded to a repository just for the purposes of satisfying publishing requirements, but not using one earlier probably because they did not understand the benefits of doing so.
-
-Using a standard repository for lodging a paper's supporting code would help other scientists access the code easily, but not using the repository for developing and maintaining the code means the author of the paper misses out on many helpful features of repositories, such as version control, open-source development and review, actions and other approaches for automating development, sharing workload, and so on.
-
-In summary, there is a lop-sided emphasis on data in science.
-In fact, data is useless without code, and code must be used to analyze data and can manipulate it in arbitrary ways --- some ways, like renormalization and other transformations, deliberately, and some accidentally caused by bugs and unwittingly misunderstanding programming techniques.
+This is consistent with the finished code being uploaded to a repository just for the purposes of satisfying publishing requirements, but not using one earlier probably because they did not understand the benefits of doing so --- not using a repository \emph{during\/} the research process means the author of the paper misses out on the many helpful features of repositories, such as version control, review, actions, and other approaches for automating development, sharing workload, and so on.
+Using repositories during research means that other people can more easily help review the approach, in much the same way that papers are routinely circulated for peer review before submitting to a formal journal.
+
+There is a lop-sided emphasis on data in science.
+In fact, data is useless without code, and code must be used to manipulate and analyze it.
 Often code is used to extrapolate data, so the code itself effectively generates more data, or the code eliminates outliers so it effectively deletes data.
-Data is routinely formatted in standard ways (else code could not process it), but code is generally very architecture- and version-specific, so --- unless properly managed --- code goes obsolete faster than data.
-And so on.
+Data is routinely formatted in simple or standard ways, but code, in contrast, is architecture- and version-specific, so --- unless properly managed --- code goes obsolete faster than data.
 In short: the integrity of code and its availability to scrutiny is in fact both harder and more important than the usual requirements put on data.
 
-\section{Extending RAP to \RAPstar}\label{RAPstar-section}
-Code is usually seen as an independent set of files that are run to generate results, typically to be copied into a paper.
-But we can go further: a paper can itself embed code or become code \citeeg{relit,Executable-Paper}.
-This supports the generalization of RAP to form \RAPstarp.
-Essentially, \RAPstar\ is the recognition that computing is not just about programming computers (which results in RAP) but is about applying computational thinking \cite{computational-thinking-wing,computational-thinking} that supports and constructively analyzes any process, and in particular human and scientific processes that may not normally be computer-based at all.
-
-Once processes in the pipeline are automated, this means that there is code to run the steps again.
-Once there is code, it can be managed in a version control system.
-A version control system then provides an audit trail for free, as well as many advantages such as being able to backtrack to an earlier version to review earlier edits.
-Importantly, code can also perform sanity checks on the process --- a very simple example is automatic bibliography systems that check that journal names, DOIs are correct, and so forth.
-(They also allow the bibliographic data to be pooled and curated with other scientists, which improves its scope and quality.)
-
-GitHub is a tool that provides \emph{actions\/}, which are named specifications to run RAP's analytic pipelines, or workflows in GitHub's terms.
-GitHub happens to specify actions in the language YAML, which, being a textual notation, in turn means that all the helpful features of GitHub --- open-source, version control, and so on --- can be applied to the pipelined processes as well.
-Research pipelines can thus be made explicit, documented, shared and improved with open collaboration.
-
-Almost the entire scientific process can be automated (even with its interaction with the world automated by sensors and robots).
-There are many ways to do this; for example, \emph{Mathematica\/} makes the analysis of the data and the calculations and the paper ``the same thing'' in its integrated notebook user interface.
-The many alternatives include R Markdown, an approach based in the open mathematical system R \cite{RMarkdown}; a system, Lepton \cite{lepton}, which allows a \LaTeX\ document to execute and include arbitrary code, and language-independent notebook systems like Jupyter; and so on.
-There are also variations of literate programming, notably relit \cite{relit}, that enable papers themselves to generate the code they rely on and may describe, which therefore ensures the code run to support papers is exactly the code the author ran to support the claims.
-
-In all such systems, running the computational paper creates the publication.
-Indeed, every time the paper is run, the authors are likely to check the results after they have been re-computed, so the RAP process actively helps reduce errors.
-
-Here is the insight: the code, just like the paper is text, so the code \emph{itself\/} can fully form part of the pipeline, made reproducible and benefit from all the RAP benefits.
-To date, this critical point has been overlooked.
-Since using RAP to improve code, rather than just the pure scientific pipeline, has not been mentioned previously, we call it \RAPstar\ to make it clear this is a new and important generalization.
-\RAPstar\ helps improve code quality, for the same reasons RAP improves the quality of the paper and the science.
-
-Software engineers have many tools for automatic code development (such as Unix's \texttt{make}) but the idea that these tools can be used to integrate and help automate code authoring as well as its documentation \emph{and\/} paper authoring is radical.
-Amongst other things, it means that the \emph{entire\/} research and development process of the paper \emph{including\/} all its underlying code can be reproduced and reused by others.
-The present paper is a modest example of \RAPstarp; more details are given in the \supplement.
-
-Note that as \RAPstar\ objectifies how science is done to a standard sufficient to enable a computer to run it.
-\RAPstar\ enables all of the methodologies of Software Engineering to be brought to bear \emph{on the science itself}.
-\RAPstar\ means the normally tacit, manual, and undocumented processes of science become explicit code.
-Code can then be scrutinized, optimized, and ensured correct by standard Software Engineering practice --- thus \RAPstar\ does not just automate science for reproduction, it makes the automation explicit so the doing of science itself can be reasoned about --- not just mentally but supported by sophisticated tools, such as theorem proving and AI\@.
-Science will be improved.
-
-\section{The paper as a scientific laboratory}\label{paper-as-lab}
-The conventional view of science is that experiments are done \emph{then\/} written up. 
-%RAP and \RAPstar\ support and enhance this process with computational tools; in particular, the archival results of research are not just the reports and publications, but include the executable processes --- the pipelines in RAP terms --- of the research.
-However, it is more productive to think of the paper itself as an active laboratory, not just as a record of finished work.
-Viewing the paper itself as a laboratory encourages authors to copy and adopt laboratory best practice (such as keeping records, as RAP and \RAPstar\ suggest) into the preparation of the paper itself; it also encourages authors to see writing as a scientifically --- not just expressive --- creative act, and not just as the final summary of a period of scientific creativity that is not written down.
-In short, seen as a laboratory, writing the paper is no longer a reactive write-up of finished work, but it is an active part of doing the science itself.
-Writing a paper thus explores the space of scientific possibility as constructively as working in the field or on a bench: the paper is a laboratory.
-
-With a computational paper, authors can literally experiment \emph{in\/} the paper, exploring the effectiveness of ideas and explanations.
-Furthermore, they can experiment with hypotheses: for example, they can make a clear claim that at that point in the lifecycle of the paper is but a wish rather than an established fact --- sketching a direction they plan to go in and develop supporting arguments and evidence for.
-So they then do the experiments or calculations (including consulting other scientists and peer reviewers) to establish a justification and other details.
-Typically, the evidence they generate or the criticism they receive may not be quite what they expected, so they then revise the claim to be correct, or change it to be a more realistic claim still as yet to be fully substantiated, or they could delete it if it just turns out to be a mess, or they could develop some altogether much better approach to the work as a result of the exploration.
-
-Typically, many ideas in the paper will be linked to data from conventional experiments.
-For example a computational paper might calculate that the statistical power of an experiment is too low (there is an unacceptable risk of committing Type II errors), so the authors will decide to improve the experiments.
-
-If there is code in the paper, every time it is typeset, the code will be run.
-Therefore the authors of the paper proof-reading the results of the code will have opportunities to try to debug and improve the code.
-Again, the paper itself is acts like a laboratory, helping the authors refine the science.
-
-Note that systems capable of handling computational papers (including \LaTeX) can create conditional documents: for example, there could be a flag \texttt{publish}.
-If \texttt{publish} is false, the author can see all their private work and thinking, including all their experimental thinking and workings; but if \texttt{publish} is true, the paper would be typeset as for a submitted paper with the detailed working concealed to ensure a clean and concise presentation.
-In principle, also submitting to the journal with \texttt{publish} false (hence with data and workings visible) could satisfy journals' code and data requirements.
-Of course, when there is a large body of data or code, they need not all be an explicit in the computational paper: they would be made available separately in the usual way.
-
-The more we view the paper as a proactive scientific laboratory, the more we gain from the \RAPstar\ perspective, and the more science gains from improved, conceptually broadened, reliable, and reproducible science.
-The more we will want to engage mature Software Engineering standards too, because the quality and creativity of future science relies on them.
-
-\section{Call to action}\label{summary}
-Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for the same reasons: computer bugs accidentally cultured in one laboratory can infect research and policy worldwide.
-
-Inadequate code can be extremely problematic.
-Incorrect results might be used for supporting science, modeling COVID-19, informing public health policy, medical research, or used in other \emph{critical software}.
+\section{Rethinking science that uses code}\label{summary}
+Computer programs are the laboratories of modern scientists, and should be used with a comparable level of care that virologists use in their laboratories --- lab books and all \cite{notebooks} --- and for exactly the same reasons: bugs, whether computer bugs or biological bugs, accidentally cultured in one laboratory can infect research, ideas, and policy worldwide.
+
+Inadequate scientific code can be problematic.
+Incorrect results might be used for supporting science, modeling pandemics or informing public health policy, informing medical research, or adopted for use in other critical software, such as medical diagnostics, credit checking, or any other impactful use.
 Professional critical software development, as used in critical industries such as aviation and the nuclear power, is (put briefly) based on \emph{correct by construction\/} \cite{cbc}, effectively: design it right first time, supported by numerous rigorous techniques, such as Formal Methods, to manage error.
-Not coincidentally, such are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable, as required for supporting quality science.
+Not coincidentally, these are \emph{exactly\/} the right methods to ensure code is both dependable and scrutable, as is required for supporting reproducibility and quality science more generally.
 Conversely, not following these practices undermines the rigor of science.
 
-An analogous situation arises in ethics.
-
+\subsection{Software Engineering Boards}
 Misuse of data, exploiting vulnerable people, and not obtaining informed consent are typical ethical problems.
-Planned research may be ethically unacceptable, but few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
+Planned research may be ethically unacceptable in ways the investigators do not anticipate: few people have the objectivity and ethical expertise to make sound ethical judgements, particularly when it comes to assessing their own work.
 National funders, and others, therefore require Ethics Boards to formally review ethical quality.
-Medical journals will not publish research that has not undergone appropriate formal ethical review.
-
-Analogously, and supplementing Ethics Boards, it is proposed that Software Engineering Boards would authorize as well as provide advice to guide the implementation of high-quality Software Engineering to support research and publication processes.
-Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to  all scientific papers and funded research being required to include Software Engineering Board support and clearance statements as appropriate.
-Note that Software Engineers themselves have a code of ethics that applies to \emph{their\/} work \cite{ethics-code}.
-
-Some journals have policies that code is available (see the \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
-Ideally, the authors should test a clean deployed build of their code and save the results.
-Presumably a paper's authors must have run their code successfully on \emph{some\/} data (if any, but see section \ref{on-code-data-publication} in the \supplement) at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of  code underpinning \emph{any\/} scientific claims.
+Medical journals will not publish research that has not undergone independent formal ethical review.
+
+Analogously, and supplementing Ethics Boards, it is argued here that Software Engineering Boards (SEBs) would authorize as well as provide advice to guide the implementation of quality Software Engineering to support research and publication processes.
+Just as journals require conflicts of interest statements, data availability statements, and ethics board clearance, we should move to scientific papers and funded research being required to include formal Software Engineering Board statements.
+Note that Software Engineers themselves have a code of ethics that applies to their own work \cite{ethics-code}.
+
+Some journals have policies that code is to be made available (see \supplement), but they should require that code is not just available in principle but \emph{actually works\/} on the relevant data.
+The authors should test a clean deployed build of their code and save the results.
+Presumably a paper's authors must have run their code successfully on \emph{some\/} data at least once, so preparing the code and data in a way that is reproducible should be a routine and uncontentious part of the rigorous development of code underpinning \emph{any\/} scientific claims.
 This requirement is no more unreasonable than requesting good statistics, as discussed earlier.
-And the solution is the same: relevant experts --- whether statisticians or Software Engineers --- need to be routinely available and engaged with the science.
-Software Engineering Boards would be a straight forward way of helping achieve this.
-
-There need to be many Software Engineering Boards (SEBs) to ensure convenient access and oversight, potentially at least one per university.
-Active, professional Software Engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the true state of the art.
-There are many high-quality software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.
-
-Open-source generally improves the quality of software.
-SEBs will take account of the fact that open-source software enables contributors to be located anywhere, typically without a formal contractual relationship with the leading researchers.
-Where appropriate, then, SEBs might require \emph{local\/} version control, unit testing, static analysis and other quality control methods for which the lead scientist and Software Engineer remain responsible, and may even need to sign off (and certainly be signed off by the SEB\@).
-Software Engineering publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software for use in peer-reviewed publications \cite{acm-artifacts}.
+And the solution is the same: that relevant experts --- statisticians or Software Engineers --- need to be available and engaged with the science.
+Software Engineering Board statements would be a straight forward way of helping achieve this and showing that it has been done adequately.
+
+There need to be many SEBs to ensure convenient access, potentially at least one per university.
+Active, professional Software Engineers should be on these SEBs; this is not a job for people who are not qualified and experienced in the area or who are not actively connected with the state of the art.
+There are many high-quality university computer science departments and software companies (especially those in safety-critical areas like aviation and nuclear power) who would be willing and competent to help.
+
+As appropriate, SEBs might require version control, unit testing, static analysis, and other quality control methods.
+Within the field of Software Engineering itself, publishers are already developing rigorous badging initiatives to indicate the level of formal review of the quality of software \cite{acm-artifacts}.
 
 A potential argument against SEBs is that they may become onerous, onerous to run, and onerous to comply with their requirements.
-A more mature view is that SEBs need their processes to be adaptable and proportionate; indeed, few people consider Ethics Boards to be disproportionately onerous.
+A more balanced view is that SEBs need their processes to be adaptable and proportionate; indeed, few people consider Ethics Boards to be disproportionately onerous.
 If software being developed is of low risk, then less stringent engineering is required than if the software could cause frequent and critical outcomes, say in their impact on public health policy for a nation.
@@
 There are standard ways to do this, such as following IEC 61508:2010 \cite{redmill,iec61508} or similar.
-Following a risk analysis (based on safety assurance cases, controlled documents and so on, if appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.
+Following a risk analysis (based on safety assurance cases, controlled documents and so on as appropriate to the domain), the Board would focus scrutiny where it is beneficial without obstructing routine science.
 
@@
 SEBs could be quickly established to provide direct access to mature Software Engineering expertise for both researchers and for journals seeking competent peer-reviewers.
-In addition, particularly during a pandemic, SEBs would provide direct access to their expertise for Governments and public policy organizations.
+In addition, particularly during a pandemic or other disasters, SEBs would provide direct access to their expertise for Governments and public policy organizations.
 Given the urgency, this paper recommends that \emph{ad hoc\/} SEBs should be established for this purpose.
@@
 An analogous proposal to SEBs has been made for Methods Review Boards \cite{mrb}, to help scientists ensure the methods they use are appropriate for addressing their research questions.
-The Methods Review Boards was motivated by an Ethics Board member noticing that often experimental methodologies are inadequate, and will waste time that cannot be corrected until the flaws are spotted too late during peer-review.
-The paper \cite{mrb} raises many of the same trade-offs that SEBs also face; indeed one would hope that Methods Review Boards would include Software Engineers or have SEBs as sub-boards --- as this paper has argued, Software Engineering is now a key methodology of science.
-As with SEBs, the goal is not to gatekeep, but to improve.
+Methods Review Boards were motivated by an Ethics Board member noticing that often experimental methodologies are inadequate, which will waste time that will not be corrected until the flaws are raised, usually too late, typically during peer-review --- creating technical debt (discussed below, in section \ref{technical-debt}).
+As with SEBs, the goal of Methods Review Boards is not to gatekeep, but to improve.
+The paper \cite{mrb} raises many of the same trade-offs that SEBs also face; indeed one would hope that Methods Review Boards would include Software Engineers or have SEBs as sub-boards or \emph{vice versa\/}: Software Engineering is now a key methodology of science.
+
+\subsection{Extending RAP to \RAPstar}\label{RAPstar-section}
+Code is usually seen as an independent set of files that are used to generate results, typically to be copied into a paper; code is usually seen as a passive part of science.
+In reality, code is very creative.
+A paper can itself embed code or become code \citeeg{relit,Executable-Paper}, as discussed in section \ref{computable-papers}: code then becomes a driver for the research.
+This view supports the generalization of RAP to form \RAPstarp.
+Essentially, \RAPstar\ is the recognition that coding is not just about programming computers (which results in RAP) but is about applying computational thinking \cite{computational-thinking-wing,computational-thinking} that supports and constructively analyzes any process, in particular the creative scientific processes of doing science and creating archival publications.
+
+Once workflow steps in the pipeline are automated, then there is code to run the steps again.
+Once there is code, it can be managed in a version control system.
+A version control system then provides an audit trail for free, as well as many advantages such as being able to backtrack to an earlier version, for instance to review earlier edits.
+Importantly, code can also perform sanity checks on the process.
+A very simple example is automatic bibliography systems that check that journal names and DOIs are correct, and so forth.
+(Bibliographic systems also allow the bibliographic data to be pooled and curated with other scientists, which improves its scope and quality.) But \RAPstar\ goes far beyond bibliographies; there is far more of the scientific workflow that can be partly or fully automated --- and with corresponding benefits.
+
+GitHub is a tool that provides \emph{actions\/} that are named specifications to run analytic pipelines, \emph{workflows\/} in GitHub's terminology.
+GitHub happens to specify actions in the language YAML, which, being a textual notation, in turn means that the features of GitHub --- open-source, version control, and so on --- can be applied to the research workflow as well.
+Research pipelines can thus be made explicit, documented, shared and, most importantly, critiqued and improved.
+
+The entire scientific process can be supported by automation (especially with its interaction with the world automated by sensors, AI, and robots).
+There are many ways to do this; for example, \emph{Mathematica\/} makes the analysis of the data and the calculations and the paper ``the same thing'' in its integrated notebooks.
+The many alternatives include R Markdown, an approach based in the open mathematical system R \cite{RMarkdown}; a system, Lepton \cite{lepton}, which allows a \LaTeX\ document to execute and include arbitrary code, and language-independent notebook systems like Jupyter; and so on (section \ref{computable-papers}).
+
+In all such systems, running the computational paper creates the publication.
+Indeed, every time the paper is run, the authors are likely to check the results and fix any problems, so an explicit RAP workflow actively helps reduce errors.
+
+Here is the insight: the paper's code, just like the paper itself is text, so the code \emph{itself\/} can fully form part of the pipeline, made reproducible, and benefit from all the usual RAP benefits.
+To date, this critical point has been overlooked.
+Since using RAP to integrate code, rather than just the conventional ``pure'' scientific workflow, has not been mentioned previously, we call it \RAPstar\ to make it clear this is a new and important generalization of RAP\@.
+\RAPstar\ helps improve code quality, for the same reasons RAP improves the quality of the science.
+Improving code quality improves the science and its reproducibility.
+
+Software engineers have many tools for automatic code development (such as Unix's \texttt{make}) but the idea that these tools can be used to integrate and help automate code authoring as well as its documentation \emph{and\/} paper authoring is radical.
+It means that the \emph{entire\/} research and development process of the paper \emph{including\/} all its underlying code can be reproduced, reused, and developed by others.
+The present paper is an example of \RAPstarp; more details are given in the \supplement.
+
+By definition, \RAPstar\ objectifies how science is done to a standard sufficient to enable a computer to run it.
+\RAPstar\ therefore enables all of the methodologies of Software Engineering to be brought to bear \emph{on the science itself}.
+\RAPstar\ means the normally tacit, manual, and undocumented processes of science (especially the coding) become explicit.
+Code can then be scrutinized, optimized, and ensured correct by standard Software Engineering practice; thus \RAPstar\ does not automate science just for easier reproduction, it makes the automation explicit so the doing of science itself can be reasoned about --- not just by scientists but supported by sophisticated tools, such as theorem proving and AI\@.
+Science will be improved by \RAPstar.
+
+\subsection{The paper as a scientific laboratory}\label{paper-as-lab}
+The conventional view of science is that experiments are done \emph{then\/} written up. 
+%RAP and \RAPstar\ support and enhance this process with computational tools; in particular, the archival results of research are not just the reports and publications, but include the executable processes --- the pipelines in RAP terms --- of the research.
+However, it is more productive to think of the paper itself as an active laboratory, not just as a record of finished work.
+The view of the paper as a scientific laboratory is explicit in computational papers (section \ref{computable-papers}): ideas are written down, and their validity is tested by the sense they make or fail to make; the ideas are then revised --- writing is an active experiment to find and develop ideas that are worth saying.
+Viewing the paper as a laboratory encourages authors to copy and adopt laboratory best practice (such as keeping records, as RAP and \RAPstar\ suggest) into the processes of writing the paper itself; viewing the paper as a laboratory also encourages authors to see writing as a scientifically --- not just expressive --- act, and not just as the final summary of a period of scientific creativity.
+In short, seen as a laboratory, the paper is no longer a reactive write-up of finished work, but it is an active part of doing good science.
+Writing a paper explores the space of scientific possibility as constructively as working in the field or on a lab bench.
+The computable paper is now the scientific laboratory.
+
+With a computational paper, authors can literally experiment \emph{in\/} the paper, exploring the effectiveness of ideas and explanations.
+Furthermore, they can experiment with hypotheses: for example, authors can make a clear claim that at that point in the lifecycle of the paper is but a wish rather than an established theory or fact --- the wish enables them to sketch a direction they plan to go in and to explore possible supporting arguments and evidence for it.
+So they then do the experiments or calculations, including consulting the literature and other scientists, to establish a justification and other details.
+The evidence they generate or the criticism they receive may not be quite what they expected, so they then revise the claim to be correct, or change it to be a more realistic claim, or they could delete it if it just turns out to be a mess, or they could develop some altogether much better approach to the work as a result of the exploration.
+
+Typically, many ideas in a paper will be linked to conventional experiments.
+For example a computational paper might calculate the statistical power of an experiment is too low (there is an unacceptable risk of committing Type II errors), so the authors will decide to improve the experiments.
+The computational paper approach allows such calculations to be made before, during, or after the experiments.
+
+If there is code in the paper, every time it is typeset, the code will be run.
+Therefore the authors of the paper proof-reading the paper and the results of the code will have opportunities to debug and improve the code.
+Again, the paper itself is acts like a laboratory, helping the authors refine the science.
+
+Note that systems capable of handling computational papers (including \LaTeX) can create conditional documents: for example, there could be a flag \texttt{publish}.
+If \texttt{publish} is false, the author can see all their private work and thinking, including all their experimental thinking and workings; but when the author sets the flag \texttt{publish} to true, the paper would be typeset for a submitted paper with the detailed workings concealed to ensure a clean and concise presentation.
+In principle, also submitting to the journal a separate version of the paper with \texttt{publish} set to false would make the data and workings visible and thus could satisfy journal code and data requirements.
+Of course, when there is a large body of data or code, they need not all be an explicit in the computational paper: they would be made available separately in the usual way --- the flag \texttt{publish} rather than being true/false might instead be a number setting the degree of disclosure.
+
+The more we view the paper as a proactive scientific laboratory environment, the more we gain from the \RAPstar\ perspective, and the more science gains from improved, conceptually broadened, reliable, and reproducible science.
+The more we will also want to engage mature Software Engineering (and computational thinking) ideas too, because the quality and creativity of future science relies on them, after all, Software Engineering is about how to tell a computer how to do something reliably, and, as Knuth said, science is what we understand well enough to explain to a computer, and so Software Engineering can now directly help ensure we are not fooling ourselves about what we understand.
 
@@
 \setcounter{footnote}{1}
-Code is only part of science, and only one critical factor in the wider reproducibility crisis: SEBs must work with --- and be engaged by --- other reproducibility initiatives, such as TOP and Methods Review Boards.
-
-Relying on SEBs \emph{alone\/} would continue one of the current besetting problems about the role of code.
-The conventional view is that scientists do the hard work compared to the ``easy'' coding work (sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}) so they just need to tell programmers what to do.
-This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, which he explores at some length, is that we need to spend more effort in working out what computers should do (i.e., do the science better) and then \emph{just\/} tell programmers to do \emph{that}.
-
-On the contrary, competent Software Engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and how that needs refining or optimizing.
-In other words, the Software Engineers can bring important insights into the science, hence improving or changing the questions and assumptions the science relies on.
-This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but it's ill-conditioned, there is no good answer to that question.'' Ideally, then, it is not  a simple sequential process
-
-\begin{center}\sf\fbox{science specifies} $\rightarrow$ \fbox{code up} $\rightarrow$ \fbox{get results}, 
+Code is more than a scientific instrument, more than a thermometer or test tube, as code makes, informs and changes decisions; indeed, code can actually \emph{do\/} science.
+Still, code is only a part of science, so relying on SEBs \emph{alone\/} would continue one of the besetting problems about the role of code in science.
+
+The conventional view is that scientists do the hard work compared to the ``easy'' coding work (sections \ref{deceptive-simplicity-of-code} \& \ref{central-role-of-code}) so they just need to tell coders what to do.
+This is the view expressed by Landauer in his classic book \emph{The Trouble with Computers\/} \cite{landauer,thimbleby-landauer}, where he argues that the trouble with computers, an idea he explores at some length, is that we need to spend more effort in working out what computers should do (here, do the science better) and then \emph{just\/} tell programmers to do \emph{that}.
+
+On the contrary, competent Software Engineers have insights into the logic, coherence, complexity, and computability of what they are asked to do, and how it needs refining or optimizing --- or the question changing.
+In other words, Software Engineers can bring important insights into the science, hence improving or changing the questions and assumptions the science relies on.
+This insight was widely recognized in the specialist area of numerical computation: ``here is a formula I want you to just code up''~\ldots\ ``but it's ill-conditioned, there is no good answer to that question.'' It is not a simple sequential workflow with the expert initiative all on the left:
+
+\begin{center}\sf\fbox{science specifies} $\rightarrow$ \fbox{code up} $\rightarrow$ \fbox{get results}
 \end{center}
 
-\noindent but an iterative cycle of collaboration and growing understanding, informed by Software Engineering best practice (via SEBs), and implemented using papers as laboratories.
+\noindent but an iterative cycle of mutual collaboration and growing understanding, informed by Software Engineering best practice (via SEBs) \emph{and\/} science, and implemented and tested-out using papers as laboratories.
 
@@
 Software engineers can help improve the science, so it is not just a matter of asking a SEB whether some coding practices (like documentation) are satisfactory, but whether the SEB has insights into the science itself too.
-Most effectively, this requires interdisciplinary working practices (science plus Software Engineering) with mutual respect for their contributing expertise.
+The SEB idea requires interdisciplinary working practices (science plus Software Engineering) with mutual respect for their contributing expertise.
 
 \subsection{Methodological statements}
-Many science journals, conferences, workshops, videos, books, etc, and funders require explicit statements how the authors have conformed to appropriate methodological standards covering issues such as conflicts of interest, ethics, data access, consent, authoring, funders and other acknowledgements of support, and so forth.
-Conformance to PRISMA, discussed above in section \ref{PRISMA-statement}, is one such methodological statement; like many journals, the high-profile journal \emph{PLOS ONE\/} has a list of similar methodologies it expects authors to comply with as appropriate.
-
+Scientific publishers (journals, conferences, workshops, videos, books, etc, and funders) often require an explicit methodology discussion, yet they rarely require the methodology to discuss the computational methodology --- which, to the extent that anything relies on code, impacts all the other methodology and results discussions.
+
+Many science publishers require explicit statements how the authors have conformed to appropriate methodological standards covering issues such as conflicts of interest, ethics, data access, consent, authoring, funders and other acknowledgements of support, and so forth.
+Conformance to PRISMA (see section \ref{PRISMA-statement}) is one such methodological standard.
 It would be easy for journals and funders to require equivalent types of statements on the quality of code, that is on the quality of its Software Engineering.
@@
 Journals and funders should provide support for hosting data and code (and any other relevant data, such as qualitative data, video, etc), and the review process must check that authors actually provide the material as they claim in the methodological and access statements.
-Conversely, scientists should be able to access funding to ensure data and code access, and, as appropriate, funding for on-going maintenance of the databases, which will typically require funding beyond the end of the normal funding period.
+Conversely, scientists should be able to access funding to ensure data and code access, as well as funding for on-going maintenance of the databases, which will typically require funding beyond the end of the normal funding period.
 
@@
 Assurance arguments provide a concise, high-level argument that the system does what is claimed.
-Assurance arguments can be more or less detailed, and more or less formal in their approach; we envisage referees would have views on the level of detail and formality required for any specific contribution.
-
-Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access statements apply to all information, regardless of any arbitrary classification of it as code or data (or documentation, assurance case, etc).
+Assurance arguments can be more or less detailed, and more or less formal in their approach; editors and referees would have views on the level of detail and formality required for any specific contribution.
+
+Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access and methodology statements apply to all information, regardless of arbitrary classification of it as code or data (or documentation, assurance case, etc).
+
+\subsection{Training to reduce technical debt}\label{technical-debt}
+Science has to work for other people in other places at other times, otherwise they cannot be sure they are studying, developing, or correcting the intended ideas reliably, but while working on a research project, the requirements of reproducibility are tempting to postpone or ignore altogether.
+It seems more expedient to ``just'' get on and do the science without regard for the extra effort of ensuring reproducibility.
+This creates the problem known as \emph{technical debt} \cite{debt}: the savings in effort now increase the future cost of reproduction.
+That is, a debt arises as the authors' savings now create higher costs for scientists later.
+The authors of a paper may create debt for themselves as shortcuts now increase the effort of retro-fitting reproducibility later.
+Indeed, what if the \emph{post hoc\/} rigors of reproducibility expose previously unknown problems where the earlier shortcuts have created a now too-late-to-avoid cost authors with integrity will be obliged to pay?
+
+There is a trade-off to balance when and how much effort to put into reproducibility.
+The trade-off is comparable to trade-offs in using statistics --- the author may realize at a later stage that a ``significant'' result depends on designing the experiment appropriately for the intended statistical claims, but now making the claim rigorously requires revising the methodology and probably improving the type of analysis too.
+What was an easier route to take earlier now causes a challenging and costly revision.
+However, since mature scientists recognize the importance of correct statistics, statistics is positioned at the forefront of their work rather than delayed and sorted out at greater cost later.
+Statistics is routinely in the undergraduate syllabus, so most scientists consciously make appropriate trade-offs minimizing statistical technical debt.
+
+In contrast to statistics, reproducibility has only recently become an explicit issue, and computational tools to support it are developing rapidly; unlike statistics, very little of best reproducibility practice will be in scientists' background training.
+While systems like Jupyter facilitate reproducible science, realizing this may only come after much work has been done.
+Unfortunately, retro-fitting the science into a reproducibility tool is a steep learning curve --- no less than learning statistics from scratch.
+As with rigorous statistics, if the benefits of reproducibility processes were not realized at the start of a project and they have to be retrofitted, then the reworking of the science will be costly.
+Worse, Jupyter and similar tools are not ``just'' computational notebook systems that are easy to use: they work with a raft of inter-related technologies, such as Binder, Docker, Python, MyST, Sphinx\ \ldots\ as well as field-specific environments like Neurolibre, as well as with (or despite) costly tools the scientists may be relying on, including proprietary systems, high performance computing resources, and subscription services.
+
+Jupyter, and its many alternatives, are effectively lifestyle choices with dauntingly steep learning curves when they are learned on the job during research.
+The solution, as with statistics, is to push back learning about reproducibility and the benefits of tools to earlier in the scientific career, at the latest to the undergraduate curriculum before reproducibility-related technical debt can arise.
 
 \subsection{Benefits beyond science}\label{benefits-beyond-science}
-Science increasingly recognizes the key supporting roles of code and computation, but many fields do not recognize computation as such, as a skilled discipline, and therefore they are missing out on the leverage that comes with recognizing computation as a first class player in their  activities.
-
-Healthcare is supported by computers, yet medical research papers remain traditional and do not refer to code.
-Yet clinical practice relies on computer analysis, so inevitably must use code unrelated to the code developed in research, which is unlikely to be developed to Software Engineering standards.
-Conversely, the critical issues (including patient safety) that assume code is more reliable than that needed for scientific research do not get evaluated by researchers.
+Science increasingly recognizes the key supporting roles of code and computation, but many fields do not recognize computation as such as a skilled discipline, and therefore they are missing out on the leverage that comes with recognizing computation as a first class player in their activities.
+
+For example, healthcare research is supported by computers and code, yet medical research papers remain in a traditional pre-digital culture and do not refer to code, as if code has no influence in the methodology of the science.
+Yet clinical practice relies on computer code (e.g., for diagnostics), so inevitably practice must use code unrelated to the code developed in research.
+In other words, the culture of not discussing and sharing code in research reduces its impact: putting research into practice is a reproducibility question, and if code has been down-played in the research it will be reproduced unreliably.
+Conversely, the critical issues (including patient safety) that tacitly assume code is more reliable than required for scientific research code do not get evaluated by researchers.
 The gap is wide.
 The problems and missed opportunities of under-valued and poorly-managed code are ubiquitous in healthcare \cite{fixit}.
-Lasting solutions might well be initiated through SEBs or equivalent.
-
-Numerous problems in finance have been facilitated if not precipitated by computational na\" \i vety.
-JP Morgan Chase (JPM) lost over \$6 billion in a credit derivatives trade \cite{jpm}, in a costly parody of bad science --- compare the discussion here with figure \ref{excelFigure}, in a series of figures in the \supplement, which illustrates the same problem as it presents in many scientific papers.
-
-As reported in \cite{jpm}, the traders did not understand the trades, did not monitor them, doubled down when results were poor, and did not communicate the extent of their losses.
+Solutions might well be initiated through SEBs.
+
+Another example is that numerous problems in finance have been precipitated by similar computational cultural na\" \i vety.
+JP Morgan Chase (JPM) lost over \$6 billion in a credit derivatives trade \cite{jpm}, in a costly parody of bad science.
+As reported in \cite{jpm}, traders did not understand the trades, did not monitor them, doubled down when results were poor, and did not communicate the extent of their losses.
 They were using manual coding methods; as the report says:
@@
 
-The report does not detail how the Excel spreadsheets were specified or coded, seemingly as unawqre of Software Engineering as the traders.
-However, one infers from the brief discussion of data handling (which is easier to automate) that it was all an unconsciously incompetent process.
+Compare the discussion here with figure \ref{excelFigure}, in a series of figures in the \supplement, which illustrates the problem as it presents in many scientific papers.
+
+The report \cite{jpm} does not detail how the Excel spreadsheets were specified or coded, seemingly as unaware of Software Engineering as the traders. It was an unconsciously incompetent process.
 
@@
 Then there was a merry-go-round of blame: ``the information communicated to the Risk Policy Committee \ldots\ did not suggest any significant problems \ldots\ there was no robust debate with the right facts at the right level about the portfolio risk.'' UK and US governments are now investigating fraud.
-Again, lasting solutions might well be initiated through SEBs or equivalent.
-
-Without taking the lessons of improving Software Engineering to other fields, including improving and broadening the recognition and career paths for developers, there will continue to be an unfortunate and unnecessary disconnect between competent coding and actual practice.
-Everything, from healthcare to finance --- not just science --- will suffer because the critical contributions of dependable code, quality Software Engineering, and competent Computational Thinking are not recognized, understood, valued, or required.
-
-\ignore{\subsection{Suggestions for further work}
-Although this study of Software Engineering in scientific research (specifically, in peer-reviewed publications) was necessarily interpretive in nature, the corpus of data (namely, the selected \the\dataN\ papers) was rigorously gathered so that it could be later updated or extended in size or scope.
-However, the insights appear to be general.
-
-Further work to extend the scope of the survey beyond the basic requirements of the present paper is of course worthwhile, but the following cases (listed in the next few paragraphs) suggest that the problem is widespread.
-We argue, then, that we should be focusing effort on avoiding problems and considering proposed solutions (see section \ref{summary}), not just assessing the problems highlighted in this paper with increasing scale or rigor.
-}
+Again, lasting solutions might well have been initiated through SEBs or equivalent; involving SEBs might well avoid future financial fiascos.
+
+Without taking the lessons of improving Software Engineering to other fields, including improving and broadening the recognition and career paths for developers, there will continue to be unfortunate and unnecessary disconnects between competent software engineering and actual scientific (or medical or financial, etc) practice.
+Everything, from healthcare to finance --- not just science --- will continue to suffer because the critical contributions of dependable code, quality Software Engineering, and competent Computational Thinking are not yet recognized, understood, valued, or required.
+
+\subsection{Approaches to further work}
+Encouraging and informing the improvement of science and, specifically the reproducibility of science that relies on code, were the main aims of this paper.
+Further work should research the efficiency, effectiveness, and quality of the various ideas proposed, such as \RAPstar\ and Software Engineering Boards, and propose and evaluate more ideas.
+
+Further work to extend the reach and scope of the survey with increasing scale, subject coverage, and rigor beyond the exploratory requirements of the present paper might seem worthwhile.
+If people feel our analysis of the problem is inadequate, better surveys may be appropriate, but recognizing that there is a problem (regardless of arguing over its scale) in practice it is more important to explore what direction to travel in.
+We should be focusing effort on acknowledging, understanding, assessing, managing and avoiding scientific problems --- including poor reproducibility.
+This requires practical solutions that scientists can adopt, which itself relies on further work to examine rigorously what effective ``practical solutions'' might entail.
+
+Being primarily concerned with reproducibility, this paper avoided assessing the correctness of code used in science, not least because that without reproducibility correctness is moot --- how code is managed and made available is more relevant than exactly what it does.
+Indeed, since none of the papers reviewed provided code specifications, it is not obvious what correctness means to practicing scientists.
+The balance, then, between practical correctness and formal correctness is an important research topic to pursue \cite{hoare-unreasonable}.
+
+The present paper did not assess the correctness of code used in papers for several reasons.
+Code was not documented well-enough (even with high-level discussion in the papers) to know what ``correct'' would mean, and no papers performed adequate tests, let alone provided adequate test material for independent verification (see Table \ref{table-summary}); moreover, installing the software environments to build and test the systems --- different environments for each paper --- was too onerous, even when those environments were specified.
+The implication is that people within the relevant fields, especially referees, should promote standardized software environments to help increase the rate of reproduction and verification of results.
+As a matter for further research, then, it is important to develop, assess, and promote effective shared online (e.g., cloud) environments, perhaps with discipline-specific solutions, so that development and test environments are standardized, powerful enough, and sufficiently accessible.
 
 \section{Conclusions}
-In a pandemic, epidemiological modeling (discussed in this paper), track and trace \cite{excel-fiasco}, modeling mutation pressures against vaccine shortages and delays between vaccinations \cite{science-delays}, and so on, drive public policy and have a direct impact, whether positively or negatively, on the quality of life.
-
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID modeling, the wider evidence reviewed here suggests that scientific coding practice is inadequate in every field.
+A pandemic creates unprecedented pressure and exposes problems in scientific methodology.
+During the COVID crisis, code led epidemiological modeling, implemented track and trace and caused problems \cite{excel-fiasco}, modeled mutation pressures against vaccine shortages \cite{science-delays}, and more.
+Code drove public policy.
+Code had a direct impact on the quality of life.
+
+While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID pandemic modeling, the evidence reviewed here suggests that scientific coding practice is inadequate in every field, but particularly worrying in the context of the extreme pressures of managing a pandemic in real time.
 As science becomes more and more reliant on computers, we need to correspondingly improve the quality of code, the quality of code policies, the quality of Software Engineering, and the quality of all scientists' understanding of computation and how to manage its unlimited complexity.
@@
 Computer code should be explicit, accessible (well-structured, etc), and adequately documented.
-Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed more generally about the standards of science \cite{whitty}.
+Papers should be explicit on their software methodologies, limitations and weaknesses, just as Whitty expressed about the standards of science generally \cite{whitty}.
 Professional software methodologies should not be ignored.
@@
 \item 
-To use computation to help make scientific processes explicit, so that they can be reproduced, scrutinized and improved.
-RAP is an increasingly popular way to help do some of this, but as this paper points out, RAP should be generalized to \RAPstar\ to help the computational parts of science as well, leading to a virtuous circle.
+To use computation to help make scientific workflows and processes explicit, so that they can be reproduced, scrutinized, and improved.
+RAP is an increasingly popular way to help do this, but as this paper points out, RAP can be generalized to \RAPstar\ to help the computational parts of science as well, leading to a virtuous circle.
 
@@
 \item
-To find effective ways to promote professional software engineers being recognized and participating  fully in scientific research, like professional statisticians routinely support quality research.
+To find effective ways to promote professional software engineers being recognized and participating fully in scientific research, just as professional statisticians routinely support quality research (see section \ref{statistics-computation-analogy}).
 \end{enumerate}
 
-While programming seems easy, and is often taken for granted and done casually, programming \emph{well\/} is very difficult \cite{fixit}.
+While programming seems easy and is often taken for granted and done casually, programming \emph{well\/} is very difficult \cite{fixit}. Science needs coding to be done well.
+
 We know from software research that ordinary programming is very buggy and unreliable.
 Without adequately specified and documented code and data, research is not open to scrutiny, let alone proper review, and its quality is suspect.
-Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion: computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not magically make it more reliable \cite{reproducibility,relit,popper-conjectures-refutations}.
-
-Software Engineering Boards, as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve computer-based science.
+Some have argued that availability of code and data ensure research is reproducible, but that is na\"\i ve criterion: computer programs are easy to run and reproduce results, but being able to reproduce something of low quality does not magically make it more reliable \cite{reproducibility,relit,popper-conjectures-refutations} (see section \ref{reproducibility-concerns}).
+
+Software Engineering Boards (SEBs), as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve computer-based science.
+If nothing else, the idea of SEBs is something to criticize and improve.
 
@@
 \ack{The author is very grateful for comments from: \csref Ross Anderson, \csref Nicholas Beale, \csref Ann Blandford, \csref Paul Cairns, \csref Rod Chapman, \csref Jos\'e Corr\'ea de~S\`a, \csref Paul Curzon, \csref Jeremy Gibbons, \csref Richard Harvey, \csref Will Hawkins, \csref Konrad Hinsen, \csref Ben Hocking, \csref Daniel Jackson, \csref Peter Ladkin, \csref Bev Littlewood, \csref Paolo Masci, Stephen Mason, \csref Robert Nachbar, \csref Martin Newby, \csref Patrick Oladimeji, \csref Claudia Pagliari, \csref Simon Robinson, \csref Jonathan Rowanhill, \csref John Rushby, \csref Susan Stepney, \csref Isaac Thimbleby, \csref Prue Thimbleby, \csref Will Thimbleby, \csref Martyn Thomas, and \csref Ben Wilson.
-The author also thanks the anonymous \emph{Computer Journal\/} referees who also contributed to the quality of this paper.}
+The author also thanks anonymous referees who also contributed to the quality of this paper.}
 %\the\csrefcount\ computer science reviewers.

### 2023-05-12 paper-seb-supplementary-material.tex
Revisions for Computer Journal

@@
 A important role of documentation is to cover configuration: how to get code to work --- without configuration, code is generally useless.
-The most basic is a README file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
+The most basic is a \texttt{README} file, which explains how to get going; more useful approaches to configuration include make files, which are programs that do the configuration automatically.
 
@@
 In other words, a journal's data policies and code policies should be the identical --- and the conventionally stricter data policies should also apply to code.
-It is baffling that some journals have data policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
+It is baffling that some journals have code policies that are weaker than their data policies; it is certainly indefensible to have no code policies at all.
 
@@
 The main paper was motivated by selected epidemiological papers and their problematic computational models (which are discussed in the paper).
-Are the issues illustrated by these case studies typical of science more broadly?
-We undertook, then, some selected studies of papers in different fields, then undertook a randomized, stratified pilot survey covering several leading peer reviewed journals.
+Although concerning in their own right, especially for informing national public health policies during a pandemic, the wider question is: are the problems illustrated by these case studies typical of science more broadly?
+We undertook, then, some selected studies of papers in a variety of fields, then undertook a randomized, stratified pilot survey covering several leading peer reviewed journals.
 The point was not to establish the frequency of problems, so much as to sign whether the problematic case study was exceptional or typical.
+It is typical.
 
@@
 
-The pilot study follows the RAP methodology.
+The pilot study itself follows the RAP methodology.
 The data and code for this paper's pilot survey (and all other analysis used in the paper and in this \supplement) are available on GitHub.
@@
 
-The \RAPstar\ approach cannot enforce the truth of such as summary, but it very significantly increases the chances that the summary is correct and up to date.
+The \RAPstar\ approach cannot enforce the truth of such as summary, but it (and the opportunity to read and review it easily) very significantly increases the chances that the summary is correct and up to date.
 For example, if refactoring leads to an option being deleted, then it will also disappear \emph{with no further work\/} from the table above.
 Also, since the summary was proof-read at the same time as proof-reading this \supplement, following \RAPstar\ also increases the chances that any errors or functionality omissions or issues in the makefile have been detected and corrected.
+data.js
 
@@
 Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
-At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.}
+At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``\ldots\ code coming soon~\ldots,'' despite \the\pubdelayinmonths\ months having already elapsed since the  paper had claimed the code could be accessed in its repository.}
 \label{table-repo-summary}\globalTableLabel{supplementaryTableRepoSummary}
@@
 There was only one example of very basic Software Engineering methods, namely independent coding, and even then the independent code used for testing was not uploaded to the paper's code repository, so the independent testing is not available for reviewers or readers of the paper.
-(See \supplement\ for more details.)
 
@@
 A JavaScript program sanity checks the JSON data.
-The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
-Some papers that had had no code available during the first assessment had uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a six month gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data; the sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
-
-Note that since JSON data is JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single file.
+The sanity checks found a few errors (e.g., it checks that if there are comments of any sort then there must be some accessible code in order to have any comments; it checks the DOI is accessible, etc), which led to a productive double-checking of all the facts of the original papers --- and correcting all the errors.
+Some papers that had had no code available during the first assessment subsequently uploaded code by the time of the double-checking.\footnote{{Note that double-checking was performed by the same person as the first assessment, though with the benefit of a long gap to bring a degree of independence}.} A field \texttt{doubleChecked} was added to supplement the original data field \texttt{accessed} to track the process of double-checking the data.
+Further sanity checks then of course checked all \texttt{doubleChecked} fields were completed.
+And so on.
+Note that since the checking was done automatically in code, whenever any data was modified and whenever the paper was typeset, the entire body of checks were easily rerun.
+
+Since JSON data is effectively JavaScript code, it was convenient to combine the data, the data sanity checking, and the analysis all in a single JavaScript file for more convenient maintenance.
 Hence, running the data generates the core human-readable information used in this paper.
@@
 
-so the named value is then available for the author to use when the paper is typeset.
+so the named value (here, \texttt{dataVariableCount}) is then available for the author to use any way they wish when the paper is typeset.
 
@@
 This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2348, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
+For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2\hskip .3em 348, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
 The approach turns a possibly-hard-to-spot \emph{single\/} error into \emph{multiple\/} errors spread throughout the paper into different contexts, thus increasing the chances of noticing the error.
@@
 For the present paper, a potential problem is if the paper is mistakenly typeset \emph{before\/} the latest JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense (e.g., from a debugging run of \texttt{data.js}).
-Variables might be checked as follows:
+Although the following text only shows the result of checking (and not the calculations), in general all generated variables can be sanity checked in \LaTeX\ or in the generating programs like \texttt{data.js}:
 
@@
 \noindent
-\vbox{\noindent Legend:\vskip 1ex
+\begin{center}
+\sf
 \input generated/legend.tex
-}
+\end{center}
 %\newpage

### 2023-05-12 response-to-referees/response 13 april 2023.tex
Revisions for Computer Journal

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                   %\ldots\  or a4paper or a5paper or\ldots\  
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}\usepackage{url}\usepackage{xcolor}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcount\noten
\noten=1
\long\def\note#1{

\noindent\color{red}{\begin{tabular}{@{}r|p{4.7in}}
\hphantom{\hskip -.25in}\ifnum \noten<10 \hphantom{0}\fi{\sf \the\noten}& #1
\end{tabular}}\color{black}
\global\advance\noten by 1}

\def\newnote#1#2{\vskip 1.5ex\noindent 
\textbf{%\ifnum \noten<10 \hphantom{\sf\textbf{0}}\fi
{(\sf\textbf{\the\noten})}} #1
\note{#2}}

\def\nonote#1#2{\vskip 1.5ex\noindent\color{blue}**** #1\color{black}\color{green}#2\color{black}}

\def\mylabel#1{\global\expandafter\edef\csname #1\endcsname{\the\noten}}

\def\myref#1{\csname #1\endcsname}

\title{COMPJ-2022-05-0325: Response to referees}
\author{Harold Thimbleby}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}

\noindent 
\color{black}
%\begin{quote}\emph{I have sent an email requesting a short extension please --- I've finished revising the main paper, but not yet finished the Supplemental Material (it has been Christmas and New Year~\ldots). I should finish early next week. I uploaded these notes on my revisions in case the computer system destroys the old submission if I do not meet the deadline.}\end{quote}

\noindent Thank you for this opportunity to respond to reviewers' comments. The reviewers' comments and queries, which I have really appreciated, have resulted in many helpful improvements to the paper, as detailed below.

I've also reread the entire paper and taken this opportunity to improve the language and clarity throughout.

I have written my responses to the reviewers' comments in red below. If it helps in referring to them in future correspondence, I have numbered reviewers' points and my specific responses.

\color{black}

\vskip .5cm\hrule height 2pt \vskip .5cm

\noindent

\newpage

\section{REVIEWERS' COMMENTS:}

\subsection{Reviewer: 1}

\newnote{I find this paper to be much improved from the initial submission, and it (and the author's comments) have convinced me that it is valuable to publish, at least to spark more discussion of the issues it raises.}{Thanks for this positive feedback!}

However, I still think there are some concerns that should be resolved, as detailed below.

\newnote{page 1, lines 12-16 --- slight changes in text
Scientists rarely ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny, published science will be unreliable, because results --- data, graphs, images, inferences, etc --- generated by the code will themselves be unreliable.}{Suggested corrections made.}

\newnote{page 1, line 17 --- slight changes in text
This paper proposes and justifies\ldots}{Yes! I like the improved order}

\newnote{The idea of a table of contents in the body of the paper seems odd, but I assume this will be taken care of in the journal production process}{I left the table of contents. If it is not required in the Journal style, it's easier to delete than to reinstate.}

\newnote{page 2, column 1, line 40
I don't believe that code is generally incorrect, at least not to point of leading to incorrect results. I accept the code is generally informal and inaccessible, and that code might have bugs, but many bugs don't lead to significantly incorrect results. (The study in Table 2, while valuable, doesn't discuss incorrect code, and is also quite small. (Note that a large-scale study of incorrect code would be valuable.) The studies mentioned in 2.a [15-16] do not study the quality of code itself, but rather reproducibility of results more broadly, including issues related to finding correct versions of dependencies, input files, platforms, etc.; this is a longer discussion than belongs in this review.}{I've re-instated a brief ``further work'' section that mentions these issues for further work. I've also added citations to some papers that discuss automatically finding dependencies etc.}

\newnote{page 4, column 2, lines 11-13 --- slight change in text:
While chemists are trained in reliable methodologies, taking quality glassware for granted is reasonable.}{Thanks; fixed}

\newnote{page 4, column 2, lines 42-44
references [9, 17-20, 20] doesn't make sense with 20 listed twice.}{Oops! Thanks; fixed}

\newnote{page 4, column 2, lines 41-49
I note reviewer 2 mentioned JOSS, which I think should be mentioned here in addition to ReScience, as it specifically does focus on peer review of code, including code quality, at least to some extent, though not to the extent the author calls for.}{I now mention both journals.}

\newnote{page 5, column 1, lines 11-13
I don't think TOP is stricter than FAIR, but rather, has a somewhat different focus. In terms of code quality, it is stricter, but in other aspects (e.g. findability), FAIR is stricter.}{Thanks. I've improved the wording now.}

\newnote{The discussion in 2.b about computable papers is reasonable but limited to the cases where the computing is can be done on purely local resources, or freely available resources. Where more computing is needed (e.g., HPC codes, this idea becomes much more complex very quickly.)  Also, I suggest that Binder be mentioned in the Jupyter bullet.}{Thanks, I've now mentioned HPC, Binder, etc, in a new section on technical debt (6.g) in the newly-restructured section 6.}

\newnote{In 2.c, it's unclear how specific RAP is meant to be. Is this a general description of a process, which can be implemented by different people using different specific tools, or is this a description of a specific set of tools used in a specified manner?}{Table 4 (which I've improved a bit to help address this concern), in particular, makes it clear that RAP is a generic workflow.}

\newnote{Table 5 isn't really a table --- again, the journal should specific what this kind of object should be called.}{I've now made the contents look more like a table! I've also ensured all tables have a consistent design.}

\newnote{page 10, column 2, line 48
Something is wrong with the grammar in ``used comment to inactivate code,''}{Thanks; fixed}

\newnote{Regarding the note about section 3.a on page 11, what happens if the company that sells Mathematic goes out of business? How would this notebook then be usable in the future? This also applied to the example about [33] in 2.b. What if the author had chosen a technology in 1999 that was dropped by its commercial vendor in 2001? How likely is it that there would be a way to run that notebook 22 years later?}{These are important points! I have addressed them in a footnote in the earlier section, 2.b.}

\newnote{page 13, column 1, line 24
Please explain what the Code of Hammurabi has to do with building houses}{Short footnote added to explain}

\newnote{\mylabel{code-data}
page 13, column 2, lines 28
data and code are not equivalent or interchangeable. For example, one is creative whole the other is not. This is like saying numbers and literature are equivalent. While I completely agree that there should be policies on code, and that these policies should be strict, it's not because code and data are the same, but because both are important to research.}{Thanks for this observation. I've rewritten the section, explaining the issues more clearly, and making the argument more evidence-based and hence less provocative.}

\newnote{I feel that some of sections 4.b and 4.c don't really match their titles, nor how they examples are divided. For example, the last paragraph of 4.b is about the status of code in publications, not about the status of coding.}{I've deleted the last paragraph of 4.b.}

\newnote{And much of 4.c is about PRISMA and how papers change over versions, not really about coding.}{I've rewritten 4.c to make it clear that PRISMA is a leading example: the section is not about PRISMA, but about publication policies and standards, of which PRISMA is a leading (and up to date) example.}

\newnote{Supplemental material, page 41, line 49 --- have data policies that are weaker should be have code policies that are weaker.}{Thanks; fixed}

\newnote{In 4.d, saying that code and data are formally equivalent (and see comment above) only makes sense in a limited context (that discussed in the supplementary material), but this is overstated here.}{See point \myref{code-data} and my response to it above, which I think is the same point.}

\newnote{In 4.d, ``bugs naive bugs'' has an extra ``bugs''}{Thanks; fixed}

\newnote{In 4.d, the last paragraph's discussion of intentional bugs is confusing terminology. It would be much cleaner to talk about bugs in translating from theory to model and in translating from models to code. I don't think that either kind is actually intentional.}{I've rewritten this paragraph to make it clearer; it also cites material that provides further discussion of the term used.}

\newnote{page 17, column 1, line 37
this is only surprising in the sense that gambling in Rick's Cafe was shocking}{I think it's surprising that I only say surprising once in the entire \hbox{paper~\ldots}}

\newnote{I am still unhappy with the use of ``repository'' in 5.d to refer to both Dryad and GitHub, since one is an archival system that requires metadata and involves curation, while the other has none of these properties, and is a social coding platform build on top of a version control system.}{Dryad and Github certainly differ in detail, but both describe themselves as repositories --- indeed Dryad started out life as DRIADE, The Digital \textbf{Repository} of Information and Data for Evolution.}

\newnote{The next paragraph talks about using the repository for maintaining the code, which cannot be done in Dryad, but could be done in GitHub.}{I've added a footnote: ``Dryad \url{datadryad.org} curates raw, unprocessed data. At the time of writing, Dryad excludes code; however, it uses a separate organization, Zenodo \url{zenodo.org}, to host code and other relevant information. This arbitrary separation is unfortunate as it increases management problems, increases reproducibility problems, limits using RAP, and most seriously limits how scientists can structure their data and code to best suit their research (see section 4.d).''}

\newnote{Note that the final paragraph of this section talks about differences between code and data, which are correct but seems to make the earlier comments about these being formally equivalent incorrect without more discussion earlier.}{Thanks to your ideas, I improved the earlier discussion, and think this fully addresses this point.}

\newnote{The abrupt change of topic that comes with section 6 is confusing. I would prefer to have seen some general discussion of what the author thinks is needed, with RAP+ then discussed as a way to implement this, rather than it being pitched as The Solution. The discussion in Section 7 seems to disconnect code from the resources used to run the code, and assumes that running code is simple and has no cost. This carries over into section 8 as well. While section 8 is the call to action, this is really just a continuation of the more opinion/vision part of the paper that started in section 6. Some restructuring or changed titles might be useful here. Section 6 could be future vision, with the contents of 6, 7, and 8 being subsections.}{I've rearranged and grouped all these sections under a new main heading ``Rethinking science that uses code.'' (The table of contents on page 2 makes clear what I've done.) I've also done minor rewording to improve the flow to be consistent with the new improved structure. Thanks for this observation.}

\subsection{Reviewer: 2}

The revised manuscript has improved in many ways. In addition to addressing many detailed comments by both reviewers, it is now clearly an opinion article. I still find it a bit lengthy and lacking actionable proposals that could initiate a transition towards the better future described by the author.

Two minor comments:

\newnote{Page 20, line 16: Almost the entire scientific process can be automated\ldots\ If ``scientific process'' means all of science, then this is clearly not true. Specific protocols for data acquisition, data analysis, etc. can be automated, but not all of science! Otherwise\ldots\ I would happily let computers review manuscripts instead of doing it myself!}{good point! thanks; fixed. I've also replaced the word ``process'' (where relevant) with ``workflow'' as this word is more specific, without including ``thinking'' that may be implied by ``process.''}

\newnote{Page 20, line 40: To date, this critical point has been overlooked.}{(Not sure what the point is here --- this is what the paper originally said.)}

\newnote{While this is true for the majority of computational scientists, there is some prior work in this space. See for example Maneage (https://maneage.org/), or the work by Court\`es (https://rescience.github.io/bibliography/Courtes\_2020.html) that extends the Guix package manager. Both approaches are very similar to the RAP* proposal.}{Thanks for these exciting references. I added a citation to the impressive system Maneage, and mentioned that it also includes a good review in its appendix. I've also added a couple of paragraphs about Court\`es's impressive work in the section on RAP\@.}

A few typos I noticed:

\newnote{- Page 14, line 12: bugs nai\"{\i}ve bugs $\Rightarrow$ na\"{\i}ve bugs}{Thanks; fixed}

\newnote{- Page 16, line 46: Ferguson's reported science unlikely to be reliable. (insert ``is'')}{Thanks; fixed}

\newnote{- Page 41, line 49:\ldots\ some journals have data policies that are weaker than their data policies\ldots\ $\Rightarrow$ code policies}{Thanks; fixed}

\newnote{Konrad Hinsen, CNRS, France
(Note to editor: I have decided to sign all my reviews. Please don't remove my signature.)}{Thanks so much Konrad. I've really appreciated our correspondence, your critical checking, and you pushing me to improve the reproducibility of my own work. Thanks.}

\subsection{Reviewer: 3}

\newnote{There are a few remaining typos}{Thanks. I've reviewed the entire paper very carefully, and of course benefitted from all the referees' careful reading and pointing out of typos, all of which I have fixed.}

\nonote{\ldots\ including one sentence where you may have omitted a ``not'', thereby reversing what you intended to say. (I regret that I have been unable to find this again --- please check. I think it was toward the bottom of a paragraph in the left-hand column.)}{}

\newnote{On p16 column 1, the sentence starting ``Alternatively) near line 13 lacks a verb.}{Thanks; fixed}

\newnote{The sentence starting ``Ferguson'' at line 46 omits ``is''.}{Thanks; fixed}

%\color{red}
%\section{SUMMARY OF CHANGES}

\end{document}

### 2023-05-12 programs/data.js

@@ -23,9 +23,19 @@ var descriptionsOfFields = {
         documentation: "Authors",
         essential: 1
     },
+    codeAssertions: {
+        documentation: "Code in repository uses assertions",
+        essential: 0,
+        flag: "$\\sf C_a$"
+    },
     codeComment: {
         documentation: "Evaluator's notes about code",
         essential: 1
+    },
+    codeConditions: {
+        documentation: "Code in repository uses pre- or post-conditions or similar",
+        essential: 0,
+        flag: "$\\sf C_c$"
     },
     codeURL: {
         documentation: "URL (typically GitHub) for code repository"
@@ -90,8 +100,8 @@ var descriptionsOfFields = {
         flag: "$\\sf S_{\\mbox{\\scriptsize NONE}}$"
     },
     hasNoComment: {
-        documentation: "Code has no non-trivial comments",
-        flag: "$\\sf C_0$"
+        documentation: "Code only has trivial or obvious comments",
+        flag: "$\\sf C_1$"
     },
     hasOpenSourceDevelopment: {
         documentation: "Team or open source development",
@@ -105,16 +115,16 @@ var descriptionsOfFields = {
         documentation: "Use of RAP/RAP* principles"
     },
     hasSubstantialcomment: {
-        documentation: "Code has substantial, useful comments and documentation",
-        flag: "$\\sf C_{+}$"
+        documentation: "Code has substantial, useful comments, and documentation",
+        flag: "$\\sf C_3$"
     },
     hasToolBasedDevelopment: {
         documentation: "Evidence of any tool-based development",
         flag: "$\\sf S_{{\\mbox{\\scriptsize tools}}}$"
     },
     hasTrivialComment: {
-        documentation: "Code only has trivial comments (e.g., copyright)",
-        flag: "$\\sf C_1$"
+        documentation: "Code only has comments unrelated to code intent (e.g., copyright)",
+        flag: "$\\sf C_0$"
     },
     journal: {
         documentation: "Journal",
@@ -158,7 +168,7 @@ var data = [{
         number: 95,
         journal: "Nature Digital Medicine",
         doi: "10.1038/s41746-020-0300-0",
-        dataComment: "On request",
+        dataComment: "On request.",
         hasCodeInPrinciple: 1,
         codeComment: "``Code is available upon request from the corresponding author'' (requested)",
         pages: 3
@@ -171,16 +181,17 @@ var data = [{
[…]

### 2023-05-12 paper-seb-main.tex
Made final reference column balance

@@
 
+\indented 2
+``Software Engineering Boards,'' analogous to Ethics or Institutional Review Boards, should be instigated and used.
+
 \indented 3
-``Software Engineering Boards,'' analogous to Ethics or Institutional Review Boards, should be instigated and used.
-
-\indented 4
 The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp.
@@
 
-\bibnewcolumn = 112 % bibentry number of last entry on (final) left hand column
+\bibnewcolumn = 122 % bibentry number of last entry on (final) left hand column
 % Hint: set it to a very large number, so the final column is as unbalanced as it gets
@@
 % depending on the largest bibitem number width in the column. As it is, it's set
-% to the width of 999 for all bibitems even though it takes a while for bibitems to get 
+% to the width of 999 for all bibitems even though it takes a few pages for bibitems to get 
 % up to 3 figures. Sigh

### 2023-05-12 response-to-referees/response 12 May 2023.tex
Made final reference column balance

--- 2023-05-12 response-to-referees/response 13 april 2023.tex
@@
 \newnote{There are a few remaining typos}{Thanks.
-I've reviewed the entire paper very carefully, and of course benefitted from all the referees' careful reading and pointing out of typos, all of which I have fixed.}
+I've reviewed the entire paper very carefully, and of course benefitted from all the referees' careful reading and pointing out of typos, all of which I have fixed --- I trust this covers all the typos you were concerned about.}
 
-\nonote{\ldots\ including one sentence where you may have omitted a ``not'', thereby reversing what you intended to say.
+\newnote{\ldots\ including one sentence where you may have omitted a ``not'', thereby reversing what you intended to say.
 (I regret that I have been unable to find this again --- please check.
-I think it was toward the bottom of a paragraph in the left-hand column.)}{}
+I think it was toward the bottom of a paragraph in the left-hand column.)}{In addition to implementing all the corrections the referees suggested, I have re-read my paper carefully several times, but I have not found this problem.}
 

### 2023-05-12 paper-seb-supplementary-material.tex
Small typos fixed in supplemental material

@@
 There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc.
-(See the \supplement\ for a list of standard methods.)
+No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc (see section \ref{supplementary-best-practice} in this \supplement).
 

### 2023-05-14 generated/over-fitting-code-section.tex
Basic updates

@@
 $\hat{y}(0)=-41.8$,
- even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.
-
-While over-fitting data is a well-known problem, the point for this paper is that \emph{code\/} itself can easily be over-fitted.
-Code can of course be over-fitted in more complex ways than can be illustrated with elementary polynomials, as here.
-Code over-fitting is much harder to recognize because there may be no simple graph plot, like figure \ref{fig-overfit}, to highlight the problems.
-Furthermore, almost all code is far more intricate than the two trivial polynomials used to illustrated figure \ref{fig-overfit}.
-
-Unfortunately, code in published science \emph{is\/} often over-fitted, and over-fitted in a way that is very hard to scrutinize.
-For example, in epidemiology (which is considered in section \ref{section-pandemic-modeling}) it is routine to use very complicated, large dynamical models parameterized with numerous social, cultural, health, demographic and geographical data.
-The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases.
-Indeed, many of the programs in the survey used comments to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countUsesVersionControlRepository\ papers in the pilot survey that reported use of code repositories (covering \input generated/kloc.tex
-\totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
-The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\remembernamevalue{numbernamesexamplefootnote}{\thefootnote}}
-
-Furthermore, scientific support code is rarely documented well enough to know what it should have been doing, which should be answered by a specification.
-With no clear specification and documentation, the code can be arbitrarily hacked to get any convenient results, since no particular specification for it has been defined that it should adhere to.
-Thus we risk doing and promoting substandard science because we --- the scientists and the publication process --- are not managing the unlimited adaptability and complexity of code that science has come to rely~on.
-This is over-fitting of the worst kind --- in conventional over-fitting one can at least hope to see that the fitting is over-parameterized for the data, but in code over-fitting the code and specification are not visible, therefore not adequately scrutinized, and --- worse --- the ``data'' the code over-fits includes the entire conceptual contribution of the paper.
-
-Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
-
-%\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
-\vskip 1ex
-\noindent 
-{\newdimen\boxwidth
-\boxwidth = \columnwidth
-\advance\boxwidth by - 1em
-\begin{tabular}{@{}l@{}}
-\framebox[\columnwidth][r]{\hfill\parbox{\boxwidth}{This section (\ref{over-fit}) of this paper, including figure \ref{fig-overfit}, was generated as a computational paper in a \emph{Mathematica\/} notebook, then exported to \LaTeX\ to include here.
-The \supplement\ provides a more complete explanation, as well as the data and details of the fitted polynomials (also generated automatically).
-This paper's repository includes the notebook.
-
-%\hskip 1em Generated text and data presented elsewhere in this paper were created by a \texttt{make} file running shell scripts and JavaScript code analyzing repositories, JSON survey data, etc.
-Results were exported to files of \LaTeX\ markup and definitions, then imported into this paper and expanded where needed.
-}\hfill}\end{tabular}}

### 2023-05-14 programs/obsolete javascript data fitting models/fittingData.js

@@ -1,176 +1,0 @@
-/*
- Author: Harold Thimbleby, 2022
- 
- Given an array of (x,y) pairs as data, this simple program finds the best fit linear model (using 
- least squares) and an exact polynomial model (a Lagrange polynomial) to fit the data.
- 
- The simple algorithms used are taken straight off Wikipedia:
- 
- 	Least Squares: https://en.wikipedia.org/wiki/Simple_linear_regression
-	Lagrange polynomial: https://en.wikipedia.org/wiki/Lagrange_polynomial
-	
-Both algorithms generate a vector of polynomial coefficients, and the function polynomialToString converts the coefficients to a string representation of the polynomial in x. Thus, for example, LeastSquares(data) generates the numerical vector [a,b] which represents the linear polynomial a+bx.
-
- */
-
-var data = [{
-    x: 2.2,
-    y: 4.9
-}, {
-    x: 4.7,
-    y: 8.2
-}, {
-    x: 2.6,
-    y: 5.7
-}, {
-    x: 1.,
-    y: 2.7
-}, {
-    x: 0.9,
-    y: 1.5
-}];
-
-/* Typical shell command (using node to run the Javascript) and resulting output:
-
-    % node p.js
-    Given data:
-    [
-      { x: 2.2, y: 4.9 },
-      { x: 4.7, y: 8.2 },
-      { x: 2.6, y: 5.7 },
-      { x: 1, y: 2.7 },
-      { x: 0.9, y: 1.5 }
-    ]
-    Model coefficients to 4 significant figures:
-    
-      Linear least squares fit: y =  0.8232 +1.656 x
-      Linear evaluated at 5:  9.1032
-      Linear evaluated at 5 in LaTeX:
-        \def\lineary{$y(5)= 9.1 $}
-    
-      Lagrange polynomial:      y = -32.04 +69.63 x -46.72 x^2 +13.08 x^3 -1.257 x^4
-      Lagrange evaluated at 5:  -2.5149999999999864
-      Lagrange evaluated at 5 in LaTeX:
-        \def\exacty{$y(5)= -2.5 $}
-    
-    
-    ---
-    
-    In Mathematica format:
-    
-    data = {{2.2,4.9},{4.7,8.2},{2.6,5.7},{1,2.7},{0.9,1.5}};    
-    linear[x_] :=  0.8232 +1.656 x;
[…]

### 2023-05-25 paper-seb-main.tex
Sorting out git-prep in makefile
Removed git temp files

@@
 
-Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant.
+Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having that consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant.
 Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
@@
 \newcount \bibnewcolumn 
-
+w
 \bibnewcolumn = 122 % bibentry number of last entry on (final) left hand column

### 2023-05-25 generated/over-fitting-code-section.tex
fixing final bugs

--- 2023-05-12 generated/over-fitting-code-section.tex
@@
 \subsection{The siren call of over-fitting code}\label{over-fit}
-%Science is a methodical process that constructs principles and theories that describe the world, and competitive peer reviewed publication is the reliable record of that.
-We do the science, and we are in the world, so there are challenging meta-theoretical issues as well as practical experimental and resourcing issues.
-Karl Popper is one of many philosophers of science to develop principles and theories about science \cite{popper-three-worlds,popper-conjectures-refutations}.
-
-%Popper was concerned with the conventional sciences, where the world, Nature, is assumed fixed and in principle knowable.
-Writing just six years after Popper, Herbert Simon introduced the ``sciences of the artificial'' \cite{sciences-artificial}, which instead focus on understanding artificial worlds, most typically computer programs.
-Programs effectively create worlds, and the sciences of the artificial seek to understand and improve those worlds.
-Crucially, there is no philosophical or other reason why artificial worlds should be based on principles, let alone be simple or elegant.
-Conventional science's \ae sthetics need not apply.
-
-%Science to date has been inevitably driven by human \ae sthetics: that is, simple, elegant, understandable, beautiful, brief, exciting, intelligible \ldots\ all are human properties that are applied to doing, funding, defining, and selecting ``good'' science.
-
-\noindent
 Poor code undermines scientific quality: poor code can generate plausible and possibly misleading results from \emph{any\/} data, including even fictional science.
-A temptation is that %artificial world \cite{sciences-artificial}, built on
-adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena. 
+A temptation is that %artificial world \cite{sciences-artificial}, built on adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena.
 % wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed!
@@
 Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
-
-%\vfill % the vfill is only needed to put the box at the bottom of a column if that makes it look better
-\vskip 1ex
-\noindent 
-{\newdimen\boxwidth
-\boxwidth = \columnwidth
-\advance\boxwidth by - 1em
-\begin{tabular}{@{}l@{}}
-\framebox[\columnwidth][r]{\hfill\parbox{\boxwidth}{This section (\ref{over-fit}) of this paper, including figure \ref{fig-overfit}, was generated as a computational paper in a \emph{Mathematica\/} notebook, then exported to \LaTeX\ to include here.
-The \supplement\ provides a more complete explanation, as well as the data and details of the fitted polynomials (also generated automatically).
-This paper's repository includes the notebook.
-
-%\hskip 1em Generated text and data presented elsewhere in this paper were created by a \texttt{make} file running shell scripts and JavaScript code analyzing repositories, JSON survey data, etc.
-Results were exported to files of \LaTeX\ markup and definitions, then imported into this paper and expanded where needed.
-}\hfill}\end{tabular}}

### 2023-05-25 paper-seb-main.tex
fixing final bugs

@@
 
-%\color{blue}
 \input generated/over-fitting-code-section.tex
-%\color{black}
 
@@
 \item Is the code valid: does it do what the paper claims?\footnote{Of course, the underlying science may be wrong to, so it is useful to distinguish \emph{internal validity\/} and \emph{external validity}.
-Internal validity occurs if the code does what the paper claims; external validity occurs if the code represents correct science (which the paper may have interpreted incorrectly.}
+Internal validity occurs if the code does what the paper claims; external validity occurs if the code represents correct science which the paper may have interpreted incorrectly.}
 \item Do other scientists, including reviewers and the authors, understand the code?
@@
 
-Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is now opeve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
+Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{It is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
 The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
@@
 \subsection{Action must be interdisciplinary}
-\setcounter{footnote}{1}
 Code is more than a scientific instrument, more than a thermometer or test tube, as code makes, informs and changes decisions; indeed, code can actually \emph{do\/} science.
@@
 
+\label{mentionPLOSandIEEE}
 Journal policies could start to explicitly encourage computationally reproducible science using RAP and \RAPstar\ techniques.
@@
 This paper's \supplement\ summarizes relevant Software Engineering good practice that Software Engineering Boards would draw on, including discussing how and why Software Engineering helps improve code reliability, dependability, and quality.
-
-\makeatletter
-\immediate\write\@auxout{\string\def\string\continueSectionNumbers{\arabic{section}}}
-\makeatother
 
@@
 
-%\nolinenumbers
-
 {\raggedright
@@
 \def\urlprefix{}
+
 \initialiseBibliography{References}{1}{}
@@
 \newcount \bibnewcolumn
-w
+
 \bibnewcolumn = 122 % bibentry number of last entry on (final) left hand column
-% Hint: set it to a very large number, so the final column is as unbalanced as it gets
+% Hint: first set it to a very large number, so the final column is as unbalanced as it gets
 % then, if it's worth fixing the final page, lower the number to some citation on the 
@@
 
-%\label{LastPage}
-%save Latex register    {linenumber}
-%\makeatletter
-%\immediate\write\@auxout{\string\def\string\MainPaperLastLine{\arabic{linenumber}}}
-%\makeatother
+\makeatletter
+\immediate\write\@auxout{\string\def\string\continueSectionNumbers{\arabic{section}}}
+\immediate\write\@auxout{\string\def\string\continueTableNumbers{\arabic{table}}}
+\immediate\write\@auxout{\string\def\string\continueFigureNumbers{\arabic{figure}}}
+\immediate\write\@auxout{\string\def\string\continueFootnoteNumbers{\arabic{footnote}}}
+\immediate\write\@auxout{\string\def\string\continuePageNumbers{\arabic{page}}}
+\makeatother
+
 \end{document}

### 2023-05-25 paper-seb-supplementary-material.tex
fixing final bugs

@@
 
-%\rfoot{\supplement\ \thepage/\SupplementalLastPage}
 \lfoot{\today}
@@
 
-%\makeatletter
-%% saves name-value pairs in both the supplementary material .aux file and the info for the main file
-%\def\remembernamevalue#1#2{%
-%	\immediate\write\@auxout{\@backslashchar gdef\@backslashchar #1{#2}}%
-%	\immediate\write\infofileForMain{\@backslashchar gdef\@backslashchar #1{#2}}%
-%}
-%\makeatother
-
 \input generated/info-for-supplement.tex
@@
 \def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-% writing to the aux file, this file generates labels like
-% \newlabel{supplementary-best-practice}{{2}{5}{Software engineering best practice}{section.2}{}}
-% whereas the main file using IEEEtrans class generates
-% \newlabel{table-summary}{{I}{5}} or \newlabel{jvs-policy}{{V}{4}} etc
 
@@
 
-% start supplementary material's page counters to continue on from the paper's page numbering
-\setcounterpageref{page}{LastPage}
+% start supplementary material's various counters to continue on from the paper's page numbering
+\setcounter{section}{\continueSectionNumbers}
+\setcounter{table}{\continueTableNumbers}
+\setcounter{figure}{\continueFigureNumbers}
+\setcounter{footnote}{\continueFootnoteNumbers}
+\setcounter{page}{\continuePageNumbers}
 \addtocounter{page}{1} % otherwise we start with the same page that the main paper ended on (we want to be on the next page:-)!
-
-%\setcounter{section}{0}
-\setcounter{section}{\continueSectionNumbers}
-\def\n#1{{\sf\bfseries ({#1})}}
-
-\newdimen \www 
-\setbox0=\hbox{\n{viii}~}
-\www=\wd0 \advance \www by .4em
-\def\nnInContents#1#2{\hbox to \www {\hfill\n{#1}~\hskip .4em} #2\\}
-
-\setbox0=\hbox{\sfdefault\large(00)}
-\newdimen\howwide \howwide=\wd0 \howwide=2.5em
-
-\def\sec#1#2#3{\setbox0=\hbox{\n{#1}~~}\leftskip \wd0
-%\begin{flushleft}
-\subsubsection*{\setbox0=\hbox{\n{#1}~~}\hskip -\wd0\copy0 \bfseries
-\immediate\write\minicontentsfile{\bslash nnInContents{#1}{#3}}%
-{#2}\starredsubsectioncontents{\hbox to \howwide{\hfill (#1)} #3}}
-%\end{flushleft}
-\leavevmode
-\vskip 1ex
-\par\leftskip 0in}
-
-\def\highlightsec#1#2#3{\sec{#1}{{#2}}{{#3}}}
-\def\highlightmysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysubsection#1{\subsection{{#1}}\subsectioncontents{{#1}}}
-\def\changemysection#1{\vskip 2ex\section{{#1}}\sectioncontents{{#1}}}
-\def\highlightmysection#1{\section{{#1}}\sectioncontents{{#1}}}
-\def\mysection#1{\section{#1}\sectioncontents{#1}}
-\def\mysubsection#1{\subsection{#1}\subsectioncontents{#1}}
 
@@
 
-\def\sectioncontents#1{\immediate\write\contentsfile{\bslash contents{\arabic{section} #1}{\thepage}}}
-\def\subsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{\alph{subsection}\ #1}{\thepage}}}
-\def\starredsubsectioncontents#1{\immediate\write\contentsfile{\bslash subcontents{#1}{\thepage}}}
-
-\noindent\vbox{% The table of contents
-\def\contents#1#2{\vskip 0mm \noindent{\bfseries #1} \dotfill\ #2\\}
-\def\subcontents#1#2{\hbox to 1em{}{#1} \dotfill\ #2\\}
-\noindent\small
-\input generated/supplementary.toc
-}
-
-\newwrite\contentsfile
-\immediate\openout\contentsfile=generated/supplementary.toc
-
 \newpage
 \begin{center}
-\emph{Pagination, section, and citation numbering continues from the main paper}
+\emph{Section, figure, table, citation, and pagination numbering continue from the main paper}
 \end{center}
-\mysection{Further issues for Software Engineering Boards (SEBs)}
-
-\highlightmysubsection{Brief definition}
+
+\setcounter{tocdepth}{3}
+
+\tableofcontents
+\newpage
+
+\section{Further issues for Software Engineering Boards (SEBs)}
+
 Software Engineering Boards, henceforth SEBs, will be used to help and assure that critical code, including epidemic modeling, is of high standard, to provide assurance for scientific papers, Government public health and other policies, etc, that the code used is of appropriate quality for its intended uses.
@@
 
-\mysubsection{Relationships of SEBs to Ethics Boards}
+\subsection{Relationships of SEBs to Ethics Boards}
 
@@
 
-\changemysubsection{SEBs are necessary but not sufficient}
+\subsection{SEBs are necessary but not sufficient}
 The main paper provides evidence and argues that SEBs (or equivalent) are necessary to help improve the quality of science, specifically science relying, explicitly or implicitly, on tools or methods based in software.
@@
 
-\mysection{Software engineering best practice}
+\section{Software engineering best practice}
 \label{supplementary-best-practice}
-\mysubsection{Introduction and standard references}
+%\subsection{Introduction and standard references}
 
@@
 
-\mysubsection{Essential components of best practice}
-Software Engineering includes the following topics, which are discussed at more length below:
-\vskip 2ex
-
-\vbox{\noindent
-      \input generated/minicontents.tex
-      }
-
-\newwrite\minicontentsfile
-\immediate\openout\minicontentsfile=generated/minicontents.tex
-
-\sec{0}{Without defining requirements, not enough skilled effort will be put into designing and implementing reliable software --- or excess effort will be wasted}{Requirements}
+%\subsection{Essential components of best practice}
+Software Engineering includes the following topics:
+
+\subsection{Requirements}
 It is not always necessary to program well if the code to be produced is for fun, experimenting, or for demonstrations.
@@
 
-\highlightsec{1}{Without formal methods, there is no rigorous and checked specification of a program, so nobody --- including its developers --- will know exactly what it is supposed to do}{Formal methods}
+\subsection{Formal methods}
 
@@
 
-\sec{2}{Without defensive programming, any errors --- in data, code, hardware, or in use --- will go unnoticed and be uncorrected}{Defensive programming} Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
+\subsection{Defensive programming} 
+Defensive programming is based on a range of methods, including error checking, independent calculation (using multiple implementations written by independent programmers), assertions, regression testing, etc.
 Notoriously, what are often unconsciously dismissed as trivial concerns frequently lead to the hardest to diagnose errors, such as buggy handling of ``well-known, trivial'' things like numbers \cite{numerals}.
@@
 
-\highlightsec{3}{Using inappropriate programming languages undermines reliability}{Using dependable programming languages}
+\subsection{Using inappropriate programming languages undermines reliability}
 Many popular languages are popular because they are easy to use, which is not the same as being reliable to use.
@@
 
-\sec{4}{Version control and open source organizes and helps software development}{Open source and version control}
+\subsection{Open source and version control}
 It is appreciated that the models may change and be adapted as new data and insights become available.
@@
 
-\sec{5}{Without professional testing, there is no acceptable evidence that a program works under real conditions}{Rigorous testing}
+\subsection{Rigorous testing}
 In poorly-run software development it is very easy to miss bugs, because the flawed thinking that inserted bugs in the code is going to be the same flawed thinking with the same misconceptions that tries to detect them.
@@
 
-\highlightsec{6}{Without documentation and record keeping, nobody --- least of all the programmer --- knows what code is supposed to do or how to get it to do it}{Good documentation and record keeping}
+\subsection{Good documentation and record keeping}
 Documentation covers internal documentation (how code works), developer (how to include it in other programs), configuration (how to configure and compile the code in different environments), external documentation (how the code is used), and help (documentation available while using the program).
@@
 
-\sec{7}{If code is not usable, even if it is ``correct'' it will not be used and interpreted correctly}{Usability}
+\subsection{Usability}
 \globalLabel{supplementary-material-makefiles}
@@
 
-\sec{8}{Without using existing solutions (libraries, APIs, etc) reinventing code merely reinvents bugs}{Reusing quality solutions}
+\subsection{Reusing quality solutions}
 Reusing quality code (mathematical functions, database operations, user interface features, connectivity, etc) avoids having to develop it oneself, saves time and avoids the risks of introducing new bugs.
@@
 
-\sec{9}{Poor programmers often fix bugs rather than the causes of bugs: complexity and obfuscation}{Simplicity} 
+\subsection{Simplicity} 
 When a program doesn't quite do what is wanted, it is tempting to add more features or variables, or to treat the problem as an ``exception'' and program around it --- which inserts more code and, almost certainly, more bugs.
@@
 
-\highlightsec{10}{International standards have been developed to support critical software development}{Compliance with standards}
+\subsection{Compliance with standards}
 
@@
 
-\sec{11}{Effective multidisciplinary teamwork is essential because no individual has the capacity to develop non-trivial reliable software}{Effective multidisciplinary teamwork}
+\subsection{Effective multidisciplinary teamwork}
 
@@
 
-\highlightsec{12}{Computing technologies are advancing rapidly, and best practice in Software Engineering is continually evolving}{Continuous Professional Development (CPD)}
+\subsection{Continuous Professional Development (CPD)}
 
@@
 
-\highlightsec{13}{Other factors \ldots}{Security and other factors}
+\subsection{Security and other factors}
 Of course, there are many other factors to be considered for the professional development of critical code, such as using appropriate methods to ensure cybersecurity \cite{security-engineering,cyber-cacm}, particularly while also being able to up- and download secure updates.
 
-For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example, one of the surveyed paper's code \cite{example-numerical-error} uses literal numbers at far too high a precision for the chosen language to be able to represent correctly (conformant implementations use IEEE 754 double precision 64-bit floating point).
+For pandemic modeling specifically, understanding the limitations of numerical methods (in particular, how numerical methods are affected by the choice of programming language and style of programming) is critical.\footnote{{For example, code from one of the surveyed paper \cite{example-numerical-error} uses literal numbers at far too high a precision for the chosen language to be able to represent correctly (conformant implementations use IEEE 754 double precision 64-bit floating point).
 Such an error typically has an undefined impact on results, and unfortunately is easy to overlook as the program almost certainly ignores the error when running.
@@
 
-\highlightsec{14}{Human Factors are the foundation of everything}{Software is a human activity}
+\subsection{Software is a human activity}
 
@@
 
-Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic Software Engineering.
+Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge --- in the present case, knowledge of basic Software Engineering.
+
 Dunning and Kruger go on to say,
@@
 
-\mysection{Code, data, and publication}
+\section{Code, data, and publication}
 \globalLabel{on-code-data-publication}
@@
 
-\mysubsection{When magic numbers become magic code}
+\subsection{When magic numbers become magic code}
 Data often controls the flow of code. For example, data summarizing patients may include their gender, but the program processes males and females differently. Then data becomes code.
@@
 
-\mysubsection{When data is code}
+\subsection{When data is code}
 {Many computer programs blur the simplistic code/data distinctions deliberately, to create virtual machines.
@@
 
-\mysubsection{Exploiting code as data for more reliable science}
+\subsection{Exploiting code as data for more reliable science}
 In the present paper, we knowingly built on this blur between data and code, a special case of \RAPstarp.
@@
 
-\mysubsection{When data is text: Exploiting code for reliable publication}
+\subsection{When data is text: Exploiting code for reliable publication}
 
@@
 
-\mysubsection{Data and polynomials used in the paper}
+\subsection{Data and polynomials used in the paper}
 The data and polynomials used in the main paper's section \ref{over-fit}, and illustrated in the paper's figure \ref{fig-overfit}, is presented below, as generated in \LaTeX\ by the same \emph{Mathematica\/} notebook that generated section \ref{over-fit}.
-Of course, in the paper itself, the data was not necessary to make the point, but if there is any need to replicate it or otherwise scrutinize the argument in the paper --- as there would be for more complex arguments in typical papers --- the data used and results are shown here.
-In a similar way, data and computed results could be presented and made available in other scientific papers.
+Of course, in the paper itself, the specific data was not necessary to make the point, but if there is any need to replicate it or otherwise scrutinize the arguments in the paper --- as there would be for more complex arguments in typical papers --- the data used and results are shown below.
+
+The table below is a \LaTeX\ table generated by \emph{Mathematica}, and is exactly the data used in the paper.\footnote{Barring coding or other errors of course, which here we checked against manually by comparing this table typeset in \LaTeX\ against the raw data in the original \emph{Mathematica\/} data table, but in general might better be done by an automatic round trip --- though that would not easily spot \LaTeX\ errors.} In a similar way, data and computed results could be presented and made available in other scientific papers.
 
@@
 
-\highlightmysubsection{Comparing conventional and RAP approaches}
+\subsection{Comparing conventional and RAP approaches}
 \globalLabel{RAP-diagrams}
@@
 
-\changemysection{The Spiegelhalter trustworthiness questions}
+\section{The Spiegelhalter trustworthiness questions}
 \globalLabel{supplementary-Spiegelhalter-section}
@@
 
-\def\questionsection#1{\changemysubsection{#1}}
 \def\question#1{\item \emph{#1\/}}
@@
 
-\questionsection{How trustworthy are the numbers?}
+\subsection{How trustworthy are the numbers?}
 \newcounter{resumeCounter}
@@
 \end{enumerate}
-\questionsection{How trustworthy is the source?}
+
+\subsection{How trustworthy is the source?}
 \begin{enumerate}
@@
 \end{enumerate}
-\questionsection{How trustworthy is the interpretation?}
+
+\subsection{How trustworthy is the interpretation?}
 \begin{enumerate}
@@
 
-\highlightmysection{A pilot survey of computational science}
+\section{A pilot survey of computational science}
 \globalLabel{survey-section}
@@
 Also, since the summary was proof-read at the same time as proof-reading this \supplement, following \RAPstar\ also increases the chances that any errors or functionality omissions or issues in the makefile have been detected and corrected.
-data.js
-
-\highlightmysubsection{Selected journal case studies}
+
+\subsection{Selected journal case studies}
+There are numerous case studies to be made from journals and their relation to code and data. Here, we select just two.
+
+\subsubsection{\emph{The Lancet\/}}
 
@@
 
+\subsubsection{\emph{Journal of Vascular Surgery}}\label{vascular-surgery-note}
 In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms.
@@
 
-\highlightmysubsection{Pilot paper sample}
+\subsection{Pilot paper sample}
 A sample of \plural{\dataN}{recent paper} covering a broad range of science were sampled from the leading journals \journalBreakdown.
@@
 %
-\caption{Overview of the peer-reviewed paper sample.}
+\caption{Overview of the peer-reviewed paper sample. (Convenient copy of main paper's table \ref{table-overview}.)}
 \label{supplement-table-overview}
@@
 
-\caption{Summary of survey results.}
+\caption{Summary of survey results. (Convenient copy of main paper's table \ref{table-summary}.)}
 \label{supplement-table-summary}
@@
 
-\highlightmysubsection{Summary of results}
+\subsection{Summary of results}
 \label{summary-of-results}
@@
 
-\highlightmysubsection{Current code policies of sampled journals}
+\subsubsection{Current code policies of sampled journals}
 \globalLabel{supplementary-journal-policies-section}
@@
 It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{detecting-against-error} and \ref{restrictive-policies}.
-In particular, for all the papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
-
-Note that journal policies were first accessed on 29 July 2020, close after the period covered by the pilot survey, thus presumably fairly closely reflecting the policies in use for the papers in the survey on the dates when they were each submitted.
+The main paper, section \ref{mentionPLOSandIEEE}, additionally mentions \emph{PLOS ONE\/} and \emph{IEEE Transactions on Software Engineering}.
+
+For all the surveyed papers that had accessible code, the code included explicit (and relevant) data that was not archived \emph{as\/} data in the journal repositories.
+
+Note that journal policies relevant to the survey were first accessed on 29 July 2020, close after the period covered by the pilot survey, thus presumably fairly closely reflecting the policies in use for the papers in the survey on the dates when they were each submitted.
 Unfortunately for the survey, the journals do not make clear what exact policies were applied to each paper when the papers were submitted (on the other hand, the survey shows that policies are not rigorously enforced).
@@
 \tabwidth=\textwidth
-\advance \tabwidth by -4em
-\def\specifyurl{\vskip 3mm
-\begingroup\raggedleft\scriptsize
-} 
-\def\specifyurletc#1{\endgroup\\
-\hbox{ }\hfill\begin{tabular}{|p{\tabwidth}@{}}
-#1
+\advance \tabwidth by -1em
+\def\specifyurl{\vskip 0mm\noindent\begingroup\small} 
+\def\specifyaccess#1{\endgroup\par\noindent
+\begin{tabular}{|p{\tabwidth}@{}}#1
 \end{tabular}\vskip .3cm
@@
 
-{\sf
 \noindent\textbf{Extract from \emph{Royal Society Open Science\/} author guidelines} \\ 
---- It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
+{\sf``It is a condition of publication that authors make the primary data, materials (such as statistical tools, protocols, software) and code publicly available.
 These must be provided at the point of submission for our Editors and reviewers for peer-review, and then made publicly available at acceptance.
@@
 Data and code should be deposited in a form that will allow maximum reuse.
-As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.}
-
-\specifyurl
-\url{royalsociety.org/journals/authors/author-guidelines/#data}
-\specifyurletc{Since first accessed 29 July 2020, the policy has been revised (undated, accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
+As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.''}\specifyurl\url{royalsociety.org/journals/authors/author-guidelines/#data}\specifyaccess{Since first accessed 29 July 2020, the policy has been revised (undated, accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
 The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
@@
 \noindent\textbf{Extract from \emph{Nature Digital Medicine\/} author guidelines}\\
-{\sf --- A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
-[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.}
-
-\specifyurl
-\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}
-\specifyurletc{Accessed 29 July 2020; since updated (accessed 2 February 2022) to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
+{\sf``A condition of publication in a Nature Research journal is that authors are required to make materials, data, code, and associated protocols promptly available to readers without undue qualifications.
+[\ldots] A condition of publication in a Nature Research journal is that authors are required to make unique materials promptly available to others without undue qualifications.''}\specifyurl\url{www.nature.com/nature-research/editorial-policies/reporting-standards#availability-of-data}\specifyaccess{Accessed 29 July 2020; since updated (accessed 2 February 2022) to require [in part] ``Upon publication, Nature Portfolio journals consider it best practice to release custom computer code in a way that allows readers to repeat the published results.
 Code should be deposited in a DOI-minting repository such as Zenodo, Gigantum or Code Ocean and cited in the reference list following the guidelines described here.''
@@
 \noindent\textbf{\emph{Lancet Digital Health\/} author guidelines}\\
-Journal has detailed data policies, but no code policy.
-
-\specifyurl
-\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}
-\specifyurletc{Accessed 29 July 2020. Still no code policy when accessed 2 February 2022.}
+Journal has detailed data policies, but no code policy.\specifyurl\url{marlin-prod.literatumonline.com/pb-assets/Lancet/authors/tldh-info-for-authors.pdf}\specifyaccess{Accessed 29 July 2020.
+Still no code policy when accessed 2 February 2022.}
 
@@
 The \emph{Journal of Vascular Surgery\/} has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed, the following statement on data policies is relevant:
-
-\vskip 3mm \noindent {\sf --- The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
-[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.}
-\specifyurl
-\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}
-\specifyurletc{Accessed 29 July 2020. Policy unchanged when accessed 2 February 2022.}
+While no \emph{Journal of Vascular Surgery\/} papers were surveyed (but see section \ref{vascular-surgery-note}), the following statement on data policies is relevant:
+
+\vskip 3mm \noindent {``The authors are required to produce the data on which the manuscript is based for examination by the Editors or their assignees, should they request it.
+[\ldots] The authors should consider including a footnote in the manuscript indicating their willingness to make the original data available to other investigators through electronic media to permit alternative analysis and/or inclusion in a meta-analysis.''}\specifyurl\url{www.editorialmanager.com/jvs/account/JVS_Instructions%20for%20Authors2020.pdf}\specifyaccess{Accessed 29 July 2020.
+Policy unchanged when accessed 2 February 2022.}
+
+\subsubsection{Sample assessment and scoring}
+
+\def\flagStyle#1{\textcolor{blue}{\sf #1}}
+
+\noindent
+Assessment flags are \flagStyle{highlighted in color} to be clearer in the following tables.\\
+
+\noindent
+\begin{center}
+\sf
+\input generated/legend.tex
+\end{center}
+%\newpage
+
+{
+\def\citenum#1{\cite{#1}}
+\begin{longtable}{@{}cp{2.8in}p{2.8in}@{}}
+\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
+\input generated/assessments.tex
+\end{longtable}
+}
 
 \label{data-methods-section}
-\highlightmysubsection{Assessment criteria and methods}
+\subsection{Assessment criteria and methods}
 \label{data-methods-subsection}
@@
 
-% This commented-out sentiment is repeated later...
-% Finally, because of their approach to automatic typesetting, some journals require that no files are explicitly \texttt{input} into \LaTeX, so a simple program is provided in the GitHub repository to recursively expand all \texttt{input} commands before submission.
-If the source files are available for download after this paper is accepted, they will therefore contain no \texttt{input} commands, but they will contain comments at the appropriate points in the expanded source code explaining the sources of the \texttt{input} data.
-It is curious that the journals cannot do this for themselves.
-
-\highlightmysubsection{Detecting and defending against error}
+\subsubsection{Detecting and defending against error}
 \label{detecting-against-error}
@@
 For the present paper, a potential problem is if the paper is mistakenly typeset \emph{before\/} the latest JSON data has been analyzed; in which case, none of the variables, like \texttt{\bslash countUsesVersionControlRepository}, will have been correctly set and their values could be undefined or nonsense (e.g., from a debugging run of \texttt{data.js}).
-Although the following text only shows the result of checking (and not the calculations), in general all generated variables can be sanity checked in \LaTeX\ or in the generating programs like \texttt{data.js}:
+
+Although the following text only shows the automatic result of checking final values (and not the calculations that led to them), in general all generated variables can easily be sanity checked in \LaTeX\ or in the generating programs like the present paper's \texttt{data.js} code:
 
@@
 		\ifnum \countUsesVersionControlRepository<\min
-			\else
-								\textbf{This is \emph{automatic\/} confirmation that $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore falls within pre-defined sanity limits set for this paper.}
+			\else% it is not too big and not too small, so confirm it's OK...
+					\vskip 4mm\begin{tabular}{|p{6in}}
+					\sf\textbf{This is \emph{automatic\/} confirmation that $ \mbox{\tt countUsesVersionControlRepository} = {\the\countUsesVersionControlRepository} $ so $$\the\min \leq \mbox{\tt countUsesVersionControlRepository} \leq \the\max$$ and therefore {\tt countUsesVersionControlRepository} falls within the pre-defined sanity limits set for this paper.}\end{tabular}\vskip 4mm
 		\fi
@@
 
-The corresponding error messages would not normally be printed in a paper like this --- they would normally be reported by stopping a \LaTeX\ run, that is before the paper can be distributed and cause confusion.
-Note that failing a sanity check indicates a problem that needs to be fixed, but passing a sanity check does not prove a paper correct, but the more sanity checks that are passed (and the harsher the checks) the more confidence we can have that the data has been processed correctly.
+The corresponding error (or success) messages would not normally be printed in a paper like this --- they would normally be reported before a \LaTeX\ run, that is before the paper can be distributed and potentially cause confusion.
+Note that failing a sanity check indicates a problem that needs to be fixed, but passing a sanity check does not prove a paper correct, but the more sanity checks that are passed (and the harsher those checks) the more confidence we can have that the data has been processed correctly.
 Of course, when formal methods are employed in the software development process, the confidence in correctness can be very high.
 
-\highlightmysubsection{Defending against system problems}
+\subsubsection{Defending against system problems}
 Code can become obsolete as programming languages develop and compilers are improved.
@@
 
-\highlightmysubsection{Problems of restrictive journal policies}
+\subsubsection{Problems of restrictive journal policies}
 \globalLabel{restrictive-policies}
@@
 Such restrictive publishing policies undermine reproducibility.
-
-\highlightmysubsection{Sample assessment and scoring}
-
-\def\flagStyle#1{\textcolor{blue}{\sf #1}}
-
-\noindent
-Assessment flags are \flagStyle{highlighted in color} to be clearer in the following tables.\\
-
-\noindent
-\begin{center}
-\sf
-\input generated/legend.tex
-\end{center}
-%\newpage
-
-{
-\def\citenum#1{\cite{#1}}
-\begin{longtable}{@{}cp{2.8in}p{2.8in}@{}}
-\sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated/assessments.tex
-\end{longtable}
-}
 
@@
 \renewenvironment{thebibliography}[1]{%
-	\section*{\uppercase{\refname}}
+	\section{\unnumberedrefname}
     \startBibliography
-    \sectioncontents{\unnumberedrefname}
     \raggedright
@@
 \immediate\closeout\infofileForMain
-\immediate\closeout\contentsfile
-\immediate\closeout\minicontentsfile
 \immediate\closeout\infofileForSupplement

### 2023-05-25 generated/make-help.tex
Revisions for Computer Journal
Fixing makefile and deleteing obsolete files on Git

@@
 % 7 - make expand
-% 8 - make help
-% 9 - make help-brief
-% 10 - make mathematica
-% 11 - make mathematica-open
-% 12 - make one-file
-% 13 - make pdf
-% 14 - make push
-% 15 - make readme
-% 16 - make really-tidyup
-% 17 - make tidyup
-% 18 - make zip
-% 19 - make zip-data
+% 8 - make git-prep
+% 9 - make help
+% 10 - make help-brief
+% 11 - make mathematica
+% 12 - make mathematica-open
+% 13 - make one-file
+% 14 - make pdf
+% 15 - make push
+% 16 - make readme
+% 17 - make really-tidyup
+% 18 - make tidyup
+% 19 - make zip
+% 20 - make zip-data
 
@@
 
-\texttt{make help}&
-      Explain how to use \texttt{make}, and list all available options for using \texttt{make}.\\
+\texttt{make git-prep}&
+      What's on Git that we've lost, or stuff we have got locally but probably don't want on Git, so you can delete it or move it out the way or whatever.\\
    \multicolumn{1}{l@{\vdots}}{}&\\
 
-\texttt{make tidyup}&
-      Tidyup before doing a Git commit.
-Remove all easily generated files, and the large Git repositories needed for the pilot survey.
-Do not remove the main PDFs, or the \LaTeX\ data include files.
-Do not remove the .aux files, as \LaTeX\ runs much more smoothly with them.\\
+\texttt{make really-tidyup}&
+      More thorough than \texttt{make tidyup} --- remove \emph{all} files that can be recreated.
+(This will mean next time you run \LaTeX\ you will have to ignore errors as the .aux files are re-created.)\\
 \end{tabular}}

### 2023-05-25 generated/over-fitting-code-section.tex
Fixing makefile and deleteing obsolete files on Git

@@
 
-Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant example how pointing out the problem of over-fitting has improved science.
+Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant paper pointing out how the problems of over-fitting has improved science.

### 2023-05-25 paper-seb-main.tex
Fixing makefile and deleteing obsolete files on Git

@@
 Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with computable paper calculations. Here, as a simple example, we calculate that
-$10!$ is equal to \factorial{10} just by writing the \LaTeX\ code \verb|\factorial{10}| \emph{in\/} this paper to calculate arbitrary factorials and group the result into conventional blocks of three digits by inserting small spaces.
+$10!$ is equal to \factorial{10} just by writing the \LaTeX\ code \verb|\factorial{10}| \emph{in\/} this paper, which can calculate arbitrary factorials and group the result into conventional blocks of three digits by inserting small spaces.
 Note that although \texttt{factorial} is not a standard \LaTeX\ function, it was readily defined by code also \emph{in\/} the present paper, so it can be easily modified by the author for other purposes as required.
@@
 Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time.
-Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done (more realistically) in this paper's footnote~\numbernamesexamplefootnote.
+Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote~\numbernamesexamplefootnote).
 
@@
 
-\bibnewcolumn = 122 % bibentry number of last entry on (final) left hand column
+\bibnewcolumn = 123 % bibentry number of last entry on (final) left hand column
 % Hint: first set it to a very large number, so the final column is as unbalanced as it gets

### 2023-05-27 generated/assessments.tex
Lots of simplifications to using multiple Latex files

@@
-\citenum{ref-1} & On request.\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations.''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
+\cite{ref-1} & On request.\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations.''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
 Documented overview, but only trivial comments.
 Assertions in code only relate to UI.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}$\sf C_a$\hskip 3pt{}}\\
-\citenum{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available.''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available.''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors).\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request.''\flagStyle{ } & Poor commenting, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team.''\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
+\cite{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available.''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available.''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors).\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request.''\flagStyle{ } & Poor commenting, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team.''\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
 However, the data are available from the corresponding author on reasonable request.''\flagStyle{ } & Empty GitHub repository: ``Code coming soon \ldots'' paper says.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
+\cite{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
 Data from UCSF are not available at this time as they have not been legally certified as being De-Identified, however, this process is underway and the data may be available by the time of publication by contacting the authors.
 Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply.''\flagStyle{ } & Basic documentation, very little comment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request.''\flagStyle{ } & Trivial comments, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available.''\flagStyle{ } & ``We used only free and open-source software'' --- some of which is unspecified.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments only.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{paper-usesRMarkdown} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
-\citenum{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
+\cite{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request.''\flagStyle{ } & Trivial comments, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available.''\flagStyle{ } & ``We used only free and open-source software'' --- some of which is unspecified.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments only.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{paper-usesRMarkdown} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\cite{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
 Currently (apparently) for private view.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
@@
 Essentially identical results were obtained.'' --- but the contrasting method is not available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-16} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-17} & On request.\flagStyle{ } & R, lightly commented.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ethics-paper} & No data required.\flagStyle{ } & Unrunnable incomplete code fragment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-19} & Data embedded in PDF.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-20} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
-\citenum{ref-21} & Partial data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including model's manual.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
-\citenum{ref-22} & No data required.\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-23} & Data cited, not all available.\flagStyle{ } & Trivial documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{example-numerical-error} & On Figshare.\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
+\cite{ref-16} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\cite{ref-17} & On request.\flagStyle{ } & R, lightly commented.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ethics-paper} & No data required.\flagStyle{ } & Unrunnable incomplete code fragment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-19} & Data embedded in PDF.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-20} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\cite{ref-21} & Partial data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including model's manual.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\cite{ref-22} & No data required.\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-23} & Data cited, not all available.\flagStyle{ } & Trivial documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{example-numerical-error} & On Figshare.\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
 Helpful features to make usable for third parties.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-25} & Data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-26} & Data on various web sites.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-27} & Data on request.\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-28} & Data on request.\flagStyle{ } & Python scripts can be requested.\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
-\citenum{ref-29} & Unspecified location on large website requiring registration.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
-\citenum{ref-30} & Available to researchers who meet criteria for access to confidential data.\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
-\citenum{ref-31} & Data access conditional on approved study proposal.\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
-\citenum{ref-32} & Unspecified locations on several large websites.\flagStyle{ } & Python used and apparently on GitHub, but --- an oversight?
+\cite{ref-25} & Data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-26} & Data on various web sites.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-27} & Data on request.\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-28} & Data on request.\flagStyle{ } & Python scripts can be requested.\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
+\cite{ref-29} & Unspecified location on large website requiring registration.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-30} & Available to researchers who meet criteria for access to confidential data.\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-31} & Data access conditional on approved study proposal.\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\cite{ref-32} & Unspecified locations on several large websites.\flagStyle{ } & Python used and apparently on GitHub, but --- an oversight?
 --- no code is available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\

### 2023-05-27 generated/over-fitting-code-section.tex
Lots of simplifications to using multiple Latex files

@@
 \subsection{The siren call of over-fitting code}\label{over-fit}
-Poor code undermines scientific quality: poor code can generate plausible and possibly misleading results from \emph{any\/} data, including even fictional science.
-A temptation is that %artificial world \cite{sciences-artificial}, built on adjusting the code's results gets more attention than the theoretical basis of the code's overall faithfulness to the experimental phenomena.
+Poor code can generate plausible and possibly misleading results from \emph{any\/} data or theory, including fraudulent science.
+A temptation is that developing code to get ``good results'' becomes more important than the code's overall faithfulness to the real scientific phenomena, theoretical or empirical.
 % wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed!
@@
 
-In modeling terms, superficially successful computer programs tend to \emph{over-fit\/} phenomena \cite{over-fit}.
-Instead of using code to test the models, we tinker with the code adapting it to be closer and closer to our prejudices.
-The code then apparently confirms our science, since we fitted it to our understanding.
+In conventional modeling terms, successful computer programs are often \emph{over-fitted\/} to phenomena \cite{over-fit}.
+That is, instead of using code to rigorously challenge, test, and develop our models, we tinker with code adapting it to generate results closer and closer to our prejudices.
+The code then apparently confirms our science, since we fitted it to our preconceptions but not to the science.
 
@@
 \totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
-The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\remembernamevalue{numbernamesexamplefootnote}{\thefootnote}}
+The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\label{name-example-footnote}}
 

### 2023-05-27 generated/saved-files.tex
Lots of simplifications to using multiple Latex files

@@
 summary-table.tex & Short summary table\\
-supplementary-references.tex & Reference list for Supplementary Material\\
+supplementary-references.bbl & Reference list for Supplementary Material (bbl file)\\

### 2023-05-27 paper-seb-main.tex
Lots of simplifications to using multiple Latex files

@@
 \input generated/page-lengths.tex
+
+\makeatletter
+      \input paper-seb-supplementary-material.aux 
+\makeatother
 
@@
 \noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors.
-The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}\vskip 1ex
+The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplement-table-repo-summary}} in the \supplement).}}\vskip 1ex
 
@@
 Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time.
-Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote~\numbernamesexamplefootnote).
+Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote \ref{name-example-footnote}).
 
@@
 The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
-Additional schematics are provided in section \ref{RAP-diagrams} in the \supplement.}
+Additional schematics are provided in section \ref{supplement-RAP-diagrams} in the \supplement.}
 \label{fig-pipeline}
@@
 
-Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Spiegelhalter-section}).
+Spiegelhalter \cite{Spiegelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplement-Spiegelhalter-section}).
 Results should not be uncritically accepted just because they are claimed.
@@
 In reality, if code is inadequate, any so-called ``insights'' will be potentially flawed, and flawed in unknown ways.
-Indeed, none of the papers sampled (described in \supplement\ section~\ref{survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
+Indeed, none of the papers sampled (\supplement\ section~\ref{supplement-survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
 
@@
 
-%An example is the documentation at the code repository \cite{BayesWaveCode}, which requires three high-profile \emph{{Physics Review D}} papers \cite{BayesWave1,BayesWave2,BayesWave3} be cited to acknowledge any use, but only one of them, \cite{BayesWave3}, mentions the code, and even then it fails to acknowledge the contributions of its 10 programmers (despite 6 being co-authors).
+%An example is the documentation at the code repository \xcite{BayesWaveCode}, which requires three high-profile \emph{{Physics Review D}} papers \xcite{BayesWave1,BayesWave2,BayesWave3} be cited to acknowledge any use, but only one of them, \xcite{BayesWave3}, mentions the code, and even then it fails to acknowledge the contributions of its 10 programmers (despite 6 being co-authors).
 The paper does cite the repository URL, but unlike the papers it cites conventionally, it omits all the code's authors.
@@
 Program code contains data.
-Almost all code contains ``magic numbers'' --- that is, data masquerading as code (see \supplement, section \ref{on-code-data-publication}).
+Almost all code contains ``magic numbers'' --- that is, data masquerading as code (see \supplement, section \ref{supplement-on-code-data-publication}).
 This common practice ensures that published data is rarely all of the relevant data because it omits the magic numbers embedded in the code.
@@
 
-Although convenience and convention treat data and code differently, ultimately, data and code are formally equivalent (see \supplement, section \ref{on-code-data-publication}) in the important sense that there is no pre-determined boundary between them or formal criteria to distinguish them: different scientific projects can draw their boundaries differently and as they see best fits their work.
+Although convenience and convention treat data and code differently, ultimately, data and code are formally equivalent (see \supplement, section \ref{supplement-on-code-data-publication}) in the important sense that there is no pre-determined boundary between them or formal criteria to distinguish them: different scientific projects can draw their boundaries differently and as they see best fits their work.
 Indeed, all of the papers reviewed here where code was available include critical data in their code.
@@
 That is, the research's methodology itself may be a mixture of data and code.
-As this paper's \supplement\ shows in section \ref{restrictive-policies}, many journals (e.g., \emph{PLOS ONE\/} and, ironically, \emph{IEEE Transactions on Software Engineering\/}) and repositories have policies that make RAP much harder or just counter-productive at the last step.
+As this paper's \supplement\ shows in section \ref{supplement-restrictive-policies}, many journals (e.g., \emph{PLOS ONE\/} and, ironically, \emph{IEEE Transactions on Software Engineering\/}) and repositories have policies that make RAP much harder or just counter-productive at the last step.
 
@@
 
-Compare the discussion here with figure \ref{excelFigure}, in a series of figures in the \supplement, which illustrates the problem as it presents in many scientific papers.
+Compare the discussion here with figure \ref{supplement-fig:excel}, in a series of figures in the \supplement, which illustrates the problem as it presents in many scientific papers.
 
@@
 
-\dataaccess{There is an extended discussion of the methodology of this paper in the \supplement, section \ref{on-code-data-publication}.
+\dataaccess{There is an extended discussion of the methodology of this paper in the \supplement, section \ref{supplement-on-code-data-publication}.
 The \supplement\ also presents all raw data in tabular form.

### 2023-05-27 paper-seb-supplementary-material.tex
Lots of simplifications to using multiple Latex files

@@
 
+% pageref gives us a hyperlink which looks unsightly, so we do it this way...
 \def\href#1{#1}
@@
 
-% pageref gives us a hyperlink which looks unsightly, so we do it this way...
 \makeatletter
@@
 
-\makeatletter
-\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalLabel: #1}%
-\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalTableLabel: #1}%
-\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-% as well as in the current document....
-\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
-\typeout{Global globalLabelValue: #1 := #2}%
-\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
-}
-\makeatother
+%\makeatletter
+%\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
+%\typeout{Global globalLabel: #1}%
+%\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
+%% as well as in the current document....
+%\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
+%}
+%\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
+%\typeout{Global globalTableLabel: #1}%
+%\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
+%% as well as in the current document....
+%\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
+%}
+%\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
+%\typeout{Global globalLabelValue: #1 := #2}%
+%\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
+%}
+%\makeatother
 
@@
 \section{Software engineering best practice}
-\label{supplementary-best-practice}
-%\subsection{Introduction and standard references}
+\label{supplement-best-practice}
 
@@
 \subsection{Usability}
-\globalLabel{supplementary-material-makefiles}
+\label{supplement-material-makefiles}
 Usability is an important consideration: \cite{hci1,hci2} is the program usable by its intended users so they can obtain correct results?
@@
 \section{Code, data, and publication}
-\globalLabel{on-code-data-publication}
+\label{supplement-on-code-data-publication}
 
@@
 
-More of such techniques for improving reliability are discussed in section \ref{data-methods-subsection}, where they are applied to accurately and reliably reporting the survey reported in the main paper.
+More of such techniques for improving reliability are discussed in section \ref{supplement-data-methods-subsection}, where they are applied to accurately and reliably reporting the survey reported in the main paper.
 
@@
 \subsection{Comparing conventional and RAP approaches}
-\globalLabel{RAP-diagrams}
-The similarities and differences between the conventional copy-and-paste approach to filling in data and diagrams in publications, the improved systematic RAP process, and using notebooks (such as \emph{Mathematica\/} or Jupyter), are illustrated in the sequence of schematics of figures \ref{fig:basic}, \ref{fig:excel}, \ref{fig:rap}, and \ref{fig:notebook}.
+\label{supplement-RAP-diagrams}
+The similarities and differences between the conventional copy-and-paste approach to filling in data and diagrams in publications, the improved systematic RAP process, and using notebooks (such as \emph{Mathematica\/} or Jupyter), are illustrated in the sequence of schematics of figures \ref{supplement-fig:basic}, \ref{supplement-fig:excel}, \ref{supplement-fig:rap}, and \ref{supplement-fig:notebook}.
 
@@
 Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code.}
-   \label{fig:basic}
+   \label{supplement-fig:basic}
 \end{figure}
@@
 }
-   \globalLabelValue{excelFigure}{\thefigure}
-   \label{fig:excel}
+   \label{supplement-fig:excel}
 \end{figure}
@@
    \includegraphics[width=5in]{generated/import.\imageFileSuffix} 
-   \caption{Improving over the normal approach (figure \ref{fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+   \caption{Improving over the normal approach (figure \ref{supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
 In the present paper, this was done by generating text files of \LaTeX\ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX\ names for all of the code-generated values.
@@
 Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen.}
-   \label{fig:rap}
+   \label{supplement-fig:rap}
 \end{figure}
@@
 Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc).}
-   \label{fig:notebook}
+   \label{supplement-fig:notebook}
 \end{figure}
@@
 \section{The Spiegelhalter trustworthiness questions}
-\globalLabel{supplementary-Spiegelhalter-section}
+\label{supplement-Spiegelhalter-section}
 
@@
 \sequestion{How rigorously has the Software Engineering been done?
-Section \ref{supplementary-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
+Section \ref{supplement-best-practice} in the \supplement\ provides a list of important issues that must be addressed for any reliable software.}
 
@@
 Are there discussions of invariants, independent checks for errors, and so on?
-Again, \supplement\ section \ref{supplementary-best-practice} provides further discussion of such issues.}
+Again, \supplement\ section \ref{supplement-best-practice} provides further discussion of such issues.}
 
@@
 The Software Engineering analogy is: are the claims backed up by a sufficiently detailed discussion of the algorithms and Software Engineering that justify the appropriateness of the chosen software implementation?
-The \supplement\ list in section \ref{supplementary-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
+The \supplement\ list in section \ref{supplement-best-practice} provides examples of expected explanations for the trustworthiness of running some code.}
 
@@
 \section{A pilot survey of computational science}
-\globalLabel{survey-section}
+\label{supplement-survey-section}
 
@@
 Typically, research offers new theories or results, points out previous errors, or uses statistical or qualitative methods to reduce the variability in previously reported results.
-In any case, the intention is to provide reliable knowledge and to bolster consensus \cite{ziman}.
+In any case, the intention is to provide reliable knowledge and to bolster consensus \xcite{ziman}.
 
@@
 For example, as standard practice, a Unix makefile is used to make it easy to analyze and generate all data, typeset the paper, and more.
-Table \ref{table:options} shows the options provided.
+Table \ref{supplement-table:options} shows the options provided.
 
@@
 }
-\label{table:options}
+\label{supplement-table:options}
 \end{table}
@@
 
-\subsubsection{\emph{Journal of Vascular Surgery}}\label{vascular-surgery-note}
+\subsubsection{\emph{Journal of Vascular Surgery}}\label{supplement-vascular-surgery-note}
 In 2020 the \emph{Journal of Vascular Surgery\/} published a research paper \cite{jvs1}, which had to be retracted on ethical grounds \cite{jvs2,jvs3}: it was a na\"\i ve study and the editorial process was unaware of digital norms.
 Notably, the paper fails to provide access to its anonymized data (with or without qualification), and fails to define the data anonymization algorithm, and also fails to even mention the code that it developed and used to perform its study.
-The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see  section \ref{supplementary-journal-policies-section}).
+The journal's data policy is itself very weak (the authors ``should consider'' including a footnote to offer limited access to the data) and, despite basic statistics policies, it has no policy at all for code (see section \ref{supplement-journal-policies-section}).
 Ironically, the retracted article \cite{jvs1} is still online (as of August 2020) with no reference to any editorial statement to the effect that it has been retracted, despite this being trivial --- and necessary --- to achieve in the widely-accessed online medium.
@@
 At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``\ldots\ code coming soon~\ldots,'' despite \the\pubdelayinmonths\ months having already elapsed since the  paper had claimed the code could be accessed in its repository.}
-\label{table-repo-summary}\globalTableLabel{supplementaryTableRepoSummary}
+\label{supplement-table-repo-summary}
 \end{table*}
@@
 
-\item the code/data policies of the \plural{\numberOfJournals}{journal} condoned at the time of the sample \emph{and continue to condone\/} poor practice at the time of writing the present paper (\monthname{\month} \the\year) --- for specific details and further explanation of the problems, see \supplement\ section \ref{supplementary-journal-policies-section};
+\item the code/data policies of the \plural{\numberOfJournals}{journal} condoned at the time of the sample \emph{and continue to condone\/} poor practice at the time of writing the present paper (\monthname{\month} \the\year) --- for specific details and further explanation of the problems, see \supplement\ section \ref{supplement-journal-policies-section};
 
@@
 \subsection{Summary of results}
-\label{summary-of-results}
+\label{supplement-summary-of-results}
 The sample selection criteria necessarily identified scientific research with Software Engineering contributions.
@@
 There was no evidence of any critical assessment of code, suggesting that scientists writing papers take it for granted that their code works as they intend.
-No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc (see section \ref{supplementary-best-practice} in this \supplement).
+No competent programmer would take it for granted that their code was correct without following rigorous methods, such as formal methods, regression testing, test driven design, etc (see section \ref{supplement-best-practice} in this \supplement).
 
@@
 
-With the one minor exception, no papers reported anything on any Software Engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{table-repo-summary}).
-The papers themselves, typically only a few published pages, are very brief compared to the substantial code they rely on (see table \ref{table-repo-summary}).
+With the one minor exception, no papers reported anything on any Software Engineering methodologies, which is astonishing given the scale of some of the software effort supporting the papers (table \ref{supplement-table-repo-summary}).
+The papers themselves, typically only a few published pages, are very brief compared to the substantial code they rely on (see table \ref{supplement-table-repo-summary}).
 
@@
 
-\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplementary-journal-policies-section}), but actual publishing practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
+\emph{Nature Digital Medicine\/} and \emph{Royal Society Open Science\/} have clear data and code policies (see \supplement\ section \ref{supplement-journal-policies-section}), but actual publishing practice falls short: \the\countHasBreach\ out of the \plural{\countHasPolicy}{paper} (\pc{\countHasBreach}{\countHasPolicy}) published in them and sampled in the survey manifestly breach their code policies.
 In contrast, \emph{Lancet Digital Health\/}, despite substantial data policies, has no code policy at all to breach.
@@
 Code repositories were used by \plural{\countUsesVersionControlRepository}{paper} (\pc{\countUsesVersionControlRepository}{\dataN}), though \plural{\countNoCodeInRepo}{paper} in the survey claimed to have code on GitHub but there was no code in the repository, only the comment ``Code coming soon\ldots'' (checked at the time of doing the review, then double-checked as detailed in the references in the \supplement, as well as most recently on \clonedate\
-while checking table \ref{table-repo-summary}): in other words, the repository had never been used and the code could never have been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
-
-The sizes of all GitHub repositories are summarized in table \ref{table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
+while checking table \ref{supplement-table-repo-summary}): in other words, the repository had never been used and the code could never have been looked at, let alone reviewed.\footnote{GitHub records show that it had not been deleted after paper submission.} This is a pity because GitHub provides help and targeted warnings and hints like ``No description, website, or topics provided [\ldots] no releases published.'' The lack of code is ironic: the paper concerned \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has as its title ``\emph{Development and validation\/} of a deep neural network model [\ldots]'' (our emphasis), yet it provides no code or development processes for the runnable model it claims to validate, so nobody else (including referees) can check any of the paper's specific claims.
+
+The sizes of all GitHub repositories are summarized in table \ref{supplement-table-repo-summary} (since many papers not using GitHub do not have all code available, non-GitHub code sizes are not easily compared and are not listed).
 
@@
 \subsubsection{Current code policies of sampled journals}
-\globalLabel{supplementary-journal-policies-section}
-
-It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{detecting-against-error} and \ref{restrictive-policies}.
+\label{supplement-journal-policies-section}
+
+It is noteworthy that none of the journals sampled permit any reliable style of managing data in published papers, such as described above in sections \ref{supplement-detecting-against-error} and \ref{supplement-restrictive-policies}.
 The main paper, section \ref{mentionPLOSandIEEE}, additionally mentions \emph{PLOS ONE\/} and \emph{IEEE Transactions on Software Engineering}.
@@
 As part of our open data policy, we ask that data and code are hosted in a public, recognized repository, with an open licence (CC0 or CC-BY) clearly visible on the landing page of your dataset.''}\specifyurl\url{royalsociety.org/journals/authors/author-guidelines/#data}\specifyaccess{Since first accessed 29 July 2020, the policy has been revised (undated, accessed 2 February 2022) but retains the same principles; full policy now available via a DOI \cite{RS-policy}.
-The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{on-code-data-publication}).}
+The policy still retains an emphasis on data accessibility, and continues a lack of awareness that code and data are equivalent and often mixed (see section \ref{supplement-on-code-data-publication}).}
 
@@
 The \emph{Journal of Vascular Surgery\/} has detailed data policies, but no code policy.
-While no \emph{Journal of Vascular Surgery\/} papers were surveyed (but see section \ref{vascular-surgery-note}), the following statement on data policies is relevant:
+While no \emph{Journal of Vascular Surgery\/} papers were surveyed (but see section \ref{supplement-vascular-surgery-note}), the following statement on data policies is relevant:
 
@@
 {
-\def\citenum#1{\cite{#1}}
 \begin{longtable}{@{}cp{2.8in}p{2.8in}@{}}
@@
 
-\label{data-methods-section}
 \subsection{Assessment criteria and methods}
-\label{data-methods-subsection}
+\label{supplement-data-methods-subsection}
 A survey sampled of recent papers that were published online in July 2020, accepted for publication after peer review in \the\numberOfJournals\ high-profile, highly competitive leading peer-reviewed journals, namely \journalBreakdown.
@@
 Some of the files generated from the JSON data are Unix shell scripts.
-For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (as it happens, using \texttt{awk} scripts), e.g., to generate table \ref{table-repo-summary} for this \supplement.
+For example, details of all the papers with GitHub repositories are automatically collected into a shell script so the repositories can be cloned locally and then measured (as it happens, using \texttt{awk} scripts), e.g., to generate table \ref{supplement-table-repo-summary} for this \supplement.
 
@@
 \subsubsection{Detecting and defending against error}
-\label{detecting-against-error}
+\label{supplement-detecting-against-error}
 % check it works for other numbers 
@@
 \subsubsection{Problems of restrictive journal policies}
-\globalLabel{restrictive-policies}
+\label{supplement-restrictive-policies}
 Automatically generated variables are used throughout the paper and this \supplement.
@@
 
-For example, the caption of table \ref{table-repo-summary} in the main paper calculates its ``\plural{\pubdelayinmonths}{month}'' figure from the generated variables recording the repository date of cloning used to provide the data to construct the table.
+For example, the caption of table \ref{supplement-table-repo-summary} in the main paper calculates its ``\plural{\pubdelayinmonths}{month}'' figure from the generated variables recording the repository date of cloning used to provide the data to construct the table.
 The number of months will of course be correctly updated if the paper's repository \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ is subsequently checked again:
@@
 
-\input generated/supplementary-references.tex
+\input generated/supplementary-references.bbl
 

### 2023-05-27 programs/data.js

@@ -1138,7 +1138,7 @@ for (var i = 0; i < data.length; i++) {
     }
     documentFlags = "\\flagStyle{" + documentFlags + "}";
     codeFlags = "\\flagStyle{" + codeFlags + "}";
-    s += "\\citenum{" + (d.reference == "reference" && typeof (d.reference) == "number" ? "R" : "") + d.reference + "} & " + bibdata(d, "dataComment") + documentFlags + " & " + bibdata(d, "codeComment") + codeFlags + "\\\\\n";
+    s += "\\cite{" + (d.reference == "reference" && typeof (d.reference) == "number" ? "R" : "") + d.reference + "} & " + bibdata(d, "dataComment") + documentFlags + " & " + bibdata(d, "codeComment") + codeFlags + "\\\\\n";
 }
 saveFile("assessments.tex", s, "Main summary table for Supplementary Material");
 
@@ -1189,7 +1189,7 @@ for (var i = 0; i < data.length; i++) {
     s += bibdata(d, "year") + ". DOI \\texttt{" + bibdata(d, "doi").trim() + "}" + (bibdata(d, "codeURL") != undefined && bibdata(d, "codeURL") != "" ? (" {Code \\url{" + bibdata(d, "codeURL").trim() + "}}") : "") + "\\\\\\hfill{Accessed " + bibdata(d, "accessed") + ".}\\ " + (!definedq(bibdata(d, "doubleChecked")) || bibdata(d, "doubleChecked") == "" ? "\\textcolor{red}{Not double-checked}" : ("{" +
         "Double-checked " + bibdata(d, "doubleChecked") + "}")) + ".}\\bibskip\n\n";
 }
-saveFile("supplementary-references.tex", s, "Reference list for Supplementary Material");
+saveFile("supplementary-references.bbl", s, "Reference list for Supplementary Material (bbl file)");
 
 // generate summary table
 var t = {};

### 2023-05-31 paper-seb-main.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo

@@
 \input paper-seb-macros.tex
-\input generated/info-for-main.tex
 \input generated/page-lengths.tex
@@
 \makeatother
-
-\long\def\ignore#1{}
 
@@
 
-%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
-%&\texttt{harold@thimbleby.net}\\
-%&ORCID 0000-0003-2222-4243
-%\end{tabular}
-
 %\orcid{0000-0003-2222-4243}
@@
 \justifying
-%\color{red}
-%This is a placeholder for the revisions that are almost complete.
-I have uploaded it now in case the computer system would lose all record of the submission --- I have emailed to ask for a short extension, but with the offices shut, there is no chance of any negotiation.
-Thanks for your understanding.
-%\color{black}\vskip .2cm
-
 \noindent
@@
 \hskip 1em This paper proposes and justifies  ways to improve research and science relying on code:
-
-%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available.
-Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects.
-Code is fallible, so any interpretation that relies on code is also fallible.
-When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized.
-In turn, scientific claims cannot be properly scrutinized.
-
-%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions.
-%The paper reports a pilot survey of peer-reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
-
-%Code can be improved using Software Engineering. This paper argues for specific solutions:
-
-%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.
-
-%\noindent\emph{Solutions:}
 
@@
 
-%\begin{quote}
 \def\seprule{\vskip 2ex\hrule\vskip 2ex}
@@
 \seprule
-%\end{quote}
 \end{abstract}
@@
 
-This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the Covid pandemic.
+This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research because of its significance for informing and driving public health responses to the COVID-19 pandemic.
 A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey.
@@
 Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having that consensus in turn depended on reproducible thermometer measurements; for example, scientists working in different places needed to know they were working with the ``same temperatures'' yet there was for a long period no consensus on what that meant.
-Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
+Contributing to reliable science depended on a thorough understanding of principles, including gradually fixing the confounding factors that were misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the volume of mercury also increases with temperature (but at a different rate), so the volume measurement is inaccurate.}
 Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, that follow international quality standards.
@@
 In the ``old days'' such numbers (\the\dataN; \the\countAuthors; etc) quoted above would have been manually worked out, eyeballed, typed up, and then (hopefully) double-checked by the authors.
-Instead, in this paper, those numbers (and many more) were computed automatically.
-They were inserted into the text of this paper automatically, and are auditable back to their sources.
+Instead, in this paper, those numbers (and many others) were computed automatically, and were then inserted into the text of this paper automatically.
+They are auditable back to their sources.
 Furthermore, they will update automatically --- with no further work from the author --- when the data or calculations change.
@@
 
-where the \the\countAuthors\ is the calculated number, save that line of text to a file, which is then import it into the paper where it is typeset in the normal way.
-Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time.
-Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote \ref{name-example-footnote}).
-
-The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper also update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct \emph{with no further work from the authors}.
+where the \the\countAuthors\ is some number calculated from the relevant data.
+
+Such a line of text is then saved to a file, which is then imported into the paper so the value is named and can be used easily and reliably.
+Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the actual value is at the time --- it will update automatically whenever the data or analysis is improved.
+Some easy \LaTeX\ coding can then present the numbers in the author's or publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11}, \ldots, \digitBlock{1000},  \digitBlock{1001}, etc, as done more realistically elsewhere in this paper (for instance, see footnote \ref{name-example-footnote}).
+
+The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results written up in the paper \emph{also\/} update automatically, and all the statistics, graphs, and analysis reported in the paper update and remain correct \emph{with no further work from the authors}.
 Indeed, if an author corrects a mistake, the correction will apply automatically to all future revisions.
@@
 
-A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes (in an appendix) a substantial and useful literature review.
-Here four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:
+A taut review of such systems is Perkel \cite{perkel-review}, whereas \cite{maneage} discusses in-depth the design trade-offs of one powerful approach, Maneage; the paper \cite{maneage} includes  a substantial and useful literature review (in its appendix).
+Here, four representative tools (WEB, \emph{Mathematica}, Jupyter, and knitr) serve to sample the variety of approaches that are available:
 
@@
 What happens if $2000$ is used for some other purposes as well?
-What happens if some of the $2000$ values are written as $2,000$ or as $2\hskip 0.3em 000.0$?
+What happens if some of the $2000$ values are written as $2,000$ or as $2\hskip 0.3em 000.0$, or if a number containing the digits 2000, like 12000, is changed?
 What happens if some $2000$ are year dates and are changed incorrectly?
+
 Then there are the Human Factors: slips and errors will happen in this workflow anyway \cite{engima}.
@@
 Similar iterative revision cycles happen with any paper, not just with systematic reviews.
-PRISMA ignores the methodological problems of using code (such as the example earlier in this paragraph).
-
-The irony is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented, despite this being one of PRISMA's explicit motivations.
+
+PRISMA, like many such standards, ignores the methodological problems of using code such as the issues raised above.
+
+The irony, then, is that PRISMA says nothing about how to ensure the results of a survey are correctly and reliably presented, despite this being one of PRISMA's explicit motivations.
 PRISMA explicitly warns about methodological issues, but ignores that poorly managed code raises methodological issues that will undermine the validity of research it supports.
@@
 \label{section-pandemic-modeling}
+For an excellent review of the extreme pressures under which scientists were working during the COVID-19 pandemic (but nothing about the role of computers!), see \cite{preventable-book} --- which, while referring to some failed mRNA vaccines, makes an important point on recovering from ``failed'' science:
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0
+Such is the beauty of science: even failed attempts are a step towards more information and progress forward.''} \cite{preventable-book}
+\end{quote}
+
+But progress, if any, would be chaotic if it was not possible to scrutinize exactly what science has failed.
+It undermines progress if science and the code it relies on are not both accessible and adequately defined.
+It is as misleading as getting the statistics or mathematics wrong.
+
+By focusing on influential science and coding undertaken to inform public health policies during a pandemic emergency, this section now focuses on an area where reliable and high quality science using code open to interdisciplinary scrutiny was very obviously required.
+Note that pandemic model correctness is a secondary concern here; correctness is not addressed as without informed scrutiny correctness cannot even be assessed.
+
 A review of epidemic modeling \cite{science-review} says, ``we use the words `computational modeling' loosely,'' and then, curiously, the review discusses exclusively mathematical modeling, implying that for the authors, and for the peer-reviewers, there is no role for code or computation as such.
@@
 
-A systematic review \cite{diagnosis-reviews} of published COVID models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
+A systematic review \cite{diagnosis-reviews} of published COVID-19 models for individual diagnosis and prognosis in clinical care, including apps and online tools, noted the common failure to follow standard TRIPOD guidelines \cite{tripod}.
 The review \cite{diagnosis-reviews} itself ignored the mapping from models to their implementation, yet if code is unreliable, the model \emph{cannot\/} be reliably used, and cannot be reliably interpreted regardless of whether TRIPOD guidelines are followed.
@@
 
-It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are programs intended for human use.
+It should be noted that flowcharts, which the review \cite{diagnosis-reviews} did consider, are graphical programs intended for human use.
 Flowcharts, too, should be designed as carefully as code, for exactly the same reasons.
 
-A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel} uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand, when it did not model air travel or other factors required for later western COVID-19 modeling.
+A high-profile 2020 COVID-19 model \cite{nature-summary,ICmodel}, which influenced UK COVID-19 public health strategies, uses a modified 2005 computer program \cite{avianFluModel,originalICmodel} originally developed for modeling H5N1 in Thailand, when it did not model air travel or other factors required for later western COVID-19 modeling.
 The 2020 model forms part of a series of papers \cite{ICmodel,avianFluModel,originalICmodel} none of which provide details of their code.
 
-A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented C code. As Ferguson, the original code author, noted in an interview,
+A co-author disclosed \cite{tweet} that the code was thousands of lines long and was undocumented code. As Ferguson, the original code author, noted in an interview,
 
@@
 
-This comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications, and a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
+The admission above is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}
+
+The comment was made by a respected, influential world-leading scientist, with many peer-reviewed publications involving computational modeling, with a respectable $h$-index\footnote{$h$-index: the largest value of $h$ such that at least $h$ papers by the author have each been cited at least $h$ times.
 The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
-(Typical $h$ values vary by discipline.)} of 93.
-Ferguson should be well aware of the standards of coding used in his own field.
-Arguably, then, this comment quoted above is representative of the culture of the field.
-
-Ferguson's admission is tantamount to saying that the published scientific findings are and need not be reproducible.\footnote{A constructive discussion of Software Engineering approaches to reproducibility can be found in \cite{basic-reproducibilty}.}
-
-Lack of reproducibility is problematic, especially as the code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
+(Typical $h$ values vary by discipline.)} of 93, and ``one of the top scientists advising the government on its response to the coronavirus crisis'' working in the UK's Scientific Advisory Group for Emergencies (SAGE) group \cite{ferguson-resign}.
+
+Ferguson's code must be representative of best practice at a time when the stakes were high and reliability was known to be essential; and if not representative of best practice, at least representative of accepted practice both in Ferguson's team, the field of epidemiology more widely, as well as with members of the high-powered interdisciplinary SAGE group.
+It is therefore instructive to explore the larger story around this science that uses code.
+
+Lack of reproducibility is problematic, especially as the model code would have required many non-trivial modifications to update it for COVID-19 with its different assumptions; moreover, the code would have had to have been updated very rapidly in response to the urgent COVID-19 crisis.
 
@@
 
-If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results; worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
-In contrast, if code is properly documented, the documentation defines the original intentions (including formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
+If code is undocumented, there would be an understandable temptation to modify it arbitrarily to get desired results (i.e., over-fitting, see section \ref{over-fit}); worse, without documentation and proper commenting, it is methodologically impossible to distinguish legitimate attempts at debugging from merely fudging the results.
+In contrast, if code is properly documented, the documentation defines the original intentions (including, where appropriate, formally using mathematics to do so), and therefore any modifications will need to be justified and explained --- or the theory revised.
 
@@
 
-Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compliers, libraries, or hardware.
+Moreover, C code is not portable, which limits making it available for other scientists to use reliably: C notoriously gets different results with different compilers, libraries, or hardware.
 In fact, in any area where reliable programming is required in a C-like language, a special dialect such as MISRA C is preferred: MISRA C manages the serious design flaws of C that otherwise make it too unreliable \cite{misra}.
@@
 \hskip -\wd0\box0 
-However, none of the criticisms of the code affects the mathematics or science of the simulation''} \cite{thumbs-up}
+However, none of the criticisms of the code affects the mathematics or science of the simulation.''} \cite{thumbs-up}
 \end{quote}
@@
 
-The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirmed that the code could be run again and produced comparable results (compared using an Excel spreadsheet).
-This weak test would pass provided the runs gave the same invalid answers --- it is not usefully a stronger test than just checking that the code compiles.
-
-Running code to obtain results claimed in a paper is a very weak test, and anyway one that should have been checked automatically during paper preparation and submission.
-If you do not know what you are reproducing, as is the case in this \emph{Nature\/} paper, there is little scientific value in doing so.
+The code in \cite{nature-summary,ICmodel} has been ``reproduced,'' as reported in \emph{Nature\/} \cite{codecheck,thumbs-up}, but this so-called reproduction merely confirms that the code can be run again and produce comparable results.
+As Eglen says,
+
+\begin{quote}\raggedright
+{\sf\setbox0=\hbox{``}
+\hskip -\wd0\box0 Each run generated a tab-delimited file in the output folder.
+Two R scripts provided by Prof Ferguson were used to summarise these runs into two summary files [\ldots] These files were compared against the values generated by Prof Ferguson [\ldots] The results were found to be identical.
+Inserting my results into his Excel spreadsheet generated the same pivot tables.
+The codecheck found that: `Small variations (mostly under 5\%) in the numbers were observed [\ldots]'\,''} \cite{codecheck}
+\end{quote}
+
+This test would pass provided the runs gave the same answers regardless of whether the answers are correct --- it is not a usefully stronger test than just checking that the code compiles.
+The comparison relied on running (apparently) unchecked R code to summarize the data, which is potentially misleading unless the published results \cite{ICmodel} exclusively relied on the same summary code.
+In general, reproducing code results, even done formally, does not scrutinize the science, as  \cite{codecheck-certificate} makes clear.
+
+Running code just to obtain results claimed in a paper is a weak test, and anyway one that should be checked routinely during paper preparation and submission.
+However, in this case the reproduction involved a community effort that also refactored and improved the code, which added value to the code and usefully improved its generality \cite{codecheck}.
+The reproduction effort was also certified \cite{codecheck-certificate}, which is the sort of evidence of quality assurance processes that arguably should be required before publication, particularly for critical code such as public health modeling.
+The publicity of this story and the certificate will certainly raise the profile of the scientific value of independent review of code.
 
@@
 In \cite{codecheck,thumbs-up} the reproduction amounted to just re-running the original code.
-It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper; if the original code runs this confirms a basic level of access for the wider scientific community.
+It is certainly essential to establish that a paper's code can be run, as non-working code cannot support any claims in a paper; if the original code runs this confirms a basic level of access for the wider scientific community, and this can be formally certified \cite{codecheck-certificate} so it is appreciated by the community.
 But a more realistic criterion than basic reproduction in this sense is whether an \emph{independently\/} developed model developed from the same paper(s) specifications produces equivalent results (called $N$-version programming, a standard Software Engineering practice \cite{NVP}) like public health surely requires as, indeed, Ferguson's own influenza paper \cite{nvp-ferguson} argues.
-However, much stronger scrutiny of code than ``reproduction'' is required to answer essential questions including:
+
+In general, much stronger scrutiny of code than ``reproduction'' is required to answer essential questions (numbered below for reference) including:
 
@@
 \item Does the code implement the methods described in the paper?
-\item Has the code been over-fitted or tweaked to support specific claims in the paper?'
+\item Has the code been over-fitted or tweaked to support specific claims in the paper?
 \item Is there a definitive version of code?
@@
 \item Is the code documented adequately, so we know what it is trying to do, and how?
-\item \ldots\ and so forth --- questions are numbered for reference. 
+\item \ldots\ and so forth. 
 \end{enumerate}
@@
 
-It is well-known that documenting code helps impro
+It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place, since so many years' opportunity to improve the code have been lost.
+The revised code has now been published, and it too has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.
+
 Some papers \citeeg{pseudo} publish models in pseudo-code, a simplified form of programming.
@@
 
-Epidemiology has a high profile because of the COVID pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
+Epidemiology has a high profile because of the COVID-19 pandemic, but the problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication.
 But other examples that were not rushed include a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model}, which provides no details of its code.
@@
 There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact.
-Climate change modeling is one such example that will have an impact long beyond the COVID pandemic.
+Climate change modeling is one such example that will have an impact long beyond the COVID-19 pandemic.
 
@@
 However, since mature scientists recognize the importance of correct statistics, statistics is positioned at the forefront of their work rather than delayed and sorted out at greater cost later.
-Statistics is routinely in the undergraduate syllabus, so most scientists consciously make appropriate trade-offs minimizing statistical technical debt.
+Because of its recognized role, statistics is routinely in the undergraduate syllabus, so most scientists consciously make appropriate trade-offs minimizing statistical technical debt.
 
@@
 Encouraging and informing the improvement of science and, specifically the reproducibility of science that relies on code, were the main aims of this paper.
+This paper raised problems and suggested some possible solutions: there are solutions, and better ones may yet be found.
+Although further work is desirable, any contributions can help improve science; not everything needs doing before we start.
+
 Further work should research the efficiency, effectiveness, and quality of the various ideas proposed, such as \RAPstar\ and Software Engineering Boards, and propose and evaluate more ideas.
@@
 
-The present paper did not assess the correctness of code used in papers for several reasons.
+The present paper did not assess the correctness of code used in papers (explicitly so in section \ref{section-pandemic-modeling}) for several reasons.
 Code was not documented well-enough (even with high-level discussion in the papers) to know what ``correct'' would mean, and no papers performed adequate tests, let alone provided adequate test material for independent verification (see Table \ref{table-summary}); moreover, installing the software environments to build and test the systems --- different environments for each paper --- was too onerous, even when those environments were specified.
@@
 A pandemic creates unprecedented pressure and exposes problems in scientific methodology.
-During the COVID crisis, code led epidemiological modeling, implemented track and trace and caused problems \cite{excel-fiasco}, modeled mutation pressures against vaccine shortages \cite{science-delays}, and more.
+During the COVID-19 crisis, code led epidemiological modeling, implemented track and trace and caused problems \cite{excel-fiasco}, modeled mutation pressures against vaccine shortages \cite{science-delays}, and more.
 Code drove public policy.
@@
 
-While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID pandemic modeling, the evidence reviewed here suggests that scientific coding practice is inadequate in every field, but particularly worrying in the context of the extreme pressures of managing a pandemic in real time.
+While this paper was originally motivated by Ferguson's public statements \citeeg{tweet,ferguson-interview} about his high-profile COVID-19 pandemic modeling, the evidence reviewed here suggests that scientific coding practice is inadequate in every field, but particularly worrying in the context of the extreme pressures of managing a pandemic in real time.
 As science becomes more and more reliant on computers, we need to correspondingly improve the quality of code, the quality of code policies, the quality of Software Engineering, and the quality of all scientists' understanding of computation and how to manage its unlimited complexity.
@@
 
-Software Engineering Boards (SEBs), as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve computer-based science.
+Software Engineering Boards (SEBs), as proposed in this paper, are an initial, straightforward, constructive, and practical way to support and improve code- and computer-based science.
 If nothing else, the idea of SEBs is something to criticize and improve.
@@
 \long\def\ack#1{\paragraph*{Acknowledgments}#1}
-\long\def\dataaccess#1{\paragraph*{Data and code access}#1}
+
+\newdimen\saveparindent \saveparindent=\parindent
+\newdimen\highlightwidth \highlightwidth=\columnwidth
+\advance \highlightwidth by -.5em
+
+\long\def\highlight#1{\vskip 1em\noindent\colorbox{yellow}{\noindent\begin{minipage}{\highlightwidth}
+\parindent=\saveparindent
+#1\end{minipage}}}
+\long\def\dataaccess#1{\highlight{\paragraph*{Data and code access}#1}}
+
 \long\def\aucontribute#1{\paragraph*{Author contribution}#1}
@@
 
-\dataaccess{There is an extended discussion of the methodology of this paper in the \supplement, section \ref{supplement-on-code-data-publication}.
-The \supplement\ also presents all raw data in tabular form.
-All material is also available for download at \url{github.com/haroldthimbleby/Software-Enginering-Boards},\footnote{\color{red}This is a temporary URL before meeting publication repository requirements for accepted papers.\color{black}} which has been tested in a clean build, etc.
-
-The data is encoded in JSON\@.
-JavaScript code, conveniently in the same file as the JSON data, checks (with \the\JSONerrorCount\ possible classes of error report) and converts the JSON data into \LaTeX\ number registers and summary tables, etc, thus making it trivial to typeset all results reliably in this paper and in its \supplement\ \emph{directly\/} from the automatic data analysis.
-
-In addition, a standard CSV file is generated from the JSON in case this is more convenient, for instance to browse directly as a spreadsheet or to import easily into other programs.}
+\dataaccess{The \supplement\ presents all data in human-readable form, and there is an extended discussion of methodology in section~\ref{supplement-on-code-data-publication}.
+Additionally, all data, code and documentation, including the \LaTeX\ source, is available online at \url{github.com/haroldthimbleby/improving-science}
+
+The raw data is encoded in JSON\@.
+JavaScript code checks against \the\JSONerrorCount\ possible classes of error, and converts the JSON data into \LaTeX, thus making it trivial to typeset results reliably in this paper and the \supplement\ \emph{directly\/} from the analysis.
+In addition, a standard CSV file is generated in case this is convenient, for instance to browse directly in a spreadsheet or to import into other programs.}
 
@@
 
-\bibnewcolumn = 123 % bibentry number of last entry on (final) left hand column
+\bibnewcolumn = 126 % bibentry number of last entry on (final) left hand column
 % Hint: first set it to a very large number, so the final column is as unbalanced as it gets
@@
 % column. Hopefully the righthand column will be of equal length!
+% Set to -1 (not 0) if you don't want this feature
 % PS Don't forget to check there is nothing on the following page!
@@
 \end{document}
-
-%\immediate\write\@auxout{\string \newcount \string \savedlastReferenceNumber
-%\string \savedlastReferenceNumber=\the\savedlastReferenceNumber}
-
-\end{document}

### 2023-05-31 paper-seb-supplementary-material.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo

@@
-\iftrue
-    \PassOptionsToPackage{hyphens}{url}
 \documentclass[10pt,a4paper]{article}
@@
 \usepackage{url}
+\PassOptionsToPackage{hyphens}{url}
 \usepackage{doi}
@@
 \usepackage{fancyhdr}    
-
 \usepackage{graphicx}
 \PassOptionsToPackage{dvipdfmx}{graphicx}
-
 \usepackage{amssymb}
 \usepackage{epstopdf}
+
 \def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
@@
 \input paper-seb-macros.tex
-\fi
 
@@
       \renewcommand{\thefootnote}{\arabic{footnote}}
+      % count the number of bibcite{name}{value} in paper-seb-main.aux
       \let\oldbibcite=\bibcite
-      % example: \bibcite{bad-code}{{31}{2020}{{Richards and Boudnik}}{{Richards and Boudnik}}}
       \newcount\mainNumberOfReferences \mainNumberOfReferences=0
@@
       \input paper-seb-main.aux 
-      %\resetlinenumber[\MainPaperLastLine]
       \let\bibcite=\oldbibcite
@@
 
-\makeatletter
-
-\def\rememberpage{\immediate\write\@auxout{\bslash gdef\bslash SupplementalLastPage{\thepage}}}
-
-% putting \typeout in the right footer will get repeated warnings (which I hope will be noticed!)
-% to complement the red text in the footer itself on every page
-\@ifundefined{SupplementalLastPage}{\rfoot{\supplement\ \thepage\ \textcolor{red}{\bf(Rerun to get number of pages)}\typeout{* Rerun to get correct number of pages in the running footer}}}{\rfoot{\supplement\ \thepage/\SupplementalLastPage}}
-
-\makeatother
-
-\lfoot{\today}
-\cfoot{}
-\lhead{}
-\rhead{}
+\rfoot{}
+\lfoot{}
+\cfoot{\thepage}
+\lhead{\mytitle: \supplement}
+\rhead{Harold Thimbleby}
 \renewcommand{\headrulewidth}{0pt}
@@
 
-% this will be used to save cross references that can be used in the main paper too
-\newwrite\infofileForMain
-\immediate\openout\infofileForMain=generated/info-for-main.tex
-
 \input generated/info-for-supplement.tex
@@
 \immediate\openout\infofileForSupplement=generated/info-for-supplement.tex
-
-\def\dotalph#1{\ifnum\csname c@#1\endcsname>0 .\alph{#1}\fi}
-
-%\makeatletter
-%\def\globalLabel#1{% we also need journal code-policies to be used in the main paper...
-%\typeout{Global globalLabel: #1}%
-%\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{section}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-%% as well as in the current document....
-%\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{section}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-%}
-%\def\globalTableLabel#1{% we also need journal code-policies to be used in the main paper...
-%\typeout{Global globalTableLabel: #1}%
-%\immediate\write\infofileForMain{\bslash newlabel{#1}{{\arabic{table}}{\thepage}{DUMMYA}{DUMMYB}{DUMMYC}}}
-%% as well as in the current document....
-%\immediate\write\@auxout{\bslash newlabel{#1}{{\bslash arabic{table}\dotalph{subsection}}{\thepage}{AAA}{BBB}{CCC}}}
-%}
-%\def\globalLabelValue#1#2{% we also need journal code-policies to be used in the main paper...
-%\typeout{Global globalLabelValue: #1 := #2}%
-%\immediate\write\infofileForMain{\bslash newlabel{#1}{{#2}{\thepage}{AAA}{BBB}{CCC}}}
-%}
-%\makeatother
 
@@
 \item Ethics Boards also have to deal with software, and it is clear that they often fail to do this effectively.
-The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in the main paper, are cases in point.
+The case of the retraction of a peer reviewed articles for \emph{The Lancet\/} \cite{science-lancet1,science-lancet2,lancet-learning} and the \emph{Journal of Vascular Surgery\/} \cite{jvs1,jvs2,jvs3}, discussed in section
+\ref{supplement-vascular-surgery-note}, are cases in point.
 \end{itemize}
@@
 If version control is not used, one has no idea what the current program actually is.
-{Version control is essential for \emph{reproducibility\/}: \cite{basic-reproducibilty,reproducibility} it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
+{Version control is essential for \emph{reproducibility\/} \cite{basic-reproducibilty,reproducibility} (see also section \ref{reproducibility-concerns}): it enables efforts to duplicate work to start with the exact version that was used in any published paper, provided that the published paper discloses the version and a URL for the relevant repository.
 Note that version control should also be used for data and web site data used by code, otherwise the results reported are not replicable.}
@@
 Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable.
-Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}.
+Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{viruses}.
 The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
@@
 
-{Significant cyber-vulnerabilities result from there being no difference between code and data.
-For example: an email arrives, which brings data to a user.
-The user opens an attachment, perhaps a word processor text document, which is more data.
+Significant cyber-vulnerabilities result from there being no difference between code and data.
+For example: an email arrives, which brings a paper to work on or other data to a user.
+The user opens the attachment, perhaps a word processor text document, which is more data.
 The word processor runs macros in the text document --- but now it is code.
-The macros move data onto the user's disc.
-The data there then runs as code, and corrupts the user's data across the disc --- which includes both data and code stored in files.
-And so on.
-Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.}
+The macros move data onto the user's disk.
+The data there then runs as code, and corrupts the user's data across the disk --- which includes both data and code stored in files.
+And so on, spreading around the world when this user emails their work to a colleague.
+Each step of a computer virus infection crosses over non-existent ``boundaries'' between data and code \cite{viruses}.
 
@@
 
-\immediate\write\infofileForMain{\bslash def\bslash MaxMainPaperCitationNumber{\the\temporary}}
-
 {\newcount \bibitemn \bibitemn = 0
@@
 
+\catcode`\%=11
+\immediate\write\infofileForSupplement{% map Git repo names to their citation numbers}
+\immediate\write\infofileForSupplement{% generated by paper-seb-supplementary-material.tex}
+\catcode`\%=14
 \def\maprepo#1{\immediate\write\infofileForSupplement{\bslash expandafter\bslash def \bslash csname cite-#1\bslash endcsname{[\the\bibciten]}}}
@@
 
-\immediate\closeout\infofileForMain
 \immediate\closeout\infofileForSupplement
-\rememberpage
 \end{document}

### 2023-05-31 generated/allGitRepos.sh

@@ -1,3 +1,4 @@
+# Generated by programs/data.js
 # run this shell script in directory models
 rm -rf "git-MetricSelectionFramework"
 git clone "http://github.com/ChristophKanzler/MetricSelectionFramework" "git-MetricSelectionFramework"

### 2023-05-31 generated/open-doi.sh

@@ -1,3 +1,4 @@
+# Generated by programs/data.js
 open "http://www.doi.org/10.1038/s41746-020-0300-0"
 open "http://www.doi.org/10.1038/s41746-020-0286-7"
 open "http://www.doi.org/10.1038/s41746-020-0287-6"

### 2023-05-31 programs/data.js

@@ -744,6 +744,19 @@ var generatedFiles = [];
 
 function saveFile(fileName, buffer, description) {
     var fullFileName = "generated/" + fileName;
+    
+    if( fileName.endsWith(".tex") || fileName.endsWith(".bbl") )
+    	buffer = "% Generated by programs/data.js\n"+buffer;
+    else if( fileName.endsWith(".html") )
+    	buffer = "<!-- Generated by programs/data.js -->\n"+buffer;
+    else if( fileName.endsWith(".sh") )
+    	buffer = "# Generated by programs/data.js\n"+buffer;
+    else if( fileName.endsWith(".nb") )
+    	buffer = "(* Generated by programs/data.js *)\n"+buffer;
+    else if( fileName.endsWith(".csv") )
+    	buffer = "Generated by programs/data.js\n"+buffer;
+	else console.log("      *** File name suffix in "+fileName+" not recognized and therefore not signed...");
+    
     fs.open(fullFileName, 'w', function (err, fd) {
         if (err) {
             if (err.code === 'EEXIST') {
@@ -1451,7 +1464,7 @@ for (var j in descriptionsOfFields)
             unuseddata = 1;
         }
         if (!definedq(bibliographyNames[j])) warnCount++;
-        console.log((definedq(bibliographyNames[j]) ? "      " : "") + "    " + j + " not used " + (definedq(bibliographyNames[j]) ? "except in data summary table and/or bibliography" : "in generated data"));
+        console.log((definedq(bibliographyNames[j]) ? "          " : "") + "    " + j + " not used " + (definedq(bibliographyNames[j]) ? "except in data summary table and/or bibliography" : "in generated data"));
     }
 
     // since so much noise is generated summmarize error warnings...

### 2023-06-06 expanded-appendix.tex
Preparing to upload to Computer Journal

--- 2023-05-31 paper-seb-supplementary-material.tex
@@
+% 
+% This file has been expanded from LaTeX file (appendix.tex)
+%   to make this straight forward LaTeX file (expanded-appendix.tex)
+%     that depends on nothing (it inputs no files apart from the document class files).
+%       Each required file's text was input in situ - search for % expand
+% 
+% (Otherwise the file appendix.tex would use BibTeX, a macro file, tables of contents, its own and the other files's .aux files, and includes various files in generated/*, etc.)
+%
+
+\makeatletter
+\relax 
+\providecommand\hyper@newdestlabel[2]{}
+\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
+\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
+\global\let\oldnewlabel\newlabel
+\gdef\newlabel#1#2{\newlabelxx{#1}#2}
+\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
+\AtEndDocument{\ifx\hyper@anchor\@undefined
+\let\newlabel\oldnewlabel
+\fi}
+\fi}
+\global\let\hyper@last\relax 
+\gdef\HyperFirstAtBeginDocument#1{#1}
+\providecommand\HyField@AuxAddToFields[1]{}
+\providecommand\HyField@AuxAddToCoFields[2]{}
+\citation{vancouver}
+\citation{hamming}
+\citation{fixit}
+\citation{cosmos}
+\@writefile{toc}{\contentsline {section}{\numberline {8}Further issues for Software Engineering Boards (SEBs)}{38}{section.8}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {8.a}Relationships of SEBs to Ethics Boards}{38}{subsection.8.1}\protected@file@percent }
+\citation{ethics-paper}
+\citation{ai-ethics}
+\citation{science-lancet1,science-lancet2,lancet-learning}
+\citation{jvs1,jvs2,jvs3}
+\citation{sommerville,knight}
+\citation{cosmos}
+\citation{swebok}
+\citation{turing-way}
+\citation{wpf}
+\citation{humphrey}
+\citation{managers}
+\@writefile{toc}{\contentsline {subsection}{\numberline {8.b}SEBs are necessary but not sufficient}{39}{subsection.8.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {9}Software engineering best practice}{39}{section.9}\protected@file@percent }
+\newlabel{supplement-best-practice}{{9}{39}{Software engineering best practice}{section.9}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.a}Requirements}{39}{subsection.9.1}\protected@file@percent }
+\citation{differential-equations}
+\citation{cbc}
+\citation{spark}
+\citation{misra}
+\citation{hoare}
+\citation{hoare}
+\citation{numerals}
+\citation{heedless}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.b}Formal methods}{40}{subsection.9.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.c}Defensive programming}{40}{subsection.9.3}\protected@file@percent }
+\citation{onlyPaperWithChecks}
+\citation{tweet,plos}
+\citation{heedless}
+\citation{spark}
+\citation{haskell}
+\citation{basic-reproducibilty,reproducibility}
+\citation{NVP}
+\citation{avianFluModel}
+\citation{ICmodel}
+\citation{flu-model}
+\citation{excel1,excel2}
+\citation{npm}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.d}Using inappropriate programming languages undermines reliability}{41}{subsection.9.4}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.e}Open source and version control}{41}{subsection.9.5}\protected@file@percent }
+\citation{open-source}
+\citation{sommerville,knight,NVP}
+\citation{assurance-case}
+\citation{LP}
+\citation{relit}
+\citation{hci1,hci2}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.f}Rigorous testing}{42}{subsection.9.6}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.g}Good documentation and record keeping}{42}{subsection.9.7}\protected@file@percent }
+\citation{flexibility}
+\citation{iso}
+\citation{178C}
+\citation{fixit}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.h}Usability}{43}{subsection.9.8}\protected@file@percent }
+\newlabel{supplement-material-makefiles}{{9.h}{43}{Usability}{subsection.9.8}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.i}Reusing quality solutions}{43}{subsection.9.9}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.j}Simplicity}{43}{subsection.9.10}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.k}Compliance with standards}{43}{subsection.9.11}\protected@file@percent }
+\citation{peerReview,knight}
+\citation{ssi-report,ssi-url}
+\citation{acm-artifacts}
+\citation{security-engineering,cyber-cacm}
+\citation{example-numerical-error}
+\citation{hamming}
+\citation{Schneier}
+\citation{electronic-evidence}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.l}Effective multidisciplinary teamwork}{44}{subsection.9.12}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.m}Continuous Professional Development (CPD)}{44}{subsection.9.13}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.n}Security and other factors}{44}{subsection.9.14}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.o}Software is a human activity}{44}{subsection.9.15}\protected@file@percent }
+\citation{fixit}
+\citation{popper-conjectures-refutations}
+\citation{simon-artificial}
+\citation{dunning-kruger,dunning-kruger-numeracy}
+\citation{dunning-kruger}
+\citation{fixit}
+\citation{enigma}
+\citation{harking}
+\citation{viruses}
+\@writefile{toc}{\contentsline {section}{\numberline {10}Code, data, and publication}{45}{section.10}\protected@file@percent }
+\newlabel{supplement-on-code-data-publication}{{10}{45}{Code, data, and publication}{section.10}{}}
+\citation{viruses}
+\citation{strachey}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.a}When magic numbers become magic code}{46}{subsection.10.1}\protected@file@percent }
+\citation{henderson}
+\citation{paper-usesRMarkdown}
+\citation{RMarkdown}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.b}When data is code}{47}{subsection.10.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.c}Exploiting code as data for more reliable science}{47}{subsection.10.3}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.d}When data is text: Exploiting code for reliable publication}{47}{subsection.10.4}\protected@file@percent }
+\citation{Spiegelhalter}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.e}Data and polynomials used in the paper}{48}{subsection.10.5}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.f}Comparing conventional and RAP approaches}{48}{subsection.10.6}\protected@file@percent }
+\newlabel{supplement-RAP-diagrams}{{10.f}{48}{Comparing conventional and RAP approaches}{subsection.10.6}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {11}The Spiegelhalter trustworthiness questions}{48}{section.11}\protected@file@percent }
+\newlabel{supplement-Spiegelhalter-section}{{11}{48}{The Spiegelhalter trustworthiness questions}{section.11}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {2}{\ignorespaces The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
+Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code.}}{49}{figure.2}\protected@file@percent }
+\newlabel{supplement-fig:basic}{{2}{49}{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
+Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code}{figure.2}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {3}{\ignorespaces When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
+The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
+Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
+In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication.
+}}{49}{figure.3}\protected@file@percent }
+\newlabel{supplement-fig:excel}{{3}{49}{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
+The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
+Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
+In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication}{figure.3}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {4}{\ignorespaces Improving over the normal approach (figure \ref {supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+In the present paper, this was done by generating text files of \LaTeX  \ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX  \ names for all of the code-generated values.
+Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
+Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen.}}{50}{figure.4}\protected@file@percent }
+\newlabel{supplement-fig:rap}{{4}{50}{Improving over the normal approach (figure \ref {supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+In the present paper, this was done by generating text files of \LaTeX \ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX \ names for all of the code-generated values.
+Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
+Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen}{figure.4}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.a}How trustworthy are the numbers?}{50}{subsection.11.1}\protected@file@percent }
+\@writefile{lof}{\contentsline {figure}{\numberline {5}{\ignorespaces In a notebook system, such as \emph  {Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
+The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
+Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
+In well-engineered code, the notebook will directly contain full documentation.
+Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc).}}{51}{figure.5}\protected@file@percent }
+\newlabel{supplement-fig:notebook}{{5}{51}{In a notebook system, such as \emph {Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
+The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
+Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
+In well-engineered code, the notebook will directly contain full documentation.
+Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc)}{figure.5}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.b}How trustworthy is the source?}{51}{subsection.11.2}\protected@file@percent }
+\citation{lancet-retracted}
+\citation{science-lancet1,science-lancet2}
+\citation{lancet-learning}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.c}How trustworthy is the interpretation?}{52}{subsection.11.3}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}\protected@file@percent }
+\newlabel{supplement-survey-section}{{12}{52}{A pilot survey of computational science}{section.12}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{52}{subsection.12.1}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.1}\emph  {The Lancet\/}}{52}{subsubsection.12.1.1}\protected@file@percent }
+\citation{jvs1}
+\citation{jvs2,jvs3}
+\citation{jvs1}
+\citation{numerals}
+\citation{fda}
+\@writefile{lot}{\contentsline {table}{\numberline {6}{\ignorespaces Conforming to the RAP{\tt  +}\ methodology, the abbreviated summary above was generated automatically, by using \texttt  {make data}.
+A full list of \texttt  {make} options is generated by \texttt  {make} or \texttt  {make help} when run on the Unix command line.
+}}{53}{table.6}\protected@file@percent }
+\newlabel{supplement-table:options}{{6}{53}{Conforming to the \RAPstar \ methodology, the abbreviated summary above was generated automatically, by using \texttt {make data}.
+A full list of \texttt {make} options is generated by \texttt {make} or \texttt {make help} when run on the Unix command line}{table.6}{}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.2}\emph  {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}\protected@file@percent }
+\newlabel{supplement-vascular-surgery-note}{{12.a.2}{53}{\emph {Journal of Vascular Surgery}}{subsubsection.12.1.2}{}}
+\citation{ICmodel}
+\citation{ICmodel}
+\citation{ICmodel}
+\@writefile{lot}{\contentsline {table}{\numberline {7}{\ignorespaces Overview of the peer-reviewed paper sample.
+(Convenient copy of main paper's table \ref {table-overview}.)}}{54}{table.7}\protected@file@percent }
+\newlabel{supplement-table-overview}{{7}{54}{Overview of the peer-reviewed paper sample. (Convenient copy of main paper's table \ref {table-overview}.)}{table.7}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {8}{\ignorespaces Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}}{54}{table.8}\protected@file@percent }
+\newlabel{supplement-table-summary}{{8}{54}{Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}{table.8}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{54}{subsection.12.2}\protected@file@percent }
+\citation{nature-review}
+\@writefile{lot}{\contentsline {table}{\numberline {9}{\ignorespaces Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt  {covid-sim} \cite  {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the Supplemental Material), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+At the time of cloning and checking all repositories in May 2023, paper [182]\ still had nothing in its repository except a single file still saying ``\ldots  \ code coming soon\nobreakspace  {}\ldots  ,'' despite 47\ months having already elapsed since the paper had claimed the code could be accessed in its repository.}}{55}{table.9}\protected@file@percent }
+\newlabel{supplement-table-repo-summary}{{9}{55}{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt {covid-sim} \cite {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement ), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+At the time of cloning and checking all repositories in \clonewhen , paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname \ still had nothing in its repository except a single file still saying ``\ldots \ code coming soon~\ldots ,'' despite \the \pubdelayinmonths \ months having already elapsed since the paper had claimed the code could be accessed in its repository}{table.9}{}}
+\citation{open-source}
+\citation{stability}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.c}Summary of results}{56}{subsection.12.3}\protected@file@percent }
+\newlabel{supplement-summary-of-results}{{12.c}{56}{Summary of results}{subsection.12.3}{}}
+\citation{RS-policy}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.1}Current code policies of sampled journals}{57}{subsubsection.12.3.1}\protected@file@percent }
+\newlabel{supplement-journal-policies-section}{{12.c.1}{57}{Current code policies of sampled journals}{subsubsection.12.3.1}{}}
+\citation{ref-1}
+\citation{ref-2}
+\citation{ref-3}
+\citation{ref-4}
+\citation{ref-5}
+\citation{ref-6}
+\citation{ref-7}
+\citation{ref-8}
+\citation{ref-9}
+\citation{ref-10}
+\citation{ref-11}
+\citation{ref-12}
+\citation{paper-usesRMarkdown}
+\citation{ref-14}
+\citation{onlyPaperWithChecks}
+\citation{ref-16}
+\citation{ref-17}
+\citation{ethics-paper}
+\citation{ref-19}
+\citation{ref-20}
+\citation{ref-21}
+\citation{ref-22}
+\citation{ref-23}
+\citation{example-numerical-error}
+\citation{ref-25}
+\citation{ref-26}
+\citation{ref-27}
+\citation{ref-28}
+\citation{ref-29}
+\citation{ref-30}
+\citation{ref-31}
+\citation{ref-32}
+\gdef \LT@i {\LT@entry 
+    {2}{26.55563pt}\LT@entry 
+    {1}{214.35622pt}\LT@entry 
+    {1}{208.35622pt}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{58}{subsubsection.12.3.2}\protected@file@percent }
+\citation{prisma}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.d}Assessment criteria and methods}{60}{subsection.12.4}\protected@file@percent }
+\newlabel{supplement-data-methods-subsection}{{12.d}{60}{Assessment criteria and methods}{subsection.12.4}{}}
+\citation{fixit}
+\citation{fixit}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.1}Detecting and defending against error}{62}{subsubsection.12.4.1}\protected@file@percent }
+\newlabel{supplement-detecting-against-error}{{12.d.1}{62}{Detecting and defending against error}{subsubsection.12.4.1}{}}
+\citation{excel-fiasco}
+\citation{latex-calculus}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.2}Defending against system problems}{63}{subsubsection.12.4.2}\protected@file@percent }
+\bibstyle{revisedCompJ}
+\bibdata{appendix.bib}
+\bibcite{vancouver}{{{130}{}{{}}{{}}}}
+\bibcite{cosmos}{{{131}{}{{}}{{}}}}
+\bibcite{ai-ethics}{{{132}{}{{}}{{}}}}
+\bibcite{science-lancet1}{{{133}{}{{}}{{}}}}
+\bibcite{science-lancet2}{{{134}{}{{}}{{}}}}
+\bibcite{lancet-learning}{{{135}{}{{}}{{}}}}
+\bibcite{jvs1}{{{136}{}{{}}{{}}}}
+\bibcite{jvs2}{{{137}{}{{}}{{}}}}
+\bibcite{jvs3}{{{138}{}{{}}{{}}}}
+\bibcite{swebok}{{{139}{}{{}}{{}}}}
+\bibcite{wpf}{{{140}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{64}{subsubsection.12.4.3}\protected@file@percent }
+\newlabel{supplement-restrictive-policies}{{12.d.3}{64}{Problems of restrictive journal policies}{subsubsection.12.4.3}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
+\bibcite{humphrey}{{{141}{}{{}}{{}}}}
+\bibcite{managers}{{{142}{}{{}}{{}}}}
+\bibcite{differential-equations}{{{143}{}{{}}{{}}}}
+\bibcite{numerals}{{{144}{}{{}}{{}}}}
+\bibcite{heedless}{{{145}{}{{}}{{}}}}
+\bibcite{plos}{{{146}{}{{}}{{}}}}
+\bibcite{excel1}{{{147}{}{{}}{{}}}}
+\bibcite{excel2}{{{148}{}{{}}{{}}}}
+\bibcite{npm}{{{149}{}{{}}{{}}}}
+\bibcite{hci1}{{{150}{}{{}}{{}}}}
+\bibcite{hci2}{{{151}{}{{}}{{}}}}
+\bibcite{flexibility}{{{152}{}{{}}{{}}}}
+\bibcite{iso}{{{153}{}{{}}{{}}}}
+\bibcite{178C}{{{154}{}{{}}{{}}}}
+\bibcite{peerReview}{{{155}{}{{}}{{}}}}
+\bibcite{ssi-report}{{{156}{}{{}}{{}}}}
+\bibcite{ssi-url}{{{157}{}{{}}{{}}}}
+\bibcite{security-engineering}{{{158}{}{{}}{{}}}}
+\bibcite{cyber-cacm}{{{159}{}{{}}{{}}}}
+\bibcite{Schneier}{{{160}{}{{}}{{}}}}
+\bibcite{electronic-evidence}{{{161}{}{{}}{{}}}}
+\bibcite{simon-artificial}{{{162}{}{{}}{{}}}}
+\bibcite{dunning-kruger}{{{163}{}{{}}{{}}}}
+\bibcite{dunning-kruger-numeracy}{{{164}{}{{}}{{}}}}
+\bibcite{enigma}{{{165}{}{{}}{{}}}}
+\bibcite{harking}{{{166}{}{{}}{{}}}}
+\bibcite{viruses}{{{167}{}{{}}{{}}}}
+\bibcite{strachey}{{{168}{}{{}}{{}}}}
+\bibcite{henderson}{{{169}{}{{}}{{}}}}
+\bibcite{lancet-retracted}{{{170}{}{{}}{{}}}}
+\bibcite{fda}{{{171}{}{{}}{{}}}}
+\bibcite{stability}{{{172}{}{{}}{{}}}}
+\bibcite{RS-policy}{{{173}{}{{}}{{}}}}
+\bibcite{latex-calculus}{{{174}{}{{}}{{}}}}
+\bibcite{ref-1}{{{175}{}{{}}{{}}}}
+\bibcite{ref-2}{{{176}{}{{}}{{}}}}
+\bibcite{ref-3}{{{177}{}{{}}{{}}}}
+\bibcite{ref-4}{{{178}{}{{}}{{}}}}
+\bibcite{ref-5}{{{179}{}{{}}{{}}}}
+\bibcite{ref-6}{{{180}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
+\bibcite{ref-7}{{{181}{}{{}}{{}}}}
+\bibcite{ref-8}{{{182}{}{{}}{{}}}}
+\bibcite{ref-9}{{{183}{}{{}}{{}}}}
+\bibcite{ref-10}{{{184}{}{{}}{{}}}}
+\bibcite{ref-11}{{{185}{}{{}}{{}}}}
+\bibcite{ref-12}{{{186}{}{{}}{{}}}}
+\bibcite{paper-usesRMarkdown}{{{187}{}{{}}{{}}}}
+\bibcite{ref-14}{{{188}{}{{}}{{}}}}
+\bibcite{onlyPaperWithChecks}{{{189}{}{{}}{{}}}}
+\bibcite{ref-16}{{{190}{}{{}}{{}}}}
+\bibcite{ref-17}{{{191}{}{{}}{{}}}}
+\bibcite{ethics-paper}{{{192}{}{{}}{{}}}}
+\bibcite{ref-19}{{{193}{}{{}}{{}}}}
+\bibcite{ref-20}{{{194}{}{{}}{{}}}}
+\bibcite{ref-21}{{{195}{}{{}}{{}}}}
+\bibcite{ref-22}{{{196}{}{{}}{{}}}}
+\bibcite{ref-23}{{{197}{}{{}}{{}}}}
+\bibcite{example-numerical-error}{{{198}{}{{}}{{}}}}
+\bibcite{ref-25}{{{199}{}{{}}{{}}}}
+\bibcite{ref-26}{{{200}{}{{}}{{}}}}
+\bibcite{ref-27}{{{201}{}{{}}{{}}}}
+\bibcite{ref-28}{{{202}{}{{}}{{}}}}
+\bibcite{ref-29}{{{203}{}{{}}{{}}}}
+\bibcite{ref-30}{{{204}{}{{}}{{}}}}
+\bibcite{ref-31}{{{205}{}{{}}{{}}}}
+\bibcite{ref-32}{{{206}{}{{}}{{}}}}
+\gdef \@abspage@last{34}
+
+\makeatother
 \documentclass[10pt,a4paper]{article}
@@
 \def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}    
-\input paper-seb-macros.tex
+% expand macros.tex
+% due to finding out that PDF doesn't work everywhere, here's where we define the image type we're using for imaged created by the Mathematica notebooks
+\def\imageFileSuffix{jpg}
+
+\usepackage[hyphens]{url}
+\usepackage{hyperref}
+\usepackage{xcolor}
+\usepackage{longtable}
+\usepackage{cite}
+
+\renewcommand\thesubsection{\thesection.\alph{subsection}}
+\renewcommand\thesubsubsection{\thesection.\alph{subsection}.\arabic{subsubsection}}
+
+\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
+\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
+
+\makeatletter
+    \def\do@url@hyp{\do/}
+    \usepackage{moreverb}
+\makeatother
+
+\DeclareUrlCommand\doi{\def\UrlLeft{{\textrm{DOI}\ } }}
+\DeclareUrlCommand\url{\def\UrlLeft{{\textrm{URL}\ } }}
+\DeclareUrlCommand\midurl{\def\UrlLeft{}\allowbreak }
+
+\renewcommand{\thefootnote}{\arabic{footnote}}
+
+\def\mytitle{Improving science that uses code}
+\def\supplement{Supplemental Material}
+
+\def\citeeg#1{[e.g., \citenum{#1}]}
+
+% calculate rounded percentages
+\newcount \t
+\newcount \tu
+\def\pc#1#2{\t = #1%
+\multiply \t by 100000%
+\divide \t by #2% ?? 0 times percentage
+\tu = \t
+\divide \t by 1000% percentage, but no idea of remainder
+\multiply \t by 1000% 1000 times percentage, with 00 as bottom digits
+\advance \tu by -\t% bottom two digits
+\divide \t by 1000\relax%
+% round to nearest integer
+% if ending in > .5 round up
+% if ending in exactly .5 round towards nearest even number
+\ifnum \tu > 500 % round up
+	\advance \t by 1 
+\else
+	\ifnum \tu = 500 
+		\ifodd \t % it's odd, so round up to even number 
+			\advance \t by 1
+		\else % leave rounded to even number 	
+		\fi
+	\fi
+\fi
+\the\t\%}
+
+% code to make digits into blocks of 3 (as in 1000 -> 1 000)
+\def\xBlockSep{\textcolor{red}{\hskip 0.3em :}\hskip 0.3em} % International Standards are to use a small space. UK typically uses a comma
+% NOTE currently, the few decimal numbers (eg 0.000 5) use hskip explicitly, since defining \xBlockSep as a comma would definitely be wrong for decimals
+\newcount\xdigitBlockn
+\newcount\xdigitBlocktmp
+\def\xdigitBlock#1{%
+        \xdigitBlockn=#1\relax
+        \ifnum \xdigitBlockn < 0
+                $-$%
+                \xdigitBlockn = -\xdigitBlockn
+        \fi
+        \let \xdigitBlockpptemp \empty
+        \loop \ifnum \xdigitBlockn > 999
+                \xdigitBlocktmp = \xdigitBlockn
+                \divide \xdigitBlockn by 1000
+                \xdigitBlocktmp = \numexpr \xdigitBlocktmp - 1000*\xdigitBlockn \relax
+                \edef \xdigitBlockpptemp{\xBlockSep\ifnum \xdigitBlocktmp < 100  0\ifnum \xdigitBlocktmp < 10 0\fi\fi \the\xdigitBlocktmp \xdigitBlockpptemp}%
+        \repeat
+        \the\xdigitBlockn
+        \xdigitBlockpptemp
+}
+
+\def\digitBlock#1{\hbox{\xdigitBlock{#1}}}
+
+% simple macro to use a number register to pluralize (add an s) to words....
+% \plural{\t}{fred} => \the\t\ freds if plural, or => one fred if singular
+\def\plural#1#2{\ifnum #1=1 
+	one #2%
+\else
+	\the#1\ #2s%
+\fi}
+
+% initialiseBibliography <title> <starting number> <introductory text>
+\newcount \bibciten \bibciten=0
+\def\initialiseBibliography#1#2#3{% title, starting citation number, an initial paragraph
+	\global\def\refname{\arabic{section} #1}
+    \global\def\unnumberedrefname{#1}
+    \global\bibciten=#2
+    \global\def\startBibliography{#3}
+}
+
+\def\bibskip{\vskip 1ex} % the gap between bib items generated by data.js
+
+% read in any constants defined from the JSON or Git data
+% expand generated/constants.tex
+% Generated by programs/data.js
+\def\journalBreakdown{\emph{Lancet Digital Health\/} ($N=6$), \emph{Nature Digital Medicine\/} ($N=12$) and \emph{Royal Society Open Science\/} ($N=14$)}
+\def\tabularJournalBreakdown{&\hbox to 3em {\hfill 6}\hskip 1em \emph{Lancet Digital Health}\\&\hbox to 3em {\hfill 12}\hskip 1em \emph{Nature Digital Medicine}\\&\hbox to 3em {\hfill 14}\hskip 1em \emph{Royal Society Open Science}\\}
+\global\newcount \dataN \global\dataN=32
+\global\newcount \countAuthors \global\countAuthors=264
+\global\newcount \countHasBreach \global\countHasBreach=11
+\global\newcount \countHasPolicy \global\countHasPolicy=26
+\global\newcount \countUsesVersionControlRepository \global\countUsesVersionControlRepository=10
+\global\newcount \counthasDataRepository \global\counthasDataRepository=9
+\global\newcount \countNoCodeInRepo \global\countNoCodeInRepo=1
+\global\newcount \numberOfJournals \global\numberOfJournals=3
+\global\newcount \countCodetested \global\countCodetested=0
+\global\newcount \hasDevelopedRigorously \global\hasDevelopedRigorously=0
+\global\newcount \countFields \global\countFields=36
+\global\newcount \countCommentedOut \global\countCommentedOut=7
+\global\newcount \countNotCommentedOut \global\countNotCommentedOut=2
+\global\newcount \countMissingCode \global\countMissingCode=1
+\global\def \averagePagesPerAuthor{1.2916666666666667}
+% end expanding generated/constants.tex
+% expand generated/metadata.tex
+% shell script (make-metadata.sh) generated LaTeX variable values counting errors detected by data.js
+\newcount \JSONerrorCount
+\JSONerrorCount = 30
+\newcount \dataVariableCount
+\dataVariableCount = 63
+% end expanding generated/metadata.tex
+% expand generated/covidsimSummary.tex
+% data calculated from the Ferguson covid-sim repo
+\def \covidsimkLOC {25}
+\def \covidsimfiles {229}
+\def \covidsimdata{734 Mb}
+% end expanding generated/covidsimSummary.tex
+% expand generated/clone-date.tex
+% date generated by running generated/allGitRepos.sh downloading Git repos
+\def\clonedate {\ignorespaces 25 May 2023}
+\def\cloneyear {2023}
+\def\clonemonth {05}
+% end expanding generated/clone-date.tex
+
+\newcount \pubdelayinmonths
+\pubdelayinmonths=\cloneyear
+\advance \pubdelayinmonths by -2019 % submitted in 2019
+\multiply \pubdelayinmonths by 12
+\advance \pubdelayinmonths by -6 % submitted in June
+\advance \pubdelayinmonths by \clonemonth
+
+\def\monthname#1{\ifcase#1 \or 
+January\or 
+February\or 
+March\or 
+April\or 
+May\or 
+June\or 
+July\or 
+August\or 
+September\or 
+October\or 
+November\or 
+December\fi}
+
+\def\numbertoname#1{\ifcase #1 zero\or one\or two\or three\or four\or five\or six\or seven\or eight\or nine\else #1\fi}
+\def\Numbertoname#1{\ifcase #1 Zero\or One\or Two\or Three\or Four\or Five\or Six\or Seven\or Eight\or Nine\else #1\fi}
+
+\def\clonewhen{\monthname{\clonemonth} \cloneyear}
+
+\def\RAPstar{RAP{\tt +}}
+\def\RAPstars{RAPs{\tt +}}
+\def\RAPstarp#1{RAP{\tt +}#1}
+\def\italicRAPstar{RAP{\tt +}}
+% end expanding macros.tex
 
@@
       \renewcommand{\thefootnote}{\arabic{footnote}}
-      % count the number of bibcite{name}{value} in paper-seb-main.aux
+      % count the number of bibcite{name}{value} in paper.aux
       \let\oldbibcite=\bibcite
@@
       \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux 
+      % expand paper.aux 
+\relax 
+\providecommand\hyper@newdestlabel[2]{}
+\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
+\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
+\global\let\oldnewlabel\newlabel
+\gdef\newlabel#1#2{\newlabelxx{#1}#2}
+\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
+\AtEndDocument{\ifx\hyper@anchor\@undefined
+\let\newlabel\oldnewlabel
+\fi}
+\fi}
+\global\let\hyper@last\relax 
+\gdef\HyperFirstAtBeginDocument#1{#1}
+\providecommand\HyField@AuxAddToFields[1]{}
+\providecommand\HyField@AuxAddToCoFields[2]{}
+\citation{a=b}
+\citation{write-for-life}
+\citation{abelson}
+\citation{nature-chatgpt}
+\newlabel{FirstPage}{{}{1}{}{Doc-Start}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {1}Introduction}{1}{section.1}\protected@file@percent }
+\citation{sommerville,knight}
+\citation{computational-thinking-wing,computational-thinking}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{reproducibility-crisis,rescience}
+\@writefile{toc}{\contentsline {section}{\numberline {2}Background}{2}{section.2}\protected@file@percent }
+\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Overview of peer-reviewed paper sample, broken down further in Table \ref {table-summary}.
+Survey methodology and data is provided in the Supplemental Material.
+(The survey does not include the motivating papers \cite  {ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref {section-pandemic-modeling}.)}}{3}{table.1}\protected@file@percent }
+\newlabel{table-overview}{{1}{3}{Overview of peer-reviewed paper sample, broken down further in Table \ref {table-summary}.
+Survey methodology and data is provided in the \supplement .
+(The survey does not include the motivating papers \cite {ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref {section-pandemic-modeling}.)}{table.1}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {2}{\ignorespaces Breakdown of pilot survey of peer-reviewed science papers relying on code. }}{3}{table.2}\protected@file@percent }
+\newlabel{table-summary}{{2}{3}{Breakdown of pilot survey of peer-reviewed science papers relying on code}{table.2}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {3}{\ignorespaces The TOP committee's recommended levels for journal article code transparency.
+Level 0 is provided for a comparison that does not meet any TOP requirements.
+Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref {reproducibility-concerns}.}}{3}{table.3}\protected@file@percent }
+\newlabel{TOPtable}{{3}{3}{The TOP committee's recommended levels for journal article code transparency.
+Level 0 is provided for a comparison that does not meet any TOP requirements.
+Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref {reproducibility-concerns}}{table.3}{}}
+\citation{temperature}
+\citation{form}
+\citation{acm-algorithms}
+\citation{hoare-correctness}
+\citation{jupyter-study}
+\citation{trisovic}
+\citation{jupyter-study}
+\citation{trisovic}
+\citation{trisovic}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.a}Code quality concerns}{4}{subsection.2.1}\protected@file@percent }
+\newlabel{related-work}{{2.a}{4}{Code quality concerns}{subsection.2.1}{}}
+\citation{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis}
+\citation{joss}
+\citation{rescience}
+\citation{TOP}
+\citation{TOP-wiki}
+\citation{jupyter-study}
+\citation{MMR-fraud}
+\citation{misconduct-issues}
+\citation{machine-learning-reproducibility}
+\citation{elns}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.b}Computable papers}{5}{subsection.2.2}\protected@file@percent }
+\newlabel{computable-papers}{{2.b}{5}{Computable papers}{subsection.2.2}{}}
+\citation{ICmodel}
+\citation{perkel-review}
+\citation{maneage}
+\citation{maneage}
+\citation{LP,LP-book}
+\citation{relit}
+\citation{notebook-patent}
+\citation{jupyter-history}
+\citation{knitr}
+\citation{spec-led-design}
+\citation{RAPstandard}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.c}RAP: Reproducible Analytical Pipelines}{7}{subsection.2.3}\protected@file@percent }
+\newlabel{RAP-section}{{2.c}{7}{RAP: Reproducible Analytical Pipelines}{subsection.2.3}{}}
+\citation{rap,goldacre,turing-way}
+\@writefile{lot}{\contentsline {table}{\numberline {4}{\ignorespaces A simplified schematic of the publication pipeline.
+For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
+The RAP and RAP{\tt  +}\ approaches encode the normally manual steps in the pipeline workflow so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The encoded RAP{\tt  +}\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
+Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the Supplemental Material.}}{8}{table.4}\protected@file@percent }
+\newlabel{fig-pipeline}{{4}{8}{A simplified schematic of the publication pipeline.
+For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
+The RAP and \RAPstar \ approaches encode the normally manual steps in the pipeline workflow so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The encoded \RAPstar \ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
+Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the \supplement }{table.4}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {5}{\ignorespaces A minimum standard of RAP, based on the UK Statistics Authority summary \cite  {RAPstandard}.}}{8}{table.5}\protected@file@percent }
+\newlabel{figure-RAP-summary}{{5}{8}{A minimum standard of RAP, based on the UK Statistics Authority summary \cite {RAPstandard}}{table.5}{}}
+\citation{MMR-fraud}
+\citation{a=b}
+\citation{Courtes}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{9}{subsubsection.2.3.1}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.2}RAP as research}{9}{subsubsection.2.3.2}\protected@file@percent }
+\citation{Courtes}
+\citation{Courtes}
+\citation{apa-statistics,Mariusz-reporting,science-statistics-recommendations}
+\citation{example-stats}
+\citation{science-statistics-recommendations}
+\citation{Spiegelhalter}
+\citation{paul-hci,pnas-stats}
+\citation{paul-hci,pnas-stats}
+\citation{Ben,se-bias}
+\@writefile{toc}{\contentsline {section}{\numberline {3}The statistics/code analogy}{10}{section.3}\protected@file@percent }
+\newlabel{statistics-computation-analogy}{{3}{10}{The statistics/code analogy}{section.3}{}}
+\citation{sas9.2}
+\citation{whitty}
+\newlabel{section-vicious-cycles}{{3}{11}{The statistics/code analogy}{section.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {3.a}The siren call of over-fitting code}{11}{subsection.3.1}\protected@file@percent }
+\newlabel{over-fit}{{3.a}{11}{The siren call of over-fitting code}{subsection.3.1}{}}
+\citation{over-fit}
+\citation{over-fit}
+\citation{may-simple}
+\citation{dyson}
+\newlabel{name-example-footnote}{{5}{12}{}{Hfootnote.5}{}}
+\citation{assurance-case}
+\@writefile{lof}{\contentsline {figure}{\numberline {1}{\ignorespaces Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
+Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
+The over-fitted polynomial model fits the sample \emph  {exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}}{13}{figure.1}\protected@file@percent }
+\newlabel{fig-overfit}{{1}{13}{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
+Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
+The over-fitted polynomial model fits the sample \emph {exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data}{figure.1}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {4}The conventional role of code}{13}{section.4}\protected@file@percent }
+\citation{assessing-quality}
+\citation{fixit}
+\citation{Hammurabi}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.a}The deceptive simplicity of code}{14}{subsection.4.1}\protected@file@percent }
+\newlabel{deceptive-simplicity-of-code}{{4.a}{14}{The deceptive simplicity of code}{subsection.4.1}{}}
+\citation{cite-code,TOP}
+\citation{prisma}
+\citation{engima}
+\citation{critiques,diagnosis-reviews}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.b}The low status of coding}{15}{subsection.4.2}\protected@file@percent }
+\newlabel{lowCodeStatus}{{4.b}{15}{The low status of coding}{subsection.4.2}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.c}The critical role of code is often ignored}{15}{subsection.4.3}\protected@file@percent }
+\newlabel{central-role-of-code}{{4.c}{15}{The critical role of code is often ignored}{subsection.4.3}{}}
+\newlabel{PRISMA-statement}{{4.c}{15}{The critical role of code is often ignored}{subsection.4.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{15}{subsection.4.4}\protected@file@percent }
+\newlabel{knowledge}{{4.d}{15}{Bugs, code and programming}{subsection.4.4}{}}
+\citation{knuth}
+\citation{essence-of-software,fixit}
+\citation{popper-conjectures-refutations}
+\citation{preventable-book}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}\protected@file@percent }
+\citation{preventable-book}
+\citation{science-review}
+\citation{diagnosis-reviews}
+\citation{tripod}
+\citation{diagnosis-reviews}
+\citation{diagnosis-reviews}
+\citation{nature-summary,ICmodel}
+\citation{avianFluModel,originalICmodel}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{tweet}
+\citation{ferguson-interview}
+\citation{basic-reproducibilty}
+\citation{ferguson-resign}
+\@writefile{toc}{\contentsline {section}{\numberline {5}State of the art}{17}{section.5}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.a}Case study: Pandemic modeling}{17}{subsection.5.1}\protected@file@percent }
+\newlabel{section-pandemic-modeling}{{5.a}{17}{Case study: Pandemic modeling}{subsection.5.1}{}}
+\citation{tweet}
+\citation{forsyth,Normal-Distributions}
+\citation{misra}
+\citation{spark}
+\citation{haskell}
+\citation{thumbs-up}
+\citation{nature-summary,ICmodel}
+\citation{codecheck,thumbs-up}
+\citation{codecheck}
+\citation{ICmodel}
+\citation{codecheck-certificate}
+\citation{codecheck}
+\citation{codecheck-certificate}
+\newlabel{discuss-covid-sim}{{5.a}{18}{Case study: Pandemic modeling}{subsection.5.1}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.b}Concerns with reproducibility}{18}{subsection.5.2}\protected@file@percent }
+\newlabel{reproducibility-concerns}{{5.b}{18}{Concerns with reproducibility}{subsection.5.2}{}}
+\citation{codecheck,thumbs-up}
+\citation{codecheck-certificate}
+\citation{NVP}
+\citation{nvp-ferguson}
+\citation{essence-of-software}
+\citation{notebooks,popper-conjectures-refutations,tripod,prisma,nih-policy,nih-nature}
+\citation{refactoring}
+\citation{bad-code}
+\citation{bad-code}
+\citation{pseudo}
+\citation{chinese}
+\citation{relit}
+\citation{flu-model}
+\citation{nature-review}
+\citation{nature-review}
+\citation{psychological-modeling}
+\citation{psychological-modeling}
+\citation{actr}
+\citation{psychological-modeling}
+\citation{Ben,nature-review,se-bias}
+\citation{ourReview}
+\citation{hamming}
+\citation{NVP}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.c}Beyond pandemic modeling}{20}{subsection.5.3}\protected@file@percent }
+\newlabel{section-science-beyond-pandemic-modeling}{{5.c}{20}{Beyond pandemic modeling}{subsection.5.3}{}}
+\citation{unfinished}
+\citation{unfinished,lancet-unfinished}
+\citation{hoare}
+\citation{machine-learning-reproducibility}
+\citation{relit}
+\citation{preclinical-reproducibility}
+\citation{fair,fair-principles}
+\citation{nih-policy}
+\citation{nih-nature}
+\newlabel{section-discussion}{{5.c}{21}{Beyond pandemic modeling}{subsection.5.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.d}The lop-sided emphasis on data}{21}{subsection.5.4}\protected@file@percent }
+\newlabel{critique-fair}{{5.d}{21}{The lop-sided emphasis on data}{subsection.5.4}{}}
+\citation{notebooks}
+\citation{cbc}
+\citation{ethics-code}
+\@writefile{toc}{\contentsline {section}{\numberline {6}Rethinking science that uses code}{22}{section.6}\protected@file@percent }
+\newlabel{summary}{{6}{22}{Rethinking science that uses code}{section.6}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.a}Software Engineering Boards}{22}{subsection.6.1}\protected@file@percent }
+\citation{acm-artifacts}
+\citation{redmill,iec61508}
+\citation{hoare-unreasonable}
+\citation{basic-reproducibilty,open-source}
+\citation{ABCs-SE}
+\citation{mrb}
+\citation{mrb}
+\citation{relit,Executable-Paper}
+\citation{computational-thinking-wing,computational-thinking}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.b}Extending RAP to RAP{\tt  +}}{23}{subsection.6.2}\protected@file@percent }
+\newlabel{RAPstar-section}{{6.b}{23}{Extending RAP to \RAPstar }{subsection.6.2}{}}
+\citation{RMarkdown}
+\citation{lepton}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.c}The paper as a scientific laboratory}{24}{subsection.6.3}\protected@file@percent }
+\newlabel{paper-as-lab}{{6.c}{24}{The paper as a scientific laboratory}{subsection.6.3}{}}
+\citation{landauer,thimbleby-landauer}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.d}Action must be interdisciplinary}{25}{subsection.6.4}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.e}Methodological statements}{25}{subsection.6.5}\protected@file@percent }
+\citation{no-raw-data}
+\citation{data-access}
+\citation{assurance-case}
+\citation{debt}
+\newlabel{mentionPLOSandIEEE}{{6.e}{26}{Methodological statements}{subsection.6.5}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{26}{subsection.6.6}\protected@file@percent }
+\newlabel{technical-debt}{{6.f}{26}{Training to reduce technical debt}{subsection.6.6}{}}
+\citation{fixit}
+\citation{jpm}
+\citation{jpm}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.g}Benefits beyond science}{27}{subsection.6.7}\protected@file@percent }
+\newlabel{benefits-beyond-science}{{6.g}{27}{Benefits beyond science}{subsection.6.7}{}}
+\citation{jpm-report}
+\citation{jpm}
+\citation{hoare-unreasonable}
+\citation{excel-fiasco}
+\citation{science-delays}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {7}Conclusions}{28}{section.7}\protected@file@percent }
+\citation{tweet,ferguson-interview}
+\citation{whitty}
+\citation{fixit}
+\citation{reproducibility,relit,popper-conjectures-refutations}
+\citation{parliamentary-evidence,my-parliamentary-evidence}
+\bibstyle{revisedCompJ}
+\bibdata{paper.bib}
+\bibcite{a=b}{1}
+\bibcite{write-for-life}{2}
+\bibcite{abelson}{3}
+\bibcite{nature-chatgpt}{4}
+\bibcite{sommerville}{5}
+\bibcite{knight}{6}
+\bibcite{computational-thinking-wing}{7}
+\bibcite{computational-thinking}{8}
+\bibcite{ICmodel}{9}
+\bibcite{avianFluModel}{10}
+\bibcite{originalICmodel}{11}
+\bibcite{reproducibility-crisis}{12}
+\bibcite{rescience}{13}
+\bibcite{temperature}{14}
+\bibcite{form}{15}
+\bibcite{acm-algorithms}{16}
+\bibcite{hoare-correctness}{17}
+\bibcite{jupyter-study}{18}
+\bibcite{trisovic}{19}
+\bibcite{machine-learning-reproducibility}{20}
+\bibcite{no-raw-data}{21}
+\bibcite{data-access}{22}
+\bibcite{reproducible-manifesto}{23}
+\bibcite{joss}{24}
+\bibcite{TOP}{25}
+\bibcite{TOP-wiki}{26}
+\bibcite{MMR-fraud}{27}
+\bibcite{misconduct-issues}{28}
+\bibcite{elns}{29}
+\bibcite{perkel-review}{30}
+\bibcite{maneage}{31}
+\bibcite{LP}{32}
+\bibcite{LP-book}{33}
+\bibcite{relit}{34}
+\bibcite{notebook-patent}{35}
+\bibcite{jupyter-history}{36}
+\bibcite{knitr}{37}
+\bibcite{spec-led-design}{38}
+\bibcite{RAPstandard}{39}
+\bibcite{rap}{40}
+\bibcite{goldacre}{41}
+\bibcite{turing-way}{42}
+\bibcite{Courtes}{43}
+\bibcite{apa-statistics}{44}
+\bibcite{Mariusz-reporting}{45}
+\bibcite{science-statistics-recommendations}{46}
+\bibcite{example-stats}{47}
+\bibcite{Spiegelhalter}{48}
+\bibcite{paul-hci}{49}
+\bibcite{pnas-stats}{50}
+\bibcite{Ben}{51}
+\bibcite{se-bias}{52}
+\bibcite{sas9.2}{53}
+\bibcite{whitty}{54}
+\bibcite{over-fit}{55}
+\bibcite{may-simple}{56}
+\bibcite{dyson}{57}
+\bibcite{assurance-case}{58}
+\bibcite{assessing-quality}{59}
+\citation{assessing-quality-conference}
+\bibcite{fixit}{60}
+\bibcite{Hammurabi}{61}
+\bibcite{cite-code}{62}
+\bibcite{prisma}{63}
+\bibcite{engima}{64}
+\bibcite{critiques}{65}
+\bibcite{diagnosis-reviews}{66}
+\bibcite{knuth}{67}
+\bibcite{essence-of-software}{68}
+\bibcite{popper-conjectures-refutations}{69}
+\bibcite{preventable-book}{70}
+\bibcite{science-review}{71}
+\bibcite{tripod}{72}
+\bibcite{nature-summary}{73}
+\bibcite{tweet}{74}
+\bibcite{ferguson-interview}{75}
+\bibcite{basic-reproducibilty}{76}
+\bibcite{ferguson-resign}{77}
+\bibcite{forsyth}{78}
+\bibcite{Normal-Distributions}{79}
+\bibcite{misra}{80}
+\bibcite{spark}{81}
+\bibcite{haskell}{82}
+\bibcite{thumbs-up}{83}
+\bibcite{codecheck}{84}
+\bibcite{codecheck-certificate}{85}
+\bibcite{NVP}{86}
+\bibcite{nvp-ferguson}{87}
+\bibcite{notebooks}{88}
+\bibcite{nih-policy}{89}
+\bibcite{nih-nature}{90}
+\bibcite{refactoring}{91}
+\bibcite{bad-code}{92}
+\bibcite{pseudo}{93}
+\bibcite{chinese}{94}
+\bibcite{flu-model}{95}
+\bibcite{nature-review}{96}
+\bibcite{psychological-modeling}{97}
+\bibcite{actr}{98}
+\bibcite{ourReview}{99}
+\bibcite{hamming}{100}
+\bibcite{unfinished}{101}
+\bibcite{lancet-unfinished}{102}
+\bibcite{hoare}{103}
+\bibcite{preclinical-reproducibility}{104}
+\bibcite{fair}{105}
+\bibcite{fair-principles}{106}
+\bibcite{cbc}{107}
+\bibcite{ethics-code}{108}
+\bibcite{acm-artifacts}{109}
+\bibcite{redmill}{110}
+\bibcite{iec61508}{111}
+\bibcite{hoare-unreasonable}{112}
+\bibcite{open-source}{113}
+\bibcite{ABCs-SE}{114}
+\bibcite{mrb}{115}
+\bibcite{Executable-Paper}{116}
+\bibcite{RMarkdown}{117}
+\bibcite{lepton}{118}
+\bibcite{landauer}{119}
+\bibcite{thimbleby-landauer}{120}
+\bibcite{debt}{121}
+\bibcite{jpm}{122}
+\bibcite{jpm-report}{123}
+\bibcite{excel-fiasco}{124}
+\bibcite{science-delays}{125}
+\bibcite{reproducibility}{126}
+\bibcite{parliamentary-evidence}{127}
+\bibcite{my-parliamentary-evidence}{128}
+\citation{parliamentary-evidence}
+\bibcite{assessing-quality-conference}{129}
+\def\continueSectionNumbers{7}
+\def\continueTableNumbers{5}
+\def\continueFigureNumbers{1}
+\def\continueFootnoteNumbers{13}
+\def\continuePageNumbers{35}
+\newlabel{LastPage}{{}{35}}
+\gdef \@abspage@last{35}
+% end expanding paper.aux 
       \let\bibcite=\oldbibcite
@@
 
-\input generated/page-lengths.tex
+% expand generated/page-lengths.tex
+% Generated by programs/data.js
+\expandafter\def\csname pagelength-MetricSelectionFramework\endcsname{17}
+\expandafter\def\csname pagelength-PENet\endcsname{9}
+\expandafter\def\csname pagelength-PostoperativeOutcomes.RiskNet\endcsname{10}
+\expandafter\def\csname pagelength-philter-ucsf\endcsname{8}
+\expandafter\def\csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname{6}
+\expandafter\def\csname pagelength-SiameseChange\endcsname{9}
+\expandafter\def\csname pagelength-lactModel\endcsname{13}
+\expandafter\def\csname pagelength-LRM\endcsname{22}
+\expandafter\def\csname pagelength-manifold-ga\endcsname{7}
+\expandafter\def\csname pagelength-blast-ct\endcsname{8}
+\newcount \gitPages \gitPages=109
+\newcount \totalPages \totalPages=341
+% end expanding generated/page-lengths.tex
 
@@
 
-\input generated/info-for-supplement.tex
+% expand generated/info-for-supplement.tex
+% map Git repo names to their citation numbers
+% generated by appendix.tex
+\expandafter\def \csname cite-MetricSelectionFramework\endcsname{[176]}
+\expandafter\def \csname cite-PENet\endcsname{[180]}
+\expandafter\def \csname cite-PostoperativeOutcomes.RiskNet\endcsname{[182]}
+\expandafter\def \csname cite-philter-ucsf\endcsname{[183]}
+\expandafter\def \csname cite-AI-CDSS-Cardiovascular-Silo\endcsname{[184]}
+\expandafter\def \csname cite-SiameseChange\endcsname{[186]}
+\expandafter\def \csname cite-dryad.1g1jwstrw\endcsname{[187]}
+\expandafter\def \csname cite-rsos.192210\endcsname{[188]}
+\expandafter\def \csname cite-ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ\endcsname{[189]}
+\expandafter\def \csname cite-rsos.200566\endcsname{[190]}
+\expandafter\def \csname cite-?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c\endcsname{[191]}
+\expandafter\def \csname cite-dryad.vx0k6djnr\endcsname{[194]}
+\expandafter\def \csname cite-lactModel\endcsname{[195]}
+\expandafter\def \csname cite-LRM\endcsname{[197]}
+\expandafter\def \csname cite-1\endcsname{[198]}
+\expandafter\def \csname cite-manifold-ga\endcsname{[203]}
+\expandafter\def \csname cite-blast-ct\endcsname{[205]}
+% end expanding generated/info-for-supplement.tex
 \newwrite\infofileForSupplement
@@
 
-\tableofcontents
+% expand table of contents appendix.toc
+\makeatletter
+  \section*{\contentsname
+    \@mkboth{\MakeUppercase\contentsname}{\MakeUppercase\contentsname}}
+\makeatother\contentsline {section}{\numberline {8}Further issues for Software Engineering Boards (SEBs)}{38}{section.8}%
+\contentsline {subsection}{\numberline {8.a}Relationships of SEBs to Ethics Boards}{38}{subsection.8.1}%
+\contentsline {subsection}{\numberline {8.b}SEBs are necessary but not sufficient}{39}{subsection.8.2}%
+\contentsline {section}{\numberline {9}Software engineering best practice}{39}{section.9}%
+\contentsline {subsection}{\numberline {9.a}Requirements}{39}{subsection.9.1}%
+\contentsline {subsection}{\numberline {9.b}Formal methods}{40}{subsection.9.2}%
+\contentsline {subsection}{\numberline {9.c}Defensive programming}{40}{subsection.9.3}%
+\contentsline {subsection}{\numberline {9.d}Using inappropriate programming languages undermines reliability}{41}{subsection.9.4}%
+\contentsline {subsection}{\numberline {9.e}Open source and version control}{41}{subsection.9.5}%
+\contentsline {subsection}{\numberline {9.f}Rigorous testing}{42}{subsection.9.6}%
+\contentsline {subsection}{\numberline {9.g}Good documentation and record keeping}{42}{subsection.9.7}%
+\contentsline {subsection}{\numberline {9.h}Usability}{43}{subsection.9.8}%
+\contentsline {subsection}{\numberline {9.i}Reusing quality solutions}{43}{subsection.9.9}%
+\contentsline {subsection}{\numberline {9.j}Simplicity}{43}{subsection.9.10}%
+\contentsline {subsection}{\numberline {9.k}Compliance with standards}{43}{subsection.9.11}%
+\contentsline {subsection}{\numberline {9.l}Effective multidisciplinary teamwork}{44}{subsection.9.12}%
+\contentsline {subsection}{\numberline {9.m}Continuous Professional Development (CPD)}{44}{subsection.9.13}%
+\contentsline {subsection}{\numberline {9.n}Security and other factors}{44}{subsection.9.14}%
+\contentsline {subsection}{\numberline {9.o}Software is a human activity}{44}{subsection.9.15}%
+\contentsline {section}{\numberline {10}Code, data, and publication}{45}{section.10}%
+\contentsline {subsection}{\numberline {10.a}When magic numbers become magic code}{46}{subsection.10.1}%
+\contentsline {subsection}{\numberline {10.b}When data is code}{47}{subsection.10.2}%
+\contentsline {subsection}{\numberline {10.c}Exploiting code as data for more reliable science}{47}{subsection.10.3}%
+\contentsline {subsection}{\numberline {10.d}When data is text: Exploiting code for reliable publication}{47}{subsection.10.4}%
+\contentsline {subsection}{\numberline {10.e}Data and polynomials used in the paper}{48}{subsection.10.5}%
+\contentsline {subsection}{\numberline {10.f}Comparing conventional and RAP approaches}{48}{subsection.10.6}%
+\contentsline {section}{\numberline {11}The Spiegelhalter trustworthiness questions}{48}{section.11}%
+\contentsline {subsection}{\numberline {11.a}How trustworthy are the numbers?}{50}{subsection.11.1}%
+\contentsline {subsection}{\numberline {11.b}How trustworthy is the source?}{51}{subsection.11.2}%
+\contentsline {subsection}{\numberline {11.c}How trustworthy is the interpretation?}{52}{subsection.11.3}%
+\contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}%
+\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{52}{subsection.12.1}%
+\contentsline {subsubsection}{\numberline {12.a.1}\emph {The Lancet\/}}{52}{subsubsection.12.1.1}%
+\contentsline {subsubsection}{\numberline {12.a.2}\emph {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}%
+\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{54}{subsection.12.2}%
+\contentsline {subsection}{\numberline {12.c}Summary of results}{56}{subsection.12.3}%
+\contentsline {subsubsection}{\numberline {12.c.1}Current code policies of sampled journals}{57}{subsubsection.12.3.1}%
+\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{58}{subsubsection.12.3.2}%
+\contentsline {subsection}{\numberline {12.d}Assessment criteria and methods}{60}{subsection.12.4}%
+\contentsline {subsubsection}{\numberline {12.d.1}Detecting and defending against error}{62}{subsubsection.12.4.1}%
+\contentsline {subsubsection}{\numberline {12.d.2}Defending against system problems}{63}{subsubsection.12.4.2}%
+\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{64}{subsubsection.12.4.3}%
+\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}%
+\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}%
+
+% end expanding table of contents appendix.toc
+
 \newpage
@@
 
-\input generated/over-fitting-data.tex
+% expand generated/over-fitting-data.tex
+\begin{center}\sf\begin{tabular}{|c|r|r|r|r|r|}\hline
+$x={}$&1.00&1.10&2.20&2.60&4.50\\
+$y={}$&1.50&2.70&4.90&5.70&8.20\\
+\hline\end{tabular}
+\end{center}
+
+Showing both polynomials with coefficients to 2 decimal places, the linear least squares model to fit this data is:
+
+$$
+\hat{y}=0.49+1.80x
+$$
+
+and the Lagrange polynomial model (an exact fit to the 5 data points) is
+
+$$
+\hat{y}=-41.79+85.46x-56.30x^2+15.65x^3-1.51x^4
+$$
+
+These polynomials are the ones shown in the main paper's figure \ref{fig-overfit}.
+
+The name of the relevant \emph{Mathematica\/} notebook file where all data and code for this section (and for generating the paper's section \ref{over-fit} and its figure \ref{fig-overfit}) can be found in \texttt{programs/over-fitting-section.nb}, which is included in the Git repository for the paper.
+% end expanding generated/over-fitting-data.tex
 
@@
 \begin{center}
-\input generated/make-help.tex
+% expand generated/make-help.tex
+% Edit makefile to change the selected make command choices, as follows:
+% 1 - make all
+% 2 - make archive
+% 3 - make check-same
+% 4 - make check-update
+% 5 - make check-versions
+% 6 - make data
+% 7 - make expand
+% 8 - make git-prep
+% 9 - make help
+% 10 - make help-brief
+% 11 - make mathematica
+% 12 - make mathematica-open
+% 13 - make one-file
+% 14 - make pdf
+% 15 - make push
+% 16 - make readme
+% 17 - make really-tidyup
+% 18 - make tidyup
+% 19 - make zip
+% 20 - make zip-data
+
+{\sf\begin{tabular}{rp{4.5in}}
+
+\texttt{make all}&
+      Analyze the data, then typeset the main PDF files (\texttt{paper-seb-main.pdf} and \texttt{paper-seb-supplementary-material.pdf}).\\
+   \multicolumn{1}{l@{\vdots}}{}&\\
+
+\texttt{make data}&
+      Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc.
+This \texttt{make} option runs \texttt{node programs/data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
+Note that downloading all the repositories in a reasonable time needs decent internet bandwidth.\\
+   \multicolumn{1}{l@{\vdots}}{}&\\
+
+\texttt{make git-prep}&
+      What's on Git that we've lost, or stuff we have got locally but probably don't want on Git, so you can delete it or move it out the way or whatever.\\
+   \multicolumn{1}{l@{\vdots}}{}&\\
+
+\texttt{make really-tidyup}&
+      More thorough than \texttt{make tidyup} --- remove \emph{all} files that can be recreated.
+(This will mean next time you run \LaTeX\ you will have to ignore errors as the .aux files are re-created.)\\
+\end{tabular}}
+
+% end expanding generated/make-help.tex
 \end{center}
@@
 \begin{center}\normalsize
-\input generated/summary-table.tex
+% expand generated/summary-table.tex
+% Generated by programs/data.js
+\begin{tabular}{|rrrc|}\hline
+Number of papers sampled relying on code&32&100\%&\\\hline\hline
+\multicolumn{4}{|l|}{\textbf{Access to code}}\\
+Some or all code available&12&38\%&\\
+Some or all code in principle available on request&8&25\%&\\
+Requested code actually made available (within 2 years 11 months${}^{\star}$)&0&0\%& \\\hline
+ \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
+Evidence program designed rigorously&{0}&{0\%}&\\
+Evidence source code properly tested&{0}&{0\%}&\\
+Evidence of any tool-based development&{0}&{0\%}&\\
+Team or open source based development&{0}&{0\%}&\\
+Other methods, e.g., independent coding methods&{1}&{3\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
+Substantial code documentation and comments&2&6\%&\\
+Comments explain some code intent&3&9\%&\\
+Procedural comments (e.g., author, date, copyright)&10&31\%&\\
+No usable comments&17&53\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Repository use}}\\
+Used code repository (e.g., GitHub)&9&28\%&\\
+Used data repository (e.g., Dryad or GitHub)&9&28\%&\\
+Empty repository&1&3\%&\\\hline
+\multicolumn{4}{|l|}{{\textbf{Evidence of documented processes}}}\\
+Evidence of RAP/\RAPstar\ or any other principles in use to support scrutiny&{0}&{0\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
+Papers published in journals with code policies&26&81\%&\\
+Clear breaches of journal code policy (if any) & 11&42\%&(\emph{N}~=~26)\\
+\hline\end{tabular}\\
+\vskip 1ex ${}^{\star}$Time of 2 years 11 months is wait between code request and date of generating this table.% end expanding generated/summary-table.tex
 \end{center}
@@
 
-\input generated/repos.tex
+% expand generated/repos.tex
+\begin{tabular}{|l|r|rrr@{\hskip .75ex}l|}\hline
+&\multicolumn{1}{c|}{\bf PDF paper}&\multicolumn{4}{c|}{\bf Repository code \& data}\\ \cline{2-6}
+\multicolumn{1}{|c|}{\bf Github repository}&\multicolumn{1}{c|}{\bf Number}&\multicolumn{1}{c}{\bf Number}&\multicolumn{1}{c}{\bf Code}&\multicolumn{2}{c|}{\bf Data}\\
+\multicolumn{1}{|c|}{\bf and paper citation}&\multicolumn{1}{c|}{\bf of pages}&\multicolumn{1}{c}{\bf of files}&\multicolumn{1}{c}{\bf kLOC}&\multicolumn{2}{c|}{\bf bytes} \\ \hline\hline
+\reponame{AI-CDSS-Cardiovascular-Silo}{AI-CDSS-Cardiovascular-Silo} & \csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname & \digitBlock{206} & 143 & 64&Mb \\
+\reponame{blast-ct}{blast-ct} & \csname pagelength-blast-ct\endcsname & \digitBlock{54} & 3 & 238&Mb \\
+\reponame{covid-sim}{covid-sim} & \csname pagelength-covid-sim\endcsname & \digitBlock{229} & 25 & 734&Mb \\
+\reponame{lactModel}{lactModel} & \csname pagelength-lactModel\endcsname & \digitBlock{20} & 2 & 165&kb \\
+\reponame{LRM}{LRM} & \csname pagelength-LRM\endcsname & \digitBlock{125} & 8 & 2&Mb \\
+\reponame{manifold-ga}{manifold-ga} & \csname pagelength-manifold-ga\endcsname & \digitBlock{11} & 1 & \multicolumn{2}{c|}{---} \\
+\reponame{MetricSelectionFramework}{MetricSelectionFramework} & \csname pagelength-MetricSelectionFramework\endcsname & \digitBlock{44} & 4 & 236&kb \\
+\reponame{PENet}{PENet} & \csname pagelength-PENet\endcsname & \digitBlock{117} & 8 & 4&Mb \\
+\reponame{philter-ucsf}{philter-ucsf} & \csname pagelength-philter-ucsf\endcsname & \digitBlock{1987} & 13 & 32&Mb \\
+\reponame{PostoperativeOutcomes\_RiskNet}{PostoperativeOutcomes.RiskNet} & \csname pagelength-PostoperativeOutcomes.RiskNet\endcsname & \digitBlock{1} & --- & \multicolumn{2}{c|}{---} \\
+\reponame{SiameseChange}{SiameseChange} & \csname pagelength-SiameseChange\endcsname & \digitBlock{5} & 1 & 1&kb \\
+% Numbers below are rounded, but more precise averages are 254.454545 -- 18824.000000 -- 97771458.363636 = 1075486042 / 11
+\hline \multicolumn{1}{|r|}{{\bf Average} ($N=11$)}&\makeAverage{\the\gitPages}{11}&254& 19& 98&Mb\\
+\hline \end{tabular}
+% end expanding generated/repos.tex
 \vskip 1ex
@@
 \sf
-\input generated/legend.tex
+% expand generated/legend.tex
+% Generated by programs/data.js
+\begin{tabular}{lrp{4.5in}}
+&\bfseries{Paper}&\\\bfseries{Flag}&\bfseries{count}&\bfseries{Meaning}\\ \hline 
+\flagStyle{$\sf P_c$}&26&Journal has a code policy (see section~\ref{supplementary-journal-policies-section})\\ 
+\flagStyle{$\sf P_{{\mbox{\scriptsize c-breach}}}$}&11&Paper breaches journal code policy (see section~\ref{supplementary-journal-policies-section})\\ 
+\flagStyle{$\sf R_c$}&10&Paper uses a code repository (e.g., GitHub)\\ 
+\flagStyle{$\sf R_{\mbox{\scriptsize c-empty}}$}&1&Code repository contains no code\\ 
+\flagStyle{$\sf R_d$}&9&Paper uses a data repository (e.g., Dryad, Figshare, GitHub)\\ 
+\flagStyle{$\sf S_{\mbox{\scriptsize NONE}}$}&12&No code available at all (note: code is not expected for standard models, systems or statistical methods)\\ 
+\flagStyle{$\sf S_p$}&8&Paper says source code is available in principle\\ 
+\flagStyle{$\sf S_{+}$}&12&Paper or URL provides source code\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize rigorous}}}$}&0&Evidence that source code was developed rigorously\\ 
+\flagStyle{$\sf S_{\mbox{\scriptsize tested}}$}&0&Evidence that source code has been run with a clean build and tested\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize tools}}}$}&0&Evidence of any tool-based development\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize open source}}}$}&0&Team or open source development\\ 
+\flagStyle{$\sf S_{{\mbox{\scriptsize otherSE}}}$}&1&Other evidence of good practice; see details in summary table\\ 
+\flagStyle{$\sf C_0$}&10&Code only has comments unrelated to code intent (e.g., copyright)\\ 
+\flagStyle{$\sf C_1$}&1&Code only has trivial or obvious comments\\ 
+\flagStyle{$\sf C_2$}&3&Helpful comments explaining code intent, rather than rephrasing the code\\ 
+\flagStyle{$\sf C_3$}&2&Code has substantial, useful comments, and documentation\\ 
+\flagStyle{$\sf C_a$}&1&Code in repository uses assertions\\ 
+\flagStyle{$\sf C_c$}&0&Code in repository uses pre- or post-conditions or similar\\ 
+\end{tabular}
+% end expanding generated/legend.tex
 \end{center}
@@
 \sf\bfseries Ref&\sf\bfseries Data&\sf\bfseries Code \\ \hline \endhead 
-\input generated/assessments.tex
+% expand generated/assessments.tex
+% Generated by programs/data.js
+\cite{ref-1} & On request.\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-2} & ``The datasets used in the current study are available from the corresponding author upon reasonable request and under consideration of the ethical regulations.''\flagStyle{ $\sf R_d$\hskip 3pt{}} & Matlab.
+Documented overview, but only trivial comments.
+Assertions in code only relate to UI.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}$\sf C_a$\hskip 3pt{}}\\
+\cite{ref-3} & ``In accordance with Twitter policies of data sharing, data used in the generation of the algorithm for this study will not be made publicly available.''\flagStyle{ } & ``Due to the sensitive and potentially stigmatizing nature of this tool, code used for algorithm generation or implementation on individual Twitter profiles will not be made publicly available.''\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-4} & ``The datasets generated during and/or analyzed during the current study are available from the corresponding author on reasonable request.''\flagStyle{ } & ``This code would be made available upon reasonable request.'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-5} & Nothing available\flagStyle{ } & Nothing available (despite building two voice-based virtual counselors).\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-6} & ``The datasets generated and analyzed during the study are not currently publicly available due to HIPAA compliance agreement but are available from the corresponding author on reasonable request.''\flagStyle{ } & Poor commenting, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-7} & ``The dataset generated and analyzed for this study will not be made publicly available due to patient privacy and lack of informed consent to allow sharing of patient data outside of the research team.''\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-8} & ``The datasets generated during and/or analyzed during the current study are not publicly available due to institutional restrictions on data sharing and privacy concerns.
+However, the data are available from the corresponding author on reasonable request.''\flagStyle{ } & Empty GitHub repository: ``Code coming soon \ldots'' paper says.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf R_{\mbox{\scriptsize c-empty}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-9} & ``The i2b2 data that support the findings of this study are available from i2b2 but restrictions apply to the availability of these data, which require signed safe usage and research-only.
+Data from UCSF are not available at this time as they have not been legally certified as being De-Identified, however, this process is underway and the data may be available by the time of publication by contacting the authors.
+Requesters identity as researchers will need to be confirmed, safe usage guarantees will need to be signed, and other restrictions may apply.''\flagStyle{ } & Basic documentation, very little comment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-10} & ``Not available due to restrictions in the ethical permit, but may be available on request.''\flagStyle{ } & Trivial comments, no documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-11} & ``The data that support the findings of this study are available in a deidentified form from Cleveland Clinic, but restrictions apply to the availability of these data, which were used under Cleveland Clinic data policies for the current study, and so are not publicly available.''\flagStyle{ } & ``We used only free and open-source software'' --- some of which is unspecified.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-12} & ``The i-ROP cohort study data for ROP is not publicly available due to patient privacy restrictions, though potential collaborators are directed to contact the study investigators \ldots''\flagStyle{ } & Not all code on GitHub, minor comments only.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{paper-usesRMarkdown} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\cite{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
+Currently (apparently) for private view.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code available for private view, though some code available with minor comments.
+Paper describes using two contrasting methods to help confirm correctness, ``As an additional check, I also coded the calculation of D based on a probabilistic approach, using genotype frequencies in each population to calculate the expected frequencies of each possible two-genotype combination (electronic supplementary material, table S1).
+Essentially identical results were obtained.'' --- but the contrasting method is not available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf S_{{\mbox{\scriptsize otherSE}}}$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\cite{ref-16} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Reasonaby commented code on Dryad, but code is not complete and presumably never checked.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\cite{ref-17} & On request.\flagStyle{ } & R, lightly commented.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ethics-paper} & No data required.\flagStyle{ } & Unrunnable incomplete code fragment.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\
+\cite{ref-19} & Data embedded in PDF.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-20} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Some comments, some code in Matlab.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}$\sf C_2$\hskip 3pt{}}\\
+\cite{ref-21} & Partial data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Documented R, including model's manual.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
+\cite{ref-22} & No data required.\flagStyle{ } & ``We constructed a bioeconomic model for an RSSF [restricted fishing effort small-scale fishery] using game theory'' for which results are discussed, yet no code is available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-23} & Data cited, not all available.\flagStyle{ } & Trivial documentation.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{example-numerical-error} & On Figshare.\flagStyle{ $\sf R_d$\hskip 3pt{}} & On Figshare, large amount of disorganised and undocumented code.
+Helpful features to make usable for third parties.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-25} & Data on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-26} & Data on various web sites.\flagStyle{ } & No code available.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-27} & Data on request.\flagStyle{ } & ``The coding used to train the artificial intelligence model are dependent on annotation, infrastructure, and hardware, so cannot be released.'' (!) Algorithm (not source code) available on request.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-28} & Data on request.\flagStyle{ } & Python scripts can be requested.\flagStyle{ $\sf S_p$\hskip 3pt{}}\\
+\cite{ref-29} & Unspecified location on large website requiring registration.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Has overall documentation but poorly commented Matlab code on GitHub.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-30} & Available to researchers who meet criteria for access to confidential data.\flagStyle{ } & Despite the paper being a ``deep learning algorithm'' the code is not available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+\cite{ref-31} & Data access conditional on approved study proposal.\flagStyle{ } & Almost completely uncommented Python, but does have a basic setup script.\flagStyle{ $\sf R_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_1$\hskip 3pt{}}\\
+\cite{ref-32} & Unspecified locations on several large websites.\flagStyle{ } & Python used and apparently on GitHub, but --- an oversight?
+--- no code is available.\flagStyle{ $\sf S_{\mbox{\scriptsize NONE}}$\hskip 3pt{}}\\
+% end expanding generated/assessments.tex
 \end{longtable}
@@
 \vbox{%
-  \input generated/example-data.tex
+  % expand generated/example-data.tex
+% Generated by programs/data.js
+\noindent\texttt{
+\hspace*{\codeIndent}\{\hspace*{\codeIndent}\newline
+\hspace*{2\codeIndent} accessed: "14 July 2020",\newline
+\hspace*{2\codeIndent} doubleChecked: "17 January 2021",\newline
+\hspace*{2\codeIndent} authors: "Callahan A, Steinberg E, Fries JA, Gomba \ldots,\newline
+\hspace*{2\codeIndent} year: 2020,\newline
+\hspace*{2\codeIndent} title: "Estimating the efficacy of symptom-based \ldots,\newline
+\hspace*{2\codeIndent} volume: 3,\newline
+\hspace*{2\codeIndent} number: 95,\newline
+\hspace*{2\codeIndent} journal: "Nature Digital Medicine",\newline
+\hspace*{2\codeIndent} doi: "10.1038/s41746-020-0300-0",\newline
+\hspace*{2\codeIndent} dataComment: "On request.",\newline
+\hspace*{2\codeIndent} hasCodeInPrinciple: 1,\newline
+\hspace*{2\codeIndent} codeComment: "``Code is available upon request from th \ldots,\newline
+\hspace*{2\codeIndent} pages: 3\newline
+\hspace*{\codeIndent}\}
+}
+% end expanding generated/example-data.tex
 }
@@
 \def\endofbibliography{ \egroup }
-  \bibliography{paper-seb-supplementary-material.bib}
+  % expand bibliography appendix.bbl
+\begin{thebibliography}{10}
+
+\bibitem{vancouver}
+{Vancouver Group} (1997) ``Uniform requirements for manuscripts submitted to
+  biomedical journals,'' {\em JAMA}, \textbf{277}(11):927--934.
+\newblock \doi{10.1001/jama.1997.03540350077040}
+
+\bibitem{cosmos}
+Stepney, S. \emph{et al.} (2018) {\em Engineering Simulations as Scientific
+  Instruments: A Pattern Language}, Springer.
+
+\bibitem{ai-ethics}
+Misselhorn, C. (2018) ``Artificial morality. {Concepts}, issues and
+  challenges,'' {\em Social Science and Public Policy}, \textbf{55}:161--169.
+\newblock \doi{10.1007/s12115-018-0229-y}
+
+\bibitem{science-lancet1}
+Servick, K. and Enserink, M. (2020) ``A mysterious company's coronavirus papers
+  in top medical journals may be unraveling,'' {\em Science}.
+\newblock \doi{10.1126/science.abd1337}
+
+\bibitem{science-lancet2}
+Servick, K. (2020) ``{COVID}-19 data scandal prompts tweaks to elite journal's
+  review process,'' {\em Science}.
+\newblock \doi{10.1126/science.abe8656}
+
+\bibitem{lancet-learning}
+{The Editors} (2020) ``Learning from a retraction,'' {\em The Lancet},
+  \textbf{396}:799.
+\newblock \doi{10.1016/S0140-6736(20)31958-9}
+
+\bibitem{jvs1}
+Hardouin, S., Cheng, T.~W., Mitchell, E.~L., Raulli, S.~J., Jones, D.~W.,
+  Siracuse, J.~J. and Farber, A. (2020) ``Prevalence of unprofessional social
+  media content among young vascular surgeons,'' {\em Journal of Vascular
+  Surgery}, \textbf{72}(2):667--671.
+\newblock \doi{10.1016/j.jvs.2019.10.069}
+
+\bibitem{jvs2}
+Baumann, J. (29 July, 2020) ``\#{MedBikini} backlash exposes research ethics
+  boards' digital gaps,'' {\em Bloomberg Law}.
+\newblock
+  \url{news.bloomberglaw.com/}\midurl{pharma-}\midurl{and-}\midurl{life-}\midurl{sciences/}\midurl{medbikini-}\midurl{backlash-}\midurl{exposes-}\midurl{research-}\midurl{ethics-}\midurl{boards-}\midurl{digital-}\midurl{gaps}
+
+\bibitem{jvs3}
+{The Editors {(of \emph{Journal of Vascular Surgery\/})}} (2020) ``{Editors'
+  Statement Regarding ``Prevalence of unprofessional social media content among
+  young vascular surgeons''},'' {\em FaceBook}.
+\newblock
+  \url{www.facebook.com/TheJVascSurg/photos/a.611986142331744/1381024778761206/?type=3&theater},
+  and copied to journal home page \url{www.jvascsurg.org} Accessed 30 July
+  2020.
+
+\bibitem{swebok}
+Bourque, P. and Fairley, R., eds. (2014) {\em Guide to the Software Engineering
+  Body of Knowledge}, (Version 3.0), IEEE Computer Society.
+\newblock \url{www.swebok.org}
+
+\bibitem{wpf}
+Zeller, A. (2006) {\em Why Programs Fail: A Guide to Systematic Debugging},
+  Morgan Kaufmann.
+
+\bibitem{humphrey}
+Humphrey, W.~S. (2005) {\em PSP: A Self-Improvement Process for Software
+  Engineers}, Addison Wesley.
+
+\bibitem{managers}
+\leavevmode\vrule height 2pt depth -1.6pt width 23pt\  (2001) {\em Winning with
+  Software: An Executive Strategy}, Addison-Wesley Professional.
+
+\bibitem{differential-equations}
+Boldo, S., Cl{\'{e}}ment, F., Filli{\^{a}}tre, J.-C., Mayero, M., Melquiond, G.
+  and Weis, P. (2014) ``Trusting computations: {A} mechanized proof from
+  partial differential equations to actual program,'' {\em Computers and
+  Mathematics with Applications}, \textbf{68}:325--352.
+\newblock \doi{10.1016/j.camwa.2014.06.004}
+
+\bibitem{numerals}
+Thimbleby, H. and Cairns, P. (2017) ``Interactive numerals,'' {\em Royal
+  Society Open Science}, \textbf{4}(4):160903.
+\newblock \doi{10.1098/rsos.160903}
+
+\bibitem{heedless}
+Thimbleby, H. (2012) ``Heedless programming: Ignoring detectable error is a
+  widespread hazard,'' {\em Software --- Practice {\&} Experience},
+  \textbf{42}(11):1393--1407.
+\newblock \doi{10.1002/spe.1141}
+
+\bibitem{plos}
+Chao, D.~L., Halloran, M.~E., Obenchain, V.~J. and {Longini Jr}, I.~M. (2010)
+  ``{FluTE}, a publicly available stochastic influenza epidemic simulation
+  model,'' {\em PLOS Computational Biology}, \textbf{6}(1):e1000656.
+\newblock Source code available at \url{GitHub.com/dlchao/FluTE},
+  \doi{10.1371/journal.pcbi.1000656}
+
+\bibitem{excel1}
+Abir, M., Nelson, C., Chan, E.~W., Al-Ibrahim, H., Cutter, C., Patel, K. and
+  Bogart, A. (2020) {\em {RAND} Critical Care Surge Response Tool: {An}
+  Excel-Based Model for Helping Hospitals Respond to the {COVID-19} Crisis},
+  RAND Corporation.
+\newblock
+  \url{www.rand.org/}\midurl{pubs/}\midurl{tools/}\midurl{TLA164-}\midurl{1.html}
+
+\bibitem{excel2}
+Alvarez, N.~M., Gonzalez-Gonzalez, E. and Trujillo-de Santiago, G. (2020)
+  ``Modeling {COVID-19} epidemics in an excel spreadsheet: {Democratizing} the
+  access to first-hand accurate predictions of epidemic outbreaks,'' {\em
+  MedRxiv}.
+\newblock Preprint, \doi{10.1101/2020.03.23.20041590}
+
+\bibitem{npm}
+Schlueter, I.~Z. (23 March 2016) {\em Blog: kik, left-pad, and npm}, NPM Blog.
+\newblock Accessed 10 April 2020,
+  \url{blog.npmjs.org/}\midurl{post/}\midurl{141577284765/}\midurl{kik-}\midurl{left-}\midurl{pad-}\midurl{and-}\midurl{npm}
+
+\bibitem{hci1}
+Shneiderman, B., Plaisant, C., Cohen, M. and {\emph{et al}} (2016) {\em
+  Designing the User Interface: Strategies for Effective Human-Computer
+  Interaction}, Pearson, 6th~ed.
+\newblock \url{www.cs.umd.edu/}\midurl{hcil/}\midurl{DTUI6}
+
+\bibitem{hci2}
+Thimbleby, H. (2007) {\em Press On: Principles of Interaction Programming}, MIT
+  Press.
+
+\bibitem{flexibility}
+Hanson, C. and Sussman, G.~J. (2021) {\em Software design for flexibility: How
+  to avoid programming yourself into a corner}, MIT Press.
+
+\bibitem{iso}
+{ISO/IEC JTC 1/SC 7 Software and systems engineering Committees} (2015) {\em
+  Software engineering --- Guidelines for the application of ISO 9001:2015 to
+  computer software}, International Organization for Standardization (ISO).
+\newblock \url{www.iso.org/}\midurl{standard/}\midurl{74348.html}
+
+\bibitem{178C}
+{RTCA Committee SC-205} (2011) {\em DO-178C --- Software Considerations in
+  Airborne Systems and Equipment Certification}, RTCA.
+\newblock \url{my.rtca.org/}\midurl{NC__Product?id=a1B36000001IcmqEAC}
+
+\bibitem{peerReview}
+Baum, T., {Le\ss mann}, H. and Schneider, K. (2017) ``The choice of code review
+  process: A survey on the state of the practice,'' {\em in Lecture Notes in
+  Computer Science}, vol.~\textbf{10611} of Product-Focused Software Process
+  Improvement: 18th International Conference, pp.~111--127.
+\newblock \doi{10.1007/978-3-319-69926-4_9}
+
+\bibitem{ssi-report}
+Brett, A., Croucher, M., Haines, R., Hettrick, S., Hetherington, J., Stillwell,
+  M. and Wyatt, C. (2017) {\em State of the Nation Report for Research Software
+  Engineers}, Research Software Engineer Network.
+\newblock \url{zenodo.org/}\midurl{record/}\midurl{495360#.Xyfuyi2ZOCM},
+  \doi{10.5281/zenodo.495360}
+
+\bibitem{ssi-url}
+{Software Sustainability Institute} (2020) {\em Web site}.
+\newblock Accessed 3 August 2020, \url{software.ac.uk}
+
+\bibitem{security-engineering}
+Anderson, R. (2020) {\em Security Engineering}, Wiley, 3rd~ed.
+
+\bibitem{cyber-cacm}
+Shostack, A. and Zurko, M.~E. (2020) ``Secure development tools and techniques
+  need more research that will increase their impact and effectiveness in
+  practice,'' {\em Communications of the ACM}, \textbf{63}(5):39--41.
+\newblock \doi{10.1145/3386908}
+
+\bibitem{Schneier}
+Schneier, B. (2018) {\em Click Here To Kill Everybody --- Security and Survival
+  in a Hyper-connected World}, W. W. Norton {\&} Company, Inc, New York, United
+  States.
+
+\bibitem{electronic-evidence}
+Mason, S. and Seng, D. (2017) {\em Electronic Evidence}, Humanities Digital
+  Library, 4th~ed.
+\newblock (NB See
+  \url{humanities-digital-library.org/index.php/hdl/catalog/book/electronicevidence}
+  until DOI is resolved), \doi{10.14296/517.9781911507079}
+
+\bibitem{simon-artificial}
+Simon, H.~A. and Laird, J. (2019) {\em Sciences of the Artificial}, MIT Press,
+  3rd~ed.
+
+\bibitem{dunning-kruger}
+Kruger, J. and Dunning, D. (1999) ``Unskilled and unaware of it: {How}
+  difficulties in recognizing one's own incompetence lead to inflated
+  self-assessments,'' {\em Journal of Personality and Social Psychology},
+  \textbf{77}(6):1121--1134.
+\newblock \doi{10.1037/0022-3514.77.6.1121}
+
+\bibitem{dunning-kruger-numeracy}
+Nuhfer, E., Fleisher, S., Cogan, C., Wirth, K. and Gaze, E. (2017) ``How random
+  noise and a graphical convention subverted behavioral scientists'
+  explanations of self-assessment data: {Numeracy} underlies better
+  alternatives,'' {\em Numeracy}, \textbf{10}(1):4.
+\newblock \doi{10.5038/1936-4660.10.1.4}
+
+\bibitem{enigma}
+Thimbleby, H. (2016) ``{Human Factors} and missed solutions to {Enigma} design
+  weaknesses,'' {\em Cryptologia}, \textbf{40}(2):177--202.
+\newblock \doi{10.1080/01611194.2015.1028680}
+
+\bibitem{harking}
+Kerr, N.~L. (1998) ``{HARKing}: {H}ypothesizing after the results are known,''
+  {\em Personality and Social Psychology Review}, \textbf{2}(3):196--217.
+\newblock \doi{10.1207/s15327957pspr0203\_4}
+
+\bibitem{viruses}
+Thimbleby, H., Anderson, S.~O. and Cairns, P. (1999) ``A framework for
+  modelling {Trojans} and computer virus infection,'' {\em Computer Journal},
+  \textbf{41}(7):444--458.
+\newblock \doi{10.1093/comjnl/41.7.444}
+
+\bibitem{strachey}
+Strachey, C. (2000) ``Fundamental concepts in programming languages,'' {\em
+  Higher-Order and Symbolic Computation}, \textbf{13}:11--49.
+\newblock \doi{10.1023/A:1010000313106}
+
+\bibitem{henderson}
+Henderson, P. (1980) {\em Functional Programming: Application and
+  Implementation}, Prentice-Hall International.
+
+\bibitem{lancet-retracted}
+Mehra, M.~R., Desai, S.~S., Ruschitzka, F. and Patel, A.~N. (2020)
+  ``{RETRACTED}: {Hydroxychloroquine} or chloroquine with or without a
+  macrolide for treatment of {COVID}-19: {A} multinational registry analysis,''
+  {\em The Lancet}, \textbf{{online}}:1--10.
+\newblock \doi{10.1016/S0140-6736(20)31180-6}
+
+\bibitem{fda}
+Zhang, Y., Masci, P., Jones, P. and Thimbleby, H. (2019) ``User interface
+  software errors in medical devices,'' {\em Biomedical Instrumentation {\&}
+  Technology}, \textbf{53}(3):182--194.
+\newblock \doi{10.2345/0899-8205-53.3.182}
+
+\bibitem{stability}
+Salama, M., Bahsoon, R. and Lago, P. (2021) ``Stability in software
+  engineering: Survey of the state-of-the-art and research directions,'' {\em
+  IEEE Transactions on Software Engineering}, \textbf{47}(7):1468--1510.
+\newblock \doi{10.1109/TSE.2019.2925616}
+
+\bibitem{RS-policy}
+Hurst, P. (2022) {\em Data sharing and mining}, Royal Society.
+\newblock \doi{10.25504/FAIRsharing.dIDAzV}
+
+\bibitem{latex-calculus}
+Fuster, R. (2012) ``The {\texttt{calculator}} and {\texttt{calculus}} packages:
+  {Arithmetic} and functional calculations inside {\LaTeX},'' {\em TUGboat},
+  \textbf{33}(3):265--271.
+
+\endofbibliography\end{thebibliography}
+
+% end expanding bibliography appendix.bbl
+
 }
@@
 \immediate\write\infofileForSupplement{% map Git repo names to their citation numbers}
-\immediate\write\infofileForSupplement{% generated by paper-seb-supplementary-material.tex}
+\immediate\write\infofileForSupplement{% generated by appendix.tex}
 \catcode`\%=14
@@
 
-\input generated/supplementary-references.bbl
+% expand generated/survey-references.bbl
+% Generated by programs/data.js
+\vbox{\bibitem{ref-1}
+{\sc A.
+Callahan, E.
+Steinberg, J.
+A.
+Fries, S.
+Gombar, B.
+Patel, C.
+K.
+Corbin, and N.
+H.
+Shah}, ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp, 2020.
+DOI \texttt{10.1038/s41746-020-0300-0}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{MetricSelectionFramework}\bibitem{ref-2}
+{\sc C.
+M.
+Kanzler, M.
+D.
+Rinderknecht, A.
+Schwarz, I.
+Lamers, C.
+Gagnon, J.
+P.
+O.
+Held, P.
+Feys, A.
+R.
+Luft, R.
+Gassert, and O.
+Lambercy}, ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp, 2020.
+DOI \texttt{10.1038/s41746-020-0286-7} {Code \url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-3}
+{\sc A.
+Roy, K.
+Nikolitch, R.
+McGinn, S.
+Jinah, W.
+Klement, and Z.
+A.
+Kaminsky}, ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp, 2020.
+DOI \texttt{10.1038/s41746-020-0287-6}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-4}
+{\sc D.
+M.
+Levine, Z.
+Co, L.
+P.
+Newmark, A.
+R.
+Groisser, A.
+J.
+Holmgren, J.
+A.
+Haas, and D.
+W.
+Bates}, ``Design and testing of a mobile health application rating tool.,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp, 2020.
+DOI \texttt{10.1038/s41746-020-0268-9}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-5}
+{\sc T.
+Kannampallil, J.
+M.
+Smyth, S.
+Jones, P.
+R.
+O.
+Payne, and J.
+Ma}, ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp, 2020.
+DOI \texttt{10.1038/s41746-020-0278-7}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{PENet}\bibitem{ref-6}
+{\sc S.
+Huang, T.
+Kothari, I.
+Banerjee, C.
+Chute, R.
+L.
+Ball, N.
+Borus, A.
+Huang, B.
+N.
+Patel, P.
+Rajpurkar, J.
+Irvin, J.
+Dunnmon, J.
+Bledsoe, K.
+Shpanskaya, A.
+Dhaliwal, R.
+Zamanian, A.
+Y.
+Ng, and M.
+P.
+Lungren}, ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0266-y} {Code \url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-7}
+{\sc S.
+S.
+Dhruva, J.
+S.
+Ross, J.
+G.
+Akar, B.
+Caldwell, K.
+Childers, W.
+Chow, L.
+Ciaccio, P.
+Coplan, J.
+Dong, H.
+J.
+Dykhoff, S.
+Johnston, T.
+Kellogg, C.
+Long, P.
+A.
+Noseworthy, K.
+Roberts, A.
+Saha, A.
+Yoo, and N.
+D.
+Shah}, ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
+{\sc I.
+S.
+Hofer, C.
+Lee, E.
+Gabel, P.
+Baldi, and M.
+Cannesson}, ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp, 2020.
+DOI \texttt{10.1038/s41746-020-0248-0} {Code \url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
+{\sc B.
+Norgeot, K.
+Muenzen, T.
+A.
+Peterson, X.
+Fan, B.
+S.
+Glicksberg, G.
+Schenk, E.
+Rutenberg, B.
+Oskotsky, M.
+Sirota, J.
+Yazdany, G.
+Schmajuk, D.
+Ludwig, T.
+Goldstein, and A.
+J.
+Butte}, ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0258-y} {Code \url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
+{\sc D.
+Choi, J.
+J.
+Park, T.
+Ali, and S.
+Lee}, ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp, 2020.
+DOI \texttt{10.1038/s41746-020-0261-3} {Code \url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-11}
+{\sc C.
+B.
+Hilton, A.
+Milinovich, C.
+Felix, N.
+Vakharia, T.
+Crone, C.
+Donovan, A.
+Proctor, and A.
+Nazha}, ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp, 2020.
+DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
+{\sc M.
+D.
+Li, K.
+Chang, B.
+Bearce, B.
+Y.
+Chang, A.
+J.
+Huang, J.
+P.
+Campbell, J.
+M.
+Brown, P.
+Singh, K.
+V.
+Hoebel, D.
+Erdo{\u g}mu\c{s}, S.
+Ioannidis, W.
+Palmer, M.
+F.
+Chiang, and J.
+Kalpathy-Cramer}, ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp, 2020.
+DOI \texttt{10.1038/s41746-020-0255-1} {Code \url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
+{\sc J.
+I.
+Hoffman, R.
+Nagel, V.
+Litzke, D.
+A.
+Wells, and W.
+Amos}, ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp, 2020.
+DOI \texttt{10.1098/rsos.200419} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
+{\sc P.
+Gr\"onquist, P.
+Panchadcharam, D.
+Wood, A.
+Menges, M.
+R\"uggeberg, and F.
+K.
+Wittel}, ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp, 2020.
+DOI \texttt{10.1098/rsos.192210} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{onlyPaperWithChecks}
+{\sc W.
+Amos}, ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp, 2020.
+DOI \texttt{10.1098/rsos.191900} {Code \url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
+{\sc M.
+Gordon, D.
+Viganola, M.
+Bishop, Y.
+Chen, A.
+Dreber, B.
+Goldfedder, F.
+Holzmeister, M.
+Johannesson, Y.
+Liu, C.
+Twardy, J.
+Wang, and T.
+Pfeiffer}, ``Are replication rates the same across academic fields?
+Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp, 2020.
+DOI \texttt{10.1098/rsos.200566} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
+{\sc D.
+Evans, and A.
+P.
+Field}, ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp, 2020.
+DOI \texttt{10.1098/rsos.200422} {Code \url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ethics-paper}
+{\sc N.
+Beale, H.
+Battey, A.
+C.
+Davison, and R.
+S.
+MacKay}, ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp, 2020.
+DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-19}
+{\sc A.
+A.
+Cherevko, T.
+S.
+Gologush, I.
+A.
+Petrenko, V.
+V.
+Ostapenko, and V.
+A.
+Panarin}, ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp, 2020.
+DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
+{\sc A.
+A.
+Soczawa-Stronczyk, and M.
+Bocian}, ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp, 2020.
+DOI \texttt{10.1098/rsos.200622} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{lactModel}\bibitem{ref-21}
+{\sc S.
+Duruz, E.
+Vajana, A.
+Burren, C.
+Flury, and S.
+Joost}, ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp, 2020.
+DOI \texttt{10.1098/rsos.200638} {Code \url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-22}
+{\sc E.
+Z.
+D.
+de Azevedo, D.
+V.
+Dantas, and F.
+G.
+Daura-Jorge}, ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp, 2020.
+DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{LRM}\bibitem{ref-23}
+{\sc A.
+M.
+Abdolhosseini-Qomi, S.
+H.
+Jafari, A.
+Taghizadeh, N.
+Yazdani, M.
+Asadpour, and M.
+Rahgozar}, ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp, 2020.
+DOI \texttt{10.1098/rsos.191928} {Code \url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{1}\bibitem{example-numerical-error}
+{\sc J.
+Webster, and M.
+Amos}, ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp, 2020.
+DOI \texttt{10.1098/rsos.200307} {Code \url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-25}
+{\sc Y-l.
+Zhu, C-J.
+Wang, F.
+Gao, Z-x.
+Xiao, P-l.
+Zhao, and J-y.
+Wang}, ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp, 2020.
+DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-26}
+{\sc B.
+Yu, C.
+J.
+Scott, X.
+Xue, X.
+Yue, and X.
+Dou}, ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp, 2020.
+DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-27}
+{\sc K.
+Joon-myoung, C.
+Younghoon, J.
+Ki-Hyun, C.
+Soohyun, K.
+Kyung-Hee, B.
+Seung D, J.
+Soomin, P.
+Jinsik, and O.
+Byung-Hee}, ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-28}
+{\sc H.
+Zhu, C.
+Cheng, H.
+Yin, X.
+Li, P.
+Zuo, J.
+Ding, F.
+Lin, J.
+Wang, B.
+Zhou, Y.
+Li, S.
+Hu, Y.
+Xiong, B.
+Wang, G.
+Wan, X.
+Yang, and Y.
+Yuan}, ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
+{\sc R.
+Fung, J.
+Villar, A.
+Dashti, L.
+C.
+Ismail, E.
+Staines-Urias, E.
+O.
+Ohuma, L.
+J.
+Salomon, C.
+G.
+Victora, F.
+C.
+Barros, A.
+Lambert, M.
+Carvalho, A.
+Jaffer Y, J.
+A.
+Noble, M.
+G.
+Gravett, M.
+Purwar, R.
+Pang, E.
+Bertino, S.
+Munim, A.
+M.
+Min, R.
+McGready, S.
+A.
+Norris, Z.
+A.
+Bhutta, S.
+H.
+Kennedy, A.
+T.
+Papageorghiou, and A.
+Ourmazd}, ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30131-X} {Code \url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-30}
+{\sc C.
+Sabanayagam, D.
+Xu, D.
+S.
+W.
+Ting, S.
+Nusinovici, R.
+Banu, H.
+Hamzah, C.
+Lim, Y-C.
+Tham, C.
+Y.
+Cheung, E.
+S.
+Tai, X.
+Y.
+Wang, J.
+B.
+Jonas, C-Y.
+Cheng, M.
+L.
+Lee, W.
+Hsu, and T.
+Y.
+Wong}, ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30063-7}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{blast-ct}\bibitem{ref-31}
+{\sc M.
+Monteiro, V.
+F.
+Newcombe, F.
+Mathieu, K.
+Adatia, K.
+Kamnitsas, E.
+Ferrante, T.
+Das, D.
+Whitehouse, D.
+Rueckert, D.
+K.
+Menon, and B.
+Glocker}, ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30085-6} {Code \url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-32}
+{\sc K-L.
+Liu, T.
+Wu, P-T.
+Chen, M.
+Tsai Y, H.
+Roth, M-S.
+Wu, W-C.
+Liao, and W.
+Wang}, ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp, 2020.
+DOI \texttt{10.1016/S2589-7500(20)30078-9}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
+
+% end expanding generated/survey-references.bbl
 

### 2023-06-06 expanded-paper.tex
Preparing to upload to Computer Journal

--- 2023-05-31 paper-seb-main.tex
@@
+% 
+% This file has been expanded from LaTeX file (paper.tex)
+%   to make this straight forward LaTeX file (expanded-paper.tex)
+%     that depends on nothing (it inputs no files apart from the document class files).
+%       Each required file's text was input in situ - search for % expand
+% 
+% (Otherwise the file paper.tex would use BibTeX, a macro file, tables of contents, its own and the other files's .aux files, and includes various files in generated/*, etc.)
+%
+
+\makeatletter
+\relax 
+\providecommand\hyper@newdestlabel[2]{}
+\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
+\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
+\global\let\oldnewlabel\newlabel
+\gdef\newlabel#1#2{\newlabelxx{#1}#2}
+\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
+\AtEndDocument{\ifx\hyper@anchor\@undefined
+\let\newlabel\oldnewlabel
+\fi}
+\fi}
+\global\let\hyper@last\relax 
+\gdef\HyperFirstAtBeginDocument#1{#1}
+\providecommand\HyField@AuxAddToFields[1]{}
+\providecommand\HyField@AuxAddToCoFields[2]{}
+\citation{a=b}
+\citation{write-for-life}
+\citation{abelson}
+\citation{nature-chatgpt}
+\newlabel{FirstPage}{{}{1}{}{Doc-Start}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {1}Introduction}{1}{section.1}\protected@file@percent }
+\citation{sommerville,knight}
+\citation{computational-thinking-wing,computational-thinking}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{reproducibility-crisis,rescience}
+\@writefile{toc}{\contentsline {section}{\numberline {2}Background}{2}{section.2}\protected@file@percent }
+\@writefile{lot}{\contentsline {table}{\numberline {1}{\ignorespaces Overview of peer-reviewed paper sample, broken down further in Table \ref {table-summary}.
+Survey methodology and data is provided in the Supplemental Material.
+(The survey does not include the motivating papers \cite  {ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref {section-pandemic-modeling}.)}}{3}{table.1}\protected@file@percent }
+\newlabel{table-overview}{{1}{3}{Overview of peer-reviewed paper sample, broken down further in Table \ref {table-summary}.
+Survey methodology and data is provided in the \supplement .
+(The survey does not include the motivating papers \cite {ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref {section-pandemic-modeling}.)}{table.1}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {2}{\ignorespaces Breakdown of pilot survey of peer-reviewed science papers relying on code. }}{3}{table.2}\protected@file@percent }
+\newlabel{table-summary}{{2}{3}{Breakdown of pilot survey of peer-reviewed science papers relying on code}{table.2}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {3}{\ignorespaces The TOP committee's recommended levels for journal article code transparency.
+Level 0 is provided for a comparison that does not meet any TOP requirements.
+Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref {reproducibility-concerns}.}}{3}{table.3}\protected@file@percent }
+\newlabel{TOPtable}{{3}{3}{The TOP committee's recommended levels for journal article code transparency.
+Level 0 is provided for a comparison that does not meet any TOP requirements.
+Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref {reproducibility-concerns}}{table.3}{}}
+\citation{temperature}
+\citation{form}
+\citation{acm-algorithms}
+\citation{hoare-correctness}
+\citation{jupyter-study}
+\citation{trisovic}
+\citation{jupyter-study}
+\citation{trisovic}
+\citation{trisovic}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.a}Code quality concerns}{4}{subsection.2.1}\protected@file@percent }
+\newlabel{related-work}{{2.a}{4}{Code quality concerns}{subsection.2.1}{}}
+\citation{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis}
+\citation{joss}
+\citation{rescience}
+\citation{TOP}
+\citation{TOP-wiki}
+\citation{jupyter-study}
+\citation{MMR-fraud}
+\citation{misconduct-issues}
+\citation{machine-learning-reproducibility}
+\citation{elns}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.b}Computable papers}{5}{subsection.2.2}\protected@file@percent }
+\newlabel{computable-papers}{{2.b}{5}{Computable papers}{subsection.2.2}{}}
+\citation{ICmodel}
+\citation{perkel-review}
+\citation{maneage}
+\citation{maneage}
+\citation{LP,LP-book}
+\citation{relit}
+\citation{notebook-patent}
+\citation{jupyter-history}
+\citation{knitr}
+\citation{spec-led-design}
+\citation{RAPstandard}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.c}RAP: Reproducible Analytical Pipelines}{7}{subsection.2.3}\protected@file@percent }
+\newlabel{RAP-section}{{2.c}{7}{RAP: Reproducible Analytical Pipelines}{subsection.2.3}{}}
+\citation{rap,goldacre,turing-way}
+\@writefile{lot}{\contentsline {table}{\numberline {4}{\ignorespaces A simplified schematic of the publication pipeline.
+For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
+The RAP and RAP{\tt  +}\ approaches encode the normally manual steps in the pipeline workflow so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The encoded RAP{\tt  +}\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
+Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the Supplemental Material.}}{8}{table.4}\protected@file@percent }
+\newlabel{fig-pipeline}{{4}{8}{A simplified schematic of the publication pipeline.
+For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
+The RAP and \RAPstar \ approaches encode the normally manual steps in the pipeline workflow so that they can be run automatically, and hence reproduce the results that underpin the final paper.
+The encoded \RAPstar \ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
+Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the \supplement }{table.4}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {5}{\ignorespaces A minimum standard of RAP, based on the UK Statistics Authority summary \cite  {RAPstandard}.}}{8}{table.5}\protected@file@percent }
+\newlabel{figure-RAP-summary}{{5}{8}{A minimum standard of RAP, based on the UK Statistics Authority summary \cite {RAPstandard}}{table.5}{}}
+\citation{MMR-fraud}
+\citation{a=b}
+\citation{Courtes}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{9}{subsubsection.2.3.1}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.2}RAP as research}{9}{subsubsection.2.3.2}\protected@file@percent }
+\citation{Courtes}
+\citation{Courtes}
+\citation{apa-statistics,Mariusz-reporting,science-statistics-recommendations}
+\citation{example-stats}
+\citation{science-statistics-recommendations}
+\citation{Spiegelhalter}
+\citation{paul-hci,pnas-stats}
+\citation{paul-hci,pnas-stats}
+\citation{Ben,se-bias}
+\@writefile{toc}{\contentsline {section}{\numberline {3}The statistics/code analogy}{10}{section.3}\protected@file@percent }
+\newlabel{statistics-computation-analogy}{{3}{10}{The statistics/code analogy}{section.3}{}}
+\citation{sas9.2}
+\citation{whitty}
+\newlabel{section-vicious-cycles}{{3}{11}{The statistics/code analogy}{section.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {3.a}The siren call of over-fitting code}{11}{subsection.3.1}\protected@file@percent }
+\newlabel{over-fit}{{3.a}{11}{The siren call of over-fitting code}{subsection.3.1}{}}
+\citation{over-fit}
+\citation{over-fit}
+\citation{may-simple}
+\citation{dyson}
+\newlabel{name-example-footnote}{{5}{12}{}{Hfootnote.5}{}}
+\citation{assurance-case}
+\@writefile{lof}{\contentsline {figure}{\numberline {1}{\ignorespaces Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
+Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
+The over-fitted polynomial model fits the sample \emph  {exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}}{13}{figure.1}\protected@file@percent }
+\newlabel{fig-overfit}{{1}{13}{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
+Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
+The over-fitted polynomial model fits the sample \emph {exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data}{figure.1}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {4}The conventional role of code}{13}{section.4}\protected@file@percent }
+\citation{assessing-quality}
+\citation{fixit}
+\citation{Hammurabi}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.a}The deceptive simplicity of code}{14}{subsection.4.1}\protected@file@percent }
+\newlabel{deceptive-simplicity-of-code}{{4.a}{14}{The deceptive simplicity of code}{subsection.4.1}{}}
+\citation{cite-code,TOP}
+\citation{prisma}
+\citation{engima}
+\citation{critiques,diagnosis-reviews}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.b}The low status of coding}{15}{subsection.4.2}\protected@file@percent }
+\newlabel{lowCodeStatus}{{4.b}{15}{The low status of coding}{subsection.4.2}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.c}The critical role of code is often ignored}{15}{subsection.4.3}\protected@file@percent }
+\newlabel{central-role-of-code}{{4.c}{15}{The critical role of code is often ignored}{subsection.4.3}{}}
+\newlabel{PRISMA-statement}{{4.c}{15}{The critical role of code is often ignored}{subsection.4.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{15}{subsection.4.4}\protected@file@percent }
+\newlabel{knowledge}{{4.d}{15}{Bugs, code and programming}{subsection.4.4}{}}
+\citation{knuth}
+\citation{essence-of-software,fixit}
+\citation{popper-conjectures-refutations}
+\citation{preventable-book}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}\protected@file@percent }
+\citation{preventable-book}
+\citation{science-review}
+\citation{diagnosis-reviews}
+\citation{tripod}
+\citation{diagnosis-reviews}
+\citation{diagnosis-reviews}
+\citation{nature-summary,ICmodel}
+\citation{avianFluModel,originalICmodel}
+\citation{ICmodel,avianFluModel,originalICmodel}
+\citation{tweet}
+\citation{ferguson-interview}
+\citation{basic-reproducibilty}
+\citation{ferguson-resign}
+\@writefile{toc}{\contentsline {section}{\numberline {5}State of the art}{17}{section.5}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.a}Case study: Pandemic modeling}{17}{subsection.5.1}\protected@file@percent }
+\newlabel{section-pandemic-modeling}{{5.a}{17}{Case study: Pandemic modeling}{subsection.5.1}{}}
+\citation{tweet}
+\citation{forsyth,Normal-Distributions}
+\citation{misra}
+\citation{spark}
+\citation{haskell}
+\citation{thumbs-up}
+\citation{nature-summary,ICmodel}
+\citation{codecheck,thumbs-up}
+\citation{codecheck}
+\citation{ICmodel}
+\citation{codecheck-certificate}
+\citation{codecheck}
+\citation{codecheck-certificate}
+\newlabel{discuss-covid-sim}{{5.a}{18}{Case study: Pandemic modeling}{subsection.5.1}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.b}Concerns with reproducibility}{18}{subsection.5.2}\protected@file@percent }
+\newlabel{reproducibility-concerns}{{5.b}{18}{Concerns with reproducibility}{subsection.5.2}{}}
+\citation{codecheck,thumbs-up}
+\citation{codecheck-certificate}
+\citation{NVP}
+\citation{nvp-ferguson}
+\citation{essence-of-software}
+\citation{notebooks,popper-conjectures-refutations,tripod,prisma,nih-policy,nih-nature}
+\citation{refactoring}
+\citation{bad-code}
+\citation{bad-code}
+\citation{pseudo}
+\citation{chinese}
+\citation{relit}
+\citation{flu-model}
+\citation{nature-review}
+\citation{nature-review}
+\citation{psychological-modeling}
+\citation{psychological-modeling}
+\citation{actr}
+\citation{psychological-modeling}
+\citation{Ben,nature-review,se-bias}
+\citation{ourReview}
+\citation{hamming}
+\citation{NVP}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.c}Beyond pandemic modeling}{20}{subsection.5.3}\protected@file@percent }
+\newlabel{section-science-beyond-pandemic-modeling}{{5.c}{20}{Beyond pandemic modeling}{subsection.5.3}{}}
+\citation{unfinished}
+\citation{unfinished,lancet-unfinished}
+\citation{hoare}
+\citation{machine-learning-reproducibility}
+\citation{relit}
+\citation{preclinical-reproducibility}
+\citation{fair,fair-principles}
+\citation{nih-policy}
+\citation{nih-nature}
+\newlabel{section-discussion}{{5.c}{21}{Beyond pandemic modeling}{subsection.5.3}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.d}The lop-sided emphasis on data}{21}{subsection.5.4}\protected@file@percent }
+\newlabel{critique-fair}{{5.d}{21}{The lop-sided emphasis on data}{subsection.5.4}{}}
+\citation{notebooks}
+\citation{cbc}
+\citation{ethics-code}
+\@writefile{toc}{\contentsline {section}{\numberline {6}Rethinking science that uses code}{22}{section.6}\protected@file@percent }
+\newlabel{summary}{{6}{22}{Rethinking science that uses code}{section.6}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.a}Software Engineering Boards}{22}{subsection.6.1}\protected@file@percent }
+\citation{acm-artifacts}
+\citation{redmill,iec61508}
+\citation{hoare-unreasonable}
+\citation{basic-reproducibilty,open-source}
+\citation{ABCs-SE}
+\citation{mrb}
+\citation{mrb}
+\citation{relit,Executable-Paper}
+\citation{computational-thinking-wing,computational-thinking}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.b}Extending RAP to RAP{\tt  +}}{23}{subsection.6.2}\protected@file@percent }
+\newlabel{RAPstar-section}{{6.b}{23}{Extending RAP to \RAPstar }{subsection.6.2}{}}
+\citation{RMarkdown}
+\citation{lepton}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.c}The paper as a scientific laboratory}{24}{subsection.6.3}\protected@file@percent }
+\newlabel{paper-as-lab}{{6.c}{24}{The paper as a scientific laboratory}{subsection.6.3}{}}
+\citation{landauer,thimbleby-landauer}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.d}Action must be interdisciplinary}{25}{subsection.6.4}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.e}Methodological statements}{25}{subsection.6.5}\protected@file@percent }
+\citation{no-raw-data}
+\citation{data-access}
+\citation{assurance-case}
+\citation{debt}
+\newlabel{mentionPLOSandIEEE}{{6.e}{26}{Methodological statements}{subsection.6.5}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{26}{subsection.6.6}\protected@file@percent }
+\newlabel{technical-debt}{{6.f}{26}{Training to reduce technical debt}{subsection.6.6}{}}
+\citation{fixit}
+\citation{jpm}
+\citation{jpm}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.g}Benefits beyond science}{27}{subsection.6.7}\protected@file@percent }
+\newlabel{benefits-beyond-science}{{6.g}{27}{Benefits beyond science}{subsection.6.7}{}}
+\citation{jpm-report}
+\citation{jpm}
+\citation{hoare-unreasonable}
+\citation{excel-fiasco}
+\citation{science-delays}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {7}Conclusions}{28}{section.7}\protected@file@percent }
+\citation{tweet,ferguson-interview}
+\citation{whitty}
+\citation{fixit}
+\citation{reproducibility,relit,popper-conjectures-refutations}
+\citation{parliamentary-evidence,my-parliamentary-evidence}
+\bibstyle{revisedCompJ}
+\bibdata{paper.bib}
+\bibcite{a=b}{1}
+\bibcite{write-for-life}{2}
+\bibcite{abelson}{3}
+\bibcite{nature-chatgpt}{4}
+\bibcite{sommerville}{5}
+\bibcite{knight}{6}
+\bibcite{computational-thinking-wing}{7}
+\bibcite{computational-thinking}{8}
+\bibcite{ICmodel}{9}
+\bibcite{avianFluModel}{10}
+\bibcite{originalICmodel}{11}
+\bibcite{reproducibility-crisis}{12}
+\bibcite{rescience}{13}
+\bibcite{temperature}{14}
+\bibcite{form}{15}
+\bibcite{acm-algorithms}{16}
+\bibcite{hoare-correctness}{17}
+\bibcite{jupyter-study}{18}
+\bibcite{trisovic}{19}
+\bibcite{machine-learning-reproducibility}{20}
+\bibcite{no-raw-data}{21}
+\bibcite{data-access}{22}
+\bibcite{reproducible-manifesto}{23}
+\bibcite{joss}{24}
+\bibcite{TOP}{25}
+\bibcite{TOP-wiki}{26}
+\bibcite{MMR-fraud}{27}
+\bibcite{misconduct-issues}{28}
+\bibcite{elns}{29}
+\bibcite{perkel-review}{30}
+\bibcite{maneage}{31}
+\bibcite{LP}{32}
+\bibcite{LP-book}{33}
+\bibcite{relit}{34}
+\bibcite{notebook-patent}{35}
+\bibcite{jupyter-history}{36}
+\bibcite{knitr}{37}
+\bibcite{spec-led-design}{38}
+\bibcite{RAPstandard}{39}
+\bibcite{rap}{40}
+\bibcite{goldacre}{41}
+\bibcite{turing-way}{42}
+\bibcite{Courtes}{43}
+\bibcite{apa-statistics}{44}
+\bibcite{Mariusz-reporting}{45}
+\bibcite{science-statistics-recommendations}{46}
+\bibcite{example-stats}{47}
+\bibcite{Spiegelhalter}{48}
+\bibcite{paul-hci}{49}
+\bibcite{pnas-stats}{50}
+\bibcite{Ben}{51}
+\bibcite{se-bias}{52}
+\bibcite{sas9.2}{53}
+\bibcite{whitty}{54}
+\bibcite{over-fit}{55}
+\bibcite{may-simple}{56}
+\bibcite{dyson}{57}
+\bibcite{assurance-case}{58}
+\bibcite{assessing-quality}{59}
+\citation{assessing-quality-conference}
+\bibcite{fixit}{60}
+\bibcite{Hammurabi}{61}
+\bibcite{cite-code}{62}
+\bibcite{prisma}{63}
+\bibcite{engima}{64}
+\bibcite{critiques}{65}
+\bibcite{diagnosis-reviews}{66}
+\bibcite{knuth}{67}
+\bibcite{essence-of-software}{68}
+\bibcite{popper-conjectures-refutations}{69}
+\bibcite{preventable-book}{70}
+\bibcite{science-review}{71}
+\bibcite{tripod}{72}
+\bibcite{nature-summary}{73}
+\bibcite{tweet}{74}
+\bibcite{ferguson-interview}{75}
+\bibcite{basic-reproducibilty}{76}
+\bibcite{ferguson-resign}{77}
+\bibcite{forsyth}{78}
+\bibcite{Normal-Distributions}{79}
+\bibcite{misra}{80}
+\bibcite{spark}{81}
+\bibcite{haskell}{82}
+\bibcite{thumbs-up}{83}
+\bibcite{codecheck}{84}
+\bibcite{codecheck-certificate}{85}
+\bibcite{NVP}{86}
+\bibcite{nvp-ferguson}{87}
+\bibcite{notebooks}{88}
+\bibcite{nih-policy}{89}
+\bibcite{nih-nature}{90}
+\bibcite{refactoring}{91}
+\bibcite{bad-code}{92}
+\bibcite{pseudo}{93}
+\bibcite{chinese}{94}
+\bibcite{flu-model}{95}
+\bibcite{nature-review}{96}
+\bibcite{psychological-modeling}{97}
+\bibcite{actr}{98}
+\bibcite{ourReview}{99}
+\bibcite{hamming}{100}
+\bibcite{unfinished}{101}
+\bibcite{lancet-unfinished}{102}
+\bibcite{hoare}{103}
+\bibcite{preclinical-reproducibility}{104}
+\bibcite{fair}{105}
+\bibcite{fair-principles}{106}
+\bibcite{cbc}{107}
+\bibcite{ethics-code}{108}
+\bibcite{acm-artifacts}{109}
+\bibcite{redmill}{110}
+\bibcite{iec61508}{111}
+\bibcite{hoare-unreasonable}{112}
+\bibcite{open-source}{113}
+\bibcite{ABCs-SE}{114}
+\bibcite{mrb}{115}
+\bibcite{Executable-Paper}{116}
+\bibcite{RMarkdown}{117}
+\bibcite{lepton}{118}
+\bibcite{landauer}{119}
+\bibcite{thimbleby-landauer}{120}
+\bibcite{debt}{121}
+\bibcite{jpm}{122}
+\bibcite{jpm-report}{123}
+\bibcite{excel-fiasco}{124}
+\bibcite{science-delays}{125}
+\bibcite{reproducibility}{126}
+\bibcite{parliamentary-evidence}{127}
+\bibcite{my-parliamentary-evidence}{128}
+\citation{parliamentary-evidence}
+\bibcite{assessing-quality-conference}{129}
+\def\continueSectionNumbers{7}
+\def\continueTableNumbers{5}
+\def\continueFigureNumbers{1}
+\def\continueFootnoteNumbers{13}
+\def\continuePageNumbers{35}
+\newlabel{LastPage}{{}{35}}
+\gdef \@abspage@last{35}
+
+\makeatother
 \documentclass{comjnl}
@@
 
-\input paper-seb-macros.tex
-\input generated/page-lengths.tex
+% expand macros.tex
+% due to finding out that PDF doesn't work everywhere, here's where we define the image type we're using for imaged created by the Mathematica notebooks
+\def\imageFileSuffix{jpg}
+
+\usepackage[hyphens]{url}
+\usepackage{hyperref}
+\usepackage{xcolor}
+\usepackage{longtable}
+\usepackage{cite}
+
+\renewcommand\thesubsection{\thesection.\alph{subsection}}
+\renewcommand\thesubsubsection{\thesection.\alph{subsection}.\arabic{subsubsection}}
+
+\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
+\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}
 
 \makeatletter
-      \input paper-seb-supplementary-material.aux 
+    \def\do@url@hyp{\do/}
+    \usepackage{moreverb}
+\makeatother
+
+\DeclareUrlCommand\doi{\def\UrlLeft{{\textrm{DOI}\ } }}
+\DeclareUrlCommand\url{\def\UrlLeft{{\textrm{URL}\ } }}
+\DeclareUrlCommand\midurl{\def\UrlLeft{}\allowbreak }
+
+\renewcommand{\thefootnote}{\arabic{footnote}}
+
+\def\mytitle{Improving science that uses code}
+\def\supplement{Supplemental Material}
+
+\def\citeeg#1{[e.g., \citenum{#1}]}
+
+% calculate rounded percentages
+\newcount \t
+\newcount \tu
+\def\pc#1#2{\t = #1%
+\multiply \t by 100000%
+\divide \t by #2% ?? 0 times percentage
+\tu = \t
+\divide \t by 1000% percentage, but no idea of remainder
+\multiply \t by 1000% 1000 times percentage, with 00 as bottom digits
+\advance \tu by -\t% bottom two digits
+\divide \t by 1000\relax%
+% round to nearest integer
+% if ending in > .5 round up
+% if ending in exactly .5 round towards nearest even number
+\ifnum \tu > 500 % round up
+	\advance \t by 1 
+\else
+	\ifnum \tu = 500 
+		\ifodd \t % it's odd, so round up to even number 
+			\advance \t by 1
+		\else % leave rounded to even number 	
+		\fi
+	\fi
+\fi
+\the\t\%}
+
+% code to make digits into blocks of 3 (as in 1000 -> 1 000)
+\def\xBlockSep{\textcolor{red}{\hskip 0.3em :}\hskip 0.3em} % International Standards are to use a small space. UK typically uses a comma
+% NOTE currently, the few decimal numbers (eg 0.000 5) use hskip explicitly, since defining \xBlockSep as a comma would definitely be wrong for decimals
+\newcount\xdigitBlockn
+\newcount\xdigitBlocktmp
+\def\xdigitBlock#1{%
+        \xdigitBlockn=#1\relax
+        \ifnum \xdigitBlockn < 0
+                $-$%
+                \xdigitBlockn = -\xdigitBlockn
+        \fi
+        \let \xdigitBlockpptemp \empty
+        \loop \ifnum \xdigitBlockn > 999
+                \xdigitBlocktmp = \xdigitBlockn
+                \divide \xdigitBlockn by 1000
+                \xdigitBlocktmp = \numexpr \xdigitBlocktmp - 1000*\xdigitBlockn \relax
+                \edef \xdigitBlockpptemp{\xBlockSep\ifnum \xdigitBlocktmp < 100  0\ifnum \xdigitBlocktmp < 10 0\fi\fi \the\xdigitBlocktmp \xdigitBlockpptemp}%
+        \repeat
+        \the\xdigitBlockn
+        \xdigitBlockpptemp
+}
+
+\def\digitBlock#1{\hbox{\xdigitBlock{#1}}}
+
+% simple macro to use a number register to pluralize (add an s) to words....
+% \plural{\t}{fred} => \the\t\ freds if plural, or => one fred if singular
+\def\plural#1#2{\ifnum #1=1 
+	one #2%
+\else
+	\the#1\ #2s%
+\fi}
+
+% initialiseBibliography <title> <starting number> <introductory text>
+\newcount \bibciten \bibciten=0
+\def\initialiseBibliography#1#2#3{% title, starting citation number, an initial paragraph
+	\global\def\refname{\arabic{section} #1}
+    \global\def\unnumberedrefname{#1}
+    \global\bibciten=#2
+    \global\def\startBibliography{#3}
+}
+
+\def\bibskip{\vskip 1ex} % the gap between bib items generated by data.js
+
+% read in any constants defined from the JSON or Git data
+% expand generated/constants.tex
+% Generated by programs/data.js
+\def\journalBreakdown{\emph{Lancet Digital Health\/} ($N=6$), \emph{Nature Digital Medicine\/} ($N=12$) and \emph{Royal Society Open Science\/} ($N=14$)}
+\def\tabularJournalBreakdown{&\hbox to 3em {\hfill 6}\hskip 1em \emph{Lancet Digital Health}\\&\hbox to 3em {\hfill 12}\hskip 1em \emph{Nature Digital Medicine}\\&\hbox to 3em {\hfill 14}\hskip 1em \emph{Royal Society Open Science}\\}
+\global\newcount \dataN \global\dataN=32
+\global\newcount \countAuthors \global\countAuthors=264
+\global\newcount \countHasBreach \global\countHasBreach=11
+\global\newcount \countHasPolicy \global\countHasPolicy=26
+\global\newcount \countUsesVersionControlRepository \global\countUsesVersionControlRepository=10
+\global\newcount \counthasDataRepository \global\counthasDataRepository=9
+\global\newcount \countNoCodeInRepo \global\countNoCodeInRepo=1
+\global\newcount \numberOfJournals \global\numberOfJournals=3
+\global\newcount \countCodetested \global\countCodetested=0
+\global\newcount \hasDevelopedRigorously \global\hasDevelopedRigorously=0
+\global\newcount \countFields \global\countFields=36
+\global\newcount \countCommentedOut \global\countCommentedOut=7
+\global\newcount \countNotCommentedOut \global\countNotCommentedOut=2
+\global\newcount \countMissingCode \global\countMissingCode=1
+\global\def \averagePagesPerAuthor{1.2916666666666667}
+% end expanding generated/constants.tex
+% expand generated/metadata.tex
+% shell script (make-metadata.sh) generated LaTeX variable values counting errors detected by data.js
+\newcount \JSONerrorCount
+\JSONerrorCount = 30
+\newcount \dataVariableCount
+\dataVariableCount = 63
+% end expanding generated/metadata.tex
+% expand generated/covidsimSummary.tex
+% data calculated from the Ferguson covid-sim repo
+\def \covidsimkLOC {25}
+\def \covidsimfiles {229}
+\def \covidsimdata{734 Mb}
+% end expanding generated/covidsimSummary.tex
+% expand generated/clone-date.tex
+% date generated by running generated/allGitRepos.sh downloading Git repos
+\def\clonedate {\ignorespaces 25 May 2023}
+\def\cloneyear {2023}
+\def\clonemonth {05}
+% end expanding generated/clone-date.tex
+
+\newcount \pubdelayinmonths
+\pubdelayinmonths=\cloneyear
+\advance \pubdelayinmonths by -2019 % submitted in 2019
+\multiply \pubdelayinmonths by 12
+\advance \pubdelayinmonths by -6 % submitted in June
+\advance \pubdelayinmonths by \clonemonth
+
+\def\monthname#1{\ifcase#1 \or 
+January\or 
+February\or 
+March\or 
+April\or 
+May\or 
+June\or 
+July\or 
+August\or 
+September\or 
+October\or 
+November\or 
+December\fi}
+
+\def\numbertoname#1{\ifcase #1 zero\or one\or two\or three\or four\or five\or six\or seven\or eight\or nine\else #1\fi}
+\def\Numbertoname#1{\ifcase #1 Zero\or One\or Two\or Three\or Four\or Five\or Six\or Seven\or Eight\or Nine\else #1\fi}
+
+\def\clonewhen{\monthname{\clonemonth} \cloneyear}
+
+\def\RAPstar{RAP{\tt +}}
+\def\RAPstars{RAPs{\tt +}}
+\def\RAPstarp#1{RAP{\tt +}#1}
+\def\italicRAPstar{RAP{\tt +}}
+% end expanding macros.tex
+% expand generated/page-lengths.tex
+% Generated by programs/data.js
+\expandafter\def\csname pagelength-MetricSelectionFramework\endcsname{17}
+\expandafter\def\csname pagelength-PENet\endcsname{9}
+\expandafter\def\csname pagelength-PostoperativeOutcomes.RiskNet\endcsname{10}
+\expandafter\def\csname pagelength-philter-ucsf\endcsname{8}
+\expandafter\def\csname pagelength-AI-CDSS-Cardiovascular-Silo\endcsname{6}
+\expandafter\def\csname pagelength-SiameseChange\endcsname{9}
+\expandafter\def\csname pagelength-lactModel\endcsname{13}
+\expandafter\def\csname pagelength-LRM\endcsname{22}
+\expandafter\def\csname pagelength-manifold-ga\endcsname{7}
+\expandafter\def\csname pagelength-blast-ct\endcsname{8}
+\newcount \gitPages \gitPages=109
+\newcount \totalPages \totalPages=341
+% end expanding generated/page-lengths.tex
+
+\makeatletter
+      % expand appendix.aux 
+\relax 
+\providecommand\hyper@newdestlabel[2]{}
+\providecommand\HyperFirstAtBeginDocument{\AtBeginDocument}
+\HyperFirstAtBeginDocument{\ifx\hyper@anchor\@undefined
+\global\let\oldnewlabel\newlabel
+\gdef\newlabel#1#2{\newlabelxx{#1}#2}
+\gdef\newlabelxx#1#2#3#4#5#6{\oldnewlabel{#1}{{#2}{#3}}}
+\AtEndDocument{\ifx\hyper@anchor\@undefined
+\let\newlabel\oldnewlabel
+\fi}
+\fi}
+\global\let\hyper@last\relax 
+\gdef\HyperFirstAtBeginDocument#1{#1}
+\providecommand\HyField@AuxAddToFields[1]{}
+\providecommand\HyField@AuxAddToCoFields[2]{}
+\citation{vancouver}
+\citation{hamming}
+\citation{fixit}
+\citation{cosmos}
+\@writefile{toc}{\contentsline {section}{\numberline {8}Further issues for Software Engineering Boards (SEBs)}{38}{section.8}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {8.a}Relationships of SEBs to Ethics Boards}{38}{subsection.8.1}\protected@file@percent }
+\citation{ethics-paper}
+\citation{ai-ethics}
+\citation{science-lancet1,science-lancet2,lancet-learning}
+\citation{jvs1,jvs2,jvs3}
+\citation{sommerville,knight}
+\citation{cosmos}
+\citation{swebok}
+\citation{turing-way}
+\citation{wpf}
+\citation{humphrey}
+\citation{managers}
+\@writefile{toc}{\contentsline {subsection}{\numberline {8.b}SEBs are necessary but not sufficient}{39}{subsection.8.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {9}Software engineering best practice}{39}{section.9}\protected@file@percent }
+\newlabel{supplement-best-practice}{{9}{39}{Software engineering best practice}{section.9}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.a}Requirements}{39}{subsection.9.1}\protected@file@percent }
+\citation{differential-equations}
+\citation{cbc}
+\citation{spark}
+\citation{misra}
+\citation{hoare}
+\citation{hoare}
+\citation{numerals}
+\citation{heedless}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.b}Formal methods}{40}{subsection.9.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.c}Defensive programming}{40}{subsection.9.3}\protected@file@percent }
+\citation{onlyPaperWithChecks}
+\citation{tweet,plos}
+\citation{heedless}
+\citation{spark}
+\citation{haskell}
+\citation{basic-reproducibilty,reproducibility}
+\citation{NVP}
+\citation{avianFluModel}
+\citation{ICmodel}
+\citation{flu-model}
+\citation{excel1,excel2}
+\citation{npm}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.d}Using inappropriate programming languages undermines reliability}{41}{subsection.9.4}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.e}Open source and version control}{41}{subsection.9.5}\protected@file@percent }
+\citation{open-source}
+\citation{sommerville,knight,NVP}
+\citation{assurance-case}
+\citation{LP}
+\citation{relit}
+\citation{hci1,hci2}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.f}Rigorous testing}{42}{subsection.9.6}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.g}Good documentation and record keeping}{42}{subsection.9.7}\protected@file@percent }
+\citation{flexibility}
+\citation{iso}
+\citation{178C}
+\citation{fixit}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.h}Usability}{43}{subsection.9.8}\protected@file@percent }
+\newlabel{supplement-material-makefiles}{{9.h}{43}{Usability}{subsection.9.8}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.i}Reusing quality solutions}{43}{subsection.9.9}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.j}Simplicity}{43}{subsection.9.10}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.k}Compliance with standards}{43}{subsection.9.11}\protected@file@percent }
+\citation{peerReview,knight}
+\citation{ssi-report,ssi-url}
+\citation{acm-artifacts}
+\citation{security-engineering,cyber-cacm}
+\citation{example-numerical-error}
+\citation{hamming}
+\citation{Schneier}
+\citation{electronic-evidence}
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.l}Effective multidisciplinary teamwork}{44}{subsection.9.12}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.m}Continuous Professional Development (CPD)}{44}{subsection.9.13}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.n}Security and other factors}{44}{subsection.9.14}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {9.o}Software is a human activity}{44}{subsection.9.15}\protected@file@percent }
+\citation{fixit}
+\citation{popper-conjectures-refutations}
+\citation{simon-artificial}
+\citation{dunning-kruger,dunning-kruger-numeracy}
+\citation{dunning-kruger}
+\citation{fixit}
+\citation{enigma}
+\citation{harking}
+\citation{viruses}
+\@writefile{toc}{\contentsline {section}{\numberline {10}Code, data, and publication}{45}{section.10}\protected@file@percent }
+\newlabel{supplement-on-code-data-publication}{{10}{45}{Code, data, and publication}{section.10}{}}
+\citation{viruses}
+\citation{strachey}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.a}When magic numbers become magic code}{46}{subsection.10.1}\protected@file@percent }
+\citation{henderson}
+\citation{paper-usesRMarkdown}
+\citation{RMarkdown}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.b}When data is code}{47}{subsection.10.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.c}Exploiting code as data for more reliable science}{47}{subsection.10.3}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.d}When data is text: Exploiting code for reliable publication}{47}{subsection.10.4}\protected@file@percent }
+\citation{Spiegelhalter}
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.e}Data and polynomials used in the paper}{48}{subsection.10.5}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsection}{\numberline {10.f}Comparing conventional and RAP approaches}{48}{subsection.10.6}\protected@file@percent }
+\newlabel{supplement-RAP-diagrams}{{10.f}{48}{Comparing conventional and RAP approaches}{subsection.10.6}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {11}The Spiegelhalter trustworthiness questions}{48}{section.11}\protected@file@percent }
+\newlabel{supplement-Spiegelhalter-section}{{11}{48}{The Spiegelhalter trustworthiness questions}{section.11}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {2}{\ignorespaces The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
+Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code.}}{49}{figure.2}\protected@file@percent }
+\newlabel{supplement-fig:basic}{{2}{49}{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
+Note that the publication, including the results, can be edited arbitrarily, and typically the results published will have been edited and modified (if only for typographical purposes) from those actually generated by the code}{figure.2}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {3}{\ignorespaces When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
+The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
+Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
+In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication.
+}}{49}{figure.3}\protected@file@percent }
+\newlabel{supplement-fig:excel}{{3}{49}{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
+The data copied \& pasted into a publication may or may not be calculated from data in other cells (such as column totals).
+Records are rarely taken of these manual processes, and, anyway, typically it is impossible to be certain exactly what has been copied unless very great care is taken.
+In consequence, if a spreadsheet is modified, it is haphazard what results are updated and corrected in the publication}{figure.3}{}}
+\@writefile{lof}{\contentsline {figure}{\numberline {4}{\ignorespaces Improving over the normal approach (figure \ref {supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+In the present paper, this was done by generating text files of \LaTeX  \ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX  \ names for all of the code-generated values.
+Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
+Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen.}}{50}{figure.4}\protected@file@percent }
+\newlabel{supplement-fig:rap}{{4}{50}{Improving over the normal approach (figure \ref {supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
+In the present paper, this was done by generating text files of \LaTeX \ definitions (represented by boxes in the central column of the schematic), hence providing \LaTeX \ names for all of the code-generated values.
+Since the copying of results into a publication is automated and easy, any improvements to the results (such as correcting errors) are made by improving the code --- which therefore contributes to improving all future-generated results too.
+Not shown, but in well-engineered code, documentation will also be generated, such as by using tools such as JavaDoc or Doxygen}{figure.4}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.a}How trustworthy are the numbers?}{50}{subsection.11.1}\protected@file@percent }
+\@writefile{lof}{\contentsline {figure}{\numberline {5}{\ignorespaces In a notebook system, such as \emph  {Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
+The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
+Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
+In well-engineered code, the notebook will directly contain full documentation.
+Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc).}}{51}{figure.5}\protected@file@percent }
+\newlabel{supplement-fig:notebook}{{5}{51}{In a notebook system, such as \emph {Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
+The code generates results and images that are automatically (and reliably) inserted in place into the notebook, usually right after the code that generates them.
+Some parts of the notebook are marked or selected to be printed as the publication, thus allowing most if not all code to be hidden from the final publication.
+In well-engineered code, the notebook will directly contain full documentation.
+Note that in most systems, notebooks can import and generate arbitrary data (images, sounds, sensor data, etc)}{figure.5}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.b}How trustworthy is the source?}{51}{subsection.11.2}\protected@file@percent }
+\citation{lancet-retracted}
+\citation{science-lancet1,science-lancet2}
+\citation{lancet-learning}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.c}How trustworthy is the interpretation?}{52}{subsection.11.3}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}\protected@file@percent }
+\newlabel{supplement-survey-section}{{12}{52}{A pilot survey of computational science}{section.12}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{52}{subsection.12.1}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.1}\emph  {The Lancet\/}}{52}{subsubsection.12.1.1}\protected@file@percent }
+\citation{jvs1}
+\citation{jvs2,jvs3}
+\citation{jvs1}
+\citation{numerals}
+\citation{fda}
+\@writefile{lot}{\contentsline {table}{\numberline {6}{\ignorespaces Conforming to the RAP{\tt  +}\ methodology, the abbreviated summary above was generated automatically, by using \texttt  {make data}.
+A full list of \texttt  {make} options is generated by \texttt  {make} or \texttt  {make help} when run on the Unix command line.
+}}{53}{table.6}\protected@file@percent }
+\newlabel{supplement-table:options}{{6}{53}{Conforming to the \RAPstar \ methodology, the abbreviated summary above was generated automatically, by using \texttt {make data}.
+A full list of \texttt {make} options is generated by \texttt {make} or \texttt {make help} when run on the Unix command line}{table.6}{}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.2}\emph  {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}\protected@file@percent }
+\newlabel{supplement-vascular-surgery-note}{{12.a.2}{53}{\emph {Journal of Vascular Surgery}}{subsubsection.12.1.2}{}}
+\citation{ICmodel}
+\citation{ICmodel}
+\citation{ICmodel}
+\@writefile{lot}{\contentsline {table}{\numberline {7}{\ignorespaces Overview of the peer-reviewed paper sample.
+(Convenient copy of main paper's table \ref {table-overview}.)}}{54}{table.7}\protected@file@percent }
+\newlabel{supplement-table-overview}{{7}{54}{Overview of the peer-reviewed paper sample. (Convenient copy of main paper's table \ref {table-overview}.)}{table.7}{}}
+\@writefile{lot}{\contentsline {table}{\numberline {8}{\ignorespaces Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}}{54}{table.8}\protected@file@percent }
+\newlabel{supplement-table-summary}{{8}{54}{Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}{table.8}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{54}{subsection.12.2}\protected@file@percent }
+\citation{nature-review}
+\@writefile{lot}{\contentsline {table}{\numberline {9}{\ignorespaces Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt  {covid-sim} \cite  {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the Supplemental Material), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+At the time of cloning and checking all repositories in May 2023, paper [182]\ still had nothing in its repository except a single file still saying ``\ldots  \ code coming soon\nobreakspace  {}\ldots  ,'' despite 47\ months having already elapsed since the paper had claimed the code could be accessed in its repository.}}{55}{table.9}\protected@file@percent }
+\newlabel{supplement-table-repo-summary}{{9}{55}{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt {covid-sim} \cite {ICmodel} for comparison.
+Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement ), so choices were made in the survey to avoid double-counting.
+Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
+At the time of cloning and checking all repositories in \clonewhen , paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname \ still had nothing in its repository except a single file still saying ``\ldots \ code coming soon~\ldots ,'' despite \the \pubdelayinmonths \ months having already elapsed since the paper had claimed the code could be accessed in its repository}{table.9}{}}
+\citation{open-source}
+\citation{stability}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.c}Summary of results}{56}{subsection.12.3}\protected@file@percent }
+\newlabel{supplement-summary-of-results}{{12.c}{56}{Summary of results}{subsection.12.3}{}}
+\citation{RS-policy}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.1}Current code policies of sampled journals}{57}{subsubsection.12.3.1}\protected@file@percent }
+\newlabel{supplement-journal-policies-section}{{12.c.1}{57}{Current code policies of sampled journals}{subsubsection.12.3.1}{}}
+\citation{ref-1}
+\citation{ref-2}
+\citation{ref-3}
+\citation{ref-4}
+\citation{ref-5}
+\citation{ref-6}
+\citation{ref-7}
+\citation{ref-8}
+\citation{ref-9}
+\citation{ref-10}
+\citation{ref-11}
+\citation{ref-12}
+\citation{paper-usesRMarkdown}
+\citation{ref-14}
+\citation{onlyPaperWithChecks}
+\citation{ref-16}
+\citation{ref-17}
+\citation{ethics-paper}
+\citation{ref-19}
+\citation{ref-20}
+\citation{ref-21}
+\citation{ref-22}
+\citation{ref-23}
+\citation{example-numerical-error}
+\citation{ref-25}
+\citation{ref-26}
+\citation{ref-27}
+\citation{ref-28}
+\citation{ref-29}
+\citation{ref-30}
+\citation{ref-31}
+\citation{ref-32}
+\gdef \LT@i {\LT@entry 
+    {2}{26.55563pt}\LT@entry 
+    {1}{214.35622pt}\LT@entry 
+    {1}{208.35622pt}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{58}{subsubsection.12.3.2}\protected@file@percent }
+\citation{prisma}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.d}Assessment criteria and methods}{60}{subsection.12.4}\protected@file@percent }
+\newlabel{supplement-data-methods-subsection}{{12.d}{60}{Assessment criteria and methods}{subsection.12.4}{}}
+\citation{fixit}
+\citation{fixit}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.1}Detecting and defending against error}{62}{subsubsection.12.4.1}\protected@file@percent }
+\newlabel{supplement-detecting-against-error}{{12.d.1}{62}{Detecting and defending against error}{subsubsection.12.4.1}{}}
+\citation{excel-fiasco}
+\citation{latex-calculus}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.2}Defending against system problems}{63}{subsubsection.12.4.2}\protected@file@percent }
+\bibstyle{revisedCompJ}
+\bibdata{appendix.bib}
+\bibcite{vancouver}{{{130}{}{{}}{{}}}}
+\bibcite{cosmos}{{{131}{}{{}}{{}}}}
+\bibcite{ai-ethics}{{{132}{}{{}}{{}}}}
+\bibcite{science-lancet1}{{{133}{}{{}}{{}}}}
+\bibcite{science-lancet2}{{{134}{}{{}}{{}}}}
+\bibcite{lancet-learning}{{{135}{}{{}}{{}}}}
+\bibcite{jvs1}{{{136}{}{{}}{{}}}}
+\bibcite{jvs2}{{{137}{}{{}}{{}}}}
+\bibcite{jvs3}{{{138}{}{{}}{{}}}}
+\bibcite{swebok}{{{139}{}{{}}{{}}}}
+\bibcite{wpf}{{{140}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{64}{subsubsection.12.4.3}\protected@file@percent }
+\newlabel{supplement-restrictive-policies}{{12.d.3}{64}{Problems of restrictive journal policies}{subsubsection.12.4.3}{}}
+\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
+\bibcite{humphrey}{{{141}{}{{}}{{}}}}
+\bibcite{managers}{{{142}{}{{}}{{}}}}
+\bibcite{differential-equations}{{{143}{}{{}}{{}}}}
+\bibcite{numerals}{{{144}{}{{}}{{}}}}
+\bibcite{heedless}{{{145}{}{{}}{{}}}}
+\bibcite{plos}{{{146}{}{{}}{{}}}}
+\bibcite{excel1}{{{147}{}{{}}{{}}}}
+\bibcite{excel2}{{{148}{}{{}}{{}}}}
+\bibcite{npm}{{{149}{}{{}}{{}}}}
+\bibcite{hci1}{{{150}{}{{}}{{}}}}
+\bibcite{hci2}{{{151}{}{{}}{{}}}}
+\bibcite{flexibility}{{{152}{}{{}}{{}}}}
+\bibcite{iso}{{{153}{}{{}}{{}}}}
+\bibcite{178C}{{{154}{}{{}}{{}}}}
+\bibcite{peerReview}{{{155}{}{{}}{{}}}}
+\bibcite{ssi-report}{{{156}{}{{}}{{}}}}
+\bibcite{ssi-url}{{{157}{}{{}}{{}}}}
+\bibcite{security-engineering}{{{158}{}{{}}{{}}}}
+\bibcite{cyber-cacm}{{{159}{}{{}}{{}}}}
+\bibcite{Schneier}{{{160}{}{{}}{{}}}}
+\bibcite{electronic-evidence}{{{161}{}{{}}{{}}}}
+\bibcite{simon-artificial}{{{162}{}{{}}{{}}}}
+\bibcite{dunning-kruger}{{{163}{}{{}}{{}}}}
+\bibcite{dunning-kruger-numeracy}{{{164}{}{{}}{{}}}}
+\bibcite{enigma}{{{165}{}{{}}{{}}}}
+\bibcite{harking}{{{166}{}{{}}{{}}}}
+\bibcite{viruses}{{{167}{}{{}}{{}}}}
+\bibcite{strachey}{{{168}{}{{}}{{}}}}
+\bibcite{henderson}{{{169}{}{{}}{{}}}}
+\bibcite{lancet-retracted}{{{170}{}{{}}{{}}}}
+\bibcite{fda}{{{171}{}{{}}{{}}}}
+\bibcite{stability}{{{172}{}{{}}{{}}}}
+\bibcite{RS-policy}{{{173}{}{{}}{{}}}}
+\bibcite{latex-calculus}{{{174}{}{{}}{{}}}}
+\bibcite{ref-1}{{{175}{}{{}}{{}}}}
+\bibcite{ref-2}{{{176}{}{{}}{{}}}}
+\bibcite{ref-3}{{{177}{}{{}}{{}}}}
+\bibcite{ref-4}{{{178}{}{{}}{{}}}}
+\bibcite{ref-5}{{{179}{}{{}}{{}}}}
+\bibcite{ref-6}{{{180}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
+\bibcite{ref-7}{{{181}{}{{}}{{}}}}
+\bibcite{ref-8}{{{182}{}{{}}{{}}}}
+\bibcite{ref-9}{{{183}{}{{}}{{}}}}
+\bibcite{ref-10}{{{184}{}{{}}{{}}}}
+\bibcite{ref-11}{{{185}{}{{}}{{}}}}
+\bibcite{ref-12}{{{186}{}{{}}{{}}}}
+\bibcite{paper-usesRMarkdown}{{{187}{}{{}}{{}}}}
+\bibcite{ref-14}{{{188}{}{{}}{{}}}}
+\bibcite{onlyPaperWithChecks}{{{189}{}{{}}{{}}}}
+\bibcite{ref-16}{{{190}{}{{}}{{}}}}
+\bibcite{ref-17}{{{191}{}{{}}{{}}}}
+\bibcite{ethics-paper}{{{192}{}{{}}{{}}}}
+\bibcite{ref-19}{{{193}{}{{}}{{}}}}
+\bibcite{ref-20}{{{194}{}{{}}{{}}}}
+\bibcite{ref-21}{{{195}{}{{}}{{}}}}
+\bibcite{ref-22}{{{196}{}{{}}{{}}}}
+\bibcite{ref-23}{{{197}{}{{}}{{}}}}
+\bibcite{example-numerical-error}{{{198}{}{{}}{{}}}}
+\bibcite{ref-25}{{{199}{}{{}}{{}}}}
+\bibcite{ref-26}{{{200}{}{{}}{{}}}}
+\bibcite{ref-27}{{{201}{}{{}}{{}}}}
+\bibcite{ref-28}{{{202}{}{{}}{{}}}}
+\bibcite{ref-29}{{{203}{}{{}}{{}}}}
+\bibcite{ref-30}{{{204}{}{{}}{{}}}}
+\bibcite{ref-31}{{{205}{}{{}}{{}}}}
+\bibcite{ref-32}{{{206}{}{{}}{{}}}}
+\gdef \@abspage@last{34}
+% end expanding appendix.aux 
 \makeatother
@@
 \received{30 April 2022}
-%\revised{--}
+\revised{8 June 2022, 3 January 2023, 31 January 2023 \& 12~May~2023} 
+\volumenumber{00} \issuenumber{0} \volumeyear{2023}
 
@@
 \noindent
-Scientists rarely rigorously ensure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny.
-Even if available, scientists rarely provide adequate documentation to understand or use their code reliably.
-When code used in science avoids adequate scrutiny, published science will be unreliable and unverifiable because results --- data, graphs, images, inferences, and so forth --- informed or generated by the code will themselves be unreliable and unverifiable.
-
-\hskip 1em This paper proposes and justifies  ways to improve research and science relying on code:
+As code is now an inextricable part of science it should be supported by competent Software Engineering, analogously to statistical claims being properly supported by competent statistics.
+
+If and when code avoids adequate scrutiny, science becomes unreliable and unverifiable because results --- text, data, graphs, images, \emph{etc\/} --- depend on untrustworthy code.
+
+Currently, scientists rarely assure the quality of the code they rely on, and rarely make it accessible for scrutiny.
+Even when available, scientists rarely provide adequate documentation to understand or use it reliably.
+
+\hskip 1em This paper proposes and justifies  ways to improve science using code:
 
@@
 \noindent
-The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input.
 The paper's \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ workflows.
@@
 \vskip 1.2ex % align top of columns!
-\tableofcontents
+% expand table of contents paper.toc
+\makeatletter
+  \section*{\contentsname
+    \@mkboth{\MakeUppercase\contentsname}{\MakeUppercase\contentsname}}
+\makeatother\contentsline {section}{\numberline {1}Introduction}{1}{section.1}%
+\contentsline {section}{\numberline {2}Background}{2}{section.2}%
+\contentsline {subsection}{\numberline {2.a}Code quality concerns}{4}{subsection.2.1}%
+\contentsline {subsection}{\numberline {2.b}Computable papers}{5}{subsection.2.2}%
+\contentsline {subsection}{\numberline {2.c}RAP: Reproducible Analytical Pipelines}{7}{subsection.2.3}%
+\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{9}{subsubsection.2.3.1}%
+\contentsline {subsubsection}{\numberline {2.c.2}RAP as research}{9}{subsubsection.2.3.2}%
+\contentsline {section}{\numberline {3}The statistics/code analogy}{10}{section.3}%
+\contentsline {subsection}{\numberline {3.a}The siren call of over-fitting code}{11}{subsection.3.1}%
+\contentsline {section}{\numberline {4}The conventional role of code}{13}{section.4}%
+\contentsline {subsection}{\numberline {4.a}The deceptive simplicity of code}{14}{subsection.4.1}%
+\contentsline {subsection}{\numberline {4.b}The low status of coding}{15}{subsection.4.2}%
+\contentsline {subsection}{\numberline {4.c}The critical role of code is often ignored}{15}{subsection.4.3}%
+\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{15}{subsection.4.4}%
+\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}%
+\contentsline {section}{\numberline {5}State of the art}{17}{section.5}%
+\contentsline {subsection}{\numberline {5.a}Case study: Pandemic modeling}{17}{subsection.5.1}%
+\contentsline {subsection}{\numberline {5.b}Concerns with reproducibility}{18}{subsection.5.2}%
+\contentsline {subsection}{\numberline {5.c}Beyond pandemic modeling}{20}{subsection.5.3}%
+\contentsline {subsection}{\numberline {5.d}The lop-sided emphasis on data}{21}{subsection.5.4}%
+\contentsline {section}{\numberline {6}Rethinking science that uses code}{22}{section.6}%
+\contentsline {subsection}{\numberline {6.a}Software Engineering Boards}{22}{subsection.6.1}%
+\contentsline {subsection}{\numberline {6.b}Extending RAP to RAP{\tt +}}{23}{subsection.6.2}%
+\contentsline {subsection}{\numberline {6.c}The paper as a scientific laboratory}{24}{subsection.6.3}%
+\contentsline {subsection}{\numberline {6.d}Action must be interdisciplinary}{25}{subsection.6.4}%
+\contentsline {subsection}{\numberline {6.e}Methodological statements}{25}{subsection.6.5}%
+\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{26}{subsection.6.6}%
+\contentsline {subsection}{\numberline {6.g}Benefits beyond science}{27}{subsection.6.7}%
+\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}%
+\contentsline {section}{\numberline {7}Conclusions}{28}{section.7}%
+
+% end expanding table of contents paper.toc
+
 \vskip 3mm \hrule
@@
 \begin{center}\sf\normalsize
-\input generated/summary-table.tex
+% expand generated/summary-table.tex
+% Generated by programs/data.js
+\begin{tabular}{|rrrc|}\hline
+Number of papers sampled relying on code&32&100\%&\\\hline\hline
+\multicolumn{4}{|l|}{\textbf{Access to code}}\\
+Some or all code available&12&38\%&\\
+Some or all code in principle available on request&8&25\%&\\
+Requested code actually made available (within 2 years 11 months${}^{\star}$)&0&0\%& \\\hline
+ \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
+Evidence program designed rigorously&{0}&{0\%}&\\
+Evidence source code properly tested&{0}&{0\%}&\\
+Evidence of any tool-based development&{0}&{0\%}&\\
+Team or open source based development&{0}&{0\%}&\\
+Other methods, e.g., independent coding methods&{1}&{3\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Documentation and comments}}\\
+Substantial code documentation and comments&2&6\%&\\
+Comments explain some code intent&3&9\%&\\
+Procedural comments (e.g., author, date, copyright)&10&31\%&\\
+No usable comments&17&53\%&\\\hline
+\multicolumn{4}{|l|}{\textbf{Repository use}}\\
+Used code repository (e.g., GitHub)&9&28\%&\\
+Used data repository (e.g., Dryad or GitHub)&9&28\%&\\
+Empty repository&1&3\%&\\\hline
+\multicolumn{4}{|l|}{{\textbf{Evidence of documented processes}}}\\
+Evidence of RAP/\RAPstar\ or any other principles in use to support scrutiny&{0}&{0\%}&\\\hline
+\multicolumn{4}{|l|}{\textbf{Adherence to journal code policy (if any)}}\\
+Papers published in journals with code policies&26&81\%&\\
+Clear breaches of journal code policy (if any) & 11&42\%&(\emph{N}~=~26)\\
+\hline\end{tabular}\\
+\vskip 1ex ${}^{\star}$Time of 2 years 11 months is wait between code request and date of generating this table.% end expanding generated/summary-table.tex
 \end{center}\vskip -2ex
@@
 The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths.
-The Society of Research Software Engineering (\url{https://society-rse.org}) has been established to further promote research software engineer interests.
+The Society of Research Software Engineering (\url{society-rse.org}) has been established to further promote research software engineer interests.
 
@@
 While pure plagiarism, which is a form of misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have deceptive properties, the results are likely to be destructive.
-The Wakefield MMR fraud claiming to link vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted; this misconduct has been extraordinarily destructive \cite{MMR-fraud}.
+The notorious Wakefield MMR fraud claiming to link vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted; this misconduct has been extraordinarily destructive \cite{MMR-fraud}.
 
@@
 \begin{itemize}\raggedright
-
+\itemsep=1.5ex
 \item
@@
 \hskip -\wd0\box0
-Science is what we understand well enough to explain to a computer'' }
+Science is what we understand well enough to explain to a computer.'' }
 \\ \hfill from the foreword to $A=B$ \cite{a=b}
@@
 \begin{enumerate}\raggedright
+\itemsep=1.5ex
 \item Critical algorithmic steps may never be codified.
@@
 
-\input generated/over-fitting-code-section.tex
+% expand generated/over-fitting-code-section.tex
+\subsection{The siren call of over-fitting code}\label{over-fit}
+Poor code can generate plausible and possibly misleading results from \emph{any\/} data or theory, including fraudulent science.
+A temptation is that developing code to get ``good results'' becomes more important than the code's overall faithfulness to the real scientific phenomena, theoretical or empirical.
+% wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed!
+\def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
+\def\figureStarDetails{*}
+
+% wolframscript (if being used) unnecessarily prints the next 2 lines (generated by CellPrint), but they are needed because the text also ends up in notebook cells!
+\def\imageWidthCalculation{\multiply \imageWidth by 21 \divide \imageWidth by 28}
+\def\figureStarDetails{*}
+
+\begin{figure\figureStarDetails}[t]
+{ \newdimen \imageWidth 
+  \imageWidth=\textwidth
+  \imageWidthCalculation
+  \begin{center}\includegraphics[width=\imageWidth]{
+generated/mathematicaplot.jpg
+}\end{center}}
+\caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
+Shown here is notional sample of experimental 2D data (the dots), a linear least squares regression, and an exact polynomial model.
+The over-fitted polynomial model fits the sample \emph{exactly\/}, but since the experimental data is presumably subject to random error (indicated by the confidence interval, itself estimated) the linear model would generally be considered a better description of the experimental data.}
+\label{fig-overfit}
+\end{figure\figureStarDetails}
+
+In conventional modeling terms, successful computer programs are often \emph{over-fitted\/} to phenomena \cite{over-fit}.
+That is, instead of using code to rigorously challenge, test, and develop our models, we tinker with code adapting it to generate results closer and closer to our prejudices.
+The code then apparently confirms our science, since we fitted it to our preconceptions but not to the science.
+
+In general, an \emph{over-fitted model\/} fits a set of data closely, but contains more parameters than can be justified by the science.
+An over-fitted model fails to reliably predict results beyond the scope of the data it has been fitted to.
+Over-fitting is a well-known problem, but the point is that when code is used, over-fitting is done unconsciously by programmers adjusting the code --- its parameters, its structure, its embedded data, the calculations it performs --- that specifies the model.
+``A model over-fits if it is  more complex than another model that fits equally well'' \cite{over-fit}, is a criterion that describes almost every program!
+Programmers without the discipline and experience to manage the unlimited adaptability of code debug, alter, and extend their code to make it do what they think it should do.
+This becomes a vicious circle as the idea of what the science is becomes driven by the code.
+Rather than debugging code by improving its fit to the actual science, it gets debugged and extended to fit the expectations.
+
+The problems of over-fitting data may be visualized using a Real$\rightarrow$Real function of one variable (figure \ref{fig-overfit}).
+The code that generates an over-fitted curve seems to work very well: in the example shown, the over-fitted curve fits the sampled data exactly; indeed, the code used here will fit any new data exactly as well.
+But the code has a negligible ability to predict new data or to describe theories of the data, which is the point of modeling.
+The fact that over-fitted code seems to work well is deceptive.
+
+Looking specifically at the data plotted in figure \ref{fig-overfit} if, for the sake of argument, we assume the error in the data is normally distributed then the values the over-fitted code generates outside the range of the sample are improbable.
+For example, the basic linear model predicts
+$\hat{y}(0)=0.5$, 
+versus the extreme value predicted by the over-fitting code, 
+$\hat{y}(0)=-41.8$, 
+ even lies far outside the plot region shown in the figure.\footnote{The bounds of the confidence interval illustrated in figure \ref{fig-overfit} depend on assumptions about the distribution of the data; in this example, we are \emph{assuming}, perhaps because we know something more about the experiment, or thanks to Occam's Razor that a linear function is more likely than a high order polynomial.} Similar problems happen with interpolation rather than extrapolation, for instance around $x=4$.
+
+While over-fitting data is a well-known problem, the point for this paper is that \emph{code\/} itself can easily be over-fitted.
+Code can of course be over-fitted in more complex ways than can be illustrated with elementary polynomials, as here.
+Code over-fitting is much harder to recognize because there may be no simple graph plot, like figure \ref{fig-overfit}, to highlight the problems.
+Furthermore, almost all code is far more intricate than the two trivial polynomials used to illustrated figure \ref{fig-overfit}.
+
+Unfortunately, code in published science \emph{is\/} often over-fitted, and over-fitted in a way that is very hard to scrutinize.
+For example, in epidemiology (which is considered in section \ref{section-pandemic-modeling}) it is routine to use very complicated, large dynamical models parameterized with numerous social, cultural, health, demographic and geographical data.
+The parameterization is mixed between data files, data written explicitly into the code, and with conditionals and other structuring in the code to cover special cases.
+Indeed, many of the programs in the survey used comments to inactivate code, presumably indicating an unfinished tinkering approach to code development.\footnote{Of the \the\countUsesVersionControlRepository\ papers in the pilot survey that reported use of code repositories (covering % expand generated/kloc.tex
+\def\totalkLOC{182}%
+% end expanding generated/kloc.tex
+\totalkLOC\ thousand lines of code --- so this is not a trivial amount of programming effort), \numbertoname{\the\countMissingCode} provided an empty repository with no code at all (effectively commenting out all their code!), and \numbertoname{\the\countCommentedOut} repositories explicitly commented out chunks of workable code.
+The \numbertoname{\the\countNotCommentedOut} remaining non-trivial repositories with no commented-out code consisted of straightforward, short code files with few comments of any sort.\label{name-example-footnote}}
+
+Furthermore, scientific support code is rarely documented well enough to know what it should have been doing, which should be answered by a specification.
+With no clear specification and documentation, the code can be arbitrarily hacked to get any convenient results, since no particular specification for it has been defined that it should adhere to.
+Thus we risk doing and promoting substandard science because we --- the scientists and the publication process --- are not managing the unlimited adaptability and complexity of code that science has come to rely~on.
+This is over-fitting of the worst kind --- in conventional over-fitting one can at least hope to see that the fitting is over-parameterized for the data, but in code over-fitting the code and specification are not visible, therefore not adequately scrutinized, and --- worse --- the ``data'' the code over-fits includes the entire conceptual contribution of the paper.
+
+Reference \cite{may-simple} shows that even trivial code (in the case cited, implementing simple difference equations) with very few parameters can have very complex results, and reference \cite{dyson} is a historically significant paper pointing out how the problems of over-fitting has improved science.
+% end expanding generated/over-fitting-code-section.tex
 
@@
 In reality, if code is inadequate, any so-called ``insights'' will be potentially flawed, and flawed in unknown ways.
-Indeed, none of the papers sampled (\supplement\ section~\ref{supplement-survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
+Indeed, none of the papers sampled (see \supplement\ section~\ref{supplement-survey-section}) claimed their papers were using code for insight; all papers claimed, explicitly or implicitly, that their code outputs were integral to their peer-reviewed results.
 
@@
 The figure cited for Ferguson was obtained from Google Scholar on 20 January 2022.
-(Typical $h$ values vary by discipline.)} of 93, and ``one of the top scientists advising the government on its response to the coronavirus crisis'' working in the UK's Scientific Advisory Group for Emergencies (SAGE) group \cite{ferguson-resign}.
-
-Ferguson's code must be representative of best practice at a time when the stakes were high and reliability was known to be essential; and if not representative of best practice, at least representative of accepted practice both in Ferguson's team, the field of epidemiology more widely, as well as with members of the high-powered interdisciplinary SAGE group.
+(Typical $h$ values vary by discipline.)} of 93, and at the time ``one of the top scientists advising the government on its response to the coronavirus crisis'' working in the UK's Scientific Advisory Group for Emergencies (SAGE) group \cite{ferguson-resign}.
+
+Ferguson's code must be representative of best practice when the stakes were high and reliability was known to be essential; and if not representative of best practice, at least representative of accepted practice both in Ferguson's team, the field of epidemiology more widely, as well as with members of the high-powered interdisciplinary SAGE group.
 It is therefore instructive to explore the larger story around this science that uses code.
@@
 \begin{enumerate}\raggedright
+\itemsep=1.5ex
+
 \item 
@@
 
-\bibliography{paper-seb-main.bib}
+% expand bibliography paper.bbl
+\begin{thebibliography}{100}
+
+\bibitem{a=b}
+Petkovsek, M., Wilf, H. and Zeilberger, D. (1996) {\em {$A=B$}}, A~K~Peters.
+  Ltd.
+\newblock \url{www2.math.upenn.edu/}\midurl{~wilf/}\midurl{AeqB.html}
+
+\bibitem{write-for-life}
+Quindlen, A. (2022) {\em Write for your life}, Random House.
+
+\bibitem{abelson}
+Abelson, R.~P. (1995) {\em Statistics as Principled Argument}, Lawrence Erlbaum
+  Associates.
+
+\bibitem{nature-chatgpt}
+Editorial (2023) ``Tools such as {ChatGPT} threaten transparent science; here
+  are our ground rules for their use,'' {\em Nature}, \textbf{613}:612.
+\newblock \doi{10.1038/d41586-023-00191-1}
+
+\bibitem{sommerville}
+Sommerville, I. (2015) {\em Software Engineering}, Pearson, 10th~ed.
+
+\bibitem{knight}
+Knight, J. (2012) {\em Fundamentals of Dependable Computing for Software
+  Engineers}, CRC Press.
+
+\bibitem{computational-thinking-wing}
+Wing, J.~M. (2008) ``Computational thinking and thinking about computing,''
+  {\em Philosophical Transactions of the {Royal Society A}: {M}athematical,
+  Physical and Engineering Sciences}, \textbf{366}(1881):3717--3725.
+\newblock \doi{10.1098/rsta.2008.0118}
+
+\bibitem{computational-thinking}
+McOwen, P.~W. and Curzon, P. (2020) {\em The Power of Computational Thinking},
+  World Scientific Publishing.
+
+\bibitem{ICmodel}
+Ferguson, N.~M. \emph{et al.} (16 March 2020) {\em Impact of non-pharmaceutical
+  interventions ({NPI}s) to reduce {COVID}-19 mortality and healthcare demand}.
+\newblock
+  \url{www.imperial.ac.uk/}\midurl{media/}\midurl{imperial-}\midurl{college/}\midurl{medicine/}\midurl{sph/}\midurl{ide/}\midurl{gida-}\midurl{fellowships/}\midurl{Imperial-}\midurl{College-}\midurl{COVID19-}\midurl{NPI-}\midurl{modelling-}\midurl{16-}\midurl{03-}\midurl{2020.pdf}
+
+\bibitem{avianFluModel}
+Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.
+  and Burke, D.~S. (2005) ``Strategies for containing an emerging influenza
+  pandemic in {Southeast Asia},'' {\em Nature}, \textbf{437}:209--214.
+\newblock \doi{10.1038/nature04017}
+
+\bibitem{originalICmodel}
+Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.
+  and Burke, D.~S. (2006) ``Strategies for mitigating an influenza pandemic,''
+  {\em Nature}, \textbf{442}:448--452.
+\newblock \doi{10.1038/nature04795}
+
+\bibitem{reproducibility-crisis}
+Baker, M. (2016) ``1,500 scientists lift the lid on reproducibility,'' {\em
+  Nature}, \textbf{533}(7604):452--454.
+\newblock \doi{10.1038/533452a}
+
+\bibitem{rescience}
+Rougier, N.~P. \emph{et al.} (2017) ``Sustainable computational science: the
+  {ReScience} initiative,'' {\em PeerJ Computer Science}, \textbf{3}(e142).
+\newblock \doi{10.7717/peerj-cs.142}
+
+\bibitem{temperature}
+Chang, H. (2007) {\em Inventing Temperature: Measurement and Scientific
+  Progress}, Oxford Studies in the Philosophy of Science.
+
+\bibitem{form}
+von Hippel, M. (2022) ``Crucial computer program for particle physics at risk
+  of obsolescence,'' {\em Quanta Magazine}.
+\newblock
+  \url{www.quantamagazine.org/}\midurl{crucial-}\midurl{computer-}\midurl{program-}\midurl{for-}\midurl{particle-}\midurl{physics-}\midurl{at-}\midurl{risk-}\midurl{of-}\midurl{obsolescence-}\midurl{20221201}
+
+\bibitem{acm-algorithms}
+Bemer, R.~W. (1958) ``Techniques department: Policy statement,'' {\em
+  Communications of the ACM}, \textbf{1}(1):5--7.
+
+\bibitem{hoare-correctness}
+Hoare, C. A.~R. (2007) ``The ideal of program correctness: {Third} {Computer}
+  {Journal} lecture,'' {\em The Computer Journal}, \textbf{50}(3):254--260.
+\newblock \doi{10.1093/comjnl/bxl078}
+
+\bibitem{jupyter-study}
+Pimentel, J.~F., Murta, L., Braganholo, V. and Freire, J. (2019) ``A
+  large-scale study about quality and reproducibility of {Jupyter} notebooks,''
+  {\em in 2019 IEEE/ACM 16th International Conference on Mining Software
+  Repositories {(MSR)}}, pp.~507--517.
+\newblock \doi{10.1109/MSR.2019.00077}
+
+\bibitem{trisovic}
+Trisovic, A., Lau, M.~K., Pasquier, T. and Crosas, M. (2022) ``A large-scale
+  study on research code quality and execution nature research,'' {\em
+  Scientific Data}, \textbf{9}(60).
+\newblock \doi{10.1038/s41597-022-01143-6}
+
+\bibitem{machine-learning-reproducibility}
+Thimbleby, H. (9 May, 2004) ``Give your computer's {IQ} a boost ---
+  {{\emph{{Journal of Machine Learning Research}}}},'' {\em Times Higher
+  Education Supplement}.
+\newblock
+  \url{www.timeshighereducation.co.uk/}\midurl{story.asp?sectioncode=26&storycode=176549}
+
+\bibitem{no-raw-data}
+Miyakawa, T. (2022) ``No raw data, no science: {Another} possible source of the
+  reproducibility crisis,'' {\em Molecular Brain}, \textbf{13}(24):1--6.
+\newblock \doi{10.1186/s13041-020-0552-2}
+
+\bibitem{data-access}
+Gabelica, M., Boj{\v c}i{\' c}, R. and Puljak, L. (2020) ``Many researchers
+  were not compliant with their published data sharing statement: mixed-methods
+  study,'' {\em Journal of Clinical Epidemiology}.
+\newblock \doi{0.1016/j.jclinepi.2022.05.019}
+
+\bibitem{reproducible-manifesto}
+Munaf{\`o}, M.~R., Nosek, B.~A., Bishop, D. V.~M., Button, K.~S., Chambers,
+  C.~D., Percie~du Sert, N., Simonsohn, U., Wagenmakers, E.-J., Ware, J.~J. and
+  Ioannidis, J. P.~A. (2017) ``A manifesto for reproducible science,'' {\em
+  Nature Human Behaviour}, \textbf{1}(1):0021.
+\newblock \doi{10.1038/s41562-016-0021}
+
+\bibitem{joss}
+Smith, A.~M. \emph{et al.} (2018) ``Journal of open source software ({JOSS}):
+  Design and first-year review,'' {\em PeerJ Computer Science},
+  \textbf{4}(e147).
+\newblock \doi{10.7717/peerj-cs.147}
+
+\bibitem{TOP}
+Nosek, B.~A. \emph{et al.} (2015) ``Promoting an open research culture:
+  {A}uthor guidelines for journals could help to promote transparency,
+  openness, and reproducibility,'' {\em Science},
+  \textbf{348}(6242):1422--1425.
+\newblock \doi{10.1126/science.aab2374}
+
+\bibitem{TOP-wiki}
+Alter, G. \emph{et al.} (2022) {\em Guidelines for {Transparency} and
+  {Openness} {Promotion} {(TOP)} in journal policies and practices ``{The}
+  {TOP} {Guidelines}''}.
+\newblock \url{osf.io/}\midurl{9f6gx/}\midurl{wiki/}\midurl{Guidelines}
+
+\bibitem{MMR-fraud}
+Godlee, F., Smith, J. and Marcovitch, H. (2011) ``Wakefield's article linking
+  {MMR} vaccine and autism was fraudulent,'' {\em BMJ}, \textbf{342}(c7452).
+\newblock \doi{10.1136/bmj.c7452}
+
+\bibitem{misconduct-issues}
+Fanelli, D. (2009) ``How many scientists fabricate and falsify research? {A}
+  systematic review and meta-analysis of survey data,'' {\em {PLOS ONE}},
+  \textbf{4}(5):e5738.
+\newblock \doi{10.1371/journal.pone.0005738}
+
+\bibitem{elns}
+Machina, H.~K. and Wild, D.~J. (2013) ``Electronic laboratory notebooks
+  progress and challenges in implementation,'' {\em Journal of Laboratory
+  Automation}, \textbf{18}(4):264--268.
+\newblock \doi{0.1177/2211068213484471}
+
+\bibitem{perkel-review}
+Perkel, J.~M. (6 May 2021) ``Reactive, reproducible, collaborative:
+  Computational notebooks evolve,'' {\em Nature}, \textbf{593}:156--157.
+\newblock \doi{10.1038/d41586-021-01174-w}
+
+\bibitem{maneage}
+Akhlaghi, M., Infante-Sainz, R., Roukema, B.~F., Khellat, M., Valls-Gabaud, D.
+  and Baena-Gall{\'e}, R. (May 2021) ``Toward long-term and archivable
+  reproducibility,'' {\em Computing in Science {\&} Engineering},
+  \textbf{23}(3):82--91.
+\newblock \url{https:/}\midurl{/}\midurl{maneage.org},
+  \doi{10.1109/mcse.2021.3072860}
+
+\bibitem{LP}
+Knuth, D.~E. (1984) ``Literate programming,'' {\em The Computer Journal},
+  \textbf{27}(2):97--111.
+\newblock \doi{10.1093/comjnl/27.2.97}
+
+\bibitem{LP-book}
+\leavevmode\vrule height 2pt depth -1.6pt width 23pt\  (1992) {\em Literate
+  programming}, vol.~\textbf{27} of {CSLI} Lecture Notes, Center for the Study
+  of Language and Information Publication, Stanford, {CA}.
+
+\bibitem{relit}
+Thimbleby, H. and Williams, D. (2018) ``A tool for publishing reproducible
+  algorithms {\&} {A} reproducible, elegant algorithm for sequential
+  experiments,'' {\em Science of Computer Programming}, \textbf{156}:45--67.
+\newblock \url{GitHub.com/}\midurl{haroldthimbleby/}\midurl{relit},
+  \doi{10.1016/j.scico.2017.12.010}
+
+\bibitem{notebook-patent}
+Gray, T.~W. and Wolfram, S. (26 March 2013) {\em Method and system for
+  presenting input expressions and evaluations of the input expressions on a
+  workspace of a computational system}, US patent no{\@.} 8,407,580~B2.
+
+\bibitem{jupyter-history}
+Granger, B.~E. and P{\'e}rez, F. (2021) ``Jupyter: {Thinking} and storytelling
+  with code and data,'' {\em Computing in Science {\&} Engineering},
+  \textbf{23}(2):7--14.
+\newblock \doi{10.1109/MCSE.2021.3059263}
+
+\bibitem{knitr}
+Xie, Y. (2015) {\em Dynamic Documents with {R} and knitr}, CRC, second~ed.
+
+\bibitem{spec-led-design}
+Thimbleby, H. (1999) ``Specification-led design for interface simulation,
+  collecting use-data, interactive help, writing manuals, analysis, comparing
+  alternative designs, etc,'' {\em Personal Technologies},
+  \textbf{4}(2):241--254.
+\newblock \url{harold.thimbleby.net/}\midurl{ansim}, \doi{10.1007/BF01885563}
+
+\bibitem{RAPstandard}
+{Office for National Statistics} (2022) {\em Using reproducible analytical
+  pipelines ({RAP}) to improve statistics}.
+\newblock
+  \url{https:/}\midurl{/}\midurl{code.statisticsauthority.gov.uk/}\midurl{case-}\midurl{studies/}\midurl{using-}\midurl{reproducible-}\midurl{analytical-}\midurl{pipelines-}\midurl{rap-}\midurl{to-}\midurl{improve-}\midurl{statistics/}\midurl{}
+
+\bibitem{rap}
+Upson, M. (2017) {\em Reproducible analytical pipelines}.
+\newblock
+  \url{dataingovernment.blog.gov.uk/}\midurl{2017/}\midurl{03/}\midurl{27/}\midurl{reproducible-}\midurl{analytical-}\midurl{pipeline}
+
+\bibitem{goldacre}
+Goldacre, B. (2022) {\em Better, broader, safer: {Using} health data for
+  research and analysis}, Department of Health and Social Care.
+\newblock
+  \url{www.gov.uk/}\midurl{government/}\midurl{publications/}\midurl{better-}\midurl{broader-}\midurl{safer-}\midurl{using-}\midurl{health-}\midurl{data-}\midurl{for-}\midurl{research-}\midurl{and-}\midurl{analysis}
+
+\bibitem{turing-way}
+Ainsworth, R. \emph{et al.} (July 2022) {\em {The Turing Way: A Handbook for
+  Reproducible Data Science}}, vol.~\textbf{v1.0.3}, Zenodo.
+\newblock \url{the-}\midurl{turing-}\midurl{way.netlify.app/}\midurl{welcome},
+  \doi{10.5281/zenodo.3233853}
+
+\bibitem{Courtes}
+Court{\`e}s, L. (2020) ``[re] storage tradeoffs in a collaborative backup
+  service for mobile devices,'' {\em Rescience C}, \textbf{6}(1:{\#}6):10.
+\newblock
+  \url{https:/}\midurl{/}\midurl{gitlab.inria.fr/}\midurl{lcourtes-}\midurl{phd/}\midurl{edcc-}\midurl{2006-}\midurl{redone},
+  \doi{10.5281/zenodo.3886739}
+
+\bibitem{apa-statistics}
+Glen, S. (2022) {\em Reporting Statistics APA Style}, Statistics How To.
+\newblock
+  \url{www.statisticshowto.com/}\midurl{probability-}\midurl{and-}\midurl{statistics/}\midurl{reporting-}\midurl{statistics-}\midurl{apa-}\midurl{style}
+
+\bibitem{Mariusz-reporting}
+Cicho{\'n}, M. (2020) ``Reporting statistical methods and outcome of
+  statistical analyses in research articles,'' {\em Pharmacological Reports},
+  \textbf{72}(3):481--485.
+\newblock \doi{10.1007/s43440-020-00110-5}
+
+\bibitem{science-statistics-recommendations}
+ (2022) {\em Science journals: editorial policies}.
+\newblock
+  \url{https:/}\midurl{/}\midurl{www.science.org/}\midurl{content/}\midurl{page/}\midurl{science-}\midurl{journals-}\midurl{editorial-}\midurl{policies}
+
+\bibitem{example-stats}
+Richards, D. \emph{et al.} (2020) ``A pragmatic randomized waitlist-controlled
+  effectiveness and cost-effectiveness trial of digital interventions for
+  depression and anxiety,'' {\em Nature Digital Medicine}, \textbf{3}(85).
+\newblock \doi{10.1038/s41746-020-0293-8}
+
+\bibitem{Spiegelhalter}
+Spiegelhalter, D. (2019) {\em The Art of Statistics: Learning from Data},
+  Pelican Books.
+
+\bibitem{paul-hci}
+Cairns, P. (2007) ``{HCI}{\ldots} not as it should be: inferential statistics
+  in {HCI} research,'' {\em in BCS-HCI '07: Proceedings of the 21st {British}
+  {HCI} {Group} Annual {Conference on People and Computers}: {HCI}{\ldots} but
+  not as we know it}, vol.~\textbf{1}, pp.~195--201.
+
+\bibitem{pnas-stats}
+Johnson, V.~E. (2013) ``Revised standards for statistical evidence,'' {\em
+  Proceedings National Academy of Sciences}, \textbf{110}(48):19313--19317.
+\newblock \doi{10.1073/pnas.1313476110}
+
+\bibitem{Ben}
+Shneiderman, B. (2016) ``Opinion: {The} dangers of faulty, biased, or malicious
+  algorithms requires independent oversight,'' {\em Proceedings National
+  Academy of Sciences}, \textbf{113}(48):13538--13540.
+\newblock \doi{10.1073/pnas.1618211113}
+
+\bibitem{se-bias}
+Friedman, B. and Nissenbaum, H. (1996) ``Bias in computer systems,'' {\em ACM
+  Transactions on Information Systems}, \textbf{14}(3):330--347.
+\newblock \doi{10.1145/230538.230561}
+
+\bibitem{sas9.2}
+Laurain, E., Ayav, C., Erpelding, M.-L., Kessler, M., Brian{\c{c}}on, S.,
+  Brunaud, L. and Frimat, L. (2022) ``Targets for parathyroid hormone in
+  secondary hyperparathyroidism: is a ``one-size-fits-all'' approach
+  appropriate? {A} prospective incident cohort study,'' {\em BMC nephrology},
+  \textbf{15}:132.
+\newblock \doi{10.1186/1471-2369-15-132}
+
+\bibitem{whitty}
+Whitty, C. J.~M. (2015) ``What makes an academic paper useful for health
+  policy?'' {\em BMC Medicine}, \textbf{13}:301.
+\newblock \doi{10.1186/s12916-015-0544-8}
+
+\bibitem{over-fit}
+Hawkins, D.~M. (2004) ``The problem of overfitting,'' {\em Journal of Chemical
+  Information and Modeling}, \textbf{44}:1--12.
+\newblock \doi{10.1021/ci0342472}
+
+\bibitem{may-simple}
+May, R.~M. (1976) ``Simple mathematical models with very complicated
+  dynamics,'' {\em Nature}, \textbf{261}:459--467.
+\newblock \doi{10.1038/261459a0}
+
+\bibitem{dyson}
+Dyson, F. (2004) ``A meeting with {Enrico Fermi},'' {\em Nature},
+  \textbf{427}:297.
+\newblock \doi{10.1038/427297a}
+
+\bibitem{assurance-case}
+Habli, I., Alexander, R., Hawkins, R., Sujan, M., McDermid, J., Picardi, C. and
+  Lawton, T. (2020) ``Enhancing {COVID-19} decision making by creating an
+  assurance case for epidemiological models,'' {\em BMJ Health {\&} Care
+  Informatics}, \textbf{27}(e100165):1--5.
+\newblock \doi{10.1136/bmjhci-2020-100165}
+
+\bibitem{assessing-quality}
+Kelly, D. and Sanders, R. (2008) {\em Assessing the Quality of Scientific
+  Software}, First International Workshop on Software Engineering For
+  Computational Science and Engineering {(see
+  \cite{assessing-quality-conference}}), Leipzig.
+\newblock
+  \url{citeseerx.ist.psu.edu/}\midurl{viewdoc/}\midurl{summary?doi=10.1.1.526.5076}
+
+\bibitem{fixit}
+Thimbleby, H. (2021) {\em Fix IT: How to see and solve the problems of digital
+  healthcare}, Oxford University Press.
+
+\bibitem{Hammurabi}
+Roth, M.~T. (2022) {\em Laws of Hammurabi}, Open Educational Resources for the
+  Ancient Near East, University of Central Florida.
+\newblock
+  \url{https:/}\midurl{/}\midurl{stars.library.ucf.edu/}\midurl{cgi/}\midurl{viewcontent.cgi?article=1133&=&context=ancientneareast&=&sei-}\midurl{redir=1&referer=https%253A%252F%252Fscholar.google.com%252Fscholar%253Fhl%253Den%2526as_sdt%253D0%25252C5%2526q%253DIf%252Ba%252Bbuilder%252Bconstructs%252Ba%252Bhouse%252Bhammurabi%2526btnG%253D#search=%22If%20builder%20constructs%20house%20hammurabi%22}
+
+\bibitem{cite-code}
+Katz, D.~S. \emph{et al.} (2021) ``Recognizing the value of software: {A}
+  software citation guide [version 2; peer review: 2 approved] {Previously}
+  titled: ``{The} importance of software citation''{\hskip 0pt},'' {\em
+  F1000Research}, \textbf{9}(1257).
+\newblock \doi{10.12688/f1000research.26932.2}
+
+\bibitem{prisma}
+Page, M.~J. \emph{et al.} (2021) ``The {PRISMA} 2020 statement: {An} updated
+  guideline for reporting systematic reviews,'' {\em Systematic Reviews},
+  \textbf{10}(89):1--11.
+\newblock \doi{10.1186/s13643-021-01626-4}
+
+\bibitem{engima}
+Thimbleby, H. (2016) ``Human factors and missed solutions to {Enigma} design
+  weaknesses,'' {\em Cryptologia}, \textbf{40}(2):177--202.
+\newblock \doi{10.1080/01611194.2015.1028680}
+
+\bibitem{critiques}
+Sayburn, A. (2020) ``Covid-19: {Experts} question analysis suggesting half {UK}
+  population has been infected,'' {\em BMJ}, \textbf{368}:m1216.
+\newblock \doi{10.1136/bmj.m1216}
+
+\bibitem{diagnosis-reviews}
+Wynants, L. \emph{et al.} (2020) ``Prediction models for diagnosis and
+  prognosis of covid-19 infection: {Systematic} review and critical
+  appraisal,'' {\em BMJ}, \textbf{369}(m1328).
+\newblock \doi{10.1136/bmj.m1328}
+
+\bibitem{knuth}
+Knuth, D.~E. (1998) {\em The Art of Computer Programming (Seminumerical
+  Algorithms)}, vol.~\textbf{2}, Addison-Wesley, 3rd~ed.
+
+\bibitem{essence-of-software}
+Jackson, D. (2021) {\em The Essence of Software}, Princeton University Press.
+
+\bibitem{popper-conjectures-refutations}
+Popper, K.~R. (2002) {\em Conjectures and Refutations: {The} Growth of
+  Scientific Knowledge}, Routledge, 2nd~ed.
+
+\bibitem{preventable-book}
+Sridhar, D. (2022) {\em Preventable: {How} a Pandemic Changed the World {\&}
+  How to Stop the Next One}, Viking.
+
+\bibitem{science-review}
+Heesterbeek, H. \emph{et al.} (2015) ``Modeling infectious disease dynamics in
+  the complex landscape of global health,'' {\em Science},
+  \textbf{347}(6227):265--270.
+\newblock \doi{10.1126/science.aaa4339}
+
+\bibitem{tripod}
+Moons, K.~G., Altman, D.~G., Reitsma, J.~B., Ioannidis, J.~P., Macaskill, P.,
+  Steyerberg, E.~W., Vickers, A.~J., Ransohoff, D.~F. and Collins, G.~S. (2015)
+  ``Transparent reporting of a multivariable prediction model for individual
+  prognosis or diagnosis ({TRIPOD}): {Explanation} and elaboration,'' {\em
+  Annals of Internal Medicine}, \textbf{162}(1):W1--73.
+\newblock \doi{10.7326/M14-0698}
+
+\bibitem{nature-summary}
+Adam, D. (2020) ``Modelling the pandemic: {The} simulations driving the world's
+  response to {COVID}-19,'' {\em Nature}, \textbf{580}:316--318.
+\newblock \doi{10.1038/d41586-020-01003-6}
+
+\bibitem{tweet}
+Ferguson, N. (22 March 2020) {\em Tweet}.
+\newblock
+  \url{twitter.com/}\midurl{neil_ferguson/}\midurl{status/}\midurl{1241835454707699713}
+
+\bibitem{ferguson-interview}
+Leake, J. (29 March 2020) ``{Neil Ferguson} interview: {No} 10's infection guru
+  recruits game developers to build coronavirus pandemic model,'' {\em The
+  Sunday Times}.
+\newblock
+  \url{www.thetimes.co.uk/}\midurl{article/}\midurl{neil-}\midurl{ferguson-}\midurl{interview-}\midurl{no-}\midurl{10s-}\midurl{infection-}\midurl{guru-}\midurl{recruits-}\midurl{game-}\midurl{developers-}\midurl{to-}\midurl{build-}\midurl{coronavirus-}\midurl{pandemic-}\midurl{model-}\midurl{zl5rdtjq5}
+
+\bibitem{basic-reproducibilty}
+Hinsen, K. (2013) ``Software development for reproducible research,'' {\em
+  Computing in Science {\&} Engineering}, \textbf{15}(4):60--63.
+\newblock \doi{10.1109/MCSE.2013.91}
+
+\bibitem{ferguson-resign}
+Smith, B. (6 May 2020) ``{SAGE} adviser {Neil Ferguson }quits over coronavirus
+  lockdown breach,'' {\em Civil Service World}.
+\newblock
+  \url{www.civilserviceworld.com/}\midurl{professions/}\midurl{article/}\midurl{sage-}\midurl{adviser-}\midurl{neil-}\midurl{ferguson-}\midurl{quits-}\midurl{over-}\midurl{coronavirus-}\midurl{lockdown-}\midurl{breach#:~:text=SAGE%20adviser%20Neil%20Ferguson%20quits%20over%20coronavirus%20lockdown%20breach,-}\midurl{Epidemiologist%20says%20he&text=Professor%20Neil%20Ferguson%2C%20one%20of,resigned%20after%20breaching%20lockdown%20rules}
+
+\bibitem{forsyth}
+Ahrens, J.~H. and Dieter, U. (1973) ``Extensions of {Forsythe's} method for
+  random sampling from the normal distribution,'' {\em Mathematics of
+  Computation}, \textbf{27}:927--937.
+\newblock \doi{10.1090/S0025-5718-1973-0329190-8}
+
+\bibitem{Normal-Distributions}
+Ahrens, J.~H. and Dieter, U. (1972) ``Computer methods for sampling from the
+  exponential and normal distributions,'' {\em Communications of the ACM},
+  \textbf{15}(10):873--882.
+\newblock \doi{10.1145/355604.361593}
+
+\bibitem{misra}
+{The MISRA Consortium Limited} (2020) {\em {MISRA} Compliance: 2020 --
+  Achieving compliance with {MISRA} Coding Guidelines}, MISRA Consortium.
+\newblock \url{www.misra.org.uk}
+
+\bibitem{spark}
+Barnes, J. (2003) {\em High Integrity Software: The {SPARK} Approach to Safety
+  and Security}, Addison Wesley.
+
+\bibitem{haskell}
+O'Sullivan, B., Stewart, D. and Goerzen, J. (2008) {\em Real World Haskell},
+  O'Reilly Media.
+\newblock \url{book.realworldhaskell.org}
+
+\bibitem{thumbs-up}
+Chawla, D.~S. (8 June 2020) ``Critiqued coronavirus simulation gets thumbs up
+  from code-checking efforts,'' {\em Nature}, \textbf{582}:323--324.
+\newblock
+  \url{www.nature.com/}\midurl{articles/}\midurl{d41586-}\midurl{020-}\midurl{01685-}\midurl{y}
+
+\bibitem{codecheck}
+Scheuber, A. and {van Elsland}, S.~L. (1 June 2020) {\em Codecheck confirms
+  reproducibility of {COVID-19} model results}.
+\newblock
+  \url{www.imperial.ac.uk/}\midurl{news/}\midurl{197875/}\midurl{codecheck-}\midurl{confirms-}\midurl{reproducibility-}\midurl{covid-}\midurl{19-}\midurl{model-}\midurl{results}
+
+\bibitem{codecheck-certificate}
+Eglen, S.~J. (2020) {\em {CODECHECK} certificate 2020-010 for paper: {Report}
+  9: impact of non-pharmaceutical interventions ({NPIs}) to reduce {COVID-19}
+  mortality and healthcare demand}.
+\newblock
+  \url{github.com/}\midurl{codecheckers/}\midurl{covid-}\midurl{report9},
+  \doi{10.5281/zenodo.3865491}
+
+\bibitem{NVP}
+{Hatton}, L. and {Roberts}, A. (1994) ``How accurate is scientific software?''
+  {\em IEEE Transactions on Software Engineering}, \textbf{20}(10):785--797.
+\newblock \doi{10.1109/32.328993}
+
+\bibitem{nvp-ferguson}
+Halloran, M.~E. \emph{et al.} (2008) ``Modeling targeted layered containment of
+  an influenza pandemic in the {United States},'' {\em Proceedings of the
+  National Academy of Sciences}, \textbf{105}(12):4639--4644.
+\newblock
+  \url{www.pnas.org/}\midurl{content/}\midurl{105/}\midurl{12/}\midurl{4639},
+  \doi{10.1073/pnas.0706849105}
+
+\bibitem{notebooks}
+Schnell, S. (2015) ``Ten simple rules for a computational biologist's
+  laboratory notebook,'' {\em PLoS Computational Biology},
+  \textbf{11}(9):e1004385.
+\newblock \doi{10.1371/journal.pcbi.1004385}
+
+\bibitem{nih-policy}
+{National Institutes of Health} (2020; effective date January 25, 2023) {\em
+  Final {NIH} Policy for Data Management and Sharing},
+  vol.~\textbf{NOT-OD-21-013}, Office of The Director, National Institutes of
+  Health.
+\newblock
+  \url{grants.nih.gov/}\midurl{grants/}\midurl{guide/}\midurl{notice-}\midurl{files/}\midurl{NOT-}\midurl{OD-}\midurl{21-}\midurl{013.html}
+
+\bibitem{nih-nature}
+Kozlov, M. (16 February 2022) ``{NIH} issues a seismic mandate: {Share} data
+  publicly,'' {\em Nature}.
+\newblock
+  \url{www.nature.com/}\midurl{articles/}\midurl{d41586-}\midurl{022-}\midurl{00402-}\midurl{1}
+
+\bibitem{refactoring}
+Ferguson, N. (22 March 2020) {\em Tweet}.
+\newblock
+  \url{twitter.com/}\midurl{neil_ferguson/}\midurl{status/}\midurl{1241835456947519492}
+
+\bibitem{bad-code}
+Richards, D. and Boudnik, K. (16 May 2020) ``{Neil} {Ferguson's} {Imperial}
+  model could be the most devastating software mistake of all time,'' {\em The
+  Telegraph}.
+\newblock
+  \url{www.telegraph.co.uk/}\midurl{technology/}\midurl{2020/}\midurl{05/}\midurl{16/}\midurl{neil-}\midurl{fergusons-}\midurl{imperial-}\midurl{model-}\midurl{could-}\midurl{devastating-}\midurl{software-}\midurl{mistake}
+
+\bibitem{pseudo}
+Zlojutro, A., Rey, D. and Gardner, L. (2019) ``A decision-support framework to
+  optimize border control for global outbreak mitigation,'' {\em Nature
+  Scientific Reports}, \textbf{9}(2216).
+\newblock \doi{10.1038/s41598-019-38665-w}
+
+\bibitem{chinese}
+Thimbleby, H. (2003) ``The {Directed Chinese Postman} problem,'' {\em Software
+  --- Practice {\&} Experience}, \textbf{33}(11):1081--1096.
+\newblock \url{harold.thimbleby.net/}\midurl{cpp/}\midurl{index.html},
+  \doi{10.1002/spe.540}
+
+\bibitem{flu-model}
+Sander, B., Nizam, A., {Garrison Jr}, L.~P., Postma, M.~J., Halloran, M.~E. and
+  {Longini Jr}, I.~M. (2009) ``Economic evaluation of influenza pandemic
+  mitigation strategies in the {US} using a stochastic microsimulation
+  transmission model,'' {\em Value Health}, \textbf{12}(2):226--233.
+\newblock \doi{10.1111/j.1524-4733.2008.00437.x}
+
+\bibitem{nature-review}
+Perkel, J.~M. (2022) ``How to fix your scientific coding errors,'' {\em
+  Nature}, \textbf{602}:172--173.
+\newblock \doi{10.1038/d41586-022-00217-0}
+
+\bibitem{psychological-modeling}
+Guest, O. and Martin, A.~E. (2021) ``How computational modeling can force
+  theory building in psychological science,'' {\em Perspectives on
+  Psychological Science}, \textbf{16}(4):789--802.
+\newblock \doi{10.1177/1745691620970585}
+
+\bibitem{actr}
+{ACT-R Research Group} (2022) {\em {ACT-R}}.
+\newblock \url{act-}\midurl{r.psy.cmu.edu/}\midurl{software}
+
+\bibitem{ourReview}
+Ladkin, P.~B., Littlewood, B., Thimbleby, H. and Thomas, M. (2020) ``The {Law
+  Commission} presumption concerning the dependability of computer evidence,''
+  {\em Digital Evidence and Electronic Signature Law Review},
+  \textbf{17}:1--14.
+\newblock \doi{10.14296/deeslr.v17i0.5143}
+
+\bibitem{hamming}
+Hamming, R.~W. (1987) {\em Numerical Methods for Scientists and Engineers},
+  Dover Publications Inc.
+
+\bibitem{unfinished}
+Kissler, S.~M., Tedijanto, C., Goldstein, E., Grad, Y.~H. and Lipsitch, M.
+  (2020) ``Projecting the transmission dynamics of {SARS-CoV-2} through the
+  postpandemic period,'' {\em Science}, \textbf{368}(6493):860--868.
+\newblock \doi{10.1126/science.abb5793}
+
+\bibitem{lancet-unfinished}
+Verity \emph{et al.} (2020) ``Estimates of the severity of coronavirus disease
+  2019: {A} model-based analysis,'' {\em Lancet}, \textbf{20}(6):669--677.
+\newblock \doi{10.1016/S1473-3099(20)30243-7}
+
+\bibitem{hoare}
+Hoare, C. A.~R. (1969) ``An axiomatic basis for computer programming,'' {\em
+  Communications of the ACM}, \textbf{12}(10):576--580.
+\newblock \doi{10.1145/363235.363259}
+
+\bibitem{preclinical-reproducibility}
+Freedman, L.~P., Cockburn, I.~M. and Simcoe, T.~S. (2015) ``The economics of
+  reproducibility in preclinical research,'' {\em PLOS Biology},
+  \textbf{13}(6):e1002165.
+\newblock \doi{10.1371/journal.pbio.1002165}
+
+\bibitem{fair}
+Wilkinson, M.~D. \emph{et al.} (2016) ``The {FAIR} guiding principles for
+  scientific data management and stewardship,'' {\em Scientific Data},
+  \textbf{3}(160018):1--9.
+\newblock \doi{10.1038/sdata.2016.18}
+
+\bibitem{fair-principles}
+Wood-Charlson, E.~M., Crockett, Z., Erdmann, C., Arkin, A.~P. and Robinson,
+  C.~B. (2022) ``Ten simple rules for getting and giving credit for data,''
+  {\em PLoS Computational Biology}, \textbf{18}(9):e1010476.
+\newblock \doi{10.1371/ journal.pcbi.1010476}
+
+\bibitem{cbc}
+Woodcock, J. C.~P., Larsen, P.~G., Bicarregui, J.~C. and Fitzgerald, J.~S.
+  (2009) ``Formal methods: {Practice} and experience,'' {\em ACM Computing
+  Surveys}, \textbf{41}(4):1--36.
+\newblock \doi{10.1145/1592434.1592436}
+
+\bibitem{ethics-code}
+ACM (2020) {\em Code of Ethics and Professional Conduct}, ACM.
+\newblock Accessed 23 April 2020,
+  \url{www.acm.org/}\midurl{code-}\midurl{of-}\midurl{ethics}
+
+\bibitem{acm-artifacts}
+\leavevmode\vrule height 2pt depth -1.6pt width 23pt\  (2020) {\em Artifact
+  Review and Badging --- Current}, vol.~\textbf{Version 1.1}, ACM.
+\newblock
+  \url{www.acm.org/}\midurl{publications/}\midurl{policies/}\midurl{artifact-}\midurl{review-}\midurl{and-}\midurl{badging-}\midurl{current}
+
+\bibitem{redmill}
+Redmill, F. (2000) ``Understanding the use, misuse and abuse of safety
+  integrity levels,'' {\em in Lessons in System Safety, Eighth Safety-critical
+  Systems Symposium}.
+\newblock Revised,
+  \url{homepages.cs.ncl.ac.uk/}\midurl{felix.redmill/}\midurl{publications/}\midurl{1%20SILs.pdf}
+
+\bibitem{iec61508}
+{IEC Technical Committee TC 65} (2010) {\em {IEC 61508:2010 CMV} commented
+  version, functional safety of electrical/electronic/programmable electronic
+  safety-related systems}.
+\newblock \url{webstore.iec.ch/}\midurl{publication/}\midurl{22273}
+
+\bibitem{hoare-unreasonable}
+Hoare, C. A.~R. (1996) ``How did software get so reliable without proof?'' {\em
+  in Lecture Notes in Computer Science}, vol.~\textbf{1051}, Springer,
+  pp.~1--17.
+\newblock \doi{10.1007/3-540-60973-3_77}
+
+\bibitem{open-source}
+Fomel, S. (2015) ``Reproducible research as a community effort: {Lessons} from
+  the {Madagascar Project},'' {\em Computing in Science {\&} Engineering},
+  \textbf{17}(1):20--26.
+\newblock \doi{10.1109/MCSE.2014.94}
+
+\bibitem{ABCs-SE}
+Stol, K.-J. and Fitzgerald, B. (2018) ``The {ABC} of software engineering
+  research,'' {\em ACM Transactions on Software Engineering and Methodology},
+  \textbf{27}(3):1--51.
+\newblock \doi{10.1145/3241743}
+
+\bibitem{mrb}
+Lakens, D. (2023) ``Methods-review boards could avert wasted research,'' {\em
+  Nature}, \textbf{613}.
+
+\bibitem{Executable-Paper}
+Gabriela, A. and Capone, R. (2011) ``Executable paper {Grand Challenge}
+  workshop,'' {\em Procedia Computer Science}, \textbf{4}:577--578.
+\newblock \doi{10.1016/j.procs.2011.04.060}
+
+\bibitem{RMarkdown}
+Xie, Y., Allaire, J.~J. and Grolemund, G. (2020) {\em R Markdown: The
+  Definitive Guide}, Chapman {\&} Hall/CRC.
+
+\bibitem{lepton}
+Li-Thiao-T{\' e}, S. (2012) ``Literate program execution for reproducible
+  research and executable papers,'' {\em Procedia Computer Science},
+  \textbf{9}:439--448.
+\newblock International Conference on Computational Science, ICCS 2012,
+  \doi{10.1016/j.procs.2012.04.047}
+
+\bibitem{landauer}
+Landauer, T.~K. (1995) {\em The Trouble with Computers: {Usefulness},
+  Usability, and Productivity}, MIT Press.
+
+\bibitem{thimbleby-landauer}
+Thimbleby, H. (1996) ``The trouble with computers: {Usefulness}, usability, and
+  productivity (by {Thomas} {K.~Landauer}),'' {\em Computational Linguistics},
+  \textbf{22}(2):265--276.
+
+\bibitem{debt}
+Falessi, D. and Kruchten, P. (2015) ``Five reasons for including technical debt
+  in the software engineering curriculum,'' {\em in Proceedings of the 2015
+  European Conference on Software Architecture Workshops}, ECSAW'15, ACM,
+  pp.~28:1--28:4.
+\newblock \doi{10.1145/2797433.2797462}
+
+\bibitem{jpm}
+{Heineman, Jr.}, B.~W. (2013) ``The {JP Morgan} ``{Whale}'' report and the
+  ghosts of the financial crisis,'' {\em Harvard Business Review}.
+\newblock
+  \url{https:/}\midurl{/}\midurl{hbr.org/}\midurl{2013/}\midurl{01/}\midurl{the-}\midurl{jp-}\midurl{morgan-}\midurl{whale-}\midurl{report-}\midurl{and}
+
+\bibitem{jpm-report}
+Cavanagh, M., ed. (2013) {\em Report of JPMorgan Chase {\&} Co{\@}. Management
+  Task Force Regarding 2012 CIO Losses}, JPMorgan Chase {\&} Co{\@}.
+
+\bibitem{excel-fiasco}
+Thimbleby, H. (2020) ``The problem isn't {Excel}, it's unprofessional software
+  engineering,'' {\em BMJ}, \textbf{371}(m4181).
+\newblock \doi{10.1136/bmj.m4181}
+
+\bibitem{science-delays}
+Wadman, M. (2021) ``Could too much time between doses drive the coronavirus to
+  outwit vaccines?'' {\em Science}.
+\newblock \doi{10.1126/science.abg5655}
+
+\bibitem{reproducibility}
+Benureau, F. C.~Y. and Rougier, N.~P. (2018) ``Re-run, repeat, reproduce,
+  reuse, replicate: {Transforming} code into scientific contributions,'' {\em
+  Frontiers in Neuroinformatics}, \textbf{11}(69).
+\newblock \doi{10.3389/fninf.2017.00069}
+
+\bibitem{parliamentary-evidence}
+{House of Commons Science and Technology Committee} (16 December 2020) {\em The
+  {UK} response to covid-19: {Use} of scientific advice}.
+\newblock
+  \url{committees.parliament.uk/}\midurl{publications/}\midurl{4165/}\midurl{documents/}\midurl{41300/}\midurl{default}
+
+\bibitem{my-parliamentary-evidence}
+Thimbleby, H. (29 April 2020) {\em Written Evidence Submitted by {Harold
+  Thimbleby} to The {UK} response to covid-19: {Use} of scientific advice},
+  (C190005), {House of Commons Science and Technology Committee} {(see
+  \cite{parliamentary-evidence})}.
+\newblock
+  \url{committees.parliament.uk/}\midurl{work/}\midurl{91/}\midurl{default/}\midurl{publications/}\midurl{written-}\midurl{evidence/}\midurl{?SearchTerm=thimbleby}
+
+\bibitem{assessing-quality-conference}
+Carver, J.~C. (2009) ``First international workshop on software engineering for
+  computational science {\&} engineering,'' {\em Computing in Science {\&}
+  Engineering}, \textbf{11}(2):7--11.
+\newblock \doi{10.1109/MCSE.2009.30}
+
+\endofbibliography\end{thebibliography}
+
+% end expanding bibliography paper.bbl
 
@@
 \end{document}
-
-\def \allInOne {define this so we don't include header in supplemental material}
-\input paper-seb-supplementary-material.tex
-\end{document}

### 2023-06-06 generated/saved-files.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo
Preparing to upload to Computer Journal

--- 2023-05-12 generated/saved-files.tex
@@
+% Generated by programs/data.js
 allGitRepos.sh & Shell script generated to download all available 10 GitHub repositories\\
@@
 summary-table.tex & Short summary table\\
-supplementary-references.tex & Reference list for Supplementary Material\\
+survey-references.bbl & Reference list for Supplementary Material (bbl file)\\

### 2023-06-06 generated/summary-table.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo
Preparing to upload to Computer Journal

--- 2023-02-27 generated/summary-table.tex
@@
+% Generated by programs/data.js
 \begin{tabular}{|rrrc|}\hline
@@
 Some or all code in principle available on request&8&25\%&\\
-Requested code actually made available (within 2 years 8 months${}^{\star}$)&0&0\%& \\\hline
+Requested code actually made available (within 2 years 11 months${}^{\star}$)&0&0\%& \\\hline
  \multicolumn{4}{|l|}{{\textbf{Evidence of any software engineering practice}}}\\
@@
 \hline\end{tabular}\\
-\vskip 1ex ${}^{\star}$Time of 2 years 8 months is wait between code request and date of generating this table.
+\vskip 1ex ${}^{\star}$Time of 2 years 11 months is wait between code request and date of generating this table.

### 2023-06-06 paper.tex
Preparing to upload to Computer Journal

--- 2023-05-31 paper-seb-main.tex
@@
 
-\input paper-seb-macros.tex
+\input macros.tex
 \input generated/page-lengths.tex
@@
 \makeatletter
-      \input paper-seb-supplementary-material.aux 
+      \input appendix.aux 
 \makeatother
@@
 
-\bibliography{paper-seb-main.bib}
+\bibliography{paper.bib}
 

### 2023-06-06 programs/data.js

@@ -1202,7 +1202,7 @@ for (var i = 0; i < data.length; i++) {
     s += bibdata(d, "year") + ". DOI \\texttt{" + bibdata(d, "doi").trim() + "}" + (bibdata(d, "codeURL") != undefined && bibdata(d, "codeURL") != "" ? (" {Code \\url{" + bibdata(d, "codeURL").trim() + "}}") : "") + "\\\\\\hfill{Accessed " + bibdata(d, "accessed") + ".}\\ " + (!definedq(bibdata(d, "doubleChecked")) || bibdata(d, "doubleChecked") == "" ? "\\textcolor{red}{Not double-checked}" : ("{" +
         "Double-checked " + bibdata(d, "doubleChecked") + "}")) + ".}\\bibskip\n\n";
 }
-saveFile("supplementary-references.bbl", s, "Reference list for Supplementary Material (bbl file)");
+saveFile("survey-references.bbl", s, "Reference list for Supplementary Material (bbl file)");
 
 // generate summary table
 var t = {};

### 2023-06-06 programs/expand.js

@@ -1,9 +1,24 @@
-// Harold Thimbleby, 2020-2022
-// Javascript code to expand TeX \input commands
-
-var TeXFiles = ["paper-seb-main.tex", "paper-seb-supplementary-material.tex"];
-
+// Harold Thimbleby, 2020-2023
+// Javascript code to recursively expand TeX \input commands in files TeXFiles[] (see below for value)
+// run as node generated/expand.js
 'use strict';
+
+var TeXFiles = ["paper.tex", "appendix.tex"];
+
+function makenote(vector) { 
+    var files = "", expandedfiles = "";
+    for( var i = 0; i < vector.length; i++ )
+    {	files += i == 0? "": i < vector-1? ", ": ", and "; 
+        files += vector[i];
+        expandedfiles += i == 0? "": i < vector-1? ", ": ", and "; 
+        expandedfiles += "expanded-"+vector[i];
+    }
+	var plural = vector.length > 1;
+	return ("LaTeX file-noun ("+files+")\n  to make "+(plural?"":"this ")+"straight forward LaTeX file-noun ("+expandedfiles+")\n    that depend-verb on nothing ("+(plural?"they":"it")+" input-verb no files apart from the document class files).\n      Each required file's text was input in situ - search for % expand\n\n(Otherwise the file-noun "+files+" would use BibTeX, a macro file, tables of contents, "+(plural?"their":"its")+" own and the other files's .aux files, and include-verb various files in generated/*, etc.)\n").replace(/-noun/mg, plural?"s":"").replace(/-verb/mg, plural?"":"s");
+}
+
+console.log("This app expands complex "+makenote(TeXFiles));
+// console.log("This app expands the complex "+makenote([TeXFiles[0]]));
 
 var fs = require('fs');
 
@@ -35,15 +50,21 @@ function getFile(file) {
     }
 }
 
-var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}");
+var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}"), recontents=RegExp("\\\\tableofcontents");
 
-var bufferExpand, bufferInsert, auxfile;
+var contentsStuff = "\\makeatletter\n  \\section*{\\contentsname\n    \\@mkboth{\\MakeUppercase\\contentsname}{\\MakeUppercase\\contentsname}}\n\\makeatother";
 
[…]

### 2023-06-07 expanded-appendix.tex
Final updates for Computer Journal

@@
+% Please do not edit this file, expanded-appendix.tex!
+% It was generated automatically from its master file appendix.tex, by programs/expand.js
+% Get everything from https://github.com/haroldthimbleby/improving-science
 % 
-% This file has been expanded from LaTeX file (appendix.tex)
-%   to make this straight forward LaTeX file (expanded-appendix.tex)
-%     that depends on nothing (it inputs no files apart from the document class files).
-%       Each required file's text was input in situ - search for % expand
+% This file expands the LaTeX file (appendix.tex), which has complex dependencies,
+%   to make this simple LaTeX file (expanded-appendix.tex)
+%     that depends on nothing, apart from its standard document class.
+%       This expanded file depends on no other files, as instead
+%         each required file's text was input and expanded recursively - search for '% expand' in the expanded file below.
 % 
-% (Otherwise the file appendix.tex would use BibTeX, a macro file, tables of contents, its own and the other files's .aux files, and includes various files in generated/*, etc.)
+% (The original file is dependent on BibTeX (.bst and .bbl files), a macro file, table of contents,
+%   its own and other files' .aux files,
+%     as well as various data files in the directory generated
+%       etc. All these file dependencies have been removed.)
 %
@@
 \citation{fda}
+\citation{ICmodel}
+\citation{ICmodel}
+\citation{ICmodel}
 \@writefile{lot}{\contentsline {table}{\numberline {6}{\ignorespaces Conforming to the RAP{\tt  +}\ methodology, the abbreviated summary above was generated automatically, by using \texttt  {make data}.
@@
 \newlabel{supplement-vascular-surgery-note}{{12.a.2}{53}{\emph {Journal of Vascular Surgery}}{subsubsection.12.1.2}{}}
-\citation{ICmodel}
-\citation{ICmodel}
-\citation{ICmodel}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{53}{subsection.12.2}\protected@file@percent }
 \@writefile{lot}{\contentsline {table}{\numberline {7}{\ignorespaces Overview of the peer-reviewed paper sample.
@@
 \newlabel{supplement-table-summary}{{8}{54}{Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}{table.8}{}}
-\@writefile{toc}{\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{54}{subsection.12.2}\protected@file@percent }
 \citation{nature-review}
@@
 Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.
-At the time of cloning and checking all repositories in May 2023, paper [182]\ still had nothing in its repository except a single file still saying ``\ldots  \ code coming soon\nobreakspace  {}\ldots  ,'' despite 47\ months having already elapsed since the paper had claimed the code could be accessed in its repository.}}{55}{table.9}\protected@file@percent }
+At the time of cloning and checking all repositories in June 2023, paper [182]\ still had nothing in its repository except a single file still saying ``\ldots  \ code coming soon\nobreakspace  {}\ldots  ,'' despite 48\ months having already elapsed since the paper had claimed the code could be accessed in its repository.}}{55}{table.9}\protected@file@percent }
 \newlabel{supplement-table-repo-summary}{{9}{55}{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt {covid-sim} \cite {ICmodel} for comparison.
@@
 \@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.2}Defending against system problems}{63}{subsubsection.12.4.2}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{63}{subsubsection.12.4.3}\protected@file@percent }
+\newlabel{supplement-restrictive-policies}{{12.d.3}{63}{Problems of restrictive journal policies}{subsubsection.12.4.3}{}}
 \bibstyle{revisedCompJ}
@@
 \bibcite{wpf}{{{140}{}{{}}{{}}}}
-\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{64}{subsubsection.12.4.3}\protected@file@percent }
-\newlabel{supplement-restrictive-policies}{{12.d.3}{64}{Problems of restrictive journal policies}{subsubsection.12.4.3}{}}
-\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
 \bibcite{humphrey}{{{141}{}{{}}{{}}}}
 \bibcite{managers}{{{142}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
 \bibcite{differential-equations}{{{143}{}{{}}{{}}}}
@@
 \bibcite{ref-5}{{{179}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
 \bibcite{ref-6}{{{180}{}{{}}{{}}}}
-\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
 \bibcite{ref-7}{{{181}{}{{}}{{}}}}
@@
 \newcount \dataVariableCount
-\dataVariableCount = 63
+\dataVariableCount = 62
 % end expanding generated/metadata.tex
@@
 % date generated by running generated/allGitRepos.sh downloading Git repos
-\def\clonedate {\ignorespaces 25 May 2023}
+\def\clonedate {\ignorespaces 7 June 2023}
 \def\cloneyear {2023}
-\def\clonemonth {05}
+\def\clonemonth {06}
 % end expanding generated/clone-date.tex
@@
 \contentsline {subsubsection}{\numberline {12.a.2}\emph {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}%
-\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{54}{subsection.12.2}%
+\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{53}{subsection.12.2}%
 \contentsline {subsection}{\numberline {12.c}Summary of results}{56}{subsection.12.3}%
@@
 \contentsline {subsubsection}{\numberline {12.d.2}Defending against system problems}{63}{subsubsection.12.4.2}%
-\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{64}{subsubsection.12.4.3}%
+\contentsline {subsubsection}{\numberline {12.d.3}Problems of restrictive journal policies}{63}{subsubsection.12.4.3}%
 \contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}%
@@
 \texttt{make all}&
-      Analyze the data, then typeset the main PDF files (\texttt{paper-seb-main.pdf} and \texttt{paper-seb-supplementary-material.pdf}).\\
+      Download and analyze the data, then typeset the main files (\texttt{paper.pdf}, \texttt{appendix.pdf} and \texttt{all.pdf}) that import all the data, as well as make the self-contained expanded \texttt{.tex} (and \texttt{.PDF}) files that do not depend on the separate data files in \texttt{data/*}.\\
    \multicolumn{1}{l@{\vdots}}{}&\\
@@
 
-\texttt{make git-prep}&
-      What's on Git that we've lost, or stuff we have got locally but probably don't want on Git, so you can delete it or move it out the way or whatever.\\
-   \multicolumn{1}{l@{\vdots}}{}&\\
-
-\texttt{make really-tidyup}&
-      More thorough than \texttt{make tidyup} --- remove \emph{all} files that can be recreated.
-(This will mean next time you run \LaTeX\ you will have to ignore errors as the .aux files are re-created.)\\
+\texttt{make help}&
+      Explain how to use \texttt{make}, and list all available options for using it.\\
 \end{tabular}}
@@
 \cite{paper-usesRMarkdown} & Data available on Dryad.\flagStyle{ $\sf R_d$\hskip 3pt{}} & Code and example runs available in R Markdown.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_3$\hskip 3pt{}}\\
-\cite{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
+\cite{ref-14} & Data directly written into program code.\flagStyle{ } & Basic Matlab with routine comments.
+Supplemental files Rsos192210supp1.docx and Rsos192210supp2.docx have links but were not accessible at time of double checking.\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf P_{{\mbox{\scriptsize c-breach}}}$\hskip 3pt{}$\sf S_{+}$\hskip 3pt{}$\sf C_0$\hskip 3pt{}}\\
 \cite{onlyPaperWithChecks} & Data available on Dryad plus publicly available data from the 1000 genomes project.
@@
 \bibitem{science-lancet1}
-Servick, K. and Enserink, M. (2020) ``A mysterious company's coronavirus papers
-  in top medical journals may be unraveling,'' {\em Science}.
+Servick, K., and Enserink, M. (2020) ``A mysterious company's coronavirus
+  papers in top medical journals may be unraveling,'' {\em Science}.
 \newblock \doi{10.1126/science.abd1337}
@@
 Hardouin, S., Cheng, T.~W., Mitchell, E.~L., Raulli, S.~J., Jones, D.~W.,
-  Siracuse, J.~J. and Farber, A. (2020) ``Prevalence of unprofessional social
+  Siracuse, J.~J., and Farber, A. (2020) ``Prevalence of unprofessional social
   media content among young vascular surgeons,'' {\em Journal of Vascular
@@
 \bibitem{swebok}
-Bourque, P. and Fairley, R., eds. (2014) {\em Guide to the Software Engineering
-  Body of Knowledge}, (Version 3.0), IEEE Computer Society.
+Bourque, P., and Fairley, R., eds. (2014) {\em Guide to the Software
+  Engineering Body of Knowledge}, (Version 3.0), IEEE Computer Society.
 \newblock \url{www.swebok.org}
@@
 \bibitem{differential-equations}
-Boldo, S., Cl{\'{e}}ment, F., Filli{\^{a}}tre, J.-C., Mayero, M., Melquiond, G.
-  and Weis, P. (2014) ``Trusting computations: {A} mechanized proof from
+Boldo, S., Cl{\'{e}}ment, F., Filli{\^{a}}tre, J.-C., Mayero, M., Melquiond,
+  G., and Weis, P. (2014) ``Trusting computations: {A} mechanized proof from
   partial differential equations to actual program,'' {\em Computers and
@@
 \bibitem{numerals}
-Thimbleby, H. and Cairns, P. (2017) ``Interactive numerals,'' {\em Royal
+Thimbleby, H., and Cairns, P. (2017) ``Interactive numerals,'' {\em Royal
   Society Open Science}, \textbf{4}(4):160903.
@@
 \bibitem{plos}
-Chao, D.~L., Halloran, M.~E., Obenchain, V.~J. and {Longini Jr}, I.~M. (2010)
+Chao, D.~L., Halloran, M.~E., Obenchain, V.~J., and {Longini Jr}, I.~M. (2010)
   ``{FluTE}, a publicly available stochastic influenza epidemic simulation
@@
 \bibitem{excel1}
-Abir, M., Nelson, C., Chan, E.~W., Al-Ibrahim, H., Cutter, C., Patel, K. and
+Abir, M., Nelson, C., Chan, E.~W., Al-Ibrahim, H., Cutter, C., Patel, K., and
   Bogart, A. (2020) {\em {RAND} Critical Care Surge Response Tool: {An}
@@
 \bibitem{excel2}
-Alvarez, N.~M., Gonzalez-Gonzalez, E. and Trujillo-de Santiago, G. (2020)
+Alvarez, N.~M., Gonzalez-Gonzalez, E., and Trujillo-de Santiago, G. (2020)
   ``Modeling {COVID-19} epidemics in an excel spreadsheet: {Democratizing} the
@@
 \bibitem{hci1}
-Shneiderman, B., Plaisant, C., Cohen, M. and {\emph{et al}} (2016) {\em
+Shneiderman, B., Plaisant, C., Cohen, M., and {\emph{et al}} (2016) {\em
   Designing the User Interface: Strategies for Effective Human-Computer
@@
 \bibitem{flexibility}
-Hanson, C. and Sussman, G.~J. (2021) {\em Software design for flexibility: How
+Hanson, C., and Sussman, G.~J. (2021) {\em Software design for flexibility: How
   to avoid programming yourself into a corner}, MIT Press.
@@
 \bibitem{peerReview}
-Baum, T., {Le\ss mann}, H. and Schneider, K. (2017) ``The choice of code review
-  process: A survey on the state of the practice,'' {\em in Lecture Notes in
-  Computer Science}, vol.~\textbf{10611} of Product-Focused Software Process
-  Improvement: 18th International Conference, pp.~111--127.
+Baum, T., {Le\ss mann}, H., and Schneider, K. (2017) ``The choice of code
+  review process: A survey on the state of the practice,'' {\em in Lecture
+  Notes in Computer Science}, vol.~\textbf{10611} of Product-Focused Software
+  Process Improvement: 18th International Conference, pp.~111--127.
 \newblock \doi{10.1007/978-3-319-69926-4_9}
@@
 Brett, A., Croucher, M., Haines, R., Hettrick, S., Hetherington, J., Stillwell,
-  M. and Wyatt, C. (2017) {\em State of the Nation Report for Research Software
-  Engineers}, Research Software Engineer Network.
+  M., and Wyatt, C. (2017) {\em State of the Nation Report for Research
+  Software Engineers}, Research Software Engineer Network.
 \newblock \url{zenodo.org/}\midurl{record/}\midurl{495360#.Xyfuyi2ZOCM},
@@
 \bibitem{cyber-cacm}
-Shostack, A. and Zurko, M.~E. (2020) ``Secure development tools and techniques
+Shostack, A., and Zurko, M.~E. (2020) ``Secure development tools and techniques
   need more research that will increase their impact and effectiveness in
@@
 \bibitem{electronic-evidence}
-Mason, S. and Seng, D. (2017) {\em Electronic Evidence}, Humanities Digital
+Mason, S., and Seng, D. (2017) {\em Electronic Evidence}, Humanities Digital
   Library, 4th~ed.
@@
 \bibitem{simon-artificial}
-Simon, H.~A. and Laird, J. (2019) {\em Sciences of the Artificial}, MIT Press,
+Simon, H.~A., and Laird, J. (2019) {\em Sciences of the Artificial}, MIT Press,
   3rd~ed.
@@
 \bibitem{dunning-kruger}
-Kruger, J. and Dunning, D. (1999) ``Unskilled and unaware of it: {How}
+Kruger, J., and Dunning, D. (1999) ``Unskilled and unaware of it: {How}
   difficulties in recognizing one's own incompetence lead to inflated
@@
 \bibitem{dunning-kruger-numeracy}
-Nuhfer, E., Fleisher, S., Cogan, C., Wirth, K. and Gaze, E. (2017) ``How random
-  noise and a graphical convention subverted behavioral scientists'
+Nuhfer, E., Fleisher, S., Cogan, C., Wirth, K., and Gaze, E. (2017) ``How
+  random noise and a graphical convention subverted behavioral scientists'
   explanations of self-assessment data: {Numeracy} underlies better
@@
 \bibitem{viruses}
-Thimbleby, H., Anderson, S.~O. and Cairns, P. (1999) ``A framework for
+Thimbleby, H., Anderson, S.~O., and Cairns, P. (1999) ``A framework for
   modelling {Trojans} and computer virus infection,'' {\em Computer Journal},
@@
 \bibitem{lancet-retracted}
-Mehra, M.~R., Desai, S.~S., Ruschitzka, F. and Patel, A.~N. (2020)
+Mehra, M.~R., Desai, S.~S., Ruschitzka, F., and Patel, A.~N. (2020)
   ``{RETRACTED}: {Hydroxychloroquine} or chloroquine with or without a
@@
 \bibitem{fda}
-Zhang, Y., Masci, P., Jones, P. and Thimbleby, H. (2019) ``User interface
+Zhang, Y., Masci, P., Jones, P., and Thimbleby, H. (2019) ``User interface
   software errors in medical devices,'' {\em Biomedical Instrumentation {\&}
@@
 \bibitem{stability}
-Salama, M., Bahsoon, R. and Lago, P. (2021) ``Stability in software
+Salama, M., Bahsoon, R., and Lago, P. (2021) ``Stability in software
   engineering: Survey of the state-of-the-art and research directions,'' {\em
@@
 \vbox{\bibitem{ref-1}
-{\sc A.
-Callahan, E.
-Steinberg, J.
-A.
-Fries, S.
-Gombar, B.
-Patel, C.
-K.
-Corbin, and N.
-H.
-Shah}, ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp, 2020.
+{Callahan, A., Steinberg, E., Fries, J.
+A., Gombar, S., Patel, B., Corbin, C.
+K., and Shah, N.
+H.} (2020) ``Estimating the efficacy of symptom-based screening for COVID-19,'' \emph{Nature Digital Medicine}, \textbf{3}(95):3pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1038/s41746-020-0300-0}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
@@
 \vbox{\maprepo{MetricSelectionFramework}\bibitem{ref-2}
-{\sc C.
-M.
-Kanzler, M.
-D.
-Rinderknecht, A.
-Schwarz, I.
-Lamers, C.
-Gagnon, J.
+{Kanzler, C.
+M., Rinderknecht, M.
+D., Schwarz, A., Lamers, I., Gagnon, C., Held, J.
 P.
-O.
-Held, P.
-Feys, A.
-R.
-Luft, R.
-Gassert, and O.
-Lambercy}, ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp, 2020.
-DOI \texttt{10.1038/s41746-020-0286-7} {Code \url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+O., Feys, P., Luft, A.
+R., Gassert, R., and Lambercy, O.} (2020) ``A data-driven framework for selecting and validating digital health metrics: use-case in neurological sensorimotor impairments,'' \emph{Nature Digital Medicine}, \textbf{3}(80):17pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0286-7}
+
+{\url{github.com/ChristophKanzler/MetricSelectionFramework}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
 \vbox{\bibitem{ref-3}
-{\sc A.
-Roy, K.
-Nikolitch, R.
-McGinn, S.
-Jinah, W.
-Klement, and Z.
-A.
-Kaminsky}, ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp, 2020.
+{Roy, A., Nikolitch, K., McGinn, R., Jinah, S., Klement, W., and Kaminsky, Z.
+A.} (2020) ``A machine learning approach predicts future risk to suicidal ideation from social media data,'' \emph{Nature Digital Medicine}, \textbf{3}(78):12pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1038/s41746-020-0287-6}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
@@
 \vbox{\bibitem{ref-4}
-{\sc D.
-M.
-Levine, Z.
-Co, L.
-P.
-Newmark, A.
-R.
-Groisser, A.
-J.
-Holmgren, J.
-A.
-Haas, and D.
-W.
-Bates}, ``Design and testing of a mobile health application rating tool.,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp, 2020.
+{Levine, D.
+M., Co, Z., Newmark, L.
+P., Groisser, A.
+R., Holmgren, A.
+J., Haas, J.
+A., and Bates, D.
+W.} (2020) ``Design and testing of a mobile health application rating tool.,'' \emph{Nature Digital Medicine}, \textbf{3}(74):7pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1038/s41746-020-0268-9}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
@@
 \vbox{\bibitem{ref-5}
-{\sc T.
-Kannampallil, J.
-M.
-Smyth, S.
-Jones, P.
-R.
-O.
-Payne, and J.
-Ma}, ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp, 2020.
+{Kannampallil, T., Smyth, J. M., Jones, S., Payne, P. R. O., and Ma, J.} (2020) ``Cognitive plausibility in voice-based AI health counselors,'' \emph{Nature Digital Medicine}, \textbf{3}(72):4pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1038/s41746-020-0278-7}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
@@
 \vbox{\maprepo{PENet}\bibitem{ref-6}
-{\sc S.
-Huang, T.
-Kothari, I.
-Banerjee, C.
-Chute, R.
-L.
-Ball, N.
-Borus, A.
-Huang, B.
-N.
-Patel, P.
-Rajpurkar, J.
-Irvin, J.
-Dunnmon, J.
-Bledsoe, K.
-Shpanskaya, A.
-Dhaliwal, R.
-Zamanian, A.
-Y.
-Ng, and M.
-P.
-Lungren}, ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp, 2020.
-DOI \texttt{10.1038/s41746-020-0266-y} {Code \url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+{Huang, S., Kothari, T., Banerjee, I., Chute, C., Ball, R.
+L., Borus, N., Huang, A., Patel, B.
+N., Rajpurkar, P., Irvin, J., Dunnmon, J., Bledsoe, J., Shpanskaya, K., Dhaliwal, A., Zamanian, R., Ng, A.
+Y., and Lungren, M.
+P.} (2020) ``PENet a scalable deep-learning model for automated diagnosis of pulmonary embolism using volumetric CT imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(61):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0266-y}
+
+{\url{github.com/marshuang80/PENet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
 
 \vbox{\bibitem{ref-7}
-{\sc S.
+{Dhruva, S.
+S., Ross, J.
+S., Akar, J.
+G., Caldwell, B., Childers, K., Chow, W., Ciaccio, L., Coplan, P., Dong, J., Dykhoff, H.
+J., Johnston, S., Kellogg, T., Long, C., Noseworthy, P.
+A., Roberts, K., Saha, A., Yoo, A., and Shah, N.
+D.} (2020) ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
+{Hofer, I.
+S., Lee, C., Gabel, E., Baldi, P., and Cannesson, M.} (2020) ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0248-0}
+
+{\url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
+{Norgeot, B., Muenzen, K., Peterson, T.
+A., Fan, X., Glicksberg, B.
+S., Schenk, G., Rutenberg, E., Oskotsky, B., Sirota, M., Yazdany, J., Schmajuk, G., Ludwig, D., Goldstein, T., and Butte, A.
+J.} (2020) ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0258-y}
+
+{\url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
+{Choi, D., Park, J. J., Ali, T., and Lee, S.} (2020) ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0261-3}
+
+{\url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-11}
+{Hilton, C.
+B., Milinovich, A., Felix, C., Vakharia, N., Crone, T., Donovan, C., Proctor, A., and Nazha, A.} (2020) ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
+
+\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
+{Li, M.
+D., Chang, K., Bearce, B., Chang, B.
+Y., Huang, A.
+J., Campbell, J.
+P., Brown, J.
+M., Singh, P., Hoebel, K.
+V., Erdo{\u g}mu\c{s}, D., Ioannidis, S., Palmer, W., Chiang, M.
+F., and Kalpathy-Cramer, J.} (2020) ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1038/s41746-020-0255-1}
+
+{\url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
+{Hoffman, J.
+I., Nagel, R., Litzke, V., Wells, D.
+A., and Amos, W.} (2020) ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200419}
+
+{\url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
+{Gr\"onquist, P., Panchadcharam, P., Wood, D., Menges, A., R\"uggeberg, M., and Wittel, F.
+K.} (2020) ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.192210}
+
+{\url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{onlyPaperWithChecks}
+{Amos, W.} (2020) ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.191900}
+
+{\url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
+{Gordon, M., Viganola, D., Bishop, M., Chen, Y., Dreber, A., Goldfedder, B., Holzmeister, F., Johannesson, M., Liu, Y., Twardy, C., Wang, J., and Pfeiffer, T.} (2020) ``Are replication rates the same across academic fields?
+Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200566}
+
+{\url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
+{Evans, D., and Field, A.
+P.} (2020) ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200422}
+
+{\url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ethics-paper}
+{Beale, N., Battey, H., Davison, A. C., and MacKay, R. S.} (2020) ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-19}
+{Cherevko, A.
+A., Gologush, T.
+S., Petrenko, I.
+A., Ostapenko, V.
+V., and Panarin, V.
+A.} (2020) ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
+{Soczawa-Stronczyk, A. A., and Bocian, M.} (2020) ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200622}
+
+{\url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{lactModel}\bibitem{ref-21}
+{Duruz, S., Vajana, E., Burren, A., Flury, C., and Joost, S.} (2020) ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200638}
+
+{\url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-22}
+{de Azevedo, E.
+Z.
+D., Dantas, D.
+V., and Daura-Jorge, F.
+G.} (2020) ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{LRM}\bibitem{ref-23}
+{Abdolhosseini-Qomi, A.
+M., Jafari, S.
+H., Taghizadeh, A., Yazdani, N., Asadpour, M., and Rahgozar, M.} (2020) ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.191928}
+
+{\url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{1}\bibitem{example-numerical-error}
+{Webster, J., and Amos, M.} (2020) ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200307}
+
+{\url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-25}
+{Zhu, Y-l., Wang, C-J., Gao, F., Xiao, Z-x., Zhao, P-l., and Wang, J-y.} (2020) ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-26}
+{Yu, B., Scott, C.
+J., Xue, X., Yue, X., and Dou, X.} (2020) ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-27}
+{Joon-myoung, K., Younghoon, C., Ki-Hyun, J., Soohyun, C., Kyung-Hee, K., Seung D, B., Soomin, J., Jinsik, P., and Byung-Hee, O.} (2020) ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-28}
+{Zhu, H., Cheng, C., Yin, H., Li, X., Zuo, P., Ding, J., Lin, F., Wang, J., Zhou, B., Li, Y., Hu, S., Xiong, Y., Wang, B., Wan, G., Yang, X., and Yuan, Y.} (2020) ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
+{Fung, R., Villar, J., Dashti, A., Ismail, L.
+C., Staines-Urias, E., Ohuma, E.
+O., Salomon, L.
+J., Victora, C.
+G., Barros, F.
+C., Lambert, A., Carvalho, M., Jaffer Y, A., Noble, J.
+A., Gravett, M.
+G., Purwar, M., Pang, R., Bertino, E., Munim, S., Min, A.
+M., McGready, R., Norris, S.
+A., Bhutta, Z.
+A., Kennedy, S.
+H., Papageorghiou, A.
+T., and Ourmazd, A.} (2020) ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1016/S2589-7500(20)30131-X}
+
+{\url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
+
+\vbox{\bibitem{ref-30}
+{Sabanayagam, C., Xu, D., Ting, D.
 S.
-Dhruva, J.
-S.
-Ross, J.
-G.
-Akar, B.
-Caldwell, K.
-Childers, W.
-Chow, L.
-Ciaccio, P.
-Coplan, J.
-Dong, H.
-J.
-Dykhoff, S.
-Johnston, T.
-Kellogg, C.
-Long, P.
-A.
-Noseworthy, K.
-Roberts, A.
-Saha, A.
-Yoo, and N.
-D.
-Shah}, ``Aggregating multiple real-world data sources using a patient-centered health-data-sharing platform,'' \emph{Nature Digital Medicine}, \textbf{3}(60):9pp, 2020.
-DOI \texttt{10.1038/s41746-020-0265-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
-
-\vbox{\maprepo{PostoperativeOutcomes.RiskNet}\bibitem{ref-8}
-{\sc I.
-S.
-Hofer, C.
-Lee, E.
-Gabel, P.
-Baldi, and M.
-Cannesson}, ``Development and validation of a deep neural network model to predict postoperative mortality, acute kidney injury, and reintubation using a single feature set,'' \emph{Nature Digital Medicine}, \textbf{3}(58):10pp, 2020.
-DOI \texttt{10.1038/s41746-020-0248-0} {Code \url{github.com/cklee219/PostoperativeOutcomes_RiskNet}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
-
-\vbox{\maprepo{philter-ucsf}\bibitem{ref-9}
-{\sc B.
-Norgeot, K.
-Muenzen, T.
-A.
-Peterson, X.
-Fan, B.
-S.
-Glicksberg, G.
-Schenk, E.
-Rutenberg, B.
-Oskotsky, M.
-Sirota, J.
-Yazdany, G.
-Schmajuk, D.
-Ludwig, T.
-Goldstein, and A.
-J.
-Butte}, ``Protected Health Information filter (Philter): accurately and securely de-identifying free-text clinical notes,'' \emph{Nature Digital Medicine}, \textbf{3}(57):8pp, 2020.
-DOI \texttt{10.1038/s41746-020-0258-y} {Code \url{github.com/BCHSI/philter-ucsf}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
-
-\vbox{\maprepo{AI-CDSS-Cardiovascular-Silo}\bibitem{ref-10}
-{\sc D.
-Choi, J.
-J.
-Park, T.
-Ali, and S.
-Lee}, ``Artificial intelligence for the diagnosis of heart failure,'' \emph{Nature Digital Medicine}, \textbf{3}(54):6pp, 2020.
-DOI \texttt{10.1038/s41746-020-0261-3} {Code \url{github.com/ubiquitous-computing-lab/AI-CDSS-Cardiovascular-Silo}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-11}
-{\sc C.
-B.
-Hilton, A.
-Milinovich, C.
-Felix, N.
-Vakharia, T.
-Crone, C.
-Donovan, A.
-Proctor, and A.
-Nazha}, ``Personalized predictions of patient outcomes during and after hospitalization using artificial intelligence,'' \emph{Nature Digital Medicine}, \textbf{3}(51):8pp, 2020.
-DOI \texttt{10.1038/s41746-020-0249-z}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 17 January 2021}.}\bibskip
-
-\vbox{\maprepo{SiameseChange}\bibitem{ref-12}
-{\sc M.
-D.
-Li, K.
-Chang, B.
-Bearce, B.
-Y.
-Chang, A.
-J.
-Huang, J.
-P.
-Campbell, J.
-M.
-Brown, P.
-Singh, K.
-V.
-Hoebel, D.
-Erdo{\u g}mu\c{s}, S.
-Ioannidis, W.
-Palmer, M.
-F.
-Chiang, and J.
-Kalpathy-Cramer}, ``Siamese neural networks for continuous disease severity evaluation and change detection in medical imaging,'' \emph{Nature Digital Medicine}, \textbf{3}(48):9pp, 2020.
-DOI \texttt{10.1038/s41746-020-0255-1} {Code \url{github.com/QTIM-Lab/SiameseChange}}\\\hfill{Accessed 14 July 2020.}\ {Double-checked 19 January 2021}.}\bibskip
-
-\vbox{\maprepo{dryad.1g1jwstrw}\bibitem{paper-usesRMarkdown}
-{\sc J.
-I.
-Hoffman, R.
-Nagel, V.
-Litzke, D.
-A.
-Wells, and W.
-Amos}, ``Genetic analysis of \emph{Boletus edulis\/} suggests that intra-specific competition may reduce local genetic diversity as a woodland ages,'' \emph{Royal Society Open Science}, \textbf{7}(200419):13pp, 2020.
-DOI \texttt{10.1098/rsos.200419} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.1g1jwstrw}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{rsos.192210}\bibitem{ref-14}
-{\sc P.
-Gr\"onquist, P.
-Panchadcharam, D.
-Wood, A.
-Menges, M.
-R\"uggeberg, and F.
-K.
-Wittel}, ``Computational analysis of hygromorphic self-shaping wood gridshell structures,'' \emph{Royal Society Open Science}, \textbf{7}(192210):9pp, 2020.
-DOI \texttt{10.1098/rsos.192210} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}\bibitem{onlyPaperWithChecks}
-{\sc W.
-Amos}, ``Signals interpreted as archaic introgression appear to be driven primarily by faster evolution in Africa,'' \emph{Royal Society Open Science}, \textbf{7}(191900):9pp, 2020.
-DOI \texttt{10.1098/rsos.191900} {Code \url{datadryad.org/stash/share/ichHKrWj7hqlznOaR6NQVzITgp40dlqWvWAgAxyafiQ}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{rsos.200566}\bibitem{ref-16}
-{\sc M.
-Gordon, D.
-Viganola, M.
-Bishop, Y.
-Chen, A.
-Dreber, B.
-Goldfedder, F.
-Holzmeister, M.
-Johannesson, Y.
-Liu, C.
-Twardy, J.
-Wang, and T.
-Pfeiffer}, ``Are replication rates the same across academic fields?
-Community forecasts from the DARPA SCORE programme,'' \emph{Royal Society Open Science}, \textbf{7}(200566):7pp, 2020.
-DOI \texttt{10.1098/rsos.200566} {Code \url{royalsocietypublishing.org/doi/suppl/10.1098/rsos.200566}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{?view.only=87ae173f775b40d79d6cd0fdcf6d4a9c}\bibitem{ref-17}
-{\sc D.
-Evans, and A.
-P.
-Field}, ``Predictors of mathematical attainment trajectories across the primary-to-secondary education transition: parental factors and the home environment,'' \emph{Royal Society Open Science}, \textbf{7}(200422):20pp, 2020.
-DOI \texttt{10.1098/rsos.200422} {Code \url{osf.io/a5xsz/?view_only=87ae173f775b40d79d6cd0fdcf6d4a9c}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ethics-paper}
-{\sc N.
-Beale, H.
-Battey, A.
-C.
-Davison, and R.
-S.
-MacKay}, ``An unethical optimization principle,'' \emph{Royal Society Open Science}, \textbf{7}(200462):11pp, 2020.
-DOI \texttt{10.1098/rsos.200462}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-19}
-{\sc A.
-A.
-Cherevko, T.
-S.
-Gologush, I.
-A.
-Petrenko, V.
-V.
-Ostapenko, and V.
-A.
-Panarin}, ``Modelling of the arteriovenous malformation embolization optimal scenario,'' \emph{Royal Society Open Science}, \textbf{7}(191992):16pp, 2020.
-DOI \texttt{10.1098/rsos.191992}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{dryad.vx0k6djnr}\bibitem{ref-20}
-{\sc A.
-A.
-Soczawa-Stronczyk, and M.
-Bocian}, ``Gait coordination in overground walking with a virtual reality avatar,'' \emph{Royal Society Open Science}, \textbf{7}(200622):19pp, 2020.
-DOI \texttt{10.1098/rsos.200622} {Code \url{datadryad.org/stash/dataset/doi:10.5061/dryad.vx0k6djnr}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{lactModel}\bibitem{ref-21}
-{\sc S.
-Duruz, E.
-Vajana, A.
-Burren, C.
-Flury, and S.
-Joost}, ``Big dairy data to unravel effects of environmental, physiological and morphological factors on milk production of mountain-pastured Braunvieh cows,'' \emph{Royal Society Open Science}, \textbf{7}(200638):13pp, 2020.
-DOI \texttt{10.1098/rsos.200638} {Code \url{github.com/SolangeD/lactModel}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-22}
-{\sc E.
-Z.
-D.
-de Azevedo, D.
-V.
-Dantas, and F.
-G.
-Daura-Jorge}, ``Risk tolerance and control perception in a game-theoretic bioeconomic model for small-scale fisheries,'' \emph{Royal Society Open Science}, \textbf{7}(200621):11pp, 2020.
-DOI \texttt{10.1098/rsos.200621}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{LRM}\bibitem{ref-23}
-{\sc A.
-M.
-Abdolhosseini-Qomi, S.
-H.
-Jafari, A.
-Taghizadeh, N.
-Yazdani, M.
-Asadpour, and M.
-Rahgozar}, ``Link prediction in real-world multiplex networks via layer reconstruction method,'' \emph{Royal Society Open Science}, \textbf{7}(191928):22pp, 2020.
-DOI \texttt{10.1098/rsos.191928} {Code \url{github.com/UT-NSG/LRM}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{1}\bibitem{example-numerical-error}
-{\sc J.
-Webster, and M.
-Amos}, ``A Turing test for crowds,'' \emph{Royal Society Open Science}, \textbf{7}(200307):12pp, 2020.
-DOI \texttt{10.1098/rsos.200307} {Code \url{figshare.com/collections/Supplementary_information_for_Webster_J_and_Amos_M_A_Turing_Test_for_Crowds_/4859118/1}}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-25}
-{\sc Y-l.
-Zhu, C-J.
-Wang, F.
-Gao, Z-x.
-Xiao, P-l.
-Zhao, and J-y.
-Wang}, ``Calculation on surface energy and electronic properties of CoS${}_2$,'' \emph{Royal Society Open Science}, \textbf{7}(191653):12pp, 2020.
-DOI \texttt{10.1098/rsos.191653}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-26}
-{\sc B.
-Yu, C.
-J.
-Scott, X.
-Xue, X.
-Yue, and X.
-Dou}, ``Derivation of global ionospheric Sporadic E critical frequency ($f_o$Es) data from the amplitude variations in GPS/GNSS radio occultations,'' \emph{Royal Society Open Science}, \textbf{7}(200320):15pp, 2020.
-DOI \texttt{10.1098/rsos.200320}\\\hfill{Accessed 22 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-27}
-{\sc K.
-Joon-myoung, C.
-Younghoon, J.
-Ki-Hyun, C.
-Soohyun, K.
-Kyung-Hee, B.
-Seung D, J.
-Soomin, P.
-Jinsik, and O.
-Byung-Hee}, ``A deep learning algorithm to detect anaemia with ECGs: a retrospective, multicentre study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
-DOI \texttt{10.1016/S2589-7500(20)30108-4}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-28}
-{\sc H.
-Zhu, C.
-Cheng, H.
-Yin, X.
-Li, P.
-Zuo, J.
-Ding, F.
-Lin, J.
-Wang, B.
-Zhou, Y.
-Li, S.
-Hu, Y.
-Xiong, B.
-Wang, G.
-Wan, X.
-Yang, and Y.
-Yuan}, ``Automatic multilabel electrocardiogram diagnosis of heart rhythm or conduction abnormalities with deep learning: a cohort study,'' \emph{Lancet Digital Health}, \textbf{2}(7):9pp, 2020.
-DOI \texttt{10.1016/S2589-7500(20)30107-2}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\maprepo{manifold-ga}\bibitem{ref-29}
-{\sc R.
-Fung, J.
-Villar, A.
-Dashti, L.
-C.
-Ismail, E.
-Staines-Urias, E.
-O.
-Ohuma, L.
-J.
-Salomon, C.
-G.
-Victora, F.
-C.
-Barros, A.
-Lambert, M.
-Carvalho, A.
-Jaffer Y, J.
-A.
-Noble, M.
-G.
-Gravett, M.
-Purwar, R.
-Pang, E.
-Bertino, S.
-Munim, A.
-M.
-Min, R.
-McGready, S.
-A.
-Norris, Z.
-A.
-Bhutta, S.
-H.
-Kennedy, A.
-T.
-Papageorghiou, and A.
-Ourmazd}, ``Achieving accurate estimates of fetal gestational age and personalised predictions of fetal growth based on data from an international prospective cohort study: a population-based machine learning study,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
-DOI \texttt{10.1016/S2589-7500(20)30131-X} {Code \url{github.com/ki-analysis/manifold-ga}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
-
-\vbox{\bibitem{ref-30}
-{\sc C.
-Sabanayagam, D.
-Xu, D.
-S.
-W.
-Ting, S.
-Nusinovici, R.
-Banu, H.
-Hamzah, C.
-Lim, Y-C.
-Tham, C.
-Y.
-Cheung, E.
-S.
-Tai, X.
-Y.
-Wang, J.
-B.
-Jonas, C-Y.
-Cheng, M.
-L.
-Lee, W.
-Hsu, and T.
-Y.
-Wong}, ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp, 2020.
+W., Nusinovici, S., Banu, R., Hamzah, H., Lim, C., Tham, Y-C., Cheung, C.
+Y., Tai, E.
+S., Wang, X.
+Y., Jonas, J.
+B., Cheng, C-Y., Lee, M.
+L., Hsu, W., and Wong, T.
+Y.} (2020) ``A deep learning algorithm to detect chronic kidney disease from retinal photographs in community-based populations,'' \emph{Lancet Digital Health}, \textbf{2}(7):7pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1016/S2589-7500(20)30063-7}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 26 January 2021}.}\bibskip
@@
 \vbox{\maprepo{blast-ct}\bibitem{ref-31}
-{\sc M.
-Monteiro, V.
-F.
-Newcombe, F.
-Mathieu, K.
-Adatia, K.
-Kamnitsas, E.
-Ferrante, T.
-Das, D.
-Whitehouse, D.
-Rueckert, D.
-K.
-Menon, and B.
-Glocker}, ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp, 2020.
-DOI \texttt{10.1016/S2589-7500(20)30085-6} {Code \url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
+{Monteiro, M., Newcombe, V.
+F., Mathieu, F., Adatia, K., Kamnitsas, K., Ferrante, E., Das, T., Whitehouse, D., Rueckert, D., Menon, D.
+K., and Glocker, B.} (2020) ``Multiclass semantic segmentation and quantification of traumatic brain injury lesions on head CT using deep learning: an algorithm development and multicentre validation study,'' \emph{Lancet Digital Health}, \textbf{2}(7):8pp.
+
+\leftskip=\labelwidth
+DOI \texttt{10.1016/S2589-7500(20)30085-6}
+
+{\url{github.com/biomedia-mira/blast-ct}}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip
 
 \vbox{\bibitem{ref-32}
-{\sc K-L.
-Liu, T.
-Wu, P-T.
-Chen, M.
-Tsai Y, H.
-Roth, M-S.
-Wu, W-C.
-Liao, and W.
-Wang}, ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp, 2020.
+{Liu, K-L., Wu, T., Chen, P-T., Tsai Y, M., Roth, H., Wu, M-S., Liao, W-C., and Wang, W.} (2020) ``Deep learning to distinguish pancreatic cancer tissue from non-cancerous pancreatic tissue: a retrospective study with cross-racial external validation,'' \emph{Lancet Digital Health}, \textbf{2}(7):10pp.
+
+\leftskip=\labelwidth
 DOI \texttt{10.1016/S2589-7500(20)30078-9}\\\hfill{Accessed 24 July 2020.}\ {Double-checked 27 January 2021}.}\bibskip

### 2023-06-07 expanded-paper.tex
Final updates for Computer Journal

@@
+% Please do not edit this file, expanded-paper.tex!
+% It was generated automatically from its master file paper.tex, by programs/expand.js
+% Get everything from https://github.com/haroldthimbleby/improving-science
 % 
-% This file has been expanded from LaTeX file (paper.tex)
-%   to make this straight forward LaTeX file (expanded-paper.tex)
-%     that depends on nothing (it inputs no files apart from the document class files).
-%       Each required file's text was input in situ - search for % expand
+% This file expands the LaTeX file (paper.tex), which has complex dependencies,
+%   to make this simple LaTeX file (expanded-paper.tex)
+%     that depends on nothing, apart from its standard document class.
+%       This expanded file depends on no other files, as instead
+%         each required file's text was input and expanded recursively - search for '% expand' in the expanded file below.
 % 
-% (Otherwise the file paper.tex would use BibTeX, a macro file, tables of contents, its own and the other files's .aux files, and includes various files in generated/*, etc.)
+% (The original file is dependent on BibTeX (.bst and .bbl files), a macro file, table of contents,
+%   its own and other files' .aux files,
+%     as well as various data files in the directory generated
+%       etc. All these file dependencies have been removed.)
 %
@@
 \bibitem{a=b}
-Petkovsek, M., Wilf, H. and Zeilberger, D. (1996) {\em {$A=B$}}, A~K~Peters.
+Petkovsek, M., Wilf, H., and Zeilberger, D. (1996) {\em {$A=B$}}, A~K~Peters.
   Ltd.
@@
 \bibitem{computational-thinking}
-McOwen, P.~W. and Curzon, P. (2020) {\em The Power of Computational Thinking},
+McOwen, P.~W., and Curzon, P. (2020) {\em The Power of Computational Thinking},
   World Scientific Publishing.
@@
 \bibitem{avianFluModel}
-Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.
+Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.,
   and Burke, D.~S. (2005) ``Strategies for containing an emerging influenza
@@
 \bibitem{originalICmodel}
-Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.
+Ferguson, N.~M., Cummings, D. A.~T., Fraser, C., Cajka, J.~C., Cooley, P.~C.,
   and Burke, D.~S. (2006) ``Strategies for mitigating an influenza pandemic,''
@@
 \bibitem{jupyter-study}
-Pimentel, J.~F., Murta, L., Braganholo, V. and Freire, J. (2019) ``A
+Pimentel, J.~F., Murta, L., Braganholo, V., and Freire, J. (2019) ``A
   large-scale study about quality and reproducibility of {Jupyter} notebooks,''
@@
 \bibitem{trisovic}
-Trisovic, A., Lau, M.~K., Pasquier, T. and Crosas, M. (2022) ``A large-scale
+Trisovic, A., Lau, M.~K., Pasquier, T., and Crosas, M. (2022) ``A large-scale
   study on research code quality and execution nature research,'' {\em
@@
 \bibitem{data-access}
-Gabelica, M., Boj{\v c}i{\' c}, R. and Puljak, L. (2020) ``Many researchers
+Gabelica, M., Boj{\v c}i{\' c}, R., and Puljak, L. (2020) ``Many researchers
   were not compliant with their published data sharing statement: mixed-methods
@@
 Munaf{\`o}, M.~R., Nosek, B.~A., Bishop, D. V.~M., Button, K.~S., Chambers,
-  C.~D., Percie~du Sert, N., Simonsohn, U., Wagenmakers, E.-J., Ware, J.~J. and
-  Ioannidis, J. P.~A. (2017) ``A manifesto for reproducible science,'' {\em
+  C.~D., Percie~du Sert, N., Simonsohn, U., Wagenmakers, E.-J., Ware, J.~J.,
+  and Ioannidis, J. P.~A. (2017) ``A manifesto for reproducible science,'' {\em
   Nature Human Behaviour}, \textbf{1}(1):0021.
@@
 \bibitem{MMR-fraud}
-Godlee, F., Smith, J. and Marcovitch, H. (2011) ``Wakefield's article linking
+Godlee, F., Smith, J., and Marcovitch, H. (2011) ``Wakefield's article linking
   {MMR} vaccine and autism was fraudulent,'' {\em BMJ}, \textbf{342}(c7452).
@@
 \bibitem{elns}
-Machina, H.~K. and Wild, D.~J. (2013) ``Electronic laboratory notebooks
+Machina, H.~K., and Wild, D.~J. (2013) ``Electronic laboratory notebooks
   progress and challenges in implementation,'' {\em Journal of Laboratory
@@
 \bibitem{maneage}
-Akhlaghi, M., Infante-Sainz, R., Roukema, B.~F., Khellat, M., Valls-Gabaud, D.
+Akhlaghi, M., Infante-Sainz, R., Roukema, B.~F., Khellat, M., Valls-Gabaud, D.,
   and Baena-Gall{\'e}, R. (May 2021) ``Toward long-term and archivable
@@
 \bibitem{relit}
-Thimbleby, H. and Williams, D. (2018) ``A tool for publishing reproducible
+Thimbleby, H., and Williams, D. (2018) ``A tool for publishing reproducible
   algorithms {\&} {A} reproducible, elegant algorithm for sequential
@@
 \bibitem{notebook-patent}
-Gray, T.~W. and Wolfram, S. (26 March 2013) {\em Method and system for
+Gray, T.~W., and Wolfram, S. (26 March 2013) {\em Method and system for
   presenting input expressions and evaluations of the input expressions on a
@@
 \bibitem{jupyter-history}
-Granger, B.~E. and P{\'e}rez, F. (2021) ``Jupyter: {Thinking} and storytelling
+Granger, B.~E., and P{\'e}rez, F. (2021) ``Jupyter: {Thinking} and storytelling
   with code and data,'' {\em Computing in Science {\&} Engineering},
@@
 \bibitem{se-bias}
-Friedman, B. and Nissenbaum, H. (1996) ``Bias in computer systems,'' {\em ACM
+Friedman, B., and Nissenbaum, H. (1996) ``Bias in computer systems,'' {\em ACM
   Transactions on Information Systems}, \textbf{14}(3):330--347.
@@
 Laurain, E., Ayav, C., Erpelding, M.-L., Kessler, M., Brian{\c{c}}on, S.,
-  Brunaud, L. and Frimat, L. (2022) ``Targets for parathyroid hormone in
+  Brunaud, L., and Frimat, L. (2022) ``Targets for parathyroid hormone in
   secondary hyperparathyroidism: is a ``one-size-fits-all'' approach
@@
 \bibitem{assurance-case}
-Habli, I., Alexander, R., Hawkins, R., Sujan, M., McDermid, J., Picardi, C. and
-  Lawton, T. (2020) ``Enhancing {COVID-19} decision making by creating an
+Habli, I., Alexander, R., Hawkins, R., Sujan, M., McDermid, J., Picardi, C.,
+  and Lawton, T. (2020) ``Enhancing {COVID-19} decision making by creating an
   assurance case for epidemiological models,'' {\em BMJ Health {\&} Care
@@
 \bibitem{assessing-quality}
-Kelly, D. and Sanders, R. (2008) {\em Assessing the Quality of Scientific
+Kelly, D., and Sanders, R. (2008) {\em Assessing the Quality of Scientific
   Software}, First International Workshop on Software Engineering For
@@
 Moons, K.~G., Altman, D.~G., Reitsma, J.~B., Ioannidis, J.~P., Macaskill, P.,
-  Steyerberg, E.~W., Vickers, A.~J., Ransohoff, D.~F. and Collins, G.~S. (2015)
-  ``Transparent reporting of a multivariable prediction model for individual
-  prognosis or diagnosis ({TRIPOD}): {Explanation} and elaboration,'' {\em
-  Annals of Internal Medicine}, \textbf{162}(1):W1--73.
+  Steyerberg, E.~W., Vickers, A.~J., Ransohoff, D.~F., and Collins, G.~S.
+  (2015) ``Transparent reporting of a multivariable prediction model for
+  individual prognosis or diagnosis ({TRIPOD}): {Explanation} and
+  elaboration,'' {\em Annals of Internal Medicine}, \textbf{162}(1):W1--73.
 \newblock \doi{10.7326/M14-0698}
@@
 \bibitem{forsyth}
-Ahrens, J.~H. and Dieter, U. (1973) ``Extensions of {Forsythe's} method for
+Ahrens, J.~H., and Dieter, U. (1973) ``Extensions of {Forsythe's} method for
   random sampling from the normal distribution,'' {\em Mathematics of
@@
 \bibitem{Normal-Distributions}
-Ahrens, J.~H. and Dieter, U. (1972) ``Computer methods for sampling from the
+Ahrens, J.~H., and Dieter, U. (1972) ``Computer methods for sampling from the
   exponential and normal distributions,'' {\em Communications of the ACM},
@@
 \bibitem{haskell}
-O'Sullivan, B., Stewart, D. and Goerzen, J. (2008) {\em Real World Haskell},
+O'Sullivan, B., Stewart, D., and Goerzen, J. (2008) {\em Real World Haskell},
   O'Reilly Media.
@@
 \bibitem{codecheck}
-Scheuber, A. and {van Elsland}, S.~L. (1 June 2020) {\em Codecheck confirms
+Scheuber, A., and {van Elsland}, S.~L. (1 June 2020) {\em Codecheck confirms
   reproducibility of {COVID-19} model results}.
@@
 \bibitem{NVP}
-{Hatton}, L. and {Roberts}, A. (1994) ``How accurate is scientific software?''
+{Hatton}, L., and {Roberts}, A. (1994) ``How accurate is scientific software?''
   {\em IEEE Transactions on Software Engineering}, \textbf{20}(10):785--797.
@@
 \bibitem{bad-code}
-Richards, D. and Boudnik, K. (16 May 2020) ``{Neil} {Ferguson's} {Imperial}
+Richards, D., and Boudnik, K. (16 May 2020) ``{Neil} {Ferguson's} {Imperial}
   model could be the most devastating software mistake of all time,'' {\em The
@@
 \bibitem{pseudo}
-Zlojutro, A., Rey, D. and Gardner, L. (2019) ``A decision-support framework to
+Zlojutro, A., Rey, D., and Gardner, L. (2019) ``A decision-support framework to
   optimize border control for global outbreak mitigation,'' {\em Nature
@@
 \bibitem{flu-model}
-Sander, B., Nizam, A., {Garrison Jr}, L.~P., Postma, M.~J., Halloran, M.~E. and
-  {Longini Jr}, I.~M. (2009) ``Economic evaluation of influenza pandemic
+Sander, B., Nizam, A., {Garrison Jr}, L.~P., Postma, M.~J., Halloran, M.~E.,
+  and {Longini Jr}, I.~M. (2009) ``Economic evaluation of influenza pandemic
   mitigation strategies in the {US} using a stochastic microsimulation
@@
 \bibitem{psychological-modeling}
-Guest, O. and Martin, A.~E. (2021) ``How computational modeling can force
+Guest, O., and Martin, A.~E. (2021) ``How computational modeling can force
   theory building in psychological science,'' {\em Perspectives on
@@
 \bibitem{ourReview}
-Ladkin, P.~B., Littlewood, B., Thimbleby, H. and Thomas, M. (2020) ``The {Law
+Ladkin, P.~B., Littlewood, B., Thimbleby, H., and Thomas, M. (2020) ``The {Law
   Commission} presumption concerning the dependability of computer evidence,''
@@
 \bibitem{unfinished}
-Kissler, S.~M., Tedijanto, C., Goldstein, E., Grad, Y.~H. and Lipsitch, M.
+Kissler, S.~M., Tedijanto, C., Goldstein, E., Grad, Y.~H., and Lipsitch, M.
   (2020) ``Projecting the transmission dynamics of {SARS-CoV-2} through the
@@
 \bibitem{preclinical-reproducibility}
-Freedman, L.~P., Cockburn, I.~M. and Simcoe, T.~S. (2015) ``The economics of
+Freedman, L.~P., Cockburn, I.~M., and Simcoe, T.~S. (2015) ``The economics of
   reproducibility in preclinical research,'' {\em PLOS Biology},
@@
 \bibitem{fair-principles}
-Wood-Charlson, E.~M., Crockett, Z., Erdmann, C., Arkin, A.~P. and Robinson,
+Wood-Charlson, E.~M., Crockett, Z., Erdmann, C., Arkin, A.~P., and Robinson,
   C.~B. (2022) ``Ten simple rules for getting and giving credit for data,''
@@
 \bibitem{cbc}
-Woodcock, J. C.~P., Larsen, P.~G., Bicarregui, J.~C. and Fitzgerald, J.~S.
+Woodcock, J. C.~P., Larsen, P.~G., Bicarregui, J.~C., and Fitzgerald, J.~S.
   (2009) ``Formal methods: {Practice} and experience,'' {\em ACM Computing
@@
 \bibitem{ABCs-SE}
-Stol, K.-J. and Fitzgerald, B. (2018) ``The {ABC} of software engineering
+Stol, K.-J., and Fitzgerald, B. (2018) ``The {ABC} of software engineering
   research,'' {\em ACM Transactions on Software Engineering and Methodology},
@@
 \bibitem{Executable-Paper}
-Gabriela, A. and Capone, R. (2011) ``Executable paper {Grand Challenge}
+Gabriela, A., and Capone, R. (2011) ``Executable paper {Grand Challenge}
   workshop,'' {\em Procedia Computer Science}, \textbf{4}:577--578.
@@
 \bibitem{RMarkdown}
-Xie, Y., Allaire, J.~J. and Grolemund, G. (2020) {\em R Markdown: The
+Xie, Y., Allaire, J.~J., and Grolemund, G. (2020) {\em R Markdown: The
   Definitive Guide}, Chapman {\&} Hall/CRC.
@@
 \bibitem{debt}
-Falessi, D. and Kruchten, P. (2015) ``Five reasons for including technical debt
-  in the software engineering curriculum,'' {\em in Proceedings of the 2015
-  European Conference on Software Architecture Workshops}, ECSAW'15, ACM,
+Falessi, D., and Kruchten, P. (2015) ``Five reasons for including technical
+  debt in the software engineering curriculum,'' {\em in Proceedings of the
+  2015 European Conference on Software Architecture Workshops}, ECSAW'15, ACM,
   pp.~28:1--28:4.
@@
 \bibitem{reproducibility}
-Benureau, F. C.~Y. and Rougier, N.~P. (2018) ``Re-run, repeat, reproduce,
+Benureau, F. C.~Y., and Rougier, N.~P. (2018) ``Re-run, repeat, reproduce,
   reuse, replicate: {Transforming} code into scientific contributions,'' {\em

### 2023-06-07 generated/assessments.tex
Revisions getting closer to sending to Computer Journal, and renaming the repo
Final updates for Computer Journal

@@
+% Generated by programs/data.js
 \cite{ref-1} & On request.\flagStyle{ } & ``Code is available upon request from the corresponding author'' (requested)\flagStyle{ $\sf P_c$\hskip 3pt{}$\sf S_p$\hskip 3pt{}}\\

### 2023-06-07 programs/data.js

@@ -403,7 +403,7 @@ var data = [{
         hasTrivialComment: 1,
         hasDirectCode: 1,
         hasBreach: 1,
-        codeComment: "Basic Matlab with routine comments.",
+        codeComment: "Basic Matlab with routine comments. Supplemental files Rsos192210supp1.docx and Rsos192210supp2.docx have links but were not accessible at time of double checking.",
         codeURL: "royalsocietypublishing.org/doi/suppl/10.1098/rsos.192210",
         pages: 9
             },
@@ -1155,37 +1155,48 @@ for (var i = 0; i < data.length; i++) {
 }
 saveFile("assessments.tex", s, "Main summary table for Supplementary Material");
 
+
+// We roughly follow Computer Journal's requirements for journal article citations: 
+// Galton, A. (1992) Logic as a formal method. Comp. J., 35, 431-440.
+// (authors, year, title of paper, title of journal, volume number, page numbers)
+// We include all authors (up to maxAuthors)
+
 function fixAuthors(s) {
     const maxAuthors = 1000; // put et al after nth author (eg if maxAuthors=3, at most 3 explicit authors)
     var a = s.split(",");
     var fa = "";
     var prefix = "";
+    var initials = "";
     for (var i = 0; i < a.length; i++) {
         a[i] = a[i].trim();
         // a[i] is in the format: Kaminsky ZA
-        // we change it to {\sc Z. A. Kaminsky}
+        // we change it to Kaminsky, Z. A.
         fa += prefix;
+        initials = "";
         var lastSpace = a[i].length - 1;
         while (lastSpace >= 0 && a[i].charAt(lastSpace) != " ") lastSpace--;
         for (var k = lastSpace + 1; k < a[i].length; k++) {
-            fa += a[i].charAt(k);
+            initials += a[i].charAt(k);
             if (a[i].charAt(k + 1) == "-") {
-                fa += a[i].charAt(++k);
+                initials += a[i].charAt(++k);
                 continue;
             }
-            fa += ". ";
+            initials += ". ";
         }
+        initials = initials.trim();
         for (var k = 0; k < lastSpace; k++)
             fa += a[i].charAt(k);
[…]

### 2023-06-07 programs/expand.js

@@ -5,7 +5,7 @@
 
 var TeXFiles = ["paper.tex", "appendix.tex"];
 
-function makenote(vector) { 
+function makenote(whatsaying, vector) { 
     var files = "", expandedfiles = "";
     for( var i = 0; i < vector.length; i++ )
     {	files += i == 0? "": i < vector-1? ", ": ", and "; 
@@ -14,11 +14,13 @@ function makenote(vector) {
         expandedfiles += "expanded-"+vector[i];
     }
 	var plural = vector.length > 1;
-	return ("LaTeX file-noun ("+files+")\n  to make "+(plural?"":"this ")+"straight forward LaTeX file-noun ("+expandedfiles+")\n    that depend-verb on nothing ("+(plural?"they":"it")+" input-verb no files apart from the document class files).\n      Each required file's text was input in situ - search for % expand\n\n(Otherwise the file-noun "+files+" would use BibTeX, a macro file, tables of contents, "+(plural?"their":"its")+" own and the other files's .aux files, and include-verb various files in generated/*, etc.)\n").replace(/-noun/mg, plural?"s":"").replace(/-verb/mg, plural?"":"s");
+	var warn = plural? "": "Please do not edit this file, "+expandedfiles+"!\nIt was generated automatically from its master file "+files+", by programs/expand.js\nGet everything from https://github.com/haroldthimbleby/improving-science\n\n";
+	return warn+("This "+whatsaying+" expands //the /LaTeX file/s// ("+files+"), which /have/has/ complex dependencies,\n  to make //this /simple LaTeX file/s// ("+expandedfiles+")\n    that depend//s/ on nothing, apart from /their/its/ standard document class.\n      /The/This/ expanded file/s// depend//s/ on no other files, as instead\n        each required file's text was input and expanded recursively - search for '% expand' in the expanded file/s//// below/.\n\n(The original file/s// /are/is/ dependent on BibTeX (.bst and .bbl files), a macro file, table/s// of contents,\n  /their/its/ own and other files' .aux files,\n    as well as various data files in the directory generated\n      etc. All these file dependencies /are/have been/ removed.)\n").replace(/\/([^/]*)\/([^/]*)\//mg, plural?"$1":"$2");
 }
 
-console.log("This app expands complex "+makenote(TeXFiles));
[…]

### 2023-06-07 programs/expand.js
Improved expand.js help text

@@ -63,7 +63,7 @@ for (var i = 0; i < TeXFiles.length; i++) {
     console.log("Recursively expand all \\input, \\tableofcontents, and \\bibliography macro calls in " + TeXFiles[i]);
     rootfile = TeXFiles[i].replace(/.tex/, "");
     auxfile = rootfile+".aux";
-    console.log("        \\input " + auxfile + " => automatically pre-inserted between \\makeatletter and \\makeatother");
+    console.log("        \\input " + auxfile + " => automatically pre-inserted between \\makeatletter and \\makeatother\n               NB "+auxfile+" is inserted after the expanded file's own .aux file is automatically loaded (if it exists).\n               ... in other words, the expanded file's own .aux file is correctly overridden by the original "+auxfile+" file.");
     
     bufferExpand = (makenote("file", [TeXFiles[i]])).replace(/^/gm, "% ")+"\n\n";
 	bufferExpand = bufferExpand + "\\makeatletter\n"+getFile(auxfile) + "\n\\makeatother\n";

### 2023-06-08 expanded-paper.tex
Improved expand.js help text
Trying to get Elsevier's Latex processing to work

@@
 \citation{popper-conjectures-refutations}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}\protected@file@percent }
 \citation{preventable-book}
-\@writefile{toc}{\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}\protected@file@percent }
 \citation{preventable-book}
@@
 \citation{Ben,nature-review,se-bias}
+\@writefile{toc}{\contentsline {subsection}{\numberline {5.c}Beyond pandemic modeling}{20}{subsection.5.3}\protected@file@percent }
+\newlabel{section-science-beyond-pandemic-modeling}{{5.c}{20}{Beyond pandemic modeling}{subsection.5.3}{}}
 \citation{ourReview}
@@
 \citation{NVP}
-\@writefile{toc}{\contentsline {subsection}{\numberline {5.c}Beyond pandemic modeling}{20}{subsection.5.3}\protected@file@percent }
-\newlabel{section-science-beyond-pandemic-modeling}{{5.c}{20}{Beyond pandemic modeling}{subsection.5.3}{}}
 \citation{unfinished}
@@
 \received{30 April 2022}
-\revised{8 June 2022, 3 January 2023, 31 January 2023 \& 12~May~2023} 
+\revised{8 June 2022, 3 January 2023, 31 January 2023, 12~May~2023 \& 7~June~2023} 
 \volumenumber{00} \issuenumber{0} \volumeyear{2023}
@@
 \maketitle
-
-%\section*{Author summary}
-%
-%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
-%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales.
-His research focuses on human error and computer system design, particularly for healthcare.
-%%In addition to over 340 peer-reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
-%Harold won the British Computer Society Wilkes Medal.
-He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder.
-His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.
-See his web site, \texttt{www.harold.thimbleby.net}, for more details.
-
-%\noindent\emph{This paper evidences widespread problems arising from scientific papers relying on inadequately defined computer and inaccessible code, which undermines peer-review and replication, and undermines scientific quality generally.
-The risks vary, depending on how critical a paper's claims may be: ranging from superficial for opinion papers, to serious for papers in domains such as medicine, public health, or policy.
-This paper contrasts the situation with the mature approach to using statistics, and goes on to make proposals to help address the problem.
-The proposals are feasible, and presented in the hope of motivating their (or better solutions) being implemented, tested, and refined.}
 
@@
   \imageWidthCalculation
-  \begin{center}\includegraphics[width=\imageWidth]{
-generated/mathematicaplot.jpg
-}\end{center}}
+  \begin{center}%
+% removed generated/ from graphics pathname
+\includegraphics[width=\imageWidth]{expanded-plot.\imageFileSuffix}
+\end{center}}
 \caption{Much computational science is concerned with finding plausible multi-dimensional models that fit models to data with the aim of extrapolating or predicting new results from them.
@@
 
-%An example is the documentation at the code repository \xcite{BayesWaveCode}, which requires three high-profile \emph{{Physics Review D}} papers \xcite{BayesWave1,BayesWave2,BayesWave3} be cited to acknowledge any use, but only one of them, \xcite{BayesWave3}, mentions the code, and even then it fails to acknowledge the contributions of its 10 programmers (despite 6 being co-authors).
-The paper does cite the repository URL, but unlike the papers it cites conventionally, it omits all the code's authors.
-Acknowledgement does not seem to matter, even to the people who did the work.
-
 Since published papers rarely mention their code, new papers contributing to the literature do not write about their code either, so the low status of code persists.
@@
 In reaction to this vicious cycle, there is a growing movement to use and cite code correctly \cite{cite-code,TOP}, because code \emph{is\/} important, particularly for the reproduction, testing, and extension of any scientific work.
-(Code also needs to be correct, not just cited correctly.)
-
-%Few journals editorial policies recognize that data and code are theoretically and in practice indistinguishable (see \supplement).
-Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.
+(Code also needs to be correct, not just cited correctly.) Few journals editorial policies recognize that data and code are in practice indistinguishable (see \supplement).
+Given that data and code are equally important in science, effectively equivalent and interchangeable, it follows that publishing policies on data handling should also apply at least as strictly to code.
 
@@
 
-%Since the symmetries in the scientific domains should be reflected in corresponding symmetries in well-designed code, this raises new possibilities of Formal Methods approaches (e.g., using automatic theorem provers) supporting advances in general science.
-
-\section{State of the art}% \\ in computational science}
+\section{State of the art}
 
@@
 The conventional view of science is that experiments are done \emph{then\/} written up.
-%RAP and \RAPstar\ support and enhance this process with computational tools; in particular, the archival results of research are not just the reports and publications, but include the executable processes --- the pipelines in RAP terms --- of the research.
 However, it is more productive to think of the paper itself as an active laboratory, not just as a record of finished work.

### 2023-06-08 expanded-copyfiles.sh

@@ -1,0 +1,6 @@
+# assuming \imageFileSuffix is defined as jpg
+cp generated/plot.jpg expanded-plot.jpg
+cp generated/basic.jpg expanded-basic.jpg
+cp generated/excel.jpg expanded-excel.jpg
+cp generated/import.jpg expanded-import.jpg
+cp generated/notebook.jpg expanded-notebook.jpg

### 2023-06-08 programs/expand.js

@@ -1,7 +1,24 @@
 // Harold Thimbleby, 2020-2023
 // Javascript code to recursively expand TeX \input commands in files TeXFiles[] (see below for value)
 // run as node generated/expand.js
+
+
 'use strict';
+
+
+var flattenGraphics = true;
+// graphics files are put in the directory generated/ but some publishers can't cope, 
+// so if flattenGraphics is true, the lines like
+//
+//    \includegraphics[width=\imageWidth]{generated/mathematicaplot.\imageFileSuffix}
+//
+// are converted to
+//
+//    \includegraphics[width=\imageWidth]{mathematicaplot.\imageFileSuffix}
+//
+// we also list all the relevant image files, as they will need copying out of the directory generated/
+
+
 
 var TeXFiles = ["paper.tex", "appendix.tex"];
 
@@ -52,7 +69,13 @@ function getFile(file) {
     }
 }
 
-var re = RegExp("\\\\input (.*)\n", "g"), rebib=RegExp("\\\\bibliography{(.*).bib}"), recontents=RegExp("\\\\tableofcontents");
+var cpFiles = "# assuming \\imageFileSuffix is defined as jpg\n",
+    cpFileCount = 0;
+
+var re = RegExp("\\\\input *(.*)\n", "g"), 
+    rebib = RegExp("\\\\bibliography *{(.*).bib}"), 
+    recontents = RegExp("\\\\tableofcontents *"),
+    regraphics = RegExp("\\\\includegraphics *\\[([^\\}]*generated/[^\\}]*)\\}");
 
 var contentsStuff = "\\makeatletter\n  \\section*{\\contentsname\n    \\@mkboth{\\MakeUppercase\\contentsname}{\\MakeUppercase\\contentsname}}\n\\makeatother";
 
@@ -91,10 +114,34 @@ for (var i = 0; i < TeXFiles.length; i++) {
             return "% expand table of contents " + rootfile + ".toc\n" + bufferInsert + "\n% end expanding table of contents " + rootfile + ".toc\n";
         });
     }
-    while (replaced);
+    while( replaced );
+    if( flattenGraphics )
+    {	do {   
+        	replaced = false;
+			cpFileCount++;
+    		// remove generated/ from \includegraphics[width=\imageWidth]{generated/mathematicaplot.\imageFileSuffix}
+        	bufferExpand = bufferExpand.replace(regraphics, function (match, graphicsDetails) {
+            	replaced = true;
[…]

### 2023-06-08 expanded-appendix.tex
Trying to get Elsevier's Latex processing to work
Fixing Elsevier quirks

@@
 \@writefile{toc}{\contentsline {subsection}{\numberline {11.b}How trustworthy is the source?}{51}{subsection.11.2}\protected@file@percent }
+\@writefile{lot}{\contentsline {table}{\numberline {6}{\ignorespaces Conforming to the RAP{\tt  +}\ methodology, the abbreviated summary above was generated automatically, by using \texttt  {make data}.
+A full list of \texttt  {make} options is generated by \texttt  {make} or \texttt  {make help} when run on the Unix command line.
+}}{52}{table.6}\protected@file@percent }
+\newlabel{supplement-table:options}{{6}{52}{Conforming to the \RAPstar \ methodology, the abbreviated summary above was generated automatically, by using \texttt {make data}.
+A full list of \texttt {make} options is generated by \texttt {make} or \texttt {make help} when run on the Unix command line}{table.6}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {11.c}How trustworthy is the interpretation?}{52}{subsection.11.3}\protected@file@percent }
+\@writefile{toc}{\contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}\protected@file@percent }
+\newlabel{supplement-survey-section}{{12}{52}{A pilot survey of computational science}{section.12}{}}
 \citation{lancet-retracted}
@@
 \citation{lancet-learning}
-\@writefile{toc}{\contentsline {subsection}{\numberline {11.c}How trustworthy is the interpretation?}{52}{subsection.11.3}\protected@file@percent }
-\@writefile{toc}{\contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}\protected@file@percent }
-\newlabel{supplement-survey-section}{{12}{52}{A pilot survey of computational science}{section.12}{}}
-\@writefile{toc}{\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{52}{subsection.12.1}\protected@file@percent }
-\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.1}\emph  {The Lancet\/}}{52}{subsubsection.12.1.1}\protected@file@percent }
 \citation{jvs1}
@@
 \citation{ICmodel}
-\@writefile{lot}{\contentsline {table}{\numberline {6}{\ignorespaces Conforming to the RAP{\tt  +}\ methodology, the abbreviated summary above was generated automatically, by using \texttt  {make data}.
-A full list of \texttt  {make} options is generated by \texttt  {make} or \texttt  {make help} when run on the Unix command line.
-}}{53}{table.6}\protected@file@percent }
-\newlabel{supplement-table:options}{{6}{53}{Conforming to the \RAPstar \ methodology, the abbreviated summary above was generated automatically, by using \texttt {make data}.
-A full list of \texttt {make} options is generated by \texttt {make} or \texttt {make help} when run on the Unix command line}{table.6}{}}
+\@writefile{toc}{\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{53}{subsection.12.1}\protected@file@percent }
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.1}\emph  {The Lancet\/}}{53}{subsubsection.12.1.1}\protected@file@percent }
 \@writefile{toc}{\contentsline {subsubsection}{\numberline {12.a.2}\emph  {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}\protected@file@percent }
@@
 \@writefile{toc}{\contentsline {subsection}{\numberline {12.b}Pilot paper sample}{53}{subsection.12.2}\protected@file@percent }
+\citation{nature-review}
 \@writefile{lot}{\contentsline {table}{\numberline {7}{\ignorespaces Overview of the peer-reviewed paper sample.
@@
 \newlabel{supplement-table-summary}{{8}{54}{Summary of survey results. (Convenient copy of main paper's table \ref {table-summary}.)}{table.8}{}}
-\citation{nature-review}
 \@writefile{lot}{\contentsline {table}{\numberline {9}{\ignorespaces Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt  {covid-sim} \cite  {ICmodel} for comparison.
@@
 \newlabel{supplement-journal-policies-section}{{12.c.1}{57}{Current code policies of sampled journals}{subsubsection.12.3.1}{}}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{57}{subsubsection.12.3.2}\protected@file@percent }
 \citation{ref-1}
@@
     {1}{208.35622pt}}
-\@writefile{toc}{\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{58}{subsubsection.12.3.2}\protected@file@percent }
 \citation{prisma}
@@
 \contentsline {section}{\numberline {12}A pilot survey of computational science}{52}{section.12}%
-\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{52}{subsection.12.1}%
-\contentsline {subsubsection}{\numberline {12.a.1}\emph {The Lancet\/}}{52}{subsubsection.12.1.1}%
+\contentsline {subsection}{\numberline {12.a}Selected journal case studies}{53}{subsection.12.1}%
+\contentsline {subsubsection}{\numberline {12.a.1}\emph {The Lancet\/}}{53}{subsubsection.12.1.1}%
 \contentsline {subsubsection}{\numberline {12.a.2}\emph {Journal of Vascular Surgery}}{53}{subsubsection.12.1.2}%
@@
 \contentsline {subsubsection}{\numberline {12.c.1}Current code policies of sampled journals}{57}{subsubsection.12.3.1}%
-\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{58}{subsubsection.12.3.2}%
+\contentsline {subsubsection}{\numberline {12.c.2}Sample assessment and scoring}{57}{subsubsection.12.3.2}%
 \contentsline {subsection}{\numberline {12.d}Assessment criteria and methods}{60}{subsection.12.4}%
@@
    \centering
-   \includegraphics[width=5in]{generated/basic.\imageFileSuffix} 
+   % removed generated/ from graphics pathname
+\includegraphics[width=5in]{expanded-basic.\imageFileSuffix} 
    \caption{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
@@
    \centering
-   \includegraphics[width=5in]{generated/excel.\imageFileSuffix} 
+   % removed generated/ from graphics pathname
+\includegraphics[width=5in]{expanded-excel.\imageFileSuffix} 
    \caption{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
@@
    \centering
-   \includegraphics[width=5in]{generated/import.\imageFileSuffix} 
+   % removed generated/ from graphics pathname
+\includegraphics[width=5in]{expanded-import.\imageFileSuffix} 
    \caption{Improving over the normal approach (figure \ref{supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
@@
    \centering
-   \includegraphics[width=5in]{generated/notebook.\imageFileSuffix} 
+   % removed generated/ from graphics pathname
+\includegraphics[width=5in]{expanded-notebook.\imageFileSuffix} 
    \caption{In a notebook system, such as \emph{Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
@@
 % 2 - make archive
-% 3 - make check-same
-% 4 - make check-update
-% 5 - make check-versions
-% 6 - make data
-% 7 - make expand
-% 8 - make git-prep
-% 9 - make help
-% 10 - make help-brief
-% 11 - make mathematica
-% 12 - make mathematica-open
-% 13 - make one-file
-% 14 - make pdf
-% 15 - make push
-% 16 - make readme
-% 17 - make really-tidyup
-% 18 - make tidyup
-% 19 - make zip
-% 20 - make zip-data
+% 3 - make check
+% 4 - make check-git
+% 5 - make check-same
+% 6 - make check-update
+% 7 - make check-versions
+% 8 - make data
+% 9 - make expand
+% 10 - make help
+% 11 - make help-brief
+% 12 - make mathematica
+% 13 - make mathematica-open
+% 14 - make one-file
+% 15 - make pdf
+% 16 - make push
+% 17 - make readme
+% 18 - make really-tidyup
+% 19 - make tidyup
+% 20 - make zip
+% 21 - make zip-data
 
@@
 
-\texttt{make data}&
-      Analyze the data, and generate all the data files, the Unix scripts, the CSV, and \LaTeX\ files (including the \LaTeX\ summary of this makefile), etc.
-This \texttt{make} option runs \texttt{node programs/data.js}, downloads the Git repositories used in the pilot survey, and then analyzes them.
-Note that downloading all the repositories in a reasonable time needs decent internet bandwidth.\\
+\texttt{make check-versions}&
+      Check that you have the right software and software versions to run everything.\\
    \multicolumn{1}{l@{\vdots}}{}&\\

### 2023-06-08 programs/expand.js
Minor updates to expand and makefile

@@ -72,10 +72,10 @@ function getFile(file) {
 var cpFiles = "# assuming \\imageFileSuffix is defined as jpg\n",
     cpFileCount = 0;
 
-var re = RegExp("\\\\input *(.*)\n", "g"), 
-    rebib = RegExp("\\\\bibliography *{(.*).bib}"), 
-    recontents = RegExp("\\\\tableofcontents *"),
-    regraphics = RegExp("\\\\includegraphics *\\[([^\\}]*generated/[^\\}]*)\\}");
+var REinput = RegExp("\\\\input *(.*)\n", "g"), 
+    REbib = RegExp("\\\\bibliography *{(.*).bib}"), 
+    REcontents = RegExp("\\\\tableofcontents *"),
+    REgraphics = RegExp("\\\\includegraphics *\\[([^\\}]*generated/[^\\}]*)\\}");
 
 var contentsStuff = "\\makeatletter\n  \\section*{\\contentsname\n    \\@mkboth{\\MakeUppercase\\contentsname}{\\MakeUppercase\\contentsname}}\n\\makeatother";
 
@@ -94,19 +94,19 @@ for (var i = 0; i < TeXFiles.length; i++) {
     var replaced;
     do {
         replaced = false;
-        bufferExpand = bufferExpand.replace(re, function (match, filename) {
+        bufferExpand = bufferExpand.replace(REinput, function (match, filename) {
             replaced = true;
             console.log("        \\input " + filename);
             bufferInsert = getFile(filename);
             return "% expand " + filename + "\n" + bufferInsert + "% end expanding " + filename + "\n";
         });
-        bufferExpand = bufferExpand.replace(rebib, function (match, filename) {
+        bufferExpand = bufferExpand.replace(REbib, function (match, filename) {
         	replaced = true;
             console.log("      * \\bibliography{"+filename+".bib} => replaced with " + filename+".bbl");
             bufferInsert = getFile(filename+".bbl");
             return "% expand bibliography " + filename + ".bbl\n" + bufferInsert + "\n% end expanding bibliography " + filename + ".bbl\n";
         });
-        bufferExpand = bufferExpand.replace(recontents, function (match, filename) {
+        bufferExpand = bufferExpand.replace(REcontents, function (match, filename) {
         	replaced = true;
 			insertTableOfContents = true;
             console.log("      * \\tableofcontents => replaced with "+ rootfile+".toc, after standard expansion of \\tableofcontents macro");
[…]

### 2023-06-09 EXappendix.tex
Working around bugs in publishers, sigh

--- 2023-06-08 expanded-appendix.tex
@@
-% Please do not edit this file, expanded-appendix.tex!
+% Please do not edit this file, EXappendix.tex!
 % It was generated automatically from its master file appendix.tex, by programs/expand.js
@@
 % This file expands the LaTeX file (appendix.tex), which has complex dependencies,
-%   to make this simple LaTeX file (expanded-appendix.tex)
+%   to make this simple LaTeX file (EXappendix.tex)
 %     that depends on nothing, apart from its standard document class.
@@
 % code to make digits into blocks of 3 (as in 1000 -> 1 000)
-\def\xBlockSep{\textcolor{red}{\hskip 0.3em :}\hskip 0.3em} % International Standards are to use a small space. UK typically uses a comma
+\def\xBlockSep{\hskip 0.3em} % International Standards are to use a small space. UK typically uses a comma
 % NOTE currently, the few decimal numbers (eg 0.000 5) use hskip explicitly, since defining \xBlockSep as a comma would definitely be wrong for decimals
@@
 \newcount\xdigitBlocktmp
+\newcount\xdigitBlocktmq
+
 \def\xdigitBlock#1{%
-        \xdigitBlockn=#1\relax
+        \xdigitBlockn = #1 \relax
         \ifnum \xdigitBlockn < 0
@@
         \fi
-        \let \xdigitBlockpptemp \empty
-        \loop \ifnum \xdigitBlockn > 999
+        \let \xdigitblockinternalmacro \empty
+        \loop 
+           \ifnum \xdigitBlockn > 999
                 \xdigitBlocktmp = \xdigitBlockn
                 \divide \xdigitBlockn by 1000
-                \xdigitBlocktmp = \numexpr \xdigitBlocktmp - 1000*\xdigitBlockn \relax
-                \edef \xdigitBlockpptemp{\xBlockSep\ifnum \xdigitBlocktmp < 100  0\ifnum \xdigitBlocktmp < 10 0\fi\fi \the\xdigitBlocktmp \xdigitBlockpptemp}%
+
+                % following line rewritten explicitly for old TeX 
+                %\xdigitBlocktmp = \numexpr \xdigitBlocktmp - 1000*\xdigitBlockn \relax           
+                \xdigitBlocktmq = \xdigitBlockn \relax
+                \multiply \xdigitBlocktmq by 1000
+                \advance \xdigitBlocktmp by -\xdigitBlocktmq
+
+                \edef \xdigitblockinternalmacro{%
+                	\xBlockSep
+					\ifnum \xdigitBlocktmp < 100 0\ifnum \xdigitBlocktmp < 10 0\fi \fi 
+					\the\xdigitBlocktmp
+					\xdigitblockinternalmacro
+				}%
         \repeat 
         \the\xdigitBlockn
-        \xdigitBlockpptemp
+        \xdigitblockinternalmacro
 }
@@
 \@writefile{toc}{\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}\protected@file@percent }
-\@writefile{toc}{\contentsline {section}{\numberline {7}Conclusions}{28}{section.7}\protected@file@percent }
 \citation{tweet,ferguson-interview}
@@
 \citation{reproducibility,relit,popper-conjectures-refutations}
+\@writefile{toc}{\contentsline {section}{\numberline {7}Conclusions}{29}{section.7}\protected@file@percent }
 \citation{parliamentary-evidence,my-parliamentary-evidence}
@@
 The next hard-coded value mentioned in the calculation is difficult to categorize: is the value of $\pi$ empirical data or is it part of a standard formula?
-Some programming languages like \emph{Mathematica\/} treat $\pi$ as an exact mathematical constant (e.g., \emph{Mathematica\/} calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was 62,831,853,071,796 digits %as calculated by the University of Applied Sciences, Chur, Switzerland
-\url{www.fhgr.ch/en/specialist-areas/applied-future-technologies/davis-centre/pi-challenge}}
+Some programming languages like \emph{Mathematica\/} treat $\pi$ as an exact mathematical constant (e.g., \emph{Mathematica\/} calculates $\tan\pi/4 = 1$ exactly), but $\pi$ is \emph{also\/} definitely an inexact empirical value.\footnote{A record set on 19 August 2021, the most accurate value of $\pi$ then known was
+62\xBlockSep 831\xBlockSep 853\xBlockSep 071\xBlockSep 796 digits. %as calculated by the University of Applied Sciences, Chur, Switzerland 
+See \url{www.fhgr.ch/en/specialist-areas/applied-future-technologies/davis-centre/pi-challenge}}
 
@@
 This now seems to be a more efficient program because it has avoided a test (which a modern complier would have optimized away anyway).
-Unfortunately, the previously explicit dependency of the code on the magic number \the\magicNumber\ has completely disappeared.
+Unfortunately, the previously explicit dependency of the code on specific the magic number \the\magicNumber\ has completely disappeared.
 
@@
    % removed generated/ from graphics pathname
-\includegraphics[width=5in]{expanded-basic.\imageFileSuffix} 
+\includegraphics[width=5in]{EXbasic} 
    \caption{The common basic --- error-prone and obsolete --- approach to scientific authoring is to use code to help generate analyses and diagrams, then manually copy and paste the selected results into the publication.
@@
    % removed generated/ from graphics pathname
-\includegraphics[width=5in]{expanded-excel.\imageFileSuffix} 
+\includegraphics[width=5in]{EXexcel} 
    \caption{When the code is in a spreadsheet, such as Microsoft Excel, the code is generally hidden from sight.
@@
    % removed generated/ from graphics pathname
-\includegraphics[width=5in]{expanded-import.\imageFileSuffix} 
+\includegraphics[width=5in]{EXimport} 
    \caption{Improving over the normal approach (figure \ref{supplement-fig:basic}), data and diagrams are placed in the publication using a programmed, systematic approach.
@@
    % removed generated/ from graphics pathname
-\includegraphics[width=5in]{expanded-notebook.\imageFileSuffix} 
+\includegraphics[width=5in]{EXnotebook} 
    \caption{In a notebook system, such as \emph{Mathematica\/} or Jupyter, a single document --- the notebook --- integrates the publication, the code, any notes, as well as all the results.
@@
 
-This convenience sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers,\footnote{Using Google Scholar it is estimated that over 40,000 papers meeting the title criteria were published in the month of July 2020.} but~\ldots
+This convenience sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers,\footnote{Using Google Scholar it is estimated that over \digitBlock{40000} papers meeting the title criteria were published in the month of July 2020.} but~\ldots
 
@@
 All the discussion and examples above were generated automatically, and have been checked correct for other correct values than \the\countUsesVersionControlRepository.
-This approach, too, considerably helps defend against common Human Factors errors.
-For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be 2\hskip .3em 348, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
+This approach, too, helps defend against common Human Factors errors.
+For example, if we set \texttt{\bslash countUsesVersionControlRepository}=\the\countUsesVersionControlRepository\ to be \digitBlock{2348} instead, say, then all of the subsequent sentences that mention it will say something unexpected and so have to be more carefully proof-read, significantly reducing confirmation bias.
 The approach turns a possibly-hard-to-spot \emph{single\/} error into \emph{multiple\/} errors spread throughout the paper into different contexts, thus increasing the chances of noticing the error.

### 2023-06-09 EXpaper.tex
Working around bugs in publishers, sigh

--- 2023-06-08 expanded-paper.tex
@@
-% Please do not edit this file, expanded-paper.tex!
+% Please do not edit this file, EXpaper.tex!
 % It was generated automatically from its master file paper.tex, by programs/expand.js
@@
 % This file expands the LaTeX file (paper.tex), which has complex dependencies,
-%   to make this simple LaTeX file (expanded-paper.tex)
+%   to make this simple LaTeX file (EXpaper.tex)
 %     that depends on nothing, apart from its standard document class.
@@
 \revised{8 June 2022, 3 January 2023, 31 January 2023, 12~May~2023 \& 7~June~2023} 
+\accepted{30 May 2023}
 \volumenumber{00} \issuenumber{0} \volumeyear{2023}
@@
 \contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}%
-\contentsline {section}{\numberline {7}Conclusions}{28}{section.7}%
+\contentsline {section}{\numberline {7}Conclusions}{29}{section.7}%
 
@@
 {\sf
-\newcount \enum
-\enum=1
-\renewcommand{\theenumi}{%\ifnum \enum<10 \hphantom{0}\fi
-\the\enum
-\global\advance \enum by 1}
 \begin{center}\begin{tabular}{|rp{6.5in}|}
@@
 The RAP approach would require the entire workflow to be revised, not just the final presentation in the paper.
-
 \end{enumerate}
@@
 % removed generated/ from graphics pathname
-\includegraphics[width=\imageWidth]{expanded-plot.\imageFileSuffix}
+\includegraphics[width=\imageWidth]{EXplot}
 \end{center}}
@@
 
-\newcount \enum
-\enum=1
-\renewcommand{\theenumi}{\ifnum \enum<10 \hphantom{0}\fi
-\the\enum
-\global\advance \enum by 1}
-\begin{enumerate}\raggedright
+\begin{enumerate}
+\raggedright
 \item Is the code valid: does it do what the paper claims?\footnote{Of course, the underlying science may be wrong to, so it is useful to distinguish \emph{internal validity\/} and \emph{external validity}.
@@
 
-Finally, as there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access and methodology statements apply to all information, regardless of arbitrary classification of it as code or data (or documentation, assurance case, etc).
+As there is no practical distinction between data and code (see \supplement) and methodology (thanks to RAP), and certainly no distinctions that cannot be circumvented, journal and funder policies of code and data access should be reviewed \emph{and unified\/} so that the access and methodology statements apply to all information, regardless of arbitrary classification of it as code or data (or documentation, assurance case, etc).
+
+Finally, there is still a long way to go.
+The \emph{Computer Journal}, which should be a leading journal for publishing computational science (and which has published this paper) uses poorly documented automated and buggy paper processing software, whose obscure bugs took several weeks of hard work to overcome in order to submit the present paper.
+Such problems will put authors off using computational papers and \RAPstar\ methodologies.
 
@@
 
-\enum=1
 \begin{enumerate}\raggedright
@@
 To find effective ways to promote professional software engineers being recognized and participating fully in scientific research, just as professional statisticians routinely support quality research (see section \ref{statistics-computation-analogy}).
+
+\item
+Computer science is a complex discipline, but with low barriers to entry.
+When critical code is being written or used, high quality developers and software engineers are required, and there are not many of them.
+Improving the educational and professional standards in coding, and awareness of the issues are a priority for all aspects of science, publishing --- and life generally.
 \end{enumerate}
@@
 
-\bibnewcolumn = 126 % bibentry number of last entry on (final) left hand column
+\bibnewcolumn = 123 % bibentry number of last entry on (final) left hand column
 % Hint: first set it to a very large number, so the final column is as unbalanced as it gets

### 2023-06-09 appendix.tex
Preparing to upload to Computer Journal
Working around bugs in publishers, sigh

--- 2023-05-31 paper-seb-supplementary-material.tex
@@
 \def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}    
-\input paper-seb-macros.tex
+\input macros.tex
 
@@
       \def\bibcite#1#2{\global\advance\mainNumberOfReferences by 1\oldbibcite{#1}{#2}}
-      \input paper-seb-main.aux 
+      \input paper.aux 
       \let\bibcite=\oldbibcite
@@
 \def\endofbibliography{ \egroup }
-  \bibliography{paper-seb-supplementary-material.bib}
+  \bibliography{appendix.bib}
 }
@@
 
-\input generated/supplementary-references.bbl
+\input generated/survey-references.bbl
 

### 2023-06-09 EXcopyfiles.sh

@@ -1,0 +1,6 @@
+# assuming \imageFileSuffix is defined as jpg
+cp generated/plot.jpg EXplot.jpg
+cp generated/basic.jpg EXbasic.jpg
+cp generated/excel.jpg EXexcel.jpg
+cp generated/import.jpg EXimport.jpg
+cp generated/notebook.jpg EXnotebook.jpg

### 2023-06-09 expanded-copyfiles.sh

@@ -1,6 +1,0 @@
-# assuming \imageFileSuffix is defined as jpg
-cp generated/plot.jpg expanded-plot.jpg
-cp generated/basic.jpg expanded-basic.jpg
-cp generated/excel.jpg expanded-excel.jpg
-cp generated/import.jpg expanded-import.jpg
-cp generated/notebook.jpg expanded-notebook.jpg

### 2023-06-09 programs/expand.js

@@ -2,9 +2,9 @@
 // Javascript code to recursively expand TeX \input commands in files TeXFiles[] (see below for value)
 // run as node generated/expand.js
 
-
 'use strict';
 
+var expandedPrefix = "EX";
 
 var flattenGraphics = true;
 // graphics files are put in the directory generated/ but some publishers can't cope, 
@@ -18,8 +18,11 @@ var flattenGraphics = true;
 //
 // we also list all the relevant image files, as they will need copying out of the directory generated/
 
+// Elsevier makes a mess of includegraphics, so this macro is expanded here
+var REsuffixtofix = /.\\imageFileSuffix/,
+    fixedSuffix = ".jpg";
 
-
+// the files we will expand...
 var TeXFiles = ["paper.tex", "appendix.tex"];
 
 function makenote(whatsaying, vector) { 
@@ -28,7 +31,7 @@ function makenote(whatsaying, vector) {
     {	files += i == 0? "": i < vector-1? ", ": ", and "; 
         files += vector[i];
         expandedfiles += i == 0? "": i < vector-1? ", ": ", and "; 
-        expandedfiles += "expanded-"+vector[i];
+        expandedfiles += expandedPrefix+vector[i];
     }
 	var plural = vector.length > 1;
 	var warn = plural? "": "Please do not edit this file, "+expandedfiles+"!\nIt was generated automatically from its master file "+files+", by programs/expand.js\nGet everything from https://github.com/haroldthimbleby/improving-science\n\n";
@@ -122,16 +125,17 @@ for (var i = 0; i < TeXFiles.length; i++) {
     		// remove generated/ from \includegraphics[width=\imageWidth]{generated/mathematicaplot.\imageFileSuffix}
         	bufferExpand = bufferExpand.replace(REgraphics, function (match, graphicsDetails) {
             	replaced = true;
-                console.log("      * \\includegraphics["+graphicsDetails+"} requires graphics file to be copied from generated/");
-                var gfile = graphicsDetails.replace(/.*\]\{/,"").replace(/\\imageFileSuffix/, "jpg");
-                cpFiles += "cp "+gfile+" expanded-"+(gfile.replace(/generated\//, ""))+"\n";
-                return "% removed generated/ from graphics pathname\n\\includegraphics["+(graphicsDetails.replace(/generated./, "expanded-")+"}");
[…]

### 2023-07-03 appendix.tex
Fixes to improve text for Computer Journal submission

@@
 \begin{quote}
-At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.\end{quote}
+At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``\texttt{\ldots~code coming soon~\ldots}'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.\end{quote}
 
@@
 
-Unfortunately, although using generated variables and analyses from a paper's data is a very simple technique to help make published papers more reliable, some journals and preprint servers (such as \emph{IEEE Transactions on Software Engineering}, \emph{PLOS ONE}, and \emph{arXiv\/}) do not permit papers to be submitted using \LaTeX\ source code that uses the standard \texttt{\textbackslash input}, \texttt{\textbackslash bibliography}, and other related commands.
+Unfortunately, although using generated variables and analyses from a paper's data is a very simple technique to help make published papers more reliable, some journals and preprint servers (such as \emph{IEEE Transactions on Software Engineering}, \emph{PLOS ONE}, \emph{arXiv}, and even the \emph{Computer Journal}, which published this paper) do not permit papers to be submitted using \LaTeX\ source code that uses the standard \texttt{\textbackslash input}, \texttt{\textbackslash bibliography}, and other related commands.
 Typically they also do not support running any data collection or analysis either (which the present paper does when it clones repositories).
@@
 
-Another program (\texttt{programs/expand.js}) was therefore written to recursively expand included files so the expanded version can be submitted adhering to any such restrictive policy.
-Of course, the expanded version now contains all variables as fixed constants, so the submitted paper is misleading and useless to other researchers if the data is modified --- the effort to ensure all published numbers are automatically correct is defeated.
-Such restrictive publishing policies undermine reproducibility.
+Another program (\texttt{programs/expand.js}) was therefore written to recursively expand included \LaTeX\ files so the expanded version can be submitted to any journal adhering to any such restrictive policies.
+It also renames and flattens the file hierarchy (e.g., copying images to be in the same directory as the expanded main \LaTeX\ file), as the journal web site did not cope with standard \LaTeX\ file naming conventions.
+A consequence is that the published version now contains all generated text and variables as fixed constants, so the published paper is misleading and useless to other researchers if any data is modified --- the effort to ensure all published data are automatically correct is defeated.
+Such technically limiting publishing policies undermine reproducibility.
+Fortunately, in the present case the original data and documents are still available on the Github site.
+
+Clearly, there is still a long way to go. Such problems will put authors off using computational papers and \RAPstar\ methodologies.
 
@@
 
+\section{Errata and updates to the paper}
+
+\emph{No material added yet.}
+
 \immediate\closeout\infofileForSupplement

### 2023-07-03 expanded/EXappendix.tex
Fixes to improve text for Computer Journal submission

--- 2023-06-09 EXappendix.tex
@@
 \bibcite{swebok}{{{139}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
 \bibcite{wpf}{{{140}{}{{}}{{}}}}
@@
 \bibcite{managers}{{{142}{}{{}}{{}}}}
-\@writefile{toc}{\contentsline {section}{\numberline {13}Additional references for Supplemental Material}{64}{section.13}\protected@file@percent }
 \bibcite{differential-equations}{{{143}{}{{}}{{}}}}
@@
 \bibcite{ref-4}{{{178}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
 \bibcite{ref-5}{{{179}{}{{}}{{}}}}
-\@writefile{toc}{\contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}\protected@file@percent }
 \bibcite{ref-6}{{{180}{}{{}}{{}}}}
@@
 \bibcite{ref-32}{{{206}{}{{}}{{}}}}
+\@writefile{toc}{\contentsline {section}{\numberline {15}Errata and updates to the paper}{69}{section.15}\protected@file@percent }
 \gdef \@abspage@last{34}
@@
 \citation{trisovic}
+\@writefile{toc}{\contentsline {subsection}{\numberline {2.a}Code quality concerns}{4}{subsection.2.1}\protected@file@percent }
+\newlabel{related-work}{{2.a}{4}{Code quality concerns}{subsection.2.1}{}}
 \citation{jupyter-study}
@@
 \citation{trisovic}
-\@writefile{toc}{\contentsline {subsection}{\numberline {2.a}Code quality concerns}{4}{subsection.2.1}\protected@file@percent }
-\newlabel{related-work}{{2.a}{4}{Code quality concerns}{subsection.2.1}{}}
 \citation{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis}
@@
 \citation{spec-led-design}
-\citation{RAPstandard}
 \@writefile{toc}{\contentsline {subsection}{\numberline {2.c}RAP: Reproducible Analytical Pipelines}{7}{subsection.2.3}\protected@file@percent }
 \newlabel{RAP-section}{{2.c}{7}{RAP: Reproducible Analytical Pipelines}{subsection.2.3}{}}
+\citation{RAPstandard}
 \citation{rap,goldacre,turing-way}
+\citation{MMR-fraud}
+\citation{a=b}
+\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{8}{subsubsection.2.3.1}\protected@file@percent }
+\citation{Courtes}
 \@writefile{lot}{\contentsline {table}{\numberline {4}{\ignorespaces A simplified schematic of the publication pipeline.
@@
 The encoded RAP{\tt  +}\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers.
-Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the Supplemental Material.}}{8}{table.4}\protected@file@percent }
-\newlabel{fig-pipeline}{{4}{8}{A simplified schematic of the publication pipeline.
+Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the Supplemental Material.}}{9}{table.4}\protected@file@percent }
+\newlabel{fig-pipeline}{{4}{9}{A simplified schematic of the publication pipeline.
 For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement.
@@
 Additional schematics are provided in section \ref {supplement-RAP-diagrams} in the \supplement }{table.4}{}}
-\@writefile{lot}{\contentsline {table}{\numberline {5}{\ignorespaces A minimum standard of RAP, based on the UK Statistics Authority summary \cite  {RAPstandard}.}}{8}{table.5}\protected@file@percent }
-\newlabel{figure-RAP-summary}{{5}{8}{A minimum standard of RAP, based on the UK Statistics Authority summary \cite {RAPstandard}}{table.5}{}}
-\citation{MMR-fraud}
-\citation{a=b}
-\citation{Courtes}
-\@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{9}{subsubsection.2.3.1}\protected@file@percent }
+\@writefile{lot}{\contentsline {table}{\numberline {5}{\ignorespaces A minimum standard of RAP, based on the UK Statistics Authority summary \cite  {RAPstandard}.}}{9}{table.5}\protected@file@percent }
+\newlabel{figure-RAP-summary}{{5}{9}{A minimum standard of RAP, based on the UK Statistics Authority summary \cite {RAPstandard}}{table.5}{}}
 \@writefile{toc}{\contentsline {subsubsection}{\numberline {2.c.2}RAP as research}{9}{subsubsection.2.3.2}\protected@file@percent }
@@
 \citation{paul-hci,pnas-stats}
-\citation{Ben,se-bias}
 \@writefile{toc}{\contentsline {section}{\numberline {3}The statistics/code analogy}{10}{section.3}\protected@file@percent }
 \newlabel{statistics-computation-analogy}{{3}{10}{The statistics/code analogy}{section.3}{}}
+\citation{Ben,se-bias}
 \citation{sas9.2}
@@
 \newlabel{PRISMA-statement}{{4.c}{15}{The critical role of code is often ignored}{subsection.4.3}{}}
-\@writefile{toc}{\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{15}{subsection.4.4}\protected@file@percent }
-\newlabel{knowledge}{{4.d}{15}{Bugs, code and programming}{subsection.4.4}{}}
 \citation{knuth}
@@
 \citation{popper-conjectures-refutations}
+\@writefile{toc}{\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{16}{subsection.4.4}\protected@file@percent }
+\newlabel{knowledge}{{4.d}{16}{Bugs, code and programming}{subsection.4.4}{}}
 \@writefile{toc}{\contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}\protected@file@percent }
@@
 \citation{codecheck-certificate}
-\citation{codecheck}
-\citation{codecheck-certificate}
 \newlabel{discuss-covid-sim}{{5.a}{18}{Case study: Pandemic modeling}{subsection.5.1}{}}
@@
 \newlabel{reproducibility-concerns}{{5.b}{18}{Concerns with reproducibility}{subsection.5.2}{}}
+\citation{codecheck}
+\citation{codecheck-certificate}
 \citation{codecheck,thumbs-up}
@@
 \@writefile{toc}{\contentsline {subsection}{\numberline {6.d}Action must be interdisciplinary}{25}{subsection.6.4}\protected@file@percent }
-\@writefile{toc}{\contentsline {subsection}{\numberline {6.e}Methodological statements}{25}{subsection.6.5}\protected@file@percent }
 \citation{no-raw-data}
@@
 \citation{debt}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.e}Methodological statements}{26}{subsection.6.5}\protected@file@percent }
 \newlabel{mentionPLOSandIEEE}{{6.e}{26}{Methodological statements}{subsection.6.5}{}}
-\@writefile{toc}{\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{26}{subsection.6.6}\protected@file@percent }
-\newlabel{technical-debt}{{6.f}{26}{Training to reduce technical debt}{subsection.6.6}{}}
 \citation{fixit}
@@
 \citation{jpm}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{27}{subsection.6.6}\protected@file@percent }
+\newlabel{technical-debt}{{6.f}{27}{Training to reduce technical debt}{subsection.6.6}{}}
 \@writefile{toc}{\contentsline {subsection}{\numberline {6.g}Benefits beyond science}{27}{subsection.6.7}\protected@file@percent }
@@
 \citation{hoare-unreasonable}
+\@writefile{toc}{\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}\protected@file@percent }
 \citation{excel-fiasco}
 \citation{science-delays}
-\@writefile{toc}{\contentsline {subsection}{\numberline {6.h}Approaches to further work}{28}{subsection.6.8}\protected@file@percent }
 \citation{tweet,ferguson-interview}
@@
 \contentsline {section}{\numberline {14}References for surveyed papers}{66}{section.14}%
+\contentsline {section}{\numberline {15}Errata and updates to the paper}{69}{section.15}%
 

### 2023-07-03 expanded/EXpaper.tex
Fixes to improve text for Computer Journal submission

--- 2023-06-09 EXpaper.tex
@@
 \affiliation{See Change Fellow in Digital Health, Swansea University}
-\email{harold@thimbleby.net}
+\email{h.w.thimbleby@swansea.ac.uk}
 
@@
 \accepted{30 May 2023}
+\paperdoi{10.1093/comjnl/bxad067}
 \volumenumber{00} \issuenumber{0} \volumeyear{2023}
@@
 \contentsline {subsection}{\numberline {2.c}RAP: Reproducible Analytical Pipelines}{7}{subsection.2.3}%
-\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{9}{subsubsection.2.3.1}%
+\contentsline {subsubsection}{\numberline {2.c.1}Potential code}{8}{subsubsection.2.3.1}%
 \contentsline {subsubsection}{\numberline {2.c.2}RAP as research}{9}{subsubsection.2.3.2}%
@@
 \contentsline {subsection}{\numberline {4.c}The critical role of code is often ignored}{15}{subsection.4.3}%
-\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{15}{subsection.4.4}%
+\contentsline {subsection}{\numberline {4.d}Bugs, code and programming}{16}{subsection.4.4}%
 \contentsline {subsection}{\numberline {4.e}Long-term problems of unreliable code}{16}{subsection.4.5}%
@@
 \contentsline {subsection}{\numberline {6.d}Action must be interdisciplinary}{25}{subsection.6.4}%
-\contentsline {subsection}{\numberline {6.e}Methodological statements}{25}{subsection.6.5}%
-\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{26}{subsection.6.6}%
+\contentsline {subsection}{\numberline {6.e}Methodological statements}{26}{subsection.6.5}%
+\contentsline {subsection}{\numberline {6.f}Training to reduce technical debt}{27}{subsection.6.6}%
 \contentsline {subsection}{\numberline {6.g}Benefits beyond science}{27}{subsection.6.7}%
@@
 
-Finally, there is still a long way to go.
-The \emph{Computer Journal}, which should be a leading journal for publishing computational science (and which has published this paper) uses poorly documented automated and buggy paper processing software, whose obscure bugs took several weeks of hard work to overcome in order to submit the present paper.
-Such problems will put authors off using computational papers and \RAPstar\ methodologies.
-
 \subsection{Training to reduce technical debt}\label{technical-debt}
@@
 \section*{Supporting information}
+
+%NEW
+Updates and errata are included in the \supplement, which is kept up to date at \url{github.com/haroldthimbleby/improving-science}
+
 \newcount\csrefcount \csrefcount=0

### 2023-07-03 models/psych/pizza.tex
Fixes to improve text for Computer Journal submission

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

%\usepackage{fancyvrb}
%\usepackage{relsize}
%
%\RecustomVerbatimEnvironment
%  {Verbatim}{Verbatim}
%  {fontsize=\relsize{-1},formatcom=\vskip .2cm,samepage}

\title{The Pizza Problem \\ and its illustration of risks with psychological modeling}
\author{Harold Thimbleby}
%\email{harold@thimbleby.net}
\begin{document}
\maketitle

\begin{abstract}
\noindent
A reflection on the unnoticed complexity and unreliability of psychological models (and, by implication, the unreliability of scientific models in general) as motivated by the apparently simple problem, specification, and code openly proposed and illustrated in G Olivia \& A E Martin ``How computational modeling can force theory building in Psychological Science,'' \emph{Perspectives on Psychological Science} %, \textbf{16}(4):789--802, 2021. 
\cite{pizzap}.

\begin{quote}\flushright
--- {\sf Through writing code, we debug our scientific thinking.} \\ G Olivia \& A E Martin, \cite{pizzap}
\end{quote}
\end{abstract}

\section{Introduction}
The Pizza Problem is raised in \cite{pizzap} in the context of arguing that demands of computational modeling guide us toward better science by forcing us to conceptually analyze, specify, and formalize intuitions that may otherwise remain unexamined. A specification and implementation of the Pizza Problem were provided to illustrate the paper.

Many computer models are complex, even running into tens of thousands or more lines of code \cite{harold-cj}, so discussions of modeling are often eclipsed by the complexity of the domain, the theory, the model, and their assumptions. Often domain theories are extremely complex, for instance expressed as systems of differential equations applied to large data, so refining them into executable models can only be done by making simplifications --- the pragmatic informality introduced in the coding process belies the superficial precision of the theory, and belies the apparent rigor of any published claims based on model results. In contrast, an important aspect of the Pizza Problem discussed here is that it is very simple: the problem itself can be stated in a single sentence, and the mathematics to model it is elementary.

The paper ``Improving science that uses code'' \cite{harold-cj}, within a larger discussion of science that depends on code, points out by way of an example that the Pizza Problem model presented in \cite{pizzap} is incorrect. In the present paper, we carefully examine the model specification and code, and hence discuss the fragility of modeling more generally.

An important point is that very little of the present discussion could have been developed if the original paper \cite{pizzap} had not openly presented all of the problem, specification, and complete code. Analogously, other work exploring or correcting errors in any science based on computational models is unlikely to be adequate without similar completeness and openness to sharing both specification and code.

In general, running a computer model promises to improve scrutiny and uncover weaknesses in the model, because the computer runs exactly what it is told, without being biased or motivated for the model to confirm a pet theory. (Sadly scrutiny is often limited to the research team, because the model is not openly shared for independent scrutiny.) The computer itself provides a new sort of scientific scrutiny: a computer runs \emph{exactly\/} what it is told, and human error, including misunderstanding how to correctly implement a model in code, as well as emotively motivated tweaks to the code (often casually introduced during debugging when test results diverge from what was hoped for), combine to ensure that results from running code may be misleading, even deceptive. Often the scientists' models are run on systems developed elsewhere, such as proprietary programming languages or statistics packages, and they are therefore relying on those systems being correct (and correctly documented) --- which they often are not \cite{heedless}. Furthermore, a computer is technically precise in how it runs code, and therefore implemented models may have detailed properties and behavior that require software engineering expertise to ensure is correct, without which na\" \i ve scientists may take, and perhaps write up, an artifact of running a program as an intended consequence of running a poorly-defined or poorly-implemented model.

\section{The Pizza Problem}
The discussion is motivated by a simple problem:

\def\orderA{{\sf\bfseries a}}
\def\orderB{{\sf\bfseries b}}

\begin{quote}\sf
Your favorite pizzeria has a special: two 12 inch pizzas for the price of one 18 inch pizza. Your definition of a good deal is one in which you purchase the most food. Is this a good deal?  \cite{pizzap}
\end{quote}

As the two options are the same price, you can purchase either at the same price, and hence obtain a good deal regardless, assuming at least one of the choices is a good deal. So, since you have a free choice, necessarily, this is a good deal.

A more precise definition of a good deal is perhaps, given that the offer is two choices at the same price, will you choose the larger amount of pizza and not be exploited by being tricked into buying less pizza for the same price? If this is the definition of a good deal, you are still likely to agree it is a good deal, since you \emph{feel\/} free to choose the more generous offer. However you may not realize that you could be deceived: there may be more to this deal that at first it appears. The problem is more subtle than the basic discussion taken from the paper makes it seem.

Perhaps the offer should be more blatant in the deal the pizzeria is offering --- and hoping to profit from at the customer's expense:

\begin{quote}\begin{center}\sf\framebox[2.75in]{\vbox{~\\Special offer! \\~
\\ Get \textbf{TWO} generous 12 inch pizzas\\ for the price of \textbf{ONE} 18 inch pizza!\\~\\} }
\end{center}\end{quote}

Indeed, even how the question is posed, say whether it arises in a natural encounter in an actual pizzeria, or, at another extreme, in a laboratory with white-coated investigators, would likely induce different or different distributions of answers. The paper \cite{pizzap} reports that expressing the problem on Twitter gained diverse replies, with participants expressing surprise and disbelief at the right answer; even, some tweeted that two pizzas are more than one pizza. Indeed, the question might also be presented in a peer reviewed paper (such as \cite{pizzap} itself) and thereby elicit further types of response. The specific wording and presentation of the problem is a source of different assumptions about the required answer.

Refer to figure \ref{supplement-fig:pizzas} to consider again which equally-priced choice, order \orderA\ or order \orderB, is the one ``in which you purchase  more food'' (for the same price).

\begin{figure}[t] 
   \centering
   \begin{tabular}{c@{\hskip 1cm}|@{\hskip 0.7cm}c}
   Order \orderA & Order \orderB\\
   Two 12 inch pizzas & One 18 inch pizza \\  \hline \\
   \includegraphics[angle=90,origin=c,width=1.2in]{pizza.jpg} 
   &
   \setbox0 = \hbox{\includegraphics[width=1.8in]{pizza.jpg}}
   \setbox0 = \hbox{\lower 0.9in \copy0}
   \dp0 = 0in
   \copy0\\
   \includegraphics[angle=-90,origin=c,width=1.2in]{pizza.jpg}
   \end{tabular}
   \caption{If these two pizza orders (drawn to scale in the figure) are offered at the same price, does order \orderA\ or \orderB\ provide more food? (Figure based on figure 1 in \cite{pizzap}.)}
   \label{supplement-fig:pizzas}
\end{figure}

Now the choices are visualized in a figure, it seems clear that the question is intended to be:

\begin{quote}\sf
If you can see or imagine two 12 inch pizzas offered for same the price as one 18 inch pizza (see figure \ref{supplement-fig:pizzas}), and you wish to buy the most pizza for the money, are you likely to correctly choose the greater amount of pizza and thus not be exploited by the pizzeria hoping to sell you less pizza for the same price?
\end{quote}

Evidently, the Pizza Problem raises non-trivial psychological issues, whether in vision or cognitive psychology, to say nothing of different people preferring crust and others a tomato-covered base, and so forth. Divertingly interesting as that may all be, the paper \cite{pizzap} in fact defines ``the pizza problem'' as the modeling issues that the problem succinctly illustrates, namely:

\begin{quote}\sf
Experience is seeing our ideas being executed by a computer, giving us the chance to debug scientific thinking in a very direct way. If we do not make our thinking explicit through formal modeling, and if we do not bother to execute (i.e., implement and run our specification through computational modeling), we can have massive inconsistencies in our understanding of our own model(s). We call this issue ``the pizza problem.''
\cite{pizzap}
\end{quote}

This is an insightful point, but there is much more to the Pizza Problem synecdoche than the paper suggests.

\section{Specifying and implementing the Pizza Problem}
The paper \cite{pizzap} proposes three elementary mathematical equations for a specification. Unprimed equation numbers are copied over from the original paper.

The amount of food $\phi$ per order $i$ is given as:

\begin{equation}\label{eqn-area}
\phi_i = N_i \pi R_i^2
\end{equation}

As defined above, $\phi_i$ specifies the total geometric area of $N_i$ pizzas all ordered with equal radius $R_i$. The formula has no visual or cognitive factors.

A decision rule is then given to specify the pizzeria order with most food:
\begin{equation}\label{eqn-specification}
\omega(\phi_i,\phi_j)=\left\{\begin{tabular}{ll}
$i$,& if $\phi_i>\phi_j$\\
$j$,&otherwise
\end{tabular}\right.
\end{equation}

This is incorrect, since if $\phi_i=\phi_j$ the biggest is not $i$ as in this case the orders $i$, $j$ have equal largest areas. The specification $\omega$ is also incorrect on the basis that the problem asks whether it is correct that the offer is a good deal; the answer to that question (whatever the question precisely may be, as discussed above) is a truth value not an order identifier. Notwithstanding that the formula has no psychological factors, nor any factors to disentangle the meaning of the question, one possible corrected definition of $\omega$ is:

\begin{equation}\tag{$\ref{eqn-specification}^\prime$}\label{eqn-corrected-specification}
\omega(\phi_i,\phi_j)=\left\{\begin{tabular}{ll}
$i$,& if $\phi_i>\phi_j$\\
$\{i,j\}$,&if $\phi_i=\phi_j$\\
$j$,&if $\phi_i<\phi_j$\\
\end{tabular}\right.
\end{equation}

\def\psych{\overline{\phi}}
The set $\{i,j\}$ seems over-complicated as an answer to the Pizza Problem as posed. The actual meaning of $\phi_i=\phi_j$ is that both choices are of equal value to the customer, and the answer presumably is ``yes, that is a good deal,'' not a set. Something is not quite right. One solution is to introduce a psychologically-motivated version of $\phi$, say $\psych$, then a plausible formulation of $\omega$ would be

\def\order{\mbox{order}}
\begin{equation}\tag{$\ref{eqn-specification}^{\prime\prime}$}
\omega_{i,j}= \order(\psych_i, \psych_j) = \order(\phi_i,\phi_j)
\end{equation}

\ldots\ that is, the deal is a good one when the ordering of the participant's subjective assessment of the offer total areas is equal to the objective geometric ordering of the total areas --- that is, the participant is not fooled. This formulation implicitly raises the question of modeling a mental $\overline{\order}$, since a psychological representation of $\order(\psych_i, \psych_j)$ is implausible; and so forth.

Arguably, in the original paper then, an important benefit of computer modeling has been missed: developing a rigorous model specification (equation \ref{eqn-specification}) because correct computer code requires it has apparently not led to clarifying the problem specification itself.

A computational concern raised by equation (\ref{eqn-corrected-specification}) is that $\omega$, as specified, can only have a singleton value or a set (possibly multiset) of two values. Code that only handles a finite maximum number of cases (here, the maximum is 2) is likely to be unnecessarily restrictive.

More general, and hence more robust, code will handle an unlimited number of cases. One such generalized specification of $\omega$ inspired by this scrutiny of equation (\ref{eqn-specification}) is as follows:

\begin{equation}\tag{$\ref{eqn-specification}^{\prime\prime}$}\label{eqn-set-corrected-specification}
\omega=\left\{i| \phi_i=\max(\phi)\right\}
\end{equation}

Note that this definition is `automatically' an empty set if there is no such $i$.

Equation (1) is subsequently generalized in \cite{pizzap} to allow different radius pizzas within a single order, as follows:
\begin{equation}\label{eqn-area-generalized}
\phi_i=\sum_{j=1}^N \pi R_j^2
\end{equation}

This is incorrect, since in equation (\ref{eqn-area}) $R_i$ was the radius of order $i$, not the radius of pizza $j$ in order $i$. The radius of pizza $j$ in order $i$ would be better written as $R_{i,j}$.

The summation in equation (\ref{eqn-area-generalized}) is given over $N$, but since pizza orders --- as in the Pizza Problem itself --- may have different numbers of pizzas, it should be over $N_i$.

Hence a corrected form of equation (\ref{eqn-area-generalized}) is:

\begin{equation}\tag{$\ref{eqn-area-generalized}^\prime$}\label{eqn-corrected-area-generalized}
\phi_i=\sum_{j=1}^{N_i} \pi R_{i,j}^2
\end{equation}

In \cite{pizzap}, Python code is presented to implement $\phi$ as \texttt{food} (from equation \ref{eqn-area}), and anonymous code is then presented to implement $\omega$ (from equation \ref{eqn-specification}):

\begin{verbatim}
# Decision rule (eq. 2): 
print(food(two_pizzas) > food(one_pizza))
\end{verbatim}

The code shown above yields a truth value, not $i$ or $j$ as actually specified by $\omega$ in equation (\ref{eqn-specification}). It happens to print \texttt{False}, but, even so, it does not meet the original English description of the problem as stated in the paper, since it does not address whether the offer is a good deal or not --- specifically, it calculates the truth value whether two 12 inch pizzas have a larger geometric area than one 18 inch pizza.

Psychology is not modeled here. Participants may think or be misled (e.g., by an optical illusion) that two 12 inch pizzas represent more food than one 18 inch pizza, and if so they will be fooled and not choose a good deal. To answer the question posed, if they are fooled the correct calculated answer to the question should be \texttt{False}, which \emph{only coincidentally\/} is the same answer as the paper's Python calculation whether two 12 inch pizzas have a larger area than one 18 inch pizza --- the given code is only providing the right answer because it is answering the wrong question.

Had the two pizza orders in the offer been of the same total area, the program would also have answered \texttt{False}, which in this case would be incorrect, since choosing either option would (on this assumption) have been an equally good deal. The answer the program provides is correct only because, again purely by chance, it was given data giving different areas. The code should have checked that the total areas were different since it relies on that assumption.

An interesting issue implicit in the paper's implementation is that the Python variables \verb|one_pizza| and \verb|two_pizzas| have values that are arrays of diameters of pizza orders (e.g., \texttt{[18]}, \texttt{[12,12]}), whereas the specifications for $\phi$ and $\omega$ both refer to identities of orders (e.g., $i, j$, etc). Arguably given the specification a programmer could legitimately choose that the implementation uses 1 or 2 (order numbers or other order identifiers) but \emph{not\/} the concrete details of the pizza diameters the specification has abstracted away, but which is what these Python variables denote. This point will be addressed below.

\section{A corrected implementation}
Python code implementing a corrected specification, equations (\ref{eqn-set-corrected-specification}) and (\ref{eqn-corrected-area-generalized}), is as follows.

As a result of defining $\omega$ as a set in equation (\ref{eqn-set-corrected-specification}), the code now developed works with any number of orders, including zero orders and empty orders (both orders of zero pizzas and orders of pizzas of zero diameter). Note that this corrected implementation, like the original, is \emph{a\/} model of the problem, and like the original it fails to support any psychological interpretation of experimental data. Nevertheless there are still important observations to be reported here on the computational modeling process, regardless of such details.

We start with simple housekeeping to define the value $\pi$ for the program:

{\tt\begin{tabbing}
import math\\

$\pi$ = math.pi
\end{tabbing}}

Next, we use a Python dictionary to define the mapping $\mbox{\it order} \mapsto \mbox{\it pizza-diameter}^{*}$, which is initialized with the given problem data. The dictionary, rather than the simple real-valued array variables as used in \cite{pizzap}, enables the corrected implementation to use the identifier of the greatest total area order(s) as defined in equation (\ref{eqn-set-corrected-specification}). This observation relies on an elementary but critical programming distinction resolved by Christopher Strachey in 1967 \cite{strachey}.

{\tt\begin{tabbing}
\# A map of Pizza orders\\
\# Variable orders[] maps order ID to a list of pizza diameters in inches\\ \\
orders = \{\\
\ \ \ \="\orderA": [12, 12], \=\# Order \orderA\ is two 12 inch pizzas\\
   \>"\orderB": [18] \>\# Order \orderB\ is one 18 inch pizza\\
\}
\end{tabbing}}

Note that Python will ignore if dictionary keys are repeated, so if this is a potential error it should be checked for independently of the implementation of the map shown above. If, for instance through a typo, they were both called \orderA, say, then Python would incorrectly, unfortunately without reporting any error, only assign one order to \texttt{orders[]}. In the case here there are only two orders and it is easy (but essential) to eyeball that they are correctly recorded in the Python code as two independent entries.

Implementing the sum in $\phi$ is a routine matter of translating mathematics to Python idioms, noting the quite trivial conversion of diameters to radii as required by the mathematical definition of $\phi$:

{\tt\begin{tabbing}
\# Implement $\phi$ from (\ref{eqn-corrected-area-generalized})\\
\# Tota area of all pizzas in orderID\\
def phi(orderID):\\
\ \ \ \=totalArea = 0\\
   \>for diameter in orders[orderID]:\\
   \>\hskip 1em\=totalArea += $\pi$ * (diameter/2)**2\\
   \>return totalArea
\end{tabbing}}

Given the nature of the problem involving a low number of modestly-sized pizzas, no effort is made here to check for numerical over- or under-flow. (In general, numerical checking and other sanity checking should always be undertaken in case computer errors lead to incorrect results that are not flagged as such.)

Implementing $\omega$ is slightly more complex; for example, it is an obligation to prove that a Python list correctly implements the mathematical set $\omega$ in this case. Relevant details of the methodology used can be found in Dijkstra \cite{dop} and elsewhere in the standard computer science literature. Note that Dijkstra and others make it clear that the present paper just discussing a specification (and alternatives) and then presenting the resulting code has omitted a considerable body of mathematical reasoning that has had to be taken for granted to save space and patience. We will discuss this issue more comprehensively, below, in section \ref{section:pedantry}.

{\tt\begin{tabbing}
\# Implement $\omega$ from equation (\ref{eqn-set-corrected-specification}) \\
\# Determine maximum area order(s) over all orders\\
maxArea = 0\\
maxAreaIDs = []\\
for orderID in orders :\\
\ \ \ \=area = phi(orderID)\\
   \>if maxArea == area : \\
   \>\ \ \ \=maxAreaIDs.append(orderID)\\
   \>if maxArea < area  :\\
   \>   \>maxArea = area\\
   \>   \>maxAreaIDs = [orderID]
\end{tabbing}}

After the calculation above, the corrected result, allowing for the $\phi_i=\phi_j$ case and including empty orders, zero diameter pizzas, and multiple equal greatest total areas, can be obtained by writing

{\tt\begin{tabbing}
print(maxAreaIDs)
\end{tabbing}}

If English is preferred, the result can be formatted more understandably:

{\tt\begin{tabbing}

print\=(\="Pizza order", maxAreaIDs, \\
     \> \>"has largest area =", round(maxArea, 1), "sq in"\\
     \>)
\end{tabbing}}

To improve the English to cover the grammatically distinct cases when there are no orders, a single order, or when there is more than one greatest area order, the wording can be improved to depend straight-forwardly on \texttt{N} (implementing the specification $N=|\omega|$), the number of greatest orders:

{\tt\begin{tabbing}
N = len(maxAreaIDs)\\
if N == 0 : \\
\ \ \ \=print("No pizza orders were made")\\
else :\\
      \> print\=(\="Pizza", "orders" if N > 1 else "order", \\
	  \>      \>\>maxAreaIDs, "has" if N == 1 else "have equal",\\
	  \>      \>\>"largest area =", round(maxArea, 1), "sq in"\\
	  \>      \>)
\end{tabbing}}

If concerned (for instance to copy and paste the results into a paper we intend to publish) we could also improve the formatting of the value of \texttt{maxAreaIDs} in the print statement above. As written above, \texttt{maxAreaIDs} will appear in Python notation, not in English nor even in standard mathematical set notation. It is not hard to correct this shortcoming, but to do so would be a diversion from the thread of this paper.

The output of the program is now ``more'' than required by the specification, particularly equations (\ref{eqn-specification}) or (\ref{eqn-set-corrected-specification}): in addition to correcting bugs and using English to clarify the output, this code generalizes the original specification to additionally cover all cases $N \ne 2$, as well as providing the area of the order(s) with greatest area. (Python dictionaries cannot have $N<0$ elements, so necessarily $N\ge0$.)

These generalizations are legitimate, since in general specifications logically imply correct code, so code can do more unless it is inconsistent with the specification. Note that a correct implementation of the specification has not resulted in code that provides a valid answer to the original question, because the question was not correctly specified.

Typically one would show that a projection of the code's output is implied by the specification, and one would provide an assurance case that the specification correctly captured the relevant essence of the original problem.\footnote{For an authoritative overview of assurance case practice, see \cite{assurance}.} Since in all but the simplest cases proving such an implication will be too demanding, critical parts of the code are generally constructed to be correct (so-called correct-by-construction, CBC) and an assurance case is separately made that the remainder of the code (for example, the user interface) does not affect the correctness of the required results. In the present case, for brevity, we omitted both details of constructing a correct implementation of $\omega$, as well as of the necessary testing.\footnote{Specification and refinement are subject to human error, but even if a specification and refinement to code are correct, programmers may make slips and introduce unnoticed errors in the coding process. Testing is essential to check correctness of implementations; furthermore, testing itself should be checked as it itself may contain unknown errors.} Curiously, testing correct grammatical English generation was more tedious than testing the implementation of $\omega$.

\section{Modeling the observed diverse preferences}
The specification and implementation are envisaged (in the paper's figure 2) to be part of an iterated serial process, being the middle two steps of six. In their scheme a cognitive framework (such as ACT/R \cite{actr}) comes first, and, notably, data comes last. Yet in their own discussion observed data of the Pizza Problem came first.

The modeling above shows that a correct rational ``algorithmic person'' preferring pizzas by area would consistently prefer option \orderB, as the program is clearly deterministic and produces the same result on every run. However, interest in the Pizza Problem was originally motivated by the observation of diverse answers to the question posed, thus motivating a psychological inquiry and building a model. The  code presented in \cite{pizzap} does not attempt to model the variation in the motivating data. One might then seek a more general model to explain the observed variation in participant preferences.

For example, models based on a linear combination $\alpha \phi(t)+ (1-\alpha)[\phi(r)-\phi(t)]$, for pizzas radius $r$ with toppings radius $t$ ($t\leq r$), could be hypothesized to classify data from people preferring topping or crust. Other variables are the price (or expected price) of the pizza orders, and whether the participant is answering the verbal question ``looks larger'' (which implies visual introspection) or ``is larger'' (which more likely implies a geometric and purely cognitive question to the participant); or perhaps the question is asked illustrated with scale pictures, or even pizzas to both look at and smell. And so on.

To do any such modeling would require data, which is not provided in \cite{pizzap}.

\section{Is this pedantry?}\label{section:pedantry}
Is our discussion unnecessary pedantry about one paper's specification and code? Indeed, almost uniquely in this case, the original code in the paper \cite{pizzap} was merely an illustration of a methodology the paper proposed, and nothing in the paper really depended on the code actually being correct and doing what was claimed.

On the other hand, the original paper itself sets a very high standard of rigor: it argues that by providing an explicit computational model --- they mention using the formal methods $\mbox{TLA}^{+}$ \cite{tla} or Z \cite{z}, which are themselves very rigorous formal notations. Using such methods, in the authors' own words, they would debug scientific thinking.

Ironically, both the specification and code in the original paper are incorrect, and a human reader, given that both are trivial and arguably both just sketches, might choose to gloss over the errors as pedantry. One might note that the actual points about psychological modeling the paper is discussing superficially do not depend on the accuracy, correctness, or precision of either the specification or the code in the examples given. (That is, the model shown merely illustrates the arguments of the paper; it is not a direct model the running of which implies any substantial claims made in the paper.) Yet to gloss over rigor in this process would miss the opportunity to uncover that the problem has been inadequately specified and that even the simple implementation is incorrect. The implementation provides a superficially correct answer despite invalid reasoning. How can one debug scientific thinking that is not presented clearly, and, in the present case, is sufficiently vague to conceal incorrect calculations? Potentially the paper \cite{pizzap} has missed some critical steps in its proposed methodology? Indeed, the present paper has made various suggestions, including assurance cases, testing --- also leaving a lot of basic computer science unspoken --- all of which appeared necessary to get even the simple Pizza Problem to work sensibly.

The coding issues raised in the present paper will undoubtedly feel like pedantry for people who do a lot of programming without feeling they need to worry about formal computer science issues.

Their code works.

They think.

It is a difference in disciplinary perspectives.

In my view, if you want to do coding, you may as well do it properly. Doing code properly is a rigorous discipline that leads to better and more dependable models and programs. It is important particularly if the code is critical, for instance perhaps for a psychology paper informing education or mental health interventions. Other papers criticized in \cite{harold-cj} used unreliable computer models to inform public health measures during the COVID-19 pandemic.

It is the same as an electrician `pedantically' pointing out that your extension lead needs an earth; they say it needs to be wired properly. You may well have a faulty extension lead without an earth that you have used successfully many times (musicians often remove the earth deliberately \cite{earthlift}, risking fatalities), so you think this is a pedantic demand over a trivial issue. Yet you are unnecessarily risking electrocuting yourself when doing the ironing or playing the keyboard. What is more, if you are playing with a band or in public [in the UK or EU], correct wiring is a legal requirement. An electrician's `pedantry' over a `trivial' issue will save somebody being killed, and will save you from a manslaughter conviction.

Can good music be based on sloppy wiring? Only by chance --- in that by chance you risk electrocuting a musician and having no music to listen to; by chance alone you hope the musician will not die. On the other hand, using proper professional wiring (e.g., earth bonding, protective devices, isolated supplies, extra low voltage supplies, etc) you can reduce the risk of electrocution to vanishing levels. Sadly, most of the time non-electrically qualified people cannot even tell the difference, and very likely they may not know how to tell the difference either.

Can good science be based on sloppy code, and sloppily derived code from sloppy specifications? Only by chance. And who wants to rely on chance when you can do better? Why waste time --- why waste other people's time trying to read, understand, and develop your science --- by doing sloppy science.

\section{Consider a spherical cow \ldots}
A thought-provoking joke in physics begins ``consider a spherical cow \ldots'' Superficially it seems absurd, yet abstraction is a common, and productive modeling approach in any subject. In physics, a spherical cow has the same volume and mass as an actual cow-shaped cow, and can be used for many valid calculations,\footnote{For example, Isaac Newton showed that the $F=ma$ force acting on any body of any shape was the same as if that body was mathematically reduced to a point mass (of the same mass) placed at the center of gravity of the body. In fact, the rigorous concept of center of gravity is precisely a spherical cow (or anything) of the same mass and radius zero.} yet has abstracted away complicated details of real cows that would make modeling impractical. The point is, where the abstraction works it is very valuable and the nature of the abstraction allows its validity to be quantified. For example, if you want to spin a cow, the angular momentum of a sphere and a cow are different, but a sphere's angular momentum can be used to approximate it, perhaps adequately for the purpose.

Is a model such as that used in \cite{pizzap} a ``spherical cow''?

The mathematical model is, as it is precise and abstracts away complicated details (such as cognitive issues) --- and this is very clear in \cite{pizzap}. The computational model, however, is not; the way in which it abstracted details (by implementing the mathematics) is unknown because it has bugs that undermine the validity of the abstraction, as described above.

The problem is that the refinement from the mathematical specification (equations 1--3) is tedious. Even the present paper omitted it, and referred to standard textbooks. Furthermore, it is error-prone, and considerable skill is required to assure the final computational model is correct (or correct enough).

Alternative approaches are possible, and in general are essential to ensure the quality of the science. For example Alloy \cite{alloy} is a specification language (with practical and comprehensive computer support) that allows an abstract specification to be incrementally refined. An Alloy specification can start closer to the pizza problem than the equations in \cite{pizzap} and corrected here, and it can include cognitive factors. How it is then refined is a matter for the scientist. The book \cite{alloy} also reviews other methods that may be more or less appropriate to other modeling problems.

\section{Conclusions}
This paper has shown that the Pizza Problem is a productive idea, though it is noteworthy for being simple enough to identify and address unintentional errors. Unfortunately, almost all scientific modeling is far more complex and is, one would assume, very likely to be incorrect in more subtle and less unconstructive ways.

In general, then, modeling is complex. Minor errors, such as those exhibited in the original paper \cite{pizzap}, are rarely obvious. To debug scientific thinking reliably, models need to be correct (or as correct as reasonably possible) relative to proposed specifications in the first place. We would argue that not only sharing specification and code is necessary, but a methodological step of ``assuring correctness'' is also necessary and needs to be done competently. Assuring correctness should be included in a paper's proposed methodology pathway.

\emph{Availability of code}. The Python code in this paper was extracted directly from the source \LaTeX\ by a simple \texttt{awk} script, and then run using Python version 3.9. The code is generated and run using a simple \texttt{make} file. In other words, the code presented in this paper as typeset has been checked to work as described.

\emph{Acknowledgements}
With thanks to Peter Ladkin and the authors of \cite{pizzap}, Olivia Guest and Andrea Martin, who commented on this paper and raised the spherical cow analogy.

\begin{thebibliography}{99}

\bibitem{actr}
Anderson, J. R. \emph{Rules of the Mind}, Erlbaum, 1993.

\bibitem{dop}
Dijkstra, E. W. \emph{A Discipline of Programming}, Prentice Hall, 1976.

\bibitem{pizzap}
Guest, O. \& Martin, A. E. ``How computational modeling can force theory building in Psychological Science,'' \emph{Perspectives on Psychological Science}, \textbf{16}(4):789--802, 2021. DOI 10.1177/1745691620970585

\bibitem{alloy}
Jackson, D. \emph{Software Abstractions}, MIT Press, revised edition, 2012.

\bibitem{assurance}
Knight J. C.,  Rowanhill, J. \& Rinehart, D. J. \emph{Understanding What It Means for Assurance Cases to ``Work,''}
NASA/CR–2017-219582, National Aeronautics and Space Administration
Langley Research Center, 2017.

\bibitem{tla}
Lamport, L. \emph{Specifying Systems: The TLA+ Language and Tools for Hardware and Software Engineers}, Addison-Wesley Professional, 2002.

\bibitem{z}
Spivey, J. M. \emph{Z Notation: A Reference Manual}, Prentice Hall International Series in Computer Science, 1992.

\bibitem{strachey}
Strachey, C. ``Fundamental concepts in programming languages,'' republished in \emph{Higher-Order and Symbolic Computation}, %Kluwer Academic Publishers, 
\textbf{13}(1/2):11--49, 2000. DOI 10.1023/A:1010000313106

\bibitem{heedless}
Thimbleby, H. ``Heedless programming: {I}gnoring detectable error is a widespread hazard,'' \emph{Software --- Practice {\&} Experience}, \textbf{42}(11):1393--1407, 2012.
DOI 10.1002/spe.1141

\bibitem{harold-cj}
Thimbleby, H. ``Improving science that uses code,'' \emph{Computer Journal}, in press.

\bibitem{earthlift}
Wikipedia, ``Ground Lift,''
\url{https://en.wikipedia.org/wiki/Ground_lift#cite_note-self06-1}, viewed 2023.

\end{thebibliography}

\end{document}

\section{New issues created by computer modeling}
A scientist would iteratively experiment, theorize, calculate, and write up their research. Although final papers are often written up in a structured way, there was little to constrain the process of science itself. Now, with computers, the additional steps of specification, programming, calculating (e.g., for statistics results) and executing computer models have been introduced --- along with using computers to support the writing processes.

In general, specification, programming, and running models can be done in any order, revised and iterated as corrections and new ideas are considered. However, at least for writing up, it is convenient to simplify the reality to a linear process: problem statement $\sqsupseteq$ specification $\sqsupseteq$ implementation $\sqsupseteq$ execution. The symbols $\sqsupseteq$ represent \emph{refinement}, with an abstract idea (on the left) related to its refinement into a more concrete form (on the right). Computer scientists are interested in how $\sqsupseteq$ works and how to support it with automatic tools; for instance $\sqsupseteq$  is transitive (if $u\sqsupseteq v$ and $v\sqsupseteq w$ then $u\sqsupseteq w$), but this theory and tool support is beyond the scope of this paper.

The paper \cite{pizzap} uses the word \emph{specification\/} to refer to a phase in programming, namely defining how a model should work. Interestingly, their discussion is itself a specification, a specification of psychological modeling.

Thus equation (\ref{eqn-corrected-area-generalized}) above, $\phi_i=\sum_{j=1}^{N_i} \pi R_{i,j}^2$, is more abstract than its refinement to the Python code that implements it. The Python code will necessarily introduce concrete details that are not in the specification, either because the specification permits it, Python or the programmer requires it, or because an error has been made. Thus, in the code given in \cite{pizzap}, there is nothing that correctly refines $\phi_i$ in equations (\ref{eqn-area} or \ref{eqn-corrected-area-generalized}); the parameter for \texttt{food} is the refinement of $R_i$ not $i$. Confusingly, the Python in \cite{pizzap} uses variable names (\verb|one_pizza| and \verb|two_pizzas|) to refine $R_i$, but these are names that are not values in the model --- the Python code loses the specification variable $i$. The model --- the running program --- represents two possible values of $R_i$, but it does not represent $i$ itself, and the concept of subscripting $R$ has been lost. (This paper's Python instead used a dictionary to represent pairs $(i,R_i)$ together as explicit program values.)

Given that these different refinements are legitimate, it is clear that the given specification has \emph{under-specified\/} the code; here, how an order is represented is has not been explicitly specified, and therefore it has been left open to interpretation. An easier to understand under-specification is that the programming language was not specified, leaving it open to the programmer to make any choice, and, presumably, using any programming language that can refine the equations would be acceptable.

Conversely, a specification may \emph{over-specify\/} the code. For example it might be specified that a customer's order called $x$ be represented by a variable called $x$. In \cite{pizzap}, the paper calls the orders \orderA\ and \orderB\ but the Python code happens to call them \verb|two_pizzas| and \verb|one_pizza| respectively.

### 2023-07-03 models/psych/original.py

@@ -1,0 +1,18 @@
+import numpy as np 
+import math
+
+def food(ds):
+	'''
+	Amount of food in an order as a function of the diameters per pizza (eq. 3). 
+	'''
+	return (math.pi * (ds/2)**2).sum()
+
+# Order option a in fig. 1, two 12'' pizzas:
+two_pizzas = np.array([12, 12])
+
+# Option b, one 18'' pizza: 
+one_pizza = np.array([18])
+
+# Decision rule (eq. 2): 
+print(food(two_pizzas) > food(one_pizza))
+

### 2023-07-03 models/psych/pizza.py

@@ -1,0 +1,50 @@
+# (c) Harold Thimbleby, harold@thimbleby.net, 28 June 2023 @ 14:54:35 
+print("Running the Python code extracted by awk script from pizza.tex")
+
+
+import math
+
+pi = math.pi
+
+# A map of Pizza orders
+# Variable orders[] maps order ID to a list of pizza diameters in inches\ 
+orders = {
+   "a": [12, 12], # Order a is two 12 inch pizzas
+   "b": [18] # Order b is one 18 inch pizza
+}
+
+# Implement $\phi$ from (\refeqn-corrected-area-generalized)
+# Tota area of all pizzas in orderID
+def phi(orderID):
+   totalArea = 0
+   for diameter in orders[orderID]:
+      totalArea += pi * (diameter/2)**2
+   return totalArea
+
+# Implement $\omega$ from equation (\refeqn-set-corrected-specification) 
+# Determine maximum area order(s) over all orders
+maxArea = 0
+maxAreaIDs = []
+for orderID in orders :
+   area = phi(orderID)
+   if maxArea == area : 
+      maxAreaIDs.append(orderID)
+   if maxArea < area  :
+      maxArea = area
+      maxAreaIDs = [orderID]
+
+print(maxAreaIDs)
+
+   
+print("Pizza order", maxAreaIDs, 
+      "has largest area =", round(maxArea, 1), "sq in"
+     )
+
+N = len(maxAreaIDs)
+if N == 0 : 
+   print("No pizza orders were made")
+else :
+       print("Pizza", "orders" if N > 1 else "order", 
+	        maxAreaIDs, "has" if N == 1 else "have equal",
+	        "largest area =", round(maxArea, 1), "sq in"
+	        )

### 2023-07-03 models/psych/original.py
Removed psych folder

@@ -1,18 +1,0 @@
-import numpy as np 
-import math
-
-def food(ds):
-	'''
-	Amount of food in an order as a function of the diameters per pizza (eq. 3). 
-	'''
-	return (math.pi * (ds/2)**2).sum()
-
-# Order option a in fig. 1, two 12'' pizzas:
-two_pizzas = np.array([12, 12])
-
-# Option b, one 18'' pizza: 
-one_pizza = np.array([18])
-
-# Decision rule (eq. 2): 
-print(food(two_pizzas) > food(one_pizza))
-

### 2023-07-03 models/psych/pizza.py

@@ -1,50 +1,0 @@
-# (c) Harold Thimbleby, harold@thimbleby.net, 28 June 2023 @ 14:54:35 
-print("Running the Python code extracted by awk script from pizza.tex")
-
-
-import math
-
-pi = math.pi
-
-# A map of Pizza orders
-# Variable orders[] maps order ID to a list of pizza diameters in inches\ 
-orders = {
-   "a": [12, 12], # Order a is two 12 inch pizzas
-   "b": [18] # Order b is one 18 inch pizza
-}
-
-# Implement $\phi$ from (\refeqn-corrected-area-generalized)
-# Tota area of all pizzas in orderID
-def phi(orderID):
-   totalArea = 0
-   for diameter in orders[orderID]:
-      totalArea += pi * (diameter/2)**2
-   return totalArea
-
-# Implement $\omega$ from equation (\refeqn-set-corrected-specification) 
-# Determine maximum area order(s) over all orders
-maxArea = 0
-maxAreaIDs = []
-for orderID in orders :
-   area = phi(orderID)
-   if maxArea == area : 
-      maxAreaIDs.append(orderID)
-   if maxArea < area  :
-      maxArea = area
-      maxAreaIDs = [orderID]
-
-print(maxAreaIDs)
-
-   
-print("Pizza order", maxAreaIDs, 
-      "has largest area =", round(maxArea, 1), "sq in"
-     )
-
-N = len(maxAreaIDs)
-if N == 0 : 
-   print("No pizza orders were made")
-else :
-       print("Pizza", "orders" if N > 1 else "order", 
-	        maxAreaIDs, "has" if N == 1 else "have equal",
-	        "largest area =", round(maxArea, 1), "sq in"
-	        )
